---
title: "Hyper-Reflectivity and Neovascularization in LMRI DB"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
  always_allow_html: true
  html_document:
    df_print: paged
---

```{r SETUP,echo = FALSE}
suppressWarnings(
pacman::p_load(HDF5Array, SingleCellExperiment, dplyr, stringr,tidyr, tibble, ggplot2, knitr, kableExtra, htmltools, gt, webshot, webshot2)
)
suppressMessages(
rm(list = ls())
)

invisible(
gc(verbose = FALSE)
)
```


```{r LOAD DATA & FUNCTIONS,echo = FALSE }
wd_path = "/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/analysis_versions/version013/"

# load experiment
sce <-  loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/sce006.h5"))

# my functions
source(sprintf("%sanalysis/00_MyFunctions.R", wd_path))
```


```{r make data ,echo = FALSE}

# WHAT EYES EVER HAVE NEOVASC OR HR?

d <- data.frame(t(assays(sce)$measure_vars)[, c("NEOVASC", "HR_TEMPORAL")])

d = cbind(subject = rownames(d), d)

suppressMessages(
  d <- left_join(d, select(data.frame(colData(sce)), subject,patID, patID_eye, month))
)


suppressWarnings(
  # do eyes ever develop NEOVASC or HR_TEMPROAL? 
  eye_maxes <- d %>% 
    group_by(patID_eye) %>% 
    summarize(across(c(NEOVASC, HR_TEMPORAL), ~max(., na.rm = TRUE)))
  # ^ error okay, makes only missing eyes -Inf
)

suppressWarnings(
  # do patients ever develop NEOVASC or HR_TEMPROAL? 
  pat_maxes <- d %>% 
    group_by(patID) %>% 
    summarize(across(c(NEOVASC, HR_TEMPORAL), ~max(., na.rm = TRUE)))
  # ^ error okay, makes only missing eyes -Inf
)


suppressWarnings(
  # eyes that develop either
  eye_develop <- eye_maxes %>% 
    mutate_all(~ifelse(is.infinite(.), NA, .)) %>% 
    select(NEOVASC, HR_TEMPORAL) %>% 
    table(useNA = "no")
)

suppressWarnings(
  # pats that develop either
  pat_develop <- pat_maxes %>% 
    mutate_all(~ifelse(is.infinite(.), NA, .)) %>% 
    select(NEOVASC, HR_TEMPORAL) %>% 
    table(useNA = "no")
)
```

```{r,echo = FALSE}

# # Create a narrower table
# kable(eye_develop, format = "html", table.attr = "class='table table-condensed'") %>%
#   kable_styling(full_width = FALSE) %>%
#   column_spec(1:ncol(eye_develop), width = "5em")

```

```{r Chi Square test, echo = FALSE}
eye_chi_sq_result <- chisq.test(eye_develop)
pat_chi_sq_result <- chisq.test(pat_develop)
```

```{r Format Table,echo = FALSE}

# calculate percents
eye_develop_percents <- round(eye_develop/sum(eye_develop), 3)
pat_develop_percents <- round(pat_develop/sum(pat_develop), 3)


# format df for gt
nms <- c("0", "1", "2")
make_gt_format <- function(matrix){
  df <- as.data.frame.matrix(matrix)
  rownames(df) <- nms
  colnames(df) <- nms
  df$rownames <- rownames(df)
  df$group = "NEOVASC"
  return(df)
}

eye_develop_df <- make_gt_format(eye_develop)
eye_develop_percents_df <- make_gt_format(eye_develop_percents)
pat_develop_df <- make_gt_format(pat_develop)
pat_develop_percents_df <- make_gt_format(pat_develop_percents)


table_width = 600
# make counts table
eye_counts_tbl <- gt(eye_develop_df, rowname_col = "rownames", row_group_as_column = TRUE, groupname_col  = "group") |>
  tab_header("Eyes that develop NEOVASC and/or HR_TEMPORAL", subtitle = "0: not Present; 1: Questionable; 2: Present")|>
  tab_spanner(columns =nms, label = "HR_TEMPORAL")|>
  tab_source_note(
    source_note = sprintf("X^2 test: p ~ %s; N = %s" ,formatC(eye_chi_sq_result$p.value), sum(eye_develop))
  )|>
  tab_source_note(
    source_note = "Note: These are counts of unique eyes that ever develop phenotype, not counts of eye-visits"
  )|> 
  tab_options(table.width = px(table_width))

# make pat counts table
pat_counts_tbl <- gt(pat_develop_df, rowname_col = "rownames", row_group_as_column = TRUE, groupname_col  = "group") |>
  tab_header("Patients that develop NEOVASC and/or HR_TEMPORAL", subtitle = "0: not Present; 1: Questionable; 2: Present")|>
  tab_spanner(columns =nms, label = "HR_TEMPORAL")|>
  tab_source_note(
    source_note = sprintf("X^2 test: p ~ %s; N = %s" ,formatC(pat_chi_sq_result$p.value), sum(pat_develop))
  )|>
  tab_source_note(
    source_note = "Note: These are counts of unique patients that ever develop phenotype, not counts of patient-visits"
  )|> 
  tab_options(table.width = px(table_width))

# make eye percents table
eye_percents_tbl <- gt(eye_develop_percents_df, rowname_col = "rownames", row_group_as_column = TRUE, groupname_col  = "group") |>
  tab_header("Ratio of eyes that develop NEOVASC and/or HR_TEMPORAL", subtitle = "0: not Present; 1: Questionable; 2: Present")|>
  tab_spanner(columns =nms, label = "HR_TEMPORAL")|>
  tab_source_note(
    source_note = sprintf("Ratio develop both: %f (1,1,+1,2+2,2+2,1)", sum(eye_develop_percents_df[2:3, 2:3]))
  )|>
  tab_source_note(
    source_note = sprintf("Ratio develop HR_TEMPORAL only: %f (0,1+0,2)", signif(sum(eye_develop_percents_df[1, 2:3])))
  )|>
  tab_source_note(
    source_note = sprintf("Ratio develop NEOVASC only: %f (1,0+2,0)", signif(sum(eye_develop_percents_df[2:3, 1])))
  )|> 
  tab_options(table.width = px(table_width))


# make patient percents table
pat_percents_tbl <- gt(pat_develop_percents_df, rowname_col = "rownames", row_group_as_column = TRUE, groupname_col  = "group") |>
  tab_header("Ratio of patients that develop NEOVASC and/or HR_TEMPORAL", subtitle = "0: not Present; 1: Questionable; 2: Present")|>
  tab_spanner(columns =nms, label = "HR_TEMPORAL")|>
  tab_source_note(
    source_note = sprintf("Ratio develop both: %f (1,1,+1,2+2,2+2,1)", sum(pat_develop_percents_df[2:3, 2:3]))
  )|>
  tab_source_note(
    source_note = sprintf("Ratio develop HR_TEMPORAL only: %f (0,1+0,2)", signif(sum(pat_develop_percents_df[1, 2:3])))
  )|>
  tab_source_note(
    source_note = sprintf("Ratio develop NEOVASC only: %f (1,0+2,0)", signif(sum(pat_develop_percents_df[2:3, 1])))
  )|> 
  tab_options(table.width = px(table_width))


# gtsave(eye_counts_tbl ,sprintf("%s/results/HR_NEOVASC_%s.pdf",wd_path,"eye_counts_tbl"))
# gtsave(eye_percents_tbl ,sprintf("%s/results/HR_NEOVASC_%s.pdf",wd_path,"eye_percents_tbl"))
# gtsave(pat_counts_tbl ,sprintf("%s/results/HR_NEOVASC_%s.pdf",wd_path,"pat_counts_tbl"))
# gtsave(pat_percents_tbl ,sprintf("%s/results/HR_NEOVASC_%s.pdf",wd_path,"pat_percents_tbl"))

```


```{r Percents, echo=FALSE}
eye_percent_both <- sum(eye_develop_percents_df[2:3, 2:3])*100
eye_percent_HR_only <- sum(eye_develop_percents_df[1, 2:3])*100
eye_percent_NEOVASC_only <- sum(eye_develop_percents_df[2:3, 1])*100

pat_percent_both <- sum(pat_develop_percents_df[2:3, 2:3])*100
pat_percent_HR_only <- sum(pat_develop_percents_df[1, 2:3])*100
pat_percent_NEOVASC_only <- sum(pat_develop_percents_df[2:3, 1])*100

```

### Q: What percent of MacTel patients have outer retina hyper reflectivity (ORHR) and choroidal neovascularization (CNV)?
ORHR and CNV are not variables in the database, but HR_TEMPORAL (Hyper reflectivity on the temporal side) and NEOVASC (Neovascularization) are. 

__`r eye_percent_both+eye_percent_HR_only`__ % of eyes develop HR_TEMPORAL and  __`r eye_percent_NEOVASC_only+eye_percent_both`%__ develop NEOVASC while only __`r eye_percent_both`%__ develop both. 

__`r pat_percent_both+pat_percent_HR_only`__ % of patients develop HR_TEMPORAL and  __`r pat_percent_NEOVASC_only+pat_percent_both`%__ develop NEOVASC while only __`r pat_percent_both`%__ develop both. 

These percentages are found in Table 2 and 4, respectively.

If you are interested in seeing a breakdown of phenotype presence on the level of patients and eyes, please reference the tables below. 

#### Tables
 
```{r, echo = FALSE}
eye_counts_tbl
```

__Table 1__: Counts of eyes for every combination of HR_TEMPORAL and NEOVASC.  A chi-square test was performed on these values to test whether the presence of Neovascularization on Hyper-reflectivity and vice-versa. 
 
-----------------------------------

-----------------------------------
 
```{r, echo = FALSE}
eye_percents_tbl
```

__Table 2__: Every combination of HR_TEMPORAL and NEOVASC as a ratio of total eyes ('N' in Table 1). The percentages of eyes that develop both and each individually were calculated by summing their respective cells in the table. 
 
-----------------------------------

-----------------------------------
 
```{r, echo = FALSE}
pat_counts_tbl
```

__Table 3__: Counts of patients for every combination of HR_TEMPORAL and NEOVASC. A chi-square test was performed on these values to test whether the presence of Neovascularization on Hyper-reflectivity and vice-versa.
 
-----------------------------------

-----------------------------------
 
```{r, echo = FALSE}
pat_percents_tbl
```

__Table 4__: Every combination of HR_TEMPORAL and NEOVASC as a ratio of total patients ('N' in Table 3). The percentages of patients that develop both and each individually were calculated by summing their respective cells in the table. 

