---
title: "01 database data cleaning"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---


# Packages

```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}
pacman::p_load(tidyr,dplyr,ggplot2,reshape2,purrr,tidyverse,lubridate,knitr,openxlsx,ComplexHeatmap,zoo,fuzzyjoin,data.table,patchwork,janitor, parallel)
rm(list = ls())
gc()


```

### options
```{r}

options(max.print = 300)

```

# Load Data

    these tables were downloaded from the https://qaresearch.lmri.net/#/login?redirectTo=%2Fsearch database on 24 FEB 2025. the VAA table comes from Emmes directly because they overwrote it in the past. VAA.csv only has NHOS while VAA-NHOR-NHOS concat.xlsx has NHOS and NHOR 
    
```{r}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

# my functions
source(sprintf("%sanalysis/00_MyFunctions.R", wd_path))

# get files in directory and extract csvs
parent_path <- "/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/"
dir_path <- "raw_data/06" 
raw_dat_path <- paste0(parent_path, dir_path)

# list of filenames
files <- dir(raw_dat_path)
csv_files <- files[grep("\\.csv$", files)] 

# load data into list
raw_dfs <- lapply(paste0(raw_dat_path,'/',csv_files), readr::read_csv)

# name items in list 
names(raw_dfs) <- lapply(csv_files, function(x) substring(x,1, nchar(x) - 4) )

# add other VAA table with NHOS and NHOR
VAA_lea <- read.xlsx(paste0(raw_dat_path,'/',files[grepl("\\.xlsx$",files)]))
raw_dfs$VAA_lea <- as_tibble(VAA_lea)

# add previous version VAA 
VAA_old <- read_csv("/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/raw_data/05_imageGradingAndVisualAcuity/consolidated_sheet_15_4_2024/VAA.csv")
raw_dfs$VAA_old <- as_tibble(VAA_old)

# only underscore in names
names(raw_dfs) <- gsub("-", "_", names(raw_dfs))

dfs <- raw_dfs

# # read in info
# info <- read.xlsx(path)

```

# make directories
```{r}

path <- paste0(wd_path, "results/VAA_visitID_fixing/")
if(!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
  
}
```

#__________________

# clean dfs
## clean chars from dfs in list

```{r}

# lapply(dfs, function(df){
#   glimpse(df)
# })


# remove quotes and equals
remove_chars <- function(dataframe) {
  fixed_dataframe <- dataframe %>%  
    mutate(across(everything(), ~gsub("[=,\"=,]", "", .))) %>% 
    mutate(across(everything(), ~trimws(.))) 
    
  names(fixed_dataframe) <- names(fixed_dataframe)  %>% 
    str_remove_all(.,'^.*?"|".*$')
  
  return(fixed_dataframe)
}

dfs <- lapply(dfs, remove_chars)

```
clean the characters off of values and names from each dataframe


# check df dimensions
```{r}

dimens <- lapply(dfs, dim)
row_n <- list()

row_n <- lapply(1:length(dimens), function(i){
  
 dimens[[i]][1]
  
}) %>% 
  unlist()

if(!all(row_n[1:(length(row_n)-1)] == row_n[2:length(row_n)])){ cat("!!!!!!!!!!!!!!!\nWarning: non-matching dimensions\n!!!!!!!!!!!!!!!\n")}
dimens

```
_______old_________

$CF_FFA
[1] 30389    98

$DEM
[1] 30389    13

$DIAGNOSIS
[1] 30389    10

$ENR
[1] 30389    41

$FAFGRADE
[1] 30389    19

$FAMSTATUS
[1] 30389     9

$MH
[1] 30389    63

$OCTMTCIRRUS
[1] 30389    19

$OCTMTSPECTRALIS
[1] 30389    33

$OCTMTSTRATUS
[1] 30389    21

$VAA
[1] 30389    74

_____new_____
$CF_FFA
[1] 31357    98

$DEM
[1] 1173    7

$DIAGNOSIS
[1] 31357    10

$ENRNHOR
[1] 31357    28

$ENRNHOS
[1] 31357    30

$FAFGRADE
[1] 31357    21

$FAMSTATUS
[1] 31357    11

$MH
[1] 31357    56

$OCTMAIN
[1] 31357    30

$OCTMTCIRRUS
[1] 31357    22

$OCTMTSPECTRALIS
[1] 31357    32

$OCTMTSTRATUS
[1] 31357    24

$VAA_emmes
[1] 2668   70

$VAA_lea
[1] 7858   70

$VAA_old
[1] 30389    74




# further cleaning 
## fix names
```{r}

dfs <- map(dfs, ~{
  names(.) <- str_replace_all(names(.), "-", "_")
  return(.)
})

```

## fix spectralis names
```{r}

# rename MT_SD08 to MT08_SD for spectralis names
patterns <- c(paste0("MT_SD0", 1:9))
replacements <- c(paste0(paste0("MT0", 1:9), "_SD"))
pattern_replacement <- data.frame(pattern = patterns, replacement = replacements)

#
# i = 6
#

# for each pattern and replacement, search pattern in names and replace
for (i in 1:(nrow(pattern_replacement))){
  names(dfs$OCTMTSPECTRALIS) <- gsub(pattern_replacement$pattern[i],pattern_replacement$replacement[i], names(dfs$OCTMTSPECTRALIS))
}

names(dfs$OCTMTSPECTRALIS)
```

## make MT, VOL values unique in "OCTMTCIRRUS" "OCTMTSTRATUS" "OCTMTSPECTRALIS"

```{r}
# what mt names are repeated
# mt.names <- lapply(dfs[grep("OCTMT", names(dfs))], function(df)  names(df)[order(names(df))] %>% substr(., 1, 3)) %>% unlist()
# names(table(mt.names)[table(mt.names)>1])
# "CEN" "cre" "eye" "MT0" "pat" "SCA" "SIG" "vis" "VOL"


# at locations with name containing MT0 and VOL, change names to the same with append pasted to end 
append <- "cirrus_"
append_tos <- c("MT0", "VOL", "CENTER", "CUBE")

for(an_append_to in append_tos){
  
  names(dfs$OCTMTCIRRUS)[grepl(an_append_to, names(dfs$OCTMTCIRRUS))] <- paste0(append,names(dfs$OCTMTCIRRUS)[grepl(an_append_to, names(dfs$OCTMTCIRRUS))])
  
}

append <- "stratus_"
append_tos <- c("MT0", "VOL", "CENTER", "SCANDT")

for(an_append_to in append_tos){
  
  names(dfs$OCTMTSTRATUS)[grepl(an_append_to, names(dfs$OCTMTSTRATUS))] <- paste0(append,names(dfs$OCTMTSTRATUS)[grepl(an_append_to, names(dfs$OCTMTSTRATUS))])
  
}


append <- "spectralis_"
append_tos <- c("MT0\\d$", "VOL", "_SD$", "SCANDT")

for(an_append_to in append_tos){
  
  names(dfs$OCTMTSPECTRALIS)[grepl(an_append_to, names(dfs$OCTMTSPECTRALIS))] <- paste0(append,names(dfs$OCTMTSPECTRALIS)[grepl(an_append_to, names(dfs$OCTMTSPECTRALIS))])
  
}

names(dfs$OCTMTCIRRUS)
names(dfs$OCTMTSTRATUS)
names(dfs$OCTMTSPECTRALIS)

```
MT fields made unique for join later on. they have the same names in the octmt datasets


## make grading date fields unique
what dataframes have GRADINGDT variable. Make "GRADINGDT" = "<df_name>_GRADINGDT"
 
```{r}

for(i in 1:length(dfs)){
  df <- dfs[[i]]
  df_name <- names(dfs)[i]
  
  cat(df_name,"{ ",  grep("DT|Date", names(df)  , value = TRUE, ignore.case = TRUE), "\n")

}

# set the names(df) == GRADINGDT => new name
for(i in 1:length(dfs)){
  df <- dfs[[i]]
  df_name <- names(dfs)[i]
  
  names(dfs[[i]])[grep("GRADINGDT", names(dfs[[i]]))] <- paste(df_name, "GRADINGDT", sep = "_")
  names(dfs[[i]])[grep("^DATE$", names(dfs[[i]]))] <- paste(df_name, "DATE", sep = "_")
}

```


## remove createdAt fields 
```{r}
lapply(dfs, function(x){
 x[["createdAt"]] %>% 
    substr(., 1, 10) %>% 
    unique() 
})

# All same so remove
dfs <- lapply(dfs, function(a_df){
      select(a_df, -matches("createdAt"))
})

```
> createdAt field is the date the data was uploaded to the db 

## fix no name columns if any
```{r}
lapply(dfs, function(a.df){
  # length(names(a.df) ) == ncol(a.df)
  which(nchar(names(a.df)) ==0)
})

dfs <- lapply(dfs, function(a.df){
  names(a.df)[which(nchar(names(a.df)) ==0)] <- "NONAME"
  return(a.df)
})

```

 
## fix null values; make visitID, patientId and eye concatenation
*duplicate patient ids because of two eye records and 1\< visits*
```{r}

# null values
nas <- c( "null", "NULL", "Null", "na" , "Na","", " ", "\t", "\n",
          "7","777","7777","8", "888","8888","9", "999", "9999", "removed",
          "  ")

## test for image. not making triple scores NA
# nas <- c( "null", "NULL", "Null", "na" , "Na","", " ", "\t", "  ", "removed")

## change nas to NA
dfs <- lapply(dfs, function(df){
          
  fixed_df <- df %>% 
  mutate_all(~ifelse( . %in% nas, NA, . )) 
          
  return(fixed_df)
}
)

```
changed the grading scores 6, 7, 8, 9 to NA, these score image quality. 

#### look at how fixing nulls changes nas in spectralis
```{r, eval = FALSE, include=FALSE}

dfs$OCTMTSPECTRALIS %>% 
  mutate_all(~as.numeric(!is.na(.))) %>% 
  colSums() %>% 
  t() %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("var") %>% 
  rename_with(~all_of(c("var", "count")) ) %>% 
  ggplot(aes(y = var, x= count))+
  geom_bar(stat = "identity")+
  ggtitle("Count non-missing")

```
> not treating 7... 8... 9..., removed as NA ::> all zones of spectralis have same count !NA
> zones 6:9 have worse image quality for some reason

# LOOK at number of patients, observations, etc in each df (raw data until here(inclusive))
```{r, eval = TRUE}


# function to print observation counts from dfs list of db tables
obs_m <- function(list_of_dfs= dfs){
  
  sapply(list_of_dfs, function(DF){

    both <- all(c("patientId", "PATID")%in% names(DF))
    
    patID_grep_str <- ifelse(both, "patientId", "patientId|PATID")
    
    n_patIDs <- DF %>% 
      select(any_of(c("patientId","eye","visitId", "PATID")) )%>% 
      drop_na() %>% 
      pull(any_of(matches(patID_grep_str))) %>% 
      n_distinct(na.rm = TRUE)
  
  
    n_observations <- DF %>% 
      select(any_of(c("patientId","eye","visitId", "PATID")) )%>%
      drop_na() %>%
      n_distinct(na.rm = TRUE)
    
    
    if("eye" %in% names(DF)){
      n_patID_eyes <- DF %>% 
        select(any_of(c("patientId","eye","visitId", "PATID"))) %>%
        drop_na() %>%
        select(matches(patID_grep_str), eye) %>% 
        n_distinct(na.rm = TRUE)
        
      n_patIDs_w_gr_1_vis <- DF %>% 
        select(any_of(c("patientId","eye","visitId", "PATID"))) %>%
        drop_na() %>% 
        filter(n()>1, .by = c(matches(patID_grep_str), eye)) %>% 
        pull(matches(patID_grep_str)) %>% 
        n_distinct(na.rm = TRUE)
      return(list(
        n_patIDs=n_patIDs,
        n_patID_eyes=n_patID_eyes,
        n_observations=n_observations,
        n_patIDs_w_gr_1_vis=n_patIDs_w_gr_1_vis
      ))
    }else{
      return(list(
        n_patIDs=n_patIDs,
        n_patID_eyes=NULL,
        n_observations=n_observations,
        n_patIDs_w_gr_1_vis=NULL
      ))
    }
      
      
  }) 
}

obs_m()
# ^                    CF_FFA DEM   DIAGNOSIS ENR   FAFGRADE FAMSTATUS MH    OCTMAIN OCTMTCIRRUS OCTMTSPECTRALIS OCTMTSTRATUS VAA  
# n_patIDs            4951   4951  4951      4951  4951     4951      4951  4951    4951        4951            4951         4951 
# n_patID_eyes        9902   9902  9902      9902  9902     9902      9902  9902    9902        9902            9902         9902 
# n_observations      15481  15481 15481     15481 15481    15481     15481 15481   15481       15481           15481        15481
# n_patIDs_w_gr_1_vis 647    647   647       647   647      647       647   647     647         647             647          647  

### 25 FEB 2025 ###
#                     CF_FFA DEM  DIAGNOSIS ENRNHOR ENRNHOS FAFGRADE FAMSTATUS MH    OCTMAIN OCTMTCIRRUS OCTMTSPECTRALIS OCTMTSTRATUS VAA 
# n_patIDs            5289   1158 5289      5289    5289    5289     5289      5289  5289    5289        5289            5289         1151
# n_patID_eyes        10578  NULL 10578     10578   10578   10578    10578     10578 10578   10578       10578           10578        NULL
# n_observations      15968  1158 15968     15968   15968   15968    15968     15968 15968   15968       15968           15968        1151
# n_patIDs_w_gr_1_vis 552    NULL 552       552     552     552      552       552   552     552         552             552          NULL

### 3 MAR 2025 ###
#                     CF_FFA DEM  DIAGNOSIS ENRNHOR ENRNHOS FAFGRADE FAMSTATUS MH    OCTMAIN OCTMTCIRRUS OCTMTSPECTRALIS OCTMTSTRATUS VAA  VAA_lea VAA_old
# n_patIDs            5289   1158 5289      5289    5289    5289     5289      5289  5289    5289        5289            5289         1151 5539    4809   
# n_patID_eyes        10578  NULL 10578     10578   10578   10578    10578     10578 10578   10578       10578           10578        NULL NULL    9615   
# n_observations      15968  1158 15968     15968   15968   15968    15968     15968 15968   15968       15968           15968        1151 5539    14215  
# n_patIDs_w_gr_1_vis 552    NULL 552       552     552     552      552       552   552     552         552             552          NULL NULL    513 

# function to print dimensions of dfs
dfs_dims <- function(DFs = dfs){
  for(i in 1:length(DFs)){
     df <- DFs[[i]]
     df_name = names(DFs)[i]
     cat(paste0(df_name, ": ",paste(dim(df), collapse = " "), "\n"))
  }
  
}

# function to chech if all values in each col of obs_m equal
col_equal <- function(m){
  for (i in 1:(ncol(m)-1) ) all( as.numeric(m[, i])==as.numeric(m[, i+1])) %>% print
}

dfs_dims()
# CF_FFA: 30389,97
# DEM: 30389,12
# DIAGNOSIS: 30389,9
# ENR: 30389,40
# FAFGRADE: 30389,18
# FAMSTATUS: 30389,8
# MH: 30389,62
# OCTMAIN: 30389,29
# OCTMTCIRRUS: 30389,18
# OCTMTSPECTRALIS: 30389,32
# OCTMTSTRATUS: 30389,20
# VAA: 30389,73

### 25 FEB 2025 ###
# CF_FFA: 31357 97
# DEM: 1173 7
# DIAGNOSIS: 31357 9
# ENRNHOR: 31357 27
# ENRNHOS: 31357 29
# FAFGRADE: 31357 20
# FAMSTATUS: 31357 10
# MH: 31357 55
# OCTMAIN: 31357 29
# OCTMTCIRRUS: 31357 21
# OCTMTSPECTRALIS: 31357 31
# OCTMTSTRATUS: 31357 23
# VAA: 2668 70

### 3 MAR 2025 ###
# CF_FFA: 31357 97
# DEM: 1173 7
# DIAGNOSIS: 31357 9
# ENRNHOR: 31357 27
# ENRNHOS: 31357 29
# FAFGRADE: 31357 20
# FAMSTATUS: 31357 10
# MH: 31357 55
# OCTMAIN: 31357 29
# OCTMTCIRRUS: 31357 21
# OCTMTSPECTRALIS: 31357 31
# OCTMTSTRATUS: 31357 23
# VAA: 2668 70
# VAA_lea: 7858 70
# VAA_old: 30389 73

# Look at one case
pat <- "001001"

lapply(dfs[grep("VAA", names(dfs))], function(df){
  df <- as.data.frame(df)

  if("patientId"%in%names(df)){
    df %>%
      filter(patientId == pat) %>%
      select(patientId,visitId, eye) %>%
      mutate(eye_obs = !is.na(eye)) %>% 
      select(-eye) %>% 
      arrange(patientId,visitId) %>% 
      distinct()
  }else if("VISNO"%in% names(df)){
      df %>%
        filter(PATID == pat) %>%
        select(PATID, VISNO) %>%
        arrange(PATID, VISNO)
  }else{
      df %>%
        filter(PATID == pat) %>%
        select(PATID)
  }
  
})


```
> VAA does not have as many visits for 001001 as other tables, only NHOR data??
>> likely because only some visits took VA data AND because dimensions match due to db structure that harmonizes

# ____

# VAA Mung
## fix VAA_lea date
```{r}

# Excel dates start at 1900-01-01, but due to an Excel leap year bug, R needs origin = "1899-12-30".
VAA_date_origin <-  "1899-12-30"

# fix excel date numeric 
dfs$VAA_lea$DATEVIS <- as.Date(as.numeric(dfs$VAA_lea$DATEVIS), origin = VAA_date_origin)

```

## make names visitId and patientId
```{r}

# if not there, make from visno and patid
dfs <- lapply(dfs, function(df){
 if(!"patientId"%in% names(df) & !"visitId"%in% names(df))
  names(df)[grep("PATID", names(df))] <- "patientId"
  names(df)[grep("VISNO", names(df))] <- "visitId"
 return(df)
})

# clean spaces in visitId
dfs$VAA_lea <- dfs$VAA_lea %>% 
  mutate(visitId = str_remove_all(visitId, " "))

```


## Lengthen VAA dfs
make each row a unique pat, vis, eye observation in VAA sheets
```{r}

VAA_df <- dfs$VAA_lea

# ****** FUNCTION: lengthen vaa df so that rows not pat vis, but pat, vis, eye unique ******
lengthen_vaa <- function(VAA_df){
  
  # rename VAA r/l cols to something common
  names(VAA_df) <- str_replace_all(names(VAA_df), "OD$|1AR$|AR$", "OD") %>% 
    str_replace_all(., "OS$|2AL$|AL$", "OS") 
  
  # pivot longer on od/os keying
  VAA_df_longer <- VAA_df %>%
    pivot_longer(
      # right/left columns 
      cols = matches("OD$|OS$"), 
      # base name remains as column name and values of 'eye' becomes od/os suffix
      names_to = c(".value", "eye_derived"),
      names_pattern = "(.*)(OD|OS)$"
    ) %>% 
    mutate(
      eye_derived = case_when(
        eye_derived == "OD" ~ 1,
        eye_derived == "OS" ~ 2,
        TRUE ~ NA_real_
      )
    )
  # make derived_eye >> eye if not already there
  if(!"eye"%in%names(VAA_df_longer)){
    VAA_df_longer$eye <- VAA_df_longer$eye_derived
    VAA_df_longer$eye_derived <- NULL
  }
  return(VAA_df_longer)
}
  
# lengthen each vaa table
dfs[grepl("VAA", names(dfs))] <- lapply(dfs[grepl("VAA", names(dfs))], lengthen_vaa)

dfs[grepl("VAA", names(dfs))] %>% 
  map(dim)
# $VAA_emmes
# [1] 5336   40
# 
# $VAA_lea
# [1] 15716    40
# 
# $VAA_old
# [1] 60778    43

dfs$VAA_lea$eye %>% table(useNA = 'always')

```



#### Compare VAA to db VAA
```{r}

# look at single case in all dfs
dfs[grepl("VAA", names(dfs))] %>% 
  lapply(., function(df){
    df %>% 
      # filter(patientId == "001001") %>% 
      filter(patientId == "017006") %>% 
      select(patientId, eye, visitId , any_of("_TCVISIT9"),matches("RW024|LOGMAR|DATE")) %>% 
      # arrange(get(ifelse("VISNO" %in% names(df), "VISNO", "visitId"))) %>% 
      arrange(get(grep("DATE", names(.), value = TRUE))) %>% 
      as.data.frame()
})
# ^ VAA_lea visitId restarts at 000 for new study, datevis is correct tho


#+ VAA_old is version used in version013; VAA_lea is concatenated NHOR/OS VAA table from Lea; VAA_emmes is the version sent from emmes(only NHOS)
dstnct_db_VAA <- distinct(select(dfs$VAA_old, patientId, visitId, eye) %>% drop_na())
dstnct_VAA <- distinct(select(dfs$VAA_lea, patientId, visitId, eye)%>% drop_na())

nrow(dstnct_db_VAA)
# 15481
nrow(dstnct_VAA)
# 15626>>14068

# COMPARE dfs$VAA_lea TO GRADING DATA----

x_stratus <- dfs$OCTMTSTRATUS %>% 
  filter(!visitId%in% c("VISITLESS")) %>% 
  filter(!is.na(eye) & !is.na(stratus_MT01)) %>% 
  mutate(subject = paste(patientId, visitId, eye, sep = "_"))

x_VAA <- dfs$VAA_lea %>% 
  filter(!visitId%in% c("VISITLESS")) %>% 
  filter(!is.na(eye) & !is.na(RW024)) %>% 
  mutate(subject = paste(patientId, visitId, eye, sep = "_"))
  # mutate(subject = paste(PATID, `_TCVISIT9`, eye, sep = "_"))

png(sprintf("%sresults/db_vs_new_VAA_subjects_venn.png", wd_path), height = 1000, width = 1000)
grid.newpage()
grid.draw(recording = FALSE,  VennDiagram::venn.diagram(list(VAA = x_VAA$subject, OCT_STRATUS = x_stratus$subject), 
                                    filename = NULL, 
                                    main = "Subjects in Lea VAA versus DB stratus grading data",
                                    disable.logging = T, 
                                    cex = 2, 
                                    main.cex = 2,
                                    sub.cex = 2
                            ))
dev.off()

# does VAA_lea contain all of vaa_emmes?
all(
  with(dfs$VAA_emmes, paste(patientId, eye, DATEVIS, sep = "_"))
  %in%
  with(dfs$VAA_lea, paste(patientId, eye, DATEVIS, sep = "_"))
)# ^ TRUE

# remove VAA_emmes and old because redundant and wrong
dfs$VAA_emmes <- NULL
dfs$VAA_old <- NULL

```

## Check how visitId compares to _TCVISIT9
#### look at distributions
```{r}

# look at distributions
dfs$VAA_lea %>% 
  select(matches("visit")) %>%
  map(~{
  x <- as.numeric(.)
  summary(x)
  })
# ^ ~tcvisit9 has negatives???

# fix negative _TCVISIT9s
dfs$VAA_lea %>%
  select(patientId, eye, visitId, matches("visit|DATE")) %>%
  arrange(patientId, eye, DATEVIS) %>% 
  group_by(patientId, eye) %>% 
  # filter(any(sum(TCVISIT9<0 | `_TCVISIT9`<0)==1)) %>%
  filter(any(`_TCVISIT9`<0)   ) %>%
  # View()
  ungroup()
# ^ only 40 observations with neg _TCV...
#+ when it's negative, other tcv... tends to be correct so using that

# make tcvisits numeric
dfs$VAA_lea <- dfs$VAA_lea %>% 
  mutate(across(matches("TCVIS"), ~as.numeric(.)))

# where _TCVISIT9 neg, set to other if it not neg, if both neg, set NA
whr <- dfs$VAA_lea$`_TCVISIT9`<0 & 
  dfs$VAA_lea$TCVISIT9 >=0 &
  !is.na(dfs$VAA_lea$`_TCVISIT9`) &
  !is.na(dfs$VAA_lea$TCVISIT9)
dfs$VAA_lea$`_TCVISIT9`[whr] <- dfs$VAA_lea$TCVISIT9[whr]

# now only where both negative remain
dfs$VAA_lea$`_TCVISIT9`[dfs$VAA_lea$`_TCVISIT9`<0] <- NA
  
  


```

#### look at how they agree
```{r}
# when they disagree, which is right?
dfs$VAA_lea %>% 
  select(patientId,eye, visitId, `_TCVISIT9`, DATEVIS) %>% 
  mutate(across(matches("visit"), ~suppressWarnings(as.numeric(.)))) %>% 
  arrange(patientId, eye, DATEVIS) %>% 
  group_by(patientId) %>% 
  filter(any(visitId != `_TCVISIT9`)) %>% 
  print(n = 50)
# ^ TCVISIT9 is correct when visitId wrong

# compare two tcvisit9s
dfs$VAA_lea %>% 
  select(patientId,eye, TCVISIT9, `_TCVISIT9`, DATEVIS) %>% 
  mutate(across(matches("visit"), ~suppressWarnings(as.numeric(.)))) %>% 
  arrange(patientId, eye, DATEVIS) %>% 
  group_by(patientId) %>% 
  filter(any(TCVISIT9 != `_TCVISIT9`)) %>% 
  print(n = 50)
# ^ mostly the same except _TCVISIT... is more correct
```

#### make visitID tcvisit when not equal
```{r}

# where visitId != `_TCVISIT9` and !is.na(`_TCVISIT9`), make visitId = `_TCVISIT9`
Where <- dfs$VAA_lea$visitId!=dfs$VAA_lea$`_TCVISIT9` & !is.na(dfs$VAA_lea$`_TCVISIT9`) 
dfs$VAA_lea$visitId[ Where] <- dfs$VAA_lea$`_TCVISIT9`[Where]

# round visitId to nearest multiple of 6
round_to_6 <- function(x) {
  divisor <- 6
  remainder <- x %% divisor
  ifelse(remainder < divisor/2 , x - remainder, x + (divisor - remainder))
}

# make multiple of 6 where no character to mess it up
Is.numeric <- !grepl("[a-zA-Z]", dfs$VAA_lea$visitId)
dfs$VAA_lea$visitId[Is.numeric] <- round_to_6(as.numeric(dfs$VAA_lea$visitId[Is.numeric]))

# format to have 3 numbers
dfs$VAA_lea$visitId <- sprintf("%03s", dfs$VAA_lea$visitId)

```



#____

# filter visitless and controls from all data
```{r}

# distinct patients in each df who are control or visitless
sapply(dfs, function(a_df){
  
  if("visitId"%in% names(a_df)){
    a_df[a_df$visitId %in% c("control", "Control", "VISITLESS"), "patientId"] %>% n_distinct
  }
  
})
# ^ all have 5784 


# filter visitless and controls from dfs
dfs <-  lapply(dfs, function(a_df){
    if(!"visitId"%in% names(a_df)){
      return(a_df)
    }else{
      a_df <- a_df[!a_df$visitId %in% c("control", "Control", "VISITLESS" ) ,  ]
      return(a_df)
    }
  })


(m <- obs_m());
# ^                   CF_FFA DEM   DIAGNOSIS ENR   FAFGRADE FAMSTATUS MH    OCTMAIN OCTMTCIRRUS OCTMTSPECTRALIS OCTMTSTRATUS VAA
# n_patIDs            4951   4951  4951      4951  4951     4951      4951  4951    4951        4951            4951         4951
# n_patID_eyes        9899   9899  9899      9899  9899     9899      9899  9899    9899        9899            9899         9899
# n_observations      15287  15287 15287     15287 15287    15287     15287 15287   15287       15287           15287        13723
# n_patIDs_w_gr_1_vis 550    550   550       550   550      550       550   550     550         550             550          550
## ^ only a few observations gone because visitless aren't counted in observation count, no eye to group on 

### 3 MAR 2025 ###
#                     CF_FFA DEM  DIAGNOSIS ENRNHOR ENRNHOS FAFGRADE FAMSTATUS MH    OCTMAIN OCTMTCIRRUS OCTMTSPECTRALIS OCTMTSTRATUS VAA_lea VAA_old
# n_patIDs            5289   1158 5289      5289    5289    5289     5289      5289  5289    5289        5289            5289         5141    4809   
# n_patID_eyes        10578  NULL 10578     10578   10578   10578    10578     10578 10578   10578       10578           10578        10282   9615   
# n_observations      15968  1158 15968     15968   15968   15968    15968     15968 15968   15968       15968           15968        14068   14215  
# n_patIDs_w_gr_1_vis 552    NULL 552       552     552     552      552       552   552     552         552             552          565     4809 

dfs_dims()
# CF_FFA: 24700,97
# DEM: 24700,12
# DIAGNOSIS: 24700,9
# ENR: 24700,40
# FAFGRADE: 24700,18
# FAMSTATUS: 24700,8
# MH: 24700,62
# OCTMAIN: 24700,29
# OCTMTCIRRUS: 24700,18
# OCTMTSPECTRALIS: 24700,32
# OCTMTSTRATUS: 24700,20
# VAA: 24700,74

### 3 MAR 2025 ###
# CF_FFA: 25573 97
# DEM: 1173 7
# DIAGNOSIS: 25573 9
# ENRNHOR: 25573 27
# ENRNHOS: 25573 29
# FAFGRADE: 25573 20
# FAMSTATUS: 25573 10
# MH: 25573 55
# OCTMAIN: 25573 29
# OCTMTCIRRUS: 25573 21
# OCTMTSPECTRALIS: 25573 31
# OCTMTSTRATUS: 25573 23
# VAA_lea: 15716 40
# VAA_old: 49400 43

```


# remove rows with NA eye
```{r}

# how many observations where eye missing?
sapply(dfs, function(a_df){
    if("eye" %in% names(a_df)){
      sum(is.na(a_df$eye))
    }
  })

# what data where eye missing?
# dfs[c("CF_FFA","OCTMAIN","OCTMTSTRATUS","OCTMTCIRRUS","OCTMTSPECTRALIS","FAFGRADE")] %>% 
lapply( dfs, function(a_df){
  if("eye" %in% names(a_df)){
    a_df %>% 
      filter(is.na(eye)) %>% 
      lapply(function(col){
        if(n_distinct(col,na.rm = FALSE)<20) table(col, useNA = "always") else head(col,10)
      })
  }
}) %>% unlist()
# ^ no MT, OCT, FAF, or CF data, all NA; only some MH, VAA, ENR, and DEM (most full is DEM$GENDER)

# filter out missing eyes from dfs
dfs <-  lapply(dfs, function(a_df){
  if("eye" %in% names(a_df)){
    a_df <- a_df[!is.na(a_df$eye)  ,  ]
  }
  return(a_df)
})

# look at dimensions of dfs after filtering out missing eyes
obs_m()
#^ didnt change because calculation needs eye in the first place

```

#________________________
# Dates

### set dates to date type
```{r}

# grab date columns
date_cols <- map(dfs, ~{grep("dt|date",names(.), ignore.case = TRUE, value = TRUE) %>% grep("HRD",., value = TRUE, invert = TRUE)}) %>% 
  unlist()

# ****** Function to make two formats date type ******
dynamic_date <- function(...){
  format = ifelse(any(grepl("[A-Za-z]", ...)), "%b %d %Y", "%Y-%m-%d")
  as.Date(..., format = format)
}
# dynamic_date("FEB 10 2020")
# dynamic_date("2021-10-2")


# set dates to DATE for dates identified above
dfs <- lapply(dfs, function(a_df){
  a_date_col <- intersect(names(a_df), date_cols)
  a_df <- a_df %>% 
    mutate(across(all_of(a_date_col), ~dynamic_date(.)))
  return(a_df)
})

# !!!!!!!!!!!! CHECK !!!!!!!!!!!! 
map(dfs, ~{x <- .
  x <- select(x, any_of(date_cols))
  return(x)
})
# !!!!!!!!!!!!  !!!!!!!!!!!! 


```

### Remove outlier dates
```{r}


# get all date values to find min threshold
map(dfs, ~{
  d <- .
  d %>% 
    select(any_of(date_cols)) %>% 
    unlist()
}) %>% 
  unlist() %>% 
  sort() %>% 
  as.Date() %>%
  head(50)
  # boxplot()
  # density(na.rm = T) %>% 
  # plot
# ^   "1970-01-01"  or 0  seems like the minimum

# set outliers to missing in date columns
dfs <- dfs %>% 
  map(~{ d = .
    cols <- intersect(names(d), date_cols)
    d <- d %>% 
      mutate(across(all_of(cols), ~ifelse(as.numeric(.)<=0, NA_Date_, .))) %>% 
      mutate(across(all_of(cols), ~as.Date(.)))
    return(d)
  })


# !!!!!!!!!!!! CHECK !!!!!!!!!!!! 
map(dfs, ~{x <- .
  x <- select(x, intersect(date_cols, names(x)))
  return(x)
})
# !!!!!!!!!!!!  !!!!!!!!!!!! 


```


### look at how dates overlap
```{r}

# distinct patient visit dates in each df
vars = c("patientId", "visitId", date_cols)
dates_dfs <- map(dfs, ~{ d <- .
  d %>% 
    select(all_of(intersect(names(d), vars))) %>% 
    distinct()
})

# make df of visits dates across dfs
dates <- purrr::reduce(dates_dfs, left_join) %>% 
  mutate(across(matches("date|dt"), ~{
    if(!is.Date(.)){
      format = ifelse(any(grepl("[A-Za-z]", .)), "%b %d %Y", "%Y-%m-%d")
      as.Date(., format = format)
    } else{
      .
    }
  }))

# what dates to use of all dates
date_columns <- grep("date|dt", names(dates), value = TRUE, ignore.case = TRUE) 
 
# make matrix of date_columns
m <- dates %>%
  select(all_of(date_columns)) %>%
  mutate_all(~ifelse(is.na(.), 0, 1)) %>%
  as.matrix()

m_full <- m[rowSums(!is.na(m)) != 0   ,  ]

# m_full %>% 
m %>% 
  Heatmap(.,
          show_heatmap_legend = TRUE,
          use_raster = F,
          column_title  = "dates",
          # row_title  = "",
          col = c("white" ,  "black"),
          na_col = "white",
          cluster_rows = FALSE,
          cluster_columns = FALSE,
          show_row_names = FALSE,
          show_column_names  = TRUE,
          # column_names_rot = 45,
          # column_names_gp = gpar(fontsize = 3)
          row_names_gp = gpar(fontsize = 3)
  )

```
> CF_FFA ,CF, Ffa, faf, oct, vaa all really full



### find what dates are disagreeing within each df
```{r}

# what dfs have >1 date column
sapply(dfs, function(df){
  sum(grepl(paste0("^", paste(date_columns, collapse = "$|^"), "$"), names(df)))
}) >1
# ^ CF_FFA and OCTMAIN have > 1 date column

# summary of date disagreements
sapply(dfs , function(df){
  
  # date cols in this df
  a_dates <- intersect(names(df), date_columns)
  
  # if > 1 date cols
  if(length(a_dates) >1 ){
    
    # every combination of 1:6
    which_cols <- expand.grid(1:length(a_dates),1:length(a_dates) ) %>% 
      # filter for unique combinations 
      filter(Var1 != Var2) %>% 
      filter(Var1 < Var2) %>% 
      arrange(Var1)
    
    # summarise difference between each date combination
    # i = 1
    for ( i in 1:nrow(which_cols)){
      
      idxs_row <- which_cols[i,]  
      diff_d <- df %>% 
          select(any_of(a_dates)) %>%
          select(idxs_row[[1]], idxs_row[[2]]) %>% 
          # na.omit() %>% 
          filter(if_any(any_of(a_dates), ~!is.na(.))) %>% 
          mutate(diff = difftime(.[[1]], .[[2]], units = "days") %>% as.numeric()) 
        
      
      summ <- summary(diff_d)
      count_diff <- paste("num rows w/ dates > 19 day diff:", nrow(diff_d %>% filter(abs(diff)>19)))
      return(
        list(Summary = summ, Count_Diff = count_diff )
             )
    }
    
  }else{
    return("Not applicable")
  }
  
})
# ^^ 0 observations with dates > 19 days apart
```
> within df dates don't disagree much 

### create single date field in each df
```{r}

# new_df list to work on 
new_dfs <- dfs %>% 
  lapply(., as_tibble)

# do dates with two date vars first
## make most aligned date, if na, make other  date
new_dfs$OCTMAIN <- new_dfs$OCTMAIN %>% 
  mutate(DATE = as.Date(PHOTODT_OCT))

# # what's the difference distr like
# new_dfs$CF_FFA %>%
#   select(all_of(c("PHOTODT_FFA", "PHOTODT_CF"))) %>%
#   mutate(diff = difftime(.[[1]] ,.[[2]] , units ="days") ) %>%
#   filter(abs(diff)>0) %>%
#   pull(diff) %>%
#   as.numeric() %>%
#   boxplot(main = "Difference between PHOTODT FFA and CF", ylab = "Day Difference")

# where no cf, use ffa date; when cf and ffa disagree, cf tends to agree with oct
new_dfs$CF_FFA <- new_dfs$CF_FFA %>% 
  mutate(DATE = PHOTODT_CF) %>% 
  mutate(DATE = ifelse(is.na(DATE), PHOTODT_FFA, DATE)) %>% 
  mutate(DATE = as.Date(DATE))

# do remainder
df <- new_dfs$OCTMTSTRATUS
new_dfs <- map(new_dfs, function(df){
  # Check if "DATE" column already exists
  if (!"DATE" %in% names(df)) {
    # Identify available date columns
    date_column <- intersect(names(df), date_columns)
    
    # If a date column exists, create "DATE" column
    if (!is_empty(date_column)) {
      date_column <- date_column[1]  # Ensure a single column is used
      new_df <- df%>% mutate(DATE = as.Date(.[[date_column]]))
      return(new_df)
    }
  }
  return(df)
})

map(new_dfs, ~{any(grepl("^DATE$", names(.)))}, simplify = T)

# look if dims changed
obs_m(new_dfs)
# ^ no change

```


## impute dates from visitId in each df
foward/backfill date using visitId, use lm 
```{r}

# impute DATE in dfs
new_dfs <- parallel::mclapply(names(new_dfs), function(a_df_name){
      
      a_df <- new_dfs[[a_df_name]]
      
      # if DATE andj visit ID to work with
      if(all(  c("DATE", "visitId") %in% names(a_df) )){
        
        # make a numeric visitId
        a_df <-a_df %>% 
          mutate(visitId_numeric = as.numeric(visitId))
      
          
        # remove absurd dates
        a_df.1 <- a_df %>% 
          mutate(DATE = case_when(
            DATE<0 ~ NA_Date_,   
            TRUE ~ DATE
          )) 
        
        a_df.2 <- a_df.1 
        
        # ## impute dates with lmb !!!! Does not work very well !!!! 
        # a_df.2 <- a_df.1 %>% 
        #    arrange(patientId, eye, visitId_numeric) %>%
        #    group_by(patientId, eye)  %>%
        #    mutate(n_obs = sum(!is.na(DATE))) %>% 
        #    mutate(numeric_DATE = as.numeric(DATE)) %>% 
        #    # run linear model to predict dates
        #    mutate(lm_numeric_DATE = ifelse(n_obs > 2, 
        #                        predict(lm(numeric_DATE ~ visitId_numeric, data = .), newdata = cur_data()) , 
        #                        numeric_DATE) ) %>% 
        #   mutate(lm_DATE = as.Date(lm_numeric_DATE)) %>% 
        #   mutate(lm_DATE = as.Date(ifelse(is.na(lm_DATE), DATE, lm_DATE))) %>% 
        #   ungroup()
        
        
        # for remainder of dates calculate manually
        a_df.3 <- a_df.2 %>% 
         group_by(patientId, eye) %>%
         mutate(
          # Forward fill (carry the last known date forward)
          DATE_ffill = na.locf(DATE, na.rm = FALSE),
          visitId_ffill = visitId_numeric[DATE_ffill==DATE] ,
        
          # Backward fill (carry the next known date backward)
          DATE_bfill = na.locf(DATE, na.rm = FALSE, fromLast = TRUE),
          visitId_bfill = visitId_numeric[DATE_bfill==DATE]
          ) %>%
        mutate(
          # foward fill visitId associated with foward filled date
          visitId_ffill= na.locf(visitId_ffill, na.rm = FALSE),
          # backward fill visitId associated with backward  filled date
          visitId_bfill = na.locf(visitId_bfill, na.rm = FALSE, fromLast = TRUE)
          )  %>%
        mutate(
            new_DATE =
              ifelse(
                  is.na(DATE) & is.na(DATE_bfill) ,  # if date and bfill na
                  # Use forward fill vars to calculate
                  ymd(DATE_ffill) %m+% months(visitId_numeric - visitId_ffill),  
                    ifelse(
                      is.na(DATE) & is.na(DATE_ffill), # if date and ffill na
                      # Use backfill vars to calculate
                      ymd(DATE_bfill) %m-% months(visitId_bfill - visitId_numeric), 
                        ifelse(
                          # if both fills available, choose one with non missing visitID fill
                          is.na(DATE) & !is.na(DATE_ffill) & !is.na(DATE_bfill) & is.na(visitId_ffill),
                          ymd(DATE_bfill) %m-% months(visitId_bfill - visitId_numeric), 
                            ymd(DATE_ffill) %m+% months(visitId_numeric - visitId_ffill)
                          )
                        )
                ),
                DATE  # If DATE is not missing, keep it unchanged
        ) %>%
        mutate(new_DATE = ifelse(is.na(new_DATE), DATE, new_DATE)) %>% 
        ungroup() %>% 
        select(-matches("fill|n_obs|lm")) %>%
        mutate(across(c(DATE, new_DATE), ~as.Date(.)))
        
        # make field to show if date imputed
        a_df.3$DATE_imputed <- !is.na(a_df.3$new_DATE) & is.na(a_df.3$DATE)
        
        # # set DATE to new_DATE
        # a_df.3$DATE <- a_df.3$new_DATE
        # a_df.3$new_DATE <- NULL
        # 
        
        return(a_df.3)
        
        }else{
          
        return(a_df)
          
        }
}, mc.cores = detectCores()-1)
# ^ throws warning becasue of making visitId numeric

names(new_dfs) <- names(dfs)

# check where DATE imputed
sapply(new_dfs, function(df){
  if(all(c("patientId", "DATE_imputed") %in% names(df))){
    df %>% 
      select(patientId, eye, visitId,matches("DATE")) %>% 
      arrange(patientId, eye, visitId) %>% 
      group_by(patientId) %>%
      filter(any(DATE_imputed)) %>%
      ungroup() %>% 
      print(n = 20)
    }else{
      "patientId and DATE_imputed not in df"
      }
})

# Make DATE = new_DATE
new_dfs <- new_dfs %>% 
  map(~{d <- .
    d$DATE <- d$new_DATE
    d$new_DATE <- NULL
    return(d)
  })

obs_m(new_dfs)

# sapply(names(new_dfs), function(nm){
#   corr <- tryCatch(
#     cor(as.numeric(new_dfs[[nm]]$new_DATE), as.numeric(new_dfs[[nm]]$DATE), use = "pairwise.complete.obs"),
#     error = function(e) NA,
#     warning = function(w) NA
#     )
#   return(corr)
# })
# # ^ all new dates correlate perfectly


```


#### look at difference between every combo of dates between new_dfs
```{r} 

#  subset each df for date and identifiers
new_dates_dfs <- lapply(new_dfs, function(a_df){
  
  if("DATE" %in% names(a_df)){
    
      a_df <- a_df %>% 
        select(visitId, patientId, eye, DATE) %>% 
        select(-matches("HRDEXUDATE|ENR|ICF|MH|GRADING")) %>% 
        mutate(eye = as.numeric(eye))
      
      return(a_df)
      
  }else{
    return(NA)
  }
  
})

# filter nas
new_dates_dfs <- new_dates_dfs[!is.na(new_dates_dfs)]

# make date names unique
for(df.name in names(new_dates_dfs)){
  df_date_colname <- paste(df.name, "DATE", sep = "_")
  new_dates_dfs[[df.name]][[df_date_colname]] <- new_dates_dfs[[df.name]][["DATE"]]
  new_dates_dfs[[df.name]][["DATE"]] <- NULL
  }


# make date df
new_dates <- purrr::reduce(new_dates_dfs, left_join) %>% 
  filter(if_all(c(visitId, patientId, eye), ~!is.na(.))) # filter out missing eyes
lapply(new_dates,summary)

# what new_dates to use
new_date_columns <- grep("_DATE$", names(new_dates), value = TRUE, ignore.case = FALSE) 


# every combination of 1:6
which_cols <- expand.grid(1:length(new_date_columns),1:length(new_date_columns) ) %>% 
  # filter for unique combinations 
  filter(Var1 != Var2) %>% 
  filter(Var1 < Var2) %>% 
  arrange(Var2)

# summarise difference between each date combination
# i = 2
for ( i in 1:nrow(which_cols)){
  
  row <- which_cols[i,]  
  diff_d <- new_dates %>% 
      select(all_of(new_date_columns)) %>%
      select(row[[1]], row[[2]]) %>% 
      na.omit() %>% 
      mutate(diff = difftime(.[[1]], .[[2]], units = "days") %>% as.numeric()) 
    
  print(new_date_columns[c(row[[1]], row[[2]])])
  summary(diff_d) %>% print
  print(" ")
  print(" ")
}

```
> ~20 day periods should work in fuzzy join, 5days too

# ????????  LOOK AT PATIDS PATTERNS ???????? 
some patids have NB or NA prefix??
```{r, eval = FALSE}

# look at weird prefixes
map(dfs, ~{x <- .
  x %>% 
    pull(patientId) %>% 
    grep("N", ., value = TRUE) %>% 
    substr(., 1, 3) %>% 
    unique()
})

# # PatientIds only
# ids <- unique(dfs$OCTMTSPECTRALIS$patientId)

# All IDs across all patID types
ids <- map(new_dfs, ~{
  x <- .
  x %>% 
    select(matches("patientId|patId")) %>% 
    unique()
}) %>% 
  unique() %>% 
  purrr::reduce(left_join) %>% 
  as.data.frame()

# Function to find the longest common substring among multiple strings
find_common_substring <- function(strings) {
  if (length(strings) < 2) return(strings)  # If only one string, return as-is
  
  # Convert to DNAStringSet (Biostrings function for efficient LCS computation)
  dna_strings <- DNAStringSet(strings)
  
  # Compute pairwise alignments to find longest common substring
  common_substring <- substr(strings[1], 1, nchar(strings[1]))  # Start with full first string
  for (s in strings[-1]) {
    common_substring <- matchPattern(common_substring, s)
    if (length(common_substring) == 0) return("")  # If no match, return empty string
    common_substring <- as.character(common_substring[[1]])  # Extract match
  }
  
  return(common_substring)
}

# Apply function to each cluster
common_substrings <- sapply(ids, function(id_string){
  
  string <- as.character(na.omit(id_string))
  sapply(string, find_common_substring) %>% 
    sapply(find_common_substring) %>% 
    sapply(find_common_substring)
  
})



# STRING DISTANCE ----
substr_patterns_per_id_type <- list()
for(i in 1:ncol(ids)){
  
  string <- ids[,i] %>% na.omit()
  
  # pat_str_dist <- stringdist::stringdistmatrix(ids, method = "lv")
  pat_str_dist <- stringdist::stringdistmatrix(string, method = "osa")
  # pat_str_dist <- stringdist::stringdistmatrix(string, method = "lcs")
  # pat_str_dist <- stringdist::stringdistmatrix(string, method = "qgram")
  # pat_str_dist <- stringdist::stringdistmatrix(string, method = "qgram")
  
  hc <- hclust(as.dist(pat_str_dist), method = "ward.D2")
  
  height <- 100
  plot(hc, main = sprintf("%s string clusters", names(ids)[i]))
  abline(a = height, b = 0, lty = 3)
  id_clusters <- dendextend::cutree(hc,
                                    h = height,
                                    k = 3
                                    )
  
  ids_clusts <- setNames(object = id_clusters,string)
  
  clust_eg <- sapply(seq(n_distinct(id_clusters)), function(i){
    cat(sprintf("\n%i\n", i))
    names(ids_clusts)[ids_clusts==i]
  }) %>% 
    do.call(cbind,., )
  
  colnames(clust_eg)= seq(n_distinct(id_clusters))
  
  substr_patterns <- sapply(clust_eg, find_common_substring)
  
  substr_patterns_per_id_type[[names(ids)[i]]] <- substr_patterns
  
}

```


### fix duplicate records in dfs
```{r}

map(new_dfs, ~{x <- .
  table(x$eye, useNA = "always")
})

# how many records for each pat, eye, scandDT_period_year?
lapply(new_dfs, function(a_df){
  
  # vars to key observations
  vars <- intersect(c("patientId","eye","DATE", "visitId"), names(a_df))
  # vars <- intersect(c("patientId","eye","DATE"), names(a_df))
  
  # for DATE`
  if("DATE" %in% names(a_df)){
        count_record <- a_df %>%
          select( all_of(vars)) %>%
          # select( all_of(c(vars, "visitId"))) %>%
          filter(!is.na(DATE)) %>%
          group_by(across(any_of(vars))) %>%
          mutate(count = n())

       count_record %>% filter(count>1)
  }else{
    print("DATE not present")
  }

  })
# ^ 202 rows of duplicates in VAA; CF_FFA, faf, and OCTMAIN have one patient with four records for each eye

# why does VAA have duplicate records?
VAA_count_record <- new_dfs$VAA_lea %>%
    group_by(patientId, eye, DATE, visitId) %>%
    mutate(duplicated = duplicated(across(everything()))) 

VAA_count_record %>% 
  group_by(patientId, eye, DATE) %>%
  select(patientId, eye, DATE, everything()) %>% 
  arrange(patientId, eye, DATE) %>% 
  filter(duplicated) %>% 
  glimpse()

# flag duplicates; TRUE if other record with same values, only flags second one
new_dfs<- new_dfs %>% 
  map(~{x <- .
  x <- x %>% 
    mutate(duplicated = duplicated(across(everything())))
  return(x)
  })

# # save file for lea
# write_csv(new_dfs$VAA_lea , sprintf("%sprocessed_data/VAA_lea_flagged_for_duplicate.csv", wd_path))

# ## make pat_eye_DATE variable for all dfs
# new_dfs <- new_dfs %>% 
#   map(~{
#     x <- .
#     if("DATE" %in% names(x) & "eye" %in% names(x)){
#         x$pat_eye_DATE <- paste(x$patientId, x$eye, x$DATE, sep  = "_")
#     }
#     return(x)
# })

# remove duplicates across all dfs; remove 'duplicated' (useless now)
old_nrow <- new_dfs %>% map(nrow)
new_dfs <- new_dfs %>% 
  map(~{
    x <- .
    x <- filter(x, !duplicated)
    x <- select(x, -duplicated)
    return(x)
  })


new_nrow <- new_dfs %>% map(nrow)
unlist(new_nrow) -unlist(old_nrow) %>% t()
# CF_FFA   DEM   DIAGNOSIS   ENRNHOR   ENRNHOS   FAFGRADE   FAMSTATUS   MH   OCTMAIN   OCTMTCIRRUS   OCTMTSPECTRALIS   OCTMTSTRATUS   VAA_lea
#      0   -15           0         0         0          0           0    0         0             0                 0              0       -74
# ^ lots of vaa duplicates

# Check for semi duplicates where patientId-eye-date duplicated
new_dfs$VAA_lea %>% 
  group_by(patientId, eye, DATE) %>% 
  filter(n()>1)
  # View()
# ^ 2 records for each pat,eye,date; measures only slightly different 

# average duplicate records in VAA: first make vars numeric then average them
new_dfs$VAA_lea <- new_dfs$VAA_lea %>% 
  mutate(across(-c(patientId),as_numeric_if_possible)) 

dups_averaged <- new_dfs$VAA_lea %>% 
  group_by(patientId, eye, DATE) %>% 
  mutate(across(where(is.numeric), 
                ~ifelse(!all(is.na(.)), mean(., na.rm = TRUE), .) 
                )
         ) %>% 
  ungroup()

# # look at how average works
# dups_averaged %>% 
#   group_by(patientId, eye, DATE) %>% 
#   filter(n()>1) %>% 
#   View()

# compare dims
(dups_averaged %>%
  group_by(patientId, eye, DATE) %>%
  filter(n()>1) %>%
  distinct() %>%
  nrow() )-
  (new_dfs$VAA_lea %>%
    group_by(patientId, eye, DATE) %>%
    filter(n()>1) %>%
    distinct() %>%
    nrow())
# -25 duplicate records

new_dfs$VAA_lea <- distinct(dups_averaged)

obs_m(new_dfs)
#                     CF_FFA DEM   DIAGNOSIS ENR   FAFGRADE FAMSTATUS MH    OCTMAIN OCTMTCIRRUS OCTMTSPECTRALIS OCTMTSTRATUS VAA  
# n_patIDs            4951   4951  4951      4951  4951     4951      4951  4951    4951        4951            4951         4950 
# n_patID_eyes        9899   9899  9899      9899  9899     9899      9899  9899    9899        9899            9899         9897 
# n_observations      15287  15287 15287     15287 15287    15287     15287 15287   15287       15287           15287        13485
# n_patIDs_w_gr_1_vis 550    550   550       550   550      550       550   550     550         550             550          531  
# ^ ~ 250 obs were duplicated in VAA, ~ 20 patients lost due to duplication

### MAR 5 2025 ### 
#                     CF_FFA DEM  DIAGNOSIS ENRNHOR ENRNHOS FAFGRADE FAMSTATUS MH    OCTMAIN OCTMTCIRRUS OCTMTSPECTRALIS OCTMTSTRATUS VAA_lea
# n_patIDs            5289   1158 5289      5289    5289    5289     5289      5289  5289    5289        5289            5289         5141   
# n_patID_eyes        10578  NULL 10578     10578   10578   10578    10578     10578 10578   10578       10578           10578        10282  
# n_observations      15968  1158 15968     15968   15968   15968    15968     15968 15968   15968       15968           15968        15574  
# n_patIDs_w_gr_1_vis 552    NULL 552       552     552     552      552       552   552     552         552             552          565    

### MAR 10 2025 ### 
#                     CF_FFA DEM  DIAGNOSIS ENRNHOR ENRNHOS FAFGRADE FAMSTATUS MH    OCTMAIN OCTMTCIRRUS OCTMTSPECTRALIS OCTMTSTRATUS VAA_lea
# n_patIDs            5289   1158 5289      5289    5289    5289     5289      5289  5289    5289        5289            5289         5141   
# n_patID_eyes        10578  NULL 10578     10578   10578   10578    10578     10578 10578   10578       10578           10578        10282  
# n_observations      15968  1158 15968     15968   15968   15968    15968     15968 15968   15968       15968           15968        15574  
# n_patIDs_w_gr_1_vis 552    NULL 552       552     552     552      552       552   552     552         552             552          565        


dfs_dims(new_dfs)
### MAR 5 2025 ### 
# CF_FFA: 15968 102
# DEM: 1158 8
# DIAGNOSIS: 15968 10
# ENRNHOR: 15968 32
# ENRNHOS: 15968 33
# FAFGRADE: 15968 25
# FAMSTATUS: 15968 11
# MH: 15968 60
# OCTMAIN: 15968 34
# OCTMTCIRRUS: 15968 26
# OCTMTSPECTRALIS: 15968 36
# OCTMTSTRATUS: 15968 28
# VAA_lea: 15642 45

```

## how many visitIDs in vaa where no date
these would have joined before but wont now
```{r}

new_dfs$VAA_lea %>% 
  select(patientId, eye,  visitId, DATE) %>% 
  mutate(discrepancy = is.na(DATE) & !is.na(visitId) ) %>% 
  filter(discrepancy) %>% 
  pull(visitId) %>% table()
# ^ 0 rows where no date and a visitID
```   

#____________


# **** look at how mt from new_dfs join
```{r, include=FALSE, eval = FALSE}

mt <-new_dfs[c("OCTMTSTRATUS", "OCTMTSPECTRALIS", "OCTMTCIRRUS")]


# rename vars to have table name prefix
for(name in names(mt)){
  names(mt[[name]])[!grepl("patient|visitId|eye", names(mt[[name]]))] <- 
    paste(name, names(mt[[name]])[!grepl("patient|visitId|eye", names(mt[[name]]))], sep = "_")
}

mt_joined <- mt %>%
  purrr::reduce(., left_join, by = join_by(patientId, visitId, eye), relationship = "one-to-one") %>%
  select(-matches("\\.x|\\.y"))

# m_full_across <- mt_joined[ rowSums(!is.na(mt_joined)) > 0 , ]
# 
# Heatmap(
#   # mt_joined,
#   m_full_across,
#   cluster_rows = FALSE, 
#   cluster_columns = FALSE, 
#   # col = c("white", "black")
#   na_col = "white",
#   col  = viridis(100)
#   )

sum(rowSums(!is.na(mt_joined)) > 0) 
# 13091>>15968 rows of mt data theoretically
rowSums(!is.na(mt_joined)) %>% density() %>% plot(main = "distribution of row missing count")

sum(rowSums(!is.na(mt_joined[, c("OCTMTSTRATUS_stratus_MT01", "OCTMTSPECTRALIS_spectralis_MT01", "OCTMTCIRRUS_cirrus_MT01")])) == 3)
# 29 >> 31 overlap between all three machines
sum(rowSums(!is.na(mt_joined[, c("OCTMTSTRATUS_stratus_MT01", "OCTMTSPECTRALIS_spectralis_MT01", "OCTMTCIRRUS_cirrus_MT01")])) >1)  
# 893>>903 where overlap between any two
sum(rowSums(!is.na(mt_joined[, c("OCTMTSTRATUS_stratus_MT01", "OCTMTSPECTRALIS_spectralis_MT01")])) == 2) 
# 377 >> 383 overlaps between strat and spec
sum(rowSums(!is.na(mt_joined[, c("OCTMTSTRATUS_stratus_MT01", "OCTMTCIRRUS_cirrus_MT01")]))== 2) 
# 235 >> 237 overlaps between strat and cirrus
sum(rowSums(!is.na(mt_joined[, c("OCTMTSPECTRALIS_spectralis_MT01", "OCTMTCIRRUS_cirrus_MT01")]))== 2) 
# 339 >> 345 overlaps between spec and cirrus


# tab <- data.table::fread(
# "old new
# 29 31 
# 893 903 
# 377 383
# 235 237
# 339 345") %>% as.data.frame()
# 
# tab %>% 
#   plot()
# add_lines

rm("mt_joined")

```



#_______________________

# Join Data 
## rename DATE columns to DATE_df, same for date imputed
```{r}


for(i in 1:length(new_dfs)){

    a_df <- new_dfs[[i]]
    a_df_name <- names(new_dfs)[i]

    # rename date to unique
    names(a_df)[grep("^DATE$", names(a_df))] <- paste( "DATE", a_df_name, sep =  "_")
    
    # rename date imputed to unique
    names(a_df)[grep("^DATE_imputed$", names(a_df))] <-paste( "DATE", a_df_name, "imputed", sep =  "_")
    
    new_dfs[[i]] <- a_df

}


# ****** Function to reorder cols to show id cols first ******
reorder_id_cols <- function(df){
 # lookup<-"patID_eye|patID|patientId|eye|visitId|DATE"
  lookup<- c("patID_eye","patID","patientId","eye","visitI[Dd]","DATE_coalsced","DATE")
  vars <- sapply(lookup, function(l)grep(l, names(df), value = TRUE) %>% grep("HRDEX", ., value = TRUE, invert = TRUE)) %>% 
    # sapply(`[`,1)
    unlist() %>% 
    as.character()
  select(df, all_of(vars), everything())
}

# reorder with grep first then next
new_dfs <- new_dfs %>% map(reorder_id_cols)


```

## remove comment vars from dfs
```{r}

new_dfs %>% map(~{x <- .
  grep("COM", names(x), value = TRUE, ignore.case = TRUE)  
})

# remove vars with COMM
new_dfs <- new_dfs %>% map(~{x <- .
  x <- x[, !grepl("COMM", names(x))]
  return(x)
})

```


## first check how many unique patient Ids and subjects are in each
```{r}

# make a list of df names with the number of unique patient ids in them
p_list <- list();for (i in 1:length(new_dfs)) {
  df_name <- names(new_dfs)[i]
  pats <- length(unique(new_dfs[[i]]$patientId))
  p_list[[df_name]] <- pats
}
p_list

```

## join enr tables together
```{r}

# simplify list
ENR_l <- list(NHOS = new_dfs$ENRNHOS,NHOR = new_dfs$ENRNHOR)

# what are these vars
ENR_l %>% 
  map(~{
    x <- . 
    x%>% 
      map(~{x <- .; head(unique(x))})
      # names
  })

# filter for unique patientId rows
ENR_l <-ENR_l %>% map(~{ 
  x <- . 
  x%>% 
  select(-matches("visitId"), -eye, -matches("DATE_")) %>% 
  distinct() 
})

# Make ENRDATE col unique to table
names(ENR_l$NHOS)[names(ENR_l$NHOS) == "ENRDATE"] <- "NHOS_ENRDATE"
names(ENR_l$NHOR)[names(ENR_l$NHOR) == "ENRDATE"] <- "NHOR_ENRDATE"


# Make relatproband col unique to table
names(ENR_l$NHOS)[names(ENR_l$NHOS) == "RELATPROBND"] <- "NHOS_RELATPROBND"
names(ENR_l$NHOR)[names(ENR_l$NHOR) == "RELATPROBND"] <- "NHOR_RELATPROBND"

# Make relatproband col unique to table
names(ENR_l$NHOS)[names(ENR_l$NHOS) == "PROBANDPW"] <- "NHOS_PROBANDPW"
names(ENR_l$NHOR)[names(ENR_l$NHOR) == "PROBANDPW"] <- "NHOR_PROBANDPW"

# make id cols first
ENR_l <- map(ENR_l,~{x <- .
  x <- reorder_id_cols(x)
  as.data.frame(x)
})

# check that each row a patient
ENR_l %>% 
  map(~{
    sum(table(.x$patientId)>1)
  })

# combine everything together
full_ENR <- purrr::reduce( ENR_l, full_join, by = "patientId") 

# any patients with >1 rows?
sum(table(full_ENR$patientId)>1)

# fix FAMILYID var
sum(full_ENR$FAMILYID.x != full_ENR$FAMILYID.y, na.rm = TRUE)
# ^ 0 means no disagreements

full_ENR$FAMILYID <- coalesce(full_ENR$FAMILYID.x, full_ENR$FAMILYID.y)

# remove repeat vars
names(full_ENR) <- str_remove_all(names(full_ENR), "\\.x")
full_ENR <- full_ENR[ , -grep("\\.y", names(full_ENR))]

# add to main list
new_dfs$ENR <- full_ENR


```


## filter static varirables
### filter diagnosis vars
```{r}

# check if BLRPATTERN changes over 'time'
if("visitId" %in% names(new_dfs$DIAGNOSIS)){
  new_dfs$DIAGNOSIS %>% 
      select(patientId, eye, visitId, BLRPATTERN) %>% 
      mutate(BLRPATTERN= ifelse(BLRPATTERN == "Yes", 1, ifelse(BLRPATTERN == "No", 0, BLRPATTERN))) %>% 
      group_by(patientId, eye, visitId) %>% 
      filter(n_distinct(BLRPATTERN)>1)
  # ^ no change in blrpattern so must be static; -> then visitId useless here
}

new_dfs$DIAGNOSIS <- new_dfs$DIAGNOSIS %>% 
  select(patientId, eye,  DIAGNOSIS, BLRPATTERN) %>% 
  distinct()

     
``` 


### for demographic, enrolment, medical history and famstatus data, remove all but static and salient variables 
```{r}

DEM <- new_dfs$DEM %>%  
  select(patientId, RACE, GENDER, ETHNICIT, YROFBRTH) %>% 
  unique()


(!is.na(DEM$YROFBRTH)) %>% table
# TRUE 
# 1158 

FAMSTATUS <- new_dfs$FAMSTATUS %>%  
  select(patientId, patId2, FAMILY_ID, FAMILY_STATUS) %>% 
  unique()
# table(FAMSTATUS$patientId)[table(FAMSTATUS$patientId)>1]

ENR <- new_dfs$ENR %>%  
  select(patientId, matches("ENRDATE"), matches("RELATPROBND"), matches("^NHO.ENR$"), FAMILYID, matches("PROBANDPW")) %>% 
  unique() %>% 
  mutate(NHOSENR =  case_when(
    NHOSENR == "No" ~ FALSE,
    NHOSENR == "Yes" ~ TRUE,
    NHOSENR == "Natural History Observation study" ~ TRUE
  ))
# table(ENR$patientId)[table(ENR$patientId)>1] 
# ^ 2 records per some patients because they are in NHOS and NHOR

# filter MH to have pat, vis, date
MH <- new_dfs$MH %>%  
  select(patientId, visitId, DATE_MH, DATE_MH_imputed, matches("DM|HGT|WGT|SMK")) %>% 
  unique()
# table(MH$patientId)[table(MH$patientId)>1] %>% table()
# ^ some have 2-6 rows?

# are there any duplicates, i.e. are there any eye specific vars?
MH %>% 
  group_by(patientId, visitId) %>% 
  mutate(duplicate_sum= sum(duplicated(across(everything())))) %>% 
  filter(duplicate_sum>0) 
# ^ none


# filter out rows with no data; any  b/c allows for any number of non missing
MH_dense <- MH %>% 
  group_by(patientId, visitId) %>% 
  filter(if_any(everything(), ~!is.na(.)))

  
```

### join static dfs together for easy join
```{r}

# join these first because they don't have visit Id
ENR_DEM_FAMSTATUS <- list(ENR, DEM, FAMSTATUS) %>% 
  purrr::reduce(full_join) 

ENR_DEM_FAMSTATUS %>% dim
list(ENR, DEM, FAMSTATUS) %>% 
  map(dim)
#^ all same dim

# full join; all obs in x and y are observed even if no match between them
patient_info_df <- full_join(MH_dense,  ENR_DEM_FAMSTATUS, relationship = "many-to-one") 

```


## Main Joining

### join the grading data, diagnosis, and patient info
```{r}

# what id vars does each df have
id_cols <- function(df){
    datevar <- grep("DATE", names(df), value = TRUE) %>% 
      grep("HRDEXUDATE", invert = TRUE, value = TRUE, .)
    c("patientId", "eye", "visitId", datevar)[c("patientId", "eye", "visitId", datevar) %in% names(df)] %>% 
      sort()
}
#+ ^ first join specific rows with pat, eye, visitId/date, then less specific, patientId, eye...

what_to_join <- c("CF_FFA", "OCTMTSPECTRALIS", "OCTMTSTRATUS", "OCTMAIN", "FAFGRADE","OCTMTCIRRUS")


# check keys
map(new_dfs[what_to_join], id_cols)

# join grading on visit pat eye
grading_joined  <- new_dfs[which(names(new_dfs) %in% what_to_join)] %>%
  purrr::reduce(., dplyr::full_join, by = c( "eye", "patientId", "visitId"), relationship = "one-to-one" )


# check keys
id_cols(grading_joined)
id_cols(new_dfs$DIAGNOSIS)

# join Diagnosis and cirrus
grading_diagnosis_joined <- grading_joined %>% 
  full_join(., new_dfs$DIAGNOSIS, by = c("patientId", "eye"), relationship = "many-to-one" ) 


# check keys
id_cols(grading_diagnosis_joined)
id_cols(patient_info_df)
dim(grading_diagnosis_joined)
dim(patient_info_df)

(table(with(patient_info_df,paste(patientId, visitId, sep = "_")))>1) %>% sum 
# ^ 0 pat,vis with more than one record

# then join patient info; many to many because pat,vis, eye in x and pat, vis, nhor, nhos in y
grading_diagnosis_info_joined <- grading_diagnosis_joined %>% 
  full_join(., patient_info_df, by = c("patientId", "visitId"), relationship = "many-to-one" )
 

```

#### collapse dates in grading_diagnosis_info_joined
```{r}

# SEE HOW DATES CORRELATE ----
date_m <- grading_diagnosis_info_joined %>% 
  select(matches("DATE")) %>% 
  select(-matches("HRDEXUDATE|imputed|ENR")) %>% 
  mutate_all(as.numeric) %>% 
  select(-DATE_OCTMTSTRATUS)

good_date_cols <- colnames(date_m)

# missingness in stratus date  
(!is.na(grading_diagnosis_info_joined$DATE_OCTMTSTRATUS )) %>% sum
# ^ 0 in DATE_OCTMTSTRATUS


date_cor <- date_m %>% 
  cor(use = "pairwise.complete") 

date_cor %>% 
  Heatmap(cluster_columns = TRUE, cluster_rows = TRUE, col = viridis::plasma(10))

# USE HC TO TRIAGE WHAT DATES TO COALESCE AND IN WHAT ORDER ----
date_dist <- dist(t(date_m))
date_dist <- date_dist/sum(date_dist)
date_hc <- hclust(date_dist, method = "ward.D")

plot(date_hc);abline(a = 0, b = 0, lty = 3)
# ^ coalesce in order of height of merge
#+ CF, faf, octmain, spectralis, MH, cirrus


# combine cf and faf
grading_diagnosis_info_joined$DATE_coalsced <- coalesce(grading_diagnosis_info_joined$DATE_CF_FFA,
                                                grading_diagnosis_info_joined$DATE_FAFGRADE)

# combine oct and coalesced
grading_diagnosis_info_joined$DATE_coalsced <- coalesce(grading_diagnosis_info_joined$DATE_coalsced,
                                                grading_diagnosis_info_joined$DATE_OCTMAIN)

# combine MH and spectralis
grading_diagnosis_info_joined$DATE_coalsced_mh_spec <- coalesce(grading_diagnosis_info_joined$DATE_MH,
                                                grading_diagnosis_info_joined$DATE_OCTMTSPECTRALIS)

# combine coalesced and DATE_coalsced_mh_spec
grading_diagnosis_info_joined$DATE_coalsced <- coalesce(grading_diagnosis_info_joined$DATE_coalsced,
                                                grading_diagnosis_info_joined$DATE_coalsced_mh_spec)

# combine coalesced and cirrus
grading_diagnosis_info_joined$DATE_coalsced <- coalesce(grading_diagnosis_info_joined$DATE_coalsced,
                                                grading_diagnosis_info_joined$DATE_OCTMTCIRRUS)

# remove intermediate
grading_diagnosis_info_joined$DATE_coalsced_mh_spec <- NULL

# (!is.na(grading_diagnosis_info_joined[, c(good_date_cols,"DATE_coalsced", "DATE_coalsced_mh_spec") ] ) ) %>%
#   colSums %>%
#   which.max


```



#### clean the grading-info-diagnosis
```{r}

# do vars duplciated by join all agree?
duplicate_vars <- names(grading_diagnosis_info_joined) %>% 
  grep("^i|\\.x|\\.y", ., value = TRUE)
duplicate_vars_patterns <- str_remove_all(duplicate_vars, "\\.[xy]")


# how many disagreements does each duplicate have with duplicate at i - 1? 
duplicate_vars_d <- grading_diagnosis_info_joined[, duplicate_vars]

disagreements <- list()
# for each variable pattern
for(v in duplicate_vars_patterns){
  # grab repeated vars with v pattern
  df <- duplicate_vars_d[, grep(v, names(duplicate_vars_d))]
  # for each column of df, equate to previous; save sum of disgreements to list
  sums <- c()
  for(i in 2:ncol(df) ){
    sums[i-1] <- sum(df[, i] != df[, (i -1)], na.rm = TRUE)
  }
  # names of 2:ncol for each sum
  names(sums) <- names(df)[-1]
  disagreements[[v]] <- sums
}  
disagreements
# ^ only grader id has disagreements, all others perfect duplicates

names(grading_diagnosis_info_joined)[which(names(grading_diagnosis_info_joined) == "patientId")] <- "patID"
names(grading_diagnosis_info_joined)[which(names(grading_diagnosis_info_joined) == "visitId")] <- "visitID"
names(grading_diagnosis_info_joined)[which(names(grading_diagnosis_info_joined) == "eye")] <- "eye_num"

# grep(paste(duplicate_vars_patterns, collapse = "|"), names(grading_diagnosis_info_joined), value = TRUE)

# remove x.x... and y.y..., keeps .x vars
rm_cols <- grep("^i|\\.x\\.x|\\.y" , names(grading_diagnosis_info_joined), value = TRUE) %>%
  sort()

# exclude from grading_diagnosis_info_joined
grading_diagnosis_info_joined_clean <- grading_diagnosis_info_joined[, -which(names(grading_diagnosis_info_joined) %in% rm_cols)]

# Check
setdiff(names(grading_diagnosis_info_joined), names(grading_diagnosis_info_joined_clean)) # old - new = difference ( what's in old that isn't in new )

# remove .x suffix
names(grading_diagnosis_info_joined_clean) <- str_remove(names(grading_diagnosis_info_joined_clean), "\\.x")

# create patID_eye var to loop over
grading_diagnosis_info_joined_clean$patID_eye <- paste(grading_diagnosis_info_joined_clean$patID, grading_diagnosis_info_joined_clean$eye_num, sep = "_")


# I missed some duplicate names; make the
names(grading_diagnosis_info_joined_clean) <- make.unique(names(grading_diagnosis_info_joined_clean), sep = "_")


```

##### Remove duplicates from grading_diagnosis_info_joined_clean
```{r}

dup <- grading_diagnosis_info_joined_clean %>% 
  group_by(patID_eye, DATE_coalsced) %>% 
  filter(n()>1) %>% 
  arrange(patID_eye)
# ^ looks like comment fields making them non-unique

glimpse(dup) 
#+ ^ duplicated rows are because of faf vars for some reason; where faf, no oct, cfffa


# what vars have one non-missing and one missing; !!! this depends on how many patID_eye-DATE records are duplicated !!!
dup_vars <- sapply(dup, function(col){
  sum(is.na(col)) == length(col)/2
}) %>% 
  Filter(function(x) x, .) %>% 
  names()

glimpse(dup[, c("patID_eye","DATE_coalsced", dup_vars)])
glimpse(dup)
#+ ^ faf and cffa/octx not lining up
# $ patID                        <chr> "NA8880888", "NA8880888", "NA8880888", "NA8880888"
# $ eye_num                      <chr> "1", "1", "2", "2"
# $ visitID                      <chr> "001", "000", "001", "000"
# $ DATE_CF_FFA                  <date> 2025-01-24, 2025-01-24, 2025-01-24, 2025-01-24
# $ CENTER_MICRON                <chr> NA, "4", "2", "10"
# $ CENTERMAX_MICRON             <chr> NA, "6", "2", "12"
# $ CENTERMIN_MICRON             <chr> NA, "5", "2", "11"
#+ ^ these not lined up either


# ****** FUNCTION: if non-unique values and one missing, fill with that value ******
coalesce_vec <- function(x) {
  nonNA <- x[!is.na(x)]
  if(n_distinct(nonNA) == 1){
    x[is.na(x)] <- unique(nonNA)
  }
    return(x)
}
# c(0,0,NA) %>% coalesce_vec()
# c( NA, "2", "2", NA) %>% coalesce_vec()
# c(0,1,NA, NA) %>% coalesce_vec()


# coalesce duplicate observations
grading_diagnosis_info_joined_clean_coalesced <-  grading_diagnosis_info_joined_clean %>%
  group_by(patID_eye, DATE_coalsced) %>%
  arrange(patID_eye,DATE_coalsced) %>% 
  mutate(across(everything(), coalesce_vec)) 

x <- grading_diagnosis_info_joined_clean_coalesced %>% 
  group_by(patID_eye, DATE_coalsced) %>%
  filter(n()>1)

glimpse(x)

# what var makes this not unique?; where uneven number of unique values and not one uinque value
count_unique_vars <- sapply(x, function(z) length(unique(z[!is.na(z)]))) %>% 
  sort()
# un_unique_vars <- Filter(function(n) n%%2 != 0 & n>1, count_unique_vars)
un_unique_vars <- Filter(function(n) n==2 | n%%2!=0 & n>1, count_unique_vars)

x[, names(un_unique_vars)] %>% glimpse
#   STUDYID CENTER_MICRON CENTERMAX_MICRON CENTERMIN_MICRON ...
#   <chr>   <chr>         <chr>            <chr>           
# 1 SS1-001 4             6                5               
# 2 NHOR/S  4             6                5               
# 3 SS1-001 2             2                2               
# 4 NHOR/S  10            12               11 
#+  
#+ ^ non unique 2,2,2 looks fishy and matches the sd vars too
#+ study id wierd
#+ visitID = 001 for dups

# set non-uniques to better value>> those with visitID 000 not 001
grading_diagnosis_info_joined_clean_coalesced %>% 
  filter(patID_eye == "NA8880888_2" & DATE_coalsced == as.Date("2025-01-24")) %>% 
  glimpse()

wherebad <- with(grading_diagnosis_info_joined_clean_coalesced, 
              patID_eye == "NA8880888_2" & 
              DATE_coalsced == as.Date("2025-01-24") & 
              # CENTER_MICRON == 2
              STUDYID == "SS1-001"
              )

wheregood <- with(grading_diagnosis_info_joined_clean_coalesced, 
              patID_eye == "NA8880888_2" & 
              DATE_coalsced == as.Date("2025-01-24") &
              # CENTER_MICRON != 2
              STUDYID != "SS1-001"
              )

# asign likely wrong CENTER_ values to likely correct ones
grading_diagnosis_info_joined_clean_coalesced[wherebad , c("CENTER_MICRON", "CENTERMAX_MICRON", "CENTERMIN_MICRON") ] <- 
  grading_diagnosis_info_joined_clean_coalesced[wheregood , c("CENTER_MICRON", "CENTERMAX_MICRON", "CENTERMIN_MICRON") ]


# for study ID too
grading_diagnosis_info_joined_clean_coalesced$STUDYID[grading_diagnosis_info_joined_clean_coalesced$STUDYID == "SS1-001"] <- "NHOR/S"

# and for visitID; set 001 to 000 for all obs
grading_diagnosis_info_joined_clean_coalesced <- grading_diagnosis_info_joined_clean_coalesced %>% 
  mutate(visitID = ifelse(visitID == "001","000" , visitID)) %>% 
  mutate(visitId_numeric = ifelse(visitId_numeric == 1,0, visitId_numeric))

nrow(distinct(grading_diagnosis_info_joined_clean_coalesced)) - nrow(grading_diagnosis_info_joined_clean_coalesced)
# -2 rows, a duplicate for each eye

# x <- grading_diagnosis_info_joined_clean_coalesced %>%
#   group_by(patID_eye, DATE_coalsced) %>%
#   filter(n()>1) 
# (x[1,] == x[2,]) %>% t() %>% na.omit() %>% all
# #^ all match

# save as new df
grading_diagnosis_info_joined_clean_coalesced <- distinct(grading_diagnosis_info_joined_clean_coalesced) %>% 
  ungroup()
  
```



### join VAA to grading_diagnosis_info_joined
#### make date interval and rename VAA identifiers
```{r}

# create range of dates, start, end, within which to join VAA_DATE to DATE_coalsced
days_window <- 18
new_dfs$VAA_lea <- new_dfs$VAA_lea %>% 
  mutate(start_DATE = ymd(DATE_VAA_lea) - days(days_window), 
         end_DATE = ymd(DATE_VAA_lea) + days(days_window)
         )

# create patID_eye var to loop over
new_dfs$VAA_lea$patID <- new_dfs$VAA_lea$patientId
new_dfs$VAA_lea$eye_num <- new_dfs$VAA_lea$eye
new_dfs$VAA_lea$patID_eye <- paste(new_dfs$VAA_lea$patientId, new_dfs$VAA_lea$eye_num, sep = "_")

```

#### fuzzy_left_join
```{r}

# eyes to loop over 
patID_eyes <- unique(grading_diagnosis_info_joined_clean_coalesced$patID_eye)

# ((((((((((((()))))))))))))  FUZZY JOIN  ((((((((((((())))))))))))) ----
# ******** INITIATE FIRST FUZZY JOIN ******** 

rm("join1")

# fuzzy join first patient
join1 <- grading_diagnosis_info_joined_clean_coalesced %>% 
    filter(patID_eye == patID_eyes[1]) %>% 
    fuzzy_left_join(.,
    # fuzzy_full_join(., 
                    new_dfs$VAA_lea %>% 
                      select(-matches("vis")) %>% 
                      filter(patID_eye == patID_eyes[1]), 
                    by = c(
                      "patID_eye" = "patID_eye",
                      "DATE_coalsced" = "start_DATE", 
                      "DATE_coalsced" = "end_DATE"
                    ),
                    match_fun = list(`==`, `>=`, `<=`)
                    )

# look at join dims
dim(grading_diagnosis_info_joined_clean_coalesced %>% filter(patID_eye == patID_eyes[1])) # x dims
dim(new_dfs$VAA_lea %>% select(-matches("vis")) %>% filter(patID_eye == patID_eyes[1])) # y dims
dim(join1) # x+y dims
# ^ 8 270 


# # look at how date joins
# new_dfs$VAA_lea %>% filter(patID_eye == patID_eyes[1])%>% select(patID_eye, matches("DATE"))
# 
# join1 %>% 
#   mutate(rnames = paste(patID_eye.x, as.character(DATE_coalsced), sep = "_")) %>% 
#   column_to_rownames("rnames") %>% 
#   t() %>% 
#   View()


# ******** REMAINING FUZZY JOINS ******** 

eye_groups <- split(patID_eyes[-1], cut(seq_along(patID_eyes[-1]), breaks = 11, labels = FALSE))

## loop through each patID_eye except first and fuzzy_join VAA data for that eye to main
how.long()
pat_join_l <- mclapply(eye_groups, function(pat_eyes ){
  
  a_join <- grading_diagnosis_info_joined_clean_coalesced %>% 
    filter(patID_eye %in% pat_eyes) %>%
    fuzzy_left_join(.,
    # fuzzy_full_join(., 
                    new_dfs$VAA_lea %>% 
                      select(-matches("vis")) %>% 
                      filter(patID_eye %in% pat_eyes) , 
                    by = c(
                      "patID_eye" = "patID_eye",
                      "DATE_coalsced" = "start_DATE", 
                      "DATE_coalsced" = "end_DATE"
                    ),
                    match_fun = list(`==`, `>=`, `<=`)
                    )
  
    return(a_join)
  
}, mc.cores = detectCores()-1 ) 

# combine joins
joined_raw <- do.call(rbind, pat_join_l) %>% 
  rbind(join1,.)

how.long() # ~ 27s

rm("pat_join_l")

joined <- reorder_id_cols(joined_raw) 

dim(grading_diagnosis_info_joined) # distinct X
 # 15287   238 >> 15287   238 >> 15968   253
dim(joined) # distinct XY
# 15323   287 >> 15321   281 ( reduced time window to 18 days) >> 15998   270 
dim(joined %>%  distinct(patID_eye.x, DATE_coalsced, .keep_all = TRUE)) # distinct XY eye-dates
# 15279   287 >> 15279   287 >>  15966   270

# count observed barplot
colSums(!is.na(joined)) %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  rename(count_present = "." ) %>%
  ggplot(aes(x = variable, y = count_present, fill = variable))+
  geom_bar(stat = "identity", show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Count Non-Missing")

```

##### preprocess joined
```{r}

# COALESCE IDS ----
## duplicated identifiers, e.g. patID, patID_eye
ids <- grep("\\.[xy]", names(joined), value = TRUE) %>% 
  str_remove_all(.,"\\.[xy]") %>% 
  sprintf("^%s\\.",.) %>% 
  unique()

# check where both not na if matching
for(id in ids){
  joined[, grep(id, names(joined))] %>%
    drop_na() %>% 
    {all(.[[1]]==.[[2]])} %>% 
    print()
}# ^ all TRUE good; all overlaps match

id <- ids[[1]]

# coalesce ids: make new var from grep pattern; select vars with pattern; coalesce
for(id in ids){
  new_var <- str_remove_all(id,"\\^|\\\\.")
  joined[[new_var]] <- joined %>% 
    select(matches(id))%>% 
    mutate_all(as.character) %>% 
    as.list()%>% 
    do.call(coalesce,.)
}

joined <- joined %>% 
  select(-matches("\\.[xy]"))

```


#### * QC on how join works
##### missing check that all va joined to grading
    
    * FUZZY_FULL_JOIN adds va where no match with grading so ~990 records not joining *
      
```{r, eval = TRUE}

# ANTI JOIN VA ON GRADING ----
# eye groups to loop over
eyes_buckets <- split(joined$patID_eye, cut(seq(joined$patID_eye), detectCores()-1, F))

# all rows in va not in grading
va_anti_grading_l <- mclapply(eyes_buckets, function(eyes){
  fuzzy_anti_join(filter(new_dfs$VAA_lea, patID_eye %in% eyes), filter(grading_diagnosis_info_joined_clean_coalesced, patID_eye %in% eyes),
                by = c(
                      "patID_eye" ="patID_eye",
                      "start_DATE"="DATE_coalsced",
                      "end_DATE"="DATE_coalsced"
                    ),
                    match_fun = list(`==`, `<=`, `>=`)
                    )
}, mc.cores = detectCores()-1)

# df of va records with no match in grading
va_anti_grading <- do.call(rbind,va_anti_grading_l)

# View(va_anti_grading)

table(va_anti_grading$visitId) %>% sort()
# 000S  054  078  090  120  132  066  030  042  084  036  060  048  072  024  012  018  000  006 
#    2    2    2    2    2    2    4    6    6    6    8    8   12   16   22   44  156  200  493 
# ^ lots of non 12 interval visits: 006,018

#+ e.g 001001_1 in va_anti_grading
#+ 
# patID_eye   visitId   DATEVIS      DATE_VAA_lea   DATE_VAA_lea_imputed   start_DATE   end_DATE     
# 001001_1    006       2007-03-05   2007-03-05     FALSE                  2007-02-15   2007-03-23   
#+
#+ versus 001001_1 in joined
# patID_eye   patID    eye_num   visitID   DATE_coalsced  
# 001001_1    001001   1         000       2006-09-04      
# 001001_1    001001   1         012       2007-09-09      
# 001001_1    001001   1         024       2008-09-18      
# 001001_1    001001   1         036       2009-11-05      
# 001001_1    001001   1         048       2010-11-05      
# 001001_1    001001   1         060       2011-11-05      
# 001001_1    001001   1         072       2012-09-18      
# 001001_1    001001   1         084       2013-09-09      
#+
#+ ^^^ these visits don't appear in grading data

# FOR RECORDS THAT DIDN'T JOIN, ADD WITH LARGER JOIN DAY WINDOW----
#+ joining the anti remainder from above to the already joined dataset 
#+ find what day window decreases number of non-joining records
#+ 

# # FIND OPTIMAL WINDOW ----
# rws_jnd <- c()
# sq <- seq(20,300,50)
# 
# eyes_buckets_x <- split(va_anti_grading$patID_eye, cut(seq(va_anti_grading$patID_eye), detectCores()-1, F))
# 
# for(i in sq){
#   
#   va_anti_grading_fuzzy_window = i
#   
#   va_anti_grading <- va_anti_grading %>% 
#     mutate(start_DATE = ymd(DATE_VAA_lea) - days(va_anti_grading_fuzzy_window), 
#            end_DATE = ymd(DATE_VAA_lea) + days(va_anti_grading_fuzzy_window)
#            )
#   
#   # find all rows in va not in joined
#   va_anti_grading_l_x <- mclapply(eyes_buckets_x, function(eyes){
#     fuzzy_anti_join(filter(va_anti_grading, patID_eye %in% eyes), filter(grading_diagnosis_info_joined_clean_coalesced, patID_eye %in% eyes),
#                   by = c(
#                         "patID_eye" ="patID_eye",
#                         "start_DATE"="DATE_coalsced",
#                         "end_DATE"="DATE_coalsced"
#                       ),
#                       match_fun = list(`==`, `<=`, `>=`)
#                       )
#   }, mc.cores = detectCores()-1)
#   
#   va_anti_grading_x <- do.call(rbind,va_anti_grading_l_x)
#   
#   # sprintf("%i more rows joined",nrow(va_anti_grading)-nrow(va_anti_grading_x) )
#   rws_jnd <- c(rws_jnd,nrow(va_anti_grading)-nrow(va_anti_grading_x))
# }
# plot(cbind(i = sq,rws_jnd = rws_jnd), type = "b")
# dat <- cbind(i = sq,rws_jnd = rws_jnd) %>% as.data.frame()
# 
# mod <- mgcv::gam(rws_jnd~ s(i, k = 15), data = dat)
# 
# line_d <- data.frame(i = seq(20, 300, 1))
# line_d$rws_jnd <- predict(mod, line_d)
# 
# lines(line_d, col = "blue")
# abline(v = seq(1, 300, 5), lty = 3)


va_anti_grading$sample <- with(va_anti_grading,paste(patID_eye, DATE_VAA_lea, sep = "_"))

setdiff(va_anti_grading$sample, with(dfs$CF_FFA, paste(patientId,eye, PHOTODT_CF, sep = "_")) ) %>% length()
setdiff(va_anti_grading$sample, with(dfs$FAFGRADE, paste(patientId,eye,SCANDTFAF , sep = "_")) ) %>% length()

length(va_anti_grading$sample)

```


##### Duplicate records
```{r}

# where more than one eye, date combo
joined_duplicates <- joined %>% 
  group_by(patID_eye, DATE_coalsced) %>% 
  filter(n()>1) %>% 
  ungroup()
nrow(joined_duplicates)
# ^ 0

if(nrow(joined_duplicates)!=0){

joined_duplicates_match <- sapply(unique(joined_duplicates$patID_eye), function(eye){
  eye1 <- joined_duplicates[joined_duplicates$patID_eye==eye, ][1, ]
  eye2 <- joined_duplicates[joined_duplicates$patID_eye==eye, ][2, ]
  eye1 <- as.character(eye1)
  eye2 <- as.character(eye2)
  eye1[is.na(eye1)] <- "NA"
  eye2[is.na(eye2)] <- "NA"
  eye1 == eye2
})
rownames(joined_duplicates_match) <- colnames(joined_duplicates)

rownames(joined_duplicates_match) <- colnames(joined_duplicates)

joined_duplicates %>% 
  select(patID_eye, visitID, DATE_coalsced, CYLPM, SPHPM) 
  # View()
# ^ no pattern, just choose first 

# are these duplicated in source dfs?
grading_diagnosis_info_joined_clean_coalesced %>% 
  group_by(patID_eye, DATE_coalsced) %>% 
  filter(n()>1) 

x <- new_dfs$VAA_lea %>% 
  group_by(patID_eye, DATE_VAA_lea) %>%
  arrange(patID_eye, DATE_VAA_lea) %>%
  filter(n()>1) 
glimpse(x)
# ^ unique because of visit Id

# remove duplicate rows 
joined <-joined  %>% 
  filter(!duplicated(across(c(patID_eye, DATE_coalsced))))
}
  
```

##### how do MT vars overlap
```{r}

sum(!is.na(joined$stratus_MT01) & !is.na(joined$spectralis_MT01))
# 377 >> 249 >> 381 (looped fuzzy join) >> 383
sum(!is.na(joined$cirrus_MT01) & !is.na(joined$spectralis_MT01)) 
# 339 >> 233 >> 341 >> 345
sum(!is.na(joined$stratus_MT01) & !is.na(joined$cirrus_MT01))
# 235 >> 88 >> 235 >> 237
sum(!is.na(joined$stratus_MT01) & !is.na(joined$cirrus_MT01)& !is.na(joined$spectralis_MT01))
# 29 >> 29 >> 29 >> 31


# look at dimensions  
n_patIDs <- joined %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  pull(patID ) %>% 
  n_distinct(na.rm = TRUE)

n_patID_eyes <- joined %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  select(patID, eye_num) %>% 
  n_distinct(na.rm = TRUE)

n_observations <- joined %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  n_distinct(na.rm = TRUE)

n_patIDs_w_gr_1_vis <- joined %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  filter(n()>1, .by = c(patID, eye_num)) %>% 
  pull(patID) %>% 
  n_distinct(na.rm = TRUE)

cbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis)
#  n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,]     4951         9899          15287                 550

### MARCH 10 2025 ###
#     n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,]     5289        10578          15964                 551

### 11 MARCH 2025 ### changed left_joins to full_joins
#  n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,]     5289        10578          15964                 551

dim(joined)
# 15321   281 >>  15966   270


md_patt <- mice::md.pattern(grading_diagnosis_info_joined_clean_coalesced,plot = FALSE)
dim(md_patt)

# remove counts
m <- md_patt[-nrow(md_patt),-ncol(md_patt)]

Heatmap(m,
        # col = colorRampPalette(c("white", "grey", "black"))(1000) ,
        col = c('white', "black"),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        use_raster = TRUE
        )


# # Define a smoothing kernel (Gaussian-like or simple mean)
# kernel <- matrix(1/25, nrow = 5, ncol = 5)  # 3x3 averaging kernel
# # kernel <- matrix(c(0, -1,  0,
# #                  -1,  5, -1,
# #                  0, -1,  0), nrow = 3, ncol = 3)  # 3x3 sharpening kernel
# 
# # Apply the convolution
# reduced_matrix <- EBImage::filter2(m, kernel)
# reduced_matrix<- round(reduced_matrix)
# 
# # Optionally downsample rows to further reduce size
# downsampled_matrix <- reduced_matrix[seq(1, nrow(reduced_matrix), by = 10),]
#                                      # seq(1, ncol(reduced_matrix), by = 2)]
# Heatmap(reduced_matrix, use_raster = T, cluster_columns = F, cluster_rows = F, col = c("white", "black"))

```

#______________________
# cleaning joined df
###  fix duplicated cols and exclude redundant ones
```{r}

grep( "\\.x", names(joined),value = T)

# new df
joined_clean <- joined

# # rename some variables
# names(joined) <- str_remove_all(names(joined), "\\.x")
# 
# # # lots of duplicate columns; where there was > 1 .x or .y
# rm_cols <- grep("\\.[xy]" , names(joined), value = TRUE) %>%
#   sort()
# 
# # exclude from joined
# joined_clean <- joined_clean[, -which(names(joined_clean) %in% rm_cols)]

# look at dimensions  
n_patIDs <- joined_clean %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  pull(patID ) %>% 
  n_distinct(na.rm = TRUE)

n_patID_eyes <- joined_clean %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  select(patID, eye_num) %>% 
  n_distinct(na.rm = TRUE)

n_observations <- joined_clean %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  n_distinct(na.rm = TRUE)

n_patIDs_w_gr_1_vis <- joined_clean %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  filter(n()>1, .by = c(patID, eye_num)) %>% 
  pull(patID) %>% 
  n_distinct(na.rm = TRUE)

cbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis)
#  n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,] 4951         9899          15287                 550

### MARCH 10 2025 ###
#      n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,]     5289        10578          15964                 551

dim(joined_clean)
# 15321   274 >> 15966   267

```

### remove duplicated rows
```{r}

joined_clean %>% 
  group_by(patID, eye_num, visitID, DATE_coalsced) %>% 
  filter(n()>1)
# ^ NONE

# 
# # remove records where first of two and duplicated
# ## first== false, duplicated == true:> keep first duplicated rows and everything else
# joined_clean <- joined_clean %>% 
#     group_by(patID, eye_num, visitID, DATE_coalsced ) %>% 
#     mutate(duplicated_again = ifelse(n()>1, TRUE, FALSE)) %>% 
#     mutate(first_pat_eye_DATE = row_number() == 1) %>% 
#     ungroup() %>%
#     filter(!(duplicated_again & !first_pat_eye_DATE) )


# look at dimensions  
n_patIDs <- joined_clean %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  pull(patID ) %>% 
  n_distinct(na.rm = TRUE)

n_patID_eyes <- joined_clean %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  select(patID, eye_num) %>% 
  n_distinct(na.rm = TRUE)

n_observations <- joined_clean %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  n_distinct(na.rm = TRUE)

n_patIDs_w_gr_1_vis <- joined_clean %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  filter(n()>1, .by = c(patID, eye_num)) %>% 
  pull(patID) %>% 
  n_distinct(na.rm = TRUE)

cbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis)
#  n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,] 4951         9899          15287                 550

### MARCH 10 2025 ###
#      n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,]     5289        10578          15964                 551

dim(joined_clean)
# 15287   276 >> 15966   269
```

### collapse studyIDs to one
```{r}

joined %>%
  select(matches("stu")) 


# joined <-  joined %>% 
#   # convert char vector to symbols and splice them together, then coalesce those columns together
#   mutate(STUDYID = coalesce(
#     !!!syms(
#       grep("STUDYID", names(joined), value = TRUE)
#       )
#     )
#   )

```

## remove useless columns
```{r}

joined_clean[,grep("COMM|SIGSTR|MANREF|SCAN_TYPE|PHOTOQUAL|QUALITY", names(joined_clean), value = TRUE)] <- NULL

```

## rename misnamed variables
```{r}

names(joined_clean)[which(names(joined_clean) == "ICLO3")] <- "ICL03"

```

## make DIAGNOSIS values lowercase
```{r}
joined_clean$DIAGNOSIS %>% table()
joined_clean$DIAGNOSIS <- tolower(joined_clean$DIAGNOSIS) 

```

## clean date columns
```{r}

date_cols <- grep("DT", names(joined_clean), value = TRUE)

joined_clean[ , date_cols] %>% 
  map(~{x = .
    any(grepl("T|Z", x))
  }) %>% any()
# ^ none 

# # remove characters from dates
# joined_clean[ , date_cols] <- lapply(joined_clean[, date_cols], function(column){
#   # replace T with " " and Z with "" and make date type
#   column <- gsub("T", " ", column) %>% 
#     gsub("Z", "", .) %>% 
#     as.Date(.)
#   
#   return(column)
# })


```


## check data if eye_num is missing
```{r}

table(joined_clean[["eye_num"]], useNA = "always")

# count missing where no eye
count.eye.missing <- colSums(!is.na(joined_clean[is.na(joined_clean$eye_num) , ]))

names(count.eye.missing)[count.eye.missing >0]


```
> no no eye num missing

<!-- ### remove data where eye_num is missing -->
<!-- ```{r} -->

<!-- joined_clean[is.na(joined_clean$eye_num) , ] -->
<!-- # joined_clean <- joined_clean[!is.na(joined_clean$eye_num) , ] -->

<!-- ``` -->
<!-- > no obs with missing eye_num -->

## remove patients that only have information data
```{r}
# non information vars(structural variables)
str_vars <- 
  grep("[_A-TVWXYZ]DATE|^DATE|DT|CHANGES|RW0",names(joined_clean),  value = TRUE, invert = TRUE) %>% 
  grep("\\D0\\d|ICL|ABNOR|AF|EZ|VASC|YELL|CTR|SPC|DET|QUAL|ATRO|VES|CRYST|ISOS|RET|HRD|FAZ|RAV|ONL|VOL|BLR|CENTER|HR|RPE",.,  value = TRUE) 

setdiff(names(joined_clean), str_vars)

#### WHERE MISSING STRUCTURE VARIABLES ####
# rows where str vars missing  
indx <- rowSums(!is.na(joined_clean[,str_vars])) == 0 
sum(indx)

## look at these patients data
joined_at_indx <-  joined_clean[indx , ]

# 1 for observed 0 for NA
m_binary <-  joined_at_indx %>% 
  select(all_of(str_vars), everything()) %>% 
  mutate(across(everything(), ~as.numeric(!is.na(.)))) %>% 
  as.matrix()

# what vars available when structural vars not 
colnames(m_binary)[colSums(m_binary) !=0] 
#^ only VAA famstatus diagnosis, scandates : 

Heatmap(m_binary,
        show_row_names = FALSE,
        name = "observed",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        col = c("white", "black"),
        column_names_gp = gpar(fontsize  = 5),
        )
# ^ just VAA, DEM and dates

# look at which var in df from list has more data than in join
missing_summ <- sapply(str_vars, function(var){
  # search for var
  var_pattern <- sprintf("^%s$", var)
  # in dfs list to find what df has it
  df_match <- sapply(new_dfs, function(d) any(grepl(var_pattern, names(d)) )  ) 
  # if df has var
  if(any(df_match)){
    # grab df name
    df_name <- names(df_match)[df_match]
    # select df
    df <- new_dfs[[df_name]]
    # summary of observed count in df and joined
    c(df = sum(!is.na(df[[var]])), 
         joined = sum(!is.na(joined[[var]]))
         )
  }
})

# difference between missing in df vs joined
missing_summ %>% 
  sapply(function(x){
    x[2]-x[1]
  }) %>% 
  unlist() %>% 
  Filter(function(n) n!=0,.)
# CENTER_MICRON.joined CENTERMAX_MICRON.joined CENTERMIN_MICRON.joined       BLRPATTERN.joined 
#                   -1                      -1                      -1                       4 
# ^ 1 fewer observation of CENTER_MICRON in joined than in df;  BLRPATTERN has 4 more in joined

# find if values match those in og dfs
samples <- with(joined_clean[indx, ], paste(patID_eye, DATE_coalsced, sep = "_"))


# pattern of dates in dfs
DATE_pattern <- paste(paste("DATE", names(new_dfs), sep = "_"), collapse = "$|^") %>% 
  sprintf("^%s$", .)

# filter df from list and joined for samples with missing above; select the intersection of columns and sum na for each sample in each df

df <- new_dfs$VAA_lea
sum_missing <- sapply(new_dfs, function(df){
  date_col <- grep(DATE_pattern, names(df), value = TRUE)
  
  if(all(c("patientId", "eye", date_col)%in% names(df))){
    df$my_DATE <- df[, date_col, drop = T]
    df_samps <- df %>% 
      mutate(sample = paste(patientId, eye, my_DATE, sep = "_")) %>% 
      # filter(sample%in%samples) %>% 
      select(sample,any_of(names(joined_clean))) 
    
    joined_samps <- joined_clean %>% 
      mutate(sample = paste(patID_eye, DATE_coalsced, sep = "_")) %>% 
      # filter(sample%in%samples) %>% 
      select(sample,any_of(names(df))) 
    
    # for each sample, sum missing in df and joined
    samp_nas <- sapply(samples, function(samp){
      df_na = sum(is.na( df_samps[df_samps$sample == samp,]))
      joined_na = sum(is.na( joined_samps[joined_samps$sample == samp,]))
      
      return(c(df_na= df_na, joined_na=joined_na, joined_mns_df = joined_na - df_na))
    })
    return(t(samp_nas))
  }
})

# sum_missing[1] %>% 
#    map(~{
#     x <- .
#     diff <- x[,"joined_mns_df"]
#     diff_sd <- sd(diff)
#     diff_mean <- mean(diff)
#     
#     # # how many have out of ordinary count of more missing
#     # sum( abs(diff - diff_mean) > diff_sd)
#     
#     # c(diff_mean  = diff_mean, diff_sd =   diff_sd)
#     
#     plot(density(x), xlim = c(-20,105),col = rgb(0,0,0,.5), ylim = c(0, 1))
#   })

# df missing versus joined missing as difference summary statistics
sum_missing_summ <- sum_missing %>% 
  map(~{
    x <- .
    diff <- x[,"joined_mns_df"]
    diff_sd <- sd(diff)
    diff_mean <- mean(diff)
    diff_median <- median(diff)

    # how many have out of ordinary count of more missing
    count_more <- sum( abs(diff - diff_mean) > diff_sd)
    count_less <- sum( abs(diff - diff_mean) < diff_sd)

    c(diff_mean  = diff_mean, 
      diff_median = diff_median,  
      diff_sd =   diff_sd, 
      count_less = count_less,
      count_more = count_more
      ) %>% round(2)
    
    # if(is.numeric(x)){
    #   lines(density(x), col = rgb(0,0,0,.5))
    # }
  }) %>% 
  do.call(rbind,.)
sum_missing_summ
### MARCH 11 2025 ###
#                 diff_mean diff_median diff_sd count_out_of_ord
# CF_FFA               2.00           2    0.00                0
# DEM                    NA          NA    0.00               NA
# DIAGNOSIS            3.03           3    0.17               24
# ENRNHOR              6.42           6    1.18              282
# ENRNHOS              6.08           6    0.84              325
# FAFGRADE             3.63           3    2.69               45
# FAMSTATUS            4.20           4    0.58              300
# MH                  17.21          21    7.07              152
# OCTMAIN              4.11           3    4.69               45
# OCTMTCIRRUS          5.17           2    6.14              182
# OCTMTSPECTRALIS      6.08           2    9.45              135
# OCTMTSTRATUS        18.20          18    0.58              300
# VAA_lea             36.54          38    4.64               95
# ENR                    NA          NA    0.00               NA

### `` ###
#                 diff_mean diff_median diff_sd count_out_of_ord
# CF_FFA              44.66           2   45.57              752
# DEM                    NA          NA    0.00               NA
# DIAGNOSIS            2.40           2    0.62              117
# ENRNHOR              3.57           6    7.60              142
# ENRNHOS              5.03           5    1.64              325
# FAFGRADE            10.14           3    7.58              852
# FAMSTATUS            4.44           5    0.76              166
# MH                   7.93          21   30.49              130
# OCTMAIN             14.56           3   12.06              852
# OCTMTCIRRUS         -5.45           2   36.06              214
# OCTMTSPECTRALIS      4.69           2   42.61               99
# OCTMTSTRATUS        -4.20          17   30.63              349
# VAA_lea             23.32          25   14.32              891
# ENR                    NA          NA    0.00               NA
# > 

# m <- fread(" ") 
# 
# mat <- as.matrix(m[,-1])
# rownames(mat) <- m$df
# 
# cbind(sum_missing_summ, x = NA, mat)

```


### remove rows that only VAA, DEM, DIAGNOSIS, FAMSTATUS data and nothing else... so useless :(
```{r}

# remove rows that only VAA, DEM, DIAGNOSIS, FAMSTATUS data and nothing else... so useless :(
joined_clean_new <- joined_clean[!indx, ] # subset for rows not missing structure vars
sum(indx) 
# 6 >> 3208 >> 163 >> 0 >> 847 rows with no str var observed

# how many  obs with 0 str vars observed in each df
ratio_observed <- map(dfs, ~{x <-.
  if(any(str_vars%in% names(x))){
    x <-   select(x, any_of(str_vars))
    # how many  obs with 0 str vars observed
    sum(rowSums(!is.na(x))==0)
  }
})

ratio_observed
names(dfs$DIAGNOSIS)[names(dfs$DIAGNOSIS)%in% str_vars]

# # image dattaset after removing rows
# m1 <- joined_clean_new %>%
#   mutate(across(everything(), ~as.numeric(!is.na(.)))) %>%
#   as.matrix()
# 
# 
# Heatmap(m1,
#         show_row_names = FALSE,
#         name = "observed",
#         cluster_rows = FALSE,
#         cluster_columns = FALSE,
#         col = c("white", "black"),
#         column_names_gp = gpar(fontsize  = 5),
#         )

n_patIDs <- joined_clean_new %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  pull(patID ) %>% 
  n_distinct(na.rm = TRUE)

n_patID_eyes <- joined_clean_new %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  select(patID, eye_num) %>% 
  n_distinct(na.rm = TRUE)

n_observations <- joined_clean_new %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  n_distinct(na.rm = TRUE)

n_patIDs_w_gr_1_vis <- joined_clean_new %>% 
  select(patID, visitID, eye_num) %>% 
  drop_na() %>% 
  filter(n()>1, .by = c(patID, eye_num)) %>% 
  pull(patID) %>% 
  n_distinct(na.rm = TRUE)

cbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis)
#  n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,]     4951         9899          15287                 550

### MARCH 10 2025 ###
#     n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,]     5248        10486          15204                 534

# ### `` ###
#      n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,]     5213        10413          15117                 533

dim(joined_clean_new)
# 15287   263 >> 15119   259

```


# count observed barplot
```{r}

colSums(!is.na(joined_clean_new)) %>% 
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  rename(count_present = "." ) %>% 
  ggplot(aes(x = variable, y = count_present, fill = variable))+
  geom_bar(stat = "identity", show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Count Non-Missing")

```

# Save Dataframe
```{r}

# make file name for dfs in file
fileName <-paste(names(dfs) , collapse =   "_") %>% paste0("_01.rds")
print(paste("SAVING...    ", fileName))
saveRDS(joined_clean_new, paste0(wd_path, "processed_data/", fileName))

```
CF_FFA_DEM_DIAGNOSIS_ENR_FAFGRADE_FAMSTATUS_MH_dense_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_01.rds

# save raw dfs list
```{r}

saveRDS(new_dfs, paste0(wd_path , "processed_data/dfs_list_01.rds"))

```
"~/Documents/LMRI/PRJ2024001/version015/processed_data/dfs_list_01.rds"


```{r}
m <- joined_clean_new %>% 
  select(stratus_MT01, spectralis_MT01, cirrus_MT01) %>% 
  mutate_all(~as.numeric(.)) %>% 
  as.matrix()
sum(rowSums(!is.na(m)) > 0) 
# 13091 >> 10037 >> 12922( removing duplicates after fuzzy join) >> 13769
#rows of mt data theoretically

sum(rowSums(!is.na(m)) == 3)
# 29 >> 29 >> 29( removing duplicates after fuzzy join) >>31
# overlap between all three machines

sum(rowSums(!is.na(m)) >1)  
# 893 >> 572 >> 899 >> 891( removing duplicates after fuzzy join)>>903
# where overlap between any two

sum(rowSums(!is.na(m[, c(1,2)])) == 2) 
# 377 >> 269 >> 381 >> 375( removing duplicates after fuzzy join)>>383
# overlaps between strat and spec

sum(rowSums(!is.na(m[, c(1,3)])) == 2) 
# 235 >> 93 >> 235 >> 235( removing duplicates after fuzzy join)>>237
# overlaps between strat and cirrus

sum(rowSums(!is.na(m[, c(2,3)])) == 2) 
# 339 >> 268 >> 341 >> 339 ( removing duplicates after fuzzy join)>>345
# overlaps between spec and cirrus
```


