---
title: "02.1_visit and missingness analysis"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---
# options
```{r}
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
```

# Packages

```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

pacman::p_load(tidyr,dplyr,ggplot2,reshape2,tidyverse,ggpubr,vcd,UpSetR,gridExtra,Metrics,purrr,corrplot,RColorBrewer,openxlsx,networkD3,htmlwidgets,lubridate,ComplexHeatmap,ggrepel,plotly, parallel, fuzzyjoin)

rm(list = ls())
gc()
par(mfrow = c(1,1))
```

# Load Data

```{r}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"
list.files(sprintf('%s/processed_data', wd_path))

data_name <- "CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_01.rds"

# joined data full
d <- readRDS(paste0(wd_path,"processed_data/",data_name))

# list of raw dfs with minimal cleaning
raw <- readRDS(paste0(wd_path,"processed_data/dfs_list_01.rds"))

# my functions
source(sprintf("%sanalysis/00_MyFunctions.R", wd_path))

```


# make new directory for plots
```{r}
path <- paste0(wd_path, "results/visit_missingness_analysis/")
if(!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
  
}

```

```{r, look at dims}

check_dims <- function(){
  
  cat(paste0("\ndims : ", paste(dim(d), collapse = " "),"\n"))
  
  n_patIDs <- d %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- d %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- d %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- d %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()

#   n_patIDs n_patID_eyes n_observations n_patIDs_w_gr_1_vis
# [1,]     4951         9899          15287                 550
# 
# dims : 15287 264

### MARCH 11 2025 ###
# dims : 15119 259
#                      [,1]
# n_patIDs             5213
# n_patID_eyes        10413
# n_observations      15117
# n_patIDs_w_gr_1_vis   533
# > 

```

##### looka at va that doesn't join to others (va anti d)

<!-- ```{r} -->

<!-- # eye groups to loop over -->
<!-- eyes_buckets <- split(unique(d$patID_eye), cut(seq(unique(d$patID_eye)), detectCores()-1, F)) -->

<!-- VAA <- raw$VAA_lea %>%  -->
<!--     mutate(start_DATE = ymd(DATE_VAA_lea) - days(18), -->
<!--            end_DATE = ymd(DATE_VAA_lea) + days(18) -->
<!--            ) -->

<!-- # all rows in va not in grading -->
<!-- va_anti_grading_l <- mclapply(eyes_buckets, function(eyes){ -->
<!--     fuzzy_anti_join(filter(VAA, patID_eye %in% eyes), filter(d, patID_eye %in% eyes), -->
<!--                 by = c( -->
<!--                       "patID_eye" ="patID_eye", -->
<!--                       "start_DATE"="DATE_coalsced", -->
<!--                       "end_DATE"="DATE_coalsced" -->
<!--                     ), -->
<!--                     match_fun = list(`==`, `<=`, `>=`) -->
<!--                     ) -->
<!-- }, mc.cores = detectCores()-1) -->

<!-- # df of va records with no match in grading -->
<!-- va_anti_grading <- do.call(rbind,va_anti_grading_l) -->
<!-- va_anti_grading <- reorder_id_cols(va_anti_grading) -->




<!-- ``` -->


## process data dicitonary

### LOAD AND CLEAN INFO 
```{r}
# load info
data_dict_raw <- read.xlsx("/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/raw_data/data_dictionaries/database_grading_key.xlsx")

names(data_dict_raw) <- snakecase::to_snake_case(names(data_dict_raw))

data_dict_raw$field[duplicated(data_dict_raw$field)]

# remove right/left labels; filter for distinct rows; remove duplicated fields
data_dict <- data_dict_raw %>%
  as_tibble() %>% 
  distinct() %>% 
  filter(!duplicated(field))

# remove extra number in RW vars (specifies chart)
data_dict$field <- gsub("^RW(.*?)[12]$","RW\\1" , data_dict$field)

# move zone in mt sd vars
data_dict$field[grep("[iu]s.*SD", data_dict$field)]  <- gsub("MT_SD(\\d{2})", "MT\\1_SD", data_dict$field[grep("[iu]s.*SD", data_dict$field)] )
```


### JOIN INFO FIELD TO D NAMES 
```{r}
names_to_match <- grep("start|end|ENR|pat|eye|vis|^DATE|DT", names(d), invert = TRUE, value = TRUE)

# Compute string distances
dist_matrix <- stringdist::stringdistmatrix(names_to_match, data_dict$field, method = "qgram")

# Find closest match
closest_match <- apply(dist_matrix, 1, which.min)
closest_match_dist <- apply(dist_matrix, 1, min)

# Set a reasonable distance threshold (adjust as needed)
threshold <- 4

# Assign only if distance is below threshold, else assign NA
best_matches <- ifelse(closest_match_dist <= threshold, data_dict$field[closest_match], NA)

# Create a new dataframe
info <- data.frame(
  name = names_to_match,
  matched_name = best_matches,
  dist = closest_match_dist,
  stringsAsFactors = FALSE
) 


info <- info %>% 
  left_join(data_dict, by = join_by(name == field)) 

#  add remainging names in d 
remaining_names <- setdiff(names(d),info$name)
add <- data.frame(matrix(rep(NA,ncol(info)*length(remaining_names)), ncol = ncol(info), dimnames = list(NULL, names(info))))
add$name = remaining_names
info <- info %>% 
  rbind(add)

# remove duplicates
info <- distinct(info)

info$table[grepl("spectralis|CENTER", info$name, ignore.case = T) & is.na(info$table)] <- "OCTMTSPECTRALIS"
info$table[grepl("stratus", info$name, ignore.case = T) & is.na(info$table)] <- "OCTMTSTRATUS"
info$table[grepl("cirrus", info$name, ignore.case = T) & is.na(info$table)] <- "OCTMTCIRRUS"
info$table[grepl("CF|FFA|ICL", info$name) & is.na(info$table)] <- "CF_FFA"
info$table[grepl("OCTMAIN", info$name) & is.na(info$table)] <- "OCTMAIN"
info$table[grepl("FAF", info$name) & is.na(info$table)] <- "FAFGRADE"
info$table[grepl("VAA|VAP|ACUIT", info$name) & is.na(info$table)] <- "VAA"
info$table[grepl("DM|[WH]GT|SMK|MH|BLR|DIAG", info$name) & is.na(info$table)] <- "MH/DIAGNOSIS"
info$table[grepl("RACE|ETHNI|GEND|YR|ENR|pat|FAM|NHO[SR]|DATE|DT|vis|eye|SITE", info$name) & is.na(info$table)] <- "DEM/ENR/FAM/other"


# fill in table description
table_descriptions <- sapply(unique(na.omit(info$table)), function(x) unique( na.omit(info$table_description[info$table==x]))) %>% 
  Filter(function(x) !is_empty(x),.)

for(tab in names(table_descriptions)){
  info$table_description[info$table==tab] <- table_descriptions[[tab]]
}



# join old info table
old_info  <- read.xlsx("/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/raw_data/data_dictionaries/database_grading_key_old.xlsx")

# each old's value in new
old_to_new_matches <- sapply(old_info$shorthand, function(pttrn)grep(pttrn, info$name)) %>% 
  sapply(., function(inds) info$name[inds])

# make space to add old ot new
add <- old_info[1,] %>% select(shorthand:ncol(.))
add[1,] <- NA
add <- add[rep(1,nrow(info)) , ]
rownames(add) <- NULL
info <- cbind(info, add)

# add old info  to new at each match
for(i in seq(old_to_new_matches)){
  old <- names(old_to_new_matches)[i]
  new <- old_to_new_matches[[i]]
  
  # info to add at cols
  rows_add <- old_info %>% 
    filter(shorthand==old) %>% 
    select(shorthand:ncol(.))
  
  # get to right dim
  rows_add <- rows_add[rep(1,length(new)), ]
  
  # add to info
  info[info$name %in% new , names(add)] <- rows_add
}

  
```

### add ordinality to info df
```{r}

# # write vars and manually label ordinal
# writeLines(info[info$ordinal , "name", drop = T][order(info$table)], sprintf("%sprocessed_data/df_vars.txt", wd_path))

# load ordinal txt file
ordinality <- read_delim(sprintf("%sprocessed_data/df_vars.txt", wd_path), delim = ",")
# ordinality %>% 
#   arrange(ordinal) %>% 
#   print(n = 300)

info$ordinal <- NULL
# add ordinality to info
info <- info %>% 
  left_join(ordinality, by = "name")

info$table[grepl("GRADERID", info$name)] <- "GRADING"

```



#_______________
# clean data
```{r}

### make a binary diagnosis field 
#### 1 if yes, 0 if cannot be determined or no for diagnoses not NA
d$DIAGNOSIS_binary<- NA
d$DIAGNOSIS_binary[!is.na(d$DIAGNOSIS)] <- ifelse(d$DIAGNOSIS[!is.na(d$DIAGNOSIS)] == "yes", 1, 0)

d %>% select(matches("DIAGNOSIS")) %>% unique()
d %>% 
  select(matches("DIAGNOSIS")) %>%
  group_by(DIAGNOSIS) %>% 
  summarise(count=  n())
# DIAGNOSIS            count
# 1 cannot be determined   264
# 2 no                    2339
# 3 yes                  12602
# 4 NA                      70
```

## remove records DIAGNOSIS_binary == 0
```{r}

d <- d %>%
  filter(DIAGNOSIS_binary== 1)

```

```{r, look at dims}
check_dims()
# dims : 12602 265
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12602
# n_patIDs_w_gr_1_vis   516
```

```{r}
sum(  ( !is.na(d$cirrus_MT01) |  !is.na(d$stratus_MT01)  ) &  !is.na(d$spectralis_MT01)) 
# 157 >> 583  >>575
sum(!is.na(d$cirrus_MT01) &  !is.na(d$stratus_MT01) ) 
# 197 >> 197 >>197
sum(!is.na(d$spectralis_MT01) &  !is.na(d$stratus_MT01) ) 
# 92 >> 297>> 291
sum(!is.na(d$spectralis_MT01) &  !is.na(d$cirrus_MT01) ) 
# 73 >> 294>> 292
```

## any duplicates?
```{r}

d %>% 
  group_by(patID, eye_num, visitID, DATE_coalsced) %>% 
  filter(n()>1)

```

```{r, look at dims}
check_dims()
# dims : 12602 265
#                      [,1]
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12602
# n_patIDs_w_gr_1_vis   516
```

## try numeric
```{r}

d <- d %>% 
  mutate(across(-matches("pat|vis"), as_numeric_if_possible))

```


#__________________
## How distributed are patients across visits?
### how many patients per visits
```{r }

# plot count of patients at each visit
available_visit_count_d <- d %>%
  group_by(visitID) %>%
  summarize(visit_count = sum(!is.na(visitID)))

available_visit_count_d$visitID <- as.factor(available_visit_count_d$visitID)

# divide by two becasue two records, 1 for each eye, for 1 patient which makes double visitIDs
available_visit_count_d$visit_count <- available_visit_count_d$visit_count / 2

visits_plot <- ggplot(available_visit_count_d, aes(y = visitID, x = visit_count, fill = visitID))+
  geom_bar(orientation = "y", stat = "identity", show.legend = FALSE)+
  geom_text(aes(label = visit_count), vjust = -0.5, hjust = 1, cex = 5) +
  ggtitle("Count of patients at each visit")+
  theme_bw()+
  theme(axis.text.y = element_text(size = 20))
visits_plot

ggsave(paste0(wd_path, "results/visit_missingness_analysis/visits_plot.png") , visits_plot, width = 10 , height = 14, dpi = 200)

``` 
> 000 012 024 036, etc., have most visits--\> 12 month increments

### how many visits per patient, what visits?
```{r}

patient_visit_l <- list()

# for each patient, grab what visits they had
for (pat in unique(d$patID)){

  # unique visits for each patient
  pat_visits <- unique(d$visitID[d$patID == pat & !is.na(d$eye_num)] )

  patient_visit_l[[pat]] <- pat_visits
}

# visualize how many visits for each patient
pat_visit_count_d <- data.frame(patient = rep(NA, length(patient_visit_l)), num_visits = rep(NA, length(patient_visit_l)))


# make a df of patients and the count of how many visits they had
for (i in 1:length(patient_visit_l)){

  pat_visit_count_d[i, ] <- data.frame(patient  = patient_visit_l[i] %>% names(),
                           num_visits = patient_visit_l[[i]] %>% length())

}
pat_visit_count_d <- pat_visit_count_d %>%
  arrange(num_visits)

ggplot(pat_visit_count_d,  aes( x = num_visits))+
  geom_bar()+
  # geom_density( )+
  scale_x_continuous(breaks = rep(0:10, 1))+
  ggtitle("Count of patients having a certain number of visit")



```

## Heatmap of patients x visit
```{r eval = FALSE, include = FALSE}

# for patient of patient_visit_l, make row, if visit exists TRUE else FALSE
m <- matrix(dimnames = list(unique(d$patID), unique(d$visitID)),nrow = length(unique(d$patID)), ncol = length(unique(d$visitID)))

# dataframe to order visitIDs by month number
visitID_order <- data.frame(visitID = unique(d$visitID), 
                            visitID_num = gsub("\\D", "", unique(d$visitID)) %>% 
                                                 as.numeric()
                            ) 
# if empty make NA, arrange by inbcreasing month
visitID_order <- visitID_order %>% 
  mutate(visitID_num  = visitID_num %>% ifelse(.== "", NA, .)) %>% 
  arrange(visitID_num)

vis_levels <- visitID_order$visitID
m <- m[, vis_levels]

# levels of visit order from previous hard-coded ordering
d$visitID <- factor(d$visitID, levels = vis_levels)


## fill matrix with patient's presence at visit
for(i in 1:nrow(m)){
  pat = rownames(m)[i]
  pat_visits <- unique(d$visitID[d$patID == pat])

  for(j in 1:ncol(m)){
    m[i, j] <- ifelse(colnames(m)[j] %in% pat_visits, 1, 0)
  }
}

m <- m %>%
  as.data.frame() %>%
  as.matrix()

pdf(paste0(wd_path, "results/visit_missingness_analysis/patient_presence_at_visit_heatmap.pdf"), height = 10, width = 10)
Heatmap(m,
        show_heatmap_legend = TRUE,
        use_raster = F,
        column_title  = "Patient presence at each visit (1 = present)",
        row_title  = "patient",
        name = "observed",
        col = c("white" ,  "black"),
        na_col = "white",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_row_names = FALSE,
        show_column_names  = TRUE,
        column_names_rot = 45
)
dev.off()

```

## sankey plot showing patient path from visit to next
```{r eval = FALSE, include = FALSE }

# Convert matrix to data frame and gather it into a long format
df_long <- as.data.frame(m) %>%
  rownames_to_column(var = "patient") %>%
  gather(key = "visit", value = "present", -patient)

# Filter out the rows where the patient was not present
df_present <- df_long %>%
  filter(present == 1) %>%
  arrange(patient, visit)

# Create a sequence ID for each patient's visit
df_present <- df_present %>%
  group_by(patient) %>%
  mutate(visit_sequence = row_number()) %>%
  ungroup()

# Create links by joining the table with itself on patient and the sequence
links <- df_present %>%
  left_join(df_present, by = "patient",  relationship = "many-to-many") %>%
  filter(visit_sequence.y == visit_sequence.x + 1) %>%
  select(source = visit.x, target = visit.y, value = 1)

# Aggregate the links to count the number of patients with the same path
links <- links %>%
  group_by(source, target) %>%
  summarise(value = n()) %>%
  ungroup()

# Nodes data frame with unique visits
nodes <- data.frame(name=unique(c(as.character(links$source), as.character(links$target))))

# Assign IDs to nodes
nodes$id <- 0:(nrow(nodes) - 1)

# Convert node names to IDs in the links dataframe
links <- links %>%
  left_join(nodes, by = c("source" = "name"))

links$source_id <- links$id
links$id <- NULL

links <- links %>%
  left_join(nodes, by = c("target" = "name"))

links$target_id <- links$id
links$id <- NULL

links <- links %>%
  select(source = source_id, target = target_id, value)

# The nodes and links dataframes are now ready to be used in a sankey plot function


s <- sankeyNetwork(Links = links, Nodes = nodes ,
                            Source = "source", Target = "target",
                            Value = "value", NodeID = "name")
s <- onRender(s, "
    function(el, x, data) {
      var svg = d3.select(el).select('svg');
      var g = svg.select('.chart');

      // Define zoom behavior
      var zoom = d3.zoom()
        .scaleExtent([1, 8]) // Set scale limits
        .on('zoom', zoomed);

      // Zoom function
      function zoomed() {
        g.attr('transform', d3.event.transform);
      }

      svg.call(zoom);

      // Disable double-click zoom
      svg.on('dblclick.zoom', null);
    }
  ")

# Save to an HTML file if needed
saveNetwork(s, file = paste0(wd_path, 'results/visit_missingness_analysis/sankeyNetwork.html'))
```

## do patients who have mactel ever not have mactel at past visits?
is diagnosis consistent?

### each patient's diagnosis at each visit
```{r}

# # max diagnosis of each patient at each visit
# diagnosis_m <- d %>%
#   select(visitID, patID, DIAGNOSIS_binary) %>%
#   group_by(patID, visitID) %>%
#   summarize(diagnosis = max(DIAGNOSIS_binary)) %>%
#   ungroup() %>%
#   pivot_wider(names_from = visitID, values_from = diagnosis, values_fill = list(diagnosis = NA)) %>%
#   column_to_rownames( var = "patID")


d %>%
  select(visitID, patID, DIAGNOSIS_binary) %>%
  group_by(patID, visitID) %>%
  summarize(diagnosis = max(DIAGNOSIS_binary)) %>%
  ungroup() %>% 
  ggplot( aes(x = visitID, y = diagnosis, group = patID))+
  geom_line(show.legend = FALSE)+
  theme_bw()

```
> patients that start as 1/0 continue as 1/0

# MT colsums
```{r}
d %>% 
  select(matches("MT0\\d$")) %>% 
  mutate_all(~as.numeric(!is.na(.))) %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rownames_to_column("variable") %>% 
  rename_with(~c("variable", "count_observed")) %>% 
  mutate(machine = gsub("_MT0\\d", "", variable), 
         zone = gsub("\\D", "", variable) ) %>% 
  ggplot(aes(machine, count_observed, fill =  zone)) +
    geom_bar(stat = "identity", position = "dodge")
ggsave(paste0(wd_path, 'results/visit_missingness_analysis/MT_colsums.png'), height = 10, width =10)

```

#______________________
# VisitIDs with new dates

## Baseline visits
```{r}
baseline_pats <- d %>%
 filter(visitID == "Baseline") %>%
 pull(patID)

x <- d[d$patID %in% baseline_pats , ]

# do these patients also have a visit 0?
for(pat in baseline_pats) print(as.character(unique(d$visitID[d$patID == pat])))
#  
```
> no baseline pats in this version of the dataset

### remove baseline visit rows
```{r}
d <- d[d$visitID != "Baseline"  ,  ]

d$visitID %>% levels()
```

## 6month visit
```{r, eval = FALSE}

d[d$visitID == "6 Months", ] %>% glimpse()

d %>% 
  filter(patID =="018002") %>% 
  select(visitID, matches("DT|Date", ignore.case = TRUE)) %>% 
  arrange(visitID) %>% 
  filter(visitID %in% c("006", "6 Months")) %>% 
  glimpse()



m <- d[d$visitID == "6 Months", ] %>%
  mutate_all(~ifelse(is.na(.), 0, 1)) %>%
  as.matrix()


Heatmap(m,
        show_heatmap_legend = TRUE,
        use_raster = F,
        column_title  = "dates",
        # row_title  = "",
        col = c("white" ,  "black"),
        na_col = "white",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_row_names = FALSE,
        show_column_names  = TRUE,
        # column_names_rot = 45,
        # column_names_gp = gpar(fontsize = 3)
        row_names_gp = gpar(fontsize = 3)
)

# look at patient with this visit
m <- d[d$patID == "018002", ] %>%
  mutate_all(~ifelse(is.na(.), 0, 1)) %>%
  as.matrix()

Heatmap(m,
        show_heatmap_legend = TRUE,
        use_raster = F,
        # row_title  = "",
        col = c("white" ,  "black"),
        na_col = "white",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_row_names = FALSE,
        show_column_names  = TRUE,
        # column_names_rot = 45,
        # column_names_gp = gpar(fontsize = 3)
        row_names_gp = gpar(fontsize = 3)
)

# > only dates and DEM for 6 Month visit

# remove 6 Month row

d <- d[d$visitID != "6 Months",]

```

## 000S visit
```{r, eval = FALSE}
d[d$visitID == "000S", ] %>% glimpse()

d <- d[d$visitID != "000S" , ]
```
> no visit 000S in this data


## what are unsch and 084S?
## 084S
```{r, eval = FALSE}
m <- d[d$visitID == "084S", ] %>%
  # select(stratus_SCANDT, SCANDT_period_year) %>%
  mutate(across(everything(),
                ~ifelse(is.na(.), 0,1)
                )) %>%
  filter(rowSums(across(everything())) != 0) %>%
  as.matrix()

Heatmap(m,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        col = c("white", "black"))

```
> not in this version

### removing records at 084S
```{r, eval = FALSE}

d[d$visitID == "084S" , ]
d <- d[d$visitID != "084S" , ]

```
> not in this version


## unsch
```{r}

d[d$visitID %in% grep("unsch", d$visitID, value = TRUE), ]  %>% glimpse()
# d[d$visitID == "000usch", ]   %>% glimpse()


m_d <- d[d$visitID %in% grep("unsch", d$visitID, value = TRUE), ] %>%
  mutate(visitID_unique = make.unique(as.character(visitID))) %>%
  arrange(visitID_unique)

rownames(m_d) <- NULL
rownames(m_d) <- m_d$visitID_unique

m <- mice::md.pattern(m_d, plot = FALSE)
m <- m[-nrow(m), -ncol(m)]

Heatmap(m ,
        show_heatmap_legend = TRUE,
        use_raster = F,
        column_title  = "data of unsch visits (1 = observed)",
        # row_title  = "",
        col = c("white" ,  "black"),
        na_col = "white",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_row_names = TRUE,
        show_column_names  = TRUE,
        # column_names_rot = 45,
        # column_names_gp = gpar(fontsize = 3)
        row_names_gp = gpar(fontsize = 3)
)

```
> MT, cf, and fafgrade data

### all patients' visit history
```{r}

# plot order of visits and dates visited by patients
ggplot(d, aes(x = DATE_coalsced, y = visitID, group = patID)) +
  geom_line(show.legend = FALSE, alpha = .2) +
  geom_point(show.legend = FALSE, alpha = .2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45))+
  scale_x_date()+
  labs(title = "All Patients' Visit Timeline")

```


### unsch patients visit history plot
```{r}

pats_with_unsch <- d$patID[ (d$visitID == "000unsch") & !is.na(d$DATE_coalsced)]

plotting_d <- subset(d, patID %in% pats_with_unsch) %>%
  select(DATE_coalsced, visitID, patID) %>%
  unique()

ggplot(plotting_d, aes(x = DATE_coalsced, y = visitID, group = patID, color = patID)) +
  geom_line(show.legend = FALSE, alpha = 0.4) +
  geom_point(show.legend = FALSE, alpha = 0.4) +
  theme_bw() +
  labs(x = "Visit ID", y = "DATE_coalsced", title = "Patient Visit Timeline for patients with 000unsch") +
  # scale_x_discrete(limits = c(levels(d$DATE_coalsced)[1], levels(d$DATE_coalsced)[length(levels(d$DATE_coalsced))]),
  #               date_breaks = "2 months", date_labels = "%Y-%m") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_label_repel(aes(label = as.character(DATE_coalsced)),
                   nudge_y = 0.1,
                   size =5,
                   angle = 45,
                   max.overlaps = Inf,
                   box.padding = 0.1,
                   show.legend = FALSE,
                   label.size = 0.5
                   )

# ggsave(paste0(wd_path, "results/visit_missingness_analysis/unsch_patients_visits_dates.png"), height = 7, width = 15)
# unsch_ggplotly <- ggplotly(unsch_plot)
# 
# htmlwidgets::saveWidget(unsch_ggplotly, paste0(wd_path, "results/visit_missingness_analysis/unsch_patients_visits_dates.html"))

```
> most 000unsch visits are 6 mo after previous visit ==> change 000unsch to 006 for 000unsch betwen 3 and 9 months away from previous visit

#### change unsch to months from visit 000
```{r}

# PATIENTS WHO HAVE UNSCH VISIT
pats_with_unsch 

# # nth_period_year <- d$SCANDT_period_year %>% levels()
# # first_period_year <- d$SCANDT_period_year %>% levels() 
# ##
# # function to take period_year and find difference in months from first visit
# convert_period_year_to_date <- function(period_year, year_divisor = 19){
#   
#   # parse years and periods from first and nth dates
#   year <- sapply(strsplit(period_year, "-"), "[", 2) %>% as.numeric()
#   period <- sapply(strsplit(period_year, "-"), "[", 1)%>% as.numeric()
# 
#   # convert period to day of year
#   period_day <- period * year_divisor
#   
#   # convert day and year to date
#   date <- as.Date(period_day - 1, origin = paste0(year,"-01-01"))
#   
#   return(date)
# }
# 
# ## TEST PERIOD_DAY CONVERSION AND INTERVAL CALCULATION ## 
# dates.1 <- paste(rep(1:year_divisor, 2), 2000:2039, sep = "-")
# dates.2 <- paste(rep(2:18, 2), 2001:2040, sep = "-")
# data.frame(
#   dates.1 = convert_period_year_to_date(dates.1),
#   dates.2 = convert_period_year_to_date(dates.2),
#   diff_months = interval( convert_period_year_to_date(dates.1), convert_period_year_to_date(dates.2)) %/% months(1)
# )
# ##  ##  ## 

# if unscheduled visit calculate months from first visit date
unsch_new_visitID_d <- d %>%
  select(patID_eye,  patID,visitID, DATE_coalsced) %>%
  unique() %>%
  filter(patID %in% pats_with_unsch) %>%
  arrange(patID, visitID) %>%
  group_by(patID_eye) %>%
  # mutate(first_period_year = dplyr::first(DATE_coalsced)) %>%
  # mutate(new_visitID =
  #          ifelse( visitID %in% grep("unsch", d$visitID, value = TRUE),
  #                  interval( convert_period_year_to_date(as.character(first_period_year)),
  #                            convert_period_year_to_date(as.character(DATE_coalsced))
  #                            ) %/% months(1) ,
  #                  visitID
  #                  )
  #        ) %>% 
  # select(-first_period_year)
  mutate(first_DATE_coalsced = dplyr::first(DATE_coalsced) ) %>% 
  mutate(new_visitID = 
           ifelse( visitID %in% grep("unsch", d$visitID, value = TRUE), 
                   interval(first_DATE_coalsced, DATE_coalsced) %/% months(1), 
                   NA
                   )
         ) %>% 
  ungroup() %>% 
  select(- first_DATE_coalsced)

# look at how they relate
left_join(d, unsch_new_visitID_d) %>% 
  select(patID, eye_num, visitID,new_visitID, DATE_coalsced, matches("DATE")) %>% 
  filter(patID %in% pats_with_unsch) %>% 
  arrange(patID, eye_num, visitID) %>% 
  print(n = 100)
  

# join new visitIDs to main
d <- left_join(d, unsch_new_visitID_d)

# if newVisitID present, set old to new and make 3 char long
d <- d %>%
  # if there's a newvisitID, change original
  mutate(visitID = as.character(visitID)) %>% 
  mutate(visitID = ifelse(!is.na(new_visitID), as.character(new_visitID), visitID)) %>%
  mutate(visitID =
        case_when(
          nchar(visitID) < 2 ~ paste0("00", visitID),
          nchar(visitID) < 3 ~ paste0("0", visitID),
          TRUE ~ visitID
  )) %>% 
  select(-new_visitID)

# update vis_levels
numeric_vis <- d$visitID %>% grep("\\d", ., value = TRUE) %>% unique()
numeric_vis <- numeric_vis[order(numeric_vis)]

nonNumeric_vis <- d$visitID %>% grep("\\D", ., value = TRUE) %>% unique()
nonNumeric_vis  <- setdiff(nonNumeric_vis, numeric_vis)

vis_levels <- c(numeric_vis, nonNumeric_vis)

all(vis_levels %in% d$visitID ) # good if TRUE
all(d$visitID%in% vis_levels )  # good if TRUE

d$visitID <-  factor(d$visitID, levels = vis_levels)


```


##### re-plot after fixing unsch
```{r}

# patients with SCANDT_period_year
g <- d[!is.na(d$DATE_coalsced),  ] %>%
  arrange( patID, visitID,DATE_coalsced)

# plot order of visits and dates visited by patients
ggplot(g, aes(x = DATE_coalsced, y = visitID, group = patID, color = patID)) +
  geom_line(show.legend = FALSE, alpha = .4) +
  geom_point(show.legend = FALSE, alpha = .4) +
  theme_minimal() +
  labs(x = "Visit ID", y = "Scan Date", title = "All Patients' Visit Timeline after fixing unscheduled visitIDs")

```

# remove unscheduled and control visits
```{r}

# where visitID has unsch or S remove row
bad_visits <- grep("unsch|\\dS|control", d$visitID, ignore.case = TRUE, value = TRUE) %>% unique()
bad_visits

# d[d$visitID %in% bad_visits , ] %>% nrow() # removes 224 rows if leaving the unsch visits
# d <-d[!d$visitID %in% bad_visits , ]

d[d$visitID == "control" ,] %>% nrow()
# d <- d[!d$visitID == "control" ,]
```
> all bad visits fixed
> controls already removed

```{r, look at dims}

check_dims()
# dims : 12600 265
#                      [,1]
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516

### MARCH 11 2025 ###
# dims : 12619 260
#                      [,1]
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

# # compare old to current names
# old <- readRDS(sprintf("%sprocessed_data/CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_01.rds",str_replace(wd_path,"14", "13") ))
# new <- readRDS(sprintf("%sprocessed_data/CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_01.rds",wd_path ))
# 
# 
# setdiff(
#   names(old),
#   names(new)
# )
# setdiff(
#   names(new),
#   names(old)
# )

```

# look at missingness
### make barplot of where each bar is count of filled cells for that column
```{r }

# make dataframe binary
df_binary <- d %>%
  mutate(month = as.numeric(visitID)) %>%
  arrange(month) %>%
  mutate_all(~ifelse(is.na(.), 0, 1))

# # plot where observed for cols in d ####
# pdf(paste0(wd_path, "results/visit_missingness_analysis/available_data_ordered_visits.pdf"), height = 30, width = 30, )
# Heatmap(df_binary %>% as.matrix(),
#         show_heatmap_legend = TRUE,
#         use_raster = F,
#         column_title  = "",
#         row_title  = "observations",
#         # col = c( "black",  "white" ),
#         col = c("white",   "black"), # 0 , 1
#         # na_col = "white",
#         cluster_rows = FALSE,
#         cluster_columns = FALSE,
#         show_row_names = FALSE,
#         show_column_names  = TRUE,
#         # column_names_rot = 45,
#         column_names_gp = gpar(fontsize = 5)
# )
# dev.off()

df_binary_colsums <- colSums(df_binary) %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  rename_with(~c("variable", "count_present"))



# label max !NA count
max_observed <- max(df_binary_colsums$count_present)

df_binary_colsums$order <- order(df_binary_colsums$count_present)

# mds <- cmdscale(dist(t(df_binary))) %>% 
#   as.data.frame()
# 
# mds$variable <- rownames(mds)
#   
# plot(mds[,1:2], type = "n"); text(mds[,1:2], label = mds$variable)
# 
# km <- kmeans(mds[,1:2], 8)


# cor_mat <- cor(select(d, where(is.numeric)), use = "pairwise.complete.obs")
# 
# # Compute missingness pattern correlation
# miss_mat <- as.matrix(is.na(select(d, where(is.numeric))))  # Missingness matrix
# miss_cor <- cor(miss_mat)  # Pairwise missingness correlation
# 
# # Combine correlation and missingness patterns
# dist_mat <- as.dist(1 - abs(cor_mat)) + as.dist(1 - abs(miss_cor))
# 
# Heatmap(as.matrix(dist_mat), cluster_rows = F, cluster_columns = F)
# 
# clust <- hclust(dist_mat, method = "ward.D2")
# 
# hc <- hclust(dist(t(df_binary)))
# 
# # df_binary_colsums$group = factor(km$cluster)
# df_binary_colsums$group = factor(cutree(hc, 12))

## plot ####
ggplot(df_binary_colsums, aes(x = factor(variable, levels = df_binary_colsums$variable[order]), y = count_present, fill = variable))+
  geom_bar(stat = "identity", show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Counts non-missing")
  
ggsave(paste0(wd_path, "results/visit_missingness_analysis/observed_colsums.png"), height = 10, width = 20 )

# ggplot(df_binary_colsums, aes(x = count_present, y = variable, fill = variable))+
#   geom_bar(stat = "identity", show.legend = FALSE, orientation = "y")+
#   theme(axis.text.x = element_text(angle = 90, hjust = 1))

# df_binary_colsums$variable <- factor(df_binary_colsums$variable, levels = df_binary_colsums$variable)
```
> substantially less vaa because not joining on visitID


#### order variables by dataset and update info with it
```{r, eval = FALSE}

# df of datasets and patterns for variables
info_sub <- info %>% 
  select(dataset, shorthand) %>% 
  arrange(dataset) %>% 
  mutate(shorthand = shorthand %>%   gsub("\\[1-9\\]", "", .)) # remove the number bits
  
# for each shorthand, grab variables associated with it (orders variables by dataset)
ordered_variables <- sapply(info_sub$shorthand, function(search_string){
  variables <- grep(search_string, df_binary_colsums$variable, value = TRUE)
  return(variables)
}, USE.NAMES = TRUE) %>% 
  unlist(., use.names = FALSE)

# add teh remainder of variables to beginning, they're out of order
all_variables <- c(setdiff(names(d), ordered_variables), ordered_variables) %>% 
  unique()
all(names(d) %in% all_variables) # Good = TRUE
all(all_variables %in% names(d)) # Good = TRUE


# match df to variable
ordering_df <- data.frame(dataset = NA, 
                          variable = all_variables)

# for each pattern in info, grab dataset and fill ordering df
for(i in 1:nrow(info_sub) ){
  
  # what dataset is associated with this search
  search <- info_sub[i, 2]
  
  # if search doesnt have 0 at end add $ because full name, no more in string
  if(substr(search, nchar(search), nchar(search)) != "0") search = paste0(search, "$") 
  
  datasets <- info_sub$dataset[ grep(search, info_sub$shorthand) ]
  print(paste(datasets, " ⭄  " ,search))
  dataset <- datasets %>% 
    paste(., collapse = "/")
  
  # fill dataset column in ordering df for rows that match this search
  ordering_df$dataset[ grep(search, ordering_df$variable)] <- dataset
  
}

# check that greping the right ones
  # table(c(
  # grep("RW0|RW1|TCV|AXIS|ACU|LOG|CYL|SPH|SNEL|VAA", ordering_df$variable, value = TRUE),
  # grep("FAF", ordering_df$variable, value = TRUE),
  # grep("CFF|FFA|CF|^QUAL", ordering_df$variable, value = TRUE),
  # grep("OCT|BLR|CTRPO", ordering_df$variable, value = TRUE),
  # grep("stratus", ordering_df$variable, value = TRUE),
  # grep("spectralis", ordering_df$variable, value = TRUE),
  # grep("DIAGNOSIS|GENDER|ETHN|YROF|RACE|month|EYE|pat|^VIS|subj", ordering_df$variable, value = TRUE),
  # grep("FAM", ordering_df$variable, value = TRUE),
  # grep("ENR|RELAT", ordering_df$variable, value = TRUE),
  # grep("DM|WGT|HGT", ordering_df$variable, value = TRUE)
  # ))

#### MANUALLY ENTER THE REMAINDER'S NOT HAVING A SHORTHAND IN INFO_SUB ###
ordering_df$variable[is.na(ordering_df$dataset)]

ordering_df$variable[ grepl("RW0|RW1|TCV|AXIS|ACU|LOG|CYL|SPH|SNEL|VAA|TCV|VISNO", ordering_df$variable)]
ordering_df$dataset[ grepl("RW0|RW1|TCV|AXIS|ACU|LOG|CYL|SPH|SNEL|VAA|TCV|VISNO", ordering_df$variable)& is.na(ordering_df$dataset)] <-  "VAA"
ordering_df$variable[is.na(ordering_df$dataset)]

ordering_df$variable[ grepl("FAF", ordering_df$variable)]
ordering_df$dataset[ grepl("FAF", ordering_df$variable)& is.na(ordering_df$dataset)] <-  "FAF"
ordering_df$variable[is.na(ordering_df$dataset)]

ordering_df$variable[ grepl("CFF|FFA|CF|^QUAL", ordering_df$variable)]
ordering_df$dataset[ grepl("CFF|FFA|CF|^QUAL", ordering_df$variable)& is.na(ordering_df$dataset)] <-  "CF_FFA"
ordering_df$variable[is.na(ordering_df$dataset)]

ordering_df$variable[ grepl("OCT|BLR|CTRPO", ordering_df$variable) & !grepl("STRAT|SPEC", ordering_df$variable)]
ordering_df$dataset[ grepl("OCT|BLR|CTRPO", ordering_df$variable) & !grepl("STRAT|SPEC", ordering_df$variable) & is.na(ordering_df$dataset)] <-  "OCTMAIN"
ordering_df$variable[is.na(ordering_df$dataset)]

ordering_df$variable[ grepl("stratus", ordering_df$variable, ignore.case = TRUE)]
ordering_df$dataset[ grepl("stratus", ordering_df$variable,ignore.case = TRUE)& is.na(ordering_df$dataset)] <-  "Macular_Thickness_OCT_Stratus"
ordering_df$variable[is.na(ordering_df$dataset)]

ordering_df$variable[ grepl("cirrus", ordering_df$variable,ignore.case = TRUE)]
ordering_df$dataset[ grepl("cirrus", ordering_df$variable,ignore.case = TRUE)& is.na(ordering_df$dataset)] <-  "Macular_Thickness_OCT_Cirrus"
ordering_df$variable[is.na(ordering_df$dataset)]

ordering_df$variable[ grepl("spectralis", ordering_df$variable,ignore.case = TRUE)]
ordering_df$dataset[ grepl("spectralis", ordering_df$variable,ignore.case = TRUE)& is.na(ordering_df$dataset)] <-  "Macular_Thickness_OCT_Spectralis"
ordering_df$variable[is.na(ordering_df$dataset)]

 ordering_df$variable[ grepl("DIAGNOSIS|GENDER|ETHN|YROF|RACE|month|EYE|patI|^VIS|subj", ordering_df$variable, ignore.case = TRUE)]
ordering_df$dataset[ grepl("DIAGNOSIS|GENDER|ETHN|YROF|RACE|month|EYE|pat|^VIS|subj", ordering_df$variable, ignore.case = TRUE) & is.na(ordering_df$dataset)] <-  "DIAGNOSIS/DEM/other"
ordering_df$variable[is.na(ordering_df$dataset)]


ordering_df$variable[ grepl("FAM", ordering_df$variable, ignore.case = TRUE) & is.na(ordering_df$dataset)] 
ordering_df$dataset[ grepl("FAM", ordering_df$variable, ignore.case = TRUE) & is.na(ordering_df$dataset)] <-  "FAMSTATUS"
ordering_df$variable[is.na(ordering_df$dataset)]

ordering_df$variable[ grepl("ENR|RELAT", ordering_df$variable, ignore.case = TRUE) & is.na(ordering_df$dataset)] 
ordering_df$dataset[ grepl("ENR|RELAT", ordering_df$variable, ignore.case = TRUE) & is.na(ordering_df$dataset)] <-  "ENR"
ordering_df$variable[is.na(ordering_df$dataset)]

ordering_df$variable[ grepl("DM|WGT|HGT", ordering_df$variable, ignore.case = TRUE) & is.na(ordering_df$dataset)] 
ordering_df$dataset[ grepl("DM|WGT|HGT", ordering_df$variable, ignore.case = TRUE) & is.na(ordering_df$dataset)] <-  "DM"
ordering_df$variable[is.na(ordering_df$dataset)]

ordering_df$variable[is.na(ordering_df$dataset)]
ordering_df$dataset[is.na(ordering_df$dataset)] <- "other"
ordering_df$variable[is.na(ordering_df$dataset)]


all(ordering_df$variable %in% names(d)) # good if TRUE
all(names(d) %in% ordering_df$variable)# good if TRUE

# join info and ordering df 
info_joined <- full_join(ordering_df, info, by = join_by(variable ==  shorthand))

# where not missing in y make dataset.x y
info_joined$dataset.x[!is.na(info_joined$dataset.y)] <-  info_joined$dataset.y[!is.na(info_joined$dataset.y)]

info_joined$dataset.y <- NULL
names(info_joined)[which(names(info_joined) == "dataset.x")] <- "dataset"

# for rows e.g "LRT[1-9] copy that info to LRT01...LRT09
## row value with info and pattern to search and fill in df
shorthand_pattern_df <- data.frame(shorthand=  grep("\\[1-9\\]", info_joined$variable, value = TRUE) ,
                                   pattern = grep("\\[1-9\\]", info_joined$variable, value = TRUE) %>% 
                                     gsub("\\[1-9\\]", "\\\\d", .)
                                   )  
# for each shorthand, fill rows with corresponding pattern with missing info
# i = 5
for(i in 1:nrow(shorthand_pattern_df)){

  shorthand <- shorthand_pattern_df[i, 1]
  pattern <- shorthand_pattern_df[i, 2]
  
  info_joined[grep(pattern, info_joined$variable), c(3:ncol(info_joined))] <- info_joined[info_joined$variable == shorthand, c(3:ncol(info_joined))]
  
  
}


# more info cleaning ----  

# fix octmain dataset values
info_joined <- info_joined %>% 
  mutate(dataset= dataset %>% 
           ifelse(. == "OCT_Main", "OCTMAIN", .)
           )

# remove shorthand rows
info_joined <- info_joined[!grepl("\\[1-9\\]", info_joined$variable) , ]

# remove octmain fovabnorm row: keep d where where not octmain and fovabnorm
info_joined <- info_joined[!(info_joined$dataset == "OCTMAIN" & info_joined$variable == "FOVABNORM")  ,  ]

# remove octmain fovabnorm row: keep d where where not octmain and fovabnorm
info_joined <- info_joined[!(info_joined$dataset == "OCTMAIN" & info_joined$variable == "CTR_PT_INVOLV")  ,  ]


# make original info updated one
info <- info_joined %>% 
  arrange(dataset)

```

#### plot counts of variables in order of dataset
```{r, eval = TRUE }

# join df_binary_colsums with info datasets
colsums_ordered <- df_binary_colsums %>% 
  left_join(., info[, c("name", "table")], by = join_by(variable == name)) 

# unique(info$dataset)
levels<-c(
"CF_FFA","FAFGRADE","OCTMTCIRRUS","OCTMTSPECTRALIS",
"OCTMTSTRATUS","OCTMAIN","VAA","MH/DIAGNOSIS","DEM/ENR/FAM/other"
)

colsums_ordered$table <- factor(colsums_ordered$table, levels = levels)

colsums_ordered <- colsums_ordered[order(colsums_ordered$table) , ]

colsum_plot <- ggplot(colsums_ordered,aes(count_present, reorder(variable, order(table)), fill = table))+
  geom_bar(stat = "identity", show.legend = TRUE, orientation = "y", width = .5, just = 0)+
  theme_light()+
  theme(axis.text.y = element_text( 
    size = 7
    # hjust = 1, vjust = 0,  angle = 90,
    # margin = margin(t = -100, r = 0, b = 0, l = 0))
        ))+
  geom_vline(xintercept = nrow(df_binary), col = "blue", linetype = 2)+
  annotate("text", y = 1, x = nrow(df_binary), label = paste("nrow(d):", nrow(df_binary)), hjust = 1, vjust = -0.5, color = "blue")

colsum_plot
ggsave(paste0(wd_path, "results/visit_missingness_analysis/colsum_plot.png"), width = 20  , height = 15, dpi = 150, limitsize = F)
 
### same plot but faceted for visitID ####
visit_var_count <- d %>%
  group_by(visitID) %>%
  mutate(across(all_of(names(d)[names(d) != "visitID"]),
                ~!is.na(.))
         ) %>%
  summarize(across(all_of(names(d)[names(d) != "visitID"]), sum)) %>%
  reshape2::melt(., id.vars = "visitID") %>%
  rename_with(~c("visitID", "variable", "count"))

# what visits are not visiless,control, or 000
follow_ups <- unique( visit_var_count$visitID)[!unique( visit_var_count$visitID) %in% c("000", "VISITLESS", "control")]

visit_var_count <- visit_var_count %>%
  mutate(visit_group =   case_when(
    visitID %in% follow_ups ~ ">000",
    TRUE ~ visitID
  )) %>% 
  mutate(visit_group = visit_group %>% factor(., levels = c("000", ">000")))

visit_colsum_plot <- ggplot(visit_var_count, aes(x = variable ,y = count, fill = variable))+
  geom_bar(stat = "identity", show.legend = FALSE, orientation = "x", width = .5, just = 0)+
  theme_bw()+
  theme(axis.text.x = element_text( hjust = 1, vjust = 0, size = 3, angle = 90,
                                    margin = margin(t = 0, r = 0, b = 0, l = 0)
                                    )
        )+
  coord_cartesian(ylim = c(0, nrow(d)))+
  geom_abline(intercept = nrow(df_binary), col = "blue", linetype = 2, slope = 0)+
  annotate("text", x = Inf, y = nrow(df_binary), label = paste("nrow(d):", nrow(df_binary)), hjust = 1, vjust = -0.5, color = "blue")+
  facet_wrap(~visit_group, scales = "free")
visit_colsum_plot
ggsave(paste0(wd_path, "results/visit_missingness_analysis/visit_colsum_plot.png"),visit_colsum_plot, width = 20, height = 15, dpi = 400, limitsize = F)

# for only VAA
VAA_visit_colsum_plot <- ggplot(visit_var_count %>% filter(variable %in% names(raw$VAA)), aes(y= variable , x = count, fill = variable))+
  geom_bar(stat = "identity", show.legend = FALSE, orientation = "y", width = .5, just = 0)+
  theme_bw()+
  theme(axis.text.y = element_text( hjust = 1, vjust = 0, size = 5, angle = 0,
                                    margin = margin(t = 0, r = 0, b = 0, l = 0)
                                    )
        )+
  facet_wrap(~visit_group, scales = "fixed", ncol = 4)

ggsave(paste0(wd_path, "results/visit_missingness_analysis/VAA_visit_colsum_plot.png"),VAA_visit_colsum_plot, width = 15, height = 10, dpi = 150, limitsize = F)

```

# what datasets have eye_num missing?
```{r }

map(raw, ~{x <- .
  table(x[['eye']], useNA = "always")
})
table(d$eye_num, useNA = "always")


```

# = = = = = = = = = = = 

# save info
```{r}
write.xlsx(info, paste0(wd_path,"processed_data/database_grading_key_02.xlsx"))
# "~/Documents/LMRI/PRJ2024001/version015/processed_data/database_grading_key_02.xlsx"
```

# save data
```{r}
new_data_name <- data_name %>% 
  gsub("1", "2", .)

d_path <- paste0(wd_path,"processed_data/", new_data_name)
print(paste("Saving... \n", d_path))
saveRDS(d, d_path)
# ~/Documents/LMRI/PRJ2024001/version015/processed_data/CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_02.rds
```

```{r}
sum(  ( !is.na(d$cirrus_MT01) |  !is.na(d$stratus_MT01)  ) &  !is.na(d$spectralis_MT01)) # 709 >> 575 >>583
sum(!is.na(d$cirrus_MT01) &  !is.na(d$stratus_MT01) ) # 201 >> 197>>199
sum(!is.na(d$spectralis_MT01) &  !is.na(d$stratus_MT01) ) # not recorded >> 291 >>297
sum(!is.na(d$spectralis_MT01) &  !is.na(d$cirrus_MT01) ) # not recorded >> 292 >> 296
```

```{r, look at dims}
check_dims()
# dims : 12600 265
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516

### MARCH 12 2025 ###
# dims : 12619 260
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```
