---
title: "1.5 MT modeling"
author: "Christian Anderson"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(tidyverse)
library(ggpubr)
library(vcd)
library(UpSetR)
library(gridExtra)
library(ComplexHeatmap)
library(viridis)
library(BiocGenerics)
library(mice)
library(Metrics)
library(purrr)
# library(SingleCellExperiment)
library(corrplot)
library(RColorBrewer)
library(plotly)
library(scatterplot3d)


rm(list = ls())
gc()
par(mfrow = c(1,1))
```

# Load Data

```{r}
wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

data_name <- "CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_02.rds"
data <- readRDS(paste0(wd_path,"processed_data/", data_name))
d <- data
raw <- readRDS(paste0(wd_path,"processed_data/dfs_list_01.rds"))

```

# make new directories
```{r}

mt_imputation_path <- paste0(wd_path, "results/mt_imputation/")
if(!dir.exists(mt_imputation_path)) {
  dir.create(mt_imputation_path, recursive = TRUE)
  
}

if(!dir.exists(paste0(wd_path, "processed_data/imputations"))) {
  dir.create(paste0(wd_path, "processed_data/imputations"), recursive = TRUE)
  
}

```

# \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# clean data

```{r, look at dims}

check_dims <- function(D = d){
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()

# dims : 12600 265
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516

### MARCH 12 2025 ###
# dims : 12619 260
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```

## make sample unique id
```{r}

d$DATE_coalesced <- d$DATE_coalsced
d$DATE_coalsced <- NULL

n_distinct(with(d, paste(patID_eye, DATE_coalesced, sep = "_")))== nrow(d)

d$sample <-with(d, paste(patID_eye, DATE_coalesced, sep = "_"))

```


## ensure the mt columns are numeric; check how many records overlap among the three machines
```{r }

# set mt columns numeric
d <- d %>% 
  mutate_at(vars(grep("^cirrus_.*\\d$|^stratus_.*\\d$|^spectralis_.*\\d$", names(data) )), as.numeric)

sum(  ( !is.na(d$cirrus_MT01) |  !is.na(d$stratus_MT01)  ) &  !is.na(d$spectralis_MT01)) # 709 overlaps >>583
sum(!is.na(d$cirrus_MT01) &  !is.na(d$stratus_MT01) ) # 201 overlaps>>199

```


## machine indices and conditions
```{r}

## mt columns
cirrus_cols <-  grep("cirrus_MT0\\d$", names(d), value = TRUE)
stratus_cols <-  grep("stratus_MT0\\d$", names(d), value = TRUE)
spectralis_cols <- grep("spectralis_MT0\\d$", names(d), value = TRUE)
mt_cols <- c(cirrus_cols, stratus_cols, spectralis_cols)

# logical vec of where eye_num exists and is == 2
leftEye <- d$eye_num =="2" & !is.na(d$eye_num)


```

find the mt columns and where eye is left

## map to right

### look at current mapping
```{r}

# subset for brv
rpe <- d %>% 
  select(all_of(c("DIAGNOSIS", "eye_num", "patID_eye", "visitID", grep("RPE_LK0\\d", names(d), value = TRUE)))) %>% 
  filter(!is.na(RPE_LK01) & DIAGNOSIS =="yes")
             
# flatten tidy
rpe_flat <- rpe %>%
  pivot_longer(cols = starts_with("RPE_LK"), 
               names_to = "category",
               values_to = "score")


ggplot(rpe_flat, aes(x = category, fill = factor(score) ))+ 
  geom_bar(position = "fill")+ 
  # facet_wrap(~eye_num, scales = "free_y", nrow = 1)+
   theme_bw()+
  geom_text(stat = "count", aes(label = after_stat(count), group = score), position = position_fill(vjust = 0.5))+
  ggtitle("RPE leakage")

```
> zone 3 and 5 must be assigned non-anatomically 

### map to right function
```{r}

# columns with mapping 0-9   
zonal_cols <- grep("^[a-zA-Z].*0[1-9]$", names(d), value = TRUE) %>% 
  grep("RW", ., value = TRUE, invert = TRUE)

# unique prefixes (symptoms)
unique_categories <- unique(gsub("[1-9]", "" , zonal_cols)) 

# check search
lapply(unique_categories, function(prefix){
  print(prefix)
  grep(prefix, names(d), value = TRUE)
})


### check
category_prefix = "BRV0"
dat <- d[leftEye , grep(category_prefix, names(d))] 
###

# function to use in loop through unique symptom columns after subsetting for symptom: for each set of 9 columns, change 3-5 & 5-3, 7-9 & 9-7
mapRight <- function(dat = data.frame()){
  
  col_names <- names(dat)
  
  # what zones does this phenotype have?
  zones_present <- col_names %>% 
    gsub("\\D", "", .) %>% 
    gsub("0", "", .)
  
  # rename temp dat with zones
  temp_dat <- dat
  names(temp_dat) <- paste0("tmp", zones_present)
  
  # join tmp vars to dat
  dat <- cbind(dat, temp_dat)
  
  # ## check ####
  # x <- dat
  # x[, col_names[c(3,5,7,9)]] <- dat[, names(temp_dat)[c(5, 3, 9, 7)]]
  # x <- x %>% select(-starts_with("tmp"))
  # 
  # unique(x) %>% 
  #   data.matrix() %>% 
  #   Heatmap(cluster_columns = F,cluster_rows = F, col = c("white", "black"))
  # unique(select(dat, -matches("tmp"))) %>% 
  #   data.matrix()  %>%
  #   Heatmap(cluster_columns = F,cluster_rows = F, col = c("white", "black"))
  # ## ####
  
  # where 3 make 5, and 7 make 9
  dat[, col_names[c(3,5,7,9)]] <- dat[, names(temp_dat)[c(5, 3, 9, 7)]]
  dat <- dat %>% select(-starts_with("tmp"))

return(dat)
}

```
> where zonal suffix, with those columns' prefixes, switch 5 and 3, 9 and 7


### Use map to right on df
```{r}

# feed  unique_category name subseted data into function
for (category_prefix in unique_categories){
  # for each category, for left eye rows, switch 3,5 and 7,9 with mapRight()
  d[leftEye , grep(category_prefix, names(d))] <-  d[leftEye , grep(category_prefix, names(d))] %>% mapRight(.)
    
}
# d[leftEye , grep(category_prefix, names(d), value = TRUE)] %>%  subset(., complete.cases(.))


```


### look at fixed mapping
```{r}

# subset for brv
rpe <- d %>% 
  select(all_of(c("DIAGNOSIS", "eye_num", "patID_eye", "visitID", grep("RPE_LK0\\d", names(d), value = TRUE)))) %>% 
  filter(!is.na(RPE_LK01) & DIAGNOSIS =="yes")
             
# flatten tidy
rpe_flat <- rpe %>%
  pivot_longer(cols = starts_with("RPE_LK"), 
               names_to = "category",
               values_to = "score")


ggplot(rpe_flat, aes(x = category, fill = factor(score) ))+ 
  geom_bar(position = "fill")+ 
  # facet_wrap(~eye_num, scales = "free_y", nrow = 1)+
   theme_bw()+
  geom_text(stat = "count", aes(label = after_stat(count), group = score), position = position_fill(vjust = 0.5))+
  ggtitle("RPE leakage")

```

## remove variables that don't show anything
```{r}

# variables true if only 0 = not present or NA
lame_vars <- d %>% 
  sapply(., function(x){
    return(all(ifelse(unique(x) %in% c(NA, "0"), TRUE, FALSE) ))
    }) 

lame_vars_names <- lame_vars[lame_vars] %>% names()
d[,lame_vars_names] %>% lapply(., unique) %>% as.data.frame() # check them
#  DATE_OCTMTSTRATUS DRV06 DRV07 DRV08 DRV09 ICL07 RAV06 RAV07 RAV08 RAV09 VTV06 VTV08 stratus_SCANDT VAPUDI
#               <NA>     0     0     0     0     0     0     0     0     0     0     0           <NA>     NA
#               <NA>    NA    NA    NA    NA    NA    NA    NA    NA    NA    NA    NA           <NA>     NA

# exclude vars that don't show anything
d <- d %>% 
  select(!all_of(lame_vars_names))

```

```{r}
check_dims()
# dims : 12600 247
#                      [,1]
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516

### MARCH 12 2025 ###
# dims : 12619 246
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```

# \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# Outliers

## look at outliers
```{r eval=TRUE, include=TRUE}
# 
# # these should be the thinnest values
# MT1_vals <- c(d$stratus_MT01, d$cirrus_MT01, d$spectralis_MT01)
# boxplot(MT1_vals, yaxt = 'n')
# y_range <- range(MT1_vals, na.rm = TRUE)
# ticks <- seq(from = floor(y_range[1]), to = ceiling(y_range[2]), by = 10) 
# primary_line <- ticks[seq(1, length(ticks), 2)]
# secondary_line <- ticks[seq(2, length(ticks), 2)]
# axis(side = 2, at = ticks, las = 1)
# abline(h = primary_line, col = "grey", lty = 3)
# abline(h = secondary_line, col = "grey", lty = 4)


upper = c(
  500,  500,  500,  500,  500,  500,  400,
  500,  500,  500,  400,  450,  450,  400,
  450,  400,  400,  450,  600,  600,  600,
  800,  600,  600,  600,  500,  600
  )
  
lower = c(
  10,  10,  10,  10,  10,  10,  10,  
  110,  100,  10,  10,  10,  10,  10,  
  150,  10,  10,  100,  50,  50,  50,  
  50,  50,  100,  50,  200,  100
  )
par(mfrow = c(4,7))
for(i in seq(mt_cols)){
  vals <- as.numeric( data.matrix(d[ , mt_cols[i]] ))
  cols <- ifelse(is.na(vals), "black",
                 ifelse(vals > upper[i] |vals<lower[i] , "red",
                        "blue") )
  
  boxplot(vals, main =mt_cols[i] )                      
  points(rep(1, length(vals)), vals, col = cols, pch = 16)  
}
par(mfrow = c(1,1))

# par(mfrow = c(1,3))
# for(columns in list(cirrus_cols, stratus_cols, spectralis_cols)){
#   pc <- prcomp(d[,columns] %>% drop_na %>% scale())
#   plot(pc$x, main = str_remove(columns[[1]], "0\\d"))
#   symbols(0, 0, circles = 5, inches = FALSE, add = TRUE, lty = 2,fg = "blue")
#   symbols(0, 0, circles = 10, inches = FALSE, add = TRUE, lty = 2,fg = "red")
# }
# par(mfrow = c(1,1))

```
> outliers @ ~ (>605 & < 50)


## remove outlier
where measurement in mt_cols is in not in_range, set to NA
```{r}

# check that all values integers
sapply(d[,mt_cols],function(x) {
  abs(x-round(x)) %>% unique()
})
# ^ difference between rounded value and value is 0 except for spectralis_MT04

# round all mt cols so that in_range evaluates as expected with integers
d[,mt_cols] <- round( d[,mt_cols] )

in_range <- c(50:605)
# where nonNA mt_column value outside of range
d %>%
  select(all_of(mt_cols)) %>% 
  filter(if_any(all_of(mt_cols),
                ~ !is.na(.) & 
                  !. %in% in_range)
         ) %>% 
  glimpse()

## set outlier as na for each variable's outlier range
for(i in seq(mt_cols)){
  column = mt_cols[[i]]
  where <- !is.na(d[,column]) &  (d[,column]> upper[i] | d[,column]<lower[i])
  d[where, column] <- NA
}


```


## before imputation look at image

```{r eval=TRUE, include=TRUE}

# make mtmachinezone x sample matrix
m <- d[ , c(grep("sample", names(d), value = TRUE), mt_cols)]
rownames(m) <- m$sample
m$sample <- NULL
m <- m %>% 
  t()

# select columns wehre at least 1 non-NA present
m_full <-m[, apply(m, 2, function(x) any(!is.na(x)))]

t_m_full <- t(m_full)

pdf(paste0(wd_path, "results/mt_imputation/MT_cols_pre_imputations.pdf"), height = 11, width = 10)
Heatmap(t_m_full,
        show_heatmap_legend = TRUE,
        use_raster = FALSE,
        column_title  = "MT columns before imputations",
        row_title  = "observations",
        name = "thickness",
        # col = c("white", "navy"   ),
        col = viridis(100),
        na_col = "white",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_row_names = FALSE,
        show_column_names  = TRUE,
        column_names_rot = 45
)
dev.off()
```


## look at missisngness
how many NAs are in each machine? >> spectralis has most NAs and stratus least so modeling anything to stratus measurements makes most sense
```{r eval=TRUE, include=TRUE}

# sum of NAs for each machine_zone
available_sum_cols <- d %>%
  select(all_of(mt_cols)) %>%
  summarise(across(everything(), ~sum(!is.na(.)))) %>%
  t() %>%
  as.data.frame()

names(available_sum_cols) <- "available_sum_cols"
available_sum_cols$machine <- gsub("_.*\\d", "",  rownames(available_sum_cols) )
available_sum_cols$zone <- gsub(".*_", "",  rownames(available_sum_cols) )

# plot counts of NAs for each zone in each machine
pdf(paste0(wd_path, "results/mt_imputation/count_observed_mt_cols.pdf"), height = 11, width = 10)
ggplot(available_sum_cols, aes(x = machine, fill = zone,  y = available_sum_cols))+
         geom_bar(stat = "identity", position = "dodge")+
  ggtitle("Number of available values for each machine/zone")
dev.off()

```



```{r}
check_dims()
# dims : 12600 247
#                      [,1]
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516
#^ no change


### MARCH 12 2025 ###
# dims : 12619 246
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```

#\_\_\_\_\_\_\_\_\_\_\_\_
# impute intra-machines
for each machine, impute NAs where btw 0 and 5 NAs so enough power to do so accurately

### new d for intra imputes
```{r}
intra_impute_d <- d
```

### conditions for imputation
conditions for rows of each machine where there's between 0 and 5 NAs and of rows wehre there is no NAs
```{r}

# rows where some NAs in mt_cols but not all>> between 0 and 5 NAs
cirr_NA_condition <- which(rowSums(is.na(intra_impute_d[, cirrus_cols])) %in% c(1:6))
strat_NA_condition <-  which(rowSums(is.na(intra_impute_d[, stratus_cols])) %in% c(1:6))
spec_NA_condition <-  which(rowSums(is.na(intra_impute_d[, spectralis_cols])) %in% c(1:6))

# where full rows
cirr_full_condition <- which(rowSums(is.na(intra_impute_d[, cirrus_cols])) == 0)
strat_full_condition <- which(rowSums(is.na(intra_impute_d[, stratus_cols])) == 0)
spec_full_condition <- which(rowSums(is.na(intra_impute_d[, spectralis_cols])) == 0)


```


### Cirrus

```{r,  results='hide'}
# Cirrus
## impute NAs from other cirrus values
cirrus_imp <- mice(intra_impute_d[c(cirr_full_condition, cirr_NA_condition), cirrus_cols], m = 5, method = 'pmm', seed = 098)
cirrus_imp_sum <- complete(cirrus_imp, action = 1)
for(i in 2: cirrus_imp$m){ cirrus_imp_sum <-  cirrus_imp_sum + complete(cirrus_imp, i)}
cirrus_imp_mean <- cirrus_imp_sum / cirrus_imp$m

# insert data at conditions, includes old valuesand new imputed values
intra_impute_d[c(cirr_full_condition, cirr_NA_condition), cirrus_cols] <- cirrus_imp_mean

```

### Stratus

```{r,  results='hide'}
# stratus
## impute NAs from other stratus values
stratus_imp <- mice(intra_impute_d[c(strat_full_condition, strat_NA_condition), stratus_cols], m = 5, method = 'pmm', seed = 098)
stratus_imp_sum <- complete(stratus_imp, action = 1)
for(i in 2: stratus_imp$m){ stratus_imp_sum <-  stratus_imp_sum + complete(stratus_imp, i)}
stratus_imp_mean <- stratus_imp_sum / stratus_imp$m

# insert data at conditions, includes old valuesand new imputed values
intra_impute_d[c(strat_full_condition, strat_NA_condition), stratus_cols] <- stratus_imp_mean
```

### Spectralis

```{r,  eval=TRUE, include=TRUE}

## impute NAs from other spectralis values
spectralis_imp <- mice(intra_impute_d[c(spec_full_condition, spec_NA_condition), spectralis_cols], m = 5, method = 'pmm', seed = 098)
spectralis_imp_sum <- complete(spectralis_imp, action = 1)
for(i in 2: spectralis_imp$m){ spectralis_imp_sum <-  spectralis_imp_sum + complete(spectralis_imp, i)}
spectralis_imp_mean <- spectralis_imp_sum / spectralis_imp$m

# insert data at conditions, includes old valuesand new imputed values
intra_impute_d[c(spec_full_condition, spec_NA_condition), spectralis_cols] <- spectralis_imp_mean
```


# plot boxplots
### each machine's zone measure distributions
```{r eval=TRUE, include=TRUE}
# stretch to have category of zone and camera with corresponding score
mt_long <- intra_impute_d %>%
  select(mt_cols) %>%
  gather(category, value) %>%
  filter(!is.na(category) & !is.na(value))

# extract machine factor
mt_long$machine <- mt_long$category %>%
  substr(., 1, 3) %>% as.factor()


pdf(paste0(wd_path, "results/mt_imputation/MT_cols_boxplots.pdf"), height = 11, width = 10, bg = "white")
ggplot(mt_long, aes(  y= category,x =value , fill = machine))+
  geom_boxplot(show.legend = TRUE, width = .5, position = position_dodge(width = 0.7), orientation = "y", color = "black", alpha = .4)+
  theme(axis.text.x = element_blank(),
        axis.text.y  = element_blank())+
  theme_minimal()
dev.off()
```
> stratus shifted higher than spectralis and cirrus, spectralis higher than cirrus

### comparing machine distributions where overlap betweeen machines
```{r}

# where overlap between machines
stratus_spectralis_overlap <- rowSums(is.na(intra_impute_d[, c(stratus_cols, spectralis_cols)])) == 0
# intra_impute_d[stratus_spectralis_overlap, c(stratus_cols, spectralis_cols)]

stratus_cirrus_overlap <- rowSums(is.na(intra_impute_d[, c(stratus_cols, cirrus_cols)])) == 0
# intra_impute_d[stratus_cirrus_overlap, c(stratus_cols, cirrus_cols)] %>% glimpse()

spectralis_cirrus_overlap <- rowSums(is.na(intra_impute_d[, c(spectralis_cols, cirrus_cols)])) == 0
# intra_impute_d[spectralis_cirrus_overlap, c(spectralis_cols, cirrus_cols)]%>% glimpse()

stratus_spectralis <- intra_impute_d %>% 
  select(all_of(c(stratus_cols, spectralis_cols))) %>% 
  filter(stratus_spectralis_overlap) %>% 
  reshape2::melt(.) %>% 
  mutate(overlap = "spectralis_stratus")

stratus_cirrus <- intra_impute_d %>% 
  select(all_of(c(stratus_cols, cirrus_cols))) %>% 
  filter(stratus_cirrus_overlap) %>% 
  reshape2::melt(.) %>% 
  mutate(overlap = "stratus_cirrus")

spectralis_cirrus <- intra_impute_d %>% 
  select(all_of(c(spectralis_cols, cirrus_cols))) %>% 
  filter(spectralis_cirrus_overlap) %>% 
  reshape2::melt(.) %>% 
  mutate(overlap = "spectralis_cirrus")

machine_overlap_d <- purrr::reduce(list(stratus_spectralis, stratus_cirrus, spectralis_cirrus), rbind ) %>% 
  mutate(zone = gsub("\\D", "", variable)) %>% 
  mutate(machine = sapply(str_split(variable, "_"), '[' , 1))

ggplot(machine_overlap_d, aes(x= overlap, y = value, fill = machine))+
  geom_boxplot(show.legend = TRUE, width = .5, position = position_dodge(width = 0.7), orientation = "x", color = "black", alpha = .4)+
  ggtitle("MT zone distributions where overlap between machines")

```

# image of all zones after intra-category imputation
```{r eval=TRUE, include=TRUE}

m <- intra_impute_d[ , c("sample", mt_cols)]
rownames(m) <- m$sample
m$sample <- NULL
m <- m %>% t()

# select columns wehre at least 1 non-NA present
m_full <- m[, apply(m, 2, function(x) any(!is.na(x)))]

# image matrix order l>r cirrus, stratus, spectralis
pdf(paste0(wd_path, "results/mt_imputation/MT_cols_post_intracategory_imputations.pdf"), height = 11, width = 10)
Heatmap(m_full %>% t(),
        name = "thickness",
        show_heatmap_legend = TRUE,
        use_raster = F,
        column_title  = "MT columns after intra-category imputations",
        row_title  = "observations",
        # col = c(  "white" , "darkgreen"),
        col = viridis(100),
        na_col = "white",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_row_names = FALSE,
        show_column_names  = TRUE,
        column_names_rot = 45
)
dev.off()
```

## how many overlaps

```{r}

# how many overlaps after intra_impute
overlaps <- c(
  paste(sum(!is.na(intra_impute_d$cirrus_MT01)&!is.na(intra_impute_d$stratus_MT01)), "overlap between cirrus and stratus"),
  paste(sum(!is.na(intra_impute_d$spectralis_MT01)&!is.na(intra_impute_d$cirrus_MT01)), "overlap between spectralis and cirrus"),
  paste(sum(!is.na(intra_impute_d$spectralis_MT01)&!is.na(intra_impute_d$stratus_MT01)),"overlap between spectralis and stratus")
)

print(overlaps)
# "199 overlap between cirrus and stratus"     "296 overlap between spectralis and cirrus" 
# [3] "297 overlap between spectralis and stratus"

overlap_m_empty <- matrix(dimnames = list(c("cirrus_MT01", "spectralis_MT01", "stratus_MT01"), c("cirrus_MT01", "spectralis_MT01", "stratus_MT01")),
                    nrow = 3, 
                    ncol = 3, 
                    data = NA )

overlap_m <- overlap_m_empty; for( r in rownames(overlap_m)){
  
  for( c in colnames(overlap_m)){
    
    overlap_m[r,c] <- sum(!is.na(intra_impute_d[[c]])&!is.na(intra_impute_d[[r]]))
    
  }
  
}
  

```
___old___
"204 overlap between cirrus and stratus"    
"308 overlap between spectralis and cirrus" 
"322 overlap between spectralis and stratus"
_________
___new____
"206 overlap between cirrus and stratus"
"315 overlap between spectralis and cirrus" 
"322 overlap between spectralis and stratus"
_________
___newer____
"235 overlap between cirrus and stratus"    
"339 overlap between spectralis and cirrus" 
"375 overlap between spectralis and stratus"
_________
___newest___
"197 overlap between cirrus and stratus"
"73 overlap between spectralis and cirrus" 
"92 overlap between spectralis and stratus"
____20/5/224____
"54 overlap between cirrus and stratus"     
"188 overlap between spectralis and cirrus" 
"165 overlap between spectralis and stratus"

___MARCH 12 2025___
"199 overlap between cirrus and stratus"     
"296 overlap between spectralis and cirrus" 
"297 overlap between spectralis and stratus"

# \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# new d for inter imputes 
```{r}
inter_impute_d <- intra_impute_d %>% 
  as.data.frame()
# inter_impute_d <- d
# ^ not as accurate without the intra imputation first

```
> evaluated interimpute without intra_impute first and imputing in two steps works best. The relationship among zones within eaach machine is more accurate to fill intra_machine missing. then interimpuptation has more N to work with as well (it filters out non-full rows). 

# \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# mice intercatagory imputation tuning
with less longitudinal sparsity within machines, now impute the missing values of spectralis based on prediction from cirrus and stratus

## plot correlations between stratus and cirrus pre inter-imputation
```{r}
# height and width
h = 700; w = 1000
```

```{r eval=FALSE, include=TRUE}
# plot stratus by cirrus with r value for nine zones
png(file = paste0(wd_path, "results/mt_imputation/stratus_by_cirrus_pre_interimputation.png"), height = h, width = w)
par(mfrow = c(3,3))
for ( i in 1:9){
  stratus <- inter_impute_d[, stratus_cols[i]] 
  cirrus <- inter_impute_d[, cirrus_cols[i]]
  mod <- lm(stratus~cirrus)

  title <- paste(
    stratus_cols[i] %>% substr(., nchar(.)-3, nchar(.)) ,
    " ~ " ,
    cirrus_cols[i] %>% substr(., nchar(.)-3, nchar(.)),
    "; r = ",
    round(cor( stratus,  cirrus,use = "pair"), 2)
  )

  plot(stratus ~ cirrus,
       main =  title,
       xlim = c(0, 500),
       ylim = c(0, 500))
  abline(c(0,1), col = "grey", lty = 2)
  abline(mod, col = "red")
  rug(stratus, side = 2, col = "darkgreen")
  rug(cirrus, col = "blue")
  text(x = c( 100), y = c(400), labels = c("pre-imputation"), cex = .8)

}
par(mfrow = c(1,1))
dev.off()

```
> stratus and cirrus have strong linear relationship


## plot correlations spectralis ~ stratus pre inter-imputation

```{r eval=TRUE, include=TRUE}
# plot spectralis by stratus with r value for nine zones
png(file = paste0(wd_path, "results/mt_imputation/spectralis_by_stratus_pre_interimputation.png"), height = h, width = w)
par(mfrow = c(3,3))
for ( i in 1:9){
  stratus <- inter_impute_d[, stratus_cols[i]]
  spectralis <- inter_impute_d[, spectralis_cols[i]]
  mod <- lm(spectralis~stratus)

  title <- paste(
    stratus_cols[i] %>% substr(., nchar(.)-3, nchar(.)) ,
    " ~ " ,
    spectralis_cols[i] %>% substr(., nchar(.)-3, nchar(.)),
    "; r = ",
    round(cor( stratus,  spectralis,use = "pair"), 2)
  )

  stratus_rug <- stratus[!is.na(spectralis) & !is.na(stratus)]
  spectralis_rug <- spectralis[!is.na(spectralis) & !is.na(stratus)]

  plot(spectralis~stratus,
       main =  title,
       xlim = c(0, 500),
       ylim = c(0, 500))
  abline(c(0,1), col = "grey", lty = 2)
  abline(mod, col = "red")
  rug(spectralis_rug, side = 2, col = "darkgreen")
  rug(stratus_rug, col = "blue")
  text(x = c( 100), y = c(400), labels = c("pre-imputation"), cex = .8)

}
par(mfrow = c(1,1))
dev.off()

```
>


## plot correlations spectralis ~ cirrus pre inter-imputation
```{r eval=TRUE, include=TRUE}

# plot spectralis by cirrus with r value for nine zones

png(file = paste0(wd_path, "results/mt_imputation/spectralis_by_cirrus_pre_interimputation.png"), height = h, width = w)
par(mfrow = c(3,3))
for ( i in 1:9){
  cirrus <- inter_impute_d[, cirrus_cols[i]]
  spectralis <- inter_impute_d[, spectralis_cols[i]]
  mod <- lm(spectralis~cirrus)

  title <- paste(
    cirrus_cols[i] %>% substr(., nchar(.)-3, nchar(.)) ,
    " ~ " ,
    spectralis_cols[i] %>% substr(., nchar(.)-3, nchar(.)),
    "; r = ",
    round(cor( cirrus,  spectralis,use = "pair"), 2)
  )

  cirrus_rug <- cirrus[!is.na(spectralis) & !is.na(cirrus)]
  spectralis_rug <- spectralis[!is.na(spectralis) & !is.na(cirrus)]

  plot(spectralis~cirrus,
       main =  title,
       xlim = c(0, 500),
       ylim = c(0, 500))
  abline(c(0,1), col = "grey", lty = 2)
  abline(mod, col = "red")
  rug(spectralis_rug, side = 2, col = "darkgreen")
  rug(cirrus_rug, col = "blue")
  text(x = c( 100), y = c(400), labels = c("pre-imputation"), cex = .8)

}
par(mfrow = c(1,1))
dev.off()

```
>

#-
# evaluate mice with artificial NAs
- first find best method to use - pmm, lasso.norm, etc. - number of iterations
- starting with modeling stratus to spectralis

## TRUE to run evaluation on imputation
```{r}
#
##
###
####
#####
impute_eval <- FALSE
run_iter_eval <- FALSE
#####
####
###
##
#
```

### predictor matrix
```{r}
# create predictor matrix so that cirrus_mt01 predicts stratus_mt01 etc.
predictorMatrix <- matrix(0,length(mt_cols),length(mt_cols))
colnames(predictorMatrix) <- mt_cols
rownames(predictorMatrix) <- mt_cols


d1 <- rbind(
  cbind(10:18, 19:27), 
  cbind(10:18, 1:9),   
  cbind(19:27, 1:9),   
  cbind(19:27, 10:18), 
  cbind(1:9, 10:18),   
  cbind(1:9, 19:27)    
)

# zone_n impute across machines
for(i in 1:nrow(d1)){
  predictorMatrix[d1[i,1],d1[i,2]] <- 1
}

Heatmap(predictorMatrix, 
        col = c("white", "black"), 
        show_row_names = TRUE,
        show_column_names = TRUE, 
        show_row_dend = FALSE, 
        show_column_dend = FALSE, 
        cluster_rows = FALSE,
        cluster_columns = FALSE
        )

```

## make test df and run evaluation of
```{r }
 
 if  (impute_eval){

     # stratus_cols, cirrus_cols, spectralis_cols
     # mt cols with >0 non-na
     d_truth<- inter_impute_d %>%
       select(all_of(mt_cols))

     # index where all stratus or all cirrus, and where spectralis
     cirrus_or_stratus_observed <- (rowSums(is.na(d_truth[, stratus_cols])) == 0 )  | (rowSums(is.na(d_truth[, cirrus_cols])) == 0)
     spectralis_observed <- rowSums(is.na(d_truth[, spectralis_cols]) ) == 0

     # ( stratus or cirrus ) and spectralis
     cirrus_or_stratus_and_spectralis_observed <- cirrus_or_stratus_observed & spectralis_observed

     # Truth is mt cols where either stratus or cirrus full across, and spectralis full across
     d_truth <- d_truth[cirrus_or_stratus_and_spectralis_observed , ]
     rownames(d_truth) <- 1:nrow(d_truth)

     # dataset to randomly introduce nas into
     d_test <- d_truth

     # artificially make NA rows to test imputation
   ## sample indexes of full rows
   num_na <- floor( nrow(d_test) * .2)
   set.seed(190)
   sample_idx <- sample(nrow(d_test), num_na)
   sample_idx <- sample_idx[order(sample_idx)]
   ## set those rows in the spectralis columns artificially to NA
   d_test[sample_idx,spectralis_cols] <- NA
   # test should all be NA and truth all not na
   d_test[sample_idx,spectralis_cols] %>% is.na() %>% table() # good if all TRUE
   d_truth[sample_idx,spectralis_cols] %>% is.na() %>% table()# good if all FALSE
   # switch to execute imputation or not
   if(TRUE){
       # impute for pmm and lasso.norm
       # for(method in c("lasso.norm", "pmm")){
       for(method in c("lasso.norm")){
               # impute test data for stratus from stratus
                 spectralis_imp <- mice(d_test, m = 10, method = method, seed = 3478, predictorMatrix=predictorMatrix)

                 # mice::densityplot(stratus_imp, main = method)

                 # find mean of imputed data
                 # initial value
                 imp_sum <- complete(spectralis_imp, action = 1)

                 # sum each consecutive df to last
                 for (i in 2:spectralis_imp$m){
                   imp_sum <- imp_sum + complete(spectralis_imp, i)
                 }

                 # avg of datasets
                 avg <- imp_sum / spectralis_imp$m

                 saveRDS(avg, file = paste0(wd_path,"processed_data/stratus_and_cirrus_to_spectralis_", method,"_imputedDF.rds") )
         }
     }

   lasso_norm_avg <- readRDS(paste0(wd_path,"processed_data/stratus_and_cirrus_to_spectralis_lasso.norm_imputedDF.rds"))
   pmm_avg <- readRDS(paste0(wd_path,"processed_data/stratus_and_cirrus_to_spectralis_pmm_imputedDF.rds"))
 }
```


## * plot correlations
```{r }

if  (impute_eval){
  # what method to plot
  # method = "PMM"
  # methods = c("Lasso.Norm", "PMM")
  methods = c("Lasso.Norm")

  # for eacah method of imputation, plot correlations between truth and imputed
  for(method in methods){

      RMSE_sum= 0
      rmses <- list()

      # plot
      png(file = paste0(wd_path, "results/mt_imputation/spectralis_by_stratus_and_cirrus_post_", method, "_interimputation_test.png"), height = h, width = w)# edited

      par(mfrow = c(3,3))
      for (i in 1:length(spectralis_cols)) {

        # column of original with values at sampling indices
        old <- d_truth[sample_idx, spectralis_cols[i]]


        # select which imputation method imputation average to plot against the truth
        if(method == "Lasso.Norm") new <- lasso_norm_avg[sample_idx,spectralis_cols[i]] else new <- pmm_avg[sample_idx,spectralis_cols[i]]

        # ranges of values
        y_range <- range(new)
        x_range <- range(old)

        # where text to go, top left
        text_x <- x_range[1] + .5*((x_range[2] - x_range[1])/2)
        text_y <- y_range[1] + 1.5*((y_range[2] - y_range[1])/2)

        # same range for each axis
        axis_range <- c(min(y_range[1], x_range[1]), max(y_range[2], x_range[2]))

        # Calculate Pearson correlation coefficient
        cor_value <- cor(old, new, method = "pearson")
        r <- cor_value
        rmse_i <- rmse(actual = scale(old), predicted = scale(new))

        title <-  paste0(method, " MT0", i, "; ", paste("r = ", round(r, 2)), paste("\nRMSE = ", round(rmse_i, 2)))

        # Plotting
        plot(new ~ old, main = title, xlim = axis_range, ylim = axis_range)
        text(paste("SPECTRALIS ~ STRATUS + CIRRUS", method), x = c(text_x), y = c(text_y), cex = .8 )
        abline(a=0, b=1, col="red")


        RMSE_sum <- RMSE_sum + rmse_i

        rmses[i] <-rmse_i

      }
      par(mfrow = c(1,1))
      dev.off()

      print(paste(method, "mean_RMSE : ", RMSE_sum / length(spectralis_cols) ))

  }
}

```
> lasso.norm performs better


## plot RMSE over n iterations

```{r }
if  (impute_eval){
  # dataframe to hold rmse values
  rmses <- data.frame(num_itr= integer(0),
                zone= integer(0),
                RMSE= integer(0))
  
  # params
  number_of_iterations = 20
  start = 6
  by = 6
  method = 'lasso.norm'
  
  # switch to execute imputations
  if (run_iter_eval){
      # impute 1:ith iteration and plot RMSE of each, for each zone
      for (i in seq.int(start,number_of_iterations, by = by)) {
      cat( paste("\nn(iterations):", i, ".......\n") )
  
          # impute test data for stratus from stratus
          imp <- mice(d_test, m = i, method = method, seed = 97, predictorMatrix=predictorMatrix)
  
          # find mean of imputed data
              imp_sum <- complete(imp, action = 1)
              for (i in 2:imp$m)  imp_sum <- imp_sum + complete(imp,action = i)
              avg <- imp_sum / imp$m
  
        for (j in 1:length(spectralis_cols)){
            # column of original with values at sampling indices
            old <- d_truth[sample_idx, spectralis_cols[j]] %>%
              round(. , 0)
  
            # column of imputed with values at sampling indices
            new <- avg[sample_idx,spectralis_cols[j]]%>%
              round(. , 0)
  
            rmses <- rmses %>%
              add_row(. ,num_itr = i,
                    zone = j,
                    RMSE = rmse(actual = scale(old), predicted = scale(new))
                    )
        }
  
      }
  
      # save imputation metrics
      saveRDS(rmses, paste0(wd_path, "processed_data/lasso.norm_model_rmses_50_iterations.rds"))
  
  }
  
  
  rmses <- readRDS(paste0(wd_path, "processed_data/lasso.norm_model_rmses_50_iterations.rds"))
  
  rmses$zone <- factor(rmses$zone)
  rmses %>% arrange(num_itr, zone)
  
  
  # plot rmses
  ## each line a zone x = iterations y = rmse
  my_colors <- brewer.pal(9, "Set1")
  
  ggplot(rmses, aes(group = zone, color = zone, x = num_itr, y = RMSE))+
    # geom_point()+
    geom_line()+
    geom_smooth(aes(group = 1), method = "loess", color = "black", se = TRUE) +
    scale_color_manual(values = my_colors)+
    theme_bw()
  ggsave(paste0(wd_path, "results/mt_imputation/RMSEs_over_iterations.png"), width = 5, height = 7)
  
}
```
> plotting RMSE for each zone actual vs. pred better measure of model accuracy. shows what number of iterations to use. Optimum number of iterations \~12

<!-- # Evaluate imputing each zone separately for all all machines to impute missing e.g. cirrus from spectralis -->
<!-- ## impute each zone for three machines -->
<!-- ```{r} -->

<!-- # for zone impute spec, cirr, strat, spec -->
<!-- # feed all mt into model, don't impute where entire dataset missing>> use output as new columns? -->
<!-- # or impute and reasign if original missing for all three set those rows to missing -->

<!-- # select just MT cols -->
<!-- truth_d <- inter_impute_d %>%  -->
<!--   select(matches("[cirrus|stratus|spectralis]_MT0\\d$")) -->

<!-- # for each zone, find rows where all machines NA -->
<!-- # n =  9 -->
<!-- zone_rows_na <- sapply(1:9, function(n){ -->

<!--   na_rows <- rowSums(is.na( truth_d[, grep(n, names(truth_d))] )) == 3 -->
<!--   # cbind(mt_d[, grep(n, names(mt_d))], na_rows) -->

<!--   return(na_rows) -->
<!-- }) %>%  -->
<!--   as.data.frame()  -->

<!-- # rename to zone names  -->
<!-- names(zone_rows_na) <- paste("MT0", 1:9, sep = "") -->

<!-- # row indices for originally NA  -->
<!-- was_NA <- is.na(truth_d) %>%  -->
<!--   as.data.frame() -->


<!-- # # sample observed idx and set those missing -->
<!-- # # loop through each machine and sample some observation indices -->
<!-- #  -->
<!-- # machines <- c("cirrus", "spectralis", "stratus") -->
<!-- # zones <- names(zone_rows_na) -->
<!-- #  -->
<!-- # # list of wehre observed for each col -->
<!-- # observed_idx <- lapply(machines, function(machine) { -->
<!-- #  -->
<!-- #   observations_at_machine_zone <- lapply(zones, function(zone) { -->
<!-- #     machine_zone <- paste(machine, zone, sep = "_") -->
<!-- #     return(which(!was_NA[[machine_zone]] & !zone_rows_na[[zone]])) -->
<!-- #   }) -->
<!-- #  -->
<!-- #   names(observations_at_machine_zone) <- zones -->
<!-- #   return(observations_at_machine_zone) -->
<!-- #  -->
<!-- # }) -->
<!-- # names(observed_idx) <- machines -->



<!-- #### IMPUTE each machine_zone #### -->
<!-- if(FALSE){ -->

<!-- # test df with each zone sample NA from each col -->
<!-- machine_zone <- "spectralis_MT05" -->

<!--   NA_idxs <-  imp_list <- list() ; for(machine_zone in names(truth_d)){ -->
<!--   # par(mfrow = c(2,13));for(ratio in seq(0, .5, .02)){ -->

<!--     a_machine <- strsplit(machine_zone, "_")[[1]][1] -->
<!--     a_zone <- strsplit(machine_zone, "_")[[1]][2] -->

<!--     other_machines <- names(truth_d) %>% -->
<!--       grep(a_machine, ., invert = TRUE, value = TRUE) %>% -->
<!--       strsplit(., "_") %>% -->
<!--       sapply(., '[', 1) %>% -->
<!--       unique() -->

<!--     other_machine_zones <- paste(other_machines, a_zone, sep = "_") -->

<!--     # where not na for this column -->
<!--     # where_observed <- observed_idx[[a_machine]][[a_zone]] -->
<!--     where_observed <- !is.na(truth_d[[machine_zone]]) -->
<!--     # truth_d[where_observed, ] -->

<!--     where_another <- !is.na(truth_d[[other_machine_zones[1]]]) | !is.na(truth_d[[other_machine_zones[2]]]) -->
<!--     # truth_d[where_another, ] -->

<!--     which_to_NA <- which(where_observed & where_another) -->
<!--     # truth_d[where_observed & where_another, c(machine_zone, other_machine_zones)] -->

<!--     # sampling where_observed -->
<!--     ratio = .1 -->
<!--     set.seed(2939) -->
<!--     sample_idx <- sample( which_to_NA, size = ratio*(length(which_to_NA))) -->


<!--     test_d <- truth_d -->
<!--     test_d <- test_d[, grep(a_zone, names(test_d))] -->
<!--     test_d[[machine_zone ]][ sample_idx] <- NA -->
<!--     # test_d[[machine_zone]][ sample_idx] -->

<!--     # impute zones for three machines -->
<!--     an_imp <-  mice(test_d, m = 10, method = "lasso.norm", seed = 345) -->

<!--     # average imputation dfs -->
<!--     an_imp_sum <- complete(an_imp, action = 1) ;  for (i in 2:an_imp$m) an_imp_sum <- an_imp_sum + complete(an_imp,action = i)  ;    an_avg <- an_imp_sum / an_imp$m -->

<!--     # set rows where all machines na for this zone to NA in imp_avg -->
<!--     an_avg[zone_rows_na[[a_zone]], ] <- NA -->
<!--     # an_avg[!zone_rows_na[[a_zone]], ] -->
<!--     # an_avg[zone_rows_na[[a_zone]], ] -->

<!--     # save imp and indxs to lists -->
<!--     imp_list[[machine_zone]] <- an_avg -->

<!--     NA_idxs[[machine_zone]] <- sample_idx -->

<!--   } -->
<!--   saveRDS(imp_list, paste0(wd_path,"processed_data/each_zone_imp_list.RDS")) -->
<!--   saveRDS(NA_idxs, paste0(wd_path,"processed_data/NA_idxs_list.RDS")) -->

<!-- } -->

<!-- imp_list <- readRDS(paste0(wd_path,"processed_data/each_zone_imp_list.RDS")) -->
<!-- NA_idxs <- readRDS(paste0(wd_path,"processed_data/NA_idxs_list.RDS")) -->

<!-- #### PLOT correlations #### -->
<!-- # machine_zone <- "spectralis_MT05" -->
<!-- png(paste0(wd_path, "results/mt_imputation/impute_by_zone_eval.png"), height = 11, width = 30, units = "in", res = 200) -->
<!-- par(mfrow = c(3,9));for (machine_zone in names(imp_list)){ -->

<!--   sample_idx <- NA_idxs[[machine_zone]] -->
<!--   num_nas <- length(sample_idx) -->

<!--   an_avg <- imp_list[[machine_zone]] -->
<!--   predicted <- an_avg[[machine_zone]][sample_idx] -->
<!--   actual <- truth_d[[machine_zone]][sample_idx ] -->

<!--   # ranges of values -->
<!--   y_range <- range(predicted) -->
<!--   x_range <- range(actual) -->

<!--   # where text to go, top left -->
<!--   text_x <- x_range[1] + .5*((x_range[2] - x_range[1])/2) -->
<!--   text_y <- y_range[1] + 1.5*((y_range[2] - y_range[1])/2) -->

<!--   # same range for each axis -->
<!--   axis_range <- c(min(y_range[1], x_range[1]), max(y_range[2], x_range[2])) -->

<!--   # Calculate Pearson correlation coefficient -->
<!--   cor_value <- cor(actual, predicted, method = "pearson") -->
<!--   r <- cor_value -->
<!--   rmse_i <- rmse(actual = scale(actual), predicted = scale(predicted)) -->
<!--   a_text <- paste0("r = ", round(r,2), "\nRMSE = ", round(rmse_i, 2)) -->

<!--   # Plotting -->
<!--   plot(predicted ~ actual, main = paste(machine_zone,"~ Machine_y_zone_i + Machine_z_zone_i\n(lasso.norm; n =", num_nas, ")"), xlim = axis_range, ylim = axis_range, cex.main = .8) -->
<!--   text(a_text,x = c(text_x), y = c(text_y), cex = .8 ) -->
<!--   abline(a=0, b=1, col="red") -->

<!-- };par(mfrow = c(1,1)) -->
<!-- dev.off() -->

<!-- ``` -->

# Rescale cirrus --> spectralis, stratus --> spectralis using distributions at overlap
```{r}

# where overlap between machines
stratus_spectralis_overlap <- rowSums(is.na(inter_impute_d[, c(stratus_cols, spectralis_cols)])) == 0
stratus_cirrus_overlap <- rowSums(is.na(inter_impute_d[, c(stratus_cols, cirrus_cols)])) == 0
spectralis_cirrus_overlap <- rowSums(is.na(inter_impute_d[, c(spectralis_cols, cirrus_cols)])) == 0

#___
#### rescale cirrus --> spectralis ####
#___

## calculate mean and sd for each zone when overlap between cirrus and spectralis
cirrus_means_where_spectralis <- colMeans(inter_impute_d[ spectralis_cirrus_overlap, c(cirrus_cols)])
cirrus_sds_where_spectralis <- sapply(inter_impute_d[ spectralis_cirrus_overlap, c(cirrus_cols)], sd)
spectralis_means_where_cirrus <- colMeans(inter_impute_d[ spectralis_cirrus_overlap, c(spectralis_cols)])
spectralis_sds_where_cirrus <- sapply(inter_impute_d[ spectralis_cirrus_overlap, c(spectralis_cols)], sd)

# ## whole distribution versus at overlap
# cirrus_means_where_spectralis <- colMeans(inter_impute_d[ , c(cirrus_cols)], na.rm = TRUE)
# cirrus_sds_where_spectralis <- sapply(inter_impute_d[ , c(cirrus_cols)], sd, na.rm = TRUE)
# spectralis_means_where_cirrus <- colMeans(inter_impute_d[ , c(spectralis_cols)], na.rm = TRUE)
# spectralis_sds_where_cirrus <- sapply(inter_impute_d[ , c(spectralis_cols)], sd, na.rm = TRUE)

# test_d
truth <-  inter_impute_d[, c(cirrus_cols, spectralis_cols)] %>% 
  na.omit()

# set.seed(2929)
# idxs_to_na <- sample(1:nrow(truth), .1*nrow(truth))
idxs_to_na <- 1:nrow(truth)
test <- truth
test[idxs_to_na , spectralis_cols ] <-  NA

## df of rescaled cirrus cols
cirrus_rescaled <- data.frame(matrix(nrow = nrow(test), ncol = 9, dimnames = list(NULL , cirrus_cols)))

## loop through each zone and rescale to spectralis
# a_cirrus = "cirrus_MT01"
for(a_cirrus in names(cirrus_rescaled)){
  
  a_spectralis <- grep(str_split(a_cirrus, "_", simplify = TRUE)[2], spectralis_cols, value = TRUE)
  a_cirrus_rescaled <- (scale(test[[a_cirrus]]) * spectralis_sds_where_cirrus[[a_spectralis]]) + spectralis_means_where_cirrus[[a_spectralis]]
  
  cirrus_rescaled[[a_cirrus]] <- a_cirrus_rescaled
  
}

png(paste0(wd_path, "results/mt_imputation/rescaling_to_spectralis_eval.png"), height = 500, width = 2000)
par(mfrow = c(2,9));
i = 1
# par(mfrow = c(3,3))
for(i in 1:9){
   
  predicted <- cirrus_rescaled[ idxs_to_na , i] 
  actual <-  truth[ idxs_to_na, spectralis_cols[i]]
  
  mod <- lm(predicted ~ actual)
  r <- cor(predicted , actual) %>% round(2)
  
  # ranges of values
  y_range <- range(predicted)
  x_range <- range(actual)
  
  min_range <- min(y_range[1] , x_range[1])
  max_range <- max(y_range[2] , x_range[2])
  
  # where text to go, top left
  text_x <- x_range[1] + .5*((x_range[2] - x_range[1])/1.8)
  text_y <- y_range[1] + 1.5*((y_range[2] - y_range[1])/2)

  
  plot(predicted ~ actual, 
       xlab = spectralis_cols[i], 
       ylab = paste("rescaled_cirrus_MT0", i),
       ylim = c(min_range, max_range),
       xlim =  c(min_range, max_range)
       )
  abline(c(0,1), lty = 3)
  abline(mod, col = "red")  
  text(paste0("r = ", r), pos = 2, y= text_y, x = text_x, col = "darkblue")
  # text(paste0("r^2 = ", r**2), pos = 2, y= text_y-15, x = text_x, col = "darkblue")

}; #par(mfrow = c(1,1))
  

#___
#### rescale stratus --> spectralis ####
#___
## calculate mean and sd for each zone when overlap between stratus and spectralis
stratus_means_where_spectralis <- colMeans(inter_impute_d[ stratus_spectralis_overlap, c(stratus_cols)])
stratus_sds_where_spectralis <- sapply(inter_impute_d[ stratus_spectralis_overlap, c(stratus_cols)], sd)
spectralis_means_where_stratus <- colMeans(inter_impute_d[ stratus_spectralis_overlap, c(spectralis_cols)])
spectralis_sds_where_stratus <- sapply(inter_impute_d[ stratus_spectralis_overlap, c(spectralis_cols)], sd)

## whole distribution versus  just at overlap
# stratus_means_where_spectralis <- colMeans(inter_impute_d[ , c(stratus_cols)], na.rm = TRUE)
# stratus_sds_where_spectralis <- sapply(inter_impute_d[ , c(stratus_cols)], sd, na.rm = TRUE)
# spectralis_means_where_stratus <- colMeans(inter_impute_d[ , c(spectralis_cols)], na.rm = TRUE)
# spectralis_sds_where_stratus <- sapply(inter_impute_d[ , c(spectralis_cols)], sd, na.rm = TRUE)

# test_d
truth <-  inter_impute_d[, c(stratus_cols, spectralis_cols)] %>% 
  na.omit()

set.seed(2929)
# idxs_to_na <- sample(1:nrow(truth), nrow(truth))
idxs_to_na <- 1:nrow(truth)
test <- truth
test[idxs_to_na , spectralis_cols ] <-  NA

## df of rescaled stratus cols
stratus_rescaled <- data.frame(matrix(nrow = nrow(test), ncol = 9, dimnames = list(NULL , stratus_cols)))

## loop through each zone and rescale to spectralis
# a_stratus = "stratus_MT01"
for(a_stratus in names(stratus_rescaled)){
  
  a_spectralis <- grep(str_split(a_stratus, "_", simplify = TRUE)[2], spectralis_cols, value = TRUE)
  a_stratus_rescaled <- (scale(test[[a_stratus]]) * spectralis_sds_where_stratus[[a_spectralis]]) + spectralis_means_where_stratus[[a_spectralis]]
  
  stratus_rescaled[[a_stratus]] <- a_stratus_rescaled
  
}

# par(mfrow = c(3,3));
for(i in 1:9){
  
  predicted <- stratus_rescaled[ idxs_to_na , i]
  actual <- truth[ idxs_to_na, spectralis_cols[i]]
  
  mod <- lm(predicted ~ actual)
  r <- cor(predicted , actual) %>% round(2)
  
  # ranges of values
  y_range <- range(predicted)
  x_range <- range(actual)

  min_range <- min(y_range[1] , x_range[1])
  max_range <- max(y_range[2] , x_range[2])
  
  # where text to go, top left
  text_x <- x_range[1] + .5*((x_range[2] - x_range[1])/1.8)
  text_y <- y_range[1] + 1.5*((y_range[2] - y_range[1])/2)

  
  plot(predicted ~ actual, 
       xlab = spectralis_cols[i], 
       ylab = paste("rescaled_stratus_MT0", i),
       ylim = c(min_range, max_range),
       xlim =  c(min_range, max_range)
       )
  abline(c(0,1), lty = 3)
  abline(mod, col = "red")  
  text(paste0("r = ", r), pos = 2, y= text_y, x = text_x, col = "darkblue")
  # text(paste0("r^2 = ", r**2), pos = 2, y= text_y-15, x = text_x, col = "darkblue")
  
}; par(mfrow = c(1,1))

dev.off()


```

#\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# Impute missing spectralis with optimized methods
## TRUE impute to working d
```{r}

#
##
###
####
#####
impute <- TRUE  
#####
####
###
##
#

```

### impute only spectralis using predictor matrix
<!-- ```{r} -->


<!-- # feed in the data where cirrus or stratus full accross, with spectralis there or not -->
<!-- row_idx_to_impute <- (rowSums(is.na(inter_impute_d[, stratus_cols])) == 0 ) | (rowSums(is.na(inter_impute_d[, cirrus_cols])) == 0 ) -->

<!-- # dataset to impute with: excludes wehre no cirrus or no stratus -->
<!-- d_in <- inter_impute_d[ row_idx_to_impute , mt_cols ] -->


<!-- # parameters -->
<!-- method = 'lasso.norm' -->
<!-- m = 12 -->

<!-- # true to impute values -->
<!-- if(impute){ -->
<!--   imp <- mice(d_in, m = m, method = method, seed = 2654, predictorMatrix=predictorMatrix) -->
<!--   saveRDS(imp, paste0(wd_path, "processed_data/imputations/spectralis_by_cirrus_and_stratus_final_imp.rds")) -->
<!-- } -->

<!-- imp <- readRDS(paste0(wd_path, "processed_data/imputations/spectralis_by_cirrus_and_stratus_final_imp.rds")) -->

<!-- # find mean of imputed data -->
<!-- imp_sum <- complete(imp, action = 1) ;  for (i in 2:imp$m) imp_sum <- imp_sum + complete(imp,action = i)  ;    avg <- imp_sum / imp$m -->

<!-- # set spectralis cols at idxs to imputed average spectralis cols -->
<!-- new_d <- inter_impute_d -->
<!-- new_d[row_idx_to_impute, spectralis_cols] <- avg[, spectralis_cols] -->

<!-- # add variable of whether spectralis values imputed; not rowidx b/c it includes filled spectralis -->
<!-- ## where is missing and cirrus or stratus not missing -->
<!-- where_spectralis_imputed <- is.na(inter_impute_d$spectralis_MT01) & (!is.na(inter_impute_d$cirrus_MT01) | !is.na(inter_impute_d$stratus_MT01)) -->
<!-- new_d$MT_spectralis_imputed[where_spectralis_imputed] <- TRUE -->
<!-- new_d$MT_spectralis_imputed[is.na(new_d$MT_spectralis_imputed)] <- FALSE -->
<!-- ``` -->


### impute all with predictor matrix

<!-- ```{r} -->

<!-- # d of just mt cols -->
<!-- mt_d <- inter_impute_d %>%   -->
<!--    # select(matches("[cirrus|stratus|spectralis]_MT0\\d$"))  -->
<!--   select(all_of(mt_cols)) -->

<!-- # where all machines for particular zone NA -->
<!-- zone_rows_na <- sapply(1:9, function(n){  -->

<!--    na_rows <- rowSums(is.na( mt_d[, grep(n, names(mt_d))] )) == 3  -->
<!--    # cbind(mt_d[, grep(n, names(mt_d))], na_rows) --> -->

<!--   return(na_rows)  -->
<!--   }) %>%   -->
<!--   as.data.frame()  -->
<!-- names(zone_rows_na) <- paste("MT0", 1:9, sep = "") -->
<!-- # # look at where not all zones na across all machines -->
<!-- # x <- mt_d[which(rowSums(zone_rows_na) != 9 & rowSums(!zone_rows_na) != 9), ] -->
<!-- # x <- x[, order(substr(names(x), nchar(names(x)), nchar(names(x))) )] -->
<!-- # glimpse(x) -->

<!-- # parameters -->
<!-- method = 'lasso.norm' -->
<!-- m = 12 -->

<!-- # true to impute values -->
<!-- if(impute){ -->
<!--   d_in <- inter_impute_d[, names(mt_d)] -->
<!--   imp <- mice(d_in, m = m, method = method, seed = 2654, predictorMatrix=predictorMatrix) -->
<!--   saveRDS(imp, paste0(wd_path, "processed_data/imputations/all_machines_final_imp.rds")) -->
<!-- } -->

<!-- imp <- readRDS(paste0(wd_path, "processed_data/imputations/all_machines_final_imp.rds")) -->

<!-- # find mean of imputed data -->
<!-- imp_sum <- complete(imp, action = 1) ;  for (i in 2:imp$m) imp_sum <- imp_sum + complete(imp,action = i)  ;    avg <- imp_sum / imp$m -->

<!-- # set what was NA for all machines at zone_n to NA again, these can not be imputed -->
<!-- for(zone in names(zone_rows_na)){ -->

<!--   # where all machines NA at 'zone', at 'zone' for all machines, set NA -->
<!--   avg[ zone_rows_na[[zone]] , grep(zone, names(avg))] <- NA -->

<!-- } -->

<!-- # set spectralis cols at idxs to imputed average spectralis cols -->
<!-- new_d <- inter_impute_d -->
<!-- new_d[, names(mt_d)] <- avg -->

<!-- # add variable of whether spectralis values imputed; not rowidx b/c it includes filled spectralis -->
<!-- ## where is missing in pre_impute and not in post impute -->
<!-- where_MT_not_imputed <- (  new_d[, names(mt_d)] %>%  mutate_all(~!is.na(.))  ) == ( inter_impute_d[, names(mt_d)] %>% mutate_all(~!is.na(.)) ) -->
<!-- where_MT_imputed <- data.frame(!where_MT_not_imputed) -->

<!-- # coalesce each machine's 9 zones to one col -->
<!-- where_MT_imputed_coalesced <- where_MT_imputed %>% -->
<!--   mutate( -->
<!--       cirrus_imputed = if_any(starts_with("cirrus"), ~ .x == TRUE), -->
<!--       spectralis_imputed = if_any(starts_with("spectralis"), ~ .x == TRUE), -->
<!--       stratus_imputed = if_any(starts_with("stratus"), ~ .x == TRUE) -->
<!--     ) %>% -->
<!--   select(matches("imputed")) -->
<!-- # new_d[  where_MT_imputed$stratus_imputed, names(mt_d)] -->

<!-- # add was imputed cols to main d -->
<!-- new_d <- cbind(new_d, where_MT_imputed_coalesced) -->

<!-- new_d %>% -->
<!--     select(matches("[cirrus|stratus|spectralis]_MT0\\d$|[cirrus|stratus|spectralis]_imputed")) %>% -->
<!--     # filter(cirrus_imputed & stratus_imputed & spectralis_imputed) -->
<!--     filter(cirrus_imputed & stratus_imputed & !spectralis_imputed) -->

<!-- ``` -->


## impute each zone for three machines
<!-- ```{r} -->

<!-- # for zone impute spec, cirr, strat, spec -->
<!-- # feed all mt into model, don't impute where entire dataset missing>> use output as new columns? -->
<!-- # or impute and reasign if original missing for all three set those rows to missing -->

<!-- # select just MT cols -->
<!-- mt_d <- inter_impute_d %>%  -->
<!--   select(matches("[cirrus|stratus|spectralis]_MT0\\d$")) -->

<!-- # for each zone, find rows where all machines NA -->
<!-- # n =  9 -->
<!-- zone_rows_na <- sapply(1:9, function(n){ -->

<!--   na_rows <- rowSums(is.na( mt_d[, grep(n, names(mt_d))] )) == 3 -->
<!--   # cbind(mt_d[, grep(n, names(mt_d))], na_rows) -->

<!--   return(na_rows) -->
<!-- }) %>%  -->
<!--   as.data.frame()  -->

<!-- # rename to zone names  -->
<!-- names(zone_rows_na) <- paste("MT0", 1:9, sep = "") -->

<!-- # row indices for originally NA -->
<!-- was_NA <- is.na(mt_d) %>% -->
<!--   as.data.frame() -->

<!-- # loop through each zone, impute all machinees for that zone, set NA indices where all machines NA for that zone -->
<!-- if(impute){ -->
<!--  machine_zone <- "spectralis_MT04" -->
<!--  imp_results <- data.frame() -->
<!--  for(machine_zone in names(mt_d)){ -->

<!--     cat() -->
<!--     print(machine_zone) -->
<!--     cat() -->

<!--     a_machine <- strsplit(machine_zone, "_")[[1]][1] -->
<!--     a_zone <- strsplit(machine_zone, "_")[[1]][2] -->

<!--     other_machines <- names(mt_d) %>% -->
<!--       grep(a_machine, ., invert = TRUE, value = TRUE) %>% -->
<!--       strsplit(., "_") %>% -->
<!--       sapply(., '[', 1) %>% -->
<!--       unique() -->

<!--     other_machine_zones <- paste(other_machines, a_zone, sep = "_") -->

<!--     # subset mt_d for zone -->
<!--     mt_d_at_zone <- mt_d[, grep(a_zone, names(mt_d))] -->

<!--     # impute zones for three machines -->
<!--     an_imp <-  mice(mt_d_at_zone, m = 3, method = "lasso.norm", seed = 345) -->

<!--     # average imputation dfs -->
<!--     an_imp_sum <- complete(an_imp, action = 1) ;  for (i in 2:an_imp$m) an_imp_sum <- an_imp_sum + complete(an_imp,action = i)  ;    an_avg <- an_imp_sum / an_imp$m -->


<!--     # set rows where all machines na for this zone to NA in imp_avg -->
<!--     an_avg[zone_rows_na[[a_zone]], ] <- NA -->
<!--     # View(an_avg[!zone_rows_na[[a_zone]], ] ) -->
<!--     # View(mt_d_at_zone[!zone_rows_na[[a_zone]], ]  ) -->
<!--     # View(an_avg[zone_rows_na[[a_zone]], ] ) -->
<!--     # View(mt_d_at_zone[zone_rows_na[[a_zone]], ] ) -->

<!--     # add cols to imp_results    -->
<!--     if(all(dim(imp_results)  == 0)){ -->

<!--       imp_results <- as.data.frame(an_avg %>% select(all_of(machine_zone))) -->

<!--       }else{ -->

<!--         imp_results <- cbind(imp_results, -->
<!--                              as.data.frame(an_avg %>% select(all_of(machine_zone))) -->
<!--                              ) -->

<!--       } -->

<!--  } -->
<!--   saveRDS(imp_results, paste0(wd_path,"processed_data/each_zone_imp_results.RDS")) -->
<!--  } -->

<!-- imp_results <- readRDS(paste0(wd_path,"processed_data/each_zone_imp_results.RDS")) -->

<!-- ### results df -->
<!-- #     imp_results <- data.frame() -->
<!-- #      -->
<!-- #     # zone  = "MT01" -->
<!-- #     for(zone in names(zone_rows_na)){ -->
<!-- #        -->
<!-- #       # select all machines for zone i -->
<!-- #       mt_d_at_zone <- mt_d[, grep(zone, names(mt_d))] -->
<!-- #        -->
<!-- #       # impute these zones for three machines -->
<!-- #       an_imp <-  mice(mt_d_at_zone, m = 5, method = "lasso.norm", seed = 345) -->
<!-- #        -->
<!-- #       # average imputation dfs -->
<!-- #       an_imp_sum <- complete(an_imp, action = 1) ;  for (i in 2:an_imp$m) an_imp_sum <- an_imp_sum + complete(an_imp,action = i)  ;    an_avg <- an_imp_sum / an_imp$m  -->
<!-- #       # cbind(an_avg, mt_d_at_zone,zone_rows_na$MT01) -->
<!-- #       ##^^ imputes where all machines NA -->
<!-- #      -->
<!-- #       # set rows where all machines na for this zone to NA in imp_avg -->
<!-- #       an_avg[zone_rows_na[[zone]], ] <- NA -->
<!-- #        -->
<!-- #       if(all(dim(imp_results)  == 0)){ -->
<!-- #         imp_results <- an_avg -->
<!-- #          -->
<!-- #       }else{ -->
<!-- #         imp_results <- cbind(imp_results, an_avg) -->
<!-- #       } -->
<!-- #        -->
<!-- #     } -->
<!-- # } -->

<!-- # add to new_d -->
<!-- new_d <- inter_impute_d  -->
<!-- new_d[, names(mt_d)] <- imp_results -->

<!-- a_col <- "cirrus_MT03" -->
<!-- a_zone <- strsplit(a_col, "_")[[1]][2] -->

<!-- # original versus new where was NA and !all_rows_NA -->
<!-- inter_impute_d[was_NA[[a_col]] & !zone_rows_na[[a_zone]], names(mt_d)] -->
<!-- new_d[was_NA[[a_col]] & !zone_rows_na[[a_zone]], names(mt_d)] -->
<!-- # z <- data.frame(cbind(new_d[, names(mt_d)],  inter_impute_d [, names(mt_d)]) ) -->
<!-- # z <- z[, sort(names(z))] -->
<!-- # View(z) -->

<!-- ``` -->

## rescale stratus and cirrus to spectralis based on sd and mean where overlap 
```{r}


# where overlap between machines
stratus_spectralis_overlap <- rowSums(is.na(inter_impute_d[, c(stratus_cols, spectralis_cols)])) == 0
stratus_cirrus_overlap <- rowSums(is.na(inter_impute_d[, c(stratus_cols, cirrus_cols)])) == 0
spectralis_cirrus_overlap <- rowSums(is.na(inter_impute_d[, c(spectralis_cols, cirrus_cols)])) == 0



#___
#### rescale cirrus --> spectralis ####
#___

## calculate mean and sd for each zone when overlap between cirrus and spectralis
cirrus_means_where_spectralis <- colMeans(inter_impute_d[ spectralis_cirrus_overlap, c(cirrus_cols)])
cirrus_sds_where_spectralis <- sapply(inter_impute_d[ spectralis_cirrus_overlap, c(cirrus_cols)], sd)
spectralis_means_where_cirrus <- colMeans(inter_impute_d[ spectralis_cirrus_overlap, c(spectralis_cols)])
spectralis_sds_where_cirrus <- sapply(inter_impute_d[ spectralis_cirrus_overlap, c(spectralis_cols)], sd)

## df of rescaled cirrus cols
cirrus_rescaled <- data.frame(matrix(nrow = nrow(inter_impute_d), ncol = 9, dimnames = list(NULL , cirrus_cols)))

## loop through each zone and rescale to spectralis
# a_cirrus = "cirrus_MT01"
for(a_cirrus in names(cirrus_rescaled)){
  
  a_spectralis <- grep(str_split(a_cirrus, "_", simplify = TRUE)[2], spectralis_cols, value = TRUE)
  a_cirrus_rescaled <- (scale(inter_impute_d[[a_cirrus]]) * spectralis_sds_where_cirrus[[a_spectralis]]) + spectralis_means_where_cirrus[[a_spectralis]]
  
  cirrus_rescaled[[a_cirrus]] <- as.numeric(a_cirrus_rescaled)
  
}
names(cirrus_rescaled) <- paste0("rescaled_", names(cirrus_rescaled))

#___
#### rescale stratus --> spectralis ####
#___
## calculate mean and sd for each zone when overlap between stratus and spectralis
stratus_means_where_spectralis <- colMeans(inter_impute_d[ stratus_spectralis_overlap, c(stratus_cols)])
stratus_sds_where_spectralis <- sapply(inter_impute_d[ stratus_spectralis_overlap, c(stratus_cols)], sd)
spectralis_means_where_stratus <- colMeans(inter_impute_d[ stratus_spectralis_overlap, c(spectralis_cols)])
spectralis_sds_where_stratus <- sapply(inter_impute_d[ stratus_spectralis_overlap, c(spectralis_cols)], sd)


## df of rescaled stratus cols
stratus_rescaled <- data.frame(matrix(nrow = nrow(inter_impute_d), ncol = 9, dimnames = list(NULL , stratus_cols)))

## loop through each zone and rescale to spectralis
# a_stratus = "stratus_MT01"
for(a_stratus in names(stratus_rescaled)){
  
  a_spectralis <- grep(str_split(a_stratus, "_", simplify = TRUE)[2], spectralis_cols, value = TRUE)
  a_stratus_rescaled <- (scale(inter_impute_d[[a_stratus]]) * spectralis_sds_where_stratus[[a_spectralis]]) + spectralis_means_where_stratus[[a_spectralis]]
  
  stratus_rescaled[[a_stratus]] <- as.numeric(a_stratus_rescaled )
  
}
names(stratus_rescaled) <- paste0("rescaled_", names(stratus_rescaled))



#___
#### rescale spectralis --> stratus ####
#___

## calculate mean and sd for each zone when overlap between spectralis and stratus
spectralis_means_where_stratus <- colMeans(inter_impute_d[ stratus_spectralis_overlap, c(spectralis_cols)])
spectralis_sds_where_stratus <- sapply(inter_impute_d[ stratus_spectralis_overlap, c(spectralis_cols)], sd)
stratus_means_where_spectralis <- colMeans(inter_impute_d[ stratus_spectralis_overlap, c(stratus_cols)])
stratus_sds_where_spectralis <- sapply(inter_impute_d[ stratus_spectralis_overlap, c(stratus_cols)], sd)


## df of rescaled spectralis cols
spectralis_rescaled <- data.frame(matrix(nrow = nrow(inter_impute_d), ncol = 9, dimnames = list(NULL , spectralis_cols)))

## loop through each zone and rescale to stratus
# a_spectralis = "spectralis_MT01"
for(a_spectralis in names(spectralis_rescaled)){
  
  a_stratus <- grep(str_split(a_spectralis, "_", simplify = TRUE)[2], stratus_cols, value = TRUE)
  a_spectralis_rescaled <- (scale(inter_impute_d[[a_spectralis]]) * stratus_sds_where_spectralis[[a_stratus]]) + stratus_means_where_spectralis[[a_stratus]]
  
  spectralis_rescaled[[a_spectralis]] <- as.numeric(a_spectralis_rescaled )
  
}
names(spectralis_rescaled) <- paste0("rescaled_", names(spectralis_rescaled))

# x <- na.omit(cbind(spectralis_rescaled, inter_impute_d[, stratus_cols]) )
# x <- x[, order(substr(names(x), nchar(names(x)) , nchar(names(x))))]
# 
# boxplot(x)




#___
#### join rescaled cirrus and stratus to main ####
#___

new_d <- inter_impute_d
new_d <- new_d %>% cbind(stratus_rescaled, cirrus_rescaled, spectralis_rescaled)


#___
####  calculate average of spectralis and rescaled cirrus and stratus 
#___

## cols to use for average
cols_for_average <- c(spectralis_cols, names(stratus_rescaled), names(cirrus_rescaled))

## for each zone, rowmean of spectralis, rescaled cirrus, and rescaled stratus
for(i in 1:9){
  
  vars <- grep(i, cols_for_average, value = TRUE)
  
  new_col_name <- paste0("avg_MT0", i)
  
  new_d[[new_col_name]] <- rowMeans(new_d[, vars], na.rm = TRUE)
  
}

avg_cols <- grep("avg_MT0", names(new_d), value = TRUE)

# add variable of whether average values are rescaled versions of cirrus or stratus


## where is missing in pre_impute and not in post impute
where_MT_not_imputed <- ( new_d[, spectralis_cols] %>% mutate_all(~!is.na(.)) ) == (  new_d[, avg_cols] %>%  mutate_all(~!is.na(.))  ) 
where_MT_imputed <- data.frame(!where_MT_not_imputed)

# row any of where_MT_imputed 
where_any_MT_imputed <- apply(where_MT_imputed, 1, any)

new_d[  where_any_MT_imputed , c(cols_for_average, avg_cols )] %>% glimpse()

# add was imputed cols to main d
new_d$rescaled_to_spectralis <- where_any_MT_imputed


new_d %>% 
  select(all_of(c(cols_for_average, avg_cols )) , rescaled_to_spectralis) %>% 
  filter(rescaled_to_spectralis)
```

```{r}

# plot bars of MT vars after rescaling
pdf(paste0(wd_path, "results/mt_imputation/post_rescaling_count_observed_mt_cols.pdf"), height = 11, width = 10)
new_d %>% 
  select(matches("MT0\\d$")) %>% 
  select(-matches("rescaled")) %>% 
  summarize_all(~sum(!is.na(.))) %>% 
  pivot_longer(everything()) %>% 
  mutate(machine = str_split(name, "_MT") %>% sapply(`[`,1)) %>% 
  mutate(zone = str_split(name, "_MT") %>% sapply(`[`,2)) %>% 
  mutate(machine = case_when(machine == "avg" ~ "combination", TRUE ~ machine)) %>% 
  mutate(machine = factor(machine, levels = c("cirrus", "spectralis", "stratus", "combination"))) %>% 
  ggplot(aes(x = machine, fill = factor(zone), y= value))+
  geom_bar(stat = "identity", position = position_dodge())+
  theme(axis.text.x = element_text(angle = 0))+
  labs(color = "zone")+
  ggtitle("Number of available values for each machine/zone after rescaling to Spectralis")
dev.off()
  
```


```{r}
plotting_d <- new_d %>% 
  select(all_of(mt_cols), matches("rescaled|avg_")) %>% 
  # mutate(colors = colors) %>% 
  mutate(colors = ifelse(rescaled_to_spectralis, "blue", "black")) %>% 
  # filter(!cirrus_imputed & spectralis_imputed & stratus_imputed) %>%  # cirrus
  # filter(cirrus_imputed & !spectralis_imputed & stratus_imputed) %>%  # spec
  # filter(cirrus_imputed & spectralis_imputed & !stratus_imputed) %>%  # strat
  # filter(!cirrus_imputed & !spectralis_imputed & stratus_imputed) %>%  # cirr and spec
  # filter(!cirrus_imputed & spectralis_imputed & !stratus_imputed) %>% # cirr and strat
  # filter(cirrus_imputed & !spectralis_imputed & !stratus_imputed) %>% # spec and strat
  filter()

##### plot spectralis by cirrus with r value for nine zones ####
pch = 1

pdf(file = paste0(wd_path, "results/mt_imputation/all_machines_post_lasso.norm_interimputation.pdf"), height = 15, width = 30)

par(mfrow = c(3,9))
# par(mfrow = c(3,3))
for ( i in 1:9){

  avg <- plotting_d[, avg_cols[i]]
  cirrus <- plotting_d[, cirrus_cols[i]]
  mod <- lm(avg~cirrus)

  title <- paste("MT zone", i ,  "; r = ", round(cor( avg,  cirrus,use = "pair"), 2))

  # ranges of values
  y_range <- range(avg, na.rm = TRUE)
  x_range <- range(cirrus, na.rm = TRUE)

  # where text to go, top left
  text_x <- x_range[1] + .5*((x_range[2] - x_range[1])/2)
  text_y <- y_range[1] + 1.5*((y_range[2] - y_range[1])/2)

  # same range for each axis
  axis_range <- c(min(y_range[1], x_range[1]), max(y_range[2], x_range[2]))


  avg_rug <- avg[!is.na(avg) & !is.na(cirrus)]
  cirrus_rug <- cirrus[!is.na(avg) & !is.na(cirrus)]
  
  # cbind(new_d[, grep("[cirrus|stratus|avg]_imputed", names(new_d))], colors) %>% unique
  
  plot(avg ~ cirrus,
       main = title ,
       xlim = axis_range,
       ylim = axis_range, 
       col = plotting_d$colors,
       pch = pch
       )
  
  abline(c(0,1), col = "grey", lty = 2)
  abline(mod, col = "red")
  rug(avg_rug, side = 2, col = "darkgreen")
  rug(cirrus_rug, col = "blue")
  text(x = text_x, y = text_y, labels = c("post-imputation"))

}


##### plot avg by stratus with r value for nine zones ####
# par(mfrow = c(3,3))
for ( i in 1:9){

  avg <- plotting_d[, avg_cols[i]]
  stratus <- plotting_d[, stratus_cols[i]]
  mod <- lm(avg~stratus)

  title <- paste("MT zone", i ,  "; r = ", round(cor( avg,  stratus,use = "pair"), 2))

  # ranges of values
  y_range <- range(avg, na.rm = TRUE)
  x_range <- range(stratus, na.rm = TRUE)

  # where text to go, top left
  text_x <- x_range[1] + .5*((x_range[2] - x_range[1])/2)
  text_y <- y_range[1] + 1.5*((y_range[2] - y_range[1])/2)

  # same range for each axis
  axis_range <- c(min(y_range[1], x_range[1]), max(y_range[2], x_range[2]))

  avg_rug <- avg[!is.na(avg) & !is.na(stratus)]
  stratus_rug <- stratus[!is.na(avg) & !is.na(stratus)]

  plot(avg ~ stratus,
       main = title ,
       xlim = axis_range,
       ylim = axis_range, 
       col = plotting_d$colors,
       pch = pch
       )
  abline(c(0,1), col = "grey", lty = 2)
  abline(mod, col = "red")
  rug(avg_rug, side = 2, col = "darkgreen")
  rug(stratus_rug, col = "blue")
  text(x = text_x, y = text_y, labels = c("post-imputation"))

}

##### plot spectralis by avg with r value for nine zones ####
# par(mfrow = c(3,3))
for ( i in 1:9){

  spectralis <- plotting_d[, spectralis_cols[i]]
  avg <- plotting_d[, avg_cols[i]]
  mod <- lm(spectralis~avg)

  title <- paste("MT zone", i ,  "; r = ", round(cor( spectralis,  avg,use = "pair"), 2))

  # ranges of values
  y_range <- range(spectralis, na.rm = TRUE)
  x_range <- range(avg, na.rm = TRUE)

  # where text to go, top left
  text_x <- x_range[1] + .5*((x_range[2] - x_range[1])/2)
  text_y <- y_range[1] + 1.5*((y_range[2] - y_range[1])/2)

  # same range for each axis
  axis_range <- c(min(y_range[1], x_range[1]), max(y_range[2], x_range[2]))

  spectralis_rug <- spectralis[!is.na(spectralis) & !is.na(avg)]
  avg_rug <- avg[!is.na(spectralis) & !is.na(avg)]

  plot(spectralis ~ avg,
       main = title ,
       xlim = axis_range,
       ylim = axis_range, 
       col = plotting_d$colors,
       pch = pch
       )
  
  abline(c(0,1), col = "grey", lty = 2)
  abline(mod, col = "red")
  rug(spectralis_rug, side = 2, col = "darkgreen")
  rug(avg_rug, col = "blue")
  text(x = text_x, y = text_y, labels = c("post-imputation"))

}
par(mfrow = c(1,1))

dev.off()
```


# make df of spectralis x avg correlation
```{r}

zone = "MT01"
cors <- sapply(paste0("MT0", 1:9), function(zone){
  spec <- new_d[, sprintf("spectralis_%s", zone)]
  avg <- new_d[, sprintf("avg_%s", zone)]
  cor_test <- cor.test(spec,avg, use = "complete.obs")
  c(p_value = cor_test$p.value, R = as.numeric(cor_test$estimate))
}) %>% 
  t() %>% 
  as.data.frame()
cors$variable <- rownames(cors)

# save to cors folder
if(!dir.exists(sprintf("%sprocessed_data/imputation_Rs/", wd_path))) dir.create(sprintf("%sprocessed_data/imputation_Rs/", wd_path))
write_csv(cors, sprintf("%sprocessed_data/imputation_Rs/MT_rescaling_correlations.csv", wd_path))

```
> imputation works differently for each machine because pulling from different number of overlaps 

## image of full MT zones after prediction
only rows with full across shown
```{r }

# m <- new_d[ , c(grep("sample", names(new_d)), mt_cols)]
# rownames(m) <- m$sample
# m$sample <- NULL
# m <- m %>% t()
#
# # select rows wehre at least 1 non-NA present
# m_full2 <-m[, apply(m, 2, function(x) any(!is.na(x)))]
#
# # image matrix order l>r cirrus, stratus, spectralis
# image(m_full2, col = "red", main = "image of MT columns after predictions")


m <-new_d[ , c("sample", mt_cols, avg_cols)]
rownames(m) <- m$sample
m$sample <- NULL

## order by zone
# m <- m[ order(substr(colnames(m), nchar(colnames(m)), nchar(colnames(m))) )]

m <- m %>% t()

# select columns wehre at least 1 non-NA present
m_full <- m[, apply(m, 2, function(x) any(!is.na(x)))]

# image matrix order l>r cirrus, stratus, spectralis
pdf(paste0(wd_path, "results/mt_imputation/MT_cols_post_intercategory_imputations.pdf"), height = 11, width = 10)
Heatmap(m_full %>% t(),
        show_heatmap_legend = TRUE,
        name = "thickness",
        use_raster = F,
        column_title  = "MT columns after inter-category imputations",
        row_title  = "observations",
        # col = c("white" ,  "brown"),
        col = viridis(100),
        na_col = "white",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_row_names = FALSE,
        show_column_names  = TRUE,
        column_names_rot = 45
)
dev.off()

```

# some more cleaning
<!-- ## Calculate foveal slope -->
<!-- ```{r} -->

<!-- # use avg columns -->
<!-- inner_zones <- grep("avg_MT0", names(new_d), value = TRUE)[1:5] -->

<!-- # calculation function -->
<!-- ## difference in thickness over distance to zone (convert from rad to deg) -->
<!-- calc_slope <- function(sector_of_interest, sector_one){  -->
<!--     180 *  (atan((sector_of_interest - sector_one) / 500)) / pi -->
<!--   } -->


<!-- # for each parafoveal zone, calculate slope against zone 1 -->
<!-- new_d <- new_d %>%  -->
<!--   mutate(across(all_of(inner_zones[-1]),  -->
<!--                 ~calc_slope(., avg_MT01),  -->
<!--                 .names = "{.col}_slope" -->
<!--                 ))  -->

<!-- ``` -->

## remove unwanted MT cols
```{r}
# # remove stratus and cirrus
# what_to_remove <- grep("cirrus|stratus", names(new_d), value = TRUE, ignore.case = TRUE) %>%
#   grep("SCANDT|DATE|rescaled_to", ., value = TRUE, invert = TRUE)
# 
# new_d <-new_d %>%
#   select(-all_of(what_to_remove))
```

# \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# save RDS
data > d > intra_impute_d > inter_impute_d > new_d
```{r}
new_data_name <- data_name %>% 
  gsub("2", "3", .)
new_d_path <- paste0(wd_path,"processed_data/", new_data_name) 

print(paste("Saving...", new_data_name))
saveRDS(new_d, new_d_path)
```
CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_03.rds
CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_03.rds

```{r}
check_dims(D = new_d)
# dims : 12600 288
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516


### MARCH 12 2025 ###
# dims : 12619 288
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```
