---
title: "03_EDA exploration continued"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---

# packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

library(ggplot2)
library(tidyverse)
library(dplyr)
library(reshape2)
library(gridExtra)
library(viridis)
library(DescTools)
library(SummarizedExperiment)
library(corrplot)
library(knitr)




rm(list = ls())
gc()

# my functions
source("~/Library/CloudStorage/OneDrive-LowyMedicalResearchInstitute/Database_EDA/analysis_versions/version015/analysis/00_MyFunctions.R")

```

# Load Data
```{r}
wd_path <- "~/Documents/LMRI/PRJ2024001/version015/"


data_name <- "CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_03.rds"
d <- readRDS(paste0(wd_path,"processed_data/", data_name))

raw <- readRDS(paste0(wd_path,"processed_data/dfs_list_01.rds"))

# grading nomenclature and information
path <- paste0(wd_path,"processed_data/database_grading_key_complete.xlsx")
info <- readxl::read_xlsx(path) %>% as.data.frame()

```
# = = = = = = = = = = = = = = = = = = = = = = 

# clean data

```{r, look at dims}

check_dims <- function(D = d){
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()

# dims : 12600 288
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516


### MARCH 12 2025 ###
# dims : 12619 288
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```


## fix ordinality
* subset for diagnosis
* make image quality  factor grades, maintain levels of increase, collapse scores that essentially mean the same thing,  NA for 5<
    * DRV: ?
    * YELLRND: maintain 0-4
    * NEOVASC: 2< ==>2 
    * FAZ: ?


### Fix ordinality
#### make zonal vars numeric
```{r}
# for the categories with <phenotype>01-09
## select the spatial phenotypes
zonal_vars <-  grep("\\D0\\d$", names(d), value = TRUE)

# make zonal vars numeric
d <- d %>% 
  mutate(across(all_of(zonal_vars), 
                ~ as.numeric(.)))
```


#### make separate ISOS loc vars
    
    Add ISOS_LOC. Make it dichotomous{ISOS_BREAK CENTRAL, ISOS_BREAK_JUXTAFOVEAL
```{r}

info %>% filter(grepl("ISOS", name)) %>% select(name, description, `0`:`6`)


# # check isos_loc coding
# View(read.xlsx("/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/info/Grading fields and variables_edited.xlsx", sheet = 2))

# make juxtafoveal and central isos location variables
d$ISOS_LOC_JF <- as.numeric(d$ISOS_LOC == 2)
d$ISOS_LOC_CEN <- as.numeric(d$ISOS_LOC == 1)

d %>% 
  select(matches("ISOS")) %>% 
  distinct() %>% 
  head(10)

# remove isos_loc
d$ISOS_LOC <- NULL

```


#### enforce ordinality
      
    - make other disease character/nominal
    - reorder FAZ to correlate to severity
    - cap yellowrnd at 2
    - make BLRPATTERN dichotomous
    - with vars that go over 2 {ICL_TYPE, TYP_LEAK_RPE, NEOVASC, DRV01DRV02DRV03DRV04DRV05, FAZ, IR03IR05, BRV03BRV05, LRT05, RAV05, VTV05, CRYSTALDEP, RET_ATROPH, RAV, FOVABNORM, CTR_PT_INVOLV, AF_INCREASED_1DD, LARGEAREA_DECR_AF, LOCAL_DECR_AF_VESL_END}
      - make confounding disease column that codes for where 6
      - make OTHER_DISEASE_PRESENT that combines counfouding disease with OTHER_DISEASE
      - set 6s to NA
      - set all > 2 to 2 

```{r}

info %>% 
  filter(nominal&ordinal) %>% 
  select(name,description, matches("\\d")) 

# OTHER_DISEASE
#+ make nominal
#+ name            description           0            1             2                                 3    4
#+ OTHER_DISEASE Other disease visible? Absent Macular hole Lamellar hole Macular edema unrelated to MacTel <NA>
d <- d %>% 
  mutate(OTHER_DISEASE = as.character(OTHER_DISEASE)) %>% 
  mutate(OTHER_DISEASE= case_when(
    OTHER_DISEASE == '0' ~ "0_Absent",
    OTHER_DISEASE == '1' ~ "Macular hole",
    OTHER_DISEASE == '2' ~ "Lamellar hole",
    OTHER_DISEASE == '3' ~ "Macular edema unrelated to MacTel", 
    TRUE~OTHER_DISEASE
  ))


# change these
info[info$name== "OTHER_DISEASE", c("nominal", "ordinal")] <- c(TRUE, FALSE)

# FAZ : foveal avascular zone:> enlargement of the FAZ has been found to negatively correlate with visual acuity (https://pmc.ncbi.nlm.nih.gov/articles/PMC7190136/)
## 0 absent, 1 standard, 2 Greater than standard, 3 smaller than standard, 4 very small
## XXX 1>>0, 2>>3, 3>>2, 4>>1XXX
##  4~1 3~2 1~3 2~4 (new)
d <- d %>% 
  mutate(FAZ = as.numeric(FAZ)) %>% 
  mutate(across(FAZ, ~ case_when(
    . == 4~1,
    . == 3~2,
    . == 1~3,
    . == 2~4,
    TRUE ~ .
  ))) 

# YELLOWRND : Eyes with MacTel have multiple tiny, yellow, crystalline deposits on the inner retinal surface adjacent to the telangiectasis (Figure 3). Nearly half of the patients exhibit crystalline deposits (60% bilaterally), but they do not correlate with the severity of disease (https://pmc.ncbi.nlm.nih.gov/articles/PMC9553319/)
## 0 absent, 1 Questionable,	2 Present: <250 micron,	3 Present: >250 micron,	4 Present: outside the fovea
## make anything over 1 to 2
d <- d %>% 
  mutate(YELLOWRND = as.numeric(YELLOWRND)) %>% 
  mutate(across(YELLOWRND, ~ case_when(
    . > 1 ~ 2, 
    TRUE ~ .
  )))

# make BLRPATTERN dichotomous
## Yes >> 1, No >> 0
d$BLRPATTERN %>% unique()
d <- d %>% 
  mutate(across(BLRPATTERN, ~case_when(
    . == "Yes" ~ "1", 
    . == "No"  ~ "0", 
    TRUE ~ .
  ) %>% as.numeric()
  ))

# grading vars that go > 2
grading_over_two <- d %>% 
  select(-matches("RW")) %>%
  select(any_of( info %>% filter(grepl("CF|OCT|MT|FAF", table)) %>% pull(name)  )) %>%
  map(function(x) {
    if(n_distinct(x)<20 & is.numeric(x)){
      max(x, na.rm = TRUE)>2
    }
  }) %>% 
  Filter(isTRUE, .) %>% 
  names()
#+ 
#+ ICL_TYPE TYP_LEAK_RPE NEOVASC DRV\\d YELLOWRND FAZ IR\\d BRV\\d LRT\\d RAV0\\d VTV\\d CRYSTALDEP RET_ATROPH RAV FOVABNORM CTR_PT_INVOLV AF_INCREASED_1DD LARGEAREA_DECR_AF  LOCAL_DECR_AF_VESL_END
#+ 

sapply(grading_over_two, function(x) d[[x]] %>% table)
#+ ^ max is grade 6 across these
#+ 

info %>% 
  filter(name%in% grading_over_two) %>% 
  distinct(shorthand, .keep_all = T) %>% 
  select(name, description, `0`:`6`) %>% 
  as.data.frame()
# for these, anything > 2 &<6 is redundant

# make confounding disease column, 1 when 6 and 0 otherwise
d <- d %>% 
  mutate(confounding_disease  = rowMaxs(as.matrix(across(all_of(grading_over_two)), na.rm = TRUE), na.rm = TRUE)) %>% 
  mutate(confounding_disease = as.numeric(confounding_disease == 6))

# make var for where other disease not absent and value is 6 (counfounding disease)
d$OTHER_DISEASE_PRESENT <- with(d, 
                                OTHER_DISEASE !="0_Absent" | confounding_disease
                                ) %>% as.numeric()

# set  NA due to confounding disease to missing 
d <- d %>% 
  mutate(across(all_of(grading_over_two), ~ifelse(.==6, NA, .)))


# then, for same vars with values > 2, set those to 2
d <- d %>% 
  mutate(across(all_of(grading_over_two), 
                ~ifelse(.>2, 2, .) 
                ))

```

## subset DATA
```{r subset for case}

# select for diagnosed
yes_d <- subset(d, DIAGNOSIS == "yes"& !is.na(eye_num))

```

#+++++++++++++++++++++++++++++++

# arrange data for plotting
* make data frame of counts for each categories eyes and zones  
e.g. >> df[ , (zone, score, x, y)] ... for each category

```{r make patterns to search for}

# make df of patterns and their labels
patts <- c("^IR.*\\d$","^BRV.*\\d$" , "^DRV.*\\d$", "^LRT.*\\d$", "^RAV.*\\d$", "^VTV.*\\d$", "^RPE_LK.*\\d$", "^ICL.*\\d$", "avg_MT.*\\d$",  "_slope$")

patt_names <-  c("Irregular RPE", "Blunted retinal vessels", "Dilated retinal vessels", "Loss of retinal trans", "Right angle vein", "Visible tele vessel",
                                     "RPE leakage", "Inner cap leakage",  "Macular thickness (Spectralis)",  "Foveal Slope")
pattern_df <- as.data.frame(cbind(pattern = patts , 
                                     label = patt_names))
                            
```

### melt_mean function
```{r function to summarize areas as mean }

# data = yes_d
# pattern = "^IR.*\\d$"

# for a pattern in df d, melt those variables into a new df
melt_mean <- function(data , pattern= character()){
  print(pattern)
  
    df <- data %>% 
      select(patID_eye, DATE_coalesced, matches(pattern))
    
    # make long 
    df_m <- reshape2::melt(df, value.name = "score", variable.name = "zone")
    
    # make char, not factors
    df_m  <- df_m  %>%
      mutate_all( ~as.character(.))
    
    # make num zone
    if (pattern != "_slope$"){
      df_m$zone_num <- substr(df_m$zone, nchar(df_m$zone), nchar(df_m$zone)) %>% as.numeric() 
    }else{
      df_m$zone_num <- substr(df_m$zone, nchar(df_m$zone)-7, nchar(df_m$zone)-6) %>% as.numeric() 
    }
    # subset the data for score and zone without NAs
    sub_df <-df_m %>% 
      filter(!is.na(score)) %>% 
      select(zone_num, score)
    
    # make numeric
    sub_df$score <- sub_df$score %>%
      as.numeric()
    
    # aggregate by zone
    summarized_df <- aggregate(score ~ zone_num, data = sub_df, FUN = mean)
    summarized_df$zone_num <-summarized_df$zone_num %>% 
      factor()
    
    # add NA zones for plotting
    na_df <- data.frame(zone_num = as.character(1:9),
                        score = NA)
    
    
    summarized_df_full <- na_df %>% 
      left_join(summarized_df, by = "zone_num", suffix = c(".na", ".sum")) %>% 
      mutate(score = coalesce(score.sum, score.na)) %>%
      select(zone_num, score)
      
return(summarized_df_full)
}


```


# loop through patterns, apply function and plot for patients with mactel

```{r}
# for each pattern apply function melt_mean
plots <- list();summ_dfs <- list();   for (i in 1:nrow(pattern_df)){
      
       # summarize columns with patt
      melt_mean_df <- melt_mean(yes_d, pattern_df$pattern[[i]])
      
      # save to list of dfs
      summ_dfs <- summ_dfs %>% append(list(list(name = pattern_df$label[[i]],
                                          data = melt_mean_df)))
      # plot counts for each zone to polygons resembling etdrs grid
      positions <- data.frame(
        zone_num = rep(melt_mean_df$zone_num, each = 4), # repeat zone_num for times each  
        
         x = c(-1,1,1,-1,
               -3,3,1,-1,
               1,3,3,1,
               -1,1,3,-3,
               -3,-1,-1,-3,
               -5,5,3,-3,
               3,5,5,3,
               -3,3,5,-5,
               -5,-3,-3,-5),
         
          y = c(1,1,-1,-1,
                3,3,1,1,
                1,3,-3,-1,
                -1,-1,-3,-3,
                3,1,-1,-3,
                5,5,3,3,
                3,5,-5,-3,
                -3,-3,-5,-5,
                5,3,-3,-5)
      )
      
      
      
      values <- data.frame(
        zone_num = melt_mean_df$zone_num,
        value = melt_mean_df$score
      )
      
      # merge values and shape coordinates into one 
      datapoly <- merge(values, positions, by = c("zone_num"))
      
      # make df of etdrs zone labels
      shape_labels <- data.frame(
        label = c("Cen", "Sinn", "Ninn", "Iinn", "Tinn", "Sout", "Nout", "Iout", "Tout"),
        lab_x = c(0,0,2,0,-2,0,4,0,-4),
        lab_y = c(0,2,0,-2,0,4,0,-4,0)
      )
      
      
      plot <- ggplot(datapoly, aes(x = x, y = y)) +
        geom_polygon(aes(fill = value, group = zone_num))+
        theme_void()+
        ggtitle(paste(pattern_df$label[i], "mean value mapped to right eye ETDRS grid"))+
        geom_text(data = shape_labels, aes(x = lab_x, y = lab_y, label = label), size = 4)
        
        
      # plot shapes, same scale for non mt plots
      if(patts[i] %in% c("cirrus_MT.*\\d$", "stratus_MT.*\\d$", "spectralis_MT0\\d$")){
         plot <- plot + 
                  scale_fill_viridis(begin = 0, end = 1)
         
      }else (
       plot <- plot+
         scale_fill_viridis(begin = 0, end = 1)
                  # scale_fill_viridis(begin = 1, end = 0, limits = c(0, 2))
      )
      
       plots[[i]] <-ggplotGrob(plot)
}



# make grid and save 
etdrs_grid <- gridExtra::marrangeGrob(grobs = plots, nrow = 5, ncol = 2)
ggsave(paste0(wd_path,"results/etdrs_grid.png") , etdrs_grid, width = 15, height = 15, dpi = 200)

```


<!-- # heatmap of phenotype x zone -->
<!-- ```{r} -->
<!-- pheno_cases <- data.frame() -->
<!-- for (item in summ_dfs){ -->
<!--  name <- item[1]$name -->
<!--  df <- item[[2]] -->
<!--  df_t <- df[,2] %>% t() -->
<!--  rownames(df_t) <- name -->
<!--  colnames(df_t) <- df$zone_num -->
<!--   pheno_cases <- pheno_cases %>% rbind(., df_t) -->
<!-- } -->

<!-- pheno_m_cases <- pheno_cases %>% as.matrix()  -->

<!-- heatmap(pheno_m_cases, scale = "row", main = "Average values for each zone and phenotype") -->

<!-- #~/Documents/LMRI/PRJ2024001/version015/results/area_phenotype_heatmap.png -->
<!-- ``` -->


# save data
```{r}
new_data_name <- data_name %>% 
  gsub("3", "4", .)
print(paste("Saving...", new_data_name))
saveRDS(d, paste0(wd_path,"processed_data/", new_data_name))
```
CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_04.rds

# save info
NOT EDITED
```{r}
write.xlsx(info, paste0(wd_path,"processed_data/database_grading_key_03.xlsx"))
```

```{r}
sum(is.na(d$LRT05)) # 27198 >  24547 26-3-2024 >> 1302 23-4-2024 >> 297 20-5-2024 >> 961 24-5-2024 >> 348  13-3-2025
```

```{r, look at dims}
check_dims()

# dims : 12600 288
#                      [,1]
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516


### MARCH 13 2025 ###
# dims : 12619 289
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```


<!-- ```{r} -->

<!-- X <- d %>%  -->
<!--   mutate_all(as_numeric_if_possible) %>%  -->
<!--   select(where(is.numeric), matches("DM|SMK|OTHER|CONF|WGT|HGT"), -patientId)  -->

<!-- X <- X %>%  -->
<!--   fastDummies::dummy_cols( names(X)[sapply(X, is.character)],remove_first_dummy = FALSE)%>% -->
<!--   select(where(is.numeric)) -->

<!-- corr <-vapply(X, function(x) { -->
<!--   if (is.numeric(x)) { -->
<!--     cor(X$confounding_disease, x, use = "complete.obs") -->
<!--   } else { -->
<!--     NA_real_ -->
<!--   } -->
<!-- }, numeric(1))  -->
<!-- corr <- corr[!is.na(corr)] -->
<!-- corr <- as.matrix(corr) -->

<!-- colnames(corr) <- "confounding_disease" -->

<!-- Heatmap(t(corr), cluster_rows = F, cluster_columns = TRUE) -->

<!-- x <- X$confounding_disease  # Reference vector -->
<!-- dists <- apply(X, 2, function(y) { -->
<!--   if (is.numeric(y)) sqrt(sum((x - y)^2, na.rm = TRUE)) else NA_real_ -->
<!-- }) -->

<!-- as.matrix(dists) %>%  -->
<!--   {1/sqrt(.+1)} %>%  -->
<!--   t() %>%  -->
<!--   Heatmap(col = plasma(100)) -->

<!-- X %>%  -->
<!--   drop_na() %>%  -->
<!--   nrow() -->

<!-- mdpatt <- md.pattern(X, plot = FALSE) -->

<!-- nrow(X) -->
<!-- mdpatt[nrow(mdpatt), 1:ncol(X) ] %>% sort() -->

<!-- fa <- psych::fac(X, nfactors = 10, rotate = "oblimin") -->

<!-- ``` -->

