---
title: "04_longitudinalEDA"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---
# options
```{r}
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
```


# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(purrr)
library(GGally)
library(tidyverse)
library(ggpubr)
library(vcd)
library(UpSetR)
library(gridExtra)
# library(BiocSingular)
library(ComplexHeatmap)
library(viridis)
library(knitr)
library(htmlwidgets)
library(zoo)
library(corrplot)
library(RColorBrewer)
library(openxlsx)
library(plotly)
library(akima)
# library(scater)
library(RColorBrewer)
library(mgcv)
library(zoo)
library(tools)
library(mice)
library(grDevices)
library(ggrepel)
library(UpSetR)

rm(list = ls())
gc()
```


# Load Data
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}
wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

# raw list of dfs
raw <- readRDS(paste0(wd_path, "processed_data/dfs_list_01.rds"))

# read in data
data_name <- "CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_04.rds"

data_0 <- readRDS(paste0(wd_path, "processed_data/", data_name)) # clean data with everything 
d <- data_0
dim(d) # 15271   217 >> 12619   289 12-3-2025

# grading nomenclature and information
info_path <- paste0(wd_path,"processed_data/database_grading_key_03.xlsx")
info <- readxl::read_xlsx(info_path) %>% 
  as.data.frame()

names_vec <- "rescaled_stratus_MT09"
# ****** FUnction to query name in info ******
var_summary <- function(names_vec = character()){
  patt <- paste(names_vec, collapse = "|")  
  m <- as.matrix(filter(info, grepl(patt, name, ignore.case = TRUE)))
  rownames(m) <- m[,1]
  print(kable(t(m), "simple"))
  for(nm in m[,'name']){
    cat(sprintf('\n%s\n', nm))
    print(table(d[[nm]]))
  }
}

```

# make new directorys
```{r}
path <- paste0(wd_path, "results/longitudinal_analysis/")
if(!dir.exists(path)) dir.create(path, recursive = TRUE)

```

#___________________________
# Clean Data

```{r, look at dims}
check_dims <- function(D = d){
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()
# dims : 12600 288
#                      [,1]
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516

#^ no change

### MARCH 13 2025 ###
# dims : 12619 289
#                      [,1]
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500
```

## make SEXF variable
```{r}

# orig: 1 = Male, 2 -> Female
d$SEXF <-  d$GENDER == 2
cbind(d$GENDER, d$SEXF) %>% unique()

```

## update info
```{r}

new_vars <- setdiff(names(d), info$name) %>%
  grep("DIAG|ISOS|SEX|DATE|samp|disease", ., value = T, invert = TRUE, ignore.case = TRUE) %>% 
  sort()

new_vars_description <- c(rep("Average MT across machines", length(grep("avg_MT", new_vars)) ), 
                       rep("cirrus rescaled to spectralis", length(grep("rescaled_cirrus", new_vars)) ),
                       rep("spectralis rescaled to stratus", length(grep("rescaled_spect", new_vars)) ), 
                       rep("stratus rescaled to spectralis", length(grep("rescaled_strat", new_vars)) ), 
                       "Value is average MT across machines")
cbind(new_vars, new_vars_description)

info_update <- info[1, ] %>% mutate_all(~NA)
info_update <- info_update[rep(1, length(new_vars)), ]
rownames(info_update) <- 1:nrow(info_update)

info_update$name <- new_vars
info_update$description <- new_vars_description

info_update$continuous <- ifelse(grepl("avg",info_update$name), TRUE, FALSE )
info_update$zonal <- ifelse(grepl("0\\d",info_update$name), TRUE, FALSE )

# add new rows to info
info <- info %>% 
  rbind.data.frame(., info_update)

```


## fix phenotypes that heal over time
```{r}

#  what vars have fewer than 20 unique values?
non_cont_vars <- d %>% 
  map( ~{ is.numeric(.) & n_distinct(.)<=20 }) %>% 
  Filter(function(x)x,.) %>% 
  names()

# what vars in non-cont-vars not from these datasets (filter out info and continous vars, or ones that can heal)
non_cont_vars.sub <- setdiff(non_cont_vars, info$name[info$table %in% c("VAA", "DEM/ENR/FAM/other","MH/DIAGNOSIS", "GRADING")]) 

# if TRUE, then var doesnt have non-numeric characters
non_cont_vars.sub <- sapply(non_cont_vars.sub, function(var) {
  # do any of the unique values have non numeric values?
  any(  !grepl("\\D", unique(as.character(d[[var]])) )  )
  # unique(as.character(d[[var]])) # %>% head()
})

# select vars where only numeric
non_cont_vars.sub <- non_cont_vars.sub[non_cont_vars.sub] %>% 
  names()

# remove vars like ISOS_LOC_JF/CEN because they code location; this can change back and forth
non_cont_vars.sub <- non_cont_vars.sub %>% 
  grep("FOVQUAL|ISOS_LOC_", .,invert = TRUE, value =TRUE)
# FAZ,IR01, IR02, IR03, IR04, IR05, IR06, IR07, IR08, IR09, BRV01,BRV02,BRV03,BRV04,BRV05,BRV06,BRV07,BRV08,BRV09,DRV01,DRV02,DRV03, DRV04,DRV05,ICL01,ICL02,ICL04,ICL05,ICL06,ICL08,ICL09,ICL03,LRT01, LRT02,LRT03,LRT04,LRT05,LRT06,LRT07,LRT08,LRT09,RAV01,RAV02,RAV03, RAV04,RAV05,VTV01,VTV02,VTV03,VTV04,VTV05,VTV07,
# VTV09,NEOVASC,RETHAEM, ICL_TYPE, RPE_LK01, RPE_LK02, RPE_LK03, RPE_LK04, RPE_LK05, RPE_LK06, RPE_LK07, RPE_LK08, RPE_LK09, YELLOWRND, CRYSTALDEP, HRDEXUDATE, RET_ATROPH, NOABNORMCF, NOABNORMFL, TYP_LEAK_RPE, RAV,FOVABNORM,CTR_PT_INVOLV,AF_INCREASED_1DD, LARGEAREA_DECR_AF, LOCAL_DECR_AF_VESL_END, NOABNORMFAF,
# CHANGESFAF, FOVDET, LOSS_ONL, ISOS_BREAK, HR_TEMPORAL,INR_EMPTY_SPC,OUTR_EMPTYSPC,SUBRET_NEW_VESL,CTRPOINT_INVOLVD,CTRPOINT_INVOL_2, INTRARET_PGMT_MIGR, NOABNORMOCT,CHAGNESOCT, DIAGNOSIS_binary, confounding_disease,OTHER_DISEASE_PRESENT, 

#### non-contious phenotypes can only get worse, set any that improve to previous ####
# ****** FUNCTION: ignore NAs and zeroes in  cummax calculation
my_cummax <- function(x) {
  if(sum(!is.na(x))>1){
    where_zeros <- x == 0
    where_na <- is.na(x)
    
    # make nas most negative
    x[where_na] <- -Inf
    
    # Compute cumulative max
    cmx <- cummax(x)
    
    # Restore NA values where they originally were
    cmx[where_na] <- NA
    
    # If NA is wrapped by zeros, turn it into 0
    wrapped_na <- where_na & lag(where_zeros, default = FALSE) & lead(where_zeros, default = FALSE)
    cmx[wrapped_na] <- 0
    
    return(cmx)
  }else{
    return(x)
  }
}

# # Test cases
# x1 <- c(0,0,NA,1,0,2,0)   # should be 0,0,NA,1,1,2,2
# x2 <- c(0,0,NA,0,1,0,2,0) # should be 0,0,0,0,1,1,2,2
# 
# cbind(x1, my_cummax(x1))
# cbind(x2,my_cummax(x2))

all(sapply(non_cont_vars.sub, function(x) is.numeric(d[[x]])))

# cummulative max all non-continuous vars:  resort to previous max if current less than max; carry_zero does not carry zeroes forward to replace NA
d <- d %>%
  mutate(across(all_of(non_cont_vars.sub), ~as.numeric(.))) %>% 
  arrange(DATE_coalesced) %>%
  group_by(patID_eye) %>%
  mutate(across(all_of(non_cont_vars.sub), my_cummax)) %>%
  ungroup()

```


<!-- ```{r} -->
<!-- corr <- d %>%  -->
<!--     select(OTHER_DISEASE, where(is.numeric)) %>%  -->
<!--     fastDummies::dummy_cols(names(.)[sapply(., is.character)],remove_first_dummy = FALSE) %>%  -->
<!--     select(where(is.numeric)) %>%  -->
<!--     cor(use = "pairwise.complete")  -->

<!-- corr <- corr[grepl("OTHER_DISEASE|confounding",rownames(corr)), ] -->
<!-- corr <- corr[ , colSums(is.na(corr))==0] -->

<!-- Heatmap(corr, cluster_rows = TRUE, cluster_columns = TRUE) -->

<!-- ``` -->


## remove NA rows
```{r}

d[is.na(d$sample) , ] %>% glimpse()
d <-d[!is.na(d$sample) , ]

```
> no NA samples

## make month
```{r}

d <- d %>% 
  group_by(patID_eye) %>% 
  arrange(DATE_coalesced) %>% 
  mutate(month = interval(dplyr::first(DATE_coalesced), DATE_coalesced) / months(1) 
  ) %>% 
  arrange(patID_eye, month) %>% 
  ungroup()

```

```{r, look at dims}
check_dims()
# dims : 12600 288
#                      [,1]
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516
#+ ^ added month variable


### MARCH 13 2025 ###
# dims : 12619 290
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```

#### parafovea 
###### delta from 000 of average parafovea (zones 2-5 ) and categorize patients based on quartile of that value 
```{r}

# high and low filter limit
filter_lim <- 50

```




####### AVERAGE MT
```{r}

what_parafovea_cols <- grep("avg_MT0[2345]$", names(d), value = TRUE)

# eyes that return after first visit and have data of interest
return_eyes <- d %>%
  select(patID_eye, month, all_of(what_parafovea_cols)) %>%
  na.omit() %>%
  group_by(patID_eye) %>%
  filter(n_distinct(month) > 1) %>%
  pull(patID_eye) %>%
  unique()

## filter data for completeness, visit, and mactel,  and make month number
p1 <- d %>% 
  filter(patID_eye %in% return_eyes) %>%
  select( patID, patID_eye, month, all_of(what_parafovea_cols), rescaled_to_spectralis) %>% 
  na.omit()

## average parafovea and find delta from baseline
p2 <- p1 %>% 
  arrange(patID_eye, month) %>% ##  order by eye and then month
  group_by(patID_eye) %>%  ## group by eye
  mutate(parafoveal_mean = rowMeans(across(all_of(what_parafovea_cols)))) %>% ## find mean of parafoveal zones
  mutate(delta_from_0 = parafoveal_mean - dplyr::first(parafoveal_mean) ) %>%    ## find difference at visit from first visit
  ungroup()

## categorize patient for first visit quartile 
quartiles <- p2 %>% 
    group_by(patID_eye) %>% 
    arrange(month) %>% 
    reframe(first_mean = dplyr::first(parafoveal_mean)) %>%
    pull(first_mean) %>% 
    quantile(., probs = seq(0, 1, .25))

quartile_levels <- c( "fourth", "third", "second", "first" )

p3 <- p2 %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  # mutate(quartile = ifelse(dplyr::first(parafoveal_mean) < quartiles [[2]], "first",
  #                          ifelse(dplyr::first(parafoveal_mean) < quartiles[[3]], "second",
  #                                 ifelse(dplyr::first(parafoveal_mean) < quartiles[[4]], "third", "fourth")))) %>% 
  mutate(quartile= cut(dplyr::first(parafoveal_mean),breaks = quartiles,include.lowest = T, labels = rev(quartile_levels))) %>%
  mutate(quartile = factor(quartile, levels = quartile_levels ) ) %>% 
  ungroup() %>% 
  group_by(quartile) %>% 
  mutate(n = n()) %>% 
  ungroup()

# high and low filter 
p3 <- p3 %>% 
  filter(abs(delta_from_0) < filter_lim)

## plot each quartile facet with line for each patient and overall lm
parafovea_quartile_facet_plot <- p3 %>% 
  # filter(abs(delta_from_0)<50) %>% 
  ggplot(aes(x = month, y = delta_from_0, color = patID_eye))+
  geom_point(show.legend = FALSE, alpha = .3)+
  geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
  geom_smooth(method = "lm",   show.legend = FALSE, aes(group = 1), se = TRUE)+
  geom_hline(yintercept = 0, linetype = 3, color = 'black')+
  facet_wrap(~quartile, scales = "free")+
  theme_bw()+
  ggtitle("Average MT mean parafoveal thickness delta from first visit \n(patients categorized based on first visit quartile)")+
   theme(plot.title = element_text(size = 8))

ggsave(paste0(wd_path, "results/longitudinal_analysis/average_MT_parafovea_quartile_facet_plot.png"), parafovea_quartile_facet_plot, width = 10, height = 8, dpi = 200)
 
# parafovea_plot <- ggplot(p3  , aes(x = month, y = delta_from_0, color = patID_eye))+
#   geom_point(show.legend = FALSE, alpha = .3)+
#   geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
#   geom_smooth(method = "lm",   show.legend = FALSE, aes(group = 1), se = TRUE)+
#   geom_hline(yintercept = 0, linetype = 3, color = 'black')+
#   theme_bw()+
#   ggtitle("Mean parafoveal thickness delta from first visit")+
#    theme(plot.title = element_text(size = 8))
# 
# ggsave(paste0(wd_path, "results/longitudinal_analysis/parafovea_plot.png"), parafovea_plot, width = 10, height = 8, dpi = 200)

# # PLOT QUINTILE OF SAME ----
# parafovea <- d %>% 
#   select(month, patID_eye, patID, month, avg_MT02:avg_MT05) %>% 
#   mutate(parafovea = rowMeans(across(matches("avg"))))
# 
# parafovea_quintile_d <- parafovea %>% 
#   arrange(month, patID_eye) %>% 
#   group_by(patID_eye) %>% 
#   filter(row_number() ==1) %>% 
#   ungroup() %>% 
#   mutate(
#     parafovea_quintile = 
#       cut(parafovea, quantile(parafovea, seq(0,1,.2), na.rm = TRUE), labels = 1:5)
#   ) %>% 
#   select(patID_eye, parafovea_quintile) %>% 
#   distinct()
# 
# parafovea <- parafovea %>% 
#   left_join(parafovea_quintile_d)
# 
# ggplot(parafovea %>% drop_na(), aes(month, parafovea, color = patID_eye))+
#   geom_smooth(method= "lm", aes(group = 1))+
#   # geom_point(show.legend = FALSE)+
#   # geom_line(show.legend = FALSE)+
#   facet_wrap(~factor(parafovea_quintile, levels = 5:1),ncol = 5, scales = "free_x")+
#   ggtitle("Average parafoveal macular thickness over time quantiled by BL thickness")
#   ggsave()
# >>>>>>> look at scratchpad.R for better version, uses imputed MT

```

####### SPECTRALIS
```{r}

what_parafovea_cols <- grep("spectralis_MT0[2345]$", names(d), value = TRUE)

# eyes that return after first visit and have data of interest
return_eyes <- d %>%
  select(patID_eye, month, all_of(what_parafovea_cols)) %>%
  na.omit() %>%
  group_by(patID_eye) %>%
  filter(n() > 1) %>%
  pull(patID_eye) %>%
  unique()

## filter data for completeness, visit, and mactel,  and make month number
p1 <- d %>% 
  filter(patID_eye %in% return_eyes) %>%
  select( patID, patID_eye, month, all_of(what_parafovea_cols), rescaled_to_spectralis) %>% 
  na.omit()

## average parafovea and find delta from baseline
p2 <- p1 %>% 
  arrange(patID_eye, month) %>% ##  order by eye and then month
  group_by(patID_eye) %>%  ## group by eye
  mutate(parafoveal_mean = rowMeans(across(all_of(what_parafovea_cols)))) %>% ## find mean of parafoveal zones
  mutate(delta_from_0 = parafoveal_mean - dplyr::first(parafoveal_mean) ) %>%    ## find difference at visit from first visit
  ungroup()

## categorize patient for first visit quartile 
quartiles <- p2 %>% 
    group_by(patID_eye) %>% 
    arrange(month) %>% 
    reframe(first_mean = dplyr::first(parafoveal_mean)) %>%
    pull(first_mean) %>% 
    quantile(., probs = seq(0, 1, .25))

quartile_levels <- c( "fourth", "third", "second", "first" )

p3 <- p2 %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  # mutate(quartile = ifelse(dplyr::first(parafoveal_mean) < quartiles [[2]], "first",
  #                          ifelse(dplyr::first(parafoveal_mean) < quartiles[[3]], "second",
  #                                 ifelse(dplyr::first(parafoveal_mean) < quartiles[[4]], "third", "fourth")))) %>% 
  mutate(quartile= cut(dplyr::first(parafoveal_mean),breaks = quartiles,include.lowest = T, labels = rev(quartile_levels))) %>%
  mutate(quartile = factor(quartile, levels = quartile_levels ) ) %>% 
  ungroup() %>% 
  group_by(quartile) %>% 
  mutate(n = n()) %>% 
  ungroup()

# high and low filter 
p3 <- p3 %>% 
  filter(abs(delta_from_0) < filter_lim)

## plot each quartile facet with line for each patient and overall lm
parafovea_quartile_facet_plot <- p3 %>% 
  # filter(abs(delta_from_0)<50) %>% 
  ggplot(aes(x = month, y = delta_from_0, color = patID_eye))+
  geom_point(show.legend = FALSE, alpha = .3)+
  geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
  geom_smooth(method = "lm",   show.legend = FALSE, aes(group = 1), se = TRUE)+
  geom_hline(yintercept = 0, linetype = 3, color = 'black')+
  facet_wrap(~quartile, scales = "free")+
  theme_bw()+
  ggtitle("Spectralis mean parafoveal thickness delta from first visit \n(patients categorized based on first visit quartile)")+
   theme(plot.title = element_text(size = 8))

ggsave(paste0(wd_path, "results/longitudinal_analysis/spectralis_parafovea_quartile_facet_plot.png"), parafovea_quartile_facet_plot, width = 10, height = 8, dpi = 200)
 
# parafovea_plot <- ggplot(p3  , aes(x = month, y = delta_from_0, color = patID_eye))+
#   geom_point(show.legend = FALSE, alpha = .3)+
#   geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
#   geom_smooth(method = "lm",   show.legend = FALSE, aes(group = 1), se = TRUE)+
#   geom_hline(yintercept = 0, linetype = 3, color = 'black')+
#   theme_bw()+
#   ggtitle("Mean parafoveal thickness delta from first visit")+
#    theme(plot.title = element_text(size = 8))
# 
# ggsave(paste0(wd_path, "results/longitudinal_analysis/parafovea_plot.png"), parafovea_plot, width = 10, height = 8, dpi = 200)

```


####### STRATUS
```{r}

what_parafovea_cols <- grep("stratus_MT0[2345]$", names(d), value = TRUE)

# eyes that return after first visit and have data of interest
return_eyes <- d %>%
  select(patID_eye, month, all_of(what_parafovea_cols)) %>%
  na.omit() %>%
  group_by(patID_eye) %>%
  filter(n() > 1) %>%
  pull(patID_eye) %>%
  unique()

## filter data for completeness, visit, and mactel,  and make month number
p1 <- d %>% 
  filter(patID_eye %in% return_eyes) %>%
  select( patID, patID_eye, month, all_of(what_parafovea_cols), rescaled_to_spectralis) %>% 
  na.omit()

## average parafovea and find delta from baseline
p2 <- p1 %>% 
  arrange(patID_eye, month) %>% ##  order by eye and then month
  group_by(patID_eye) %>%  ## group by eye
  mutate(parafoveal_mean = rowMeans(across(all_of(what_parafovea_cols)))) %>% ## find mean of parafoveal zones
  mutate(delta_from_0 = parafoveal_mean - dplyr::first(parafoveal_mean) ) %>%    ## find difference at visit from first visit
  ungroup()

## categorize patient for first visit quartile 
quartiles <- p2 %>% 
    group_by(patID_eye) %>% 
    arrange(month) %>% 
    reframe(first_mean = dplyr::first(parafoveal_mean)) %>%
    pull(first_mean) %>% 
    quantile(., probs = seq(0, 1, .25))

quartile_levels <- c( "fourth", "third", "second", "first" )

p3 <- p2 %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  # mutate(quartile = ifelse(dplyr::first(parafoveal_mean) < quartiles [[2]], "first",
  #                          ifelse(dplyr::first(parafoveal_mean) < quartiles[[3]], "second",
  #                                 ifelse(dplyr::first(parafoveal_mean) < quartiles[[4]], "third", "fourth")))) %>% 
  mutate(quartile= cut(dplyr::first(parafoveal_mean),breaks = quartiles,include.lowest = T, labels = rev(quartile_levels))) %>%
  mutate(quartile = factor(quartile, levels = quartile_levels ) ) %>% 
  ungroup() %>% 
  group_by(quartile) %>% 
  mutate(n = n()) %>% 
  ungroup()

# high and low filter 
p3 <- p3 %>% 
  filter(abs(delta_from_0) < filter_lim)

## plot each quartile facet with line for each patient and overall lm
parafovea_quartile_facet_plot <- p3 %>% 
  # filter(abs(delta_from_0)<50) %>% 
  ggplot(aes(x = month, y = delta_from_0, color = patID_eye))+
  geom_point(show.legend = FALSE, alpha = .3)+
  geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
  geom_smooth(method = "lm",   show.legend = FALSE, aes(group = 1), se = TRUE)+
  geom_hline(yintercept = 0, linetype = 3, color = 'black')+
  facet_wrap(~quartile, scales = "free")+
  theme_bw()+
  ggtitle("Stratus Mean parafoveal thickness delta from first visit \n(patients categorized based on first visit quartile)")+
   theme(plot.title = element_text(size = 8))

ggsave(paste0(wd_path, "results/longitudinal_analysis/stratus_parafovea_quartile_facet_plot.png"), parafovea_quartile_facet_plot, width = 10, height = 8, dpi = 200)
 
# parafovea_plot <- ggplot(p3  , aes(x = month, y = delta_from_0, color = patID_eye))+
#   geom_point(show.legend = FALSE, alpha = .3)+
#   geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
#   geom_smooth(method = "lm",   show.legend = FALSE, aes(group = 1), se = TRUE)+
#   geom_hline(yintercept = 0, linetype = 3, color = 'black')+
#   theme_bw()+
#   ggtitle("Mean parafoveal thickness delta from first visit")+
#    theme(plot.title = element_text(size = 8))
# 
# ggsave(paste0(wd_path, "results/longitudinal_analysis/parafovea_plot.png"), parafovea_plot, width = 10, height = 8, dpi = 200)

```


####### CIRRUS
```{r}

what_parafovea_cols <- grep("cirrus_MT0[2345]$", names(d), value = TRUE)

# eyes that return after first visit and have data of interest
return_eyes <- d %>%
  select(patID_eye, month, all_of(what_parafovea_cols)) %>%
  na.omit() %>%
  group_by(patID_eye) %>%
  filter(n() > 1) %>%
  pull(patID_eye) %>%
  unique()

## filter data for completeness, visit, and mactel,  and make month number
p1 <- d %>% 
  filter(patID_eye %in% return_eyes) %>%
  select( patID, patID_eye, month, all_of(what_parafovea_cols), rescaled_to_spectralis) %>% 
  na.omit()

## average parafovea and find delta from baseline
p2 <- p1 %>% 
  arrange(patID_eye, month) %>% ##  order by eye and then month
  group_by(patID_eye) %>%  ## group by eye
  mutate(parafoveal_mean = rowMeans(across(all_of(what_parafovea_cols)))) %>% ## find mean of parafoveal zones
  mutate(delta_from_0 = parafoveal_mean - dplyr::first(parafoveal_mean) ) %>%    ## find difference at visit from first visit
  ungroup()

## categorize patient for first visit quartile 
quartiles <- p2 %>% 
    group_by(patID_eye) %>% 
    arrange(month) %>% 
    reframe(first_mean = dplyr::first(parafoveal_mean)) %>%
    pull(first_mean) %>% 
    quantile(., probs = seq(0, 1, .25))

quartile_levels <- c( "fourth", "third", "second", "first" )

p3 <- p2 %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  #mutate(quartile = ifelse(dplyr::first(parafoveal_mean) < quartiles [[2]], "first",
  #                         ifelse(dplyr::first(parafoveal_mean) < quartiles[[3]], "second",
  #                                ifelse(dplyr::first(parafoveal_mean) < quartiles[[4]], "third", "fourth")))) %>% 
  mutate(quartile= cut(dplyr::first(parafoveal_mean),breaks = quartiles,include.lowest = T, labels = rev(quartile_levels))) %>%
  mutate(quartile = factor(quartile, levels = quartile_levels ) ) %>% 
  ungroup() %>% 
  group_by(quartile) %>% 
  mutate(n = n()) %>% 
  ungroup()


# high and low filter 
p3 <- p3 %>% 
  filter(abs(delta_from_0) < filter_lim)

## plot each quartile facet with line for each patient and overall lm
parafovea_quartile_facet_plot <- p3 %>% 
  # filter(abs(delta_from_0)<50) %>% 
  ggplot(aes(x = month, y = delta_from_0, color = patID_eye))+
  geom_point(show.legend = FALSE, alpha = .3)+
  geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
  geom_smooth(method = "lm",   show.legend = FALSE, aes(group = 1), se = TRUE)+
  geom_hline(yintercept = 0, linetype = 3, color = 'black')+
  facet_wrap(~quartile, scales = "free")+
  theme_bw()+
  ggtitle("Cirrus mean parafoveal thickness delta from first visit \n(patients categorized based on first visit quartile)")+
   theme(plot.title = element_text(size = 8))

ggsave(paste0(wd_path, "results/longitudinal_analysis/cirrus_parafovea_quartile_facet_plot.png"), parafovea_quartile_facet_plot, width = 10, height = 8, dpi = 200)

```

#-
### all zones thickness delta000 quartiled 
```{r}

what_parafovea_cols <- grep("avg_MT0[2345]$", names(d), value = TRUE)

## filter data for completeness, visit, and mactel,  and make month number
a <- d %>% 
  filter( patID_eye %in% return_eyes) %>%
  select(all_of(what_parafovea_cols), patID_eye, patID, eye_num, visitID) %>% ## grab zone 5
  filter(!visitID %in% unique(grep("[a-z]", d$visitID, ignore.case = TRUE, value = TRUE)) ) %>%  ## filter for non-character visits
  mutate(month = visitID %>% as.character() %>% as.numeric()) %>%  ## make visits numeric month
  na.omit()

## make average of both eyes at each visit and find difference at each visit from first visit
b <- a %>% 
  arrange(patID_eye, patID, month) %>% ##  order by patient and then month
  # group_by(patID_eye, month) %>%  ## group by patient and then month 
  # summarize(across(starts_with("spectralis"), 
  #               ~ mean(.), 
  #               .names = "zone_{str_sub(.col, 12, 12)}")
  #        ) %>% # average eyes
  # ungroup() %>% 
  group_by(patID_eye) %>% 
  mutate(across(starts_with("avg"), 
                ~ . - dplyr::first(.), 
                .names = "delta000_{.col}")
         ) %>% # calculage diff from 000
  filter(n() >1) %>% # filter for patients who return
  ungroup()

## categorize patient for first visit quartile and quintile
MT2_quartiles <- quantile(b$avg_MT02[b$month == 0])
MT3_quartiles <- quantile(b$avg_MT03[b$month == 0])
MT4_quartiles <- quantile(b$avg_MT04[b$month == 0])
MT5_quartiles <- quantile(b$avg_MT05[b$month == 0])

c <- b %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  mutate( MT2_quartile = ifelse(dplyr::first(avg_MT02) < MT2_quartiles [[2]], "first",
                           ifelse(dplyr::first(avg_MT02) < MT2_quartiles[[3]], "second",
                                  ifelse(dplyr::first(avg_MT02) < MT2_quartiles[[4]], "third", "fourth"))),
          
          MT3_quartile = ifelse(dplyr::first(avg_MT03) < MT3_quartiles [[2]], "first",
                           ifelse(dplyr::first(avg_MT03) < MT3_quartiles[[3]], "second",
                                  ifelse(dplyr::first(avg_MT03) < MT3_quartiles[[4]], "third", "fourth"))),
          
          MT4_quartile = ifelse(dplyr::first(avg_MT04) < MT4_quartiles [[2]], "first",
                           ifelse(dplyr::first(avg_MT04) < MT4_quartiles[[3]], "second",
                                  ifelse(dplyr::first(avg_MT04) < MT4_quartiles[[4]], "third", "fourth"))),
          
          MT5_quartile = ifelse(dplyr::first(avg_MT05) < MT5_quartiles [[2]], "first",
                           ifelse(dplyr::first(avg_MT05) < MT5_quartiles[[3]], "second",
                                  ifelse(dplyr::first(avg_MT05) < MT5_quartiles[[4]], "third", "fourth")))


        ) %>% 
  mutate(across(ends_with("quartile"), 
                ~factor(., levels = c( "fourth" , "third", "second", "first") ) )
         ) %>% 
  ungroup()

# tidy df of deltas
c_long <- c %>% 
  select(delta000_avg_MT02:delta000_avg_MT05, patID, patID_eye, eye_num, visitID, month, all_of(ends_with("quartile"))) %>% 
  pivot_longer(cols = starts_with("delta"), names_to = "variable", values_to = "value")

# tidy df of raw values
c_long_raw <-  c %>% 
  select(avg_MT02:avg_MT05, patID, patID_eye, eye_num, visitID, month, all_of(ends_with("quartile"))) %>% 
  pivot_longer(cols = starts_with("avg"), names_to = "variable", values_to = "value")

quartiles_ordered <- as.character(unique(c_long$MT5_quartile)[order(unique(c_long$MT5_quartile) , decreasing = TRUE)])

# grobs <- lapply(quartiles_ordered, function(a_quartile) {
#   
#   plot <-  ggplot(c_long %>% subset(quartile == a_quartile )  , aes(x = month, y = value, color = patID_eye))+
#   # geom_point(show.legend = FALSE, alpha = .3)+
#   # geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
#   geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k = 5), show.legend = FALSE, aes(group = 1), se = TRUE)+
#   facet_wrap(~variable, scales = "fixed" )+
#   theme_bw()+
#   ggtitle(paste("Macular zone thickness deltas from first visit (natural splines).\n", toTitleCase(a_quartile), "quartile"))+
#    theme(plot.title = element_text(size = 8))
#   
#  return(ggplotGrob(plot))
# })
# 
# # rearrange plots
# new_grobs <- grobs[c(1, 3, 2, 4)]
# 
# plot_grid <- gridExtra::marrangeGrob(grobs = new_grobs, nrow =2, ncol = 2 )
# 
# plot_grid
# 
# 
# ggsave(paste0(wd_path, "results/longitudinal_analysis/quartileFacet_zoneFacet.png"), plot_grid, width = 10, height = 10, dpi = 200)

#################
### of deltas ###
#################

# list of zones to loop over
zones_quartile <- data.frame(variable = c_long$variable %>% unique(), 
                             quartile_name = grep("quartile", value = TRUE, names(c_long)))
plots <- lapply(1:nrow(zones_quartile), function(i){
            
            zone_name <- zones_quartile[i, 1]
            quartile_name <- zones_quartile[i, 2]
            
            filtered_data <- c_long %>%
                    filter(variable == zone_name) %>%
                    mutate(quartile_group = get(quartile_name))
            
            plot <- ggplot(filtered_data, aes(x = month, y = value, color = patID_eye))+
                    geom_point(show.legend = FALSE, alpha = .3)+
                    geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
                    # geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k = 5), show.legend = FALSE, aes(group = 1), se = TRUE)+
                    geom_smooth(method = "lm", show.legend = FALSE, aes(group = 1), se = TRUE)+
                    facet_wrap(~quartile_group, scales = "free" , nrow = 1)+
                    theme_bw()+
                    theme(plot.title = element_text(size = 8))+
                    ggtitle(zone_name)
                    
            if (i < 4 ){
              
              plot <- plot + theme(plot.title = element_text(size = 8), 
                          axis.title.x  = element_blank())
            }
      
            
            if(i == 1) { 
              plot <- plot+ ggtitle(paste("Macular zone thickness deltas from first visit (lm).\n", zone_name))
              
            }
            
          return(ggplotGrob(plot))
      
    })

delta000_grid <- gridExtra::marrangeGrob(grobs = plots,nrow = 4, ncol = 1)
  
ggsave(paste0(wd_path, "results/longitudinal_analysis/zoneQuartileFacet_zoneFacet_parafovea.png"), delta000_grid, width = 10, height = 10, dpi = 200)

#####################
### of raw values ###
#####################
# list of zones to loop over
zones_quartile <- data.frame(variable = c_long_raw$variable %>% unique(), 
                             quartile_name = grep("quartile", value = TRUE, names(c_long_raw)))
i = 1
plots <- lapply(1:nrow(zones_quartile), function(i){
            
            zone_name <- zones_quartile[i, 1]
            quartile_name <- zones_quartile[i, 2]
            
            filtered_data <- c_long_raw %>%
                    filter(variable == zone_name) %>%
                    mutate(quartile_group = get(quartile_name))
            
            plot <- ggplot(filtered_data, aes(x = month, y = value, color = patID_eye))+
                    geom_point(show.legend = FALSE, alpha = .3)+
                    geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
                    # geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k = 5), show.legend = FALSE, aes(group = 1), se = TRUE)+
                    geom_smooth(method = "lm", show.legend = FALSE, aes(group = 1), se = TRUE)+
                    facet_wrap(~quartile_group, scales = "free" , nrow = 1)+
                    theme_bw()+
                    theme(plot.title = element_text(size = 8))+
                    ggtitle(zone_name)
                    
            if (i < 4 ){
              
              plot <- plot + theme(plot.title = element_text(size = 8), 
                          axis.title.x  = element_blank())
            }
      
            
            if(i == 1) { 
              plot <- plot+ ggtitle(paste("Macular zone thickness from first visit (lm).\n", zone_name))
              
            }
            
          return(ggplotGrob(plot))
      
    })

raw_parafovea_grid <- gridExtra::marrangeGrob(grobs = plots,nrow = 4, ncol = 1) 

ggsave(paste0(wd_path, "results/longitudinal_analysis/zoneQuartileFacet_zoneFacet_parafovea_raw_values.png"), raw_parafovea_grid, width = 10, height = 10, dpi = 200)


```


---

<!-- ### By Gender -->
<!-- ```{r} -->

<!-- ## filter data for completeness, visit, and mactel,  and make month number -->
<!-- a <- d %>%  -->
<!--   subset(., patID_eye %in% return_eyes & DIAGNOSIS == "yes") %>% ## filter for mactel and returning patients and SEXF -->
<!--   select(all_of(what_parafovea_cols), patID_eye, patID_eye, eye_num, visitID, GENDER) %>% ## grab zone 5 -->
<!--   filter(!visitID %in% unique(grep("[a-z]", d$visitID, ignore.case = TRUE, value = TRUE)) ) %>%  ## filter for non-character visits -->
<!--   mutate(month = visitID %>% as.character() %>% as.numeric()) %>%  ## make visits numeric month -->
<!--   filter( complete.cases(.))  ## complete rows -->


<!-- # rename to easier names -->
<!-- names(a)[grep("MT0", names(a))] <-  names(a)[grep("MT0", names(a))] %>%  -->
<!--   gsub(".*_MT", "MT", .) -->

<!-- ## make average of both eyes at each visit and find difference at each visit from first visit -->
<!-- b <- a %>%  -->
<!--   arrange(patID_eye, month) %>% ##  order by patient and then month -->
<!--   group_by(patID_eye) %>%  -->
<!--   mutate(across(grep("[0-9]$", names(a), value = TRUE),  -->
<!--                 ~ . - dplyr::first(.),  -->
<!--                 .names = "delta000_{.col}") -->
<!--          ) %>% # calculage diff from 000 -->
<!-- ungroup() -->
<!-- b %>% head() -->

<!-- ## categorize patient for first visit quartile  -->
<!-- # male_quartiles <- quantile(b$MT05[b$month == 0 & b$GENDER == "Male"]) -->
<!-- # female_quartiles <- quantile(b$MT05[b$month == 0 & b$GENDER == "Female"]) -->

<!-- ## categorize patient for first visit quartile  -->
<!-- female_MT2_quartiles <- quantile(b$MT02[b$month == 0 & b$GENDER == "Female"]) -->
<!-- female_MT3_quartiles <- quantile(b$MT03[b$month == 0 & b$GENDER == "Female"]) -->
<!-- female_MT4_quartiles <- quantile(b$MT04[b$month == 0 & b$GENDER == "Female"]) -->
<!-- female_MT5_quartiles <- quantile(b$MT05[b$month == 0 & b$GENDER == "Female"]) -->
<!-- male_MT2_quartiles <-   quantile(b$MT02[b$month == 0 & b$GENDER == "Male"]) -->
<!-- male_MT3_quartiles <-   quantile(b$MT03[b$month == 0 & b$GENDER == "Male"]) -->
<!-- male_MT4_quartiles <-   quantile(b$MT04[b$month == 0 & b$GENDER == "Male"]) -->
<!-- male_MT5_quartiles <-   quantile(b$MT05[b$month == 0 & b$GENDER == "Male"]) -->
<!-- # quintiles -->
<!-- female_MT2_quintiles <- quantile(b$MT02[b$month == 0 & b$GENDER == "Female"], probs = seq(0, 1, .2)) -->
<!-- female_MT3_quintiles <- quantile(b$MT03[b$month == 0 & b$GENDER == "Female"], probs = seq(0, 1, .2)) -->
<!-- female_MT4_quintiles <- quantile(b$MT04[b$month == 0 & b$GENDER == "Female"], probs = seq(0, 1, .2)) -->
<!-- female_MT5_quintiles <- quantile(b$MT05[b$month == 0 & b$GENDER == "Female"], probs = seq(0, 1, .2)) -->
<!-- male_MT2_quintiles <-   quantile(b$MT02[b$month == 0 & b$GENDER == "Male"], probs = seq(0, 1, .2)) -->
<!-- male_MT3_quintiles <-   quantile(b$MT03[b$month == 0 & b$GENDER == "Male"], probs = seq(0, 1, .2)) -->
<!-- male_MT4_quintiles <-   quantile(b$MT04[b$month == 0 & b$GENDER == "Male"], probs = seq(0, 1, .2)) -->
<!-- male_MT5_quintiles <-   quantile(b$MT05[b$month == 0 & b$GENDER == "Male"], probs = seq(0, 1, .2)) -->

<!-- # if male, categorize based on male zoneN visit zero quantiles, if female ... -->
<!-- c <- b %>%  -->
<!--   group_by(patID_eye) %>%  -->
<!--   arrange(month) %>%  -->
<!--   mutate(MT2_quartile = case_when( -->
<!--            GENDER == "Male" ~ ifelse(dplyr::first(MT02) <     male_MT2_quartiles [[2]], 1, -->
<!--                                 ifelse(dplyr::first(MT02) <   male_MT2_quartiles[[3]], 2, -->
<!--                                  ifelse(dplyr::first(MT02) <  male_MT2_quartiles[[4]], 3, 4)) -->
<!--                 ),  -->
<!--             GENDER == "Female" ~ ifelse(dplyr::first(MT02) <  female_MT2_quartiles [[2]],1, -->
<!--                              ifelse(dplyr::first(MT02) <   female_MT2_quartiles[[3]],2, -->
<!--                                ifelse(dplyr::first(MT02) < female_MT2_quartiles[[4]], 3, 4)) -->
<!--                 ),  -->
<!--          TRUE ~ NA_integer_ -->
<!--       ), -->

<!--        MT3_quartile = case_when( -->
<!--            GENDER == "Male" ~ ifelse(dplyr::first(MT03) <   male_MT3_quartiles [[2]], 1, -->
<!--                                ifelse(dplyr::first(MT03) <  male_MT3_quartiles[[3]], 2, -->
<!--                                 ifelse(dplyr::first(MT03) < male_MT3_quartiles[[4]], 3, 4)) -->
<!--                   ),  -->
<!--            GENDER == "Female" ~ ifelse(dplyr::first(MT03) <    female_MT3_quartiles [[2]],1, -->
<!--                                   ifelse(dplyr::first(MT03) <  female_MT3_quartiles[[3]],2, -->
<!--                                   ifelse(dplyr::first(MT03) <  female_MT3_quartiles[[4]], 3, 4)) -->
<!--                   ),  -->
<!--            TRUE ~ NA_integer_ -->
<!--       ), -->
<!--         MT4_quartile = case_when( -->
<!--            GENDER == "Male" ~ ifelse(dplyr::first(MT04) < male_MT4_quartiles [[2]], 1, -->
<!--                                ifelse(dplyr::first(MT04) < male_MT4_quartiles[[3]], 2, -->
<!--                                       ifelse(dplyr::first(MT04) < male_MT4_quartiles[[4]], 3, 4)) -->
<!--                   ),  -->
<!--            GENDER == "Female" ~ ifelse(dplyr::first(MT04) < female_MT4_quartiles [[2]],1, -->
<!--                                ifelse(dplyr::first(MT04) < female_MT4_quartiles[[3]],2, -->
<!--                                       ifelse(dplyr::first(MT04) < female_MT4_quartiles[[4]], 3, 4)) -->
<!--                   ),  -->
<!--            TRUE ~ NA_integer_ -->
<!--       ), -->
<!--         MT5_quartile = case_when( -->
<!--            GENDER == "Male" ~ ifelse(dplyr::first(MT05) < male_MT5_quartiles [[2]], 1, -->
<!--                                ifelse(dplyr::first(MT05) < male_MT5_quartiles[[3]], 2, -->
<!--                                       ifelse(dplyr::first(MT05) < male_MT5_quartiles[[4]], 3, 4)) -->
<!--                   ),  -->
<!--            GENDER == "Female" ~ ifelse(dplyr::first(MT05) < female_MT5_quartiles [[2]],1, -->
<!--                                ifelse(dplyr::first(MT05) < female_MT5_quartiles[[3]],2, -->
<!--                                       ifelse(dplyr::first(MT05) < female_MT5_quartiles[[4]], 3, 4)) -->
<!--                   ),  -->
<!--            TRUE ~ NA_integer_ -->
<!--       )) -->
<!-- ##test## -->
<!-- # x <- seq(100, 420, 5) -->
<!-- #  -->
<!-- # ifelse( x < male_MT5_quintiles [[2]], 1, -->
<!-- #                            ifelse(x < male_MT5_quintiles[[3]], 2, -->
<!-- #                                   ifelse(x < male_MT5_quintiles[[4]], 3, -->
<!-- #                                          ifelse( x < male_MT5_quintiles[[5]],4, 5)))) -->

<!-- c <- c %>%  -->
<!--   group_by(patID_eye) %>%  -->
<!--   arrange(month) %>%  -->
<!--   mutate( MT2_quintile = case_when( -->
<!--     GENDER == "Male" ~ ifelse(dplyr::first(MT02) < male_MT2_quintiles [[2]], 1, -->
<!--                            ifelse(dplyr::first(MT02) < male_MT2_quintiles[[3]], 2, -->
<!--                                   ifelse(dplyr::first(MT02) < male_MT2_quintiles[[4]], 3, -->
<!--                                          ifelse(dplyr::first(MT02) < male_MT2_quintiles[[5]],4, 5)))), -->
<!--     GENDER == "Female"  ~ ifelse(dplyr::first(MT02) < female_MT2_quintiles [[2]], 1, -->
<!--                            ifelse(dplyr::first(MT02) < female_MT2_quintiles[[3]], 2, -->
<!--                                   ifelse(dplyr::first(MT02) < female_MT2_quintiles[[4]], 3, -->
<!--                                          ifelse(dplyr::first(MT02) < female_MT2_quintiles[[3]],4, 5)))) -->
<!--     ), -->
<!--     MT3_quintile = case_when( -->
<!--     GENDER == "Male" ~ ifelse(dplyr::first(MT03) < male_MT3_quintiles [[2]], 1, -->
<!--                            ifelse(dplyr::first(MT03) < male_MT3_quintiles[[3]], 2, -->
<!--                                   ifelse(dplyr::first(MT03) < male_MT3_quintiles[[4]], 3, -->
<!--                                          ifelse(dplyr::first(MT03) < male_MT3_quintiles[[5]],4, 5)))), -->
<!--     GENDER == "Female"  ~ ifelse(dplyr::first(MT03) < female_MT3_quintiles [[2]], 1, -->
<!--                            ifelse(dplyr::first(MT03) < female_MT3_quintiles[[3]], 2, -->
<!--                                   ifelse(dplyr::first(MT03) < female_MT3_quintiles[[4]], 3, -->
<!--                                          ifelse(dplyr::first(MT03) < female_MT3_quintiles[[5]],4, 5)))) -->
<!--     ), -->
<!--     MT4_quintile = case_when( -->
<!--     GENDER == "Male" ~ ifelse(dplyr::first(MT04) < male_MT4_quintiles [[2]], 1, -->
<!--                            ifelse(dplyr::first(MT04) < male_MT4_quintiles[[3]], 2, -->
<!--                                   ifelse(dplyr::first(MT04) < male_MT4_quintiles[[4]], 3, -->
<!--                                          ifelse(dplyr::first(MT04) < male_MT4_quintiles[[5]],4, 5)))), -->
<!--     GENDER == "Female"  ~ ifelse(dplyr::first(MT04) < female_MT4_quintiles [[2]], 1, -->
<!--                            ifelse(dplyr::first(MT04) < female_MT4_quintiles[[3]], 2, -->
<!--                                   ifelse(dplyr::first(MT04) < female_MT4_quintiles[[4]], 3, -->
<!--                                          ifelse(dplyr::first(MT04) < female_MT4_quintiles[[5]],4, 5)))) -->
<!--     ), -->
<!--     MT5_quintile = case_when( -->
<!--     GENDER == "Male" ~ ifelse(dplyr::first(MT05) < male_MT5_quintiles [[2]], 1, -->
<!--                            ifelse(dplyr::first(MT05) < male_MT5_quintiles[[3]], 2, -->
<!--                                   ifelse(dplyr::first(MT05) < male_MT5_quintiles[[4]], 3, -->
<!--                                          ifelse(dplyr::first(MT05) < male_MT5_quintiles[[5]],4, 5)))), -->
<!--     GENDER == "Female"  ~ ifelse(dplyr::first(MT05) < female_MT5_quintiles [[2]], 1, -->
<!--                            ifelse(dplyr::first(MT05) < female_MT5_quintiles[[3]], 2, -->
<!--                                   ifelse(dplyr::first(MT05) < female_MT5_quintiles[[4]], 3, -->
<!--                                          ifelse(dplyr::first(MT05) < female_MT3_quintiles[[5]],4, 5)))) -->
<!--     )) -->


<!-- c <- c %>%  -->
<!--   mutate(across(ends_with("quartile"),  -->
<!--                 ~factor(., levels = c( 4,3,2,1) ) ) -->
<!--          ) %>% -->
<!--   mutate(across(ends_with("quintile"),  -->
<!--                 ~factor(., levels = c( 5, 4,3,2,1) ) ) -->
<!--          ) %>%  -->
<!--   ungroup() %>%  -->
<!--   select(!all_of(grep("^MT0\\d$", names(c), value = TRUE))) -->



<!-- # look at frequencies -->
<!-- # c[c(grep("quartile", names(c), value = TRUE), "GENDER")] %>% table()  -->

<!-- c_long <- c %>%  -->
<!--   select(delta000_MT02:delta000_MT05, patID_eye, patID_eye, eye_num, visitID, month, ends_with("quartile"), ends_with("quintile"), GENDER) %>%  -->
<!--   pivot_longer(cols = starts_with("delta"), names_to = "variable", values_to = "value") -->

<!-- quintiles_ordered <- as.character(unique(c_long$MT5_quintile)[order(unique(c_long$MT5_quintile))]) -->

<!-- # both <- ggplot(c_long , aes(x = month, y = value, color = patID_eye))+ -->
<!-- #   # geom_point(show.legend = FALSE, alpha = .3)+ -->
<!-- #   # geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+ -->
<!-- #   geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k = 5), show.legend = FALSE, aes(group = 1), se = TRUE)+ -->
<!-- #   facet_wrap(GENDER~variable, scales = "free" , ncol = 4)+ -->
<!-- #   theme_bw()+ -->
<!-- #   ggtitle(paste("macular zone thickness deltas from first visit (natural splines).\n"))+ -->
<!-- #    theme(plot.title = element_text(size = 8)) -->

<!-- height = 10; width = 5 -->

<!-- zones_quintile<- data.frame(variable = c_long$variable %>% unique(),  -->
<!--                              quintile_name = grep("quintile", value = TRUE, names(c_long))) -->


<!-- # Males -->
<!-- male_plots <- lapply(1:nrow(zones_quintile), function(i){ -->

<!--           # male_quintile_parafovea_facets <- ggplot(c_long %>% filter(GENDER == "Male"), aes(x = month, y = value, color = patID_eye))+ -->
<!--           #       # geom_point(show.legend = FALSE, alpha = .3)+ -->
<!--           #       # geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+ -->
<!--           #       geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k = 5), show.legend = FALSE, aes(group = 1), se = TRUE)+ -->
<!--           #       facet_wrap(variable~quintile_name, scales = "free", ncol = 16, nrow = 1 )+ -->
<!--           #       theme_bw()+ -->
<!--           #       ggtitle(paste("Male macular zone thickness deltas from first visit (natural splines).\n"))+ -->
<!--           #       theme(plot.title = element_text(size = 8)) -->
<!--           #  -->

<!--             zone_name <- zones_quintile[i, 1] -->
<!--             quintile_name <- zones_quintile[i, 2] -->

<!--             filtered_data <- c_long %>% -->
<!--                     filter(variable == zone_name) %>% -->
<!--                     filter(GENDER == "Male") %>%  -->
<!--                     mutate(quintile_group = get(quintile_name)) -->

<!--             plot <- ggplot(filtered_data, aes(x = month, y = value, color = patID_eye))+ -->
<!--                     geom_point(show.legend = FALSE, alpha = .3)+ -->
<!--                     geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+ -->
<!--                     # geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k = 5), show.legend = FALSE, aes(group = 1), se = TRUE)+ -->
<!--                     geom_smooth(method = "lm", show.legend = FALSE, aes(group = 1), se = TRUE)+ -->
<!--                     facet_wrap(~quintile_group, scales = "free" , nrow = 1)+ -->
<!--                     theme_bw()+ -->
<!--                     theme(plot.title = element_text(size = 8))+ -->
<!--                     ggtitle(zone_name) -->

<!--             if (i < 4 ){ -->

<!--               plot <- plot + theme(plot.title = element_text(size = 8),  -->
<!--                           axis.title.x  = element_blank()) -->
<!--             } -->


<!--             if(i == 1) {  -->
<!--               plot <- plot+ ggtitle(paste("Male macular zone thickness deltas from first visit (lm).\n", zone_name)) -->

<!--             } -->

<!--           return(ggplotGrob(plot)) -->
<!-- }) -->

<!-- male_quintile_facet_zone_grid <- gridExtra::marrangeGrob(male_plots, ncol = 1, nrow = 4) -->
<!-- ggsave(paste0(wd_path, "results/longitudinal_analysis/male_quintile_facet_zone_grid.png"), male_quintile_facet_zone_grid, width = width, height = height, dpi = 200) -->

<!-- # Females -->
<!-- female_plots <- lapply(1:nrow(zones_quintile), function(i){ -->

<!--             zone_name <- zones_quintile[i, 1] -->
<!--             quintile_name <- zones_quintile[i, 2] -->

<!--             filtered_data <- c_long %>% -->
<!--                     filter(variable == zone_name) %>% -->
<!--                     filter(GENDER == "Female") %>%  -->
<!--                     mutate(quintile_group = get(quintile_name)) -->

<!--             plot <- ggplot(filtered_data, aes(x = month, y = value, color = patID_eye))+ -->
<!--                     geom_point(show.legend = FALSE, alpha = .3)+ -->
<!--                     geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+ -->
<!--                     # geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k = 5), show.legend = FALSE, aes(group = 1), se = TRUE)+ -->
<!--                     geom_smooth(method = "lm", show.legend = FALSE, aes(group = 1), se = TRUE)+ -->
<!--                     facet_wrap(~quintile_group, scales = "free" , nrow = 1)+ -->
<!--                     theme_bw()+ -->
<!--                     theme(plot.title = element_text(size = 8))+ -->
<!--                     ggtitle(zone_name) -->

<!--             if (i < 4 ){ -->

<!--               plot <- plot + theme(plot.title = element_text(size = 8),  -->
<!--                           axis.title.x  = element_blank()) -->
<!--             } -->


<!--             if(i == 1) {  -->
<!--               plot <- plot+ ggtitle(paste("Female macular zone thickness deltas from first visit (lm).\n", zone_name)) -->

<!--             } -->

<!--           return(ggplotGrob(plot)) -->
<!-- }) -->

<!-- female_quintile_facet_zone_grid <- gridExtra::marrangeGrob(female_plots, ncol = 1, nrow = 4) -->
<!-- ggsave(paste0(wd_path, "results/longitudinal_analysis/female_quintile_facet_zone_grid.png"), female_quintile_facet_zone_grid, width = width, height = height, dpi = 200) -->

<!-- ``` -->

##### join quartiles and quintiles to d
```{r}
d <- d %>% 
  left_join(., c_long %>% select(patID_eye, ends_with("quartile"), ends_with("quintile")) %>% unique())

```


## parafovea raw values grid plot faceted by quintile 
```{r}

# make data ----
what_parafovea_cols <- grep("avg_MT0[2345]$", names(d), value = TRUE)

## filter data for completeness, visit, and mactel,  and make month number
parafovea_zones_d <- d %>% 
  filter( patID_eye %in% return_eyes) %>%
  select(patID_eye, patID, eye_num, visitID,all_of(what_parafovea_cols)) %>% 
  mutate(month = visitID %>% as.character() %>% as.numeric()) %>%  ## make visits numeric month
  na.omit()

## calculate quintiles of first visit zones
parafovea_zones_first_values <- parafovea_zones_d %>% 
  arrange( month) %>% 
  group_by(patID_eye) %>% 
  filter(row_number( ) == 1) %>% 
  ungroup() %>% 
  select(-month, -visitID) %>% 
  unique()

## categorize patient for first visit quintile and quintile
MT2_quintiles <- quantile(parafovea_zones_first_values$avg_MT02, seq(0,1,.2))
MT3_quintiles <- quantile(parafovea_zones_first_values$avg_MT03, seq(0,1,.2))
MT4_quintiles <- quantile(parafovea_zones_first_values$avg_MT04, seq(0,1,.2))
MT5_quintiles <- quantile(parafovea_zones_first_values$avg_MT05, seq(0,1,.2))

# quintile eyes' first visit value by first visit quintiles
parafovea_zones_quintile_d <- parafovea_zones_d %>% 
  select(patID_eye, patID, eye_num,  month, visitID, everything()) %>% 
  arrange(month) %>% 
  group_by(patID_eye) %>%
  filter(row_number( ) == 1) %>% 
  mutate( MT2_quintile = cut(avg_MT02, 
                             MT2_quintiles, 
                             labels = as.character(1:5), include.lowest = TRUE),
          MT3_quintile = cut(avg_MT03, 
                             MT3_quintiles, 
                             labels = as.character(1:5), include.lowest = TRUE),
          
          MT4_quintile = cut(avg_MT04, 
                             MT4_quintiles, 
                             labels = as.character(1:5), include.lowest = TRUE),
                         
          MT5_quintile = cut(avg_MT05 ,
                             MT5_quintiles, 
                             labels = as.character(1:5), include.lowest = TRUE)

        ) %>% 
  mutate(across(ends_with("quintile"), 
                ~factor(., levels =as.character(5:1)))) %>% 
  ungroup()

# lengthen 
parafovea_zones_quintile_d_long <- parafovea_zones_quintile_d %>% 
  pivot_longer(cols = matches("avg"), names_to = "zone")


#####################
### of raw values ###
#####################
# list of zones to loop over
zones_quintile <- data.frame(zone = parafovea_zones_quintile_d_long$zone %>% unique(), 
                             quintile_name = grep("quintile", value = TRUE, names(parafovea_zones_quintile_d_long)))
i = 4
plots <- lapply(1:nrow(zones_quintile), function(i){
            
            zone_name <- zones_quintile[i, 1]
            quintile_name <- zones_quintile[i, 2]
            
            filtered_data <- parafovea_zones_quintile_d_long %>%
                    filter(zone == zone_name) 
                    # mutate(quintile_group = get(quintile_name))
            
            # ymax is mean of highest quintile + buffer ymin is mean of lowest quintile - buffer
            ylims <- c(mean(filtered_data$value[filtered_data[[quintile_name]] == "1"])-10,  mean(filtered_data$value[filtered_data[[quintile_name]] == "5"]) + 10) 
            
            plot <- ggplot(filtered_data, aes(x = month, y = value, color = patID_eye))+
                    # geom_point(show.legend = FALSE, alpha = .3)+
                    geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+
                    # geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k = 5), show.legend = FALSE, aes(group = 1), se = TRUE)+
                    geom_smooth(method = "lm", show.legend = FALSE, aes(group = 1), se = TRUE)+
                    facet_wrap(~get(quintile_name), scales = "free_x" , nrow = 1)+
                    theme_bw()+
                    theme(plot.title = element_text(size = 8))+
                    coord_cartesian(ylim= c(ylims[1], ylims[2]))+
                    ggtitle(zone_name)
                    
            if (i < 4 ){
              
              plot <- plot + theme(plot.title = element_text(size = 8), 
                          axis.title.x  = element_blank())
            }
      
            
            if(i == 1) { 
              plot <- plot+ ggtitle(paste("Macular zone thickness from first visit (lm).\n", zone_name))
              
            }
            
          return(plot)
      
    })

plots_grobs <- sapply(plots, ggplotGrob)
raw_parafovea_grid <- gridExtra::marrangeGrob(grobs = plots,nrow = 4, ncol = 1) 

ggsave(paste0(wd_path, "results/longitudinal_analysis/zonequintileFacet_zoneFacet_parafovea_raw_values.png"), raw_parafovea_grid, width = 10, height = 10, dpi = 200)




```


#_________________

try to understand thickening first quantile by correlating to neovasc, leakage, and pigment, and hardexudates
1. 
* model using teh slope at each time point regressed against the specific phenotypes
* calculate the rate for every zone of each eye, categorize each patient based on the quantile of zone_5_rate, model thickness by thickness_quantile + rate_quantile, correlate rate and quantile
* calculate variability for all zones for each eye over time

# calculate the rate for every zone of each eye
```{r}

# find thickness delta/duration and join to d
rate_d <- d %>% 
  select(month, patID_eye, avg_MT01:avg_MT09) %>% 
  na.omit() %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  mutate(duration = dplyr::last(month) - dplyr::first(month)) %>%
  mutate(across(starts_with("avg"),
                 ~ (dplyr::last(.) - dplyr::first(.)) / duration,
                 .names = "{str_sub(.col, -4)}_rate")
          )%>%

ungroup() 

# join rate stuff to d
d <- left_join(d, rate_d %>% select(patID_eye, ends_with("rate")) %>% unique())



```

## categorize each eye by quantile of thicknening rate and thickness
```{r, eval = FALSE}

# df of quintile for each eye thickness and rate
quintile_d <- d %>% 
  filter(DIAGNOSIS_binary == 1) %>% 
  select(avg_MT05, MT05_rate, patID_eye, month, GENDER) %>% 
  filter(!is.na(avg_MT05)) %>% 
  group_by(patID_eye) %>% 
  filter(month== dplyr::first(month)) %>% 
  ungroup() %>% 
  unique() 


# quintile_d[is.na(quintile_d$GENDER) & quintile_d$duration !=0 , ]
# table(d$GENDER[d$duration == 0] %>% ifelse(is.na(.), "NA", .)) # lots of genderless at first visit


# categorize by quantile of mt05_rate and thickness
## rate quintile
#### men
male_mt05_rate_quintile <- quantile(quintile_d$MT05_rate[quintile_d$GENDER == "Male"], na.rm = TRUE, probs = seq(0, 1, .2)) %>% as.vector()
#### women
female_mt05_rate_quintile <- quantile(quintile_d$MT05_rate[quintile_d$GENDER == "Female"], na.rm = TRUE, probs = seq(0, 1, .2)) %>% as.vector()
### aSEXFual
aSEXF_mt05_rate_quintile <- quantile(quintile_d$MT05_rate[is.na(quintile_d$GENDER)], na.rm = TRUE, probs = seq(0, 1, .2)) %>% as.vector()

# ## thickness quintile
#### men
male_mt05_quintile <- quantile(quintile_d$avg_MT05[quintile_d$GENDER == "Male"], na.rm = TRUE, probs = seq(0, 1, .2)) %>% as.vector()
#### women
female_mt05_quintile <- quantile(quintile_d$avg_MT05[quintile_d$GENDER == "Female"], na.rm = TRUE, probs = seq(0, 1, .2)) %>% as.vector()
#### aSEXFual quintile
aSEXF_mt05_quintile <-  quantile(quintile_d$avg_MT05[quintile_d$GENDER %>% is.na()], na.rm = TRUE, probs = seq(0, 1, .2)) %>% as.vector()

# set min to 0 so first interval from 0 to first quintile lim
male_mt05_rate_quintile[1] <- -Inf 
female_mt05_rate_quintile[1] <- -Inf 
aSEXF_mt05_rate_quintile[1] <- -Inf 
male_mt05_quintile[1] <- 0 
female_mt05_quintile[1] <- 0 
aSEXF_mt05_quintile[1] <- 0 

x <- cbind(male_mt05_quintile, female_mt05_quintile, aSEXF_mt05_quintile, male_mt05_rate_quintile,female_mt05_rate_quintile, aSEXF_mt05_rate_quintile) %>% as_tibble( )  %>%  melt() %>% mutate(measure = ifelse(grepl("rate", variable), "rate", "thickness"))
ggplot(x, aes(x = value, y= variable, color = variable))+ geom_jitter(width = 0.2, height = 0, size = 5) + # Adjust the width for jittering
  theme_minimal()+
  facet_wrap(~measure, scales = "free")



# cataquintile by gender
quintile_d <- quintile_d %>% 
    mutate(rate_quintile = case_when(
        GENDER == "Male" ~ findInterval(MT05_rate, male_mt05_rate_quintile, left.open = TRUE, rightmost.closed = FALSE),
        GENDER == "Female" ~ findInterval(MT05_rate, female_mt05_rate_quintile, left.open = TRUE, rightmost.closed = FALSE),
        # is.na(GENDER)  ~ findInterval(MT05_rate, aSEXF_mt05_rate_quintile, left.open = TRUE, rightmost.closed = FALSE),
        TRUE ~ NA_real_
        )
    ) %>% 
    mutate(thickness_quintile = case_when(
        GENDER == "Male" ~ findInterval(avg_MT05, male_mt05_quintile, left.open = TRUE, rightmost.closed = FALSE),
        GENDER == "Female" ~ findInterval(avg_MT05, female_mt05_quintile, left.open = TRUE, rightmost.closed = FALSE),
        is.na(GENDER) ~ findInterval(avg_MT05, aSEXF_mt05_quintile, left.open = TRUE, rightmost.closed = FALSE),
        TRUE ~ NA_real_
        )
    ) 

unique(quintile_d$thickness_quintile)
unique(quintile_d$rate_quintile)

# join quintiles to patID_eye in d
d <- left_join(d, quintile_d[, c("patID_eye", "thickness_quintile", "rate_quintile")])

d[!is.na(d$avg_MT05) & is.na(d$thickness_quintile) ,  c("patID_eye", "visitID", "avg_MT05", "MT05_rate", "thickness_quintile", "rate_quintile", "GENDER", "DIAGNOSIS_binary")]
d[!is.na(d$avg_MT05) & is.na(d$rate_quintile)  ,  c("patID_eye", "visitID", "avg_MT05", "MT05_rate",  "thickness_quintile", "rate_quintile", "GENDER", "DIAGNOSIS_binary")]
```

# calculate each eye's variability at each zone and quintilize!
```{r, eval = FALSE}
# var for each eye group at each zone
d <- d %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  mutate(across(grep("avg_MT0\\d$", names(d), value = TRUE), 
                 ~ var(., na.rm = TRUE), 
                 .names = "{str_sub(.col, -4)}_variance")
          )%>% 
  ungroup() 

vars <- c("patID_eye", "visitID", "avg_MT05", "MT05_variance",  "MT05_rate", "thickness_quintile", "rate_quintile", "GENDER", "DIAGNOSIS_binary")

d %>% 
  select(any_of(vars)) %>% 
  filter(MT05_variance > 1000) %>% 
  arrange(patID_eye)
#^ lots of patients with exhorbitant variance   d$avg_MT05[d$patID_eye == "002021_1"] %>% var(., na.rm = TRUE)

plot(density(scale(d$MT05_variance), na.rm = TRUE), col = "blue" , xlim = c(-5, 14)); lines(density(scale(d$avg_MT05), na.rm = TRUE), col = "red"); 

# make quintiles
male_var_quintile <- quantile(d$MT05_variance[d$GENDER == "Male"], probs = seq(0,1, .2), na.rm = TRUE)
female_var_quintile <- quantile(d$MT05_variance[d$GENDER == "Female"], probs = seq(0,1, .2), na.rm = TRUE)
male_var_quintile[1] <- -1
female_var_quintile[1] <- -1


# plot quintiles over distributions
par(mfrow = c(2,1))
plot(density(d$MT05_variance[d$GENDER == "Male"], na.rm = TRUE), col = "blue", main = "Male variance with quintile lines", xlim = c(-1, 7000), lwd = 3); lapply(male_var_quintile, function(n){abline(v = n, col = "red", lty= 2)})
plot(density(d$MT05_variance[d$GENDER == "Female"], na.rm = TRUE), col = "blue", main = "Female variance with quintile lines", c(-1, 7000), lwd = 3); lapply(female_var_quintile, function(n){abline(v = n, col = "red", lty= 2)})
#^ variance is greatly right skewed
par(mfrow = c(1,1))

# categorize by quintile 
d <- d %>% 
  group_by(patID_eye) %>% 
  mutate(var_quintile = case_when(
        GENDER == "Male" ~ findInterval(MT05_variance, male_var_quintile, left.open = TRUE, rightmost.closed = FALSE),
        GENDER == "Female" ~ findInterval(MT05_variance, female_var_quintile, left.open = TRUE, rightmost.closed = FALSE),
        TRUE ~ NA_real_
        )
    ) %>% 
  ungroup()

d$var_quintile %>% table()

# y <- d$MT05_rate **2 %>% sqrt(.)
# y <- d$MT05_rate %>% abs(.)
y <- d$MT05_rate %>% na.omit()
x <- d$MT05_variance%>% na.omit()

# mod <- lm(x~poly(y, degree = 2))
mod <- lm(thickness_quintile ~ MT05_variance, data =d)
summary(mod)

plot.new()
plot(d[, c( "MT05_variance", "thickness_quintile")]); abline(mod, col = "red")


```

<!-- # categorize eyes for having abnormal (positive) overall mt05 rate -->
<!-- we do not expect zone 5 to thicken over time, what is the difference between patients with abnormal == true compared to those with abnormal == False -->
<!-- ```{r} -->
<!-- d$pos_MT05_rate <- ifelse(d$MT05_rate > 0, TRUE, FALSE) -->

<!-- hist(d$MT05_rate[d$pos_MT05_rate] ) -->

<!-- ``` -->

<!-- #__________________ -->
<!-- # modeling thickness and rate by phenotypes -->

<!-- ### look at delta000 over time for each quintile -->
<!-- ```{r} -->

<!-- ## filter data for completeness, visit, and mactel,  and make month number -->
<!-- b <- d %>%  -->
<!--   filter(month != 18) %>%  # ignore one 18 visit -->
<!--   filter( DIAGNOSIS_binary == 1) %>%  -->
<!--   select(avg_MT02:avg_MT05, patID, patID_eye, eye_num, visitID, GENDER, month, thickness_quintile) %>%  -->
<!--   na.omit()  -->

<!-- c <- b %>%  -->
<!--   arrange(patID, month) %>% ##  order by patient and then month -->
<!--   group_by(patID_eye) %>%  -->
<!--   # mutate(MT05_delta000= avg_MT05 - dplyr::first(avg_MT05)) %>% # calculage diff from 000 -->
<!--   mutate(across(grep("[0-9]$", names(b), value = TRUE),  -->
<!--                 ~ . - dplyr::first(.),  -->
<!--                 .names = "delta000_{str_sub(.col, -4)}") -->
<!--          ) %>%  -->
<!--   ungroup() -->

<!-- c %>% head() -->


<!-- c_long <- c %>%  -->
<!--   select(delta000_MT02:delta000_MT05,  patID_eye, month, thickness_quintile, GENDER) %>%  -->
<!--   pivot_longer(cols = starts_with("delta")) -->

<!-- M5_quintile_delta000 <- ggplot(c_long %>% filter(name == "delta000_MT05") , aes(x = month, y = value, color = patID_eye))+ -->
<!--   geom_point(show.legend = FALSE, alpha = .3)+ -->
<!--   geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .3)+ -->
<!--   geom_smooth(method = "lm", show.legend = FALSE, aes(group = 1), se = TRUE)+ -->
<!--   # geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k = 5), show.legend = FALSE, aes(group = 1), se = TRUE)+ -->
<!--   # facet_wrap(GENDER~name~thickness_quintile, scales = "free" , ncol = 5)+ -->
<!--   facet_wrap(GENDER~thickness_quintile, scales = "free" , ncol = 5)+ -->
<!--   theme_bw()+ -->
<!--   ggtitle(paste("macular zone thickness deltas from first visit (natural splines).\n"))+ -->
<!--   theme(plot.title = element_text(size = 8)) -->

<!-- ggsave( paste0(wd_path, "results/longitudinal_analysis/M5_quintile_delta000.png"), M5_quintile_delta000, width = 15, height = 10, dpi = 100) -->
<!-- ``` -->

<!-- #### regress against thickness regardless of quintile, and mixed effects -->
<!-- ```{r } -->

<!-- phenotypes <-  c("NEOVASC", "RPE_LK05", "HRDEXUDATE", "INTRARET_PGMT_MIGR", "LOCAL_DECR_AF_VESL_END", "IR05", "HR_TEMPORAL", "DRV05", "RAV05", "LRT05") # "EZLOSS" -->

<!-- ## -->
<!-- # x = "EZLOSS" -->
<!-- ## -->

<!-- phenotypes_mods <- lapply(phenotypes, function(x){ -->
<!--   # variables to scale -->
<!--   vars <- c("avg_MT05", x ) -->

<!--   a <- d %>% -->
<!--   # filter(thickness_quintile == quintile_n) %>%  -->
<!--   select( SEXF, patID ,eye_num,  month,patID_eye, all_of(vars)) %>%  -->
<!--   na.omit() %>%  -->
<!--   mutate(across(all_of(vars), ~scale(.)) -->
<!--          ) %>%  -->
<!--     as.data.frame() -->

<!--   # formula <- as.formula(paste0("avg_MT05 ~ SEXF + ", x)) -->
<!--   # formula <- as.formula(paste0("avg_MT05 ~ SEXF + month + (1|patID) +", x)) -->
<!--   # formula <- as.formula(paste0("avg_MT05 ~ SEXF +  (1|month) + (1|patID) + (1|eye_num) +", x)) -->
<!--   # formula <- as.formula(paste0("avg_MT05 ~ SEXF + (1|month/patID/eye_num)  +", x)) -->

<!--   # m_1 <-  lmerTest::lmer(formula, data = a, REML = TRUE, verbose = FALSE)  -->
<!--   formula <- as.formula(paste0("avg_MT05 ~ SEXF + month + (1|patID_eye)  + ", x)) -->




<!--   if(nrow(a) > 10){ -->
<!--   return( -->
<!--      lmerTest::lmer(formula, data = a, REML = TRUE, verbose = FALSE)  -->
<!--   ) -->

<!--   } -->
<!-- }) -->
<!-- names(phenotypes_mods) <- phenotypes -->


<!-- # exclude null models  -->
<!-- phenotypes_mods <- Filter(function(x) !is.null(x), phenotypes_mods) -->

<!-- # make model summaries -->
<!-- phenotypes_mods_summaries <- lapply(phenotypes_mods, summary) # -->

<!-- # Initialize an empty data frame to store results -->
<!-- results_df <- data.frame(Model = character(),  -->
<!--                          Term = character(),  -->
<!--                          Estimate = numeric(),  -->
<!--                          Std.Error = numeric(), -->
<!--                          t.value = numeric(), -->
<!--                          Pr = numeric(),  -->
<!--                          stringsAsFactors = FALSE) -->

<!-- # Loop through each model and extract information -->
<!-- for (model_name in names(phenotypes_mods_summaries)) { -->
<!--   model_coef <- coef(phenotypes_mods_summaries[[model_name]]) -->
<!--   model_df <- data.frame(Model = model_name,  -->
<!--                          Term = rownames(model_coef),  -->
<!--                          Estimate = model_coef[, "Estimate"], -->
<!--                          Std.Error = model_coef[, "Std. Error"], -->
<!--                          t.value = model_coef[, "t value"], -->
<!--                          Pr = model_coef[, "Pr(>|t|)"]) -->
<!--   results_df <- rbind(results_df, model_df) -->
<!-- } -->

<!-- # filter control variables from results -->
<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->
<!--   # mutate(quintile = quintile_n) -->

<!-- # calculate adjusted p value -->
<!-- results_df$q_value <- p.adjust(results_df$Pr, method = "BH") -->


<!-- thickness_SEXF_phenotypes_nonquintile_plot <- ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( Pr < .05,  Model, NA)), size = 2, -->
<!--                        nudge_y = 0.5,  # Adjust these values as needed -->
<!--                        show.legend = FALSE)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("MT05_thickness ~ SEXF + month + (1|patID_eye) + phenotype") -->


<!-- ggsave( paste0(wd_path, "results/longitudinal_analysis/thickness_SEXF_phenotypes_nonquintile_plot.png"), thickness_SEXF_phenotypes_nonquintile_plot, width = 8, height = 5, dpi = 300) -->

<!-- ``` -->

# ||||||||||||||||||||

# save info
```{r}
new_info_path <- info_path %>% gsub("_03.xlsx", "_04.xlsx",.)
write.xlsx(info, new_info_path)

```


# save d
```{r}

new_data_name <- data_name %>% 
  gsub("4", "5", .)
cat(sprintf("Saving... \n%s\n", new_data_name))
saveRDS(d, paste0(wd_path, "processed_data/", new_data_name))

```


```{r, look at dims}
check_dims()
# dims : 12600 319
#                      [,1]
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516
# ^ added MT stats vars

### MARCH 14 2025 ###
# dims : 12619 306
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500


```
