---
title: "08_functionAndStructure"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---

# options
```{r}
# 
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()
```


# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}
library(tidyr)
library(dplyr)
library(reshape2)
library(purrr)
library(GGally)
library(tidyverse)
library(ggpubr)
library(vcd)
library(UpSetR)
library(ggplot2)
library(gridExtra)
library(BiocSingular)
library(ComplexHeatmap)
library(viridis)
library(knitr)
library(htmlwidgets)
library(zoo)
library(corrplot)
library(RColorBrewer)
library(openxlsx)
library(plotly)
library(RColorBrewer)
library(mgcv)
library(zoo)
library(tools)
library(mice)
library(grDevices)
library(ggrepel)
library(networkD3)
library(htmltools)
library(grid)
library(gplots)
library(Metrics)
library(psych)
library(car)
library(stats)
library(lme4)
library(reshape2)



rm(list = ls())
gc()

source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```

Task: measure zone progression for each phenotype

# Load Data

```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}
wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

# # raw list of dfs
# raw <- readRDS(paste0(wd_path, "processed_data/dfs_list_01.rds"))

# joined cleaned joined data
data_name <- "CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_05.rds"
d_raw <- readRDS(paste0(wd_path, "processed_data/", data_name))# clean data with everything 

d <- d_raw

# grading nomenclature and information
info_path <- paste0(wd_path, "processed_data/database_grading_key_04.xlsx") 

info <- read.xlsx(info_path)
# ****** FUnction to query name in info ******
var_summary <- function(names_vec = character()){
  patt <- paste(names_vec, collapse = "|")  
  m <- as.matrix(filter(info, grepl(patt, name, ignore.case = TRUE)))
  rownames(m) <- m[,1]
  print(kable(t(m), "simple"))
  for(nm in m[,'name']){
    cat(sprintf('\n%s\n', nm))
    print(table(d[[nm]]))
  }
}

```


# make new directory for plots
```{r}
path <- paste0(wd_path, "results/VAA_analysis/")
if(!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
  
}

path <- paste0(wd_path, "results/VAA_analysis/missingness/")
if(!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
  
}

path <- paste0(wd_path, "results/VAA_analysis/imputation_testing/")
if(!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
  
}

path <- paste0(wd_path, "results/VAA_analysis/FA/")
if(!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
  
}

```

# clean Data
```{r, look at dims}
check_dims <- function(D = d){
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()
# dims : 12600 319
#                      [,1]
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516
# ^ no changes

### MARCH 14 2025 ###
# dims : 12619 306
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```


#____________________________________________
# look at VAA variables

## barplot of number of records in VAA in dataset
```{r}



d %>% 
  select(eye_num, any_of(info$name[info$table=="VAA" & !is.na(info$table)] )  ) %>% 
  group_by(eye_num) %>% 
  reframe(across(everything(), ~sum(!is.na(.)))) %>%
  pivot_longer(-eye_num) %>% 
  ggplot( aes(x= name, y = value, fill = factor(eye_num)))+
  geom_bar(stat = "identity", show.legend = TRUE, position = "dodge")+
  ggtitle("counts of available records for VAA variables")+
  theme_bw()+
  theme(axis.text.x = element_text(angle =90, hjust = 1))
ggsave(paste0(wd_path, "results/VAA_analysis/missingness/vaa_counts_labeled_bar.png"), width = 15, height = 5, dpi = 300)

```


#____________________________________________
# caluculate missing scores from others
### calculate missing logmar 
```{r}

## missing logmarod where acuit1ar? ######
LOGMAR_mod <- lm(LOGMAR ~ ACUIT, data = d)

summary(LOGMAR_mod)

# plot them with calculated line
plot(d$LOGMAR ~ d$ACUIT, main = "");abline(LOGMAR_mod, col = "blue", lwd = 3)


# where to predict logmar
where_acuit_not_logmar <- is.na(d$LOGMAR) & !is.na(d$ACUIT)

# fix missing acuity scoress: calculate where acuity observed and logmar missing
d$LOGMAR[where_acuit_not_logmar] <-   predict(LOGMAR_mod, data.frame(ACUIT = d$ACUIT[where_acuit_not_logmar] ))

```

### calculate missing snellen
missing snellen where acuit?, predict snellen where acuity 
```{r}

# TRAIN MODELS----
## model for left eye
SNELL_formula <-
  "SNELL ~ " %>% paste(., "+ I(10 ^ LOGMAR)") %>%
  # "log10(SNELL) ~ " %>% paste(., "LOGMAR") %>%
  as.formula()

SNELL_mod <- lm(SNELL_formula, data = d)
plot(10^(d$LOGMAR) , d$SNELL); add_lines("d", col = "black")
abline(SNELL_mod, col = "blue", lty  = 5, lwd =3)
legend("topleft", legend = paste(SNELL_formula[c(2,1,3)], collapse = " ") )

# (((((((((((((((((())))))))))))))))))
# fix missing acuity scores using model ----
# (((((((((((((((((())))))))))))))))))

# double check models
test <- select(d, matches("SNELL|LOG")) %>% 
  drop_na() %>% 
  mutate(pred =predict(SNELL_mod,. )) %>% 
  select(matches("pred|SNELL"))
plot(test)
  
# where to predict snellen
where_logmar_not_snell <- is.na(d$SNELL) & !is.na(d$LOGMAR)
# predict using models
d$SNELL[where_logmar_not_snell] <- predict(SNELL_mod, d[where_logmar_not_snell ,"LOGMAR" ] )



```

<!-- ### **re-plot non_missing counts bar plot  -->
<!-- ```{r} -->


<!-- vaa_count_df <- d %>%  -->
<!--   select( all_of(c(info$variable_new[info$dataset == "VAA" & info$in_d], "month", "patID", "eye_num"))) %>%  -->
<!--   group_by(month) %>%  -->
<!--   summarise(across(everything(), ~ sum(!is.na(.)), .names = "{.col}")) %>%  -->
<!--   melt(id.vars = "month") %>%  -->
<!--   group_by(variable) %>%  -->
<!--   mutate(var_total = sum(value)) %>%  -->
<!--   ungroup() %>%  -->
<!--   group_by(month) %>%  -->
<!--   mutate(month_total = sum(value)) %>%  -->
<!--   ungroup() -->

<!-- # factor variable to have order of df -->
<!-- vaa_count_df <- vaa_count_df %>%  -->
<!--   mutate(variable = factor(variable, levels = count_plot_levels)  ) %>%  -->
<!--   mutate(month_group = factor(ifelse(month > 0, "0<", "0"), levels  = c("0", "0<")) ) -->

<!-- # vAA overall counts -->
<!-- VAA_colsum_plot <-ggplot(vaa_count_df, aes(x = value ,y =variable , fill = variable))+ -->
<!--   geom_bar(stat = "identity", show.legend = FALSE, orientation = "y", width = .5, just = 0)+ -->
<!--   theme_light()+ -->
<!--   # theme(axis.text.x = element_text( hjust = 1, vjust = 0, size = 10, angle = 90, -->
<!--   #                                   margin = margin(t = -100, r = 0, b = 0, l = 0)  -->
<!--   #                                   ) -->
<!--   #       )+ -->
<!--   geom_abline(intercept = nrow(d), col = "blue", linetype = 2, slope = 0)+ -->
<!--   ggtitle("VAA available counts after derivations on logmar and snellen from acuity") -->

<!-- ggsave(paste0(wd_path, "results/VAA_analysis/missingness/VAA_postCalc_colsum_plot.png"),VAA_colsum_plot, width = 15, height = 8, dpi = 200, limitsize = F) -->

<!-- ### same plot but faceted for visitID #### -->
<!-- VAA_colsum_facetVisit_plot <- ggplot(vaa_count_df, aes(y = variable ,x= value, fill = variable))+ -->
<!--   geom_bar(stat = "identity", show.legend = FALSE, orientation = "y", width = .5, just = 0)+ -->
<!--   theme_bw()+ -->
<!--   # theme(axis.text.x = element_text( hjust = 1, vjust = 0, size = 4, angle = 90, -->
<!--   #                                   margin = margin(t = 0, r = 0, b = 0, l = 0)  -->
<!--   #                                   ) -->
<!--   #       )+ -->
<!--   geom_abline(intercept = nrow(d), col = "blue", linetype = 2, slope = 0)+ -->
<!--   facet_wrap(~month_group, scales = "fixed")+ -->
<!--   ggtitle("VAA available counts after derivations on logmar and snellen from acuity") -->
<!-- ggsave(paste0(wd_path, "results/VAA_analysis/missingness/VAA_postCalc_colsum_facetVisit_plot.png"),VAA_colsum_facetVisit_plot, width = 15, height = 10, dpi = 200, limitsize = F) -->

<!-- ``` -->

#____________________________________________
# clean VAA 

## recode VAA vars
```{r}

d_new <- d

prescrip_columns <- c("CYLPM","SPHPM")

## fix prescrip
d_new <- d_new %>% 
  mutate(across(all_of(prescrip_columns), ~case_when(
    . == "+" ~ 1, 
    . == "-" ~ -1, 
    . == "Plano" ~ 0, 
    . == "No correction needed" ~ 0, 
    TRUE ~ NA_real_
  ))
         )

```

# plot heatmap of VAA correlations
```{r}


VAA_vars <- info$name[!is.na(info$table) & info$table == "VAA"] %>% 
  grep("DATE|TH|1$", ., value = TRUE, invert = TRUE)

VAA_vars <- intersect(VAA_vars, names(d_new))

# select VAA variables; exclude 1m rowscores
vaa_m <- d_new %>% 
  select(all_of(VAA_vars)) %>% 
  ungroup()

corr_m <- cor(vaa_m, method = "pearson", use = "pairwise.complete.obs") 

# # make a filtered matrix for coloring
# to each row set zero if outside [-threshold, threshold]
# threshold = 0
# corr_m_filtered <- apply(corr_m, 1, function(x){ ifelse(x>(-1*threshold) & x< threshold,0, x)})
# corr_m_filtered <- apply(corr_m, 1, function(x){ if(x>(threshold * -1) & x < threshold) 0 else x })

# par(mfrow = c(1, 2))
# plot filtered matrix but label with original
# gplots::heatmap.2(corr_m, 
#                   Rowv = TRUE,
#                   symm = TRUE,
#                   hclustfun = function(x) hclust(x, method = "ward.D2"),
#                   dendrogram = "row", 
#                   Colv = TRUE, 
#                   col =bluered,
#                   margins = c(20, 15),
#                   trace = "none", 
#                   cellnote=corr_m, 
#                   key = FALSE)

png(sprintf("%sresults/VAA_analysis/VA_corr_hm.png", wd_path), height = 1000, width = 1000)
Heatmap(corr_m,
        column_title = "VAA Correlations", 
        name = "r", 
        cluster_rows = TRUE, 
        cluster_columns = TRUE,
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2",
        col = rev(c("red", "white", "blue"))
        ) 
dev.off()

# gplots::heatmap.2(vaa_imp_corr_m, 
#                   Rowv = TRUE,
#                   symm = TRUE,
#                   hclustfun = function(x) hclust(x, method = "ward.D2"),
#                   dendrogram = "row", 
#                   Colv = TRUE, 
#                   col =bluered,
#                   margins = c(20, 15),
#                   trace = "none", 
#                   cellnote=vaa_imp_corr_m,
#                   key = FALSE)



```

#____________________________________________
# impute VAA

## look at potential vision improvement over time
```{r}

va_over_time <- d_new %>% 
  select(patID_eye, month, matches("RW\\d.*4$")) %>%
  arrange(patID_eye, month) %>%
  group_by(patID_eye) %>% 
  mutate(across(matches("RW"), 
                ~.-dplyr::first(.))) %>% # normalize to BL
  ungroup()%>% 
  drop_na()

# make long
va_over_time_long <- va_over_time %>% 
  # filter(if_any(matches("RW"), ~.!=lead(.) & .!=lag(.)), .by = patID_eye) %>% # filter for recoreds that mark change
  # filter(patID_eye %in% va_improvers  ) %>% 
  pivot_longer(-c(month,patID_eye)) 
  
ggplot(va_over_time_long,aes(month, value))+
  geom_step( aes(group = patID_eye),color = "black", alpha = .1,show.legend = FALSE)+
  # geom_line( aes(group = name),color = "black", alpha = .3, show.legend = FALSE)+
  geom_smooth(aes(group = name), color = "darkorange")+
  # geom_smooth(aes(color = name))+
  facet_wrap(~name, scales = "free_x", ncol = 5)+
  theme_bw()+
  ggtitle("BL normalized row scores")+
  theme(aspect.ratio = .8)+
  geom_blank()

ggsave(paste0(wd_path, "results/VAA_analysis/bl_norm_rowscores_over_time.png"), height = 5, width = 15)

```



## test impute data
start by finding best method

#### 'TRUE' to run imputation evaluation chunks
```{r}
#
##
###
####
run_test_imputes <- FALSE
####
###
##
#
```


### make df to test
#### first make truth df

```{r}
#VAA_vars= "AXISA" "CYLPM" "SNELL" "SPHPM" "ACUIT"  "CYLNUMA" "LOGMAR" "RW014"  "RW024"  "RW034"  "RW044"  "RW054"  "RW064"  "RW074"  "RW084"  "RW094"  "RW104"  "RW114"  "RW124"  "RW134"  "RW144"  "SPHNUMA"
# what rows have no na

rows_to_impute <- rowSums(is.na(d_new[, VAA_vars])) ==0

# complete df to test with excluding logmar because of collinearity
truth <- d_new[rows_to_impute, ] %>%
  select(all_of(VAA_vars)) %>% 
  select(-LOGMAR) %>% 
  as.matrix()

```

#### test dataframe randomly assigning NAs
<!-- ```{r} -->
<!-- # sample nas -->
<!-- prop_na <- 0.2 -->
<!-- # Calculate the number of NAs to introduce -->
<!-- num_na <- ceiling(nrow(truth) * ncol(truth) * prop_na) -->

<!-- # Create a vector of indices for the dataframe -->
<!-- indices <- expand.grid(1:nrow(truth), 1:ncol(truth)) -->

<!-- # Sample from these indices -->
<!-- sampled_indices <- sample(1:nrow(indices), size = num_na, replace = FALSE) -->

<!-- test <- truth -->
<!-- # Assign NAs -->
<!-- for (idx in sampled_indices) { -->
<!--   test[indices[idx, 1], indices[idx, 2]] <- NA # at row column in test for row column of index df, set to NA -->
<!-- } -->

<!-- # look at missing -->
<!-- row_na <- indices[sampled_indices, 1] %>% unique() -->
<!-- col_na <- indices[sampled_indices, 2] %>% unique() -->

<!-- test[row_na, ] -->
<!-- truth[row_na, ] -->
<!-- ``` -->



#### make test df with whole rows missing from categories
* three categories: prescription, row scores, overall scores

```{r}
#VAA_vars= "ACUIT"  "AXIS"   "CYLNUM" "CYLPM"  "LOGMAR" "RW014"  "RW024"  "RW034"  "RW044"  "RW054"  "RW064" "RW074"  "RW084"  "RW094"  "RW104"  "RW114"  "RW124"  "RW134"  "RW144"  "SNELL"  "SPHNUM" "SPHPM" 

# sample nas
prop_na <- 0.3

# Calculate the number of NAs rows to introduce
num_na <- ceiling(length(truth) * prop_na)

# df to assign nas to
test <- truth

# sample missing as 1d inds
inds_1d <- sample(seq(length(truth)), num_na )
 
test[inds_1d] <- NA
inds_2d <- which(is.na(test), arr.ind = TRUE)

# # Create a vector of indices for the dataframe
# ## sample rows from each category of truth so that no na rows overlap
# prescrip_na_rows <- sample(1:nrow(truth), num_na, replace = FALSE)
# RW_na_rows <- sample( setdiff(1:nrow(truth), prescrip_na_rows),num_na, replace = FALSE)
# scores_na_rows <- sample( setdiff(1:nrow(truth), c(prescrip_na_rows, RW_na_rows)),num_na, replace = FALSE)
# 
# ## vectors of columns for each category
# prescrip_na_cols <- grep("CYL|SPH|AXIS", VAA_vars, value = TRUE)
# RW_na_cols <- grep("RW\\d", VAA_vars, value = TRUE)
# scores_na_cols <- grep("ACUI|SNEL", VAA_vars, value = TRUE)
# # setequal(c(prescrip_na_cols, RW_na_cols, scores_na_cols), VAA_vars)
# 
# # Assign NAs to row indices for respective category
# test[prescrip_na_rows, prescrip_na_cols] <- NA
# test[RW_na_rows, RW_na_cols] <- NA
# test[scores_na_rows, scores_na_cols] <- NA

## check if same number na in each
# is.na(test[c(prescrip_na_rows), prescrip_na_cols ][1] ) %>% sum()
# is.na(test[c(RW_na_rows), RW_na_cols][1]) %>% sum()
# is.na(test[c(scores_na_rows), scores_na_cols][1]) %>% sum()
# 

```

##### construct predictor matrix
```{r}

predictor_m <- matrix(0, ncol = ncol(test),nrow = ncol(test), dimnames = list(colnames(test), colnames(test)) )

predictor_m[grep("RW|SNELL", colnames(predictor_m)), "ACUIT"] <-  1 # rw and snell log and acuit
predictor_m[grep("RW|ACUIT", colnames(predictor_m)), "SNELL"] <-  1 # same
predictor_m[grep("UMA|PMA|AXIS", colnames(predictor_m)), grep("UMA|PMA|AXIS", colnames(predictor_m))] <-  1 # prescriptions
predictor_m[grep("RW", colnames(predictor_m)), grep("RW", colnames(predictor_m))] <-  1 # row scores with selff
predictor_m[grep("RW0[4-9]|RW[10-14]", colnames(predictor_m)), grep("RW0[1-3]", colnames(predictor_m))] <-  0 # high rows not correlated to low

# remove diagonal
diag(predictor_m) <- 0

# Heatmap(predictor_m)
# qpred <- quickpred(test,mincor = .1, minpuc = .6)
# Heatmap(qpred)

```


### PMM x pred matrix ####
```{r}

if(run_test_imputes){

  method = "pmm"
  pmm_no_PredM <- mice(test, method = method)
  pmm_no_PredM_sum <- complete(pmm_no_PredM, action =1) ; for (i in 1:pmm_no_PredM$m){pmm_no_PredM_sum <-pmm_no_PredM_sum + complete(pmm_no_PredM, action =i )}
  pmm_no_PredM_mean <- pmm_no_PredM_sum / pmm_no_PredM$m
  pmm_no_PredM_mean <- data.matrix(pmm_no_PredM_mean)
  #^ doesn't imput Logmar becasue of collinearity, assess snellen an acuity as proxy

  pmm_PredM <- mice(test, method = method, predictorMatrix = predictor_m);
  pmm_PredM_sum <- complete(pmm_PredM, action =1) ; for (i in 1:pmm_PredM$m){pmm_PredM_sum <-pmm_PredM_sum + complete(pmm_PredM, action =i )} ;
  pmm_PredM_mean <- pmm_PredM_sum / pmm_PredM$m
  pmm_PredM_mean <- data.matrix(pmm_PredM_mean)
  #^ doesn't imput Logmar becasue of collinearity, assess snellen an acuity as proxy

    rmse_d <- data.frame(var= character(), RMSE_1 = numeric(), RMSE_2 = numeric(), stringsAsFactors = FALSE)

    # plot
    h = 1500
    w = 1700
    png(paste0(wd_path, "results/VAA_analysis/imputation_testing/PMM_with_predMatrix.png"), height = h, width = w)
    par(mfrow = c(6,7))
    # n = "ACUIT"
    for (n in colnames(truth)){
      
      # find inds of missing for column n
      inds <- inds_2d[ inds_2d[,2] == which(colnames(truth) == n) ,]
      
      rmse_1 <- rmse(truth[inds] %>% scale(), pmm_no_PredM_mean[inds] %>% scale())
      rmse_2 <- rmse(truth[inds]%>% scale(), pmm_PredM_mean[inds]%>% scale())
      cor_1 <- cor(pmm_no_PredM_mean[inds], truth[inds])
      cor_2 <- cor(pmm_PredM_mean[inds], truth[inds])

      cex = 1
      position <-range(pmm_no_PredM_mean[inds])[1] + (range(pmm_no_PredM_mean[inds])[2] -range(pmm_no_PredM_mean[inds]) [1])/ 2
      if(position == 0){position = .4}
      plot(pmm_no_PredM_mean[inds]~ truth[inds], col = "blue", main = paste("pmm, no predictor matrix\n", n)); abline(c(0, 1), col = "blue") ;
      text(x = position, y = position, labels = paste("RMSE:", round(rmse_1, 2)), font = 2, cex = cex)
      text(x = position, y = position-position/8, labels = paste("r:", round(cor_1, 2)), font = 2, cex = cex)

      position <-range(pmm_PredM_mean[inds])[1] + (range(pmm_PredM_mean[inds])[2] -range(pmm_PredM_mean[inds]) [1])/ 2
      if(position == 0){position = .4}

      plot(pmm_PredM_mean[inds]~ truth[inds], col = "red", main = paste("pmm, predictor matrix for like variables\n", n)); abline(c(0, 1), col = "red")
      text(x = position, y = position, labels = paste("RMSE:", round(rmse_2, 2)), font = 2, cex = cex)
      text(x = position, y = position-position/8, labels = paste("r:", round(cor_2, 2)), font = 2, cex = cex)

      rmse_d[n,] <- data.frame(var = n, RMSE_1= rmse_1, RMSE_2 =rmse_2)

    }
    dev.off()
    par(mfrow = c(1,1))

    c(wo_predM  = rmse_d$RMSE_1 %>% mean(., na.rm = TRUE) ,    with_predM= rmse_d$RMSE_2 %>% mean(., na.rm = TRUE) )
      # wo_predM   with_predM
      # 0.7470695  1.2763752
      # wo_predM   with_predM 
      # 1.248221   1.290610 
      # wo_predM with_predM 
      # 0.6376694  0.7539783 
}
```
>> model doesn't impute Logmar becasue of collinearity, assess snellen and acuity as proxy
>> => calculate logmar from acui after impute*

<!-- ### lasso.norm x pred matrix #### -->
<!-- ```{r} -->
<!-- if(run_test_imputes){ -->

<!--     method = "lasso.norm" -->

<!--     lasso_no_PredM <- mice(test, method = method) -->
<!--     lasso_no_PredM_sum <- complete(lasso_no_PredM, action =1) ; for (i in 1:lasso_no_PredM$m){lasso_no_PredM_sum <-lasso_no_PredM_sum + complete(lasso_no_PredM, action =i )} -->
<!--     lasso_no_PredM_mean <- lasso_no_PredM_sum / lasso_no_PredM$m -->


<!--     lasso_PredM <- mice(test, method = method, predictorMatrix = predictor_m); -->
<!--     lasso_PredM_sum <- complete(lasso_PredM, action =1) ; for (i in 1:lasso_PredM$m){lasso_PredM_sum <-lasso_PredM_sum + complete(lasso_PredM, action =i )} ; -->
<!--     lasso_PredM_mean <- lasso_PredM_sum / lasso_PredM$m -->

<!--     # -->
<!--     # n = "ACUIT" -->
<!--     # -->

<!--     rrmse_d <- data.frame(var= character(), RMSE_1 = numeric(), RMSE_2 = numeric(), stringsAsFactors = FALSE) -->

<!--     # plot -->
<!--     h = 1500 -->
<!--     w = 1700 -->
<!--     png(paste0(wd_path, "results/VAA_analysis/imputation_testing/lasso.norm_with_predMatrix.png"), height = h, width = w) -->
<!--     par(mfrow = c(6,7)) -->
<!--     # n = "ACUIT" -->
<!--     for (n in colnames(truth)){ -->

<!--       # find inds of missing for column n -->
<!--       inds <- inds_2d[ inds_2d[,2] == which(colnames(truth) == n) ,] -->

<!--       rmse_1 <- rmse(truth[inds] %>% scale(), lasso_no_PredM_mean[inds] %>% scale()) -->
<!--       rmse_2 <- rmse(truth[inds]%>% scale(), lasso_PredM_mean[inds]%>% scale()) -->
<!--       cor_1 <- cor(lasso_no_PredM_mean[inds], truth[inds]) -->
<!--       cor_2 <- cor(lasso_PredM_mean[inds], truth[inds]) -->

<!--       cex = 1 -->
<!--       position <-range(lasso_no_PredM_mean[inds])[1] + (range(lasso_no_PredM_mean[inds])[2] -range(lasso_no_PredM_mean[inds]) [1])/ 2 -->
<!--       if(position == 0){position = .4} -->
<!--       plot(lasso_no_PredM_mean[inds]~ truth[inds], col = "blue", main = paste("lasso, no predictor matrix\n", n)); abline(c(0, 1), col = "blue") ; -->
<!--       text(x = position, y = position, labels = paste("RMSE:", round(rmse_1, 2)), font = 2, cex = cex) -->
<!--       text(x = position, y = position-position/8, labels = paste("r:", round(cor_1, 2)), font = 2, cex = cex) -->

<!--       position <-range(lasso_PredM_mean[inds])[1] + (range(lasso_PredM_mean[inds])[2] -range(lasso_PredM_mean[inds]) [1])/ 2 -->
<!--       if(position == 0){position = .4} -->

<!--       plot(lasso_PredM_mean[inds]~ truth[inds], col = "red", main = paste("lasso, predictor matrix for like variables\n", n)); abline(c(0, 1), col = "red") -->
<!--       text(x = position, y = position, labels = paste("RMSE:", round(rmse_2, 2)), font = 2, cex = cex) -->
<!--       text(x = position, y = position-position/8, labels = paste("r:", round(cor_2, 2)), font = 2, cex = cex) -->

<!--       rmse_d[n,] <- data.frame(var = n, RMSE_1= rmse_1, RMSE_2 =rmse_2) -->

<!--     };par(mfrow = c(1,1)) -->
<!--     dev.off() -->

<!--     c(lassoNorm_wo_predM = rmse_d$RMSE_1 %>% mean(., na.rm = TRUE), lassoNor_w_predM = rmse_d$RMSE_2 %>% mean(., na.rm = TRUE) ) -->
<!--     # lassoNorm_wo_predM   lassoNor_w_predM -->
<!--     #      0.7529663          1.2702938 -->
<!--     # lassoNorm_wo_predM   lassoNor_w_predM -->
<!--     #      0.9423242          1.1695645 -->
<!-- } -->
<!-- ``` -->


### lasso.norm x PMM ####
```{r}
if(run_test_imputes){

   method = c("pmm", "lasso.norm")

    pmm_imp <- mice(test, method = method[1])
    pmm_imp_sum <- complete(pmm_imp, action =1) ; for (i in 1:pmm_imp$m){pmm_imp_sum <-pmm_imp_sum + complete(pmm_imp, action =i )}
    pmm_imp_mean <- pmm_imp_sum / pmm_imp$m

    names(test)[map(test , ~var(.x, na.rm = TRUE))<.5]
    
    lasso_imp <- mice(test, method = method[2]);
    lasso_imp_sum <- complete(lasso_imp, action =1) ; for (i in 1:lasso_imp$m){lasso_imp_sum <-lasso_imp_sum + complete(lasso_imp, action =i )} ;
    lasso_imp_mean <- lasso_imp_sum / lasso_imp$m


    rmse_d <- data.frame(var= character(), RMSE_1 = numeric(), RMSE_2 = numeric(), stringsAsFactors = FALSE)

    # plot
    h = 1500
    w = 1700
    png(paste0(wd_path, "results/VAA_analysis/imputation_testing/lassoNorm_x_PMM.png"), height = h, width = w)
    par(mfrow = c(6,7))
    # n = "ACUIT"
    for (n in colnames(truth)){
      
      # find inds of missing for column n
      inds <- inds_2d[ inds_2d[,2] == which(colnames(truth) == n) ,]
      
      rmse_1 <- rmse(truth[inds] %>% scale(), pmm_imp_mean[inds] %>% scale())
      rmse_2 <- rmse(truth[inds]%>% scale(), lasso_imp_mean[inds]%>% scale())
      cor_1 <- cor(pmm_imp_mean[inds], truth[inds])
      cor_2 <- cor(lasso_imp_mean[inds], truth[inds])

      cex = 1
      position <-range(pmm_imp_mean[inds])[1] + (range(pmm_imp_mean[inds])[2] -range(pmm_imp_mean[inds]) [1])/ 2
      if(position == 0){position = .4}
      plot(pmm_imp_mean[inds]~ truth[inds], col = "blue", main = paste("pmm, no predictor matrix\n", n)); abline(c(0, 1), col = "blue") ;
      text(x = position, y = position, labels = paste("RMSE:", round(rmse_1, 2)), font = 2, cex = cex)
      text(x = position, y = position-position/8, labels = paste("r:", round(cor_1, 2)), font = 2, cex = cex)

      position <-range(lasso_imp_mean[inds])[1] + (range(lasso_imp_mean[inds])[2] -range(lasso_imp_mean[inds]) [1])/ 2
      if(position == 0){position = .4}

      plot(lasso_imp_mean[inds]~ truth[inds], col = "red", main = paste("lasso.norm, no predictor matrix\n", n)); abline(c(0, 1), col = "red")
      text(x = position, y = position, labels = paste("RMSE:", round(rmse_2, 2)), font = 2, cex = cex)
      text(x = position, y = position-position/8, labels = paste("r:", round(cor_2, 2)), font = 2, cex = cex)

      rmse_d[n,] <- data.frame(var = n, RMSE_1= rmse_1, RMSE_2 =rmse_2)


    };par(mfrow = c(1,1))
    dev.off()

    rmse_d$RMSE_1 %>% mean(., na.rm = TRUE) # 0.6529529
    rmse_d$RMSE_2 %>% mean(., na.rm = TRUE) # 0.7583347
}
```
>> pmm good 


### PMM over different num_iterations ####
```{r}

if(run_test_imputes){

    method = "pmm"
    rmse_d <- data.frame(iteration = integer(), var= character(), RMSE_1 = numeric(), stringsAsFactors = FALSE)

    # loop over m iter and impute, loop over cols and calculate rmse with true
    if(FALSE){

      for ( iter_n in seq(5,30, 2)){
          pmm_iter_imp <- mice(test, method = method, m = iter_n)
          pmm_iter_imp_sum <- complete(pmm_iter_imp, action =1) ; for (i in 1:pmm_iter_imp$m){pmm_iter_imp_sum <-pmm_iter_imp_sum + complete(pmm_iter_imp, action =i )}
          pmm_iter_imp_mean <- pmm_iter_imp_sum / pmm_iter_imp$m
          
          pmm_iter_imp_mean <- as.matrix(pmm_iter_imp_mean)

          rmse_1 <- rmse( truth[inds_1d], pmm_iter_imp_mean[inds_1d])

          rmse_d <-rmse_d %>%
                rbind(., data.frame(iteration = i, RMSE_1= rmse_1))
      }

    saveRDS(rmse_d, paste0(wd_path, "processed_data/vaa_rmse_iter.rds"))

    }


    rmse_d <- readRDS(paste0(wd_path, "processed_data/vaa_rmse_iter.rds"))

    # ggplot(rmse_d , aes( iteration, RMSE_1, color = var))+
    #   geom_smooth(se = FALSE)+
    #   scale_x_continuous(breaks = seq(5, 30, 1))

    ggplot(rmse_d, aes( iteration, RMSE_1))+
      geom_smooth(se = FALSE)+
      # geom_smooth(se = FALSE, aes(color = var))+
      # geom_line(aes(color = var))+
      scale_x_continuous(breaks = seq(5, 30, 1))+
      ggtitle("RMSE across vars over different iteration numbers for VAA imputation")
    ggsave(paste0(wd_path, "results/VAA_analysis/imputation_testing/RMSE_iterations.png"), height = 10, width= 8)
}

```
>> ~19 iterations optimizes


### plot actual versus predicted for pmm at 16 iterations with categorized na test df
```{r}

if(run_test_imputes){
    h = 1500
    w = 1700
    print("pmm")
    pmm_at_iter <- mice(test, method = "pmm", m = 19, )
    pmm_at_iter_sum <- complete(pmm_at_iter, action =1)
    for (i in 1:pmm_at_iter$m){pmm_at_iter_sum <-pmm_at_iter_sum + complete(pmm_at_iter, action =i )}
    pmm_at_iter_mean <- pmm_at_iter_sum / pmm_at_iter$m

    rmse_d <- data.frame(variable= character(), RMSE = numeric(), stringsAsFactors = FALSE)
    cor_d <- data.frame(variable= character(), R = numeric(), p_value = numeric(), stringsAsFactors = FALSE)
    
    png(paste0(wd_path, "results/VAA_analysis/imputation_testing/PMM_at_iter.png"), height = h/2, width = w)

    par(mfrow = c(3,7))
    for (n in colnames(truth)){
      
      # find inds of missing for column n
      inds <- inds_2d[ inds_2d[,2] == which(colnames(truth) == n) ,]
      
      rmse <- rmse(truth[inds] %>% scale(), pmm_at_iter_mean[inds] %>% scale())
      corr_test <- cor.test(pmm_at_iter_mean[inds], truth[inds])
      corr <- as.numeric(corr_test$estimate)

      cex = 1
      position <-range(pmm_at_iter_mean[inds])[1] + (range(pmm_at_iter_mean[inds])[2] -range(pmm_at_iter_mean[inds]) [1])/ 2
      if(position == 0){position = .4}
      plot(pmm_at_iter_mean[inds]~ truth[inds], col = "blue", 
           main = sprintf("pmm, 19 iterations\nR=%f\nRMSE=%f\n%s", corr, rmse, n)
           )
      abline(c(0, 1), col = "blue") ;
      # text(x = position, y = position, labels = paste("RMSE:", round(rmse, 2)), font = 2, cex = cex)
      # text(x = position, y = position-position/8, labels = paste("r:", round(cor, 2)), font = 2, cex = cex)

      rmse_d[n,] <- data.frame(variable = n, RMSE = rmse)
      # make df of spectralis x avg correlation
      cor_d[n,] <- data.frame(variable = n, R=as.numeric(corr_test$estimate), p_value = corr_test$p.value)


    }
    par(mfrow = c(1,1))
    dev.off()
    
    # save correlations to cors folder
    if(!dir.exists(sprintf("%sprocessed_data/imputation_Rs/", wd_path))) dir.create(sprintf("%sprocessed_data/imputation_Rs/", wd_path))
    write_csv(cor_d, sprintf("%sprocessed_data/imputation_Rs/VAA_imputation_correlations.csv", wd_path))

}




```

#___
## Actual imputation
### TRUE to do VAA imputation on working d
```{r}
#
##
###
####
#####
impute <- TRUE
#####
####
###
##
#
```

### impute
```{r}

# row inds where less than ratio of columns are missing for each row
rows_to_impute <- rowSums(is.na(d_new[, VAA_vars])) < floor(length(VAA_vars)*.75) # 25% of columns must be non missing for each row

# data to feed into mice: has < ncols of NAs in each row and only VAA cols
d_new[, VAA_vars] <-  d_new[, VAA_vars] %>%
 mutate_all(~as.numeric(.))

if(impute){
  set.seed(19462)
  # impute!
  impute_input <- d_new[rows_to_impute, VAA_vars]
  vaa_imp <- mice(impute_input, method = "pmm", m = 19)
  saveRDS(vaa_imp, file = paste0(wd_path, "processed_data/vaa_columns_imputed_01.rds"))
}

vaa_imp <- readRDS(paste0(wd_path, "processed_data/vaa_columns_imputed_01.rds"))

vaa_imp_sum <- complete(vaa_imp, action =1) 
for (i in 1:vaa_imp$m){vaa_imp_sum <-vaa_imp_sum + complete(vaa_imp, action =i )} 
vaa_imp_mean <- vaa_imp_sum / vaa_imp$m

# replace missing with imputed in a new df
d_newest <- d_new
d_newest[rows_to_impute , VAA_vars ] <-  vaa_imp_mean

d_newest[,VAA_vars] %>% na.omit() %>% dim() - d_new[,VAA_vars]  %>% na.omit() %>% dim() # ~ 9000 more , 2x more! >> 5496 >> 6110

```


### calculate logmar from Acuity where acuity available or vice versa (depends on what mice throws out due to collinearity)
```{r}

acuit <- d_newest$ACUIT
logmar <- d_newest$LOGMAR
ACUIT_by_LOGMAR_mod <- lm(ACUIT ~ LOGMAR, data = d_newest)
ACUIT_by_LOGMAR_mod %>% summary()


plot( d_newest$ACUIT~d_newest$LOGMAR, type = "n")
abline(ACUIT_by_LOGMAR_mod, col = rgb(0,.4,1,.8), lwd = 4, lty = 3)
points(d_newest$ACUIT~d_newest$LOGMAR, col = ifelse(!is.na(d_newest$ACUIT) & is.na(d_new$ACUIT), "red", "black"))

# calculate missing Logmar where acuity
where_logmar_not_acuit <- is.na(d_newest$ACUIT) & !is.na(d_newest$LOGMAR)
d_newest$ACUIT[where_logmar_not_acuit] <- predict(ACUIT_by_LOGMAR_mod, data.frame(LOGMAR = d_newest$LOGMAR[where_logmar_not_acuit]))

```

### fix prescriptions that should be discrete
```{r}

# imputation made these variables more continuous so making them 1,0,or -1 
d_newest <- d_newest %>% 
  mutate(across(all_of(matches("PM")), 
                  ~round(.)
                )
  ) 

```


<!-- #### VAA na counts after imputation and re-calculation -->

<!-- ```{r} -->

<!-- vaa_count_df <- d_newest %>% -->
<!--   select( patID_eye, month, all_of(VAA_vars)) %>% -->
<!--   group_by(month) %>% -->
<!--   summarise(across(everything(), ~ sum(!is.na(.)), .names = "{.col}")) %>% -->
<!--   reshape2::melt(id.vars = "month") %>% -->
<!--   group_by(variable) %>% -->
<!--   mutate(var_total = sum(value)) %>% -->
<!--   ungroup() %>% -->
<!--   group_by(month) %>% -->
<!--   mutate(month_total = sum(value)) %>% -->
<!--   ungroup() -->

<!-- ## factor variable to have order of df -->
<!-- # order variables to go acuity scores then other -->
<!-- count_plot_levels <-  unique(vaa_count_df$variable[order(vaa_count_df$var_total)]) -->
<!-- last <- grep("ACUIT|LOGMAR|SNELL", count_plot_levels, value = TRUE) -->
<!-- count_plot_levels <- grep("ACUIT|LOGMAR|SNELL", count_plot_levels, value = TRUE, invert = TRUE) -->
<!-- count_plot_levels <- c(count_plot_levels, last) -->

<!-- vaa_count_df <- vaa_count_df %>% -->
<!--   mutate(variable = factor(variable, levels = count_plot_levels  ) )%>% -->
<!--   mutate(month_group = factor(ifelse(month > 0, "0<", "0"), levels  = c("0", "0<")) ) -->

<!-- # vAA overall counts -->
<!-- VAA_colsum_plot <-ggplot(vaa_count_df, aes(x = value ,y =variable , fill = variable))+ -->
<!--   geom_bar(stat = "identity", show.legend = FALSE, orientation = "y", width = .5, just = 0)+ -->
<!--   theme_light()+ -->
<!--   # theme(axis.text.x = element_text( hjust = 1, vjust = 0, size = 10, angle = 90, -->
<!--   #                                   margin = margin(t = -100, r = 0, b = 0, l = 0) -->
<!--   #                                   ) -->
<!--   #       )+ -->
<!--   geom_abline(intercept = nrow(d), col = "blue", linetype = 2, slope = 0)+ -->
<!--   ggtitle("VAA available counts") -->

<!-- ggsave(paste0(wd_path, "results/VAA_analysis/missingness//VAA_colsum_plot_postImputation.png"),VAA_colsum_plot, width = 15, height = 8, dpi = 150, limitsize = F) -->

<!-- ### same plot but faceted for visitID #### -->
<!-- VAA_colsum_facetVisit_plot <- ggplot(vaa_count_df, aes(y = variable ,x= value, fill = variable))+ -->
<!--   geom_bar(stat = "identity", show.legend = FALSE, orientation = "y", width = .5, just = 0)+ -->
<!--   theme_bw()+ -->
<!--   # theme(axis.text.x = element_text( hjust = 1, vjust = 0, size = 1.5, angle = 90, -->
<!--   #                                   margin = margin(t = 0, r = 0, b = 0, l = 0) -->
<!--   #                                   ) -->
<!--   #       )+ -->
<!--   geom_abline(intercept = nrow(d), col = "blue", linetype = 2, slope = 0)+ -->
<!--   facet_wrap(~month_group, scales = "fixed")+ -->
<!--   ggtitle("VAA available counts") -->
<!-- ggsave(paste0(wd_path, "results/VAA_analysis/missingness//VAA_colsum_facetVisit_plot_postImputation.png"),VAA_colsum_facetVisit_plot, width = 20, height = 15, dpi = 400, limitsize = F) -->
<!-- ``` -->

#### barplot of all columns after imputation and recalculation
```{r}

count_df <- d_newest %>%
  group_by(month) %>%
  summarise(across(everything(), ~ sum(!is.na(.)))
            ) %>% #, .names = "{.col}")) %>%
  reshape2::melt(id.vars = "month") %>%
  group_by(variable) %>%
  mutate(var_total = sum(value)) %>%
  ungroup() %>%
  group_by(month) %>%
  mutate(month_total = sum(value)) %>%
  ungroup()

## factor variable to have order of df
# order variables to go acuity scores then other
count_plot_levels <-  unique(count_df$variable[order(count_df$var_total)])
# last <- grep("ACUIT|LOGMAR|SNELL", count_plot_levels, value = TRUE)
# count_plot_levels <- grep("ACUIT|LOGMAR|SNELL", count_plot_levels, value = TRUE, invert = TRUE)
# count_plot_levels <- c(count_plot_levels, last)

count_df <- count_df %>%
  # mutate(variable = factor(variable, levels = count_plot_levels  ) )%>%
  mutate(month_group = factor(ifelse(month > 0, ">0", "0"), levels  = c("0", ">0")) )

# vAA overall counts
colsum_plot <-ggplot(count_df, aes(x = value ,y =variable , fill = variable))+
  geom_bar(stat = "identity", show.legend = FALSE, orientation = "y", width = .5, just = 0)+
  theme_light()+
  # theme(axis.text.x = element_text( hjust = 1, vjust = 0, size = 10, angle = 90,
  #                                   margin = margin(t = -100, r = 0, b = 0, l = 0)
  #                                   )
  #       )+
  geom_abline(intercept = nrow(d), col = "blue", linetype = 2, slope = 0)+
  ggtitle("All variables available counts")

ggsave(paste0(wd_path, "results/VAA_analysis/missingness//colsum_plot_postImputation.png"),colsum_plot, width = 25, height = 20, dpi = 300, limitsize = F)

### same plot but faceted for visitID ####
colsum_facetVisit_plot <- ggplot(count_df, aes(y = variable ,x= value, fill = variable))+
  geom_bar(stat = "identity", show.legend = FALSE, orientation = "y", width = .5, just = 0)+
  theme_bw()+
  # theme(axis.text.x = element_text( hjust = 1, vjust = 0, size = 1.5, angle = 90,
  #                                   margin = margin(t = 0, r = 0, b = 0, l = 0)
  #                                   )
  #       )+
  geom_abline(intercept = nrow(d), col = "blue", linetype = 2, slope = 0)+
  facet_wrap(~month_group, scales = "fixed")+
  ggtitle("All variables available counts")

ggsave(paste0(wd_path, "results/VAA_analysis/missingness//colsum_facetVisit_plot.png"), colsum_facetVisit_plot, width = 20, height = 15, dpi = 400, limitsize = F)
```

#____________________________________________
# factor analysis of VAA
### check for high correlations
```{r}

d_newest_all_available_VA <-d_newest %>%
  select(all_of(VAA_vars)) %>% 
  na.omit()


Heatmap(cor(d_newest_all_available_VA, use = "complete.obs"), 
            row_dend_width = unit(3, "cm"),
            column_dend_height = unit(3, "cm"), 
        name =  "r") 

m <-cor(d_newest_all_available_VA)
rownames(m) <- colnames(m)

# remove noise make strong relationships stronger and weak weaker
Heatmap((m^2)*sign(m))

### metrics
KMO(d_newest_all_available_VA) # MSA = .86
MSA <- KMO(d_newest_all_available_VA)$MSAi

# barplot of msas
msa_d <- data.frame(value = MSA) %>% 
  rownames_to_column("var") %>% 
  mutate(var = factor(var, levels = var[order(value)] ))

msa_cutoff <- .65
ggplot(msa_d, aes(value,var,  fill = var))+
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_vline(xintercept = msa_cutoff, linetype = 2, alpha = .5)+
  scale_x_continuous(breaks = c(msa_cutoff,seq(0, 100, .25)))+
  ggtitle("MSA scores for VAA variables")
# # * 0.00 to 0.49 unacceptable
# # * 0.50 to 0.59 miserable
# # * 0.60 to 0.69 mediocre
# # * 0.70 to 0.79 middling
# # * 0.80 to 0.89 meritorious
# # * 0.90 to 1.00 marvelous
  
# filter for MSA msa_cutoff
good_msa_vars <- names(MSA)[MSA>=msa_cutoff]
d_newest_all_available_VA_filt <- d_newest_all_available_VA[,good_msa_vars]
# KMO(d_newest_all_available_VA_filtered) # MSA = .73
# ^ overall and individual MSA better keeping those corelated factors

# screes of filtering for vars and not 
par(mfrow = c(2,1))
# scree(d_newest_all_available_VA_filt, main = "filtered")
# scree(d_newest_all_available_VA, main = "unfiltered")
fa.parallel(d_newest_all_available_VA_filt, fa="fa", main = "filtered")
fa.parallel(d_newest_all_available_VA, fa="fa", main = "unfiltered")
par(mfrow = c(1,1))

```
> filtering out low MSA (prescriptional variables) suggests that there are 3-4 pcs/factors that describe this data

### optimize FA
```{r}



# what rows have all not NA for columns having MSA> cutoff
rows_complete <- rowSums(is.na(d_newest[,good_msa_vars])) == 0

Nfacs <- 4

# run FA on filtered full va vars
oblimin <- fa(d_newest[rows_complete, good_msa_vars], Nfacs, rotate="oblimin" ) 
# ^ lots of warnings: not ideal but ok

# loadings of varas on factors
oblimin_loads <- oblimin$loadings

# look at how transform loadings for plotting
oblimin_loads_trnsfm <- oblimin_loads %>% 
  {(abs(.)^2.5) * sign(.)} 

par(mfrow = c(2,2))
for(i in 1:4){
  x <- data.frame(cbind(fac = oblimin_loads[,i, drop = T],fac_transformed= oblimin_loads_trnsfm[,i, drop = T]))
  x <- x[order(x[,1]) , ]
  # x <- scale(x)
  plot(x, type = "b")
  add_lines()
}
par(mfrow = c(1,1))

# plot fa diagram 
png(paste0(wd_path, "results/VAA_analysis/FA/FA_diagrams.png"), height = 1000, width = 1000)
par(mfrow = c(2,1))
plot(oblimin_loads[,c("MR1","MR2")],type="n", xlim = c(-1, 1), ylim = c(-1, 1), main = "oblimin"); text(oblimin_loads,labels=names(d_newest[rows_complete, names(MSA)[MSA>msa_cutoff]]),cex=.7); abline(h = c(0), v = c(0), lty = 3)
fa.diagram(oblimin, simple = TRUE)
par(mfrow = c(1,1), mar = rep(3,4))
dev.off()

```
>> mr better metrics than pa
>> with two factors, second factor overlaps with first, expected because oblimin allows correlations between factors
>> keeping teh correlated variables increases variance captured 
>> factors load negatively with Logmar and snellen, ∴ positive loading ∝ better vision

## run FA
```{r}

lamda <-function(x, lamda = NULL){
  if(is.null(lamda)){
    print("please enter lamda scaling factor")
  }else{
    (abs(x)^lamda) * sign(x)
  }
} 

n_factors <- 4
  
fit <- fa(d_newest[rows_complete, names(MSA)[MSA>msa_cutoff]], nfactors = n_factors, rotate="oblimin", residuals = FALSE)

# look at factor distributions
for(i in 1:ncol(fit$scores)){
  b <- fit$scores[,i]
  b <- scale(b)
  # b <- lamda(b, .9)
  
  if(i == 1){
    plot(density(b, na.rm = TRUE), col = rgb(.1,0,0,.4), ylim = c(0,3))
  }else{
    lines(density(b, na.rm = TRUE), col = rgb(.1,0,0,.4))
  }
}

# scale factors
factors <- fit$scores %>%
  scale() %>% 
  as.data.frame() 
# rename them
names(factors) <- names(factors) %>% 
  gsub("\\D+", "FA", .)

subject_factors <- cbind( d_newest[rows_complete , "sample"] ,factors)

d_newest[, grep("^FA\\d$", names(d_newest))] <- NULL
d_newest <- d_newest %>%
  left_join(., subject_factors)

```

### heatmap of loadings and corr with vars
```{r}

# loadings Heatmap
png(paste0(wd_path, "results/VAA_analysis/FA/", n_factors, "_FA_loadings_heatmap.png"), height = 1000, width = 1000)
Heatmap(oblimin_loads)
dev.off()


# look at what factor correlates to
d_newest_cor <-cor(d_newest[ , c(good_msa_vars, grep("FA\\d", names(d_newest), value = TRUE) )  ], use = "complete.obs") # corr on vaa cols
rownames(d_newest_cor) <- colnames(d_newest_cor)

png(paste0(wd_path, "results/VAA_analysis/FA/", n_factors, "_FA_corr_vars_heatmap.png"), height = 1000, width = 1000)
Heatmap(d_newest_cor, 
        row_dend_width = unit(6, "cm"),
        column_dend_height = unit(6, "cm"), 
        name = "r",
        row_names_gp = gpar(fontsize = 15),
        column_names_gp = gpar(fontsize = 15)
        ) 
dev.off()


```

# Remove prescriptional vars
```{r}

d_newest <-d_newest %>%
  select(-matches("CYL|SPH|AXIS"))

```


#|||||||||||||||||||||
# Save data
```{r}
# D
new_data_name <- data_name %>% 
  gsub("5", "6", .)
print(paste("Saving... ", new_data_name))
saveRDS(d_newest, paste0(wd_path, "processed_data/", new_data_name))

# # info
# new_info_path <- info_path %>% gsub("_04", "_05", .)
# 
# write.xlsx(info, new_info_path)


```

```{r}
check_dims(d_newest)
# dims : 12600 289
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516
# ^  no changes

### MARCH 17 2025 ###
# dims : 12619 310
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```

