---
title: "08_longitudinal_function"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---

# options
```{r}
# dev.off() # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()
```


# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}
pacman::p_load(tidyr, dplyr, reshape2, purrr, GGally, tidyverse, ggpubr, vcd, UpSetR, ggplot2, gridExtra, ComplexHeatmap, viridis, knitr, htmlwidgets, zoo, corrplot, RColorBrewer, openxlsx, plotly, RColorBrewer, mgcv, zoo, tools, mice, grDevices, ggrepel, networkD3, htmltools, grid, gplots, Metrics, psych, car, stats, lme4, lmerTest, VennDiagram, ape, patchwork)

rm(list = ls())
gc()

# myfunctions
source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```

# Load Data

```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}
wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

# joined cleaned joined data
data_name <- "CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_06.rds"
d_raw <- readRDS(paste0(wd_path, "processed_data/", data_name))# clean data with everything 
d <- d_raw

# grading nomenclature and information
info_path <- paste0(wd_path, "processed_data/database_grading_key_04.xlsx")
info <- read.xlsx(info_path)

```

# make new directorys for plots
```{r}

factors_over_time_path <- paste0(wd_path, "results/factors_over_time/")
if(!dir.exists(factors_over_time_path)) {
 dir.create(factors_over_time_path, recursive = TRUE)
 
}


```

# clean data

```{r, look at dims}
check_dims <- function(D = d){
 
 cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
 
 n_patIDs <- D %>% 
 select(patID, visitID, eye_num) %>% 
 drop_na() %>% 
 pull(patID ) %>% 
 n_distinct(na.rm = TRUE)
 
 n_patID_eyes <- D %>% 
 select(patID, visitID, eye_num) %>% 
 drop_na() %>% 
 select(patID, eye_num) %>% 
 n_distinct(na.rm = TRUE)
 
 n_observations <- D %>% 
 select(patID, visitID, eye_num) %>% 
 drop_na() %>% 
 n_distinct(na.rm = TRUE)
 
 n_patIDs_w_gr_1_vis <- D %>% 
 select(patID, visitID, eye_num) %>% 
 drop_na() %>% 
 filter(n()>1, .by = c(patID, eye_num)) %>% 
 pull(patID) %>% 
 n_distinct(na.rm = TRUE)
 
 print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
 
}

check_dims()
# dims : 12600 289
# [,1]
# n_patIDs 3693
# n_patID_eyes 7386
# n_observations 12600
# n_patIDs_w_gr_1_vis 516
# ^ no changes

### MARCH 17 2025 ###
# dims : 12619 310
# n_patIDs 4034
# n_patID_eyes 8063
# n_observations 12617
# n_patIDs_w_gr_1_vis 500

```

```{r}

cols_to_remove <- grep("DT|date|DATE", names(d), value = TRUE, ignore.case = TRUE) %>% 
 grep("HRDEXUDATE|DATE_coalesced", ., value = TRUE, invert = TRUE) 

cols_to_remove <- c(cols_to_remove, 
 grep("text|manref|qual|scan_type", names(d), ignore.case = TRUE, value = TRUE),
 grep("patientId", names(d), ignore.case = TRUE, value = TRUE)
 )
 

# remove date + other columns
d <- d %>% 
 select(-all_of(cols_to_remove))

```


```{r, look at dims}
check_dims()
# dims : 12600 261
# [,1]
# n_patIDs 3693
# n_patID_eyes 7386
# n_observations 12600
# n_patIDs_w_gr_1_vis 516
# ^ removed cols_to_remove

# dims : 12619 280
# n_patIDs 4034
# n_patID_eyes 8063
# n_observations 12617
# n_patIDs_w_gr_1_vis 500

```


# = = = = = = = = = = = = = = = = = = = = = = 

# plot each factor as function of time 

## categorize eyes by acuity quintile
```{r}

acu_quintiles <- quantile(d$ACUIT, probs = seq(0, 1, .2), na.rm = TRUE)
# findInterval(test_vec, acu_quintiles)
# test_vec <- c(-1, 1, 2, 50, 65, 75, 77, 80, 85, 90, 200)

d <- d %>% 
 arrange(month) %>% 
 group_by(patID_eye) %>% 
 mutate(ACU_quintile = findInterval(dplyr::first(ACUIT), acu_quintiles)) %>% 
 ungroup()

d <- d %>% 
 mutate(FA_mean = rowMeans(across(matches("FA\\d$"))))

FA_firsts <- d %>% 
 select(patID_eye,month, matches("FA\\d|FA_mean")) %>% 
 na.omit() %>% 
 arrange(patID_eye, month) %>% 
 group_by(patID_eye) %>% 
 # mutate(across(matches("FA"), ~dplyr::first(.))) %>%
 # unique() %>% 
 reframe(across(matches("FA"), ~dplyr::first(.))) 

 
FA_mean_quintiles <- quantile(FA_firsts$FA_mean, probs = seq(0, 1, .2), na.rm = TRUE)
FA1_quintiles <- quantile(FA_firsts$FA1, probs = seq(0, 1, .2), na.rm = TRUE)
FA2_quintiles <- quantile(FA_firsts$FA2, probs = seq(0, 1, .2), na.rm = TRUE)
FA3_quintiles <- quantile(FA_firsts$FA3, probs = seq(0, 1, .2), na.rm = TRUE)
FA4_quintiles <- quantile(FA_firsts$FA4, probs = seq(0, 1, .2), na.rm = TRUE)
# test_vec <- c(-5.35, - 5.3, -1, .2, .3,.51, 1)
# findInterval(test_vec, FA_mean_quintiles)

# categorize eyes by FA_mean
d <- d %>% 
 group_by(patID_eye) %>% 
 arrange(month) %>%
 mutate(
 FA_mean_quintile = dplyr::first(FA_mean) %>% 
 findInterval(., FA_mean_quintiles),
 FA1_quintile = dplyr::first(FA1) %>% 
 findInterval(., FA1_quintiles),
 FA2_quintile = dplyr::first(FA2) %>% 
 findInterval(., FA2_quintiles),
 FA3_quintile = dplyr::first(FA3) %>%
 findInterval(., FA3_quintiles),
 FA4_quintile = dplyr::first(FA4) %>%
 findInterval(., FA4_quintiles)
 ) %>% 
 ungroup() %>% 
 mutate(across(ends_with("quintile"), ~factor(., levels = 5:1 ) ))

```

## plot factors over time
```{r}

factor_data <- d %>% 
 select(patID_eye,month, FA_mean, matches("FA\\d|FA.*quintile")) %>% 
 na.omit() 
# factor_data %>% glimpse()

factor_data_delta <- factor_data %>% 
 arrange(patID_eye, month) %>% 
 group_by(patID_eye) %>% 
 mutate(across(all_of(c("FA1", "FA2","FA_mean", "FA3", "FA4", "FA_mean")), 
 ~ . - dplyr::first(.),
 .names = "{col}_delta000")
 )%>% 
 ungroup() 



factor_data_long <- factor_data_delta %>% 
 melt(measure.vars = c( grep("FA.*_delta000", names(factor_data_delta), value = TRUE)), id.vars = c("month", "patID_eye", grep("quintile", names(factor_data), value = TRUE)))

# filter out the mean 
# factor_data_long <- factor_data_long %>% 
# filter(variable != "FA_mean_delta000")

#
# # factor_over_time_facet_quintile_plot <-
# ggplot(factor_data_long , aes(month, value, color = patID_eye))+
# # geom_point(show.legend = FALSE, aes(color = patID_eye))+
# geom_point(show.legend = FALSE, alpha = .4)+
# geom_smooth(method = "lm", se = TRUE, show.legend = FALSE, aes(group = 1))+
# geom_line(show.legend = FALSE, alpha = .4, aes(group = 1))+
# # geom_smooth(method = "loess", se = TRUE, show.legend = TRUE, aes())+
# # facet_wrap(~MT5_quintile)+
# facet_wrap(~variable+FA_mean_quintile)+
# # facet_wrap(~ACU_quintile)+
# ggtitle(paste0(var, " over time, faceted for quintile"))+
# theme_bw()
#
#
# ggsave(paste0(wd_path, "results/VAA_analysis/factor_over_time_facet_quintile_plot.png"), factor_over_time_facet_quintile_plot, width = 10, height = 12, dpi = 100)


# a_var <- "FA_mean"
grobs <- list(); for(a_var in c("FA1", "FA2", "FA3", "FA4", "FA_mean")){ 

 # make plotting data with dummy names for looped variables
 plot_data <- factor_data_long
 ## make dummy column of quintile of interest
 var_quintile = paste0(a_var, "_quintile")
 plot_data$var_quintile <- factor_data_long[[var_quintile]]

 # filter for values of var_delta000
 var = paste0(a_var, "_delta000")
 plot_data <- plot_data %>%
 filter(variable == var)
 

 plot <- ggplot(plot_data , aes(month, value, color = patID_eye))+
 # geom_point(show.legend = FALSE, aes(color = patID_eye))+
 # geom_point(show.legend = FALSE, alpha = .4)+
 geom_smooth(method = "lm", se = TRUE, show.legend = FALSE, aes(group = 1))+
 # geom_line(show.legend = FALSE, alpha = .4, aes(group = 1))+
 # geom_smooth(method = "loess", se = TRUE, show.legend = TRUE, aes())+
 # facet_wrap(~MT5_quintile)+
 # facet_wrap(~var_quintile, ncol = 5, scales = "free")+
 facet_wrap(~var_quintile, ncol = 5, scales = "fixed")+
 # coord_cartesian(ylim = ylims)+
 # facet_wrap(~ACU_quintile)+
 # ggtitle(paste0(var, " over time, faceted for quintile"))+
 ggtitle(paste0(var, " over time, faceted for quintile of factor_n"))+
 theme_bw()

 grobs[[var_quintile]] <- ggplotGrob(plot)


}

factor_over_time_facet_quintile_plot_grid <- gridExtra::marrangeGrob(grobs, ncol = 1, nrow = n_distinct(factor_data_long$variable))
ggsave(paste0(factors_over_time_path,"factor_over_time_facet_quintile_plot_grid.png"), factor_over_time_facet_quintile_plot_grid, width = 10, height = 12, dpi = 100)






# for FA_mean quintile, all factor lines
factors_over_time <- ggplot(factor_data_long %>% filter(!variable == "FA_mean_delta000") , aes(month, value, color = variable))+
 geom_smooth(method = "lm", se = TRUE, show.legend = FALSE)+
 geom_point(aes(fill = variable), show.legend = TRUE, alpha = .2) +
 # geom_point(show.legend = FALSE, aes(color = patID_eye))+
 # geom_line(show.legend = FALSE, alpha = .4, show.legend = FALSE, aes(group = 1))+
 # geom_smooth(method = "loess", se = TRUE, show.legend = TRUE, aes())+
 # facet_wrap(~MT5_quintile)+
 facet_wrap(~FA_mean_quintile, ncol = 5)+
 # facet_wrap(~ACU_quintile)+
 ggtitle(paste0("Acuity Factors delta_000, faceted for FA_mean quintile"))+
 theme_bw()+
 geom_hline(yintercept = 0, linetype = 2)

ggsave(paste0( factors_over_time_path, "factors_over_time.png"), factors_over_time, width = 10, height = 10, dpi = 100)

# 
# ggplot(factor_data_long %>% filter(!variable == "FA_mean_delta000") , aes(month, value, color = variable))+
# geom_smooth(method = "lm", se = TRUE, show.legend = TRUE)+
# # geom_point(aes(color = patID_eye), show.legend = FALSE) +
# # geom_point(show.legend = FALSE, aes(color = patID_eye))+
# # geom_line(show.legend = FALSE, alpha = .4, show.legend = FALSE, aes(group = 1))+
# # geom_smooth(method = "loess", se = TRUE, show.legend = TRUE, aes())+
# # facet_wrap(~MT5_quintile)+
# facet_wrap(~variable+FA_mean_quintile, ncol = 5)+
# # facet_wrap(~ACU_quintile)+
# ggtitle(paste0("Acuity Factors delta_000, faceted for FA_mean quintile"))+
# theme_bw()



var_names <- unique(factor_data_long$variable)
colors = c("#FF585D", "#7BAE00", "#00BFC4" , "#C77CFF", "darkorange")
names(colors) <- var_names

height_y = 6

# make a plot faceted by mean quintile for each factor
plots <- list();for(var_name in var_names){
 
 a_plot_data <- factor_data_long %>% 
 filter(variable == var_name)
 
 a_color <- colors[[var_name]]
 
 a_plot <- ggplot(a_plot_data, aes(month, value)) +
 geom_smooth(method = "lm", alpha = .3, se = TRUE, show.legend = FALSE, color = a_color) +
 # geom_point(show.legend = FALSE, aes(color = patID_eye)) +
 # geom_line(show.legend = FALSE, aes(color = patID_eye), alpha = .4) +
 facet_wrap(~FA_mean_quintile, ncol = 5, scales = "fixed") +
 ggtitle(paste0(var_name, " faceted for FA_mean quintile")) +
 theme_bw()

 ggsave(paste0(factors_over_time_path, var_name,"_over_time_mean_quintile_facet.png"), a_plot, width = height_y * 3, height = height_y, dpi = 100)
 
 plots[[var_name]] <- a_plot
}


```



### plot factor values over time, not delta000s
```{r}

factor_data <- d %>% 
 select(patID_eye, month, FA_mean, matches("FA\\d|FA\\d$|FA_|FA.*quintile")) %>% 
 na.omit() 

# filter out patients who only appear once, only have visit 000
return_pats <- factor_data %>% 
 group_by(patID_eye) %>% 
 filter(n()>1) %>% 
 pull(patID_eye)

factor_data_long <- factor_data %>% 
 filter(patID_eye %in% return_pats) %>% 
 melt(measure.vars = c( grep("FA\\d$|FA_mean$", names(factor_data), value = TRUE)), 
 id.vars = c("month", "patID_eye", grep("quintile", names(factor_data), value = TRUE) )
 ) %>% 
 mutate(value = round(as.numeric(value), 3))

# # filter out the mean 
# factor_data_long <- factor_data_long %>% 
# filter(variable != "FA_mean")

# a_var <- "FA1"

colors = c(FA1 = "#FF8783", FA2 = "#5ADAE0", FA3 = "brown", FA4 = "darkgreen", FA_mean= "#16E746")
grobs <- list(); for(a_var in c("FA1", "FA2", "FA3", "FA4","FA_mean")){

 # make plotting data with dummy names for looped variables
 plot_data <- factor_data_long
 
 ## make dummy column of quintile of interest
 var_quintile = paste0(a_var, "_quintile")
 plot_data$var_quintile <- factor_data_long[[var_quintile]]

 # filter for values of var_delta000
 plot_data <- plot_data %>%
 filter(variable == a_var) 
 

 plot <- ggplot(plot_data , aes(month, value, color = factor(a_var)) ) +
 # geom_point(show.legend = FALSE, aes(color = patID_eye), alpha = .3)+
 # geom_point(show.legend = FALSE, alpha = .4)+
 geom_smooth(method = "lm", se = TRUE, show.legend = FALSE, aes(group = 1))+
 # geom_line(show.legend = FALSE, alpha = .4)+
 # geom_smooth(method = "loess", se = TRUE, show.legend = TRUE, aes())+
 # facet_wrap(~MT5_quintile)+
 # facet_wrap(~var_quintile, ncol = 5, scales = "free")+
 facet_wrap(~var_quintile, ncol = 5, scales = "fixed")+
 # coord_cartesian(ylim = ylims)+
 # facet_wrap(~ACU_quintile)+
 # ggtitle(paste0(var, " over time, faceted for quintile"))+
 ggtitle(paste0(a_var, " over time, faceted by first visit ", a_var, " quintile"))+
 theme_bw()+ 
 scale_color_manual(values = colors)+
 geom_hline(yintercept = 0)

 grobs[[var_quintile]] <- ggplotGrob(plot)


}

factor_over_time_facet_quintile_plot_grid_actual_vals <- gridExtra::marrangeGrob(grobs, ncol = 1, nrow = n_distinct(factor_data_long$variable))
ggsave(paste0(factors_over_time_path,"factor_over_time_facet_quintile_plot_grid_actual_vals.png"), factor_over_time_facet_quintile_plot_grid_actual_vals, width = 10, height = 12, dpi = 100)

```

## plot means of rowscores over time
```{r, eval = FALSE}

# select rowsscores
rowscore_d <- d %>% 
 select(patID_eye, month, matches("RW0\\d.*4$|RW1\\d.*4$")) %>% 
 mutate(
 all_letters = rowSums(across(matches("RW\\d.*4$")), na.rm = TRUE),
 tiny_letters = rowSums(across(matches("RW1[234]")), na.rm = TRUE),
 small_letters = rowSums(across(matches("RW1[01]|RW0[89]")), na.rm = TRUE),
 medium_letters = rowSums(across(matches("RW0[4567]")), na.rm = TRUE),
 large_letters = rowSums(across(matches("RW0[123]")) , na.rm = TRUE)
 ) %>% 
 # mutate(
 # all_letters = rowMeans(across(matches("RW\\d.*4$")), na.rm = TRUE),
 # tiny_letters = rowMeans(across(matches("RW1[234]")), na.rm = TRUE),
 # small_letters = rowMeans(across(matches("RW1[01]|RW0[89]")), na.rm = TRUE),
 # medium_letters = rowMeans(across(matches("RW0[4567]")), na.rm = TRUE),
 # large_letters = rowMeans(across(matches("RW0[123]")) , na.rm = TRUE)
 # ) %>%
 # mutate(across(matches("letters"), ~as.vector(scale(.))))
 select(-matches("RW"))

# quintile
rowscore_d <- rowscore_d %>% 
 na.omit() %>% 
 mutate(all_letters_quintile = cut(all_letters, breaks = quantile(all_letters, seq(0,1,.2), na.rm = TRUE), labels = 1:5))
 # mutate(across(matches("tiny|small|medium|large"),
 # ~cut(., breaks = quantile(., seq(0,1,.2), na.rm = TRUE), labels = 1:5),
 # .names = "{.col}_quintile"
 # )
 # )


# filter out patients who only appear once, only have visit 000
return_pats <- rowscore_d %>% 
 group_by(patID_eye) %>% 
 filter(n()>1) %>% 
 pull(patID_eye)

rowscore_d <- rowscore_d %>% 
 filter(patID_eye %in% return_pats)

# rowscore_d %>% 
# melt(id.vars = c('month', 'patID_eye', 'all_letters_quintile')) %>% 
# ggplot(aes(month, value, color = factor(variable)) ) +
# geom_smooth(method = "lm", se = TRUE, show.legend = FALSE)+
# facet_wrap(~all_letters_quintile)+
# theme_bw()


grobs <- list(); for(a_var in c("FA1", "FA2", "FA_mean", "FA3", "FA4")){

 # make plotting data with dummy names for looped variables
 plot_data <- rowscore_d_long
 
 ## make dummy column of quintile of interest
 var_quintile = paste0(a_var, "_quintile")
 plot_data$var_quintile <- rowscore_d_long[[var_quintile]]

 # filter for values of var_delta000
 plot_data <- plot_data %>%
 filter(variable == a_var) 
 

 plot <- ggplot(plot_data , aes(month, value, color = factor(a_var)) ) +
 # geom_point(show.legend = FALSE, aes(color = patID_eye), alpha = .3)+
 # geom_point(show.legend = FALSE, alpha = .4)+
 geom_smooth(method = "lm", se = TRUE, show.legend = FALSE, aes(group = 1))+
 # geom_line(show.legend = FALSE, alpha = .4)+
 # geom_smooth(method = "loess", se = TRUE, show.legend = TRUE, aes())+
 # facet_wrap(~MT5_quintile)+
 # facet_wrap(~var_quintile, ncol = 5, scales = "free")+
 facet_wrap(~var_quintile, ncol = 5, scales = "fixed")+
 # coord_cartesian(ylim = ylims)+
 # facet_wrap(~ACU_quintile)+
 # ggtitle(paste0(var, " over time, faceted for quintile"))+
 ggtitle(paste0(a_var, " over time, faceted by first visit ", a_var, " quintile"))+
 theme_bw()+ 
 scale_color_manual(values = colors)+
 geom_hline(yintercept = 0)

 grobs[[var_quintile]] <- ggplotGrob(plot)
}

```



# __________________
# calculate some variables
## vision loss rate
```{r}

# rate df. calculating rate by excluding NAs so that last() visits that do not have data are not used in calculation, i.e. latest visits with data used instead
FA_rate_df <- d %>% 
 select(month, patID_eye, FA1:FA2 , FA_mean) %>% 
 na.omit() %>% 
 group_by(patID_eye) %>% 
 arrange(patID_eye, month) %>%
 mutate(duration = dplyr::last(month) - dplyr::first(month) ) %>% 
 mutate(across(all_of(starts_with("FA")),
 ~ ifelse(duration > 0, 
 (dplyr::last(.) - dplyr::first(.)) / duration, 
 NA),
 .names = "{.col}_rate"
 )
 ) %>% 
 select(-duration) %>% 
 ungroup()

# join rates to d
d <- d %>% 
 left_join(., FA_rate_df)


```



## parafoveal rate
```{r}

d <- d %>% 
  mutate(parafovea = rowMeans(across(matches("avg_MT0[2345]")), na.rm = TRUE))

# calculate rate change of average mt2-5 thicknesses, excluding visits missing value
parafoveal_rate_d <- d %>% 
 select(parafovea, patID_eye, month) %>% 
 arrange(month) %>% 
 group_by(patID_eye) %>% 
 na.omit() %>% # remove nas so that last values that are missing are not used 
 mutate(duration = dplyr::last(month) - dplyr::first(month) ) %>% 
 mutate(parafovea_rate = (dplyr::last(parafovea) - dplyr::first(parafovea) ) / duration) %>% 
 select(-duration) %>% 
 ungroup() 

# ___from 06_____
# rate_d <- d %>% 
# select(month, patID_eye, spectralis_MT01:spectralis_MT09) %>% 
# group_by(patID_eye) %>% 
# arrange(month) %>% 
# # mutate(duration = dplyr::last(month[!is.na(month)]) - dplyr::first(month[!is.na(month)]) ) %>% 
# # mutate(across(starts_with("spectralis"), 
# # ~ (dplyr::last(.) - dplyr::first(.)) / duration, 
# # .names = "{str_sub(.col, -4)}_rate")
# # )%>% 
# mutate(
# across(
# starts_with("spectralis"),
# ~ {
# valid_data <- .[!is.na(.) & !is.na(month)]
# valid_months <- month[!is.na(.) & !is.na(month)]
# if (length(valid_data) > 1) {
# (dplyr::last(valid_data) - dplyr::first(valid_data)) / (dplyr::last(valid_months) - dplyr::first(valid_months))
# } else {
# NA_real_
# }
# },
# .names = "{str_sub(.col, -4)}_rate"
# )
# ) %>%
# ungroup() 
# ________

# join these rates to d
d <- d %>% 
 left_join(., parafoveal_rate_d)

```

<!-- ## MT slope rate -->
<!-- ```{r} -->

<!-- # ### test----- -->
<!-- # slope_rate_test <- d %>%  -->
<!-- # select( patID_eye, month , ends_with("slope")) %>%  -->
<!-- # mutate(avg_MT_slope_sans3 = rowMeans(across(matches("^avg_MT0[245]_slope")))) %>% # avg parafovea except 3 because that takes a long time to be affected -->
<!-- # arrange(month) %>%  -->
<!-- # group_by(patID_eye) %>%  -->
<!-- # na.omit() %>%  -->
<!-- # mutate(duration = dplyr::last(month) - dplyr::first(month) ) %>%  -->
<!-- # mutate(across(ends_with("slope"),  -->
<!-- # ~(dplyr::last(.) - dplyr::first(.) ) / duration,  -->
<!-- # .names = "{substr(.col, nchar(.col) -9, nchar(.col))}_rate" -->
<!-- # ) -->
<!-- # ) %>%  -->
<!-- # mutate(avg_MT_slope_sans3_rate = (dplyr::last(avg_MT_slope_sans3) - dplyr::first(avg_MT_slope_sans3) ) / duration) %>%  -->
<!-- # select(-duration) %>%  -->
<!-- # ungroup() %>%  -->
<!-- # arrange(patID_eye) %>%  -->
<!-- # select(patID_eye, month, avg_MT_slope_sans3_rate,avg_MT_slope_sans3, matches("^avg_MT0[245]_slope")) -->
<!-- #  -->
<!-- # ### ---- -->

<!-- # make average of 2,4,5 zones slope; calculate rate change of average mt2-5 thicknesses, excluding visits missing value -->
<!-- slope_d <- d %>%  -->
<!--  select( patID_eye, month , ends_with("slope")) %>%  -->
<!--  mutate(avg_MT_slope_sans3 = rowMeans(across(matches("^avg_MT0[245]_slope")))) %>% # avg parafovea except 3 because that takes a long time to be affected -->
<!--  arrange(month) %>%  -->
<!--  group_by(patID_eye) %>%  -->
<!--  na.omit() %>%  -->
<!--  mutate(duration = dplyr::last(month) - dplyr::first(month) ) %>%  -->
<!--  mutate(across(ends_with("slope"),  -->
<!--  ~(dplyr::last(.) - dplyr::first(.) ) / duration,  -->
<!--  .names = "{substr(.col, nchar(.col) -9, nchar(.col))}_rate" -->
<!--  ) -->
<!--  ) %>%  -->
<!--  mutate(avg_MT_slope_sans3_rate = (dplyr::last(avg_MT_slope_sans3) - dplyr::first(avg_MT_slope_sans3) ) / duration) %>%  -->
<!--  select(-duration) %>%  -->
<!--  ungroup() %>%  -->
<!--  arrange(patID_eye) -->


<!-- # join these rates to d -->
<!-- d <- d %>%  -->
<!--  left_join(., slope_d) -->

<!-- ``` -->

<!-- #### plot avg_MT_slope_sans3 quintile plots -->
<!-- ```{r} -->

<!-- avg_MT_slope_sans3_quintiles <- d %>%  -->
<!--  arrange(month) %>%  -->
<!--  group_by(patID_eye) %>%  -->
<!--  filter(month == dplyr::first(month)) %>%  -->
<!--  pull(avg_MT_slope_sans3) %>%  -->
<!--  quantile(., -->
<!--  seq(0,1,.2),  -->
<!--  na.rm = TRUE -->
<!--  ) -->

<!-- quintiles <- d %>%  -->
<!--  select(patID_eye, month, avg_MT_slope_sans3) %>%  -->
<!--  arrange(month) %>%  -->
<!--  group_by(patID_eye) %>%  -->
<!--  filter(month == dplyr::first(month)) %>%  -->
<!--  mutate(avg_MT_slope_sans3_quintile =  -->
<!--  cut(avg_MT_slope_sans3,  -->
<!--  avg_MT_slope_sans3_quintiles,  -->
<!--  labels = 1:5,  -->
<!--  include.lowest = TRUE) -->
<!--  ) %>%  -->
<!--  mutate(avg_MT_slope_sans3_quintile = avg_MT_slope_sans3_quintile %>%  -->
<!--  factor(., levels = as.character(5:1)) -->
<!--  ) %>%  -->
<!--  select(-month, -avg_MT_slope_sans3) %>%  -->
<!--  unique() -->

<!-- slope_quintiles_d <- d %>%  -->
<!--  select(patID_eye, month, patID, avg_MT_slope_sans3) %>%  -->
<!--  left_join(quintiles) %>%  -->
<!--  group_by(patID_eye) %>%  -->
<!--  mutate(slope_delta0 =  -->
<!--  avg_MT_slope_sans3 - dplyr::first(avg_MT_slope_sans3) -->
<!--  ) %>%  -->
<!--  ungroup() -->


<!-- # slope difference from baseline over time -->
<!-- slope_delta_0 <- ggplot(slope_quintiles_d %>% na.omit(), aes(month, slope_delta0))+ -->
<!--  # geom_smooth(method = "lm") + -->
<!--  geom_smooth(method = "gam",formula = y~x +s(x, k = 3, bs = 'cs')) + -->
<!--  facet_wrap(~avg_MT_slope_sans3_quintile, ncol = 5)+ -->
<!--  geom_hline(yintercept = 0, linetype =2)+ -->
<!--  theme_bw()+ -->
<!--  ggtitle("Average of parafoveal slopes (excl. 3) difference from baseline over time") -->

<!-- # slope over time -->
<!-- slope <- ggplot(slope_quintiles_d %>% na.omit(), aes(month, avg_MT_slope_sans3))+ -->
<!--  # geom_smooth(method = "lm") + -->
<!--  geom_smooth(method = "gam",formula = y~x +s(x, k = 3, bs = 'cs')) + -->
<!--  facet_wrap(~avg_MT_slope_sans3_quintile, ncol = 5)+ -->
<!--  geom_hline(yintercept = 0, linetype =2)+ -->
<!--  theme_bw()+ -->
<!--  ggtitle("Average of parafoveal slopes (excl. 3) over time") -->

<!-- ggsave(paste0(wd_path, "results/longitudinal_analysis/slope_delta_0_and_raw_values_over_time.png"), slope_delta_0/slope ,height = 10, width = 20) -->

<!-- ``` -->


## check rates for infinities
```{r}
# check
map(d, ~{
 any(is.infinite(.))
}) %>% 
 as.logical() %>% 
 any()


```

<!-- ## calculate spherical equivent -->
<!-- * sphere power + .5(cyliner power) -->
<!-- ```{r} -->

<!-- d$SPH -->
<!-- # make spherical equivalent variable -->
<!-- d <- d %>%  -->
<!--  mutate(SPHEQ = SPHNUM + (CYLNUM / 2)) -->

<!-- # make variable for myopic: myopic if SPHNUM < 0 -->
<!-- d <- d %>%  -->
<!--  mutate(myopic = ifelse(SPHNUM<0, 1,0)) -->


<!-- # new info -->
<!-- new_row <- data.frame(name  = "SPHEQ", -->
<!--                       matched_name = NA, -->
<!--                       dist  = NA, -->
<!--                       table = "VAA", -->
<!--                       table_description = "Visual Acuity Assesment", -->
<!--                       description = "Spherical equivalent (SPHNUM + (CYLNUM/2)", -->
<!--                       shorthand  = "SPHEQ", -->
<!--                       `0` = NA, -->
<!--                       `1` = NA, -->
<!--                       `2` = NA, -->
<!--                       `3` = NA, -->
<!--                       `4` = NA, -->
<!--                       `5` = NA, -->
<!--                       `6` = NA, -->
<!--                       continuous = T, -->
<!--                       nominal = F, -->
<!--                       zonal = F, -->
<!--                       can_resolve = T, -->
<!--                       ordinal = F -->
<!--                       ) -->
<!-- names(new_row) <- str_remove_all(names(new_row), "X") -->

<!-- # add this to info -->
<!-- info <- info %>%  -->
<!--  rbind(.,new_row) -->
<!-- ``` -->

## sum zonal vars across zones
```{r}

# what phenos to sum
zonal_pheno_names <- grep("0\\d$", names(d), value = TRUE) %>% 
 grep("spectralis|MT0|RW", . , value = TRUE, invert = TRUE) %>% 
 gsub("0\\d",'0', . ) %>% 
 unique()

zonal_pheno <- "IR0"

# for each pheno name, sum the columns in model_df_first_filtered
for (zonal_pheno in zonal_pheno_names){
 # grep(zonal_pheno, names(d), value = TRUE) %>% print()

 new_name <- paste0(zonal_pheno, "_sum")
 d[[new_name]] <- d[, grep(zonal_pheno, names(d), value = TRUE)] %>%
 rowSums(.)
}
 
 
```

## average continuous variables
```{r}

names(d) %>% 
 grep("0\\d", ., value = TRUE)
# only stratus_mt and stratus_mt_slope continuouos zonal variable


# mutate(avg_MT_slope_mean = rowMeans(across(matches("MT0\\d_slope$")))) 

```

#|||||||||||||||||||||

# Save data
```{r}
# D
new_data_name <- data_name %>% 
 gsub("6", "7", .)
print(paste("Saving... ", new_data_name))

saveRDS(d, paste0(wd_path, "processed_data/", new_data_name))

```

```{r, look at dims}
check_dims()
# dims : 12600 289
# [,1]
# n_patIDs 3693
# n_patID_eyes 7386
# n_observations 12600
# n_patIDs_w_gr_1_vis 516
# ^ added FA stats, slope stuff, rates

### MARCH 17 2025 ###
# dims : 12619 312
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```
