---
title: "09_clustering"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---
# options
```{r}
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()



```

# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

pacman::p_load(tidyr,dplyr,reshape2,purrr,GGally,tidyverse,ggpubr,vcd,ggplot2,tidytext,gridExtra,ComplexHeatmap,zoo,mgcv,car,zoo,tools,mice,grDevices,ggrepel,grid,gplots,Metrics,stats,lme4,lmerTest,dendextend,magick,umap,scater,SingleCellExperiment,scran,bluster,slingshot,ggfortify,pheatmap,TSCAN,igraph,knitr, HDF5Array, parallel, openxlsx, qs)



rm(list = ls())
gc()
```

# functions
```{r}

source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```


# Load Data

```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

# raw list of dfs
# list.files(sprintf("%sprocessed_data/",wd_path)) %>% 
#   grep("dfs_list",.,value = TRUE)
raw <- readRDS(paste0(wd_path, "processed_data/dfs_list_01.rds"))

# joined cleaned joined data
data_name <- "CF_FFA_DEM_DIAGNOSIS_ENRNHOR_ENRNHOS_FAFGRADE_FAMSTATUS_MH_OCTMAIN_OCTMTCIRRUS_OCTMTSPECTRALIS_OCTMTSTRATUS_VAA_lea_07.rds"
d_raw <- readRDS(paste0(wd_path, "processed_data/", data_name))# clean data with everything 
d <- d_raw

# grading nomenclature and information
info_path <- paste0(wd_path, "processed_data/database_grading_key_04.xlsx")
info <-  read.xlsx(info_path)

```

## load Robbie's and Yue's factors
```{r}


# Robbies linear factors
factor_root = dirname(wd_path)
paths = c(
    sprintf("%s/raw_data/Linear_Factors/data_cleaned_sce_Microsoft_factorised.RData", factor_root),
    sprintf("%s/raw_data/Linear_Factors/data_cleaned_sce_Yue_factorised.RData", factor_root)
)

load(paths[[1]], verbose = TRUE)
microsoft_sce <- sce
sce <- NULL
microsoft_coldata <- colData(microsoft_sce) %>% 
  as.data.frame() %>% 
  select(patID = universal_patient_id, eye, chew_grade =Chew_grading, MacTel)

# "universal_scan_id"  "sampleID"           "MacTel_python"       "universal_patient_id" "universal_patient_id_old"
# "patientId"          "s3originpath"       "studydate"           "studytype"            "seriestype"
# "eye"                "modality"           "seriesclean"         "dicomIdentifier"      "name"
# "Patient.ID.yue"     "dicom_file_name"    "MacTel.Diagnosis"    "Patient.s.SEXF"        "Data.Shape"
# "Pixel.Spacing"      "data.dim"           "dataset_name"        "instance_path"        "patient_no"
# "ptid_json"          "patient_id"         "instance_json"       "instance_id"          "image.counts"
# "Chew_grading"       "MacTel"


load(paths[2])
yue_sce <- sce
sce <- NULL
yue_coldata <- colData(yue_sce) %>% 
  as.data.frame() %>% 
  select(patID = macID, NHOR, eye, chew_grade = emily_grade) %>% 
  mutate(visitID = sapply(strsplit(NHOR, "_"), `[`, 2)
         ) %>% 
  select( -NHOR)
# "site.ID" "macID""visitType.Code""visitTypeText""date.of.scan" NHOR""eye" "colourID" "visitID" "photoDate" "gradingDate" "graderID" "emily_grade""MacTel_visit_ID" "sampleID"

both <- full_join(microsoft_coldata, yue_coldata) %>% 
  mutate(eye = case_match(eye, 
    "R" ~ 1, 
    "L" ~ 2
  )) %>% 
  mutate(patID_eye = paste(patID, eye, sep = "_"))

both %>% 
  group_by(patID_eye, visitID) %>% 
  reframe(n_distinct_chew = n_distinct(chew_grade)) %>% 
  pull(n_distinct_chew) %>% 
  table()
# ^ no duplicates

both_sub <- both %>% 
  select(patID_eye, visitID, chew_grade) %>%
  na.omit() %>% 
  unique()

# join this to main d
d <- 
  d %>% 
  left_join( both_sub, 
             by = c("patID_eye", "visitID"), 
             relationship = "one-to-one") 


```


# make new directorys for plots
```{r}

a_path <- paste0(wd_path, "results/clustering/09_clustering")
if(!dir.exists(a_path)) {
  dir.create(a_path, recursive = TRUE)
  
}


```

# clean data


```{r, look at dims}
check_dims <- function(D = d){
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()
# dims : 12600 290
#                      [,1]
# n_patIDs             3693
# n_patID_eyes         7386
# n_observations      12600
# n_patIDs_w_gr_1_vis   516
# ^ added chew grade

### MARCH 17 2025 ###
# dims : 12619 313
# n_patIDs             4034
# n_patID_eyes         8063
# n_observations      12617
# n_patIDs_w_gr_1_vis   500

```

## round month, make sample column
```{r}
# round month 
d$month <- round(d$month, 2)

# make sample
d <- d %>% 
    mutate(sample = paste(patID_eye, month, sep = "_"))

```


# update info 
## with new vars
```{r}

# ****** FUnction to query name in info ******
var_summary <- function(names_vec = character()){
  patt <- paste(names_vec, collapse = "|")  
  m <- as.matrix(filter(info, grepl(patt, name, ignore.case = TRUE)))
  rownames(m) <- m[,1]
  print(kable(t(m), "simple"))
  for(nm in m[,'name']){
    cat(sprintf('\n%s\n', nm))
    print(table(d[[nm]]))
  }
}


# data.frame(new_var = setdiff(names(d), info$name), 
#                                 dataset  = NA) %>% print

# dput 
new_vars_datasets <- structure(list(
  name = c("DIAGNOSIS_binary", "DATE_coalesced", "sample", "ISOS_LOC_JF", "ISOS_LOC_CEN", "confounding_disease", "OTHER_DISEASE_PRESENT", "SEXF", "month", "MT2_quartile", "MT3_quartile", "MT4_quartile", "MT5_quartile", "MT01_rate", "MT02_rate","MT03_rate", "MT04_rate", "MT05_rate", "MT06_rate", "MT07_rate", "MT08_rate", "MT09_rate", "FA1", "FA3", "FA4", "FA2", "ACU_quintile", "FA_mean", "FA_mean_quintile", "FA1_quintile", "FA2_quintile", "FA3_quintile", "FA4_quintile", "FA1_rate", "FA3_rate", "FA4_rate", "FA2_rate", "FA_mean_rate", "parafovea", "parafovea_rate", "avg_MT_slope_sans3", "MT02_slope_rate", "MT03_slope_rate", "MT04_slope_rate", "MT05_slope_rate", "avg_MT_slope_sans3_rate", "SPHEQ", "myopic", "IR0_sum", "BRV0_sum", "DRV0_sum", "ICL0_sum", "LRT0_sum", "RAV0_sum", "VTV0_sum", "RPE_LK0_sum", "avg_MT_mean", "avg_MT_slope_mean", "chew_grade"), 
  table = c("MH/DIAGNOSIS", "DEM/ENR/FAM/other", "DEM/ENR/FAM/other", "OCTMAIN", "OCTMAIN", "CF_FFA/FAF", "OCTMAIN", "DEM/ENR/FAM/other", "DEM/ENR/FAM/other", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "OCTMT","OCTMT", "OCTMT", "OCTMT", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "VAA", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "OCTMT", "VAA", "VAA", "CF_FFA", "CF_FFA", "CF_FFA", "CF_FFA", "CF_FFA", "CF_FFA", "CF_FFA", "CF_FFA", "OCTMT", "OCTMT", "MH/DIAGNOSIS")), 
  row.names = c(NA, -59L), class = "data.frame")


info <- full_join(info, new_vars_datasets)


setdiff(names(d), info$name)
setdiff(info$name, names(d))

# remove rows of variables not in d
info <- info[info$name %in% names(d) , ]

```

# = = = = = = = = = = = = = = = = 

# Filtering data for SCE

## subset data for different 'assays'

### what vars
exclude:
  * factors and prescriptional vars BUT inlcude other VA
  * engineered vars EXCEPT slope :> (sums, means, etc.)
  * extremely missing vars :> ICL (later step)
  * redundant vars :> stratus, cirrus, spectralis
  * demographics
```{r}

str_vars <- names(d) %>% 
  grep(x = ., 
"I[dD]|sample|num|GENDER|ETHN|DIAG|SMK|SEXF|YR|RACE|SPH|CYL|AXIS|ACUIT|SNELL|LOGMAR|RW.*1$|VAA|FA\\d|FA_|chew|myopic|VIS|visit|_DATE|ENRDATE|coalesced|SITE|month|date|duration|OTHER|confounding|SIGSTR|rescaled_to_spectralis|stratus|cirrus|^spectralis|stratus|cirrus|spectralis|CHA[NG]{2}ES|ABNORM|sum|mean|slope|parafovea|variance|rate|RELATPROBND|NHO[SR]|DM|WGT|HGT|eye|duplicated|quintile|quartile|FAMILY", 
    value = TRUE, invert = TRUE)
  
setdiff(names(d), str_vars) %>% sort()

```

### subset d for non-engineered measures vars
```{r}
#---
# subset d for vars of interest and scale them
#---
d <- d %>% 
  mutate(across(all_of(str_vars), ~ as.numeric(.)))

## subset for measure vars
measure_vars <- d %>% 
  select(patID_eye, month, sample, all_of(str_vars)) %>% 
  as.data.frame()

rownames(measure_vars) <- measure_vars$sample
```

#### remove outlying missy variables
```{r}

(colSums(is.na(measure_vars))/nrow(measure_vars)) %>%
  summary()
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 0.00000 0.01379 0.02544 0.08058 0.09446 0.87986 

#---
# filter outliers and missy variables
#---

# filter out variables with > min_out ratio missing
# min ratio of missingness outliers for varaiables
var_min_out <- boxplot(colSums(is.na(measure_vars))/nrow(measure_vars), main = "variable missingess")$out %>% min()
# ^ minimum outlier of missingness  =  0.2462953 >> 24% missing

hist(colSums(is.na(measure_vars))/nrow(measure_vars), breaks = seq(0, 1, .01), main = "Variable Missingness")
axis(side = 1, at = seq(0, 1, 0.1))

out_vars <-  names(measure_vars)[colSums(is.na(measure_vars))/nrow(measure_vars) >= 0.2 ]


#---
# remove vars who are skewed
#---
skw <- measure_vars %>% 
  select_if(~is.numeric(.)) %>% 
  sapply(psych::skew) %>%
  abs() 
# sort(skw)

# look at how vars skewed
hist(skw, breaks = seq(0, 200, 1), main = "Variable Skews")
axis(side = 1, at = seq(0, 200, 10))

# use threshold of 20
skewed_vars <- skw[skw>30] %>% names()
# "IR06"             "IR07"             "IR08"             "IR09"             "BRV06"           
#  "BRV07"            "BRV08"            "BRV09"            "ICL01"            "ICL04"           
#  "ICL06"            "ICL09"            "LRT06"            "LRT07"            "LRT08"           
#  "LRT09"            "VTV07"            "VTV09"            "RPE_LK06"         "RPE_LK07"        
#  "RPE_LK08"         "CENTERMIN_MICRON"
#  ^ these vars are rare but do occur in pop

# look at values
sapply(skewed_vars, function(x){
  measure_vars[[x]] %>% table
})


# add combine vars that are skewed or too missy
out_vars <- unique(c(out_vars, skewed_vars))

#---
# remove vars >6
#---
zones_gr_6_vars <- colnames(measure_vars) %>%
  grep("\\D0[789]$", ., value = TRUE) %>%
  grep("avg", invert = TRUE, ., value = TRUE)
# ^ "IR07"     "IR08"     "IR09"     "BRV07"    "BRV08"    "BRV09"    "ICL08"    "ICL09"    "LRT07"    "LRT08"   
#   "LRT09"    "VTV07"    "VTV09"    "RPE_LK07" "RPE_LK08" "RPE_LK09"

# add vars of zone > 6
out_vars <- c(out_vars, zones_gr_6_vars)

# remove variables
measure_vars <- measure_vars %>% 
  select(-all_of(out_vars))
# 1] "FAZ"              "TYP_LEAK_RPE"     "EZLOSS"           "CENTER_MICRON"   
#  [5] "CENTERMAX_MICRON" "CENTERMIN_MICRON" "BLRPATTERN"       "IR06"            
#  [9] "IR07"             "IR08"             "IR09"             "BRV06"           
# [13] "BRV07"            "BRV08"            "ICL09"            "LRT06"           
# [17] "LRT07"            "LRT08"            "LRT09"            "VTV07"           
# [21] "VTV09"            "IR07"             "IR08"             "IR09"            
# [25] "BRV07"            "BRV08"            "BRV09"            "ICL08"           
# [29] "ICL09"            "LRT07"            "LRT08"            "LRT09"           
# [33] "VTV07"            "VTV09"            "RPE_LK07"         "RPE_LK08"        
# [37] "RPE_LK09"        

```

#### remove outlying missy samples
```{r}
# filter for eye_visits with < min_out ratio missing
## minimum ratio of missing among missing outliers
samp_min_out <- boxplot(rowSums(is.na(measure_vars))/ncol(measure_vars), main = "sample missinness across features")$out %>% min()
# ^ minimum outlier of missingness  = missing 25% >> 27%>>0.3367347

hist(rowSums(is.na(measure_vars))/ncol(measure_vars), breaks = seq(0, 1, .01), main = " Missingness")
axis(side = 1, at = seq(0, 1, 0.1))

table(rowSums(is.na(measure_vars))/ncol(measure_vars) >= samp_min_out)
# ^ FALSE  TRUE 
# 12401   218 

## what patients are missingness outliers
out_samples <- measure_vars$sample[rowSums(is.na(measure_vars))/ncol(measure_vars) >= samp_min_out]

# look at what data is missing for these samples
ratio_missing <- colSums(is.na(measure_vars[measure_vars$sample%in% out_samples , ]))/nrow(measure_vars[measure_vars$sample%in% out_samples , ]) 
ratio_missing %>% 
  sort() %>% 
  barplot(las = 2)
# ^ mostly CFFFA vars

# deselect outlier samples and variables from assay
measure_vars <- measure_vars[ !rownames(measure_vars) %in% out_samples , ]

```


### filter out id cols and scale
```{r}

# make measure vars numeric matrix
measure_vars <- measure_vars[, -which(colnames(measure_vars) %in% c("patID_eye", "month", "sample"))]
measure_vars <- data.matrix(measure_vars)

# scale measure vars
scaled_measure_vars <- scale(measure_vars)

measure_vars_scale_attr <- attributes(scaled_measure_vars)

```


## make single cell object
```{r}


## patient info indexed by sample, filter out out_samples that were filtered out in measured_vars
patient_info <-
  d %>% 
  filter(sample %in% rownames(scaled_measure_vars) ) %>% 
  select(-all_of(str_vars)) %>% 
  select(-matches("avg_MT|spectr|cirru|stratu|0_sum|rate|quartil")) %>% 
  as.data.frame()
rownames(patient_info) <- patient_info$sample


setdiff(names(d), names(patient_info))

## variable info filtering for variables in the measure df
# all(colnames(scaled_measure_vars) %in% info$name)
variable_info <- 
  info %>% 
  filter(name %in% colnames(scaled_measure_vars)) %>% 
  select(-matches("matched|dist"))
rownames(variable_info) <- variable_info$name


# reorder variable_info to match col order of measure var d
# variable_info$variable[match(names(scaled_measure_vars) ,variable_info$variable )] == names(scaled_measure_vars)
variable_info <- variable_info[ match(colnames(scaled_measure_vars) ,variable_info$name ), ]


# check dims
dim(scaled_measure_vars) == dim(measure_vars)
dim(patient_info) == dim(scaled_measure_vars)
nrow(variable_info) == ncol(scaled_measure_vars)
all(rownames(variable_info) == names(scaled_measure_vars))
all(rownames(patient_info) == rownames(scaled_measure_vars))
# ^ all should be TRUE

# add assays
sce <- SingleCellExperiment(assays = list(measure_vars = measure_vars %>% t(), 
                                          scaled_measure_vars = scaled_measure_vars %>% t()),
                            colData = patient_info,
                            rowData =variable_info
                            )  

# save d as metadata
metadata(sce)$d <- d

# add scale attr
metadata(sce) <- append(metadata(sce), list(measure_vars_scale_attr=measure_vars_scale_attr))

```

### Rename Avg MT cols
```{r}

rownames(sce) <- str_remove_all(rownames(sce), "avg_")
rowData(sce)$name <- str_remove_all(rowData(sce)$name, "avg_")

```


```{r, look at dims}

as.data.frame(colData(sce)) %>% 
  left_join(
    as.data.frame(t(assays(sce)$scaled_measure_vars)) %>% 
      rownames_to_column("sample")
  ) %>% 
  check_dims()
# dims : 12525 134
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ removed outlier patients and vars; and filtered out non-str/func vars

### MARCH 17 2025 ###
# dims : 12335 188
# n_patIDs             3997
# n_patID_eyes         7947
# n_observations      12334
# n_patIDs_w_gr_1_vis   490
# ^ removeing vars with var_min_out >0.2462953  missingness and samples with missingness > .3

dim(sce)

```
#_____________

# Imputation and Dimensionality Reduction
#### cluster variables by missingness
```{r}

# add missingness var to rowData
rowData(sce)$msngns_ratio <- rowSums(is.na(assays(sce)$scaled_measure_vars))/ ncol(assays(sce)$scaled_measure_vars)

m <- assays(sce)$scaled_measure_vars %>% 
  t() %>% 
  is.na() %>% 
  {.*1}

rownames(m) <-NULL

# m_dist <- dist(t(m))
# 
# hc <- hclust(m_dist, method = "ward.D2")
# 
# plot(hc)
# 
# clust <- cutree(hc, k = 5)


set.seed(10298)
hm <- Heatmap(m[sample(seq(nrow(m)), 1e3), ],clustering_method_columns = "ward.D2", clustering_method_rows = "ward.D2" , col = c("white", "black"))

pdf(sprintf("%sresults/clustering/09_clustering/imputation_QC/missingness_hm.pdf",wd_path), height = 20, width = 20)
draw(hm)
dev.off()

set.seed(10298)
var_hc <- hclust(dist(t(m)), method = "ward.D2")

plot(var_hc)

num_clust <- 5
data_group_clusters <- cutree(var_hc, k = num_clust)
data_group_clusters <- as.data.frame(data_group_clusters)

# add to rowData
rowData(sce)$msngns_clust <- data_group_clusters$data_group_clusters[match(rownames(data_group_clusters), rownames(rowData(sce)))]

(data_groups <- split(rowData(sce)$name, rowData(sce)$msngns_clust))

# plot hm and annotate by dataset type


```



#### Imputation QC##### Make test/train dataset
```{r}

# how many is actually missing?
sum(is.na(assays(sce)$scaled_measure_vars ))/length(assays(sce)$scaled_measure_vars )
# ^ ~5% of values NA


# make truth dataset: no missing
truth <- assays(sce)$scaled_measure_vars %>% 
  t() %>% 
  {.[rowSums(is.na(.))==0 , ]}

# make test dataset and set 10% missing within each data_group
test <- truth


ratio_na <- .1
row_inds_pool <- seq(nrow(test))
# a list to save indices to 
data_groups_missing_row_inds <- list()

# seeds to use in iterative sampling
set.seed(1202)
seeds <- round(runif(length(data_groups), min = 10, max = 1e6) )

# for each datagroup vector
for(i in seq(data_groups)){
  
  # grab vars of data_groups at i
  dat_group = data_groups[[i]]
  
  # sample 10% of rows
  set.seed(seeds[i])
  row_inds <- sample(row_inds_pool, floor(nrow(test)*ratio_na))
  
  # remove rows from future sampling
  row_inds_pool <- row_inds_pool[-row_inds]
  
  # set these rows missing within this data group
  test[row_inds, dat_group] <- NA
  
  # save inds
  data_groups_missing_row_inds[[i]] <- row_inds
  
}

# all inds should be same length
purrr::map(data_groups_missing_row_inds, length)

# see if artificial missingness recapitulates missinness in truth

# # visualize with hm
# m <- is.na(test[sample(seq(nrow(test)), 500), ])*1
# rownames(m) <- NULL
# Heatmap(m, col = c("white", "black"), clustering_method_columns = "ward.D2", use_raster = TRUE)

# manually hc
test_msngns_hc <- hclust(dist(t(is.na(test))))
test_msngns_clust <- cutree(test_msngns_hc, k = num_clust)

test_msngns_data_groups <- split(names(test_msngns_clust), test_msngns_clust)

# for each data grooup, check that variables all the same
for(i in seq(num_clust)){
  print(all(test_msngns_data_groups[[i]] == data_groups[[i]]))
}


# save 2-d indices for later
inds_2D <- which(is.na(test), arr.ind = TRUE)

# check that indexing missing correct 
sapply( sort(colnames(truth)), function(name){
  # rows of inds where column ind  value matches corresponding col ind of name in truth
  inds <- inds_2D[inds_2D[,2] == which(colnames(truth) == name),]
  # check that missing where should be
  cbind(all(is.na(test[inds])),all(!is.na(truth[inds])))
}) %>% 
  all()
# ^ TRUE :)

```

<!-- ```{r test how indexing missing } -->

<!-- # x <- matrix(rnorm(100),ncol = 4) -->
<!-- x <- as.matrix(iris[,1:4]) -->
<!-- y <- x -->
<!-- ratio_na <- .2 -->
<!-- set.seed(12048) -->
<!-- yinds_1D <- sample(seq(length(x)), ratio_na*length(x)) -->
<!-- y[yinds_1D] <- NA -->

<!-- # save 2-d indices for later -->
<!-- yinds_2D <- which(is.na(y), arr.ind = TRUE) -->

<!-- x -->
<!-- y -->
<!-- yinds_2D -->

<!-- ymice <- mice(y) -->

<!-- ymean <- lapply(ymice$m, function(m){complete(ymice, action = m)}) %>%  -->
<!--   Reduce("+", .) %>%  -->
<!--   {./ymice$m} %>%  -->
<!--   as.matrix() -->

<!-- par(mfrow = c(2,2)) -->
<!-- for(nm in colnames(x)){ -->
<!--   inds <- yinds_2D[ yinds_2D[,2] == which(colnames(x) == nm)] -->
<!--   plot(x[inds], ymean[inds], main = nm, xlab  = "actual", ylab = "predicted") -->
<!-- } -->
<!-- par(mfrow = c(1,1)) -->


<!-- ``` -->


##### test mice methods on 'test'
```{r}

# run imp and compare predicted to actual
# m = 5
m = 11
meth <- "pmm"

if(FALSE){
  how.long(T)
    imp_test <- mclapply(seq(m), function(n){
      mice(test, m = 1, method = "pmm")
    }, mc.cores = detectCores()-1)
  how.long() 
  # ^ 5min
  
  qs::qsave(imp_test, sprintf("%sprocessed_data/NA%.2f_%s_m%d_all_vars_imputation_test_imp_list.qs", wd_path, ratio_na, meth, m))
}else{
  imp_test <- qs::qread(sprintf("%sprocessed_data/NA%.2f_%s_m%d_all_vars_imputation_test_imp_list.qs", wd_path, ratio_na, meth, m))
}


# average imputations: iterate through mice list and select first complete df; reduce list with elementwise matrix sum of all; divide by m for average of matrices
imp_mean <- lapply(seq(m), function(i){
  complete(imp_test[[i]],1)
}) %>% 
  Reduce( "+", .) %>% 
  {./m}

```

##### Evaluate mice imputation
```{r}

# dataframe to save predxactual R correlation
corr_d <- data.frame(variable = character(), R = numeric(), p_value = numeric())

# make imputation directory
if(!dir.exists(sprintf("%sresults/clustering/09_clustering/imputation_QC/",wd_path))) dir.create(sprintf("%sresults/clustering/09_clustering/imputation_QC/",wd_path))

pdf(sprintf("%sresults/clustering/09_clustering/imputation_QC/%s_%i.pdf",wd_path, meth, m), height = 20 , width = 75, bg = "white")
par(mfrow = c(5, 20))
for(name in sort(colnames(imp_mean))){
  
  # rows of inds where column ind  value matches corresponding col ind of name in truth
  inds <- inds_2D[inds_2D[,2] == which(colnames(truth) == name),]
  
  # select inds where colname matches for col ind
  actual <- truth[inds]
  predicted <- imp_mean[inds]
  
  corr_test <- cor.test(actual, predicted, use = "complete.obs")  
  corr <- as.numeric(corr_test$estimate)%>% round(.,3)
  rmse <- rmse(actual, predicted) %>% round(.,3)
  
  title <-"%s;R=%.2f,RMSE=%.2f\nActual vs. Predicted %s at %i iterations" %>% 
    sprintf( ., name,corr,rmse , meth, m)
  
  plot(actual, predicted, main = title)
  add_lines()
  
  corr_d <- rbind(corr_d, data.frame(variable = name, R = as.numeric(corr_test$estimate), p_value = corr_test$p.value))
}
par(mfrow = c(1,1))
dev.off()

#  add missingness variable
ratio_missing <-  apply(assays(sce)$scaled_measure_vars, 1, function(x){sum(is.na(x))/length(x)})
ratio_missing <- as.data.frame(cbind(ratio_missing,variable = names(ratio_missing)))
corr_d<- corr_d %>% 
  left_join(ratio_missing)

# save imputation correlations
(corr_d <- corr_d[order(corr_d$R),])

# # save corr_d to file
# write_csv(corr_d, sprintf("%sprocessed_data/imputation_Rs/all_vars_impute_Rs.csv", wd_path))


# make common var to join on 
corr_d$name = corr_d$variable
corr_d$variable <- NULL

# add to rowdata
corr_d <- corr_d[match(rownames(sce) , corr_d$name), ]

if(all(rownames(sce) == corr_d$name)) rowData(sce) <- cbind(rowData(sce), corr_d[, -which(names(corr_d) == "name")]) else stop("check corr_d name order")

```

###### Corr bar
```{r}

# make significance bar
corr_d$signif <- with(corr_d,
                      ifelse(p_value<.001, "***", ifelse(p_value<.01, "**", ifelse(p_value<.05, "*", "not significant"))) %>%
                        factor(., levels = c("***", "**", "*", "not significant"))
)

ggplot(corr_d, aes(R,reorder(name, R), fill = signif ))+
  geom_bar(stat = "identity", show.legend = TRUE, position = "dodge", alpha = .8)+
  geom_text(aes(label = sprintf("%s%%", round(as.numeric(ratio_missing), 3)*100) ))+
  ggtitle(sprintf("Imputation performance across variables for between data type imputation (%s;m=%s)", meth, m), "Percentages indicate percentage missing in full dataset")+
  scale_fill_viridis_d(begin = 1, end = 0)+
  ylab("Feature")+
  theme_bw()
ggsave(sprintf("%sresults/clustering/09_clustering/imputation_QC/between_imputations_R_bar.pdf", wd_path), height = 12, width = 10)

ggsave(sprintf("%sresults/publication_figures/between_imputations_R_bar.pdf", wd_path), height = 12, width = 10)

```

###### Average rho (fisher transformation)
    
    Pearson is nonlinear so i used the fisher transformation (atanh(rho)) to calculate the mean:
    # Raw pearsons
      rho.mean 0.2280444
      rho.sd 0.1911418
      rho.CI.lwr -0.1461240
      rho.CI.upr 0.5451321
    # With transformation:
      z.mean 0.2485974
      z.sd 0.2392857
      z.CI.lwr -0.2206646
      z.CI.upr 0.6244028
```{r}

# fisher Transformation of R

# z = .5*ln((1+rho) / (1-rho)) = artanh(r), where artanh is the inverse hyperbolic tangent function. 
fisher_transformation = function(r){
  # .5*log( (1+r) / (1-r) )
  atanh(r)
}

rho = corr_d$R
z = fisher_transformation(rho)

plot(rho, z)

rho_mean = mean(rho, na.rm = TRUE)
rho_sd = sd(rho, na.rm = TRUE)
rho_CI = setNames(rho_mean +c(-1,1) * 1.96 *rho_sd , c("lwr", "upr")) 

z_mean = mean(z, na.rm = TRUE)
z_sd = sd(z, na.rm = TRUE)
z_CI = setNames(z_mean +c(-1,1) * 1.96 *z_sd , c("lwr", "upr")) 

# transform back to R
res <- c(
  "z.mean.backtransformed" = tanh(z_mean), 
  "z.sd.backtransformed" = tanh(z_sd), 
  "z.CI.backtransformed" = tanh(z_CI),
  "rho.mean" = rho_mean, 
  "rho.sd" = rho_sd, 
  "rho.CI" = rho_CI
  )

t(t(res))

```


# Actual imputation

## ** switch [ TRUE to run the imputation
```{r}
##
####
######
########
run_imputation <- FALSE
########
######
####
##
```

### impute 
```{r}

# params for imputation
imp.method = "pmm"
# imp.m = 5
imp.m = 11 #(12 cores total)

# run imp on scaled measure vars
if(run_imputation){

  # pred <- quickpred(assay(sce , "measure_vars") %>% t())
  scaled_pred <- quickpred(assay(sce , "scaled_measure_vars") %>% t())
  
  # parralel impute imp.m times
  how.long(T)
    imp_of_scaled_l <- mclapply(seq(imp.m), function(n){
      set.seed(23939)
      mice(assay(sce , "scaled_measure_vars") %>% t(), m = 1, method = imp.method, predictorMatrix = scaled_pred)
    },mc.set.seed = F, mc.cores = detectCores()-1)
  how.long() # ~5min

  qs::qsave(imp_of_scaled_l, sprintf("%s/processed_data/%s_m%d_scaled_measure_vars_imp_list.qs", wd_path, meth, m))
  
}else{
  imp_of_scaled_l <- qs::qread(sprintf("%s/processed_data/%s_m%d_scaled_measure_vars_imp_list.qs", wd_path, meth, m))
}

# average imputations: iterate through mice list and select first complete df; reduce list with elementwise matrix sum of all; divide by m for average of matrices
imp_of_scaled_avg <- lapply(seq(imp.m), function(i){
  complete(imp_of_scaled_l[[i]],1)
}) %>% 
  Reduce( "+", .) %>% 
  {./m}


## look at how imputation works
boxplot(imp_of_scaled_avg %>% scale, na.omit = TRUE, las  = 2)

# add impute average to sce
assay(sce, "pmm_imp_of_scaled") <- imp_of_scaled_avg %>% t


```

## fix some variables 

## ensure ISOS_BREAK and ISOS_LOC correct
    Set ISOS_LOC_CEN/JF to 0 if ISOS_BREAK 0
```{r}
 

# look at how loc impuuted
isos <- assays(sce)$pmm_imp_of_scaled %>%
  t() %>%
  data.frame() %>%
  select(matches("ISOS"))



sds <- measure_vars_scale_attr$`scaled:scale`
means <- measure_vars_scale_attr$`scaled:center`

# rescale to original distr: mutiply each row elementwise by sds, do same summing with mean
isos_rescaled <- isos %>%
  sweep(2, sds[names(isos)], "*") %>%
  sweep(2, means[names(isos)], "+")

# plot distribution before and after unscale
par(mar = c(5,5,3,3))
isos %>% unlist() %>%density() %>% plot(xlim = c(-3, 3), ylim = c(0, 2)) # txtdensity(width = 100, height = 10) 
isos_rescaled %>% unlist() %>%
  density() %>% 
  lines(col = "blue", lw = 2) # txtdensity(width = 100, height = 10) 


isos_rescaled%>%
  mutate_all(round) %>%
  group_by(across(everything())) %>% 
  mutate(count = n()) %>% 
  distinct() %>% 
  arrange(desc(count), desc(ISOS_BREAK)) %>% 
  ungroup() %>% 
  rowwise() %>% 
  mutate(bar = paste( rep("█",  max(1,ceiling(count/90)) ), collapse = "" )) %>% 
  print(n = 100)
# ISOS_BREAK ISOS_LOC_JF ISOS_LOC_CEN count bar                                                          
#          2           0            1  5487 █████████████████████████████████████████████████████████████
#          2           1            0  4195 ███████████████████████████████████████████████              
#          0           0            0  1864 █████████████████████                                        
#          1           1            0   618 ███████                                                      
#          1           0            1    83 █                                                            
#          1           0            0    19 █     # set to missing? no, set to JF, assuming they are early on        
#          2           0            0    10 █     # set to missing? no, set to JF, assuming they are early on        


# round to id values to replace
isos_rescaled_round <- isos_rescaled %>% 
  apply(.,2,round) 

# grab samps to make JF 1 where curretnly it is 0
make_1_samps <- isos_rescaled_round[isos_rescaled_round[, 1] %in% c(2,1) & isos_rescaled_round[,2]== 0 & isos_rescaled_round[,3]== 0,] %>% rownames()

isos[,"ISOS_LOC_JF"] %>%density() %>% plot(xlim = c(-3, 3), ylim = c(0, 3.5)) # txtdensity(width = 100, height = 10) 
isos_rescaled[,"ISOS_LOC_JF"] %>% 
  density() %>% 
  lines(col = "blue", lw = 2) # txtdensity(width = 100, height = 10) 

# scale 1 back to assay distr to add back
val_1 <- (1-means["ISOS_LOC_JF"])/sds["ISOS_LOC_JF"]

# set samps at ISOS_LOC_JF to val
assays(sce)$pmm_imp_of_scaled["ISOS_LOC_JF",make_1_samps] <- val_1



```

### Calculate foveal slope and average parafoveal zones
```{r}

assay <- assays(sce)$pmm_imp_of_scaled %>% 
  t() %>% 
  as.data.frame()

# average parafovea
parafovea <- rowMeans(assay[, grep("MT0[2345]", colnames(assay))])

# add to assay
assay <- cbind(assay, parafovea)

# ****** Function: calculate slope angle in degrees from zone 1 up to zone 2,3,4,5 ******
## difference in thickness over distance to zone (convert from rad to deg)
calc_slope <- function(sector_of_interest, sector_one){
    180 *  (atan((sector_of_interest - sector_one) / 500)) / pi
  }

# use avg columns
inner_zones <- grep("MT0[12345]", colnames(assay), value = TRUE)

# scales for rescaling
## rename 
names(attributes(scaled_measure_vars )$`scaled:scale`) <- str_remove_all(names(attributes(scaled_measure_vars )$`scaled:scale`), "avg_")
names(attributes(scaled_measure_vars )$`scaled:center`) <- str_remove_all(names(attributes(scaled_measure_vars )$`scaled:center`), "avg_")

sds <- attributes(scaled_measure_vars )$`scaled:scale`
means <- attributes(scaled_measure_vars )$`scaled:center`


# nm <- "ISOS_LOC_CEN"
# rescale to original; 
for(colnm in inner_zones){
  where <- which(names(sds) == colnm)
  assay[,colnm] <-  (assay[where]*sds[where] )+ means[where]
}


# calculate slope against zone 1
assay <- assay %>% 
  mutate(across(all_of(inner_zones[-1]),
                ~calc_slope(., MT01),
                .names = "{.col}_slope"
                )) %>% 
  mutate(across(matches("MT0"), ~as.numeric(scale(.))))

assay <- data.matrix(t(assay))

# add new vars to rowdata
# dput(data.frame(rowData(sce))[0,])
new_rowdat <-structure(list(name = "parafovea", table = NA, table_description = "OCTMT", description = "Average of parafoveal zones 2-5", shorthand = NA, 
    X0 = NA, X1 = NA, X2 = NA, X3 = NA, X4 = NA, X5 = NA, X6 = NA, continuous = TRUE, nominal = FALSE, zonal = FALSE, can_resolve = FALSE, 
    ordinal = FALSE, msngns_ratio = NA, msngns_clust = NA, R = NA, p_value = NA, ratio_missing = NA), class = "data.frame", row.names = c(NA, -1L))

other_new_rowdat <- new_rowdat[rep(1,length(inner_zones)-1)  , ]

other_new_rowdat$name <- paste(inner_zones[-1], "slope", sep = "_")
other_new_rowdat$description <- "Slope in degrees from zone 1 to parafoveal zone"

both_new_rowdat <- rbind(new_rowdat, other_new_rowdat)
rownames(both_new_rowdat) <- both_new_rowdat$name

rowdat <- as.data.frame(rowData(sce) )

rowdat <- rbind(rowdat, both_new_rowdat)

# make new experiment with added data

na_rows <- rbind(parafovea = rep(NA, ncol(sce)), 
                 MT02_slope = rep(NA, ncol(sce)),
                 MT03_slope = rep(NA, ncol(sce)),
                 MT04_slope = rep(NA, ncol(sce)),
                 MT05_slope= rep(NA, ncol(sce))
                 )
colnames(na_rows) <- colnames(sce)

measure_vars_new = rbind(assays(sce)$measure_vars, na_rows)
scaled_measure_vars_new = rbind(assays(sce)$scaled_measure_vars, na_rows)

rowdat <- rowdat[match(rowdat$name, rownames(assay)) , ]

all(rownames(rowdat) ==rownames(assay))

purrr::map(list(measure_vars_new = measure_vars_new,scaled_measure_vars_new=scaled_measure_vars_new,  assay=assay,colData= colData(sce), rowdat=rowdat), dim)


# construct new experiment
new_sce <- SingleCellExperiment(assays = list(measure_vars = measure_vars_new, 
                                          scaled_measure_vars = scaled_measure_vars_new, 
                                          pmm_imp_of_scaled = assay),
                            colData = colData(sce),
                            rowData =rowdat
                            )  

metadata(new_sce) <- metadata(sce)

sce <- new_sce

```



# filter out variables 
## that didnt impute well
```{r}

# variables with non-significant correraltion btw actual and predicted and missingness > 5%
(rm_vars <- corr_d$name[!is.na(corr_d$p_value) & corr_d$p_value>.05 & corr_d$ratio_missing>.05])
# "ICL02" "ICL06" "RW014" "MT06"  "MT07"  "MT08"  "MT09" 

sce <- sce[-which(rownames(sce) %in% rm_vars) , ]

```

## remove parafovea 
```{r}

# og parafovea now redundant, replace with new one
colData(sce)$parafovea <- NULL

colData(sce)$parafovea <- assays(sce)$pmm_imp_of_scaled["parafovea",]

sce <- sce[-which(rownames(sce) == "parafovea") , ]


```


#___________________________
# calculate reduced dims and distance
### ** switch [ TRUE to calculate distance and run reduced dims
```{r}
##
####
######
########
run_rdims <- TRUE
########
######
####
##
```

### calculate distance run PCA, TSNE, and UMAP
```{r}


# SET PARAMS----

# umap_params <- list(x = sce, dimred =  "PCA", metric = "euclidean", spread = 1.2, min_dist = 0.005, n_trees = 25, n_neighbors = 15, search_k = 750) # pre isos_loc_JF fix
# umap_params <- list(x = sce,   dimred = "PCA", metric = "cosine", spread = 0.8, min_dist = 0.005, n_trees = 25, n_neighbors = 10, search_k = 500) # post isos_loc_JF fix
# umap_params <- list(x = sce,   dimred = "PCA", metric = "cosine", spread = 0.8, min_dist = 0.005, n_trees = 25, n_neighbors = 10, search_k = 500) # post isos_loc_JF fix

# umap_params <- list( dimred = "PCA", metric = "cosine", spread = 1.2, min_dist = 0.005, n_trees = 25, n_neighbors = 20, search_k = 1000, seed = 29209)
umap_params <- list( dimred = "PCA",  metric = "euclidean", spread = 1.2, min_dist = 0.02, n_trees = 50, n_neighbors = 20, search_k = 2000, seed = 29209)


# set.seed(29209)
# exag <- 20 # 12
# tsne_par <- list(x = sce, dimred = structure(1L, levels = "PCA", class = "factor"),
#               perplexity = 400, # 50, # increases sphericity of points
#               momentum = 0.9, # 0.9,
#               final_momentum = 0.5, #0.9,
#               eta = floor(ncol(sce)/exag),  #100,
#               exaggeration_factor = exag #12
#               )
# 
# tsne_par <- list(x = sce, dimred = structure(1L, levels = "PCA", class = "factor"), 
#               perplexity = 50, # 50, 
#               momentum = 0.9, # 0.9, 
#               final_momentum = 0.9, #0.9,
#               eta = 100,  #100, 
#               exaggeration_factor = 12 #12
#               )
#+ ^ rules of thumb i infered: hi momentum, high perplexity, low eta, high final momentum, high momentum, medium exageration factor, 
  
tsne_params <- list(dimred =  "PCA",perplexity = 50,final_momentum = 0.9,momentum =0.5,eta = 100,exaggeration_factor = 4 , seed  = 29209)

# RUN RDIMS ----
if(run_rdims){
  
  set.seed(29209)
  sce <- runPCA(sce,exprs_values = "pmm_imp_of_scaled" )
  
  umap_par <- append(umap_params[-which(names(umap_params) == "seed")], list(x = sce))
  set.seed(umap_params$seed)
  sce <- do.call(runUMAP, umap_par)
  
  tsne_par <- append(tsne_params[-which(names(tsne_params) == "seed")], list(x = sce))
  set.seed(tsne_params$seed)
  sce <- do.call(runTSNE, tsne_par)
  
  # save separate sce that contains reduced dims
  qsave(sce, paste0(wd_path, "processed_data/sce_reduced_dims.qs"))
  
}else{
  # put reducued dims dims into sce
  sce_reduced_dims <- qread(paste0(wd_path, "processed_data/sce_reduced_dims.qs"))
  
  # metadata(sce)$distance <- metadata(sce_reduced_dims)$distance
  reducedDims(sce)$PCA <- reducedDims(sce_reduced_dims)$PCA
  reducedDims(sce)$TSNE <- reducedDims(sce_reduced_dims)$TSNE
  reducedDims(sce)$UMAP <- reducedDims(sce_reduced_dims)$UMAP
  rm("sce_reduced_dims")
}

# add params to metadata
metadata(sce) <- append(metadata(sce), list(params = list(tsne = tsne_params, umap = umap_params)))

sce

# pmm_pca <- reducedDim(sce)
# 
# attributes(pmm_pca)$rotation %>%
#   as.data.frame() %>%
#   rownames_to_column("var") %>%
#   select(var, PC1:PC3) %>%
#   pivot_longer(matches("PC"), names_to = "PC") %>%
#   ggplot(aes(y = reorder_within(var, value, PC), x= value, fill = var))+
#   geom_bar(stat = "identity", orientation = "y", show.legend = FALSE)+
#   facet_wrap(~PC, scales = "free_y")+
#   scale_y_reordered() +
#   theme(axis.text.y = element_text(size = 4))+
#   ggtitle("Imputed assay (scaled) PCA loadings\n(free_y)")
# 
# if(!dir.exists(paste0(wd_path, "results/clustering/09_clustering/dimred"))) dir.create(paste0(wd_path, "results/clustering/09_clustering/dimred"))
# 
# ggsave(paste0(wd_path, "results/clustering/09_clustering/dimred/PCA_rotations.png"), height = 10, width = 20 )


```



# = = = = = = = = = = = 
# save data
```{r}

metadata(sce)$info <- info
metadata(sce)$raw <- raw

# save SCE
saveHDF5SummarizedExperiment(sce, paste0(wd_path, "processed_data/sce001.h5"), replace = T)

# notify.me("09_clustering_imp_and_dimred.Rmd complete")

```



```{r, look at dims}

as.data.frame(colData(sce)) %>% 
  left_join(
    as.data.frame(t(assays(sce)$scaled_measure_vars)) %>% 
      rownames_to_column("sample")
  ) %>% 
  check_dims()
# dims : 12525 143
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ added clustering variables and letter averages

### MARCH 19 2025 ###
# dims : 12276 169
# n_patIDs             3971
# n_patID_eyes         7888
# n_observations      12275
# n_patIDs_w_gr_1_vis   492

dim(sce)

```
