---
title: "09_clustering"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---
# options
```{r}
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()



```

# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

pacman::p_load(tidyr,dplyr,reshape2,purrr,GGally,tidyverse,ggpubr,vcd,ggplot2,tidytext,gridExtra,ComplexHeatmap,zoo,mgcv,car,zoo,tools,mice,grDevices,ggrepel,grid,gplots,Metrics,stats,lme4,lmerTest,dendextend,magick,umap,scater,SingleCellExperiment,scran,bluster,slingshot,ggfortify,pheatmap,TSCAN,igraph,knitr, qs, patchwork, parallel,txtplot, HDF5Array)

rm(list = ls())
gc()
```

# functions
```{r}

source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```


# Load Data

```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"


sce <- HDF5Array::loadHDF5SummarizedExperiment(sprintf("%sprocessed_data/sce001.h5", wd_path))

```

## mung data

### update rowdata
```{r}


# rowData(sce) %>%
#   as.data.frame() %>%
#   select(name:description) %>%
#   filter(if_any(everything(),is.na))

# add description to mt
rowData(sce)$table_description[grepl("MT0\\d$", rowData(sce)$name)] <- "OCTMT"

# add description to isos_loc
rowData(sce)$description[grepl("ISOS_LOC", rowData(sce)$name)] <- "Juxtafoveal/central inner-segment/outer-segment break"

# add description to isos_loc
rowData(sce)$table_description[grepl("ISOS_LOC", rowData(sce)$name)] <- "OCTMAIN"

```

### update colData
```{r}

# add original missingness to coldata
msngns_ratio <- colSums(is.na(assays(sce)$measure_vars))/ nrow(assays(sce)$measure_vars)
txtplot::txtdensity(msngns_ratio, pch = '˙', height= 20, xlab = "Sample Missingness")
txtplot::txtboxplot(msngns_ratio, width = 49, legend = TRUE)
#   +----+-----------+------------+-----------+-----------+------------+------------+
#    |   ˙                                                                           |
# 30 +   ˙˙                                                                          +
#    |    ˙                                                                          |
#    |    ˙                                                                          |
#    |    ˙                                                                          |
#    |     ˙                                                                         |
# 20 +     ˙                                                                         +
#    |     ˙                                                                         |
#    |     ˙                                                                         |
#    |      ˙                                                                        |
#    |      ˙                                                                        |
# 10 +      ˙                                                                        +
#    |       ˙                                                                       |
#    |       ˙                         ˙˙˙˙                                          |
#    |        ˙˙     ˙˙˙˙˙˙˙˙        ˙˙˙  ˙˙˙˙˙˙˙˙˙                                  |
#  0 +         ˙˙˙˙˙˙˙      ˙˙˙˙˙˙˙˙˙˙            ˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙˙    +
#    +----+-----------+------------+-----------+-----------+------------+------------+
#       0.05         0.1         0.15         0.2        0.25          0.3            
#                                   Sample Missingness                                

#       0.05         0.1        0.15         0.2       
#      |--+-----------+-----------+-----------+-------|
#        +----------------+                            
#     1  |                |------------------------    
#        +----------------+                        

all(colnames(assays(sce)$measure_vars)==rownames(colData(sce)))
# ^ TRUE

# add to colData
colData(sce) <- cbind(colData(sce), msngns_ratio)


# add sample to colData
colData(sce)$sample <- rownames(colData(sce))

```

### fix CTR pt vars names
```{r}

old <- grep("CTR", rownames(sce), value = TRUE)

rowData(sce) %>% 
  as.data.frame() %>% 
  filter(name %in% old) %>% 
  select(name, description) 

name_change <- setNames(c("LARGEAREA_DECR_AF_CTR_PT_INVOLV",'INR_EMPTY_SPC_CTR_PT_INVOLV','OUTR_EMPTYSPC_CTR_PT_INVOLV'), c("CTR_PT_INVOLV",    "CTRPOINT_INVOLVD","CTRPOINT_INVOL_2"))
  
# corr <- assays(sce)$pmm_imp_of_scaled %>% 
#   as.matrix() %>% 
#   t() %>% 
#   cor()
# 
# corr[!grepl("CTR", colnames(corr)), grepl("CTR", colnames(corr))] %>% 
#   {(abs(.)^3)*sign(.)} %>% 
#   Heatmap()
# 
# rowData(sce) %>% 
#   as.data.frame() %>% 
#   filter(grepl("AF|IN|OUTR",name) )

rownames(sce)[match(names(name_change), rownames(sce))] <- name_change


```



#___
# Clustering
## KMEANS then NN
```{r}

# using kmeans then graph nn methods ----------
## OG seed = 43933; centers = 1e3; k = 7
# 
# # original
# set.seed(43933)
# km_nn <- clusterCells(sce, use.dimred="PCA",
#     BLUSPARAM=TwoStepParam(
#         first=KmeansParam(centers=1e3),
#         second=NNGraphParam(k=7)
#     ), full = TRUE
# )

## for when 2 ISOS_LOC vars :: 12 clusters
# set.seed(1010); # 1010 (11) >> 1010(15)>> 1010(11)>>60989(14)
# km_nn <- clusterCells(sce, use.dimred="PCA",
#   BLUSPARAM=TwoStepParam(
#       first=KmeansParam(centers=1440), # 1440 (11)>> 1430(15)>> 1430(11)>>1125(14)
#       second=NNGraphParam(k=10) # 12 (11)>>12 (15) >> 13 (11)>>8(14)
#   ), full = TRUE
# )

km_nn_params <- list(seed = 1010, centers = 1440, k = 10)

set.seed(km_nn_params$seed)
km_nn <- clusterCells(sce, use.dimred="PCA",
  BLUSPARAM=TwoStepParam(
      first=KmeansParam(centers=km_nn_params$centers), # 1440 (11)>> 1430(15)>> 1430(11)>>1125(14)
      second=NNGraphParam(k=km_nn_params$k) # 12 (11)>>12 (15) >> 13 (11)>>8(14)
  ), full = TRUE
)

table(km_nn$clusters)

# add hclusters as labels
sce$km_nn_cluster <- km_nn$clusters

# add params
metadata(sce)$params <- append(metadata(sce)$params, list(km_nn = km_nn_params))

# (file <- tempfile())
# write_csv(data.frame(sample = sce$sample,cluster = sce$km_nn_cluster), file)

# set.seed(2920)
# sce <- runTSNE(sce, dimred="PCA")


k_nn_pca <- plotReducedDim(sce, dimred="PCA", text_by = "km_nn_cluster",color_by = "km_nn_cluster")+ 
  ggtitle("k-means, nearest neighbor PCA")
(k_nn_tsne <- plotReducedDim(sce, dimred="TSNE",text_by = "km_nn_cluster", color_by = "km_nn_cluster") + 
    ggtitle("k-means, nearest neighbor TSNE"))
k_nn_umap <- plotReducedDim(sce, dimred="UMAP",text_by = "km_nn_cluster", color_by = "km_nn_cluster")+
  ggtitle("k-means, nearest neighbor UMAP")

ggsave(paste0(wd_path, "results/clustering/09_clustering/km_nn_clusters.png"),  
       k_nn_pca+k_nn_tsne+k_nn_umap, 
        height = 10, width = 30)


```

### permute km params
#### try combinations of parameters
```{r, eval = FALSE}

params <- expand.grid(seed = c(1010, 1120,102903,12099, 123040, 60989),
                     centers = c(1440,1430, 1435, 1445, 1425, 1450  ), 
                     k = c(12,10, 11,  13, 9, 14 ), KEEP.OUT.ATTRS = F)

params_filled <- params

how.long(T)
params_filled$num_clust <- mclapply(seq(nrow(params)), function(i){
  
  set.seed(params$seed[i]); 
  km_nn <- clusterCells(sce, use.dimred="PCA",
    BLUSPARAM=TwoStepParam(
        first=KmeansParam(centers=params$centers[i]),
        second=NNGraphParam(k=params$k[i])
    ), full = TRUE
  )
  
  n_distinct(km_nn$clusters)
}, mc.cores = detectCores()-1) %>% 
  unlist()
how.long() # 1.2 min


params_filled$num_clust%>% txtdensity()


params_sub <- params_filled[params_filled$num_clust>11, ]
params_sub <- params_sub[order(params_sub$num_clust) , ]

sprintf("%sresults/clustering/09_clustering/clustering_iter/", wd_path) %>% 
  {if(!dir.exists(.)){ dir.create(.)}}

how.long(T)
plots <- mclapply(seq(nrow(params_sub)), function(i){
  
  set.seed(params_sub$seed[i]); 
  km_nn <- clusterCells(sce, use.dimred="PCA",
    BLUSPARAM=TwoStepParam(
        first=KmeansParam(centers=params_sub$centers[i]),
        second=NNGraphParam(k=params_sub$k[i])
    ), full = TRUE
  )
  
  pars <- unlist(params_sub[ i , ])
  title = paste(paste(names(pars), pars), collapse = "; ")
  
  UMAP <- plotUMAP(sce, color_by = I(km_nn$clusters))+
    ggtitle(title)+
    theme(aspect.ratio = 1)
  
  TSNE = plotTSNE(sce, color_by = I(km_nn$clusters))+
    ggtitle(title)+
    theme(aspect.ratio = 1)
  
  ggsave(sprintf("%sresults/clustering/09_clustering/clustering_iter/%s.png", wd_path, paste(paste(names(pars), pars, sep = "_"), collapse = "_")), 
         TSNE+UMAP, height  = 7, width = 14)
  
}, mc.cores = detectCores()-1)  
how.long() # ~2 min


```

#### use seed with best results with more combinations of center and k
```{r, eval  =FALSE}

params <- expand.grid(seed = 60989,
                     centers = seq(1120, 1150, 5), 
                     k = seq(6,15,2), KEEP.OUT.ATTRS = F)


sprintf("%sresults/clustering/09_clustering/clustering_iter2/", wd_path) %>% 
  {if(!dir.exists(.)){ dir.create(.)}}

how.long(T)
plots <- mclapply(seq(nrow(params)), function(i){
  
  set.seed(params$seed[i]); 
  km_nn <- clusterCells(sce, use.dimred="PCA",
    BLUSPARAM=TwoStepParam(
        first=KmeansParam(centers=params$centers[i]),
        second=NNGraphParam(k=params$k[i])
    ), full = TRUE
  )
  
  pars <- unlist(params[ i , ])
  title = paste(paste(names(pars), pars), collapse = "; ")
  
  UMAP <- plotUMAP(sce, color_by = I(km_nn$clusters))+
    ggtitle(title)+
    theme(aspect.ratio = 1)
  
  TSNE = plotTSNE(sce, color_by = I(km_nn$clusters))+
    ggtitle(title)+
    theme(aspect.ratio = 1)
  
  ggsave(sprintf("%sresults/clustering/09_clustering/clustering_iter2/%s.png", wd_path, paste(paste(names(pars), pars, sep = "_"), collapse = "_")), 
         TSNE+UMAP, height  = 7, width = 14)
  
}, mc.cores = detectCores()-1)  
how.long()

```

#### move plots
```{r, eval = FALSE}

# move plots of interest to own folder
km_plots_files <- list.files(sprintf("%sresults/clustering/09_clustering/clustering_iter/",wd_path))
km_plots_files_sub <- grep("clust_[11-20]", km_plots_files, value = TRUE) %>% 
  grep("10.png", ., value = TRUE, invert = TRUE)

dest_dir <- sprintf("%sresults/clustering/09_clustering/clustering_iter/sub_clustering_iter",wd_path)

# make dir, clear if already there and make it , if not there, just make it
dest_dir %>% 
  {if(dir.exists(.)){
      unlink(., recursive = TRUE)
      dir.create(.)
      } else{
        dir.create(.)
      }
    }

# Construct full source and destination paths
source_paths <- file.path(sprintf("%sresults/clustering/09_clustering/clustering_iter",wd_path), km_plots_files_sub)
dest_paths <- file.path(dest_dir, km_plots_files_sub)

# Move the files
file.copy(source_paths, dest_paths)

```


# Analyse clusters

## **** COMBINE CLUSTERS THAT ARE SIMILAR ****
```{r}

# combine 12>>3, 4>>7, 9>>2
sce$my_cluster <- sce$km_nn_cluster
sce$my_cluster[sce$my_cluster == "4"] <- "7"
sce$my_cluster[sce$my_cluster == "9"] <- "2"
sce$my_cluster[sce$my_cluster == "12"] <- "3"


unique(sce$my_cluster)  %>% as.numeric()

plot(as.numeric(as.character(sce$km_nn_cluster)), as.numeric(as.character(sce$my_cluster)))
axis(side = 1, at = seq(0, 16, by = 1))
axis(side = 2, at = seq(0, 16, by = 1))
abline(h = 0:15, lty = 2, col= 'grey80')
abline(v = 0:15, lty = 2, col= 'grey80')

# refactor; set to numeric; factor levels
sce$my_cluster <- factor(as.numeric(factor(sce$my_cluster , 
  levels = as.character(c(1:3,5:8,10:11,13:16)))))

```


## aggregate across cells and plot Heatmap of cluster x variable for NN and HC clustering on assay and assay distance matrix
```{r}

# what assay to aggregate on 
# assay_to_aggregate <- "scaled_measure_vars"
assay_to_aggregate <- "pmm_imp_of_scaled"

# KM-NN AGGREGATE ------------

# aggregate clusters by mean of each variable on unscaled assay (mean of raw value versus scaled values)
agg_cell <- aggregateAcrossCells(sce , use.assay.type = assay_to_aggregate, ids = sce$my_cluster, statistics = "mean")

# scale aggregate assay
assay(agg_cell, "scaled") <- assay(agg_cell, assay_to_aggregate) %>% t() %>% scale() %>% t()

# Get cluster sizes
cluster_sizes <- table(sce$my_cluster)

# Define column annotation with cluster sizes
col_anno <- HeatmapAnnotation(
  "Cluster Size" = anno_barplot(as.matrix(cluster_sizes),
                                gp = gpar(fill = "grey"),
                                border = TRUE, 
                                # width =unit(10, 'cm') ,
                                height = unit(10, 'cm')
                                )
)


# Define colors for each data category
category_colors <- setNames(RColorBrewer::brewer.pal(n = length(unique(rowData(sce)$table_description)), "Set3"), unique(rowData(sce)$table_description))
msngness_clust_colors <- setNames(RColorBrewer::brewer.pal(n = length(unique(rowData(sce)$msngns_clust)), "Set3"), paste0("X",unique(rowData(sce)$msngns_clust)))

row_anno <-  rowAnnotation(
  data_category = rowData(sce)$table_description,
  # msngness_clust = paste0("X",rowData(sce)$msngns_clust),
  col = list(
      data_category = category_colors
      # msngness_clust = msngness_clust_colors
      )
  )

hm_m <- assay(agg_cell, "scaled")

# Define heatmap
set.seed(102)
hm <- Heatmap(
  hm_m,
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_title = "Cluster Mean Profiles",
  row_title = "Features",
  bottom_annotation = col_anno,
  left_annotation = row_anno,
  heatmap_legend_param = list(
    title = "Scaled\nValue",
    color_bar = "continuous"
  )
)


# Save plot
pdf(paste0(wd_path, "results/clustering/09_clustering/aggregated_my_clusters.pdf"), height = 20, width = 10)
set.seed(123)
draw(hm,padding = unit(c(5, 5, 5, 5), "mm"), ht_gap = unit(30, "mm"))
dev.off()

# save column/row ordering
set.seed(123)
hm_rw_order <- setNames(rownames(hm_m)[row_order(hm)], seq(rownames(hm_m)))
set.seed(123)
hm_col_order <- setNames(colnames(hm_m)[column_order(hm)], seq(colnames(hm_m)))

# save hm ordering
metadata(sce) <- append(metadata(sce), 
                        list(other = 
                               list(cluster_agg_hm_ordering = 
                                      list(col_order = hm_col_order, row_order =hm_rw_order )
                                    )
                             )
                        )


# make phylo tree
# Convert hclust to a dendrogram object
dend <- as.dendrogram(hclust(dist(t(assay(agg_cell, "scaled")))))

# Convert to a phylogenetic tree
phylo_tree <- ape::as.phylo(dend)

png(paste0(wd_path, "results/clustering/09_clustering/aggregated_my_clusters_tree.png"), height = 700, width = 700)
plot(phylo_tree, type = 'tidy', cex = 2)
dev.off()


# Transpose the matrix
hm_m_t <- t(hm_m)

# Define new column annotation (was row_anno before transpose)
col_anno_t <- HeatmapAnnotation(
  data_category = rowData(sce)$table_description,
  col = list(data_category = category_colors)
)

# Define new row annotation (was col_anno before transpose)
row_anno_t <- rowAnnotation(
  "Cluster Size" = anno_barplot(as.matrix(cluster_sizes),
                                 gp = gpar(fill = "grey"),
                                 border = TRUE, 
                                 width = unit(3, 'cm'))
)

# Draw heatmap
set.seed(123)
hm_t <- Heatmap(hm_m_t,
        column_title = "Cluster Feature Averages",
        name = "Feature Mean\nScaled",
        column_names_gp = gpar(fontsize = 10), 
        column_names_max_height  = unit(10, "cm"),
        top_annotation = col_anno_t,
        right_annotation = row_anno_t
)

# Save plot
pdf(paste0(wd_path, "results/poster_plots/aggregated_my_clusters.pdf"), height = 7, width = 20, bg = "transparent")
set.seed(123)
draw(hm_t,padding = unit(c(5, 5, 5, 5), "mm"), ht_gap = unit(30, "mm"),background = "transparent")
dev.off()


# Draw heatmap with annotation
set.seed(123)
hm <- Heatmap(hm_m,
        column_title = "Aggregated KM-NN",
        name = "cluster mean scaled\nacross clusters",
        row_names_gp = gpar(fontsize = 10),
        row_names_max_width = unit(2, "cm"),
        bottom_annotation =  col_anno, 
        right_annotation = row_anno
)



# plot differently
transformed_m <- assay(agg_cell, assay_to_aggregate) %>%
  t() %>% 
  scale() %>% 
  t() %>% 
  {(abs(.)^3) * sign(.)} 

boxplot(transformed_m)

Heatmap(transformed_m,
        cluster_rows = TRUE,
        cluster_columns = TRUE,
        column_title = "Aggregated KM-NN, transformed",
        row_names_gp = gpar(fontsize = 10),
        row_names_max_width = unit(2, "cm"), column_dend_height = unit(4, "cm")
        )
# grab vars greater than threshold 
vars_per_clust <- apply(transformed_m, 2, function(col){
  boxplot(col, plot = FALSE)$out
})

var_clusts <- split(rep(names(vars_per_clust), lengths(vars_per_clust)), unlist(lapply(vars_per_clust, names)))

var_clusts <- purrr::map(var_clusts, as.data.frame)

var_clusts <- purrr::map(names(var_clusts), ~{
  nm <- .
  x <- var_clusts[[nm]]
  names(x) <- "clust"
  x[, "var"] <- nm
  return(x)
})

var_clusts_d <- var_clusts %>% do.call(rbind,.)

# make df and write csv
write.csv(var_clusts_d, paste0(wd_path, "processed_data/outlying_vars_per_clust.csv") )

# # LOOK AT CLUSTERS WITH HC ----
# if(FALSE){
#   # hc on pcs
#   clust_hc <- hclust(dist(reducedDim(sce, "PCA")), method = "ward.D2")
#   
#   # # Cut tree into clusters
#   # clusts <- cutree(clust_hc, k = 11)
#   clusts <- sce$km_nn_cluster[clust_hc$order]
#   
#   # Convert hclust to dendrogram
#   dend <- as.dendrogram(clust_hc)
#   
#   # Define cluster colors
#   
#   # color_pal <- setNames(palKelly()[seq_along(unique(clusts))], unique(clusts))
#   color_pal <- cluster_cols
#   cols <- color_pal[as.character(clusts)][order.dendrogram(dend)]
#   
#   # # color branches
#   # dend <- color_branches(dend, col = cols)
# 
#   # Remove labels from the dendrogram
#   labels(dend) <- NULL  
#   
#   # Create an empty matrix (1 row, column names match clustering order)
#   empty_mat <- matrix(NA, nrow = 1, ncol = length(clusts))
#   colnames(empty_mat) <- seq_along(clusts)
#   
#   # Create annotation for clusters (ribbon)
#   ha <- HeatmapAnnotation(
#     Cluster = factor(clusts, levels = unique(clusts)), 
#     col = list(Cluster = color_pal),
#     which = "column",
#     show_legend = FALSE,
#     annotation_height = unit(10, "cm")
#   )
#   
#   # Create a legend manually
#   legend_list <- list(
#     Legend(
#       labels = names(color_pal),
#       labels_gp = gpar(fontsize = 20),
#       legend_gp = gpar(fill = color_pal),
#       title = "Cluster",
#       ncol = 1  # Ensures proper legend layout
#     )
#   )
#   
#   # Generate the heatmap with only the dendrogram and ribbon
#   pdf(paste0(wd_path, "results/clustering/09_clustering/PC_HC_tree.pdf"), height = 30, width = 100)
#   ht <- Heatmap(
#     empty_mat, 
#     bottom_annotation = ha, 
#     cluster_columns = dend, 
#     cluster_rows = FALSE,
#     show_column_dend = TRUE, 
#     # heatmap_height = unit(1, 'inch'),
#     show_heatmap_legend = FALSE, 
#     column_dend_side = "top",
#     column_dend_height = unit(25, 'inch'),
#     col = NA  , 
#     show_column_names = FALSE,
#     height = unit(0, "null")
#   )
#   
#   # Draw the plot with manual legend
#   draw(ht,
#        annotation_legend_side = "right", 
#        heatmap_legend_list = legend_list
#        )
#   dev.off()
# }


```

### new cluster labels hm
```{r}

# relabel clusters
xsce <- sce

## key
lookup <- tribble(
~my_cluster, ~new_cluster,
       3 ,   1,
       6 ,   2,
       4 ,   3,
       9 ,   4,
       1 ,   5,
       2 ,   6,
      12 ,   7,
      10 ,   8,
       11,   9,
        8,  10,
        5,  11,
        7,  12,
       13,  13
)

## assign new cluster based on ppmean
xsce$old_my_cluster <- xsce$my_cluster
xsce$my_cluster <- factor(lookup$new_cluster[match(as.character(xsce$my_cluster), as.character(lookup$my_cluster))])



agg_cell <- aggregateAcrossCells(xsce , use.assay.type = "pmm_imp_of_scaled", ids = xsce$my_cluster, statistics = "mean")

# scale aggregate assay
assay(agg_cell, "scaled") <- assay(agg_cell, "pmm_imp_of_scaled") %>% t() %>% scale() %>% t()

# Get cluster sizes
cluster_sizes <- table(xsce$my_cluster)

# Define column annotation with cluster sizes
col_anno <- HeatmapAnnotation(
  "Cluster Size" = anno_barplot(as.matrix(cluster_sizes),
                                gp = gpar(fill = "grey"),
                                border = TRUE, 
                                # width =unit(10, 'cm') ,
                                height = unit(10, 'cm')
                                )
)


# Define colors for each data category
category_colors <- setNames(RColorBrewer::brewer.pal(n = length(unique(rowData(xsce)$table_description)), "Set3"), unique(rowData(xsce)$table_description))
msngness_clust_colors <- setNames(RColorBrewer::brewer.pal(n = length(unique(rowData(xsce)$msngns_clust)), "Set3"), paste0("X",unique(rowData(xsce)$msngns_clust)))

row_anno <-  rowAnnotation(
  data_category = rowData(xsce)$table_description,
  col = list(
      data_category = category_colors
      )
  )

new_hm_m <- assay(agg_cell, "scaled")

# Define heatmap
set.seed(102)
new_hm <- Heatmap(
  new_hm_m,
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_title = "Cluster Profiles",
  row_title = "Features",
  bottom_annotation = col_anno,
  left_annotation = row_anno,
  heatmap_legend_param = list(
    title = "Scaled Feature\nMean",
    color_bar = "continuous"
  )
)


# Save plot
pdf(paste0(wd_path, "results/clustering/09_clustering/aggregated_new_my_clusters.pdf"), height = 20, width = 10)
set.seed(123)
draw(new_hm,padding = unit(c(5, 5, 5, 5), "mm"), ht_gap = unit(30, "mm"))
dev.off()
sprintf(
  "cp %s %s",
  paste0(wd_path, "results/clustering/09_clustering/aggregated_new_my_clusters.pdf"),
  paste0(wd_path, "results/publication_figures/aggregated_new_my_clusters.pdf")
) %>% 
  system()

```

# Trajectory analysis

## optimize TSNE and UMAP
### TSNE
```{r, eval=FALSE}

# diff sce to test on 
xsce <- sce
reducedDim(xsce, "TSNE") <- NULL
assays(xsce) <- list(pmm_imp_of_scaled = assays(xsce)$pmm_imp_of_scaled)

# permuted params to try
tsne_params <- expand.grid(KEEP.OUT.ATTRS = FALSE,
  dimred = c("PCA"),
  perplexity = c(5, 30, 50),  # Default: 30
  momentum = c(0.5, 0.9),  # Default: 0.5
  final_momentum = c(0.8, 0.9),  # Default: 0.8
  eta = c(100, 200, 500),  # Default: 200
  exaggeration_factor = c(4, 12, 20)  # Default: 12
) %>% as.data.frame()

if(!dir.exists(sprintf("%sresults/clustering/09_clustering/tsne_param_optim/",wd_path))) dir.create(sprintf("%sresults/clustering/09_clustering/tsne_param_optim/",wd_path))

# for each set of params, plot results parralelly ----
plots <- parallel::mclapply(seq(nrow(tsne_params)), function(i){

  # make params digestible by do.call
  params <- as.list(tsne_params[i, ])
  # add sce to list
  params <- append(params, list(x = xsce))


  seed = 29209
  set.seed(seed)
  xsce <- do.call(runTSNE, params)

  # MST
  colLabels(xsce) <- xsce$km_nn_cluster
  by.cluster <-  aggregateAcrossCells(xsce , use.assay.type = "pmm_imp_of_scaled", ids = xsce$label, statistics = "mean")

  ## PCA
  centroids_pca <- reducedDim(by.cluster, "PCA")

  # Set clusters=NULL as we have already aggregated above.
  mst_pca <- TSCAN::createClusterMST(centroids_pca, clusters=NULL)

  pca_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="PCA")

  mst_pca_cluster_plot <- plotPCA(xsce, colour_by="label", text_by = "label", text_colour = "black") +
      geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC2, group=edge))

  map.tscan_pca <- mapCellsToEdges(xsce, mst=mst_pca, use.dimred="PCA")
  tscan.pseudo_pca <- orderCells(map.tscan_pca, mst_pca)

  common.pseudo_pca <- averagePseudotime(tscan.pseudo_pca)

  mst_pca_PT_plot <- plotPCA(xsce, colour_by=I(common.pseudo_pca),
          text_by="label", text_colour="red") +
      geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC2, group=edge))

  ## TSNE
  tsne_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="TSNE")

  mst_tsne_cluster_plot <- plotTSNE(xsce, colour_by="label", text_by = "label", text_colour = "black") +
      geom_line(data=tsne_line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))+
      theme(aspect.ratio = 1)+
      ggtitle("TSNE attempt", paste0("seed ", seed,"; ", paste(names(params)[2:7], unlist(params[2:7]), collapse = "; ")))

  ggsave(sprintf("%sresults/clustering/09_clustering/tsne_param_optim/%d_tsne_mst_cluster.png",wd_path, i), mst_tsne_cluster_plot, height = 8, width = 9, bg = "white")

},mc.set.seed = FALSE, mc.cores = detectCores()-1)
#  dimred perplexity final_momentum momentum eta exaggeration_factor
# 1    PCA         50            0.9      0.5 100                   4
# ^^ best 



# try optimal params ----

# # seed = 29209
# # seed = 29200
# set.seed(1)
# seeds <- runif(20, min = 1, max = 1e6) %>% round()
# seeds
# 
# # 265509 372125 572854 908208 201683 898390 944675 660798 629114  61787 205975 176558 687023
# # [14] 384104 769842 497700 717619 991906 380036 777445
# # >>> best:: 205975, 629114, 944675, 201683, 572854
# 
# # plots <- mclapply(seeds, function(seed){
#     
#   set.seed(seed)
#   params <- list(x = xsce,dimred = structure(1L, levels = "PCA", class = "factor"), 
#   
#       # perplexity = 5, 
#       # perplexity = 5, 
#       perplexity = 30,
#       # momentum = 0.9, 
#       # momentum = 0.9,
#       momentum = 0.5,
#       # final_momentum = 0.8, 
#       # final_momentum = 0.8,
#       final_momentum = 0.8,
#       # eta = 100,
#       # eta = 100,
#       eta = 200,
#       # exaggeration_factor = 20
#       # exaggeration_factor = 16
#       exaggeration_factor = 12
#       )
#   
  params <- list(x = xsce,
                dimred =  "PCA", 
                perplexity = 50, # 50,
                final_momentum = 0.9, # 0.9,
                momentum =0.5, #0.9,
                eta = 100,  #100,
                exaggeration_factor = 4 #12
  )
  
  set.seed(29209)
  xsce <- do.call(runTSNE, params)

  # $momentum
  # 0.9  0.9  0.9 0.9
  #
  # $final_momentum
  # 0.5  0.5  0.9 0.4
  #
  # $eta
  # 10  613.8  613 613
  #
  # $exaggeration_factor
  # 15  20  20 20


  # MST
  colLabels(xsce) <- xsce$km_nn_cluster
  by.cluster <-  aggregateAcrossCells(xsce , use.assay.type = "pmm_imp_of_scaled", ids = xsce$label, statistics = "mean")

  ## PCA
  centroids_pca <- reducedDim(by.cluster, "PCA")

  # Set clusters=NULL as we have already aggregated above.
  mst_pca <- TSCAN::createClusterMST(centroids_pca, clusters=NULL)

  pca_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="PCA")

  mst_pca_cluster_plot <- plotPCA(xsce, colour_by="label", text_by = "label", text_colour = "black") +
      geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC2, group=edge))

  map.tscan_pca <- mapCellsToEdges(xsce, mst=mst_pca, use.dimred="PCA")
  tscan.pseudo_pca <- orderCells(map.tscan_pca, mst_pca)

  common.pseudo_pca <- averagePseudotime(tscan.pseudo_pca)

  mst_pca_PT_plot <- plotPCA(xsce, colour_by=I(common.pseudo_pca),
          text_by="label", text_colour="red") +
      geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC2, group=edge))

  ## TSNE
  tsne_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="TSNE")

  mst_tsne_cluster_plot <- plotTSNE(xsce, colour_by="label", text_by = "label", text_colour = "black") +
      geom_line(data=tsne_line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))
  mst_tsne_cluster_plot +
    theme(aspect.ratio = 1)+
    ggtitle("TSNE attempt", paste0("seed ", seed,"; ", paste(names(params)[2:7], unlist(params[2:7]), collapse = "; ")))

# }, mc.cores = detectCores()-1, mc.set.seed = FALSE)
# 
# 
# grid <- (plots[[1]]+plots[[2]]+plots[[3]]+plots[[4]]+plots[[5]]+plots[[6]]+plots[[7]]+plots[[8]]+plots[[9]]+plots[[10]])/(plots[[11]]+plots[[12]]+plots[[13]]+plots[[14]]+plots[[15]]+plots[[16]]+plots[[17]]+plots[[18]]+plots[[19]]+plots[[20]])
# 
# ggsave(paste0(wd_path, "results/clustering/09_clustering/diff_seeds_tsnes_grid.pdf"),grid)




```


### UMAP

```{r, eval=FALSE}

# diff sce to test on 
xsce <- sce
reducedDim(xsce, "UMAP") <- NULL
assays(xsce) <- list(pmm_imp_of_scaled = assays(xsce)$pmm_imp_of_scaled)

# construct parameters df
umap_params <- expand.grid( dimred = c("PCA"), 
    metric = factor(c("euclidean", "cosine"), levels = c("euclidean", "cosine")), # "euclidean" (the default)"cosine""manhattan" "hamming"
    spread =  c(.8, 1, 1.2) ,# 1
    min_dist = c(.005, .01, .02),
    n_trees = c(25, 50, 75), 
    n_neighbors = c(10, 15, 20),# 0.01,
    KEEP.OUT.ATTRS = F
    ) %>%
  as.data.frame

umap_params$search_k =with(umap_params, 2 * n_neighbors * n_trees)


# make directory for plots
sprintf("%sresults/clustering/09_clustering/umap_params_optim/",wd_path) %>% 
  {if(!dir.exists(.)) dir.create(.)}


# for each set of params, plot results ----
how.long(T)
plots <- parallel::mclapply(seq(nrow(umap_params)), function(i){
  
  # make params digestible by do.call
  params <- as.list(umap_params[i, ])
  
  # add sce to list
  params <- append(params, list(x = xsce))
  
  seed = 29209
  set.seed(seed)
  xsce <- do.call(runUMAP, params)
  
  # MST
  colLabels(xsce) <- xsce$km_nn_cluster
  by.cluster <-  aggregateAcrossCells(xsce , use.assay.type = "pmm_imp_of_scaled", ids = xsce$label, statistics = "mean")
  
  ## PCA
  centroids_pca <- reducedDim(by.cluster, "PCA")
  
  # Set clusters=NULL as we have already aggregated above.
  mst_pca <- TSCAN::createClusterMST(centroids_pca, clusters=NULL)
  
  pca_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="PCA")
  
  mst_pca_cluster_plot <- plotPCA(xsce, colour_by="label", text_by = "label", text_colour = "black") +
      geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC2, group=edge))
  
  map.tscan_pca <- mapCellsToEdges(xsce, mst=mst_pca, use.dimred="PCA")
  tscan.pseudo_pca <- orderCells(map.tscan_pca, mst_pca)
  
  common.pseudo_pca <- averagePseudotime(tscan.pseudo_pca)
  
  ## UMAP
  umap_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="UMAP")
  
  mst_umap_cluster_plot <- plotUMAP(xsce, colour_by="label", text_by = "label", text_colour = "black") +
      geom_line(data=umap_line.data, mapping=aes(x=UMAP1, y=UMAP2, group=edge))+
      theme(aspect.ratio = 1)+
      ggtitle("UMAP attempt", paste0("seed ", seed,"; ", paste(names(params)[2:length(params)-1], unlist(params[2:length(params)-1]), collapse = "; ")) )
  
  ggsave(sprintf("%sresults/clustering/09_clustering/umap_params_optim/%d_umap_mst_cluster.png",wd_path, i), mst_umap_cluster_plot, height = 8, width = 9, bg = "white")

}, mc.set.seed = FALSE, mc.cores = detectCores()-1)
how.long() # ~ 5.5
#+ seed: 29209>>
#+ metric:2,2,2
#+ spread: 1,.8,.8
#+ min_dist: 0.02, 0.005, 0.005
#+ n_trees: 50, 25,25
#+ n_neighbors: 15, 20, 20
#+ search_k: 2*n_neighbors*n_trees
# BEST: Seed 29209; metric cosine; spread 1.2; min_dist 0.005; n_trees 25; n_neighbors 20

# try different seeds ----
set.seed(1)
seeds <- runif(20, min = 1, max = 1e6) %>% round()

other_umap_params <- data.frame(
  seed = seeds,
  dimred = "PCA", 
  metric = "cosine",
  spread = 0.8, 
  min_dist = 0.005,
  n_trees = 25,
  n_neighbors = 10,
  search_k = 500
  )

plots <- mclapply(seq(nrow(other_umap_params)), function(i){
  params <- as.list(other_umap_params[i,-which(names(other_umap_params)=="seed")])
  params <- append(params, list(x = xsce))
  
  set.seed(other_umap_params$seed[i])
  xsce <- do.call(runUMAP, params)

  # MST
  colLabels(xsce) <- xsce$km_nn_cluster
  by.cluster <-  aggregateAcrossCells(xsce , use.assay.type = "pmm_imp_of_scaled", ids = xsce$label, statistics = "mean")

  ## PCA
  centroids_pca <- reducedDim(by.cluster, "PCA")

  # Set clusters=NULL as we have already aggregated above.
  mst_pca <- TSCAN::createClusterMST(centroids_pca, clusters=NULL)

  pca_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="PCA")

  mst_pca_cluster_plot <- plotPCA(xsce, colour_by="label", text_by = "label", text_colour = "black") +
      geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC2, group=edge))

  map.tscan_pca <- mapCellsToEdges(xsce, mst=mst_pca, use.dimred="PCA")
  tscan.pseudo_pca <- orderCells(map.tscan_pca, mst_pca)

  common.pseudo_pca <- averagePseudotime(tscan.pseudo_pca)

  mst_pca_PT_plot <- plotPCA(xsce, colour_by=I(common.pseudo_pca),
          text_by="label", text_colour="red") +
      geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC2, group=edge))

  
  umap_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="UMAP")

  mst_umap_cluster_plot <- plotUMAP(xsce, colour_by="label", text_by = "label", text_colour = "black") +
      geom_line(data=umap_line.data, mapping=aes(x=UMAP1, y=UMAP2, group=edge))
  mst_umap_cluster_plot +
    theme(aspect.ratio = 1)+
    ggtitle("UMAP attempt",sprintf("seed = %d", other_umap_params$seed[i]))

}, mc.cores = detectCores(), mc.set.seed = FALSE)
length(plots)
grid <-wrap_plots(plots, ncol = 5, nrow = 4)

ggsave(paste0(wd_path, "results/clustering/09_clustering/diff_seeds_umaps_grid.png"),grid,height = 15, width = 20)


# Try favorite seeds ----
other_umap_params <- data.frame(
  seed = c(29209, 794240),
  dimred = "PCA", 
  metric = "cosine",
  spread = 0.8, 
  min_dist = 0.005,
  n_trees = 25,
  n_neighbors = 10,
  search_k = 500
  )


plots <- mclapply(seq(nrow(other_umap_params)), function(i){
  params <- as.list(other_umap_params[i,-which(names(other_umap_params)=="seed")])
  params <- append(params, list(x = xsce))
  
  set.seed(other_umap_params$seed[i])
  xsce <- do.call(runUMAP, params)

  # MST
  colLabels(xsce) <- xsce$km_nn_cluster
  by.cluster <-  aggregateAcrossCells(xsce , use.assay.type = "pmm_imp_of_scaled", ids = xsce$label, statistics = "mean")

  ## PCA
  centroids_pca <- reducedDim(by.cluster, "PCA")

  # Set clusters=NULL as we have already aggregated above.
  mst_pca <- TSCAN::createClusterMST(centroids_pca, clusters=NULL)

  pca_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="PCA")

  mst_pca_cluster_plot <- plotPCA(xsce, colour_by="label", text_by = "label", text_colour = "black") +
      geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC2, group=edge))

  map.tscan_pca <- mapCellsToEdges(xsce, mst=mst_pca, use.dimred="PCA")
  tscan.pseudo_pca <- orderCells(map.tscan_pca, mst_pca)

  common.pseudo_pca <- averagePseudotime(tscan.pseudo_pca)
  
  umap_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="UMAP")

  mst_umap_cluster_plot <- plotUMAP(xsce, colour_by="label", text_by = "label", text_colour = "black") +
    geom_line(data=umap_line.data, mapping=aes(x=UMAP1, y=UMAP2, group=edge))+
    theme(aspect.ratio = 1)+
    ggtitle("UMAP attempt",sprintf("seed = %d", other_umap_params$seed[i]))

  
  map.tscan_umap <- mapCellsToEdges(xsce, mst=mst_pca, use.dimred="PCA")
  tscan.pseudo_umap <- orderCells(map.tscan_umap, mst_pca)
  
  common.pseudo_umap <- averagePseudotime(tscan.pseudo_umap)
  
  mst_umap_PT_plot <- plotUMAP(xsce, colour_by=I(common.pseudo_umap),text_by="label", text_colour="red") +
      geom_line(data=umap_line.data, mapping=aes(x=UMAP1, y=UMAP2, group=edge))+
      theme(aspect.ratio = 1)+
      ggtitle("UMAP attempt",sprintf("seed = %d", other_umap_params$seed[i]))

    
  mst_umap_PT_plot+mst_umap_cluster_plot
   
}, mc.cores = detectCores(), mc.set.seed = FALSE)

wrap_plots(plots, ncol = 1)
# ^ 29209 best 

```

### manage umap/tsne plots
     
     with parameters of 'param_df_name', cluster parameters with hc, move plots into folder by cluster
     
```{r, eval = FALSE}

# param df to cluster
param_df_name <- 'umap_params'
param_df <- get(param_df_name)


# add par clust to tsne params
par_dist <- dist(scale(data.matrix(param_df[,-1])))
hc <- hclust(par_dist, method = "ward.D2")

# Plot the dendrogram with colored labels
plot(hc, labels = FALSE)
text(x = seq_along(hc$order), 
     y = rep(-0.5, length(hc$order)), 
     labels = hc$order, 
     col = ifelse(seq(nrow(param_df))[hc$order]%in%c(60
                                                     ), "red", "black"), 
     cex = 0.7, 
     pos = 1,
     offset =1.5, 
     srt = 90
      )
clusts <- cutree(hc, k = 11)

param_df$par_clust <- clusts
write.csv(param_df, sprintf("%sresults/clustering/09_clustering/%s.csv",wd_path, param_df_name))


# Heatmap of param clusters
param_df %>% 
  select(-dimred) %>% 
  data.matrix( ) %>% 
  as.data.frame() %>% 
  # group_by( par_clust) %>% 
  # reframe(across(everything(), mean)) %>% 
  # ungroup() %>%
  mutate(rows = 1:n()) %>%
  column_to_rownames("rows") %>% 
  scale() %>%
  t() %>% 
  Heatmap()

# select move images into directories by cluster
groups <- split(seq(nrow(param_df)), paste("clust",clusts, sep = ""))

source_dir <- sprintf("%sresults/clustering/09_clustering/%s_optim/",wd_path, param_df_name)
files <- list.files(source_dir) 



# for each clsuter, move all images to same directory
for(group_name in names(groups)){
  group <- groups[[group_name]]
  
  dest_dir <- sprintf("%sresults/clustering/09_clustering/%s_optim_%s/",wd_path,param_df_name, group_name)
  
  # make dir, clear if already there and make it , if not there, just make it
  dest_dir %>% 
    {if(dir.exists(.)){
        unlink(., recursive = TRUE)
        dir.create(.)
        } else{
          dir.create(.)
        }
      }
  
  files_to_move <- grep(paste(sprintf("^%d_", group), collapse = "|"), files, value = TRUE)

  # Construct full source and destination paths
  source_paths <- file.path(source_dir, files_to_move)
  dest_paths <- file.path(dest_dir, files_to_move)

  # Move the files
  file.copy(source_paths, dest_paths)
  
}
 
# momentum .9, perplexity 50, eta 100
# hi momentum, high perplexity, low eta
# 48, 51,54,57 62,63, 65

# rownames(tsne_params) <- seq(nrow(tsne_params)) 
# 
# tsne_params[, -c(1, 7)] %>%
# # tsne_params[c(48, 51,54,57, 62,63, 65), -c(1, 7)] %>% 
#   scale() %>%
#   # {x <- .; x[which(is.nan(x), arr.ind = T)] <- 0;x} %>%
#   t() %>% 
#   Heatmap( )

group <- c(19,22,58,60,27,111,110,140, 95,97,151)
group_name = "my_picks"

groups %>% 
    purrr::map(~{(group%in%. )%>% sum()})

umap_params[group, ]

```

## MST and sling curve
```{r}


# MST ------------
# colLabels(sce) <- sce$km_nn_cluster
colLabels(sce) <- sce$my_cluster
by.cluster <-  aggregateAcrossCells(sce , use.assay.type = "pmm_imp_of_scaled", ids = sce$label, statistics = "mean")

# get pca cluster centroids
centroids_pca <- reducedDim(by.cluster, "PCA")

# calculate MST; Set clusters=NULL as we have already aggregated above.
mst_pca <- TSCAN::createClusterMST(centroids_pca, clusters=NULL)

# calculate pseudotime by mapping cells to mst edges
map.tscan <- mapCellsToEdges(sce, mst=mst_pca, use.dimred="PCA")
tscan.pseudo <- orderCells(map.tscan, mst_pca)
sce$pseudoprog <- averagePseudotime(tscan.pseudo) 

## PCA ----
pca_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="PCA")

mst_pca_cluster_plot <- plotPCA(sce, colour_by="label", text_by = "label", text_colour = "black") + 
    geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC2, group=edge))+
  theme(aspect.ratio = 1)


mst_pca_PT_plot <- plotPCA(sce, colour_by='pseudoprog', 
        text_by="label", text_colour="red") +
    geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC2, group=edge))+
  theme(aspect.ratio = 1)

## TSNE ----
tsne_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="TSNE")

mst_tsne_cluster_plot <- plotTSNE(sce, colour_by="label", text_by = "label", text_colour = "black") + 
    geom_line(data=tsne_line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))+
  theme(aspect.ratio = 1)
mst_tsne_cluster_plot

mst_tsne_PT_plot <- 
  plotTSNE(sce, colour_by='pseudoprog',
  # plotTSNE(sce, colour_by="km_nn_cluster", 
        text_by="label", text_colour="red") +
    geom_line(data=tsne_line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))+
  theme(aspect.ratio = 1)


## UMAP ----
umap_line.data <- reportEdges(by.cluster, mst=mst_pca, clusters=NULL, use.dimred="UMAP")

mst_umap_cluster_plot <- plotUMAP(sce, colour_by="label", text_by = "label", text_colour = "black") +
    geom_line(data=umap_line.data, mapping=aes(x=UMAP1, y=UMAP2, group=edge))+
  theme(aspect.ratio = 1)

mst_umap_PT_plot <- plotUMAP(sce, colour_by="pseudoprog",
        text_by="label", text_colour="red") +
    geom_line(data=umap_line.data, mapping=aes(x=UMAP1, y=UMAP2, group=edge))+
  theme(aspect.ratio = 1)

# save line data to sce
metadata(sce) <- append(metadata(sce), list(trajectory_data = list(pca_line_data = pca_line.data, umap_line_data = umap_line.data, tsne_line_data = tsne_line.data)))



# save trajectory plots
(traj_plot <- (mst_pca_cluster_plot+mst_tsne_cluster_plot+mst_umap_cluster_plot)/ (mst_pca_PT_plot + mst_tsne_PT_plot+mst_umap_PT_plot))
# draw(hm)

ggsave(  paste0( wd_path, "results/clustering/09_clustering/trajectory_plots.png"), traj_plot, height = 10, width =  15, dpi = 300)

```


### Compare versions clusters to original 
```{r, eval = FALSE}

# original where ISOS_LOC in {0, 1(central), 2(juxtafoveal)}
old_sce <- readRDS("/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/analysis_versions/version013/processed_data/sce002_original_clustering.rds")

# current sce to filter
new_sce <- sce


# # look at counts in each cluster
ncol(sce)-ncol(old_sce)
# befr_tbl_old <- table(old_sce$km_nn_cluster)
# befr_tbl_new <- table(new_sce$km_nn_cluster)

# make samples more agreeable
cnames_new <- colnames(new_sce)
cnames_old <- colnames(old_sce)

nums_new <- str_extract_all(cnames_new, "[0-9]+(?:\\.[0-9]+)?$") %>%
  unlist() %>% 
  as.numeric() %>% 
  round(1)
nums_old <- str_extract_all(cnames_old, "[0-9]+(?:\\.[0-9]+)?$") %>%
  unlist() %>% 
  as.numeric() %>% 
  round(1)

cnames_new <-str_remove_all(cnames_new, "[0-9]+(?:\\.[0-9]+)?$") %>% 
  paste(., nums_new, sep = "")

cnames_old <-str_remove_all(cnames_old, "[0-9]+(?:\\.[0-9]+)?$") %>% 
  paste(., nums_old, sep = "")

new_sce$sample <- cnames_new
old_sce$sample <- cnames_old

plot.new();VennDiagram::venn.diagram(
  x = list(new = new_sce$sample, old =old_sce$sample ),
  filename = NULL,
  disable.logging = TRUE,
  output=TRUE,
  lwd = 2,
  # lty = 'blank',
  fill = c("#E69F00", "#56B4E9"),
  cex = .9,
  fontface = "italic",
  cat.dist = c(-0.05, -0.05)
  ) %>%
  grid.draw()

new_sce <- new_sce[, new_sce$sample %in% old_sce$sample]
old_sce <- old_sce[, old_sce$sample %in% new_sce$sample]



# aftr_tbl_old <- table(old_sce$km_nn_cluster)
# aftr_tbl_new <- table(new_sce$km_nn_cluster)
# 
# clust_counts_new <- cbind(befr_tbl_new, aftr_tbl_new)
# clust_counts_old <- cbind(befr_tbl_old, aftr_tbl_old)
# 
# clust_counts_old <- cbind(clust_counts_old, (clust_counts_old[,1] - clust_counts_old[, 2])/clust_counts_old[,1])
# clust_counts_new <- cbind(clust_counts_new, (clust_counts_new[,1] - =clust_counts_new[, 2])/clust_counts_new[,1])

old_centroids <-
  reducedDims(old_sce)$TSNE %>%
  cbind(cluster = old_sce$km_nn_cluster) %>%
  as.data.frame() %>%
  summarize(across(TSNE1:TSNE2, ~median(.)), .by = cluster) %>%
  arrange(cluster)  

new_centroids <- reducedDims(new_sce)$TSNE %>%
  cbind(cluster = new_sce$my_cluster) %>%
  as.data.frame() %>%
  summarize(across(TSNE1:TSNE2, ~median(.)), .by = cluster) %>%
  arrange(cluster)  

structure <- as.data.frame(data.table::fread(
  "from to
  1 2
  2 5
  2 3
  3 11
  3 7
  3 9
  2 4
  4 10
  4 8
  4 6"
))

# Join coordinates for the 'from' and 'to' clusters
old_line_data <- structure %>%
  left_join(old_centroids, by = c("from" = "cluster")) %>%
  rename_with(~c('from', 'to', 'from_TSNE1' ,'from_TSNE2')) %>%
  left_join(old_centroids, by = c("to" = "cluster")) %>%
  rename_with(~c('from', 'to', 'from_TSNE1' ,'from_TSNE2','to_TSNE1' ,'to_TSNE2' )) 

# color new TSNE by old clusters --> object_clusters
new_old <- plotTSNE(new_sce, color_by = I(old_sce$km_nn_cluster), text_by = I(old_sce$km_nn_cluster))+
  ggtitle("New TSNE with OG clusters")+
  theme(aspect.ratio = .9)+
  geom_line(data=tsne_line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))+
  geom_label(data = new_centroids, aes(x = TSNE1, y = TSNE2, label = cluster), colour = "black", fill = "white", alpha = .5)

new_new <- plotTSNE(new_sce, color_by = "km_nn_cluster", text_by = )+
  ggtitle("New TSNE with new clusters")+
  theme(aspect.ratio = .9)+
  geom_line(data=tsne_line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))+
  geom_label(data = new_centroids, aes(x = TSNE1, y = TSNE2, label = cluster), colour = "black", fill = "white", alpha = .5)

# color old TSNE by new clusters
old_new <- plotTSNE(old_sce, color_by = I(new_sce$km_nn_cluster), text_by = I(new_sce$km_nn_cluster))+
  ggtitle("OG TSNE with new clusters")+
  theme(aspect.ratio = .9)+
  geom_segment(data = old_line_data, aes(x = from_TSNE1, y = from_TSNE2, xend = to_TSNE1, yend = to_TSNE2))+
  geom_label(data = old_centroids, aes(x = TSNE1, y = TSNE2, label = cluster), colour = "black", fill = "white", alpha = .5)

old_old <- plotTSNE(old_sce, color_by = 'km_nn_cluster')+
  ggtitle("OG TSNE with OG clusters")+
  theme(aspect.ratio = .9)+
  geom_segment(data = old_line_data, aes(x = from_TSNE1, y = from_TSNE2, xend = to_TSNE1, yend = to_TSNE2))+
  geom_label(data = old_centroids, aes(x = TSNE1, y = TSNE2, label = cluster), colour = "black", fill = "white", alpha = .5)
  

(p <- (old_old+old_new)/(new_old+new_new))


ggsave(paste0(wd_path, 'results/clustering/09_clustering/old_versus_new_clustering.png'),
       p,
       height = 10, width =10)

```

### Fudge Pseudotime to not regress
```{r}


# grab vars to fix pt
pt_d <- colData(sce)[, c("sample", "patID", "patID_eye", "month", "pseudoprog")] %>% 
  as.data.frame()

# look to see how much regression is happening // save eyes that regress
pseudoprog_regressors <- pt_d %>% 
  arrange(month) %>% 
  group_by(patID_eye) %>% 
  filter(pseudoprog<lag(pseudoprog)) %>% 
  # ggplot(aes(month, pseudoprog))+ geom_line(aes(colour =patID_eye), show.legend = FALSE)
  pull(patID_eye) %>% 
  unique()


# for each eye, ordered by month, set current to last if current < last
pt_d <- pt_d %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  mutate(regr_fxd_pseudoprog = cummax_na(pseudoprog)) %>% 
  ungroup()
  

ggplot()+
  geom_line(data = pt_d%>% filter(patID_eye %in% pseudoprog_regressors),
            aes(month, pseudoprog, group = patID_eye), color = "red", show.legend = FALSE, alpha = .3,  linewidth = .3)+
  geom_line(data = pt_d%>% filter(patID_eye %in% pseudoprog_regressors),
            aes(month, regr_fxd_pseudoprog, group = patID_eye), color = "blue", alpha = .3, show.legend = FALSE, linewidth = .3)+
  ggtitle("Fixing Pseudoprogression Regression\nblack/blue = orig./fixed")+
  theme_bw()

# add fixed pseudoprogression back to sce if ordering matches
if(all(sce$sample == pt_d$sample)) sce$regr_fxd_pseudoprog <- pt_d$regr_fxd_pseudoprog else warning("Check that ordering matches!")


# plot original and fixed on TSNE
orig <- plotReducedDim(sce, "TSNE", colour_by = "pseudoprog")+ ggtitle("Pseudoprogression")
fixed <- plotReducedDim(sce, "TSNE", colour_by = "regr_fxd_pseudoprog")+ ggtitle("Regression fixed Pseudoprogression")

orig_fixed_grid <- marrangeGrob(sapply(list(orig,fixed), ggplotGrob), ncol = 2, nrow = 1)


# make directorey for plot
paste0(wd_path, "results/clustering/09_clustering/regression_fixing/") %>% 
  {if(!dir.exists(.)) dir.create(.)}

ggsave(paste0(wd_path, "results/clustering/09_clustering/regression_fixing/original_v_fixed_pseudotime_tsnes.png"),orig_fixed_grid, height = 7, width= 14)

```


## TSNE colored by chew grade
```{r}


chew_grade_tsne <- plotTSNE(sce[ , !is.na(sce$chew_grade)], colour_by = "chew_grade" ,text_by="label", text_colour="red")
# +  geom_line(data=tsne_line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))
ggsave(paste0(wd_path, "results/clustering/09_clustering/chew_grade_tsne.png"),chew_grade_tsne, height = 5, width = 7, bg = "white")

chew_grade_pca <- plotPCA(sce[ , !is.na(sce$chew_grade)], colour_by = "chew_grade",text_by="label", text_colour="red" )+  geom_line(data=pca_line.data, mapping=aes(x=PC1, y=PC1, group=edge))
ggsave(paste0(wd_path, "results/clustering/09_clustering/chew_grade_pca.png"),chew_grade_pca, height = 5, width = 7, bg = "white")

# divide pseudoprog into 7 scores like chew
sce$pseudoprog_septile <- cut(sce$pseudoprog, 
                                breaks = quantile(sce$pseudoprog, seq(0, 1, 1/7)),
                                labels = 0:(length(seq(0, 1, 1/6)) - 1),
                                include.lowest = T) %>% 
  as.character() %>% 
  as.numeric()
 
# plotTSNE(sce[ , !is.na(sce$chew_grade)], colour_by = "chew_grade")
# plotTSNE(sce[ , !is.na(sce$chew_grade)], colour_by = "pseudoprog")
# plotTSNE(sce[ , !is.na(sce$chew_grade)], colour_by = "pseudoprog_septile")

# plot chew and pt over time
pt_v_chew_d <- colData(sce)[, c('patID_eye','month','chew_grade','pseudoprog','pseudoprog_septile') ] %>% 
  as.data.frame() %>% 
  na.omit %>% 
  pivot_longer(cols = matches("grade|prog$"), names_to = "variable")

ggplot(pt_v_chew_d %>% filter(variable != "pseudoprog_septile"), aes(month, value, color = patID_eye))+
  geom_point(show.legend = FALSE)+
  # geom_line(show.legend =FALSE)+
  geom_smooth(method = "lm", show.legend = FALSE)+
  geom_smooth(method = "lm", aes(group = 1), show.legend = FALSE)+
  facet_wrap(~variable)+
  ggtitle("Chew Grade and Pseudoprogression change over time")+
  theme_bw()



# chi square test on chew and pseudoprog
chew_pt_tab <- table(chew_grade=sce$chew_grade, pseudoprog_septile=sce$pseudoprog_septile)
chew_pt_tab
chisq.test(chew_pt_tab )
# chisq.test(chew_pt_tab, simulate.p.value = TRUE )

cor(sce$chew_grade, sce$pseudoprog, use = "complete.obs") 


```

## correlate pseudoprog to vision
```{r, eval = FALSE}

cor(sce$pseudoprog, as.matrix(t(assays(sce[ grep("RW0", rownames(sce)) ,])$pmm_imp_of_scaled)), use = "complete.obs")
#   RW034      RW044      RW054      RW064      RW074      RW084      RW094
# [1,] -0.367122 -0.4272488 -0.4901791 -0.5384068 -0.5670037 -0.5838525 -0.5823807

```


# ______________________

## !!!!! REDO CLUSTER ORDERING AND PC1 !!!!!
```{r}

xsce <- sce

# invert PC1
## rdims
pcs <- reducedDim(xsce, "PCA")
pcs[,"PC1"] <- pcs[,"PC1"] * -1
reducedDim(xsce, "PCA") <- pcs

## tree
pc_tree <- metadata(xsce)$trajectory_data$pca_line_data
pc_tree$PC1 <- pc_tree$PC1 * -1
metadata(xsce)$trajectory_data$pca_line_data <- pc_tree

# rename clusters
## order of clusteres along pseudoprog
new_cluster_names <-  data.frame(colData(xsce))%>% 
  select(my_cluster, pseudoprog) %>% 
  summarize(mean_pp = mean(pseudoprog),median_pp = median(pseudoprog), .by = my_cluster) 
# my_cluster new_cluster
#          3           1
#          6           2
#          4           3
#          9           4
#          1           5
#          2           6
#         12           7
#         10           8
#          11          9
#           8         10
#           5         11
#           7         12
#          13         13
 
with(new_cluster_names,{
     plot(mean_pp,median_pp, type = "n")
     text(mean_pp,median_pp, labels = my_cluster)
     add_lines()
  })

new_cluster_names <- new_cluster_names[order(new_cluster_names$mean_pp),]
new_cluster_names$new_cluster <- seq(nrow(new_cluster_names))

## assign new cluster based on ppmean
xsce$old_my_cluster <- xsce$my_cluster
xsce$my_cluster <- factor(new_cluster_names$new_cluster[match(as.character(xsce$my_cluster), as.character(new_cluster_names$my_cluster))])

# ****** Function to replace each part of the edge ******
replace_edge_clusters <- function(edge_strs) {
  sapply(edge_strs, function(edge_str){
    parts <- strsplit(edge_str, "--")[[1]]
    new_parts <- sapply(parts, function(x) {
      new_val <- new_cluster_names$new_cluster[match(as.integer(x), new_cluster_names$my_cluster)]
      as.character(new_val)
    })
    paste(new_parts, collapse = "--")
  })
}

## replace in trees
metadata(xsce)$trajectory_data$pca_line_data$old_edge<- metadata(xsce)$trajectory_data$pca_line_data$edge
metadata(xsce)$trajectory_data$pca_line_data$edge <- replace_edge_clusters(metadata(xsce)$trajectory_data$pca_line_data$old_edge)

metadata(xsce)$trajectory_data$tsne_line_data$old_edge<- metadata(xsce)$trajectory_data$tsne_line_data$edge
metadata(xsce)$trajectory_data$tsne_line_data$edge <- replace_edge_clusters(metadata(xsce)$trajectory_data$tsne_line_data$old_edge)

```
# Movement between clusters
## table for each cluster of what next cluster is
for each cluster, filter for eyes that have another visit »
count of what next clusters divided by number of eyes in those each respective cluster » 
### ratio of source to destination (s/d)
```{r}

# sumarize counts that move between clusters ---------
# filter for eyes that exist in >1 cluster 
pat.d <- colData(xsce) %>% 
  as.data.frame() %>% 
  arrange(patID_eye, month) %>% 
  select(patID_eye, month, my_cluster) %>% 
  mutate(my_cluster = as.numeric(as.character(my_cluster))) %>% 
  group_by(patID_eye) %>% 
  filter(n_distinct(my_cluster) >1) %>%
  ungroup()

# n per cluster : number in each cluster that moves >> dynamicness?
move_i <- table(pat.d$my_cluster) %>%
  as.matrix()

# number that move out of each cluster >> 
move_from_i <- sapply(seq(n_distinct(xsce$my_cluster)), function(i){
  pat.d %>% 
    group_by(patID_eye) %>% 
    filter(any(my_cluster == i) ) %>% 
    ungroup() %>%
    filter(my_cluster!=i & lag(my_cluster) == i) %>% 
    nrow() 
})%>% 
    as.matrix()
  
# number that move to each cluster
move_to_i <- sapply(seq(n_distinct(xsce$my_cluster)), function(i) {
  pat.d %>% 
    group_by(patID_eye) %>% 
    arrange(month) %>%
    mutate(prev = lag(my_cluster), curr = my_cluster) %>%
    ungroup() %>%
    filter(curr == i & !is.na(prev) & prev != curr) %>%
    nrow()
}) %>% 
    as.matrix()

# ratio of exiting from i over going to i >> destinationness
to_over_from <-  move_to_i /move_from_i

# ratio of >> sourceness
from_over_to <- move_from_i  / move_to_i

# outflow - inflow (net flow) 
net_flow <- move_from_i - move_to_i

# add all movement m into one
move_summary_m <- data.frame(influx=move_to_i,
                        efflux=move_from_i,
                        net_flow = net_flow, 
                        influxivity=to_over_from,
                        effluxivity= from_over_to) %>% 
  as.matrix()

move_summary_m[is.infinite(move_summary_m)] <- 0

rownames(move_summary_m) <- as.character(seq(n_distinct(xsce$my_cluster)))

move_summary_hm <- draw(
  Heatmap(move_summary_m %>% scale(),# %>% t(), 
        name = "scaled count",
        cluster_columns = FALSE, 
        cluster_rows = FALSE, 
        column_title = "Movement summary", 
        col = viridis::plasma(100), 
        column_names_rot = 45, 
        # column_names_gp = gpar(fontsize = 1)
        # heatmap_width = unit(3, "inches")
        )
)
move_summary_hm

# plot move summary
png(paste0(wd_path, "results/clustering/09_clustering/move_summary.png"), height = 900, width = 500)
move_summary_hm
dev.off()

```


#### counts of eyes that move from i to each other cluster
how many eyes that come from x go to y, y', y''', etc.
```{r}


# empty schema
empty_col_m <- matrix(0, ncol = 1, nrow = n_distinct(xsce$my_cluster), dimnames = list(as.character(seq(n_distinct(xsce$my_cluster))), NULL))

# m to fill with raw counts of destinations from i
next_clusters_m_frequencies <- matrix(NA, nrow = n_distinct(xsce$my_cluster), ncol = n_distinct(xsce$my_cluster), 
                         dimnames = list(paste("to", as.character(seq(n_distinct(xsce$my_cluster))), sep = "."),paste("from", as.character(seq(n_distinct(xsce$my_cluster))), sep = "."))
                         )  

# fill m with counts for each destination that come from i
for (i in 1:ncol(next_clusters_m_frequencies)){
  # table of counts of each cluster that originate from cluster i  
  ## for each patID_eye, filter for eyes that exist in cluster i at some point, then grab cluster of eye whose previous cluster is i
  from_i_to_other_freqs <- pat.d %>%
    group_by(patID_eye) %>% 
    arrange(month) %>% 
    filter(any(my_cluster == i) ) %>% 
    filter(lag(my_cluster) == i) %>% # filter for where previous cluster == i
    ungroup() %>% 
    pull(my_cluster) %>% 
    table() %>% 
    as.matrix()
  
  # matrix to fill with frequencies
  from_i_to_other_freqs_new <- empty_col_m
  
  # fill matrix with all clusters with frequencies from from_i_to_other_freqs
  from_i_to_other_freqs_new[rownames(from_i_to_other_freqs) , ] <- from_i_to_other_freqs
  
  # fill next_clusters_m_frequencies with raw counts 
  next_clusters_m_frequencies[,i] <-  from_i_to_other_freqs_new
}
 
# next_clusters_m_frequencies %>% Heatmap(cluster_rows = FALSE,cluster_columns = FALSE, col = plasma(20))

# # make summary m
# summary <- cbind(next_clusters_m_frequencies, rowSum = rowSums(next_clusters_m_frequencies))
# summary <- rbind(summary, colSum = colSums(summary))

# transition matrix (-self transitions) should match move_to_i matrix
x <- next_clusters_m_frequencies
diag(x) <- 0

all(rowSums(x)==  move_to_i)
# ^ TRUE

```


#### [ N(c->c') / N(c') ]--------
what ratio of destination is count of eyes that move from clust i to each cluster?
of eyes that leave cluster where do they go?
```{r}

# m to fill with ratio of destination values
next_clusters_m_ratio_of_dest <-matrix(NA, nrow = n_distinct(xsce$my_cluster), ncol = n_distinct(xsce$my_cluster), 
                         dimnames = list(paste("to", as.character(1:n_distinct(xsce$my_cluster)), sep = "."),paste("from", as.character(1:n_distinct(xsce$my_cluster)), sep = "."))
                         )  
  
# calculate ratio of destination cluster that moves to clust i'
for(i in 1:ncol(next_clusters_m_frequencies)){
  
  # count of each destination of eyes that come from i
  from_i_to_other_freqs <- next_clusters_m_frequencies[,i]
  
  # of all that move to each cluster, what ratio comes from i
  ratio <- from_i_to_other_freqs / move_to_i
  
  next_clusters_m_ratio_of_dest[, i] <- ratio
  
  }  

# set missing to 0
next_clusters_m_ratio_of_dest[is.na(next_clusters_m_ratio_of_dest)] <- 0

# remove self movement
diag(next_clusters_m_ratio_of_dest) <- 0

# remove row that don't make sense, i.e. where no movement should be going to
next_clusters_m_ratio_of_dest <- next_clusters_m_ratio_of_dest[-c(3),]

hm <- Heatmap(next_clusters_m_ratio_of_dest, 
        name = "ratio",
        column_title = "N(c->c') / ∑N(any->c')\n(# that move from c to c'\ndivided by total number that move to c')",
        cluster_rows = FALSE,
        cluster_columns = FALSE, 
        col = viridis::plasma(100), 
        column_names_gp = gpar(fontsize = 25),
        row_names_gp = gpar(fontsize = 25)
        )
draw(hm)

png(paste0(wd_path, "results/clustering/09_clustering/from_i_ratio_of_destination.png"), height = 1000, width = 1000)
draw(hm)
dev.off()

ratio_of_dest_tidy_transitions <- next_clusters_m_ratio_of_dest %>%
  as.data.frame() %>%
  tibble::rownames_to_column("to") %>%
  pivot_longer(
    cols = starts_with("from."),
    names_to = "from",
    values_to = "ratio"
  ) %>%
  mutate(
    from = as.integer(gsub("from\\.", "", from)),
    to = as.integer(gsub("to\\.", "", to))
  ) %>% 
  mutate(type = "Of eyes entering cluster, where do they come from?" )

ratio_of_dest_tile <- ggplot(ratio_of_dest_tidy_transitions %>% drop_na(), aes(factor(from), factor(to), fill = ratio))+
  geom_tile()+
  scale_fill_viridis_c(option = "plasma")+
  xlab("Origin Cluster")+
  ylab("Destination Cluster")+
  ggtitle("Of eyes that move to each cluster, how many come from each cluster")

```


#### [ N(c->c') / N(c) ] --------
what ratio of source is count of eyes that move from clust i to each cluster?
```{r}

# m to fill with ratios of source
next_clusters_m_ratio_of_source <-matrix(NA, nrow = n_distinct(xsce$my_cluster), ncol = n_distinct(xsce$my_cluster), 
                         dimnames = list(paste("to", as.character(1:n_distinct(xsce$my_cluster)), sep = "."),paste("from", as.character(1:n_distinct(xsce$my_cluster)), sep = "."))
                         )  

# fill m with ratios of source outgoing count
for (i in 1:ncol(next_clusters_m_ratio_of_source)){
  
  # of eyes in clust i, what is their next, non-i cluster
  from_i_to_other_freqs <- pat.d %>%
    group_by(patID_eye) %>% 
    arrange(month) %>% 
    filter(any(my_cluster == i) ) %>% 
    filter(my_cluster != i & lag(my_cluster) == i) %>% 
    # filter(month> max(month[my_cluster == i])) %>% 
    # filter(row_number() == 1) %>% 
    ungroup() %>% 
    pull(my_cluster) %>% 
    table() %>% 
    as.matrix()
  
  from_i_to_other_freqs_new <- empty_col_m

  # make right dimensions
  from_i_to_other_freqs_new[as.numeric(rownames(from_i_to_other_freqs)) , ] <- from_i_to_other_freqs
  
  # ratio of destination cluster that moves to clust i'
  ratio <- from_i_to_other_freqs_new / move_from_i
  
  next_clusters_m_ratio_of_source[, i] <- ratio
  
}  

next_clusters_m_ratio_of_source[is.na(next_clusters_m_ratio_of_source)] <- 0

diag(next_clusters_m_ratio_of_source) <- 0

# remove columns that dont' make sense; terminal clusteres shouldnt have movement out 
next_clusters_m_ratio_of_source <- next_clusters_m_ratio_of_source[,-c(5,7:8,10:13)]


hm <- Heatmap(next_clusters_m_ratio_of_source, 
        name = "ratio",
        column_title = "N(c->c') / ∑N(c->any)\n(#that move from c to c'\ndivided by total number that move from c)",
        cluster_rows = FALSE,
        cluster_columns = FALSE, 
        col = viridis::plasma(100),
        column_names_gp = gpar(fontsize = 25),
        row_names_gp = gpar(fontsize = 25)
        )

draw(hm)

png(paste0(wd_path, "results/clustering/09_clustering/from_i_ratio_of_source.png"), height = 1000, width = 1000)
draw(hm)
dev.off()

ratio_of_source_tidy_transitions <- next_clusters_m_ratio_of_source %>%
  as.data.frame() %>%
  tibble::rownames_to_column("to") %>%
  pivot_longer(
    cols = starts_with("from."),
    names_to = "from",
    values_to = "ratio"
  ) %>%
  mutate(
    from = as.integer(gsub("from\\.", "", from)),
    to = as.integer(gsub("to\\.", "", to))
  ) %>% 
  mutate(type = "Figure A: Of eyes leaving cluster, where do they go?" )

ratio_of_source_tile <- ggplot(ratio_of_source_tidy_transitions %>% drop_na(), aes(factor(from), factor(to), fill = ratio))+
  geom_tile()+
  scale_fill_viridis_c(option = "plasma")+
  xlab("Origin Cluster")+
  ylab("Destination Cluster")+
  ggtitle("Figure B: Of eyes that move out of each cluster, how many go to each cluster")

both <- rbind(ratio_of_source_tidy_transitions,ratio_of_dest_tidy_transitions)

ggplot(both %>% drop_na(), aes(factor(from), factor(to), fill = ratio))+
  geom_tile()+
  scale_fill_viridis_c(option = "plasma")+
  xlab("Origin Cluster")+
  ylab("Destination Cluster")+
  facet_wrap(~type,columns = 1, scales = "free")+
  theme_bw()+
  theme(aspect.ratio = 1,
        panel.grid = element_blank(),
        panel.background = element_rect(fill = "transparent", color = NA),
        plot.background = element_rect(fill = "transparent", color = NA),
        legend.background = element_rect(fill = "transparent", color = NA),
        legend.box.background = element_rect(fill = "transparent", color = NA)
        )+
  ggtitle("Eye Transitions")

ggsave(paste0(wd_path, "results/clustering/09_clustering/eye_transitions_tile.pdf"), height = 8,width = 15, bg = NULL)

```


### make text of transition percentages
```{r}

# redo clusters
z <- both

z <- z %>% 
  mutate(across(c(from, to) , ~., .names= "{.col}_old")) %>% 
  mutate(across(c(from, to) , 
                ~case_when(
                   . == 3 ~ 1,
                   . == 6 ~ 2,
                   . == 4 ~ 3,
                   . == 9 ~ 4,
                   . == 1 ~ 5,
                   . == 2 ~ 6,
                   . == 12 ~ 7,
                   . == 10 ~ 8,
                   . == 11~ 9,
                   . == 8~ 10,
                   . == 5~ 11,
                   . == 7~ 12,
                   . == 13~ 13,
                   TRUE ~ NA
                )
                ))


x <- z[z$type==unique(z$type)[1] & z$ratio>.2,]
x <- x[order(x$from, decreasing = F) , ]
txt <- with(x, 
       sprintf("\nOf eyes leaving %s %0.2f go to %s\n" , from,ratio, to)
)

x <- z[z$type==unique(z$type)[2] & z$ratio>.2,]
x <- x[order(x$to, decreasing = F) , ]

txt <- c(txt, 
         with(x, 
       sprintf("\nOf eyes entering %s %0.2f come from %s\n" , to,ratio, from)
))

cat(txt)


```



# reduced dim sequential plots
## making sequential tsne plots
```{r}

tsne_line_dat <- metadata(xsce)$trajectory_data$tsne_line_data

# dimensions to matach OG plot AR
h=6
w=9

# plain bw
plotReducedDim(xsce, "TSNE")+
   theme(aspect.ratio = 1)
ggsave(paste0(wd_path, "results/clustering/09_clustering/plain_tsne.png"), height = h, width = w, bg = "black")

# plain bw with axes
plotReducedDim(xsce, "TSNE")+
   theme(aspect.ratio = 1,
         axis.line = element_line(color = "white"),
         axis.text = element_text(color = "white"), 
         axis.title = element_text(color = "white"),
         axis.ticks = element_line(color = "white"))
ggsave(paste0(wd_path, "results/clustering/09_clustering/plain_tsne_w_axes.pdf"), height = h, width = w, bg = "black")

# color with cluster labs
p <- plotReducedDim(xsce, "TSNE", color_by = "my_cluster", text_by = "my_cluster", text_colour = "yellow")+
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1, 
        )

# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/color_tsne.png"), height = h, width = w, bg = "black")

# color and lines
p <- plotTSNE(xsce, colour_by="my_cluster", text_by = "my_cluster", text_colour = "yellow") + 
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1)+
  geom_line(data=tsne_line_dat, mapping=aes(x=TSNE1, y=TSNE2, group=edge), col = "white", linewidth = 1.5)

# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/lines_and_color_tsne.png"), height = h, width = w, bg = "black")

# PT
p <- plotTSNE(xsce, colour_by="pseudoprog", text_by = "my_cluster", text_colour = "red") +
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1)

# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/pseudoprog_tsne.png"), height = h, width = w, bg = "black", dpi = 200)# PT

p <- plotTSNE(xsce, colour_by="pseudoprog", text_by = "my_cluster", text_colour = "red") +
  ggtitle("TSNE colored by pseudoprog")+
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1,
        title = element_text(color = "white"))
# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/titled_pseudoprog_tsne.png"), height = h, width = w, bg = "black", dpi = 200)


# PT with lines
p <- plotTSNE(xsce, colour_by="pseudoprog", text_by = "my_cluster", text_colour = "red") + 
  theme(legend.background = element_rect(fill = "white", color = "white") )+
  geom_line(data=tsne_line_dat, mapping=aes(x=TSNE1, y=TSNE2, group=edge), col = "white")+
  theme(aspect.ratio = 1)

# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)


ggsave(paste0(wd_path, "results/clustering/09_clustering/pseudoprog_mst_tsne.png"), height = h, width = w, bg = "black", dpi = 200)

# tsne colored by chew grading score
p <- plotTSNE(xsce[ , !is.na(xsce$chew_grade)], colour_by="chew_grade", text_by = "my_cluster", text_colour = "red") +
  ggtitle("TSNE colored by Chew grading score")+
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1, 
        title = element_text(color = "white"))
# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/chew_grade_tsne.png"), height = h, width = w, bg = "black", dpi = 200)

# tsne colored by ACUIT
p <- plotTSNE(xsce[ , !is.na(xsce$ACUIT)], colour_by="ACUIT", text_by = "my_cluster", text_colour = "red") +
  ggtitle("TSNE colored by visual acuity")+
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1, 
        title = element_text(color = "white"))
# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/ACUIT_tsne.png"), height = h, width = w, bg = "black", dpi = 200)


```


## making sequential PCA plots

### plot
```{r}

pca_line_dat <- metadata(xsce)$trajectory_data$pca_line_data

# dimensions to matach OG plot AR
h=6
w=9

# plain bw
plotReducedDim(xsce, "PCA")+
   theme(aspect.ratio = 1)
ggsave(paste0(wd_path, "results/clustering/09_clustering/plain_pca.png"), height = h, width = w, bg = "black")

# plain bw with axes
plotReducedDim(xsce, "PCA")+
  geom_point(color = "grey", alpha = .5)+
   theme(aspect.ratio = 1,
         axis.line = element_line(color = "white",linewidth = 1.5),
         axis.ticks = element_line(color = "white", linewidth = 1.5, size = 2), 
         axis.text = element_text(color = "white"), 
         axis.title = element_text(color = "white")
         )

ggsave(paste0(wd_path, "results/clustering/09_clustering/plain_pca_w_axes.pdf"), height = h, width = w, bg = "black")

# color with cluster labs
p <- plotReducedDim(xsce, "PCA", color_by = "my_cluster", text_by = "my_cluster", text_colour = "yellow")+
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1,
         axis.line = element_line(color = "white",linewidth = 2),
         axis.ticks = element_line(color = "white", linewidth = 2, size = 2), 
         axis.text = element_text(color = "white"), 
         axis.title = element_text(color = "white")
         )

# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/color_pca.png"), height = h, width = w, bg = "black")


# color and lines
p <- plotPCA(xsce, colour_by="my_cluster", text_by = "my_cluster", text_colour = "yellow") + 
  geom_line(data=pca_line_dat, mapping=aes(x=PC1, y=PC2, group=edge), col = "white", linewidth = 1.5)+
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
         aspect.ratio = 1,
         axis.line = element_line(color = "white",linewidth = 2),
         axis.ticks = element_line(color = "white", linewidth = 2, size = 2), 
         axis.text = element_text(color = "white"), 
         axis.title = element_text(color = "white")
         )


# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/lines_and_color_pca.png"), height = h, width = w, bg = "black")

# PT
p <- plotPCA(xsce, colour_by="pseudoprog", text_by = "my_cluster", text_colour = "red") +
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1,
         axis.line = element_line(color = "white",linewidth = 1.5),
         axis.text = element_text(color = "white"), 
         axis.title = element_text(color = "white"),
         axis.ticks = element_line(color = "white",linewidth = 1.5))

# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/pseudoprog_pca.png"), height = h, width = w, bg = "black", dpi = 200)# PT

p <- plotPCA(xsce, colour_by="pseudoprog", text_by = "my_cluster", text_colour = "red") +
  ggtitle("PCA colored by pseudoprog")+
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1,
        title = element_text(color = "white"))

# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/titled_pseudoprog_pca.png"), height = h, width = w, bg = "black", dpi = 200)


# PT with lines
p <- plotPCA(xsce, colour_by="pseudoprog", text_by = "my_cluster", text_colour = "red") + 
  theme(legend.background = element_rect(fill = "white", color = "white") )+
  geom_line(data=pca_line_dat, mapping=aes(x=PC1, y=PC2, group=edge), col = "white")+
  theme(aspect.ratio = 1)
# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)


ggsave(paste0(wd_path, "results/clustering/09_clustering/pseudoprog_mst_pca.png"), height = h, width = w, bg = "black", dpi = 200)

# pca colored by chew grading score
p <- plotPCA(xsce[ , !is.na(xsce$chew_grade)], colour_by="chew_grade", text_by = "my_cluster", text_colour = "red") +
  ggtitle("PCA colored by Chew grading score")+
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1, 
        title = element_text(color = "white"))
# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/chew_grade_pca.png"), height = h, width = w, bg = "black", dpi = 200)

# pca colored by ACUIT
p <- plotPCA(xsce[ , !is.na(xsce$ACUIT)], colour_by="ACUIT", text_by = "my_cluster", text_colour = "red") +
  ggtitle("PCA colored by visual acuity")+
  theme(legend.background = element_rect(fill = "white"),
        legend.margin = ggplot2::margin(t = 4, r = 4, l = 4, b = 4), 
        aspect.ratio = 1, 
        title = element_text(color = "white"))
# Extract the data used in the plot
label_data <- ggplot_build(p)$data[[2]]  # Second layer is usually geom_text

# Rebuild plot without the geom_text layer
p$layers[[2]] <- NULL

# Add geom_label manually
p + geom_label(
  data = label_data,
  aes(x = x, y = y, label = label, color = label),
  color = "black",
  fill = "white",
  label.size = 0.2
)

ggsave(paste0(wd_path, "results/clustering/09_clustering/ACUIT_pca.png"), height = h, width = w, bg = "black", dpi = 200)


```



# = = = = = = = = = = = 
# save data
```{r}
# save SCE
saveHDF5SummarizedExperiment(sce, paste0(wd_path, "processed_data/sce002.h5"), replace = T)

```


```{r, look at dims}
check_dims <- function(D = d){
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}


as.data.frame(colData(sce)) %>% 
  left_join(
    as.data.frame(t(assays(sce)$scaled_measure_vars)) %>% 
      rownames_to_column("sample")
  ) %>% 
  check_dims()
# dims : 12525 143
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ added clustering variables and letter averages


### MARCH 24 2025 ###
# dims : 12276 174
# n_patIDs             3971
# n_patID_eyes         7888
# n_observations      12275
# n_patIDs_w_gr_1_vis   492

```
