---
title: "09.3_ClusteringRobustnessCheck"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---


# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

pacman::p_load(tidyr, dplyr, reshape2, purrr, GGally, tidyverse, ggpubr, vcd, ggplot2, gridExtra, ComplexHeatmap, viridis, mgcv, car, zoo, tools, mice, grDevices, ggrepel, grid, Metrics, stats, lme4, lmerTest, dendextend, umap, scater, SingleCellExperiment, scran, bluster, slingshot, ggfortify, pheatmap, TSCAN, igraph, knitr, qs, abind, HDF5Array, igraph, RColorBrewer, mclust, dbscan, parallel, future, magrittr, patchwork)

rm(list = ls())
gc()


```

# functions
```{r}

source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```


# Load Data

```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

sce <- HDF5Array::loadHDF5SummarizedExperiment(sprintf("%sprocessed_data/sce002.2.h5", wd_path))

```

# REDO CLUSTER ORDERING AND PC1 
```{r, eval = TRUE}
# invert PC1
## rdims
pcs <- reducedDim(sce, "PCA")
pcs[,"PC1"] <- pcs[,"PC1"] * -1
reducedDim(sce, "PCA") <- pcs
## tree
pc_tree <- metadata(sce)$trajectory_data$pca_line_data
pc_tree$PC1 <- pc_tree$PC1 * -1
metadata(sce)$trajectory_data$pca_line_data <- pc_tree

# rename clusters
## order of clusteres along pseudoprog
new_cluster_names <-  data.frame(colData(sce))%>% 
  select(my_cluster, pseudoprog) %>% 
  summarize(mean_pp = mean(pseudoprog),median_pp = median(pseudoprog), .by = my_cluster) 

with(new_cluster_names,{
     plot(mean_pp,median_pp, type = "n")
     text(mean_pp,median_pp, labels = my_cluster)
     add_lines()
  })

new_cluster_names <- new_cluster_names[order(new_cluster_names$mean_pp),]
new_cluster_names$new_cluster <- seq(nrow(new_cluster_names))

## assign new cluster based on ppmean
sce$old_my_cluster <- sce$my_cluster
sce$my_cluster <- factor(new_cluster_names$new_cluster[match(as.character(sce$my_cluster), as.character(new_cluster_names$my_cluster))])

# ****** Function to replace each part of the edge ******
replace_edge_clusters <- function(edge_strs) {
  sapply(edge_strs, function(edge_str){
    parts <- strsplit(edge_str, "--")[[1]]
    new_parts <- sapply(parts, function(x) {
      new_val <- new_cluster_names$new_cluster[match(as.integer(x), new_cluster_names$my_cluster)]
      as.character(new_val)
    })
    paste(new_parts, collapse = "--")
  })
}

## replace in trees
metadata(sce)$trajectory_data$pca_line_data$old_edge<- metadata(sce)$trajectory_data$pca_line_data$edge
metadata(sce)$trajectory_data$pca_line_data$edge <- replace_edge_clusters(metadata(sce)$trajectory_data$pca_line_data$old_edge)

metadata(sce)$trajectory_data$tsne_line_data$old_edge<- metadata(sce)$trajectory_data$tsne_line_data$edge
metadata(sce)$trajectory_data$tsne_line_data$edge <- replace_edge_clusters(metadata(sce)$trajectory_data$tsne_line_data$old_edge)

```

# ____________________   ____________________   ____________________   ____________________   ____________________

# EXPERIMENTAL trajectory analysis
## See how different clustering affects trajectories
cluster and calculate trajectories on different parameters
     
     using > or < 11 clusters changes trajectory?
     
```{r}
# ITERATE THROUGH DIFFERENT PARAMETERS{SEED,CENTERS,K}; SAVE CLUSTERS TO DF ----
iterate_through_params <- FALSE

## parameters to try; first row is original; make first row original params
params <- expand.grid(seed = c(1010,43933,29030), 
                     centers = c(1440, 500,1e3, 1.5e3 ), 
                     k = c(12,8,10, 1416), 
                     n_clusters = c(NA), 
                     clusters= c(NA)
)

# add iteration number
params$ith <-  1:nrow(params)

how.long(T);if(iterate_through_params){
  # for each set of parameters, recluster and save clustering
  param_results <-  mclapply(1:nrow(params), function(i){
    set.seed(params$seed[i])
    km_nn <- clusterCells(sce, use.dimred="PCA",
        BLUSPARAM=TwoStepParam(
            first=KmeansParam(centers= params$centers[i]),
            second=NNGraphParam(k=params$k[i])
        ), full = TRUE
    ) 
    return(list(n_clusters = n_distinct(km_nn$clusters), clusters = I(list(km_nn$clusters))))
  }, mc.cores = detectCores()-1)
    #+ fewer k centroids >> clusters
    #+ more knn centers < clusters
    
  # params$n_clusters[i] <- n_distinct(km_nn$clusters)
  # params$clusters[i] <- I(list(km_nn$clusters))
  
  params$n_clusters[1:length(param_results)]  <- sapply(param_results, `[[`, 1)
  params$clusters[1:length(param_results)]  <- sapply(param_results, `[[`, 2)
  
  # save params 
  # qsave(params, paste0(wd_path, "processed_data/diff_clustering_params_results_ISOS_LOC_CEN_JF.rds"))
  qsave(params, paste0(wd_path, "processed_data/diff_clustering_params_results_1.rds"))

};how.long() # ~1min w 48 params


# read in params filled with clusters
filled_params <- qread(paste0(wd_path, "processed_data/diff_clustering_params_results_1.rds"))

```


## ITERATE THROUGH CLUSTER ITERATIONS AND CALCULATE TREE 
```{r}

# for each iteration of filled_params, calculate tree and save to one tree df
how.long(T);
  line_data_pseudoprog_l <- mclapply(1:nrow(filled_params), function(ith){
    
    # add clusters as labels
    colLabels(sce) <- filled_params$clusters[[ith]]
    
    # find cluster means for each var
    a_by.cluster <-  aggregateAcrossCells(sce , use.assay.type = "pmm_imp_of_scaled", ids = sce$label )
    
    # PCA centroids
    a_centroids_pca <- reducedDim(a_by.cluster, "PCA")
    
    # make MST between clusters based on PCA cluster centroids
    a_mst_pca <- createClusterMST(a_centroids_pca, clusters=NULL)
    
    # edge dataframe with x,y
    a_line_data <- reportEdges(a_by.cluster, mst=a_mst_pca, clusters=NULL, use.dimred="TSNE")
    
    # map cells to nearest edge
    a_map.tscan_tsne <- mapCellsToEdges(sce, mst=a_mst_pca, use.dimred= "PCA")
    
    # order cells based on sequence of edges
    a_tscan.pseudo_tsne <- orderCells(a_map.tscan_tsne, a_mst_pca)
  
    # calculate pseudotime based on position in trajectory
    a_pseudoprog <- averagePseudotime(a_tscan.pseudo_tsne)
    
    # add var of what iteration
    a_line_data$ith <- ith
    
    # remove to save space
    rm("a_by.cluster")
    
    return(list(line_data = a_line_data, pseudoprog = a_pseudoprog))
  }, mc.cores = detectCores()-1);
how.long()# 2.3s(35parms); 5s(101params)

# add pseudotimes to filled_params
filled_params$pseudotime <- lapply(line_data_pseudoprog_l, `[[`, 2)

# make one line_data df; add n_clusters
line_data <- line_data_pseudoprog_l %>% 
  lapply(`[[`, 1) %>% 
  do.call(rbind, .) %>% 
  left_join(filled_params %>% select(n_clusters, ith) )

```

#### plot different trajectories and PPs
```{r}

# ****** FUNCTION : take a vector and normalize to lie between closed interval of 0 and 1
norm_0_1 <- function(vec, buffer = 0.01, rightclosed = TRUE, leftclosed = TRUE) {
  # Ensure the buffer is non-negative
  if (buffer < 0) stop("Buffer must be non-negative.")
  
  Min <- min(vec)
  Max <- max(vec)
  
  range_diff <- Max - Min
  adj <- (buffer * range_diff)
  
  # Adjust range to include buffer
  adj_min <- if (leftclosed) Min - adj else Min
  adj_max <- if (rightclosed) Max + adj else Max
  
  # Normalize
  normalized <- (vec - adj_min) / (adj_max - adj_min)
  
  # Ensure the output lies within [0, 1]
  normalized[normalized < 0] <- 0
  normalized[normalized > 1] <- 1
  
  return(normalized)
}
x <- sort(unique(line_data$n_clusters))
# x %>% 
#   inv() %>% 
#   norm_0_1(buffer = .3) 
table(line_data$n_clusters)

# # sample teh same amount of each cluster count
# c_inds <- sapply(as.character(unique(line_data$n_clusters)), function(c) which(line_data$n_clusters == c))
# c_inds_sample <- lapply(c_inds, function(v) sample(v, 10)) %>% unlist()
# line_data_sub <- line_data[c_inds_sample, ]
line_data$n_clusters <- as.numeric(line_data$n_clusters)
line_data$group = paste(line_data$edge, line_data$ith, sep= "_")

# plot all trees overlapping 
# trees <- 
  ggplot()+
  geom_point(data = reducedDim(sce, "TSNE"), 
             aes(TSNE1, TSNE2,
                 # color = as.numeric(original_clustering_sce$my_cluster)
             ), 
             color = "grey",
             alpha = .3)+
    # scale_color_manual(values = cluster_cols)+
  geom_line(data = line_data %>% filter(n_clusters>0)  , 
            aes(round(TSNE1,1), round(TSNE2, 1), group = group, 
                # color = sqrt(1/n_clusters)
                ),
            alpha = .1,
            linewidth = 2, linetype = 1)+
    scale_color_viridis_c(option = "plasma")+
  ggtitle(paste("MST trajectories over", max(filled_params$ith), "iterations"))+
  theme_bw()+
  theme(aspect.ratio = 1)
  
ggplot()+
  geom_density2d_filled(data = line_data, aes(TSNE1,TSNE2))+
  geom_point(data = reducedDim(sce, "TSNE"), aes(TSNE1, TSNE2), col = "grey", alpha = .3 , shape = 5)+
  ggtitle("Centroid density")+
  theme_bw()

# see how pseudotimes correlate across different clusterings
pseudotime_d <-
  filled_params %>% 
  select(pseudotime) %>% 
  t() %>% 
  sapply(unlist) %>% 
  as.data.frame() %>% 
  rename_with(~as.character(filled_params$ith)) %>% 
  rownames_to_column("sample")

# calculate stats on new PPs
pseudotime_d$rowSDs <- rowSds(as.matrix(pseudotime_d %>% select(where(is.numeric))))
pseudotime_d$rowVars <- rowVars(as.matrix(pseudotime_d %>% select(where(is.numeric))))
pseudotime_d$rowMeans <- rowMeans(as.matrix(pseudotime_d %>% select(where(is.numeric))))
pseudotime_d$rowMeds <- rowMedians(as.matrix(pseudotime_d %>% select(where(is.numeric))))

# grab tsnes and new PPs
tsne <- reducedDim(sce, "TSNE") %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  left_join(pseudotime_d)


# for each stat, plot TNSE colored by stat
plots <- lapply(c("rowMeans", "rowMeds","rowSDs", "rowVars"), function(stat){
  
  # limit <- if(stat != "rowSDs") c(0, max(as.matrix(select(pseudotime_d,matches("\\d"))) ) ) else c(min(tsne[, stat]), max(tsne[,stat]))
  plot <-  ggplot()+
    geom_point(data = tsne, aes(TSNE1, TSNE2, color = !!sym(stat)))+
    geom_density2d(data = line_data, aes(TSNE1,TSNE2), linewidth = 1, linetype = 1, col = "coral")+
    # scale_color_viridis_c(limits = limit)+
    scale_color_viridis_c()+
    theme_bw()+
    ggtitle("Centroid density with statistic of pseudotime across different iterations")+
    labs(color = stat)+
    theme(aspect.ratio = 1)
})

# save grid of stat plots
plot_grid <- (plots[[1]]+plots[[2]])/ ( plots[[3]]+plots[[4]])

sprintf("%sresults/clustering/09_clustering/robustness_test/",wd_path) %>% 
  {if(!dir.exists(.)){dir.create(.)}}

ggsave(paste0( wd_path, "results/clustering/09_clustering/robustness_test/centroid_dens_pp_stat_plots.png"),plot_grid,height = 10, width = 30 )


# plot the relationship between the original trajectory pp to all others (original is 1)
png(paste0( wd_path, "results/clustering/09_clustering/robustness_test/originalPP_versus_other_trajectories.png"), height = 2e3, width = 2e3)
par(mfrow = c(6,6))
for(i in 1:36 ){
  
  # note: tsne 1 is original clustering 
  pp <- tsne[,as.character(i)] 
  n_clusters <- filled_params$n_clusters[filled_params$ith == i]
  plot(tsne$`1`~pp, xlab = paste("iteration",i), main = paste(n_clusters, "clusters"), cex = .1, pch= 3, col = rgb(0,0,0,.4))
  
  }
par(mfrow = c(1,1))
dev.off()

```


## 'Average'  trajectories
### use kmeans to find centroid centroids and then calculate new MST
```{r}

######      ######      ######      
# GRAPH OF EYE TRAJECTORIES ----
######      ######      ######      

# longitud_edges <- as.data.frame(colData(manipulated_objects$longitudinal_obj$singleCellExperiment)) %>%
#   select(patID_eye, month, label) %>%
#   arrange(patID_eye , month) %>%
#   group_by(patID_eye) %>%
#   mutate(current_clust = as.numeric(label),
#          next_clust = as.numeric(lead(label))) %>%
#   drop_na(next_clust) %>%
#   ungroup()
# 
# 
# # make edges list
# edge_list <- longitud_edges %>%
#   select(current_clust, next_clust)
# 
# # # weighted adjacency matrix
# # longitud_adj <- as.matrix(as_adjacency_matrix(graph_from_edgelist(as.matrix(edge_list))))
# # my.map(longitud_adj)
# 
# # filter for distinct edges
# edge_list_filtered <- edge_list %>%
#   filter(current_clust!=next_clust) %>%
#   distinct() %>%
#   as.matrix()
# 
# # graph of distinct edges
# longitud_graph <- graph_from_edgelist(edge_list_filtered, directed = TRUE)
# 
# # plot graph of distinct edges
# plot(longitud_graph,
#      edge.arrow.size = .4,
#      edge.arrow.width = 1
#      )

######      ######      ######      
# PLOT ALL TREES ----

# # plot trees, excluding original
# ggplot()+
#   geom_point(data = reducedDim(sce, "TSNE"), aes(TSNE1, TSNE2, color = sce$my_cluster), alpha = .1)+
#   geom_line(data = line_data[line_data$ith != 1 , ] , 
#             aes(TSNE1, TSNE2, group = edge), col = 'black',  alpha = .4, linewidth = 3, linetype = 1)+
#   geom_line(data = manipulated_objects$longitudinal_obj$trajectoryLineData, aes(TSNE1,TSNE2, group = edge), col = "black", alpha = .3, linewidth = 3 )+
#   geom_line(data = manipulated_objects$singleton_obj$trajectoryLineData, aes(TSNE1,TSNE2, group = edge),    col = "black", alpha = .3, linewidth = 3 )+
#   ggtitle(paste("MST trajectories over", max(filled_params$ith), "iterations"))+
#   theme_bw()+
#   theme(aspect.ratio = 1)

# # combine tree data into one df; add necessary ids
# longitud_tree <- manipulated_objects$longitudinal_obj$trajectoryLineData %>% 
#   mutate (ith = max(line_data$ith)+1, n_clusters = n_distinct(TSNE1,TSNE2)) %>% 
#   mutate(type = "filtered")
# singleton_tree <- manipulated_objects$singleton_obj$trajectoryLineData %>%
#   mutate (ith = max(line_data$ith)+2, n_clusters = n_distinct(TSNE1,TSNE2)) %>%
#    mutate(type = "filtered")
# 
# # join all together; add identifier for what data was used to compute centroids
# all_trees <- line_data %>% 
#   mutate(type = "iterated") %>% 
#   rbind(longitud_tree) %>%
#   rbind(singleton_tree) %>%
#   mutate(edge2 = edge) %>% 
#   separate(edge2, into = c("node1", "node2"), sep = "--", convert = TRUE)

# look at just iterated line data
all_trees <- line_data %>% 
  mutate(type = "iterated") %>%
  mutate(edge2 = edge) %>%
  separate(edge2, into = c("node1", "node2"), sep = "--", convert = TRUE)

# all trees overlayed
ggplot()+
  geom_point(data = reducedDim(sce, "TSNE"), aes(TSNE1, TSNE2, color = sce$my_cluster), alpha = .3)+
  scale_color_manual(values = cluster_cols)+
  geom_line(data = all_trees, aes(TSNE1,TSNE2, group = group),color = "black", alpha = .3, linewidth = 1.5)+
  theme_bw()+
  ggtitle("MST of Multiple Clustering Iterations Overlayed on TSNE Clusters")+
  labs(color = "Cluster")+
  guides(color = guide_legend(override.aes = list(alpha = 1)))+
  theme(aspect.ratio = 1)
ggsave(paste0(wd_path, "results/publication_figures/mst_spider_web_overlay.pdf"),width = 10,height = 10, limitsize = TRUE)

######      ######      ######      ######      
# MAKE MST OF CLUSTER CENTROIDS KM CENTROIDS ----
######      ######      ######      ######      

# data for edge points of iteration centroids

## type of tree to plot MST on 
# tree_type = "filtered"
tree_type = "iterated"

## filter for tree type and exclude first iteration b/c original
points <- all_trees %>%
  # filter(ith != 1) %>%
  filter(type ==tree_type) %>%
  select(TSNE1, TSNE2) %>%
  mutate(id = row_number())

# kmeans on mst centroids
# seeds <- c()
# seed <- sample(1:1e6, 1)
# set.seed(seed)
# seeds <- c(seeds,seed)


# num_centers <- 8; set.seed(834) # or 642565 for 8 centroids
num_centers <- 11; set.seed(834) # or 642565 for 
# num_centers <- 8; set.seed(685349) # or 642565 for 8 centroids
# num_centers <- 7; set.seed(198323) # for 7 centroids

km_params <- data.frame(seeds = c(834, 685349, 198323, 1029, 9304, 2049, 2049, 2049, 2049, 2049), 
                        centers = c(11,8,7, 6, 12, 10,9, 15, 18, 20))
km_params <- km_params[order(km_params$centers), ]

km_plots <- lapply(1:nrow(km_params) , function(i){
  
  num_centers <- km_params$centers[i]
  
  
  set.seed(km_params$seeds[i])
  centroid_clusters <- kmeans(points[, c("TSNE1", "TSNE2")], num_centers)
  # centroid_clusters <- dbscan(dist(points[, c("TSNE1", "TSNE2")]), eps = 3,minPts = 30)
  
  # extract centroids data
  centers <- 
    centroid_clusters$centers %>%
    # cbind(points[, c("TSNE1", "TSNE2")], cluster = centroid_clusters$cluster )%>% 
    # summarize(across(TSNE1:TSNE2, ~mean(.)), .by = cluster) %>% 
    as.data.frame() %>% 
    rownames_to_column("id") %>% 
    mutate(id = as.numeric(id))
  
  # # plot km centroids over iterated centroids
  # ggplot()+
  #   geom_point(data = reducedDim(sce, "TSNE"), 
  #              aes(TSNE1, TSNE2),color = "grey", alpha = .1)+
  #   geom_density_2d(data = points, 
  #                   aes(TSNE1, TSNE2),color = "blue", alpha = .3)+
  #   geom_point(data = points, 
  #              aes(TSNE1, TSNE2, color = factor(centroid_clusters$cluster)),
  #              shape = 3, show.legend = FALSE)+
  #   geom_point(data = centers, aes(TSNE1, TSNE2), shape = 10, size = 4)+
  #   theme_bw()+
  #   theme(aspect.ratio = 1)
  
  # Calculate distance matrix 
  dist_matrix <- as.matrix(dist(centers))
  
  # Create a complete graph from the distance matrix
  g <- graph_from_adjacency_matrix(dist_matrix, mode = "undirected", weighted = TRUE)
  
  # Generate the Minimum Spanning Tree (MST)
  mst <- igraph::mst(g,algorithm = "prim")
  
  # Extract edges and format them for plotting
  edges <- igraph::as_data_frame(mst, what = "edges") %>% 
    mutate_all(~as.numeric(.))
  
  # Join the points data to get coordinates for each edge
  edges <- edges %>%
    left_join(centers, by = c("from" = "id")) %>%
    dplyr::rename(TSNE1_from = TSNE1, TSNE2_from = TSNE2)%>%
    left_join(centers, by = c("to" = "id")) %>%
    dplyr::rename(TSNE1_to = TSNE1, TSNE2_to = TSNE2)
  
  # Plot the kmcenters and the MST with branches
  plot <- ggplot() +
    geom_point(data = reducedDim(sce, "TSNE"), aes(TSNE1, TSNE2),color = "grey", alpha = .1)+
    geom_density_2d(data = all_trees %>%filter(ith != 1) %>%filter(type == tree_type), aes(x = TSNE1, y = TSNE2), alpha = .4 ) + # not colored
    # geom_point(data = all_trees %>%filter(ith != 1) %>%filter(type == tree_type), aes(x = TSNE1, y = TSNE2, color = factor(ith))) + # colored
    # geom_point(data = all_trees %>%filter(ith != 1) %>%filter(type == tree_type), aes(x = TSNE1, y = TSNE2)) + # not colored
    geom_point(data = points, aes(TSNE1, TSNE2, color = factor(centroid_clusters$cluster)))+
    # geom_point(data = all_trees %>% filter(ith == 1), aes(x = TSNE1, y = TSNE2), color = "black", size = 3, shape =4) +
    geom_segment(data = edges, aes(x = TSNE1_from, y = TSNE2_from, xend = TSNE1_to, yend = TSNE2_to), color = "darkblue") +
    geom_point(data = centers, aes(TSNE1, TSNE2), shape = 17, color = "coral", size = 6)+
    labs(title = paste("MST of", tree_type,"data centroids and their k-means centroids") ,
         color = "Cluster",
         x = "TSNE1",
         y = "TSNE2")+
    theme_bw()+
    theme(aspect.ratio = 1)+
    geom_blank()
  
  # ggsave(sprintf("%sresults/clustering/09_clustering/robustness_test/%s_km_centroids_of_centroids.png",wd_path,km_params$centers[i] ),plot, height = 10, width = 12)
  
  return(plot)
  
  })

p <- patchwork::wrap_plots(plotlist = km_plots[1:3], ncol = 3)

ggsave(sprintf("%sresults/clustering/09_clustering/robustness_test/km_centroids_of_centroids_grid.png",wd_path ),p,height = 8, width = 24)

ggsave(paste0(wd_path, "results/publication_figures/km_centroids_of_centroids_grid.pdf"), p,width = 21,height = 7, limitsize = TRUE)

```

#____________________


# Does the Bifurcation hold up when clustering with OG params and fewer points?
## FUNCTION: ORIG clustering/trajectory pipeline; + PCA and TSNE 
```{r}

tsne_params <- metadata(sce)$params$tsne
km_nn_params <- metadata(sce)$params$km_nn

# ****** FUNCTION : cluster and trajectory analysis to give original clustering/trajectory; reduce dimensions of specified assay
og_pipeline <- function(an_sce, assay_name, 
                        runPCA = FALSE, 
                        runTSNE = FALSE, 
                        tsne_final_momentum = tsne_params$final_momentum, 
                        tsne_momentum = tsne_params$momentum, 
                        tsne_eta = tsne_params$eta, 
                        tsne_exageration_factor = tsne_params$exaggeration_factor, 
                        tsne_perplexity = tsne_params$perplexity, 
                        tsne_theta = NULL, 
                        tnse_seed = tsne_params$seed,
                        clust_seed = km_nn_params$seed,
                        clust_centers = km_nn_params$centers,
                        clust_k = km_nn_params$k){
  
  how.long(T)
  # REDUCED DIMS ---
  if(runPCA){
    # clear any existing reduced dimensions
    reducedDim(an_sce, "PCA") <- NULL
    
    # calculate  PCs and TSNE
    an_sce <- runPCA(an_sce,exprs_values = assay_name )
  }
  if(runTSNE){
    # clear any existing reduced dimensions
    reducedDim(an_sce, "TSNE") <- NULL
    # dynamic argument list
    tsne_args <- list(an_sce, dimred = "PCA")
    if (!is.null(tsne_perplexity)) tsne_args$perplexity <- tsne_perplexity
    if (!is.null(tsne_theta)) tsne_args$theta <- tsne_theta
    
    # Run t-SNE with dynamically constructed arguments
    set.seed(tnse_seed)
    an_sce <- do.call(runTSNE, tsne_args)
  }
  
  # CLUSTERING ----------
  set.seed(clust_seed)
  km_nn <- clusterCells(an_sce, use.dimred="PCA",
      BLUSPARAM=TwoStepParam(
          first=KmeansParam(centers=clust_centers),
          second=NNGraphParam(k=clust_k)
      ), full = TRUE
  ) 
  
  # add clusters as labels
  colLabels(an_sce) <- km_nn$clusters
  
  # TRAJECTORY CALC ----
  # find cluster means for each var
  a_by.cluster <-  aggregateAcrossCells(an_sce , use.assay.type = assay_name, ids = an_sce$label )
  
  # PCA centroids
  a_centroids <- reducedDim(a_by.cluster, "PCA")
  
  # make MST between clusters based on PCA cluster centroids
  a_mst <- createClusterMST(a_centroids, clusters=NULL)
  
  # edge dataframe with x,y
  line_data_pca <- reportEdges(a_by.cluster, mst=a_mst, clusters=NULL, use.dimred="PCA")
  if(runTSNE){
    line_data_tsne <- reportEdges(a_by.cluster, mst=a_mst, clusters=NULL, use.dimred="TSNE")
  }else{
    line_data_tsne <-NULL
  }
  
  # map cells to nearest edge
  a_map.tscan <- mapCellsToEdges(an_sce, mst=a_mst, use.dimred= "PCA")
  
  # order cells based on sequence of edges
  a_tscan.pseudo <- orderCells(a_map.tscan, a_mst)

  # calculate pseudotime based on position in trajectory
  a_pseudoprog <- averagePseudotime(a_tscan.pseudo)
  
  # add pseudoprog to sce
  an_sce$pseudoprog <- a_pseudoprog
  
  how.long()
  
  return(list(singleCellExperiment = an_sce, 
              trajectoryLineData = list(pca = line_data_pca, tsne = line_data_tsne))
         )
  
}
# test_sce_pipeline <- og_pipeline(sce)
# test_sce <- test_sce_pipeline$singleCellExperiment
# plotReducedDim(test_sce, "TSNE", color_by = "pseudoprog")

```

## look at trajectory when filtering/manipulating assay
  
    switch to run following chunks
```{r, eval=TRUE}

run_manipulate_objects <- TRUE

```


#### FILTERING OUT EYES: SINGLETONS AND MULTIVISITS 
```{r, eval=TRUE}

# what eyes have >1 visit
longitudinal_eyes <- as.data.frame(colData(sce)) %>% 
  filter(n()>1, .by = patID_eye) %>% 
  pull(patID_eye)

# for single visits; first visits
singletons <-  as.data.frame(colData(sce)) %>% 
  select(patID_eye, sample, month) %>% 
  arrange(patID_eye,month) %>% 
  filter(row_number( ) == 1, .by = patID_eye) %>% 
  pull(sample)

# switch to run pipeline on singlton and longitudinal eyes

# run pipeline on longitudinal  eyes
if(run_manipulate_objects){
  # run pipeline
  longitudinal_obj =  og_pipeline(sce[, sce$patID_eye %in% longitudinal_eyes], assay_name = "pmm_imp_of_scaled", runPCA = TRUE)
  # save data  individually
  if(!dir.exists(paste0(wd_path, "processed_data/longitudinal_sce_objects/"))) dir.create(paste0(wd_path, "processed_data/longitudinal_sce_objects/"))
    qsave(longitudinal_obj$trajectoryLineData, paste0(wd_path, "processed_data/longitudinal_sce_objects/trajectoryLineData.rds"))
    saveHDF5SummarizedExperiment(longitudinal_obj$singleCellExperiment, paste0(wd_path, "processed_data/longitudinal_sce_objects/singleCellExperiment.h5"),replace=TRUE)
}else{  
  longitudinal_obj <- list(trajectoryLineData= qread(paste0(wd_path, "processed_data/longitudinal_sce_objects/trajectoryLineData.rds")),
                    singleCellExperiment = loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/longitudinal_sce_objects/singleCellExperiment.h5")))
  }
 # >> 22s

# run pipeline on singletons
if(run_manipulate_objects){
  # run pipeline
  singleton_obj = og_pipeline(sce[, sce$sample %in% singletons], assay_name = "pmm_imp_of_scaled", runPCA = TRUE)
  
  # save data  individually
  if(!dir.exists(paste0(wd_path, "processed_data/singleton_objects/"))) dir.create(paste0(wd_path, "processed_data/singleton_objects/"))
  qsave(singleton_obj$trajectoryLineData, paste0(wd_path, "processed_data/singleton_objects/trajectoryLineData.rds"))
  saveHDF5SummarizedExperiment(singleton_obj$singleCellExperiment, paste0(wd_path, "processed_data/singleton_objects/singleCellExperiment.h5"),replace=TRUE)
}else{  
  singleton_obj <- list(trajectoryLineData= qread(paste0(wd_path, "processed_data/singleton_objects/trajectoryLineData.rds")),
                    singleCellExperiment = loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/singleton_objects/singleCellExperiment.h5")))
  }
# >> 33s

```


#### RANDOMLY REMOVE N samples
```{r, eval=TRUE}

# number of observations to remove to remove from teh data
ratios_to_keep <- seq(.40,.90,.10)
ns <- floor(ncol(sce) * ratios_to_keep)

# make directory to save all of the objects
if(!dir.exists(paste0(wd_path, "processed_data/sampled_objects/"))) dir.create(paste0(wd_path, "processed_data/sampled_objects/"))

# for each partition of the data, sample random eyes and run pipeline on those
if(run_manipulate_objects){
  for(i in 1:length(ns)){
    # random samples: sample n amount of samples to keep in assay 
    set.seed(2038)
    random_eyes <- sample(sce$sample, ns[i], F) 
    ratio = ratios_to_keep[i]
    
    # run pipeline on sampled assay
      
      # run pipeline
      sampled_obj = og_pipeline(sce[, sce$sample %in% random_eyes], assay_name = "pmm_imp_of_scaled", runPCA = TRUE)
      
      # specific object directory by sample size
      dir_name <- paste0("processed_data/sampled_objects/",ratio,"_sampled_objects/")
      
      # save data  individually
      if(!dir.exists(paste0(wd_path, dir_name))) dir.create(paste0(wd_path, dir_name))
      
      qsave(sampled_obj$trajectoryLineData, paste0(wd_path, dir_name, "trajectoryLineData.rds"))
      saveHDF5SummarizedExperiment(sampled_obj$singleCellExperiment, paste0(wd_path, dir_name, "singleCellExperiment.h5"),replace=TRUE)
    
    print(i)
    
  } # >> 19+24+30+35+41+48 = 3.5 min
} 

rm("sampled_obj")

# read in sampled objects to list
files <- dir(paste0(wd_path, "processed_data/sampled_objects/"))

# file = "0.5_sampled_objects"

sampled_objects_list <- list()
for(file in files){
  
  # rename so no number at start of object
  new_name <- paste("sampled_obj", str_split(file, "_", simplify = TRUE)[[1]], sep = "_")
  
  a_sampled_object_list <- list(trajectoryLineData= qread(paste0(wd_path, "processed_data/sampled_objects/", file, "/trajectoryLineData.rds")),
                    singleCellExperiment = loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/sampled_objects/", file,"/singleCellExperiment.h5"))  )
  
  sampled_objects_list[[new_name]] <- a_sampled_object_list
}

```


#### ADD GAUSSIAN NOISE TO ASSAY
```{r, eval=TRUE}

# random noise
set.seed(2108)
noise <- matrix(  rnorm(length(assays(sce)$pmm_imp_of_scaled),mean = 0, sd = .5), 
  ncol = ncol(assays(sce)$pmm_imp_of_scaled), 
  nrow = nrow(assays(sce)$pmm_imp_of_scaled)
  )

dim(noise)
dim(assays(sce)$pmm_imp_of_scaled)


# add noise to object
assays(sce)$noisey <- as.matrix(assays(sce)$pmm_imp_of_scaled) + noise

# run pipeline on noisey assay
if(run_manipulate_objects){
  # run pipeline
  noisey_obj <-  og_pipeline(sce, assay_name = "noisey", runPCA = TRUE)
  
  # save data  individually
  if(!dir.exists(paste0(wd_path, "processed_data/noisey_sce_objects/"))) dir.create(paste0(wd_path, "processed_data/noisey_sce_objects/"))
  
  qsave(noisey_obj$trajectoryLineData, paste0(wd_path, "processed_data/noisey_sce_objects/trajectoryLineData.rds"))
  saveHDF5SummarizedExperiment(noisey_obj$singleCellExperiment, paste0(wd_path, "processed_data/noisey_sce_objects/singleCellExperiment.h5"),replace=TRUE)
  
}else{  
  noisey_obj <- list(trajectoryLineData= qread(paste0(wd_path, "processed_data/noisey_sce_objects/trajectoryLineData.rds")),
                    singleCellExperiment = loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/noisey_sce_objects/singleCellExperiment.h5"))
                     )
  }
 # >> 1min

```


#### REMOVE VA FROM ASSAY 
```{r, eval=TRUE}
## filter out va vars
structure_sce <- sce
structure_sce <- structure_sce[!rownames(structure_sce) %in% grep("^RW\\d.*$", rownames(structure_sce), value = TRUE) ,]

# run pipeline on structure assay
if(run_manipulate_objects){
  # run pipeline
  structure_obj <- og_pipeline(structure_sce, assay_name = "pmm_imp_of_scaled", runPCA = TRUE)
  
  # save data  individually
  if(!dir.exists(paste0(wd_path, "processed_data/structure_sce_objects/"))) dir.create(paste0(wd_path, "processed_data/structure_sce_objects/"))
  
  qsave(structure_obj$trajectoryLineData, paste0(wd_path, "processed_data/structure_sce_objects/trajectoryLineData.rds"))
  saveHDF5SummarizedExperiment(structure_obj$singleCellExperiment, paste0(wd_path, "processed_data/structure_sce_objects/singleCellExperiment.h5"),replace=TRUE)
  
}else{  
  structure_obj <- list(trajectoryLineData= qread(paste0(wd_path, "processed_data/structure_sce_objects/trajectoryLineData.rds")),
                    singleCellExperiment = loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/structure_sce_objects/singleCellExperiment.h5"))
                     )
  }
# >> 53 s

```


#### REMOVE STRUCTURE FEATURES FROM ASSAY 
<!-- ```{r, eval=FALSE} -->
<!-- -->
<!-- ## filter out va vars -->
<!-- VA_sce <- sce -->
<!-- VA_sce <- VA_sce[rownames(VA_sce) %in% grep("^RW\\d.*$", rownames(VA_sce), value = TRUE) ,] -->
<!-- -->
<!-- # run pipeline on structure assay -->
<!-- how.long(clear = TRUE) -->
<!-- if(run_VA_object){ -->
<!--   # run pipeline -->
<!--   VA_obj <- og_pipeline(VA_sce, assay_name = "pmm_imp_of_scaled", runPCA = TRUE) -->
<!-- -->
<!--   # save data  individually -->
<!--   if(!dir.exists(paste0(wd_path, "processed_data/VA_sce_objects/"))) dir.create(paste0(wd_path, "processed_data/VA_sce_objects/")) -->
<!-- -->
<!--   qsave(VA_obj$trajectoryLineData, paste0(wd_path, "processed_data/VA_sce_objects/trajectoryLineData.rds")) -->
<!--   saveHDF5SummarizedExperiment(VA_obj$singleCellExperiment, paste0(wd_path, "processed_data/VA_sce_objects/singleCellExperiment.h5"),replace=TRUE) -->
<!-- -->
<!-- }else{ -->
<!--   VA_obj <- list(trajectoryLineData= qread(paste0(wd_path, "processed_data/VA_sce_objects/trajectoryLineData.rds")), -->
<!--                     singleCellExperiment = loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/VA_sce_objects/singleCellExperiment.h5")) -->
<!--                      ) -->
<!--   } -->
<!-- how.long() # >> 1.6 min -->
<!-- ``` -->

#### PUT OBJECTS IN LIST 
```{r, eval=TRUE}

if(run_manipulate_objects){
  manipulated_objects <- list(
    original_obj = og_pipeline(sce, assay_name = "pmm_imp_of_scaled", runPCA = FALSE),
    longitudinal_obj = longitudinal_obj,
    singleton_obj =singleton_obj,
    noisey_obj = noisey_obj,
    sampled_obj_0.4 = sampled_objects_list$sampled_obj_0.4, 
    sampled_obj_0.5 = sampled_objects_list$sampled_obj_0.5, 
    sampled_obj_0.6 = sampled_objects_list$sampled_obj_0.6, 
    sampled_obj_0.7 = sampled_objects_list$sampled_obj_0.7, 
    sampled_obj_0.8 = sampled_objects_list$sampled_obj_0.8, 
    sampled_obj_0.9 = sampled_objects_list$sampled_obj_0.9, 
    structure_obj= structure_obj
  )
  
  # save  
  qsave(manipulated_objects, sprintf("%sprocessed_data/clustering_robustness_manipulated_objects_list.qs", wd_path))
  
  # remove this because in the manipulated objects list
  rm_objects <- c("sampled_objects_list", "structure_sce", "VA_sce", "noisey_obj", "structure_obj", "VA_obj", "singleton_obj", "longitudinal_obj")
  rm(list = rm_objects[rm_objects%in%ls()])
  
  
}else{
  manipulated_objects <- qread(sprintf("%sprocessed_data/clustering_robustness_manipulated_objects_list.qs", wd_path))
}

```


<!-- #### function to scale 60% plot to make hidden edge visible -->
<!-- ```{r, eval = FALSE} -->

<!-- # grab TSNE and cluster data for sample_data_.6 -->
<!-- eye_filter_name <- "60% of data" -->
<!-- tsne <- reducedDim(manipulated_objects$sampled_obj_0.6$singleCellExperiment,"TSNE")%>%  -->
<!--   as.data.frame() -->
<!-- tsne$my_cluster = colLabels(manipulated_objects$sampled_obj_0.6$singleCellExperiment) -->


<!-- # manipulate TNSE1 to make an edge more visible -->
<!-- # NULL function -->
<!-- scale_func <- function(x) return(x) -->


<!-- # # piecewise transformation stretches space where lessthn 5 by rational root -->
<!-- # scale_func <- function(x) { -->
<!-- #   # where <- abs(x)<5 -->
<!-- #   where <- rep(TRUE, length(x)) -->
<!-- #   x[where] <- sign(x[where]) * abs(x[where])^(1/1.3) -->
<!-- #   return(x) -->
<!-- # } -->
<!-- #  -->
<!-- # # hyperbolic smoothing -->
<!-- # scale_func <- function(x){ -->
<!-- #   # where <- abs(x)<1 -->
<!-- #   # x[where] <- tanh(x[where] / max(abs(x[where]))) * max(abs(x[where])) -->
<!-- #   # p <- 2 -->
<!-- #   # x <- (x^p) * sign(x)[sign(x)<0] -->
<!-- #   x <- tanh(x) -->
<!-- #   return(x) -->
<!-- #   } -->
<!-- # plot(c(-10:10), scale_func(-10:10)) -->

<!-- # logarithmic transformation stretches near zero p amount; p>1 less stretching, p<1, more stretching -->
<!-- scale_func <- function(x, p = 1.5) { -->
<!--   x <- (log(1+abs(x)))^p * sign(x) -->
<!--   return(x) -->
<!-- } -->

<!-- # Test funciton -->
<!-- par(mfrow = c(2,5)); for (p in seq(.1,2,.2)){ -->
<!--   plot(c(-10:10), scale_func(-10:10, p = p), main = sprintf("p=%g", p)) -->
<!--   }; par(mfrow = c(1,1)) -->

<!-- # plot TSNE -->
<!-- plot_60_perc <- ggplot(tsne, aes(scale_func(TSNE1),TSNE2, color = factor(my_cluster)))+ -->
<!--                 theme_bw()+ -->
<!--                 geom_point()+ -->
<!--                 scale_color_manual(values = cluster_cols)+ -->
<!--                 geom_line(data = manipulated_objects$sampled_obj_0.6$trajectoryLineData ,  -->
<!--                         aes(scale_func(TSNE1), TSNE2, group = edge), col = 'black')+ -->
<!--                 ggtitle(paste("Trajectory and Pseudoprogression with", eye_filter_name))+ -->
<!--                 labs(color = "my_cluster")+ -->
<!--                 theme(aspect.ratio = 1,  -->
<!--                       title = element_text(face = "bold", size = 12))+ -->
<!--                 geom_blank()   -->


<!-- # tsne_x <- tsne -->
<!-- # tsne_x[, c("TSNE1","TSNE2")]  <- rotate_points(as.matrix(tsne_x[, c("TSNE1","TSNE2")]) , 270)  -->
<!-- # tsne_x[, c("TSNE1","TSNE2")] <-mirror(tsne_x[, c("TSNE1","TSNE2")], over = "y") -->
<!-- # tsne_x[, c("TSNE1","TSNE2")] <-mirror(tsne_x[, c("TSNE1","TSNE2")], over = "x") -->
<!-- #  -->
<!-- # ggplot(tsne_x , aes(scale_func(TSNE1),TSNE2, color = factor(my_cluster)))+ -->
<!-- #                 theme_bw()+ -->
<!-- #                 geom_point()+ -->
<!-- #                 scale_color_manual(values = cluster_cols)+ -->
<!-- #                 # geom_line(data = manipulated_objects$sampled_obj_0.6$trajectoryLineData ,  -->
<!-- #                 #         aes(scale_func(TSNE1), TSNE2, group = edge), col = 'black')+ -->
<!-- #                 ggtitle(paste("Trajectory and Pseudoprogression with", eye_filter_name))+ -->
<!-- #                 labs(color = "my_cluster")+ -->
<!-- #                 theme(aspect.ratio = 1,  -->
<!-- #                       title = element_text(face = "bold", size = 12))+ -->
<!-- #                 geom_blank()   -->

<!-- ``` -->

### plot TSNES
```{r, eval = TRUE}
# plot main titles
eye_filter_names <- data.frame(name = names(manipulated_objects)[-which(names(manipulated_objects)=='structure_obj')],
                          title = c("original data", "longitudinal eyes", "BL visits", "data with gaussian noise", "40% of data", "50% of data", "60% of data", "70% of data", "80% of data", "90% of data"))
                                    # , "only structural features"))

eye_filter_names$subheading = rep("coloring with original clusters", nrow(eye_filter_names))


# angles to rotate plots with
# angles <- c(0,100, 0,0,170, -180, -180,100, -10, -190)
angles <- c(0,rep(0,9))
pc1_multiples <- c(1,rep(-1,9))
pc2_multiples <- c(1,rep(-1,5), 1,rep(-1,3))
# angles <- c(0,rep(180,9))

filtering_trajectory_plots <- list()
for(i in 1:nrow(eye_filter_names)){

    eye_filter_name <- eye_filter_names[i,2]
    subheading <- eye_filter_names[i,3]

    # object list
    a_manipulated_object <- manipulated_objects[[i]]

    # grab pca and tree data; add cluster to pca
    a_pca <- as.data.frame(reducedDim(a_manipulated_object$singleCellExperiment, "PCA") )
    
    # # color with manipulate data clusters
    # a_pca$my_cluster <- colLabels(a_manipulated_object$singleCellExperiment)
    
    # color with original clusters
    a_pca$my_cluster <- sce$my_cluster[match(rownames(a_pca), colnames(sce))]
    
    # grab tree
    a_tree <- a_manipulated_object$trajectoryLineData$pca
    

    # grab centroids for labels
    cluster_labels <- a_pca %>% 
      group_by(my_cluster) %>% 
      summarize_all(median)
      
    
    # BESPOKE CHANGES TO PLOTS ----
    
    # # for 60% plot, stretch x axis to make a edge more visiible
    # if(eye_filter_name == "60% of data"){
    #   a_pca[, c("PCA1","PCA2")] <-rotate_points(a_pca[, c("PCA1","PCA2")] , -120)
    #   a_tree[, c("PCA1","PCA2")] <-rotate_points(a_tree[, c("PCA1","PCA2")] , -120)
    #   cluster_labels[, c("PCA1","PCA2")] <-rotate_points(cluster_labels[, c("PCA1","PCA2")] , -120)
    # }

    
    # # for gaussian noise plot, rotate -90
    # if(eye_filter_name %in% c("data with gaussian noise", "80% of data")){
    # 
    #   # rotate points
    #   a_pca[, c("PCA1","PCA2")] <- a_pca[, c("PCA1","PCA2")] %>%
    #     mirror("y") %>%
    #     rotate_points(45)
    # 
    #   # rotate tree
    #   a_tree[, c("PCA1","PCA2")] <- a_tree[, c("PCA1","PCA2")] %>%
    #     mirror("y") %>%
    #     rotate_points(45)
    # }
    
    # # for 40% plot
    # if(eye_filter_name %in% c("40% of data")){
    #   a_pca[, c("PCA1","PCA2")] <-rotate_points(a_pca[, c("PCA1","PCA2")] , 180)
    #   a_tree[, c("PCA1","PCA2")] <-rotate_points(a_tree[, c("PCA1","PCA2")] , 180)
    #   cluster_labels[, c("PCA1","PCA2")] <-rotate_points(cluster_labels[, c("PCA1","PCA2")] , 180)
    # }
    
    # # longitudinal plot
    # if(eye_filter_name %in% c("longitudinal eyes")){
    #   a_pca[, c("PCA1","PCA2")] <-a_pca[, c("PCA1","PCA2")] %>% 
    #     rotate_points(90)
    #   a_tree[, c("PCA1","PCA2")] <-a_tree[, c("PCA1","PCA2")] %>% 
    #     rotate_points(90)
    #   cluster_labels[, c("PCA1","PCA2")] <-cluster_labels[, c("PCA1","PCA2")] %>% 
    #     rotate_points(90)
    # }
    
    # # BL plot
    # if(eye_filter_name %in% c("BL visits")){
    #   a_pca[, c("PCA1","PCA2")] <-a_pca[, c("PCA1","PCA2")] %>% 
    #     rotate_points(-90) %>% 
    #     mirror(.,'y')  
    #   a_tree[, c("PCA1","PCA2")] <-a_tree[, c("PCA1","PCA2")] %>% 
    #     rotate_points(-90) %>% 
    #     mirror(.,'y') 
    #   cluster_labels[, c("PCA1","PCA2")] <-cluster_labels[, c("PCA1","PCA2")] %>% 
    #     rotate_points(-90) %>% 
    #     mirror(.,'y') 
    # }
    
    # using angles vector
    a_pca[, c("PC1","PC2")] <-a_pca[, c("PC1","PC2")] %>%
      rotate_points(.,angles[i]) %>%
      # mirror(.,'y') %>%
      {x <- .; x[,1] <- x[,1]*pc1_multiples[i]; x} %>%
      {x <- .; x[,2] <- x[,2]*pc2_multiples[i]; x} %>% 
      {.}
    
      
    a_tree[, c("PC1","PC2")] <-a_tree[, c("PC1","PC2")] %>%
      rotate_points(.,angles[i]) %>%
      # mirror(.,'y') %>%
      {x <- .; x[,1] <- x[,1]*pc1_multiples[i]; x} %>% 
      {x <- .; x[,2] <- x[,2]*pc2_multiples[i]; x} %>% 
      {.}

    cluster_labels[, c("PC1","PC2")] <-cluster_labels[, c("PC1","PC2")] %>%
      rotate_points(angles[i]) %>%
      # mirror(.,'y') %>%
      {x <- .; x[,1] <- x[,1]*pc1_multiples[i]; x} %>%
      {x <- .; x[,2] <- x[,2]*pc2_multiples[i]; x} %>% 
      {.}

    # plot PCA----
    plot <- ggplot(a_pca, aes(PC1,PC2, color = factor(my_cluster)))+
                theme_bw()+
                geom_point(show.legend =FALSE, alpha = .6)+
                # geom_point(show.legend = FALSE, alpha = .4)+
                scale_color_manual(values = cluster_cols)+
                geom_line(data = a_tree, aes(PC1, PC2, group = edge), col = 'black')+
                geom_label(data = cluster_labels, aes(label = my_cluster, x = PC1, y = PC2), col = "black", alpha = .5)+
                ggtitle(paste(eye_filter_name), subheading)+
                labs(color = "Cluster")+
                theme(aspect.ratio = 1,
                      title = element_text(face = "bold", size = 12))+
                geom_blank()
    
    # add plot to list
    filtering_trajectory_plots[[eye_filter_name]] <- plot
}
rm("a_manipulated_object")

legend_only <- ( ggplot(a_pca, aes(PC1,PC2 , color = factor(my_cluster)))+
  scale_color_manual(values = cluster_cols)+
  labs(color = "my_cluster")+
  guides(color = guide_legend(ncol = 2))+
  geom_point()+
  theme_void() )%>% 
  cowplot::get_legend() %>% 
  cowplot::plot_grid()


filtering_trajectory_plots <- append(filtering_trajectory_plots, list(legend = legend_only))


sub_filtering_trajectory_plots <- filtering_trajectory_plots[c(1:5, 7,9,11)]

# grid of plots
filtering_trajectory_plots_grid <- patchwork::wrap_plots(sub_filtering_trajectory_plots[-c(2,3)], ncol = 3)
filtering_trajectory_plots_grid

ggsave(paste0(wd_path, "results/clustering/09_clustering/robustness_test/manipulated_sces_cluster_tree_plots.png"), filtering_trajectory_plots_grid, width = 20, limitsize = TRUE)
ggsave(paste0(wd_path, "results/publication_figures/manipulated_sces_cluster_tree_plots.pdf"), filtering_trajectory_plots_grid,  width = 15,height = 15, limitsize = TRUE)

```



# = = = = = = = = = = = 
# NOT SAVING DATA
```{r}
# # save SCE
# # saveRDS(sce, paste0(wd_path, "processed_data/sce002_robust_test.rds"))
```


<!-- ```{r} -->
<!-- # Define number of points -->
<!-- n <- 100 -->

<!-- # Generate face outline (circle) -->
<!-- theta <- seq(0, 2*pi, length.out = n) -->
<!-- face_x <- cos(theta) -->
<!-- face_y <- sin(theta) -->

<!-- # Generate left eye (small circle) -->
<!-- eye_theta <- seq(0, 2*pi, length.out = n/2) -->
<!-- left_eye_x <- -0.4 + 0.1 * cos(eye_theta) -->
<!-- left_eye_y <-  0.5 + 0.1 * sin(eye_theta) -->

<!-- # Generate right eye (small circle) -->
<!-- right_eye_x <- 0.4 + 0.1 * cos(eye_theta) -->
<!-- right_eye_y <- 0.5 + 0.1 * sin(eye_theta) -->

<!-- # Generate mouth (smiling arc) -->
<!-- mouth_theta <- seq(0, pi, length.out = n/2) -->
<!-- mouth_x <- 0.5 * cos(mouth_theta) -->
<!-- mouth_y <- 0.5 * (sin(-mouth_theta)) -->

<!-- # Combine all points into a matrix -->
<!-- smiley_face_matrix <- rbind( -->
<!--   cbind(face_x, face_y), -->
<!--   cbind(left_eye_x, left_eye_y), -->
<!--   cbind(right_eye_x, right_eye_y), -->
<!--   cbind(mouth_x, mouth_y) -->
<!-- ) -->

<!-- plot(smiley_face_matrix); add_lines("xy") -->
<!-- plot(rotate_points(smiley_face_matrix, 90)); add_lines("xy") -->
<!-- plot(rotate_points(smiley_face_matrix, 180)); add_lines("xy") -->
<!-- plot(smiley_face_matrix %>% {(abs(.)^.7)*sign(.)}); add_lines("xy") -->

<!-- # Define parameters -->
<!-- num_circles <- 5  # Number of concentric circles -->
<!-- num_points <- 1e3  # Number of points per circle -->
<!-- radii <- seq(0.2, 1.0, length.out = num_circles)  # Radii of circles -->

<!-- # Initialize empty vectors for coordinates -->
<!-- x_coords <- c() -->
<!-- y_coords <- c() -->

<!-- # Generate points for each circle -->
<!-- for (r in radii) { -->
<!--   theta <- seq(0, 2*pi, length.out = num_points) -->
<!--   x_coords <- c(x_coords, r * cos(theta)) -->
<!--   y_coords <- c(y_coords, r * sin(theta)) -->
<!-- } -->

<!-- # Combine into a matrix -->
<!-- concentric_circles_matrix <- cbind(x_coords, y_coords) -->

<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.1)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.1)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.1)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.1)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.1)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.1)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.1)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.1)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.1)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.1)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.3)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.5)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^.9)*sign(.)}, type = "l", asp = 1) -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1)*sign(.)}, type = "l"  , asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.3)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.6)*sign(.)}, type = "l", asp = 1)  -->
<!-- plot(concentric_circles_matrix %>% {(abs(.)^1.9)*sign(.)}, type = "l", asp = 1)  -->

<!-- ``` -->

