---
title: "09.2_networkFlowValidation"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---
# packages
```{r}

if("package:raster" %in% search()) detach("package:raster", unload = TRUE, character.only = TRUE)

pacman::p_load(igraph, dplyr, tidyr, knitr, SingleCellExperiment, GenSA, ComplexHeatmap, viridis, HDF5Array) 
rm(list = ls()); gc()


# functions
source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```

# load data
```{r}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"
sce <- HDF5Array::loadHDF5SummarizedExperiment(sprintf("%sprocessed_data/sce002.2.h5", wd_path))

```
 
#__________________

# REDO CLUSTER ORDERING AND PC1 
```{r, eval = TRUE}

# invert PC1
## rdims
pcs <- reducedDim(sce, "PCA")
pcs[,"PC1"] <- pcs[,"PC1"] * -1
reducedDim(sce, "PCA") <- pcs
## tree
pc_tree <- metadata(sce)$trajectory_data$pca_line_data
pc_tree$PC1 <- pc_tree$PC1 * -1
metadata(sce)$trajectory_data$pca_line_data <- pc_tree

# rename clusters
## order of clusteres along pseudoprog
new_cluster_names <-  data.frame(colData(sce))%>% 
  dplyr::select(my_cluster, pseudoprog) %>% 
  summarize(mean_pp = mean(pseudoprog),median_pp = median(pseudoprog), .by = my_cluster) 

with(new_cluster_names,{
     plot(mean_pp,median_pp, type = "n")
     text(mean_pp,median_pp, labels = my_cluster)
     add_lines()
  })

new_cluster_names <- new_cluster_names[order(new_cluster_names$mean_pp),]
new_cluster_names$new_cluster <- seq(nrow(new_cluster_names))

## assign new cluster based on ppmean
sce$old_my_cluster <- sce$my_cluster
sce$my_cluster <- factor(new_cluster_names$new_cluster[match(as.character(sce$my_cluster), as.character(new_cluster_names$my_cluster))])

# ****** Function to replace each part of the edge ******
replace_edge_clusters <- function(edge_strs) {
  sapply(edge_strs, function(edge_str){
    parts <- strsplit(edge_str, "--")[[1]]
    new_parts <- sapply(parts, function(x) {
      new_val <- new_cluster_names$new_cluster[match(as.integer(x), new_cluster_names$my_cluster)]
      as.character(new_val)
    })
    paste(new_parts, collapse = "--")
  })
}

## replace in trees
metadata(sce)$trajectory_data$pca_line_data$old_edge<- metadata(sce)$trajectory_data$pca_line_data$edge
metadata(sce)$trajectory_data$pca_line_data$edge <- replace_edge_clusters(metadata(sce)$trajectory_data$pca_line_data$old_edge)

metadata(sce)$trajectory_data$tsne_line_data$old_edge<- metadata(sce)$trajectory_data$tsne_line_data$edge
metadata(sce)$trajectory_data$tsne_line_data$edge <- replace_edge_clusters(metadata(sce)$trajectory_data$tsne_line_data$old_edge)

```

## Aggregate end clusters and remove 7
make a root and two child node clustert scheme
```{r, eval = TRUE}



condense_clusters <- function(x){
  x <- as.character(x)
  new <- case_when(
    x %in% c("12", "13") ~ '6', 
    x %in% c("9", "8", "11", "10") ~ '5',
    TRUE ~ x
  )
  return(new)
}


sce$my_cluster <- condense_clusters(sce$my_cluster )

# remove 7
sce <- sce[ , sce$my_cluster!="7" ]

plotTSNE(sce,  color_by = 'my_cluster', text_by = 'my_cluster')

```


#### define number of clusters
```{r}
num_clusts <- n_distinct(sce$my_cluster)

```


# counts of eyes that move from i to each other cluster
```{r}

# sumarize counts that move between clusters ---------
# filter for eyes that exist in >1 cluster 
pat.d <- colData(sce) %>% 
  as.data.frame() %>% 
  arrange(patID_eye, month) %>% 
  dplyr::select(patID_eye, month, my_cluster) %>% 
  mutate(my_cluster = as.numeric(as.character(my_cluster))) %>% 
  group_by(patID_eye) %>% 
  filter(n_distinct(my_cluster) >1) %>%
  ungroup()

# how many eyes that come from x go to y, y', y''', etc.

# m to fill with raw counts of destinations from i
next_clusters_m_frequencies <- matrix(NA, nrow = num_clusts, ncol = num_clusts, 
                         dimnames = list(paste("to", as.character(seq(num_clusts)), sep = "."),paste("from", as.character(seq(num_clusts)), sep = "."))
                         )  
# empty matrix 
empty_col_m <- matrix(0, ncol = 1, nrow = num_clusts, dimnames = list(as.character(seq(num_clusts)), NULL))

# fill m with counts for each destination that come from i
for (i in 1:ncol(next_clusters_m_frequencies)){
  # table of counts of each cluster that originate from cluster i  
  ## for each patID_eye, filter for eyes that exist in cluster i at some point, then grab cluster of eye whose previous cluster is i
  from_i_to_other_freqs <- pat.d %>%
    group_by(patID_eye) %>% 
    arrange(month) %>% 
    filter(any(my_cluster == i) ) %>% # filter for eyes that exist in this cluster at some point
    # filter(my_cluster!=i & lag(my_cluster) == i) %>% # filter for where current not i and previous was i
    filter(lag(my_cluster) == i) %>% # filter for where previous cluster == i
    # filter(month> max(month[my_cluster == i])) %>% 
    # filter(row_number() == 1) %>% # filter out duplicate eyes; these might have gone 1-2...1-2 at diff time points
    ungroup() %>% 
    pull(my_cluster) %>% 
    table() %>% 
    as.matrix()
  
  # matrix to fill with frequencies
  from_i_to_other_freqs_new <- empty_col_m
  
  # fill matrix with all clusters with frequencies from from_i_to_other_freqs
  from_i_to_other_freqs_new[as.numeric(rownames(from_i_to_other_freqs)) , ] <- from_i_to_other_freqs
  
  # fill next_clusters_m_frequencies with raw counts 
  next_clusters_m_frequencies[,i] <-  from_i_to_other_freqs_new
}

```

#______
### generate theoretical "unordered" random tree structure
```{r}

rand_edges <- combn(seq(num_clusts), 2) %>%
  t()

rand_g <- graph_from_edgelist(rand_edges, directed = TRUE)

rand_g <- set_vertex_attr(rand_g, index = V(rand_g), name = "color", value =  cluster_cols[V(rand_g)])

set.seed(1019)
par(bg = "black", col.main = 'white')
plot.igraph(rand_g, vertex.color = V(rand_g)$color, vertex.size = 10, bg = "black", main = "Unordered")
par(bg = "white", col.main = 'black')

pdf(sprintf("%sresults/clustering/09_clustering/09.2.4/unordered_graph.pdf", wd_path) , height = 10, width = 10)
set.seed(1019)
# par(bg = "black", col.main = 'white')
plot.igraph(rand_g, vertex.color = V(rand_g)$color, vertex.size = 10, bg = "black", main = "Unordered")
par(bg = "white", col.main = 'black')
dev.off()

sprintf("cp %sresults/clustering/09_clustering/09.2.4/unordered_graph.pdf %sresults/publication_figures/unordered_graph.pdf", wd_path, wd_path) %>% 
  system()

rand_distance_matrix <- as.matrix(distances(rand_g))

```



# GENERATE OBSERVED BIFURCATED GRAPH
```{r}

## observed transition matrix
transition_matrix <- t(next_clusters_m_frequencies)

## set diagonal to 0 so no self transitions
diag(transition_matrix) <- 0

m <- transition_matrix
dimnames(m) <- purrr::map(dimnames(m), ~str_remove_all(., "to\\.|from\\."))
observed_graph <- graph_from_adjacency_matrix(m, weighted = TRUE)
observed_graph <- set_vertex_attr(observed_graph, "color", index = V(observed_graph), cluster_cols[V(observed_graph)] )


# par(bg = "black", col.main = 'white')
# plot.igraph(observed_graph, layout = layout_as_tree(observed_graph, root = 3), vertex.color = V(observed_graph)$color, vertex.size = 20, main = "Observed")
plot.igraph(observed_graph, vertex.color = V(observed_graph)$color, vertex.size = 20, main = "Observed")
par(bg = "white", col.main = 'black')


pdf(sprintf("%sresults/clustering/09_clustering/09.2.4/observed_graph.pdf", wd_path) , height = 10, width = 10)
# par(bg = "black", col.main = 'white')
# plot.igraph(observed_graph, layout = layout_as_tree(observed_graph, root = 3), vertex.color = V(observed_graph)$color, vertex.size = 20, main = "Observed")
plot.igraph(observed_graph, vertex.color = V(observed_graph)$color, vertex.size = 20, main = "Observed")
par(bg = "white", col.main = 'black')
dev.off()

sprintf("cp %sresults/clustering/09_clustering/09.2.4/observed_graph.pdf %sresults/publication_figures/observed_graph.pdf", wd_path, wd_path) %>% 
  system()


# ## probabilities of movement to cluster equals col normralized matrix, from matrix equals row normalized matrix
# transition_matrix_prob_m <- transition_matrix %>%
#   sweep(., 1, rowSums(.), "/")
#   # sweep(., 2, colSums(.), "/")
#   # sinkhorn_knopp()

# my.map(sqrt(transition_matrix_prob_m), title = "observed bifurcated transition probability matrix")
# my.map(transition_matrix, title = "observed bifurcated transition matrix")

# make graph
bif_edges <- metadata(sce)$trajectory_data$tsne_line_data %>%
  separate(edge, c("from", "to"), sep = '--') %>%
  dplyr::select(from,to) %>%
  
  {if (exists("condense_clusters", mode = "function")) {
      mutate(., from = condense_clusters(from),
                to   = condense_clusters(to))
    } else {
      .
    }} %>% 

  distinct() %>%
  mutate_all(~as.numeric(as.character(.))) 

# remove cluster 7
bif_edges <- bif_edges %>% 
  filter(if_all(everything(), ~ .!=7 ))

bif_edges <- as.matrix(bif_edges)

bif_g <-graph_from_edgelist(bif_edges, directed = FALSE)
bif_g <- set_vertex_attr(bif_g, "color", index = V(bif_g), cluster_cols[V(bif_g)] )


# par(bg = "black", col.main = 'white')
plot.igraph(bif_g,layout = layout_as_tree(bif_g, root = 1), vertex.color = V(bif_g)$color, vertex.size = 20, main = "Bifurcated")
par(bg = "white", col.main = 'black')

pdf(sprintf("%sresults/clustering/09_clustering/09.2.4/%s.pdf",wd_path, "bifurcated_graph"), height = 10,width = 10, bg = "white")
  # par(bg = "black", col.main = 'white')
  plot.igraph(bif_g,layout = layout_as_tree(bif_g, root = 1), vertex.color = V(bif_g)$color, vertex.size = 20, main = "Bifurcated")
  par(bg = "white", col.main = 'black')
dev.off()

sprintf("cp %s %s", 
        sprintf("%sresults/clustering/09_clustering/09.2.4/%s.pdf",wd_path, 'bifurcated_graph'),
        sprintf("%sresults/publication_figures/%s.pdf",wd_path, 'bifurcated_graph')
        ) %>% 
  system()


# distance matrix based on the tree structure >> 1-2-{3-7/11/9,4-6/8/10}
bif_distance_matrix <- distances(bif_g)

# adjacency matrix
bif_adjacency_matrix <- as.matrix(as_adjacency_matrix(bif_g))

```

<!-- ### Prune for stronger of two directions -->
<!-- ```{r} -->

<!-- m <- transition_matrix %>%  -->
<!--   # {.^2} %>% -->
<!--   # {.^(1/2)} %>% -->
<!--   # {sqrt(.)} %>% -->
<!--   {} -->

<!-- # # prune graph for stronger of two directions -->
<!-- m2 <- m -->

<!-- m2[lower.tri(m2)] <- pmax(0,m[lower.tri(m)] - m[upper.tri(m)]) -->
<!-- m2[upper.tri(m2)] <- pmax(0,m[upper.tri(m)]- m[lower.tri(m)]) -->

<!-- m <- m2 -->

<!-- dimnames(m) <- purrr::map(dimnames(m), ~str_remove_all(., "to\\.|from\\.")) -->

<!-- g <- graph_from_adjacency_matrix(m, weighted = TRUE, mode = "directed") -->
<!-- g <- set_vertex_attr(g, "color", index = V(g), cluster_cols[V(g)] ) -->

<!-- # par(bg = "black", col.main = 'white') -->
<!-- set.seed(102) -->
<!-- plot.igraph(g, layout = layout_with_fr(g), vertex.color = V(g)$color, vertex.size = 20, main = "Observed transitions pruned for stronger of two directions") -->
<!-- # plot.igraph(g, layout = layout_with_gem(g), vertex.color = V(g)$color, vertex.size = 20, main = "Observed") -->
<!-- # plot.igraph(g, layout = layout_with_kk(g), vertex.color = V(g)$color, vertex.size = 20, main = "Observed") -->
<!-- # plot.igraph(g, layout = layout_with_dh(g), vertex.color = V(g)$color, vertex.size = 20, main = "Observed") -->
<!-- # plot.igraph(g, layout = layout_on_grid(g), vertex.color = V(g)$color, vertex.size = 20, main = "Observed") -->
<!-- par(bg = "white", col.main = 'black') -->

<!-- ``` -->


# does observed data fit linear structure better than structured random? (hopefully not)
assume that the theoretical graph is linear, ordered by pseudoprog and test if random fits this better than the observed

### generate linear graph 
```{r}


# PROGRAMATICALLY EXTRACT GRAPH EDGES ----
#+ given ordering of a progression var, make edges
#+ 

# cluster progression info
cluster_info<- sce %>% 
  colData() %>% 
  as.data.frame() %>% 
  # dplyr::select(my_cluster, pseudoprog,chew_grade_derived_mine_loose, matches("chew") ) %>% 
  # dplyr::select(-chew_grade) %>% 
  dplyr::select(my_cluster, pseudoprog,chew_grade_derived_mine_loose) %>% 
  mutate(my_cluster = factor(my_cluster)) 

# summarize each cluster's chew and pseudoprog
cluster_stats <- cluster_info %>%
  group_by(my_cluster) %>%
  summarise(
    across(matches("pseudoprog|chew"), ~median(., na.rm = TRUE), .names = "{.col}_median"),
    .groups = "drop"  )

# group clusters whose stat is close
cluster_stats_grouped <- cluster_stats

stat_vars <- names(cluster_stats_grouped)[grepl("median$", names(cluster_stats_grouped))]

thrshld <- 1
for (var in stat_vars) {
  # Sort by the current variable
  sorted <- cluster_stats_grouped %>%
    dplyr::select(my_cluster, all_of(var)) %>%
    arrange(.data[[var]])

  # Compute group based on thresholded diffs
  grp <- cumsum(c(TRUE, diff(sorted[[var]]) >= thrshld))

  # Join group info back by cluster
  cluster_stats_grouped[[paste0(var, "_group")]] <- grp[match(cluster_stats_grouped$my_cluster, sorted$my_cluster)]
}

# plot each stats cluster ordering coloring for cluster groups
cluster_stats_grouped %>% 
  pivot_longer(cols = -c(my_cluster, ends_with("_group"))) %>%
  mutate(
    group = case_when(
      grepl("pseudoprog_median", name) ~ pseudoprog_median_group,
      # grepl("chew_grade_derived_mine_strict_median", name) ~ chew_grade_derived_mine_strict_median_group,
      # grepl("chew_grade_median", name) ~ chew_grade_median_group,
      grepl("chew_grade_derived_mine_loose_median", name) ~ chew_grade_derived_mine_loose_median_group,
      # grepl("chew_grade_derived_yue_median", name) ~ chew_grade_derived_yue_median_group,
      # grepl("chew_grade_xgb_median", name) ~ chew_grade_xgb_median_group,
      # grepl("chew_grade_XGB_derived_median", name) ~ chew_grade_XGB_derived_median_group
    )
  ) %>%
  dplyr::select(-matches(".*_group$")) %>% 
  ggplot(aes(tidytext::reorder_within(my_cluster, value, name), value, color = factor(group))) +
  geom_point(size  = 5) +
  facet_wrap(~name, scales = "free", nrow = 1) +
  tidytext::scale_x_reordered() +
  theme(axis.text.x = element_text(angle = 90))


# generate structures for each measure of progression 
stat  = "pseudoprog_median"

# ****** Function: make edges from cluster order with groups
make_edges <- function(stat = character()){
  
  # var with stat groupsings
  group_var <- grep(sprintf("%s_group", stat), names(cluster_stats_grouped), value = TRUE)
  
  ## for stat and that stats groups, make list of groups 
  clusts <- cluster_stats_grouped$my_cluster[order(cluster_stats_grouped[[stat]])]
  groups <- cluster_stats_grouped[order(cluster_stats_grouped[[stat]]) , group_var]
  grouped <- split(clusts,groups )
  
  ## connect groups
  edges <- matrix(NA, ncol = 2, dimnames = list(NA, c("from", "to")))[0,]
  for (i in seq_along(grouped)) {
    current_group <- grouped[[i]]
    # Intra-group connections
    if (length(current_group) > 1) {
      intra <- rbind(t(combn(current_group, 2)), t(combn(rev(current_group), 2)))
      colnames(intra) <- c("from", "to")
      edges <- rbind(edges, intra)
    }
    # Inter-group connections
    if (i < length(grouped)) {
      next_group <- grouped[[i + 1]]
      inter <- expand.grid(from = current_group, to = next_group)
      edges <- rbind(edges, inter)
    }
  }
  
  edges <- unique(edges)  # Remove duplicates if needed
  edges <- as.matrix(edges)  
  
  return(edges)
}

# make edge matrix of inferred structure from ordering of stat
edge_list <- sapply(stat_vars, function(stat){
  make_edges(stat)
}, simplify = FALSE)


# add simple linear edges matrices
simple_edge_ms <- lapply(stat_vars, function(stat){
  ordered_clusts <- as.character(cluster_stats_grouped$my_cluster[order(cluster_stats_grouped[[stat]])])
  edge_m <- cbind(ordered_clusts[-length(ordered_clusts)], ordered_clusts[-1])
  colnames(edge_m) <- c("from", "to")
  rownames(edge_m) <- seq(nrow(edge_m))
  return(edge_m)
})

names(simple_edge_ms) <- paste("simple", stat_vars, sep = "_")

# add to other edge ms
edge_list <- append(edge_list,simple_edge_ms )

# MAKE LINEAR PROGRESSION GRAPH----
# ****** FUnction: make graphs from list of edges
make_graph <- function(edge_matrix =NA, plot = FALSE, main = NA){
  g <- graph_from_edgelist(edge_matrix, directed = TRUE)
  g <- set_vertex_attr(g, index = V(g), name = "color", value =  cluster_cols[names(V(g))])
  
  if(plot){
    set.seed(210409490);
    # par(bg = "black", col.main = 'white')
    plot.igraph(g, vertex.color = V(g)$color, 
                # layout = layout_as_tree(g, root = c(12,3)),
                vertex.size = 10,
                edge.arrow.size = .1, 
                main = main
                )
    par(bg = "white", col.main = 'black')
  }
  
  # numeric adjacency matrix
  adjacency_matrix <- as.matrix(as_adjacency_matrix(g)) 
  
  # # make diagonal = 1 so there is self transision
  # diag(adjacency_matrix) <- 1
  
  # graph_from_adjacency_matrix(adjacency_matrix) %>% plot
  
  # distance between nodes
  distance_matrix <- distances(g)
  
  # make sure oredered right
  distance_matrix <- distance_matrix[as.numeric(rownames(distance_matrix )) , as.numeric(colnames(distance_matrix ))]
  
  
  return(distance_matrix)
}

linear_progression_distance_matrix <- make_graph(edge_list$pseudoprog_median)

```

#__________

# Test which structure observed transitions fits best
<!-- ```{r} -->


<!-- par(mfrow =c(1,1), bg = "white") -->

<!-- # ****** function to calculate fit based on the tree structure ****** -->
<!-- fit_score <- function(observed_trans_matrix, fit_matrix) { -->

<!--   # set diagonal 0 because distance to self == 0 -->
<!--   diag(fit_matrix) <- 0 -->
<!--   diag(observed_trans_matrix) <- 0 -->
<!--   #^ severely reduces fit when including diagonal transitions (i.e. 1-1, 2-2) -->

<!--   # sum of trans * closeness -->
<!--   fit <- sum(observed_trans_matrix * fit_matrix ) -->

<!--   # larger fit =>> more closely aligning to fit matrix -->
<!--   return(fit) -->
<!-- } -->

<!-- # ****** FUNCTION: turn transition matrix into fit matrix -->
<!-- #+ fit matrix is the reciprocal of a distance matrix with a constant added to avoid division by zero error -->
<!-- #+ diag is set to zero to ignore self-transitions -->
<!-- make_fit <- function(m){ -->
<!--   m <- 1/(1+m) -->
<!--   diag(m) <- 0 -->
<!--   return(m) -->
<!-- } -->

<!-- normalize <- function(m){ -->
<!--   m/sum(m) -->
<!-- } -->


<!-- # make proximity matrices wout diagonal -->
<!-- bif_fit <- bif_distance_matrix %>% make_fit() -->
<!-- lin_fits <- lapply(edge_list, make_graph)  %>%  -->
<!--   lapply(., make_fit) -->
<!-- rand_fit <-rand_distance_matrix %>% make_fit() -->

<!-- # normalize fit matrices -->
<!-- bif_fit <- normalize(bif_fit) -->
<!-- lin_fits <- purrr::map(lin_fits,normalize) -->
<!-- rand_fit <-normalize(rand_fit) -->

<!-- all_fits <- append(list(bif_fit = bif_fit, rand_fit = rand_fit), lin_fits) -->

<!-- trans_mat <- transition_matrix -->

<!-- fits <- sapply(all_fits, function(fit_m){ -->
<!--   fit_score(trans_mat, fit_m) -->
<!-- }) -->

<!-- fits <- fits[c(1:2,5,6)] -->
<!-- names(fits) <- c("Bifurcated", "Unordered", "Pseudotime Median","Chew Grade Median") -->
<!-- barplot(sort(fits)) -->

<!-- pdf(sprintf("%sresults/clustering/09_clustering/09.2.4/%s.pdf",wd_path, "fit_bars"), height = 5,width = 7, bg = "white") -->
<!--   par(mar = c(8,5,3,3)) -->
<!--   bp <- barplot(sort(fits), main = "", las = 3, col = "white", names.arg = NA, yaxt = "none",  ylim  = c(0,max(fits)+.5)) -->
<!--   axis(2,at = seq(0,8,.5), las = 2) -->
<!--   mtext("Observed transitions fit against structures", side = 3, line = -.3, cex = 1) -->
<!--   labels <- names(sort(fits)) -->
<!--   text(x =bp, y = par("usr")[3] - 0.5, labels = labels, -->
<!--        srt = 50, adj = 1, xpd = TRUE, cex = 0.8) -->
<!--   par(mar = c(3,3,3,3)) -->
<!-- dev.off() -->

<!-- sprintf("cp %s %s", sprintf("%sresults/clustering/09_clustering/09.2.4/%s.pdf",wd_path, "fit_bars"), sprintf("%sresults/publication_figures/%s.pdf",wd_path, "fit_bars")) %>%  -->
<!--     system() -->

<!-- pdf(sprintf("%sresults/clustering/09_clustering/09.2.4/%s.pdf",wd_path, "inverted_fit_bars"), height = 5,width = 10, bg = "white") -->
<!--   par( -->
<!--     mar = c(8,5,3,3) -->
<!--     # bg = "black",         -->
<!--     # col.axis = "white",     -->
<!--     # col.lab = "white",      -->
<!--     # col.main = "white",     -->
<!--     # col.sub = "white" -->
<!--     )    -->
<!--   bp <- barplot(sort(fits), main = "", las = 3,  -->
<!--                 # col = "white",  -->
<!--                 names.arg = NA, yaxt = "none",  ylim  = c(0,max(fits)+.5)) -->
<!--   axis(2,at = seq(0,8,.5), las = 2 -->
<!--        # col = "white" -->
<!--        ) -->
<!--   mtext("Observed transitions fit against structures", side = 3, line = -.3, cex = 1 -->
<!--         # col = "white" -->
<!--         ) -->
<!--   labels <- names(sort(fits)) -->
<!--   text(x =bp, y = par("usr")[3] , labels = labels, -->
<!--        srt = 0, adj = c(.5,2), xpd = TRUE, cex = 0.8 -->
<!--        # col = 'white' -->
<!--        ) -->
<!--  par(mar = rep(3,4), -->
<!--     bg = "white",         -->
<!--     col.axis = "black",     -->
<!--     col.lab = "black",      -->
<!--     col.main = "black",     -->
<!--     col.sub = "black" -->
<!--     ) -->
<!-- dev.off() -->


<!-- graph_path <- sprintf("%sresults/clustering/09_clustering/09.2.4/linear_graphs", wd_path)  -->
<!-- graph_path %>%  -->
<!--   {if(!dir.exists(.)) dir.create(.)} -->
<!-- lapply(names(edge_list), function(nm){ -->
<!--   m <- edge_list[[nm]] -->
<!--   pdf(sprintf("%s/%s.pdf",graph_path, nm), height = 5,width = 5) -->
<!--   make_graph(m, plot = TRUE, main = stringr::str_to_title(nm)) -->
<!--   dev.off() -->
<!-- }) -->

<!-- lapply(c("simple_pseudoprog_median","simple_chew_grade_derived_mine_loose_median"), function(nm){ -->
<!--   m <- edge_list[[nm]] -->
<!--   pdf(sprintf("%sresults/clustering/09_clustering/09.2.4/%s.pdf",wd_path, nm), height = 10,width = 10) -->
<!--     make_graph(m, plot = TRUE,  -->
<!--                main = str_replace_all(nm, "_", " ") %>%stringr::str_to_title(.) -->
<!--                ) -->
<!--   dev.off() -->
<!--   sprintf("cp %s %s",  -->
<!--           sprintf("%sresults/clustering/09_clustering/09.2.4/%s.pdf",wd_path, nm), -->
<!--           sprintf("%sresults/publication_figures/%s.pdf",wd_path, nm) -->
<!--           ) %>%  -->
<!--     system() -->
<!-- }) -->

<!-- ``` -->


```{r}

# **** FUNCTIONS *****
normalize <- function(m){
  # m/sum(m)
  m %>% sweep(1, rowSums(m), "/")
}

make_prob <- function(m){
  m <- 1/(1+m)
  diag(m) <- 0
  m <- normalize(m)
  return(m)
}
g.test
compute_g_stat <- function(O, E) {
  valid_idx <- (O > 0 & E > 0) & (row(O) != col(O))
  stat <- 2 * sum(O[valid_idx] * log(O[valid_idx] / E[valid_idx]))
  return(stat)
}

# make proximity matrices 
bif_prob <- bif_distance_matrix %>% make_prob()
lin_probs <- lapply(edge_list, make_graph)  %>% 
  lapply(., make_prob)
unordered_prob <-rand_distance_matrix %>% make_prob()

all_probs <- append(list(bif_prob = bif_prob, unordered_prob = unordered_prob), lin_probs)

O <- t(next_clusters_m_frequencies)
diag(O) <- 0

Es <- lapply(all_probs, function(p){

  # row sum normalization
  E <- p %>%
    sweep(1,rowSums(O), "*")
  
  return(E)
})

# Compute G statistic (likelihood ratio test)
Stats <- sapply(Es, function(E){
  compute_g_stat(O, E)

})


stats <- Stats[c(1:2,5,6)]
names(stats) <- c("Bifurcated", "Unordered", "Pseudoprogression Median", "Chew Stage Median")

ord <- order(stats)
stats <- stats[ord]

pdf(sprintf("%sresults/clustering/09_clustering/09.2.4/%s.pdf", wd_path, "G_stat_fit_bars"),    height = 5, width = 7, bg = "white")
{
  # sub <- statistics[-which(names(stats)=="Unordered")]
  sub <- stats
  
  par(mar = c(5, 10, 4, 2))  # wide left margin for labels
  dotchart(sub,
           labels = names(sub),
           xlab = "G Statistic",
           main = "Likelihood Ratio Test of Observed versus Potential Structures",
           pch = 19,
           col = "black",
           cex = 1
           )
  par(mar = c(3, 3, 3, 3))
}
dev.off()

```

## MC emperical distribution of stats
```{r}

## SEMI RANDOM MATRIX WOUT DIAG WHERE ROWSSUMS MATCH ROWSUMS OF O
generate_random_matrix <- function(O, n_matrices = 10, seed = 123) {
  set.seed(seed)
  n <- nrow(O)
  offdiag_idx <- row(O) != col(O)

  replicate(n_matrices, {
    mat <- matrix(0, nrow = n, ncol = n)
    for (i in 1:n) {
      row_sum <- sum(O[i, ])
      if (row_sum == 0) next  # skip empty rows

      offdiag_cols <- setdiff(1:n, i)
      probs <- runif(length(offdiag_cols))
      probs <- probs / sum(probs)

      mat[i, offdiag_cols] <- rmultinom(1, size = row_sum, prob = probs)
    }
    mat
  }, simplify = FALSE)
}


### COMPLETELY RANDOM MATRIX WOUT DIAG 
# generate_random_matrix <- function(O, n_matrices = 10, seed = 123) {
#   set.seed(seed)
#   n <- nrow(O)
#   total_sum <- sum(O)
#   offdiag_idx <- row(O) != col(O)
# 
#   replicate(n_matrices, {
#     probs <- runif(sum(offdiag_idx))
#     probs <- probs / sum(probs)
#     counts <- rmultinom(1, size = total_sum, prob = probs)
# 
#     mat <- matrix(0, nrow = n, ncol = n)
#     mat[offdiag_idx] <- counts
#     mat
#   }, simplify = FALSE)
# }


n_rand <- 1e4
rEs <- generate_random_matrix(O,n_matrices = n_rand)


names(rEs) <- paste0("r", seq(n_rand))

# allEs <- append(rEs, list(Es, Observed = O))
rEs <- append(rEs, list(observed = O))

rStats <- sapply(rEs, function(E){
  compute_g_stat(O,E)
})


{
  par(mar = c(5, 5, 4, 2))  # wide left margin for labels
  
  # Plot density of random test statistics
  dens <- density(rStats[grepl("^r\\d", names(rStats))])
  plot(dens, main = "Empirical Null Distribution",
       xlab = "Test Statistic", ylab = "Density", lwd = 2,
       xlim = c(0, max(dens$x))
       )
  
  # Add vertical lines for observed fits
  abline(v = stats, 
         col = rainbow(length(stats)), lty = 1, lwd = 4)
  
  # Add legend
  legend("topleft",
         legend = names(stats),
         col = rainbow(length(stats)),
         lty =1,lwd = 4,
         box.col = NA, bg = NA)
  
  par(mar = c(3, 3, 3, 3))
}

# probability that each alignment is due to randomness?
sapply(stats, function(s){
  mean(rStats<s)
}) %>% 
  round(6)
# Bifurcated Pseudoprogression Median        Chew Stage Median                Unordered 
#     0.0118                   0.0122                   0.0122                   0.0822 


```


