---
title: "08_functionAndStructureModeling"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---
--------
--------
--------
this version uses a lagged model in rate modeling to increase number of samples. Idea: see what phenotypes have effect from one visit to next, versus from first to last
--------
--------
--------
--------

# options
```{r}
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()
```


# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}
library(tidyr)
library(dplyr)
library(reshape2)
library(purrr)
library(GGally)
library(tidyverse)
library(ggpubr)
library(vcd)
library(UpSetR)
library(ggplot2)
library(gridExtra)
# library(BiocSingular)
library(ComplexHeatmap)
library(viridis)
library(knitr)
library(htmlwidgets)
library(zoo)
library(corrplot)
library(RColorBrewer)
library(openxlsx)
library(plotly)
# library(scater)
library(RColorBrewer)
library(mgcv)
library(zoo)
library(tools)
library(mice)
library(grDevices)
# library(limma)
library(ggrepel)
library(networkD3)
library(htmltools)
library(grid)
library(gplots)
library(Metrics)
library(psych)
library(car)
library(stats)
library(lme4)
library(lmerTest)
library(VennDiagram)
library(ape)
library(SingleCellExperiment)



rm(list = ls())
gc()
```

# functions
```{r}

# function to show notification 
notify.me <- function(message = "R completed"){
  
  message <- paste0("'", message, "'")
  
  command <- paste0( "terminal-notifier -message ", message, " -title 'R Notification' -subtitle 'Completion Notice' -sound /System/Library/Sounds/Ping.aiff")
  return(system(command))
}

```


# Load Data
info and d from sce
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

# load experiment
sce <- readRDS(paste0(wd_path, "processed_data/sce003.rds"))

d <- left_join(metadata(sce)$d, as.data.frame(colData(sce)) )
info <- metadata(sce)$info
raw <- metadata(sce)$raw

```

# make new directorys for plots
```{r}
path <- paste0(wd_path, "results/modeling/")
if(!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
  
}


VA_modeling_path <- paste0(wd_path, "results/modeling/VA_new/")
if(!dir.exists(VA_modeling_path)) {
  dir.create(VA_modeling_path, recursive = TRUE)
  
}

MT_modeling_path <- paste0(wd_path, "results/modeling/MT_new/")
if(!dir.exists(MT_modeling_path)) {
  dir.create(MT_modeling_path, recursive = TRUE)
  
}

```

# clean data


# _______________________
# what phenotypes to model?
*ensure numeric
```{r}

phenotypes <-  c( "NEOVASC",  "HRDEXUDATE", "INTRARET_PGMT_MIGR", "LOCAL_DECR_AF_VESL_END", "HR_TEMPORAL", "RETHAEM","YELLOWRND","CRYSTALDEP","RET_ATROPH","TYP_LEAK_RPE","FAZ",
                  "BLRPATTERN","RAV","FOVABNORM","FOVDET","ISOS_LOC","CTR_PT_INVOLV","AF_INCREASED_1DD",      
                  "ISOS_LOC","LOSS_ONL","ISOS_BREAK","HR_TEMPORAL","INR_EMPTY_SPC" , "OUTR_EMPTYSPC","SUBRET_NEW_VESL"   ,"CTRPOINT_INVOLVD",
                  "CTRPOINT_INVOL_2","INTRARET_PGMT_MIGR","EZLOSS","CENTER_MICRON", "CENTERMAX_MICRON", "CENTERMIN_MICRON", 
                  "CYLNUMA","SPHNUMA", "AXISA","CYLPMA","SPHPMA", "SPHEQMA",  "myopic",     
                  "FA_mean", "FA_mean_rate",
                 grep("IR0\\d", names(d), value = TRUE), 
                 grep("BRV", names(d), value = TRUE), 
                 grep("DRV", names(d), value = TRUE), 
                 grep("VTV", names(d), value = TRUE), 
                 grep("LRT", names(d), value = TRUE), 
                 grep("RAV", names(d), value = TRUE), 
                 grep("RPE_LK", names(d), value = TRUE), 
                 grep("_rate", names(d), value = TRUE), 
                 grep("ICL", names(d), value = TRUE),
                 grep("_slope_rate", names(d), value = TRUE),
                 grep("avg", names(d), value = TRUE),
                 # grep("km_nn_cluster|severity_of_cluster", names(d), value = TRUE),
                 grep("_sum", names(d), value = TRUE),
                 grep("pseudoprog", names(d), value = TRUE)
                 
                 
                 ) %>% 
  
                grep("FA.*rate", ., value = TRUE, invert = TRUE) %>% 
  unique()
                 

all(sapply(d[, phenotypes] ,is.numeric))

```



#__________

<!-- #__  ___   ___   __ -->
<!-- # modeling vision loss -->
<!-- # *** -->

<!-- #$$ -->
<!-- # * raw values -->
<!-- ## VA_mean -->
<!-- #### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->


<!-- #  variables (rows of value>0) < 30 where FA_mean present -->
<!-- m <- d[!is.na(d$FA_mean),phenotypes] != 0 -->
<!-- vars <- m %>% -->
<!--   colSums(., na.rm = TRUE)  -->

<!-- # variables with > 30 obs of value>0  -->
<!-- deviant_vars <- vars[vars > 30] %>% names() -->

<!-- # how many unique months do these have where FA_mean and var available? -->
<!-- count_unq_months <-  sapply(deviant_vars, function(x){  -->
<!--                     # how many unique months where both FA and var available? -->
<!--                     return( d $month[!is.na(d$FA_mean) & !is.na(d[,x])] %>%  -->
<!--                              unique() %>% -->
<!--                              length() -->
<!--                             ) -->
<!--                   }) -->

<!-- # vars with greater than one unique month observed  -->
<!-- VA_mean_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%  -->
<!--   names() -->

<!-- # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0 -->

<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF, all_of(VA_mean_greater_than_one_month)) -->

<!-- # scale vars of interest -->
<!-- model_df[, VA_mean_greater_than_one_month[-which(VA_mean_greater_than_one_month == "FA_mean")]] <- scale(model_df[, VA_mean_greater_than_one_month[-which(VA_mean_greater_than_one_month == "FA_mean")]]) -->


<!-- ``` -->

<!-- #### model each variable against FA_mean, correcting for SEXF, and random effect eye and patient -->
<!-- ```{r} -->
<!-- # exclude the response variable, FA_mean and correction variables -->
<!-- input_phenotypes <- names(model_df)[-which(names(model_df) %in%  -->
<!--                                              c("FA_mean", "FA_mean_rate",  -->
<!--                                                 "month","patID_eye","patID","SEXF")) -->
<!--                                     ] -->

<!-- # for each variable in input_phenotypes, model FA_mean  -->
<!-- # pheno = "FAZ" -->
<!-- results_list <- list(); for (pheno in input_phenotypes){ -->
<!--   # model without month becasue variables already capture time as progression -->
<!--   formula <- as.formula( paste0("FA_mean ~ SEXF  +", pheno,  "+ (1|patID_eye) + (1 | patID)") ) -->

<!--   # model with month -->
<!--   # formula <- as.formula( paste0("FA_mean ~ SEXF + month +", pheno,  "+ (1|patID_eye)") ) -->

<!--   print(formula) -->


<!--   # results_list[[pheno]] <- summary(lm(formula = formula, data = model_df)) -->
<!--   results_list[[pheno]] <- summary(lmer(formula = formula, data = model_df)) -->

<!-- } -->

<!-- ## make results df #### -->
<!-- results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(), -->
<!--                          Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE) -->

<!-- # fill df with model values -->
<!-- for (model_name in names(results_list)) { -->

<!--   model_coef <- coef(results_list[[model_name]]) -->

<!--   a_result_df <- data.frame(Model = model_name,  -->
<!--                          Term = rownames(model_coef),  -->
<!--                          Estimate = model_coef[, "Estimate"], -->
<!--                          Std.Error = model_coef[, "Std. Error"], -->
<!--                          t.value = model_coef[, "t value"], -->
<!--                          Pr = model_coef[, "Pr(>|t|)"]) -->

<!--   results_df <- rbind(results_df, a_result_df) -->
<!-- } -->

<!-- # calculate adjusted p value -->
<!-- results_df$q_value <- p.adjust(results_df$Pr, method = "BH") -->

<!-- # filter control variables from results -->
<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->


<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Model %in% grep("avg_MT|MT0", results_df$Model, value = TRUE)) %>%  -->
<!--   filter(!Model %in% c("mean_parafoveal_rates")) -->

<!-- ### plot #### -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- VA_by_phenos_volcano <- ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_mean ~ SEXF + phenotype_i + (1|patID_eye) + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "VA_by_phenos_volcano.png"), VA_by_phenos_volcano, width = 20, height = 12, dpi = 200) -->


<!-- # volcano plot of vaa~phenotypes result -->
<!-- VA_by_phenos_volcano_filtered <- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_mean ~ SEXF + phenotype_i + (1|patID_eye) + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "VA_by_phenos_volcano_filtered.png"), VA_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200) -->

<!-- notify.me("VA by phenos done") -->

<!-- ``` -->

<!-- ## FA1  -->

<!-- #### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->


<!-- #  variables (rows of value>0) < 30 where FA1 present -->
<!-- m <- d[!is.na(d$FA1),phenotypes] != 0 -->
<!-- vars <- m %>% -->
<!--   colSums(., na.rm = TRUE)  -->

<!-- # variables with > 30 obs of value>0  -->
<!-- deviant_vars <- vars[vars > 30] %>% names() -->

<!-- # how many unique months do these have where FA1 and var available? -->
<!-- count_unq_months <-  sapply(deviant_vars, function(x){  -->
<!--                     # how many unique months where both FA and var available? -->
<!--                     return( d $month[!is.na(d$FA1) & !is.na(d[,x])] %>%  -->
<!--                              unique() %>% -->
<!--                              length() -->
<!--                             ) -->
<!--                   }) -->

<!-- # vars with greater than one unique month observed  -->
<!-- FA1_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%  -->
<!--   names() -->

<!-- # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0 -->

<!-- # # subset d with vars of interest  -->
<!-- # model_df <- d %>%  -->
<!-- #   select(month, patID_eye, patID, SEXF, FA1, all_of(FA1_greater_than_one_month)) -->
<!-- #  -->
<!-- # # scale vars of interest -->
<!-- # model_df[, FA1_greater_than_one_month] <- scale(model_df[, FA1_greater_than_one_month]) -->


<!-- #**************$$**************** -->
<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF, all_of(VA_mean_greater_than_one_month), FA1) -->

<!-- # scale vars of interest, not response variable  -->
<!-- model_df[, c( VA_mean_greater_than_one_month)] <- scale(model_df[, c( VA_mean_greater_than_one_month)]) -->
<!-- #******************************** -->
<!-- #^ modeling with VA_mean variables because some filtered out for FA1 varirable filtering  -->


<!-- ``` -->

<!-- #### model FA1, correcting for SEXF, month, and random effect eye -->
<!-- ```{r} -->
<!-- # exclude the response variable, FA1 -->
<!-- input_phenotypes <- names(model_df)[-which(names(model_df) %in% c("FA1", "FA1_rate", "month","patID_eye","patID","SEXF" ))] -->

<!-- # for each variable in input_phenotypes, model FA1  -->
<!-- # pheno = "FAZ" -->
<!-- results_list <- list(); for (pheno in input_phenotypes){ -->
<!--   # model without month becasue variables already capture time as progression -->
<!--   formula <- as.formula( paste0("FA1 ~ SEXF  +", pheno,  "+ (1|patID_eye) + (1 | patID)") ) -->

<!--   # model with month -->
<!--   # formula <- as.formula( paste0("FA1 ~ SEXF + month +", pheno,  "+ (1|patID_eye)") ) -->

<!--   print(formula) -->


<!--   # results_list[[pheno]] <- summary(lm(formula = formula, data = model_df)) -->
<!--   results_list[[pheno]] <- summary(lmer(formula = formula, data = model_df)) -->

<!-- } -->

<!-- ## make results df #### -->
<!-- results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(), -->
<!--                          Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE) -->

<!-- # fill df with model values -->
<!-- for (model_name in names(results_list)) { -->

<!--   model_coef <- coef(results_list[[model_name]]) -->

<!--   a_result_df <- data.frame(Model = model_name,  -->
<!--                          Term = rownames(model_coef),  -->
<!--                          Estimate = model_coef[, "Estimate"], -->
<!--                          Std.Error = model_coef[, "Std. Error"], -->
<!--                          t.value = model_coef[, "t value"], -->
<!--                          Pr = model_coef[, "Pr(>|t|)"]) -->

<!--   results_df <- rbind(results_df, a_result_df) -->
<!-- } -->

<!-- # filter control variables from results -->
<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->

<!-- # calculate adjusted p value -->
<!-- results_df$q_value <- p.adjust(results_df$Pr, method = "BH") -->


<!-- # filter out unwanted variables -->
<!-- results_df <- results_df %>%  -->
<!--   filter(!Model %in% c("FA_mean")) -->

<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Model %in% c(grep("avg_MT|MT0", results_df$Model, value =TRUE), "mean_parafoveal_rates" ) ) -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- FA1_by_phenos_volcano <- ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_1 ~ SEXF + phenotype_i + (1|patID_eye) + (1 | patID)") -->

<!-- ggsave(paste0( VA_modeling_path, "FA1_by_phenos_volcano.png"), FA1_by_phenos_volcano, width = 20, height = 12, dpi = 200) -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- FA1_by_phenos_volcano_filtered <- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_1 ~ SEXF + phenotype_i + (1|patID_eye) + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "FA1_by_phenos_volcano_filtered.png"), FA1_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200) -->

<!-- ``` -->



#$$
# * Rate
<!-- ## Corrplot of betas for VA rate models #### -->
<!-- run all models for each factor rate  -->
<!-- ### Subset data for modeling #### -->
<!-- ```{r} -->

<!-- factor_rate_vars <- grep("FA.*_rate", names(d), value = TRUE) -->

<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF,  all_of(factor_rate_vars), all_of(VA_mean_greater_than_one_month)) -->

<!-- # scale vars of interest, except responses: FA_mean, FA1:4 -->
<!-- model_df[, c( VA_mean_greater_than_one_month[-which(VA_mean_greater_than_one_month == "FA_mean")])] <- scale(model_df[, c(VA_mean_greater_than_one_month[ -which(VA_mean_greater_than_one_month ==  "FA_mean")]) ]) -->

<!-- # reduce model_df to patient's rate and first_visit phenotype value -->
<!-- ## for fa, grab first non na  -->
<!-- model_df_first <- model_df %>%  -->
<!--   arrange(patID_eye, month) %>%   -->
<!--   group_by(patID_eye) %>% -->
<!--   mutate(across( -->
<!--     !matches("FA"),  -->
<!--     ~dplyr::first(.), -->
<!--     .names  = "first_{.col}" -->
<!--     )) %>% -->
<!--   mutate(across( -->
<!--     matches("FA"),  -->
<!--     ~dplyr::first(.[!is.na(.)]), -->
<!--     .names  = "first_{.col}" -->
<!--     )) %>%  -->
<!--   ungroup() -->



<!-- # filter out non firsts and firsts that don't make sense, only unique rows -->
<!-- model_df_first_filtered_unique <- model_df_first %>%  -->
<!--   select( patID_eye, patID, SEXF, starts_with("first")) %>%    -->
<!--   select(!all_of(c("first_month", "first_patID", "first_SEXF"))) %>% -->
<!--   unique() -->
<!-- ``` -->


<!-- ### Model #### -->
<!-- ```{r} -->

<!-- # for each rate , for each phenotype, model rate by that phenotype and add to main results_df -->
<!-- ## vars modeled -->
<!-- model_vars <- paste("first", VA_mean_greater_than_one_month, sep = "_") -->
<!-- rate_vars <- paste("first", factor_rate_vars, sep = "_") -->

<!-- ## -->
<!-- # rate_var  = "first_FA1_rate" -->
<!-- # model_var  = "first_NEOVASC" -->
<!-- ## -->

<!-- results_df_raw <- data.frame(Model = character(),   -->
<!--                          Term = character(), -->
<!--                          Estimate = numeric(), -->
<!--                          Std.Error = numeric(), -->
<!--                          t.value = numeric(), -->
<!--                          Pr = numeric(), -->
<!--                          rate_variable = character(),  -->
<!--                          stringsAsFactors = FALSE) -->

<!-- # loop through rates, except mean, and then model vars -->
<!-- for(rate_var in rate_vars[-5]){ -->

<!--   a_rates_summaries <- list() -->

<!--   cat(paste("\n***\n", toupper(rate_var), "\n***\n\n")) -->

<!--   for(model_var in model_vars){ -->
<!--     # make formula -->
<!--     a_formula <- paste0(rate_var, " ~ SEXF + (1|patID) + " , model_var) %>%  -->
<!--       as.formula() -->
<!--     as.character(a_formula )[2:3] %>% print() -->

<!--     # model and summarize -->
<!--     a_mod_summ <- summary(lmer(a_formula, data = model_df_first_filtered_unique)) -->

<!--     # save summary to list of summaries -->
<!--     a_rates_summaries[[model_var]] <- a_mod_summ -->
<!--   } -->

<!--   # fill df with model values -->
<!--   for (model_name in names(a_rates_summaries)) { -->

<!--       model_coef <- coef(a_rates_summaries[[model_name]]) -->

<!--       a_result_df <- data.frame(Model = model_name,  -->
<!--                              Term = rownames(model_coef),  -->
<!--                              Estimate = model_coef[, "Estimate"], -->
<!--                              Std.Error = model_coef[, "Std. Error"], -->
<!--                              t.value = model_coef[, "t value"], -->
<!--                              Pr = model_coef[, "Pr(>|t|)"],  -->
<!--                              rate_variable = rate_var) -->

<!--       results_df_raw <- rbind(results_df_raw, a_result_df) -->
<!--     } -->

<!-- } -->

<!-- ### Filter results and format data #### -->

<!-- # filter out intercepts and correcting terms -->
<!-- results_df <- results_df_raw %>% -->
<!--   filter(!str_detect(Term, "Intercept|SEXF")) -->

<!-- # calculate q val -->
<!-- results_df$q_value <- p.adjust(results_df$Pr) -->

<!-- # pivot data so rate vars are cols with estimates and models as rows -->
<!-- estimates_rates_d <- results_df %>%  -->
<!--   select(rate_variable, Model, Estimate, q_value) -->

<!-- # widen data each rate col = estimates for each model -->
<!-- rate_estimates_d <- estimates_rates_d %>% -->
<!--   pivot_wider(names_from = rate_variable, values_from = Estimate,  -->
<!--               id_cols = Model, values_fn = list(Estimate = dplyr::first)) -->

<!-- # rate_qvals_d <- estimates_rates_d %>% -->
<!-- #   pivot_wider(names_from = rate_variable, values_from = q_value,  -->
<!-- #               id_cols = Model, values_fn = list(q_value = dplyr::last)) -->
<!-- ``` -->


<!-- ### Plot betas #### -->
<!-- ```{r} -->

<!-- # plot raw values of betas -->
<!-- rate_estimates_matrix <- rate_estimates_d %>% -->
<!--   column_to_rownames("Model") %>% -->
<!--   as.matrix()   -->

<!-- pdf(paste0(VA_modeling_path, "factor_rate_model_betas_heatmap.pdf"), width = 20, height = 10) -->
<!-- Heatmap(rate_estimates_matrix,  -->
<!--         na_col = "black", -->
<!--         column_title = "Betas of models for different factor rates",  -->
<!--         column_title_gp = gpar(fontsize = 20), -->
<!--         column_dend_height = unit(30, "mm"), -->
<!--         row_dend_width = unit(30, "mm"), -->
<!--         cluster_rows = hclust(dist(rate_estimates_matrix), method = "ward.D2"), -->
<!--         cluster_columns = hclust(dist(t(rate_estimates_matrix)), method = "ward.D2") -->
<!--         # cell_fun = function(j, i, x, y, width, height, fill) { -->
<!--         #               grid.text(text_annotations[i, j], x, y) -->
<!--         #       } -->
<!-- ) -->
<!-- dev.off() -->

<!-- # plot correlaation between betas -->
<!-- rate_estimates_cor <- cor(rate_estimates_d[, -1]) -->
<!-- pdf(paste0(VA_modeling_path, "factor_rate_betas_cor_heatmap.pdf"), width = 10, height = 8) -->
<!-- Heatmap(rate_estimates_cor,  -->
<!--         na_col = "black", -->
<!--         column_title = "Correlations between betas in modeling for different factor rates",  -->
<!--         column_title_gp = gpar(fontsize = 20), -->
<!--         column_dend_height = unit(30, "mm"), -->
<!--         row_dend_width = unit(30, "mm"), -->
<!--         clustering_method_columns = "ward.D2" -->
<!--         ) -->
<!-- dev.off() -->


<!-- # par(mfrow = c( 2,1)) -->
<!-- # plot(type = "fan", as.phylo(hclust(dist(t(rate_estimates_matrix), ), method = "ward.D2"))) -->
<!-- # plot(type = "fan", as.phylo(hclust(dist(rate_estimates_matrix, ), method = "ward.D2"))) -->
<!-- # # plot(hclust(dist(rate_estimates_matrix), method = "ward.D2")) -->
<!-- # # plot(hclust(dist(t(rate_estimates_matrix), ), method = "ward.D2")) -->
<!-- # par(mfrow = c(1, 1)) -->




<!-- # ### plot q_values #### -->
<!-- # # plot correlations between q-vals -->
<!-- # rate_q_value_cor <- cor(rate_qvals_d[, -1]) -->
<!-- # pdf(paste0(wd_path, "results/VAA_analysis/factor_rate_qVals_cor_heatmap.pdf"), width = 10, height = 8) -->
<!-- # Heatmap(rate_q_value_cor,  -->
<!-- #         na_col = "black", -->
<!-- #         column_title = "Correlations between q-values in modeling for different factor rates",  -->
<!-- #         column_title_gp = gpar(fontsize = 20), -->
<!-- #         column_dend_height = unit(30, "mm"), -->
<!-- #         row_dend_width = unit(30, "mm") -->
<!-- #         ) -->
<!-- # dev.off() -->
<!-- #  -->
<!-- # # plot raw values of q_values -->
<!-- # pdf(paste0(wd_path, "results/VAA_analysis/factor_rate_model_qVals_heatmap.pdf"), width = 20, height = 10) -->
<!-- # Heatmap(rate_qvals_d %>% column_to_rownames("Model") %>%  mutate_all(~-log10(.)) %>% as.matrix() %>% t(), -->
<!-- #         na_col = "black", -->
<!-- #         column_title = "-log10 q-value of models for different factor rates", -->
<!-- #         column_title_gp = gpar(fontsize = 20), -->
<!-- #         column_dend_height = unit(30, "mm"), -->
<!-- #         row_dend_width = unit(30, "mm") -->
<!-- #     ) -->
<!-- # dev.off() -->


<!-- ``` -->


## Mean Acuity lagged rate by phenotypes

### visualizing model 
```{r}

# n <- sample(1:1000000, 1); set.seed(n)
# set.seed(222649)
# set.seed(255891)
good_pat  = "011005_1"

x <- d %>% 
  # filter(patID_eye == sample(model_df_first$patID_eye, 1)) %>%
  filter(patID_eye == good_pat) %>%
  select(patID_eye,month, avg_MT04_slope, FA_mean) %>% 
  na.omit() %>% 
  arrange(month) %>% 
  mutate(across(-c(month, patID_eye), ~as.numeric(scale(.)))) 

plot(x$avg_MT04_slope~x$month,  col = "red", ylim = c(min(scale(x[, sapply(x, is.numeric)])), max(scale(x[, sapply(x, is.numeric)]))))
lines(x$avg_MT04_slope~x$month, col = "red")

points(x$FA_mean~x$month, pch = 2, col = "blue")
lines(x$FA_mean~x$month, col = "blue"); 

legend(x = mean(x$month), y = mean(x$avg_MT04_slope), legend = c("avg_MT04_slope", "FA_mean"), 
       col = c("red", "blue"), pch = c(1, 2), lty = 1 , bg = NA)
```

#### subset data to scale and model
* filter out the ones with (rows>0) < 30
```{r}
# #  variables (rows of value>0) < 30 where FA_mean_rate present
# m <- d[!is.na(d$FA_mean_rate),phenotypes] != 0
# 
# vars <- colSums(m, na.rm = TRUE) 
# 
# # variables with > 30 obs of value>0 
# deviant_vars <- vars[vars > 30] %>% names()
# 
# # how many unique months do these have where FA_mean_rate and var available?
# count_unq_months <-  sapply(deviant_vars, function(x){ 
#                     # how many unique months where both FA and var available?
#                     return( d $month[!is.na(d$FA_mean_rate) & !is.na(d[,x])] %>% 
#                              unique() %>%
#                              length()
#                             )
#                   })
# # vars with greater than one unique month observed 
# VA_mean_rate_greater_than_one_month <- count_unq_months[count_unq_months>1] %>% 
#   names()
# 
# VA_mean_rate_greater_than_one_month <- c(VA_mean_rate_greater_than_one_month, "FA_mean_rate")

# filter out vars that dont change where there is an fameanrate
vars_dont_change <- phenotypes[ sapply(d[  !is.na(d$FA_mean_rate), phenotypes], n_distinct) <= 1]

# how many unique months do these have where FA_mean_rate and var available? 
count_unq_months <- sapply(phenotypes, function(x){
  
  num_months <- d$month[!is.na(d$FA_mean_rate) & !is.na(d[[x]]) ] %>% 
      unique() %>% 
      length()
  return(num_months)
})

vars_few_month <- count_unq_months[count_unq_months<2] %>% 
  names()

vars_dont_change_and_few_month <- c(vars_dont_change, vars_few_month) %>% unique()

VA_mean_rate_greater_than_one_month <- c("FA_mean_rate", phenotypes[!phenotypes %in% vars_dont_change_and_few_month])

# subset d with vars of interest
model_df <- d %>%
  select(month, patID_eye, patID, SEXF, all_of(VA_mean_rate_greater_than_one_month))

# scale vars of interest
model_df[, VA_mean_rate_greater_than_one_month] <- scale(model_df[, VA_mean_rate_greater_than_one_month])

# 
# #**************$$****************
# # subset d with vars of interest 
# model_df <- d %>% 
#   select(month, patID_eye, patID, SEXF, all_of(VA_mean_greater_than_one_month), FA_mean_rate)
# 
# # scale vars of interest, minus response: FA_mean_rate
# model_df[, c( VA_mean_greater_than_one_month )] <- scale(model_df[, c( VA_mean_greater_than_one_month ) ])
# #********************************

# # are any first values NA?
# first_NA <- d %>%
#   filter(!is.na(eye_num)) %>% 
#   arrange(month) %>%
#   group_by(patID_eye) %>% 
#   select(month, patID_eye, all_of(VA_mean_greater_than_one_month)) %>%
#   filter(if_any(everything(), ~is.na(dplyr::first(.)) )) %>% 
#   ungroup()
# 
# first_NA %>% arrange(patID_eye)
# # ^ YES! 

## reduce model_df to patient's rate and first_visit phenotype value.
# model_df_first <- model_df %>% 
#   arrange(month) %>%  
#   group_by(patID_eye) %>%
#   mutate(across(
#     everything(), 
#         ~dplyr::first(.),
#     .names  = "first_{.col}"
#     )) %>% 
#   ungroup()


## make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows-----------
# first_values_df <- model_df %>% 
#   select(-patID, - SEXF) %>% 
#   arrange(patID_eye, month) %>%  
#   group_by(patID_eye) %>%
#   mutate(
#     across(
#         -matches("FA"), 
#         ~dplyr::first(.[!is.na(.)]),
#         .names = "first_{.col}"
#   )) %>%
#   ungroup() %>%
#   select(-matches("month"), -matches("FA_mean")) %>% 
#   select(patID_eye,  starts_with("first_")) %>% 
#   unique() %>% 
#   arrange(patID_eye)
# 
# # filter out the firsts that don't make sense
# model_df_first <- model_df %>% 
#   arrange(patID_eye) %>% 
#   select(patID_eye, patID, SEXF, FA_mean_rate) %>% 
#   na.omit() %>%
#   unique() %>%  
#   left_join(., first_values_df)


# lagged difference from last visit---------
# manipulated_vars_df <-
#   model_df %>% 
#   select(-patID, - SEXF) %>% 
#   arrange(patID_eye, month) %>%  
  # group_by(patID_eye) %>%
  # 
  # # mutate(across(
  # #   -matches("patID_eye|month|FA_mean_rate"),
  # #   ~( (. - dplyr::lag(.))  / (month - dplyr::lag(month))),
  # #   .names = "lag_rate_{.col}"
  # # )) %>%
  # 
  # mutate(across(
  #   -matches("patID_eye|month"),
  #   ~((. - dplyr::lead(.) ) / (dplyr::lead(month) - month)) ,  # value - next(value) /  next(month) - month; dpheno/dt
  #   .names = "lag_rate_{.col}"
  # ))

# # join to model_df
# model_df_first <- model_df %>% 
#   arrange(patID_eye) %>% 
#   select(patID_eye, patID, SEXF, FA_mean_rate) %>% 
#   na.omit() %>%
#   unique() %>%  
#   left_join(., manipulated_vars_df)

# lagged rate of response 

model_df_lagged_rate <- model_df %>% 
  group_by(patID_eye) %>%
  # value - next(value) /  next(month) - month; dpheno/dt
  mutate(FA_mean_lag_rate = 
           (FA_mean - dplyr::lead(FA_mean) ) / (dplyr::lead(month) - month)
         ) 


# replot concept plot
x <- model_df_lagged_rate %>% 
  filter(patID_eye == sample(model_df_lagged_rate$patID_eye, 1 ) ) %>% 
  # filter(patID_eye == good_pat) %>% 
  select(patID_eye, month, FA_mean, avg_MT04_slope,FA_mean_lag_rate ) 

plot(x$avg_MT04_slope~x$month,  col = "red", ylim = c(-2,2))
lines(x$avg_MT04_slope~x$month, col = "red")

points(x$FA_mean~x$month, pch = 2, col = "blue")
lines(x$FA_mean~x$month, col = "blue"); 

points(x$FA_mean_lag_rate~x$month, pch = 2, col = "purple")
lines(x$FA_mean_lag_rate~x$month, col = "purple"); 

legend(x = mean(x$month), y = mean(x$avg_MT04_slope), 
       legend = c("avg_MT04_slope", "FA_mean", "FA_mean_lag_rate" ), 
       col = c("red", "blue", "purple"), 
       pch = c(1, 2), lty = 1 , bg = NA)

```



#### model each variable against FA_mean_rate, using zonal phenotype averages
```{r}

  
# exclude the response variable, FA_mean_rate and correctors
input_phenotypes <- names(model_df_first) %>% 
  grep("month|patID_eye|patID|SEXF|FA_mean_rate", ., value = TRUE, invert = TRUE)  
  # grep("0_sum", ., value = TRUE, invert = TRUE) # ignore existent sum variables

# pheno = "first_FOVABNORM"
# for each variable in input_phenotypes, model FA_mean 
results_list <- list() ; for (pheno in input_phenotypes){
  
  ## linear model with random effects
  re_formula <- as.formula( paste0("FA_mean_rate ~ SEXF  + ", pheno,  " + (1|patID)") )
  print(paste("random effects: ",paste0("FA_mean_rate ~ SEXF  + ", pheno,  " + (1|patID)")))
  

  results_list[[pheno]] <- summary(lmer(formula = re_formula, data = model_df_first))
  # results_list[[pheno]] <- summary(lm(formula = lm_formula, data = model_df))
  
  
}


## make results df ####
results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
                         Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)

# fill df with model values
for (model_name in names(results_list)) {
  
  model_coef <- coef(results_list[[model_name]])
  
  a_result_df <- data.frame(Model = model_name, 
                         Term = rownames(model_coef), 
                         Estimate = model_coef[, "Estimate"],
                         Std.Error = model_coef[, "Std. Error"],
                         t.value = model_coef[, "t value"],
                         Pr = model_coef[, "Pr(>|t|)"])
  
  results_df <- rbind(results_df, a_result_df)
}

# filter control variables from results
results_df <- results_df %>% 
  filter(!Term %in% c("(Intercept)", "SEXF", "month"))

# calculate adjusted p value
results_df$q_value <- p.adjust(results_df$Pr, method = "BH")

VA_mean_rate_results_df <- results_df

# filter for models that don't make sense
results_df_filtered <- results_df %>% 
  filter(!Model   %in% grep("FA", results_df$Model, value = TRUE)) %>% 
  filter(!Model   %in% grep("avg|MT0|_.*0\\d$", results_df$Model, value = TRUE)) # exclude MT vars and zonal vars
  
### plot ####
VA_rate_by_phenos_volcano_summed_phenos <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("VA_factor_mean_rate ~ SEXF + phenotype_i + (1 | patID)")

ggsave(paste0(VA_modeling_path, "VA_rate_by_phenos_volcano_summed_phenos.png"), VA_rate_by_phenos_volcano_summed_phenos, width = 20, height = 12, dpi = 200)

VA_rate_by_phenos_volcano_summed_phenos_filtered <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("VA_factor_mean_rate ~ SEXF + phenotype_i + (1 | patID)")

ggsave(paste0(VA_modeling_path, "VA_rate_by_phenos_volcano_summed_phenos_filtered.png"), VA_rate_by_phenos_volcano_summed_phenos_filtered, width = 20, height = 12, dpi = 200)

VA_rate_by_phenos_volcano_summed_phenos_filtered_nominalP <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(Pr), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( Pr < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("VA_factor_mean_rate ~ SEXF + phenotype_i + (1 | patID)")

ggsave(paste0(VA_modeling_path, "VA_rate_by_phenos_volcano_summed_phenos_filtered_nominalP.png"), VA_rate_by_phenos_volcano_summed_phenos_filtered_nominalP, width = 20, height = 12, dpi = 200)

```


## FA1 Rate by phenotypes

#### subset data to scale and model
* filter out the ones with (rows>0) < 30
```{r}

#  variables (rows of value>0) < 30 where FA1_rate present
m <- d[!is.na(d$FA1_rate),phenotypes] != 0
vars <- m %>%
  colSums(., na.rm = TRUE) 

# variables with > 30 obs of value>0 
deviant_vars <- vars[vars > 30] %>% names()

# how many unique months do these have where FA1_rate and var available?
count_unq_months <-  sapply(deviant_vars, function(x){ 
                    # how many unique months where both mr and var available?
                    return( d $month[!is.na(d$FA1_rate) & !is.na(d[,x])] %>% 
                             unique() %>%
                             length()
                            )
                  })
# vars with greater than one unique month observed 
FA1_rate_greater_than_one_month <- count_unq_months[count_unq_months>1] %>% 
  names()

FA1_rate_greater_than_one_month <- c(FA1_rate_greater_than_one_month, "FA1_rate")

# setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0

# # subset d with vars of interest 
# model_df <- d %>% 
#   select(month, patID_eye, patID, SEXF, all_of(FA1_rate_greater_than_one_month))
# 
# # scale vars of interest
# model_df[, FA1_rate_greater_than_one_month] <- scale(model_df[, FA1_rate_greater_than_one_month])


#**************$$****************
# subset d with vars of interest 
model_df <- d %>% 
  select(month, patID_eye, patID, SEXF, all_of(VA_mean_greater_than_one_month), FA1_rate)

# scale vars of interest, minus response: FA1_rate
model_df[, c( VA_mean_greater_than_one_month) ] <- scale(model_df[, c( VA_mean_greater_than_one_month) ])
#********************************


# # reduce model_df to patient's rate and first_visit phenotype value
# model_df_first <- model_df %>% 
#   arrange(month) %>%  
#   group_by(patID_eye) %>%
#   mutate(across(
#     everything(), 
#     ~dplyr::first(.),
#     .names  = "first_{.col}"
#     )) %>% 
#   ungroup()
# 
# 
# 
# model_df_first_filtered <- model_df_first %>% 
#   select(patID_eye, patID, SEXF, FA1_rate, starts_with("first")) %>%   
#   select(!matches("first.*\\d_rate")) %>% 
#   select(!c("first_SEXF", "first_month", "first_patID")) %>% 
#   unique()


# make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows
first_values_df <- model_df %>% 
  select(-patID, - SEXF) %>% 
  arrange(patID_eye, month) %>%  
  group_by(patID_eye) %>%
  mutate(
    across(
        everything(), 
        ~dplyr::first(.[!is.na(.)]),
        .names = "first_{.col}"
  )) %>%
  ungroup() %>%
  select(-matches("month"), -matches("FA1_rate")) %>% 
  select(patID_eye,  starts_with("first_")) %>% 
  unique() %>% 
  arrange(patID_eye)

# filter out the firsts that don't make sense
model_df_first <- model_df %>% 
  arrange(patID_eye) %>% 
  select(patID_eye, patID, SEXF, FA1_rate) %>% 
  na.omit() %>%
  unique() %>%  
  left_join(., first_values_df)




```


#### model each variable against FA1_rate, using zonal phenotype averages
```{r}

# what phenos to sum
zonal_pheno_names <- grep("_.*0\\d$", names(model_df_first), value = TRUE) %>% 
    gsub("0\\d",'', . ) %>% 
    unique()

# zonal_pheno <- "first_IR"

# for each pheno name, sum the columns in model_df_first
for (zonal_pheno in zonal_pheno_names){
  
  new_name <- paste0(zonal_pheno, "_sum")
  model_df_first[[new_name]] <- model_df_first[, grep(zonal_pheno, names(model_df_first))] %>% 
    rowSums(.)
}
  
  
# exclude the response variable, FA1_rate
input_phenotypes <- names(model_df_first)[!names(model_df_first) %in% c("month", "patID_eye", "patID", "SEXF", "FA1_rate")]%>% 
  grep("0_sum", ., value = TRUE, invert = TRUE) # ignore existent sum variables


# pheno = "first_FOVABNORM"
# for each variable in input_phenotypes, model FA_mean 
results_list <- list() ; for (pheno in input_phenotypes){
  # model without month becasue variables already capture time as progression
  
  ## linear model
  # lm_formula <- as.formula( paste0("FA1_rate ~ SEXF  + month + ", pheno ))
  # lm_formula <- as.formula( paste0("FA1_rate ~ SEXF  +  ", pheno ))
  # print(paste("lm: ", paste0("FA1_rate ~ SEXF  + ", pheno )))
  
  ## linear model with random effects
  re_formula <- as.formula( paste0("FA1_rate ~ SEXF  + ", pheno,  " + (1|patID)") )
  print(paste("random effects: ",paste0("FA1_rate ~ SEXF  + ", pheno,  " + (1|patID)")))
  

  results_list[[pheno]] <- summary(lmer(formula = re_formula, data = model_df_first))
  # results_list[[pheno]] <- summary(lm(formula = lm_formula, data = model_df))
  
  
}


## make results df ####
results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
                         Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)

# fill df with model values
for (model_name in names(results_list)) {
  
  model_coef <- coef(results_list[[model_name]])
  
  a_result_df <- data.frame(Model = model_name, 
                         Term = rownames(model_coef), 
                         Estimate = model_coef[, "Estimate"],
                         Std.Error = model_coef[, "Std. Error"],
                         t.value = model_coef[, "t value"],
                         Pr = model_coef[, "Pr(>|t|)"])
  
  results_df <- rbind(results_df, a_result_df)
}
# calculate adjusted p value
results_df$q_value <- p.adjust(results_df$Pr, method = "BH")

# filter control variables from results
results_df <- results_df %>% 
  filter(!Term %in% c("(Intercept)", "SEXF", "month"))

# remove FA, overfitted
results_df <- results_df %>% 
  filter(!Model   %in% grep("FA", results_df$Model, value = TRUE)) 

FA1_rate_results_df <- results_df

# filter for models that don't make sense to include
results_df_filtered <- results_df %>% 
  filter(!Model   %in% grep("avg|MT0|_.*0\\d", results_df$Model, value = TRUE)) # exclude MT vars and zonal vars
  
### volcano plot of vaa~phenotypes result ####
FA1_rate_by_phenos_volcano_summed_phenos <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("VA_factor_1_rate ~ SEXF + phenotype_i + (1 | patID)")

ggsave(paste0(VA_modeling_path, "FA1_rate_by_phenos_volcano_summed_phenos.png"), FA1_rate_by_phenos_volcano_summed_phenos, width = 20, height = 12, dpi = 200)

FA1_rate_by_phenos_volcano_summed_phenos_filtered <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("VA_factor_1_rate ~ SEXF + phenotype_i + (1 | patID)")

ggsave(paste0(VA_modeling_path, "FA1_rate_by_phenos_volcano_summed_phenos_filtered.png"), FA1_rate_by_phenos_volcano_summed_phenos_filtered, width = 20, height = 12, dpi = 200)

FA1_rate_by_phenos_volcano_summed_phenos_filtered_nominalP <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(Pr), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( Pr < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("VA_factor_1_rate ~ SEXF + phenotype_i + (1 | patID)")

ggsave(paste0(VA_modeling_path, "FA1_rate_by_phenos_volcano_summed_phenos_filtered_nominalP.png"), FA1_rate_by_phenos_volcano_summed_phenos_filtered_nominalP, width = 20, height = 12, dpi = 200)


```


##### corrplot of betas
```{r}
VA_mean_rate_results_df
FA1_rate_results_df

FA_mean <- VA_mean_rate_results_df %>% 
  select(Term, Estimate) %>% 
  rename_with(~all_of(c("Term", "FA_mean_Estimate")))

FA1 <- FA1_rate_results_df %>% 
  select(Term, Estimate) %>% 
  rename_with(~all_of(c("Term", "FA1_Estimate")))

estimates <- merge(FA1, FA_mean) %>% 
  column_to_rownames("Term") %>% 
  as.matrix()

Heatmap(estimates)
plot(estimates, type = "n");text(estimates, labels = rownames(estimates), pos = 4, col = rainbow(length(estimates)));abline(c(0,1), lty = 3)


```


##### find what's driving crystal_dep significance in the FA_mean_rate model
```{r}


# find first value of each patient for FA rates and CRYSTALDEP
CRYSTALDEP_model_df <- d %>% 
  select(month, SEXF, patID, patID_eye, grep("FA.*_rate", names(d), value = TRUE), CRYSTALDEP) %>% 
  arrange(month) %>%  
  group_by(patID_eye) %>%
  mutate(across(
    all_of(c(grep("FA.*_rate", names(d), value = TRUE), "CRYSTALDEP") ), 
    ~dplyr::first(.[!is.na(.)]),
    .names  = "first_{.col}"
    )) %>% 
  ungroup() %>% 
  select(!c(grep("FA.*_rate", names(d), value = TRUE), "CRYSTALDEP"))%>% 
  select(-month) %>% 
  unique()

CRYSTALDEP_results_list <- list() ; for (factor_name in paste0("first_", c("FA1_rate", "FA2_rate"))){ #, "FA3_rate", "FA4_rate", "FA_mean_rate"))  ) {

  re_formula <- as.formula( paste0(factor_name, " ~ SEXF  + first_CRYSTALDEP + (1|patID)") )
  re_formula %>% print()
  

  CRYSTALDEP_results_list[[factor_name]] <- summary(lmer(formula = re_formula, data = CRYSTALDEP_model_df))
  }

CRYSTALDEP_results <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
                         Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)


for (model_name in names(CRYSTALDEP_results_list)){
  
  model_coef <- coef(CRYSTALDEP_results_list[[model_name]])
  a_result_df <- data.frame(Model = model_name, 
                         Term = rownames(model_coef), 
                         Estimate = model_coef[, "Estimate"],
                         Std.Error = model_coef[, "Std. Error"],
                         t.value = model_coef[, "t value"],
                         Pr = model_coef[, "Pr(>|t|)"])
  
  CRYSTALDEP_results <- rbind(CRYSTALDEP_results, a_result_df)
}

CRYSTALDEP_results <- CRYSTALDEP_results %>% 
  filter(Term == "first_CRYSTALDEP")

CRYSTALDEP_results$q_value <- p.adjust(CRYSTALDEP_results$Pr)

ggplot(CRYSTALDEP_results, aes(Estimate, -log10(q_value), color = Model))+
  geom_point()+
  geom_label_repel(aes(label =  Model),
                       size = 3,
                       show.legend = FALSE)+
  theme_bw()+
  ggtitle("VA_factor_i ~ SEXF + first_CRYSTALDEP + (1|patID)")

```

## regress mean rate by factors within each cluster
### vision of each cluster for each trajectory
```{r}

d <- d %>% 
  mutate(
    tiny_letters = rowMeans(across(matches("RW1[234]")), na.rm = TRUE),
    small_letters = rowMeans(across(matches("RW1[01]|RW0[89]")),na.rm = TRUE),
    medium_letters = rowMeans(across(matches("RW0[4567]")),na.rm = TRUE),
    large_letters = rowMeans(across(matches("RW0[123]")) ,na.rm = TRUE)
         )

# For row scores ---------
cluster_letter_means <- d %>% 
  select(patID_eye, month, km_nn_cluster, matches("letters")) %>% 
  pivot_longer(cols = matches("letters"), names_to = "variable") 


paths <- list(one  = c(1,2,3,11), 
                    two = c(1,2,3,9), 
                    three = c(1,2,3,7), 
                    three = c(1,2,4,8), 
                    four = c(1,2,4,6), 
                    five = c(1,2,4,8), 
                    six = c(1,2,4,10), 
                    seven = c(1,2,5)
              )

path = paths[["four"]]

# for each path plot every each letter size
cluster_letters_plots <- list(); for(path in paths){
  
  plotting_d <- cluster_letter_means %>% 
    filter(km_nn_cluster %in% path) %>% 
    mutate(km_nn_cluster = factor(km_nn_cluster, levels = path))
  
  cluster_letter_plot <- ggplot(plotting_d, aes(x = km_nn_cluster, y = value))+
    # geom_violin(aes( color = km_nn_cluster)) +
    geom_boxplot(aes( color = km_nn_cluster), show.legend = FALSE) +
    geom_point(aes( color = km_nn_cluster), alpha = .3, show.legend = FALSE)+
    # geom_jitter(aes( color = km_nn_cluster), alpha = .3)+
    geom_smooth(method = "lm", aes(color = NULL, ,group = variable ),show.legend = FALSE)+
    # geom_smooth(method = "loess", aes(color = NULL, ,group = variable ),show.legend = FALSE)+
    facet_wrap(~variable)+
    ggtitle(paste0(paste(path, collapse = "-")))
  
  cluster_letters_plots[[paste(path, collapse = "-")]] <- cluster_letter_plot
    
  
}

ggsave(paste0(VA_modeling_path, "letter_means_by_cluster_path.png"), 
       marrangeGrob(cluster_letters_plots %>% sapply(., ggplotGrob), ncol =7, nrow =1 ),
       height = 10, width = 40
       )

# For ACUIT ---------
cluster_ACUIT <- d %>% 
  select(patID_eye, month, km_nn_cluster, ACUIT) %>% 
  pivot_longer(cols = ACUIT, names_to = "variable") 


paths <- list(one  = c(1,2,3,11), 
                    two = c(1,2,3,9), 
                    three = c(1,2,3,7), 
                    three = c(1,2,4,8), 
                    four = c(1,2,4,6), 
                    five = c(1,2,4,8), 
                    six = c(1,2,4,10), 
                    seven = c(1,2,5)
              )

# path = paths[["four"]]

# for each path plot every each letter size
cluster_ACUIT_plots <- list(); for(path in paths){
  
  plotting_d <- cluster_ACUIT %>% 
    filter(km_nn_cluster %in% path) %>% 
    mutate(km_nn_cluster = factor(km_nn_cluster, levels = path))
  
  cluster_ACUIT_plot <- ggplot(plotting_d, aes(x = km_nn_cluster, y = value))+
    # geom_violin(aes( color = km_nn_cluster)) +
    geom_boxplot(aes( color = km_nn_cluster), show.legend = FALSE) +
    geom_point(aes( color = km_nn_cluster), alpha = .3, show.legend = FALSE)+
    # geom_jitter(aes( color = km_nn_cluster), alpha = .3)+
    geom_smooth(method = "lm", aes(color = NULL, ,group = variable ),show.legend = FALSE)+
    # geom_smooth(method = "loess", aes(color = NULL, ,group = variable ),show.legend = FALSE)+
    facet_wrap(~variable)+
    ggtitle(paste0(paste(path, collapse = "-")))
  
  cluster_ACUIT_plots[[paste(path, collapse = "-")]] <- cluster_ACUIT_plot
    
  
}

ggsave(paste0(VA_modeling_path, "ACUIT_by_cluster_path.png"), 
       marrangeGrob(cluster_ACUIT_plots %>% sapply(., ggplotGrob), ncol =4, nrow =2 ),
       height = 10, width = 40
       )

```

## see how rate correlates to factors when subsetting for each trajectory a -> b, b->c, etc.

# $$
# * Rate for each cluster

```{r}




```


# ____   ___   ___   _____

# modeling thickness 
#***

## MT5 thickness
### subset data to scale and model
* filter out the ones with (rows>0) < 30
```{r}

#  variables (rows of value>0) < 30 where avg_MT05 present
m <- d[!is.na(d$avg_MT05),phenotypes] != 0
vars <- m %>%
  colSums(., na.rm = TRUE) 

# variables with > 30 obs of value>0 
deviant_vars <- vars[vars > 30] %>% names()

# how many unique months do these have where FA_mean and var available?
count_unq_months <-  sapply(deviant_vars, function(x){ 
                    # how many unique months where both mr and var available?
                    return( d $month[!is.na(d$avg_MT05) & !is.na(d[,x])] %>% 
                             unique() %>%
                             length()
                            )
                  })
# vars with greater than one unique month observed 
MT5_greater_than_one_month <- count_unq_months[count_unq_months>1] %>% 
  names()

# setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0

# subset d with vars of interest 
model_df <- d %>% 
  select(month, patID_eye, patID, SEXF, all_of(MT5_greater_than_one_month))

# scale vars of interest
model_df[, MT5_greater_than_one_month[-which(MT5_greater_than_one_month == "avg_MT05")]] <- scale(model_df[, MT5_greater_than_one_month[-which(MT5_greater_than_one_month == "avg_MT05")]  ])


```

### model each variable against MT5 correcting for SEXF, and random effect of patient dependence
> modeling rate by  patient's first phenotype does not have time effect, and therefore no repeated eye measures; but there are still repeated patients because each has two eyes

```{r}

# exclude the response variable, mt5
input_phenotypes <- names(model_df)[!names(model_df) %in% c("month","patID_eye", "patID", "SEXF", "avg_MT05")]
  
# pheno <-  "FOVABNORM"

# for each variable in input_phenotypes, model paravoveal rate 
results_list <- list(); for (pheno in input_phenotypes){
  # model without month becasue variables already capture time as progression
  
  ## linear model
  # lm_formula <- as.formula( paste0("avg_MT05 ~ SEXF  +  ", pheno ))
  # print(paste("lm: ", paste0("avg_MT05 ~ SEXF  + ", pheno )))
  
  ## linear model with random effects
  re_formula <- as.formula( paste0("avg_MT05 ~ SEXF  +  ", pheno,  " + (1|patID_eye) + (1|patID)") )
  print(paste("random effects: ",paste0("avg_MT05 ~ SEXF  +  ", pheno,  "  + (1|patID_eye) + (1|patID)")))
  

  results_list[[pheno]] <- summary(lmer(formula = re_formula, data = model_df))
  # results_list[[pheno]] <- summary(lm(formula = lm_formula, data = model_df))
  
  
  
}


## make results df ####
results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
                         Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)

# fill df with model values
for (model_name in names(results_list)) {
  
  model_coef <- coef(results_list[[model_name]])
  
  a_result_df <- data.frame(Model = model_name, 
                         Term = rownames(model_coef), 
                         Estimate = model_coef[, "Estimate"],
                         Std.Error = model_coef[, "Std. Error"],
                         t.value = model_coef[, "t value"],
                         Pr = model_coef[, "Pr(>|t|)"])
  
  results_df <- rbind(results_df, a_result_df)
}

# filter control variables from results
results_df <- results_df %>% 
  filter(!Term %in% c("(Intercept)", "SEXF", "month", "patID_eye")) %>% 
  filter(!Model %in% c("patID_eye"))

# calculate adjusted p value
results_df$q_value <- p.adjust(results_df$Pr, method = "BH")


# filter out mt variables
results_df_filtered <- results_df %>% 
  filter(q_value>0) %>% 
  filter(!Model %in% 
           # c(paste("avg_MT0",c(1:4, 6:9), sep = ""), paste(paste("avg_MT0",c(1:4, 6:9), sep = ""), "_slope", sep = ""))
           
           # c(paste("avg_MT0",c(1:4, 6:9), sep = ""), paste(paste("avg_MT0",c(1:9), sep = ""), "_slope", sep = ""),  
           #   paste(paste("MT0",c(1:9), sep = ""), "_rate", sep = ""), 
           #   grep("CENTER|parafovea", results_df$Model, value = TRUE) , 
           #   paste(paste("avg_MT0",c(1:9), sep = ""), "_SD", sep = "")
           #   )
           grep("MT0|avg|CENTER|parafovea|slope", results_df$Model, value = TRUE)
         )



# volcano plot of vaa~phenotypes result
MT5_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("MT5 ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)")

ggsave(paste0(MT_modeling_path, "MT5_by_phenos_volcano.png"), MT5_by_phenos_volcano, width = 20, height = 12, dpi = 200)

MT5_by_phenos_volcano_filtered <-ggplot(results_df_filtered , aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("MT5 ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)")

ggsave(paste0(MT_modeling_path, "MT5_by_phenos_volcano_filtered.png"), MT5_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200)

```


## MT5 rate
### subset data to scale and model
* filter for the vars with (rows>0) > 30
```{r}

#  variables (rows of value != 0) where MT05_rate present
m <- d[!is.na(d$MT05_rate),phenotypes] != 0

# how many of each var is not 0
vars <- m %>%
  colSums(., na.rm = TRUE)

# variables with > 30 obs of value>0
powered_vars <- vars[vars > 30] %>% names()

# how many unique months do these have where FA_mean and var available?
count_unq_months <-  sapply(powered_vars, function(x){
                    
                    # how many unique months where both response variable and var available?
                    return( d$month[!is.na(d$MT05_rate) & !is.na(d[,x])] %>%
                             unique() %>%
                             length()
                            )
                  })

# vars with greater than one unique month observed
MT5_rate_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%
  names()


 
# # subset d with vars of interest
# model_df <- d %>%
#   select(month, patID_eye, patID, SEXF, all_of(MT5_rate_greater_than_one_month))
# 
# # scale vars of interest
# model_df[, MT5_rate_greater_than_one_month] <- scale(model_df[, MT5_rate_greater_than_one_month])

 
#************#************#************#************trying using the variables filtered for when modeling for mt5 thickness (has more vars)
# subset d with vars of interest
model_df <- d %>%
  filter(!is.na(MT05_rate) )%>% 
  select(month, patID_eye, patID, SEXF, all_of(MT5_greater_than_one_month))

# scale vars of interest
model_df[, MT5_greater_than_one_month[-which(MT5_greater_than_one_month == "MT05_rate")] ] <- scale(model_df[, MT5_greater_than_one_month[-which(MT5_greater_than_one_month == "MT05_rate")] ])
#************#************#************#************


# # reduce model_df to patient's rate and first_visit phenotype value
# model_df_first <- model_df %>% 
#   arrange(month) %>%  
#   group_by(patID_eye) %>%
#   # summarize(across(
#   #   .cols = !c("patID_eye", "month", "SEXF"), 
#   #   .fns = ~dplyr::first(.),
#   #   .names  = "first_{.col}")
#   #   ) %>%
#   mutate(across(
#     everything(), 
#     ~dplyr::first(.),
#     .names  = "first_{.col}"
#     )) 
# 
# # select columns for modeling
# model_df_first_filtered <- model_df_first %>% 
#   select( patID_eye, patID, SEXF, MT05_rate, starts_with("first")) %>%   
#   select(!matches("0[12346789]_rate")) %>% # exclude MT rates
#   select(!c("first_SEXF", "first_month", "first_patID")) %>% 
#   unique()

# make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows
first_values_df <- model_df %>% 
  select(-patID, - SEXF) %>% 
  arrange(patID_eye, month) %>%  
  group_by(patID_eye) %>%
  mutate(
    across(
        everything(), 
        ~dplyr::first(.[!is.na(.)]),
        .names = "first_{.col}"
  )) %>%
  ungroup() %>%
  select(-matches("month"), -matches("MT05_rate")) %>% 
  select(patID_eye,  starts_with("first_")) %>% 
  unique() %>% 
  arrange(patID_eye)

# filter out the firsts that don't make sense
model_df_first <- model_df %>% 
  arrange(patID_eye) %>% 
  select(patID_eye, patID, SEXF, MT05_rate) %>% 
  na.omit() %>%
  unique() %>%  
  left_join(., first_values_df)




```

### model each variable against MT5 correcting for SEXF, and random effect of patient dependence
> modeling rate by  patient's first phenotype does not have time effect, and therefore no repeated eye measures; but there are still repeated patients because each has two eyes

```{r}


# what phenos to sum
zonal_pheno_names <- grep("_.*0\\d$", names(model_df_first), value = TRUE) %>% 
    grep("avg",., value =  TRUE, invert = TRUE) %>% 
    gsub("0\\d",'', . ) %>% 
    unique()


# for each pheno name, sum the columns in model_df_first
for (zonal_pheno in zonal_pheno_names){
  
  new_name <- paste0(zonal_pheno, "_sum")
  model_df_first[[new_name]] <- model_df_first[, grep(zonal_pheno, names(model_df_first))] %>% 
    rowSums(.)
}


# exclude the response variable, mt5
input_phenotypes <- names(model_df_first)[!names(model_df_first) %in% c("month","patID_eye", "patID", "SEXF", "MT05_rate")]%>% 
  grep("0_sum", ., value = TRUE, invert = TRUE) # ignore existent sum variables

  
# pheno <-  "first_NEOVASC"

# for each variable in input_phenotypes, model paravoveal rate 
results_list <- list(); for (pheno in input_phenotypes){
  
  ## linear model with random effects
  re_formula <- as.formula( paste0("MT05_rate ~ SEXF  +  ", pheno,  "  + (1|patID)") )
  print(paste("random effects: ",paste0("MT05_rate ~ SEXF  +  ", pheno,  "  + (1|patID)")))
  
  # summary(lmer(formula = re_formula, data = model_df_first))

  results_list[[pheno]] <- summary(lmer(formula = re_formula, data = model_df_first))
  
}


## make results df ####
results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
                         Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)

# fill df with model values
for (model_name in names(results_list)) {
  
  model_coef <- coef(results_list[[model_name]])
  
  a_result_df <- data.frame(Model = model_name, 
                         Term = rownames(model_coef), 
                         Estimate = model_coef[, "Estimate"],
                         Std.Error = model_coef[, "Std. Error"],
                         t.value = model_coef[, "t value"],
                         Pr = model_coef[, "Pr(>|t|)"])
  
  results_df <- rbind(results_df, a_result_df)
}

# filter control variables from results
results_df <- results_df %>% 
  filter(!Term %in% c("(Intercept)", "SEXF", "month", "patID_eye")) %>% 
  filter(!Model %in% c("patID_eye"))

# calculate adjusted p value
results_df$q_value <- p.adjust(results_df$Pr, method = "BH")


# filter out mt variables
results_df_filtered <- results_df %>% 
  filter(!Model %in% 
           grep("avg|MT0|CENTERMEAN|parafovea", results_df$Model, value = TRUE)
         )



# volcano plot of vaa~phenotypes result
MT5_rate_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("MT5_rate ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)")

ggsave(paste0(MT_modeling_path, "MT5_rate_by_phenos_volcano.png"), MT5_rate_by_phenos_volcano, width = 20, height = 12, dpi = 200)

MT5_rate_by_phenos_volcano_filtered<-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("MT5_rate ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)")

ggsave(paste0(MT_modeling_path, "MT5_rate_by_phenos_volcano_filtered.png"), MT5_rate_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200)

MT5_rate_by_phenos_volcano_filtered_nominalP <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(Pr), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( Pr < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("MT5_rate ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)")

ggsave(paste0(MT_modeling_path, "MT5_rate_by_phenos_volcano_filtered_nominalP.png"), MT5_rate_by_phenos_volcano_filtered_nominalP , width = 20, height = 12, dpi = 200)





```





## parafoveal mean rate
### subset data to scale and model
* filter out the ones with (rows>0) < 30
```{r}


#  variables (rows of value!=0) < 30 where parafovea_rate present
m <- d[!is.na(d$parafovea_rate),phenotypes] != 0
vars <- m %>%
  colSums(., na.rm = TRUE)

# variables with > 30 obs of value != 0
deviant_vars <- vars[vars > 30] %>% names()

# how many unique months do these have where FA_mean and var available?
count_unq_months <-  sapply(deviant_vars, function(x){
                    # how many unique months where both mr and var available?
                    return( d $month[!is.na(d$parafovea_rate) & !is.na(d[,x])] %>%
                             unique() %>%
                             length()
                            )
                  })
# vars with greater than one unique month observed
parafovea_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%
  names()


# 
# # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0
# 
# # subset d with vars of interest 
# model_df <- d %>% 
#   select(month, patID_eye, patID, SEXF, all_of(parafovea_greater_than_one_month))
# 
# # scale vars of interest
# model_df[, parafovea_greater_than_one_month] <- scale(model_df[, parafovea_greater_than_one_month])


#************#************#************#************ trying using the variables filtered for when modeling for mt5
# subset d with vars of interest 
model_df <- d %>% 
  select(month, patID_eye, patID, SEXF, all_of(MT5_greater_than_one_month))

# scale vars of interest
model_df[, MT5_greater_than_one_month[-which(MT5_greater_than_one_month == "parafovea_rate")] ] <- scale(model_df[, MT5_greater_than_one_month[-which(MT5_greater_than_one_month == "parafovea_rate")]])
#************#************#************#************#

# # reduce model_df to patient's rate and first_visit phenotype value
# model_df_first <- model_df %>% 
#   arrange(month) %>%  
#   group_by(patID_eye) %>%
#   # summarize(across(
#   #   .cols = !c("patID_eye", "month", "SEXF"), 
#   #   .fns = ~dplyr::first(.),
#   #   .names  = "first_{.col}")
#   #   ) %>%
#   mutate(across(
#     everything(), 
#     ~dplyr::first(.),
#     .names  = "first_{.col}"
#     )) 
# 
# # select columns for modeling
# model_df_first_filtered <- model_df_first %>% 
#   select( patID_eye, patID, SEXF, parafovea_rate, starts_with("first")) %>%   
#   select(!matches("\\d_rate")) %>% 
#   select(!c("first_SEXF", "first_month", "first_parafovea_rate", "first_mean_parafoveal_rates", "first_patID")) %>% 
#   unique()

# make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows
first_values_df <- model_df %>% 
  select(-patID, - SEXF) %>% 
  arrange(patID_eye, month) %>%  
  group_by(patID_eye) %>%
  mutate(
    across(
        everything(), 
        ~dplyr::first(.[!is.na(.)]),
        .names = "first_{.col}"
  )) %>%
  ungroup() %>%
  select(-matches("month"), -matches("parafovea_rate")) %>% 
  select(patID_eye,  starts_with("first_")) %>% 
  unique() %>% 
  arrange(patID_eye)

# filter out the firsts that don't make sense
model_df_first <- model_df %>% 
  arrange(patID_eye) %>% 
  select(patID_eye, patID, SEXF, parafovea_rate) %>% 
  na.omit() %>%
  unique() %>%  
  left_join(., first_values_df)



```

### model each variable against parafovea rate, correcting for SEXF, and random effect of patient dependence
> modeling rate by  patient's first phenotype does not have time effect, and therefore no repeated eye measures; but there are still repeated patients because each has two eyes

```{r}

# what phenos to sum
zonal_pheno_names <- grep("_.*0\\d$", names(model_df_first), value = TRUE) %>% 
    grep("avg",., value =  TRUE, invert = TRUE) %>% 
    gsub("0\\d",'', . ) %>% 
    unique()

# zonal_pheno <- "first_IR"

# for each pheno name, sum the columns in model_df_first
for (zonal_pheno in zonal_pheno_names){
  
  new_name <- paste0(zonal_pheno, "_sum")
  model_df_first[[new_name]] <- model_df_first[, grep(zonal_pheno, names(model_df_first))] %>% 
    rowSums(.)
}

# exclude the response variable, parafovea_rate
input_phenotypes <- names(model_df_first)[!names(model_df_first) %in% c("month","patID_eye", "patID", "SEXF", "parafovea_rate")]
  
# for each variable in input_phenotypes, model paravoveal rate 
results_list <- list(); for (pheno in input_phenotypes){
  # model without month becasue variables already capture time as progression
  
  ## linear model
  # lm_formula <- as.formula( paste0("parafovea_rate ~ SEXF  +  ", pheno ))
  # print(paste("lm: ", paste0("parafovea_rate ~ SEXF  + ", pheno )))
  
  ## linear model with random effects
  re_formula <- as.formula( paste0("parafovea_rate ~ SEXF  +  ", pheno,  " + (1|patID)") )
  print(paste("random effects: ",paste0("parafovea_rate ~ SEXF  +  ", pheno,  "  + (1|patID)")))
  

  results_list[[pheno]] <- summary(lmer(formula = re_formula, data = model_df_first))
  # results_list[[pheno]] <- summary(lm(formula = lm_formula, data = model_df_first))
  
  
  
}


## make results df ####
results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
                         Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)

# fill df with model values
for (model_name in names(results_list)) {
  
  model_coef <- coef(results_list[[model_name]])
  
  a_result_df <- data.frame(Model = model_name, 
                         Term = rownames(model_coef), 
                         Estimate = model_coef[, "Estimate"],
                         Std.Error = model_coef[, "Std. Error"],
                         t.value = model_coef[, "t value"],
                         Pr = model_coef[, "Pr(>|t|)"])
  
  results_df <- rbind(results_df, a_result_df)
}

# calculate adjusted p value
results_df$q_value <- p.adjust(results_df$Pr, method = "BH")

# filter control variables from results
results_df <- results_df %>% 
  filter(!Term %in% c("(Intercept)", "SEXF", "month", "patID_eye")) %>% 
  filter(!Model %in% c("patID_eye"))

# filter non zone 5 variables from df
results_df_otherZones_filtered <- results_df %>% 
  filter(!Model %in% 
           c(
             grep(".*0[12346789]$", results_df$Model, value = TRUE)
             ,
             grep(".*0[12346789]_slope$", results_df$Model, value = TRUE)
             )
         )

# filter non zone 5 variables from df
results_df_MT_filtered <- results_df %>% 
  filter(!Model %in% grep("MT0|avg|CENTERMEAN", results_df$Model, value = TRUE ))



# volcano plot of vaa~phenotypes result
parafovea_rate_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("parafovea_rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(MT_modeling_path, "parafovea_rate_by_phenos_volcano.png"), parafovea_rate_by_phenos_volcano, width = 20, height = 12, dpi = 200)

# volcano plot of vaa~phenotypes result
parafovea_rate_by_phenos_volcano_otherZones_filtered<- ggplot(results_df_otherZones_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("parafovea_rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(MT_modeling_path, "parafovea_rate_by_phenos_volcano_otherZones_filtered.png"), parafovea_rate_by_phenos_volcano_otherZones_filtered, width = 20, height = 12, dpi = 200)

# volcano plot of vaa~phenotypes result
parafovea_rate_by_phenos_volcano_MT_filtered<- ggplot(results_df_MT_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("parafovea_rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(MT_modeling_path, "parafovea_rate_by_phenos_volcano_MT_filtered.png"), parafovea_rate_by_phenos_volcano_MT_filtered, width = 20, height = 12, dpi = 200)

parafovea_rate_by_phenos_volcano_MT_filtered_nominalP<- ggplot(results_df_MT_filtered, aes(x = Estimate, y = -log10(Pr), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( Pr < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("parafovea_rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(MT_modeling_path, "parafovea_rate_by_phenos_volcano_MT_filtered_nominalP.png"), parafovea_rate_by_phenos_volcano_MT_filtered_nominalP, width = 20, height = 12, dpi = 200)






```


### & how are the variables filtered differently among these models?
```{r eval=FALSE, include=FALSE}
MT5_rate_greater_than_one_month

MT5_greater_than_one_month
setdiff(MT5_rate_greater_than_one_month, MT5_greater_than_one_month)
setdiff(MT5_greater_than_one_month, MT5_rate_greater_than_one_month)


pheno_list <- list(MT5_rate_greater_than_one_month, MT5_greater_than_one_month, parafovea_greater_than_one_month)

display_venn <- function(x, ...){
  library(VennDiagram)
  grid.newpage()
  venn_object <- venn.diagram(x, filename = NULL, ...)
  grid.draw(venn_object)
}


display_venn(
        pheno_list,
        category.names = c("MT5_rate_greater_than_one_month", "MT5_greater_than_one_month", "parafovea_greater_than_one_month"),
        # Circles
        lwd = 2,
        lty = 'blank',
        fill = c("#999999", "#E69F00", "#56B4E9"),
        # Numbers
        cex = .9,
        fontface = "italic",
        # Set names
        cat.cex = 1,
        cat.fontface = "bold",
        cat.default.pos = "outer",
        cat.pos = c(-30, 0, 30), # Adjust these angles to better associate labels with the shapes
        cat.dist = c(0.055, 0.055, 0.055), # Adjust the distances if necessary to move labels closer or further from the circles
        cat.col = c("#999999", "#E69F00", "#56B4E9") 
)

# whats in MT5 thats not in others
setdiff(MT5_greater_than_one_month, c(MT5_rate_greater_than_one_month, parafovea_greater_than_one_month))
#  "IR02"  "BRV02" "DRV02" "DRV04"

# whats in MT5_rate thats not in other
setdiff(c(MT5_rate_greater_than_one_month), parafovea_greater_than_one_month)
# "NEOVASC"  "RETHAEM"  "IR03"     "BRV01"    "VTV02"    "ICL_TYPE"

```
> use MT5 variables


#___   ________   _____   ______
#***
# model slope

## *slope 
### subset data to scale and model
* filter out the ones with (rows>0) < 30
```{r}

#  variables (rows of value>0) < 30 where avg_MT_slope_sans3 present
m <- d[!is.na(d$avg_MT_slope_sans3),phenotypes] != 0

vars <- m %>%
  colSums(., na.rm = TRUE) 

# variables with > 30 obs of value>0 
deviant_vars <- vars[vars > 30] %>% names()

# how many unique months do these have where FA_mean and var available?
count_unq_months <-  sapply(deviant_vars, function(x){ 
                    # how many unique months where both mr and var available?
                    return( d $month[!is.na(d$avg_MT_slope_sans3) & !is.na(d[,x])] %>% 
                             unique() %>%
                             length()
                            )
                  })
# vars with greater than one unique month observed 
slope_greater_than_one_month <- count_unq_months[count_unq_months>1] %>% 
  names()

# setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0

# subset d with vars of interest 
model_df <- d %>% 
  select(month, patID_eye, patID, SEXF, all_of(slope_greater_than_one_month))

# scale vars of interest, except the response
model_df[, slope_greater_than_one_month[-which(slope_greater_than_one_month == "avg_MT_slope_sans3")] ] <- scale(model_df[, slope_greater_than_one_month[-which(slope_greater_than_one_month == "avg_MT_slope_sans3")] ])

```

### model each variable against slope correcting for SEXF, and random effect of patient dependence

```{r}

# exclude the response variable, mt5
input_phenotypes <- names(model_df)[!names(model_df) %in% c("month","patID_eye", "patID", "SEXF", "avg_MT_slope_sans3")]
  
# pheno <-  "NEOVASC"

# for each variable in input_phenotypes, model paravoveal rate 
results_list <- list(); for (pheno in input_phenotypes){
  
  ## linear model with random effects
  re_formula <- as.formula( paste0("avg_MT_slope_sans3 ~ SEXF  +  ", pheno,  " + (1|patID_eye) + (1|patID)") )
  print(paste("random effects: ",paste0("avg_MT_slope_sans3 ~ SEXF  +  ", pheno,  "  + (1|patID_eye) + (1|patID)")))
  

  results_list[[pheno]] <- summary(lmer(formula = re_formula, data = model_df))
  
}


## make results df ####
results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
                         Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)

# fill df with model values
for (model_name in names(results_list)) {
  
  model_coef <- coef(results_list[[model_name]])
  
  a_result_df <- data.frame(Model = model_name, 
                         Term = rownames(model_coef), 
                         Estimate = model_coef[, "Estimate"],
                         Std.Error = model_coef[, "Std. Error"],
                         t.value = model_coef[, "t value"],
                         Pr = model_coef[, "Pr(>|t|)"])
  
  results_df <- rbind(results_df, a_result_df)
}

# filter control variables from results
results_df <- results_df %>% 
  filter(!Term %in% c("(Intercept)", "SEXF", "month", "patID_eye")) %>% 
  filter(!Model %in% c("patID_eye"))

# calculate adjusted p value
results_df$q_value <- p.adjust(results_df$Pr, method = "BH")


# filter out mt variables
results_df_filtered <- results_df %>% 
  filter(!Model %in% grep("avg|CENTER|MT0\\d", results_df$Model, value = TRUE)
           # c(paste("avg_MT0",c(1:4, 6:9), sep = ""), paste(paste("avg_MT0",c(1:4, 6:9), sep = ""), "_slope", sep = ""))
           # c(paste("avg_MT0",c(1:4, 6:9), sep = ""), 
           #   paste(paste("avg_MT0",c(1:9), sep = ""), "_slope", sep = ""),  
           #   paste(paste("MT0",c(1:9), sep = ""), "_rate", sep = ""), 
           #   paste(paste("avg_MT0",c(1:9), sep = ""), "_SD", sep = ""), "CENTERMEAN"
           #   )
         )



# volcano plot of vaa~phenotypes result
# MT5_slope_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+
#       geom_point(show.legend = FALSE)+
#       geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
#                        size = 4,
#                        nudge_y = 0.5,
#                        show.legend = FALSE,
#                        max.overlaps = 20)+
#       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
#       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
#       theme_bw()+
#       # facet_wrap(~quintile, scales = "fixed")+
#       ggtitle("MT5 ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)")
# 
# ggsave(paste0(MT_modeling_path, "MT5_by_phenos_volcano.png"), MT5_slope_by_phenos_volcano, width = 20, height = 12, dpi = 200)

avg_MT_slope_sans3_by_phenos_volcano_filtered <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("avg_MT_slope_sans3 (-zone3) ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)")

ggsave(paste0(MT_modeling_path, "avg_MT_slope_sans3_by_phenos_volcano_filtered.png"), avg_MT_slope_sans3_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200)





```


## *slope rate 
 ### subset data to scale and model
* filter out the ones with (rows>0) < 30
```{r}


#  variables (rows of value!=0) < 30 where avg_MT_slope_sans3_rate present
m <- d[!is.na(d$avg_MT_slope_sans3_rate) , phenotypes] != 0
vars <- m %>%
  colSums(., na.rm = TRUE)

# variables with > 30 obs of value != 0
deviant_vars <- vars[vars > 30] %>% names()

# how many unique months do these have where FA_mean and var available?
count_unq_months <-  sapply(deviant_vars, function(x){
                    # how many unique months where both mr and var available?
                    return( d $month[!is.na(d$avg_MT_slope_sans3_rate) & !is.na(d[,x])] %>%
                             unique() %>%
                             length()
                            )
                  })
# vars with greater than one unique month observed
slope_rate_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%
  names()


# subset d with vars of interest 
model_df <- d %>% 
  select(month, patID_eye, patID, SEXF, all_of(slope_rate_greater_than_one_month))

# scale vars of interest
model_df[, slope_rate_greater_than_one_month[-which(slope_rate_greater_than_one_month == "avg_MT_slope_sans3_rate")]] <- scale(model_df[, slope_rate_greater_than_one_month[-which(slope_rate_greater_than_one_month == "avg_MT_slope_sans3_rate")]])

# make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows
first_values_df <- model_df %>% 
  select(-patID, - SEXF) %>% 
  arrange(patID_eye , month) %>%  
  group_by(patID_eye) %>%
  mutate(
    across(
        everything(), 
        ~dplyr::first(.[!is.na(.)]),
        .names = "first_{.col}"
  )) %>%
  ungroup() %>%
  select(-matches("month"), -matches("slope_rate")) %>% 
  select(patID_eye,  starts_with("first_")) %>% 
  unique() %>% 
  arrange(patID_eye)
  
# filter out the firsts that don't make sense
model_df_first <- model_df %>% 
  arrange(patID_eye) %>% 
  select(patID_eye, patID, SEXF, avg_MT_slope_sans3_rate) %>% 
  na.omit() %>%
  unique() %>%  
  left_join(., first_values_df)



```

### model each variable against slope rate, correcting for SEXF, and random effect of patient dependence
```{r}

# what phenos to sum
zonal_pheno_names <- grep("_.*0\\d$", names(model_df_first), value = TRUE) %>% 
    grep("avg",., value =  TRUE, invert = TRUE) %>% 
    gsub("0\\d",'', . ) %>% 
    unique()

# zonal_pheno <- "first_IR"

# for each pheno name, sum the columns in model_df_first
for (zonal_pheno in zonal_pheno_names){
  
  new_name <- paste0(zonal_pheno, "_sum")
  model_df_first[[new_name]] <- model_df_first[, grep(zonal_pheno, names(model_df_first))] %>% 
    rowSums(.)
}

# exclude the response variable, avg_MT_slope_sans3_rate
input_phenotypes <- names(model_df_first)[!names(model_df_first) %in% c("month","patID_eye", "patID", "SEXF", "avg_MT_slope_sans3_rate")]
  
# for each variable in input_phenotypes, model paravoveal rate 
results_list <- list(); for (pheno in input_phenotypes){
  
  ## linear model with random effects
  re_formula <- as.formula( paste0("avg_MT_slope_sans3_rate ~ SEXF  +  ", pheno,  " + (1|patID)") )
  print(paste("random effects: ",paste0("avg_MT_slope_sans3_rate ~ SEXF  +  ", pheno,  "  + (1|patID)")))
  

  results_list[[pheno]] <- summary(lmer(formula = re_formula, data = model_df_first))
  
  
  
  
}


## make results df ####
results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
                         Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)

# fill df with model values
for (model_name in names(results_list)) {
  
  model_coef <- coef(results_list[[model_name]])
  
  a_result_df <- data.frame(Model = model_name, 
                         Term = rownames(model_coef), 
                         Estimate = model_coef[, "Estimate"],
                         Std.Error = model_coef[, "Std. Error"],
                         t.value = model_coef[, "t value"],
                         Pr = model_coef[, "Pr(>|t|)"])
  
  results_df <- rbind(results_df, a_result_df)
}

# calculate adjusted p value
results_df$q_value <- p.adjust(results_df$Pr, method = "BH")

# filter control variables from results
results_df <- results_df %>% 
  filter(!Term %in% c("(Intercept)", "SEXF", "month", "patID_eye")) %>% 
  filter(!Model %in% c("patID_eye"))

# filter non zone 5 variables from df
results_df_otherZones_filtered <- results_df %>% 
  filter(!Model %in% 
           c(
             grep(".*0[12346789]$", results_df$Model, value = TRUE)
             ,
             grep(".*0[12346789]_slope$", results_df$Model, value = TRUE)
             )
         )

# filter non zone 5 variables from df
results_df_MT_filtered <- results_df %>% 
  filter(!Model %in% grep("MT0|avg|CENTERMEAN", results_df$Model, value = TRUE ))


# 
# volcano plot of vaa~phenotypes result
avg_MT_slope_sans3_rate_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("avg_slope rate ~ SEXF + phenotype_i + (1| patID)")

# ggsave(paste0(MT_modeling_path, "MT5_slope_rate_by_phenos_volcano.png"), MT5_slope_rate_by_phenos_volcano, width = 20, height = 12, dpi = 200)
# 
# # volcano plot of vaa~phenotypes result
# MT5_slope_rate_by_phenos_volcano_otherZones_filtered<- ggplot(results_df_otherZones_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+
#       geom_point(show.legend = FALSE)+
#       geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
#                        size = 4,
#                        nudge_y = 0.5,
#                        show.legend = FALSE,
#                        max.overlaps = 20)+
#       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
#       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
#       theme_bw()+
#       # facet_wrap(~quintile, scales = "fixed")+
#       ggtitle("MT5_slope rate ~ SEXF + phenotype_i + (1| patID)")
# 
# ggsave(paste0(MT_modeling_path, "MT5_slope_rate_by_phenos_volcano_otherZones_filtered.png"), MT5_slope_rate_by_phenos_volcano_otherZones_filtered, width = 20, height = 12, dpi = 200)

# volcano plot of vaa~phenotypes result
avg_MT_slope_sans3_rate_by_phenos_volcano_MT_filtered<- ggplot(results_df_MT_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("avg_slope rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(MT_modeling_path, "avg_MT_slope_sans3_rate_by_phenos_volcano_MT_filtered.png"), avg_MT_slope_sans3_rate_by_phenos_volcano_MT_filtered, width = 20, height = 12, dpi = 200)

avg_MT_slope_sans3_rate_by_phenos_volcano_MT_filtered_nominalP<- ggplot(results_df_MT_filtered, aes(x = Estimate, y = -log10(Pr), color = Model))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( Pr < .05,  Model, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("avg_slope rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(MT_modeling_path, "avg_MT_slope_sans3_rate_by_phenos_volcano_MT_filtered_nominalP.png"), avg_MT_slope_sans3_rate_by_phenos_volcano_MT_filtered_nominalP, width = 20, height = 12, dpi = 200)

```
