---
title: "08_functionAndStructureModeling"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---

# options
```{r}
cat("\nRUNNING 10_function_structuere_modeling\n\n")
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()
```


# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

pacman::p_load(tidyr,dplyr,reshape2,purrr,GGally,tidyverse,ggpubr,vcd,UpSetR,ggplot2,gridExtra,ComplexHeatmap,viridis,knitr,htmlwidgets,zoo,corrplot,RColorBrewer,openxlsx,plotly,RColorBrewer,mgcv,zoo,tools,mice,grDevices,ggrepel,networkD3,htmltools,grid,gplots,Metrics,psych,car,stats,lme4,lmerTest,VennDiagram,ape,SingleCellExperiment,parallel, HDF5Array)



rm(list = ls())
gc()
```

# functions
```{r}
source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```


# Load Data
info and d from sce
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

# load experiment
sce <-  HDF5Array::loadHDF5SummarizedExperiment(sprintf("%sprocessed_data/sce002.2.h5", wd_path))

raw <- metadata(sce)$raw

```



# make new directorys for plots
```{r}

path <- paste0(wd_path, "results/modeling/")
if(!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
  
}


VA_modeling_path <- paste0(wd_path, "results/modeling/VA/")
if(!dir.exists(VA_modeling_path)) {
  dir.create(VA_modeling_path, recursive = TRUE)
  
}

MT_modeling_path <- paste0(wd_path, "results/modeling/MT/")
if(!dir.exists(MT_modeling_path)) {
  dir.create(MT_modeling_path, recursive = TRUE)
  
}

```

# clean data


```{r, look at dims}

check_dims <- function(){
  
  D <- as.data.frame(colData(sce)) %>% 
  left_join(
    as.data.frame(t(assays(sce)$scaled_measure_vars)) %>% 
      rownames_to_column("sample")
  ) 
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()
# dims : 12525 143
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ no change

### APRIL 2 2025 ###
# dims : 12276 179
# n_patIDs             3971
# n_patID_eyes         7888
# n_observations      12275
# n_patIDs_w_gr_1_vis   492

```

## make year
```{r}

sce$year <- round(sce$month/12, 2)

```


## derive t2d from other DM vars
```{r}

#key of dm vars
colData(sce)[, c("DM", "DMTYPE", 'DMTYPEAU')] %>% 
  as.data.frame() %>% 
  group_by(across(everything())) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  arrange(count)
#   DM    DMTYPE  DMTYPEAU count
# 1 NA    NA      Type I       4
# 2 Yes   NA      NA          18
# 3 Yes   Type I  NA          28
# 4 NA    NA      Type II    180
# 5 Yes   Type II NA        1870
# 6 No    NA      NA        4007
# 7 NA    NA      NA        6169

colData(sce)[, grep("DM", names(colData(sce)), value = TRUE)] %>%
  as.matrix() %>% 
  mice::md.pattern(rotate.names = T)

# Start with NA
sce$t2d <- NA_integer_

# Set t2d = 1 if either field says Type II
sce$t2d[which(sce$DMTYPE == "Type II" | sce$DMTYPEAU == "Type II")] <- 1

# Set t2d = 0 if all available fields say no evidence of T2D and they are not missing
sce$t2d[which(
  (sce$DM == "No" & !is.na(sce$DM)) |
  (sce$DMTYPE != "Type II" & !is.na(sce$DMTYPE)) |
   (sce$DMTYPEAU != "Type II" & !is.na(sce$DMTYPEAU))
)] <- 0

table(sce$t2d, useNA = "always")
#    0    1 <NA> 
# 4039 2050 6187 

```


## filter relations to proband
- if probandfamily/solo keep
- if family solo with mactel keep
```{r}



# make one relatproband var in da
relatprobnd <- data.frame(colData(sce)) %>% 
  select(NHOS_RELATPROBND,NHOR_RELATPROBND) %>% 
  mutate_all( ~ifelse(grepl("\\d+", .), NA, .))%>% 
  mutate(RELATPROBND = coalesce(.[[1]] , .[[2]]))

colData(sce)$RELATPROBND <-relatprobnd$RELATPROBND

d <- assays(sce)$pmm_imp_of_scaled %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  left_join( colData(sce) %>% data.frame())

family_vars <- c("patID", "FAMILY_ID","FAMILY_STATUS", "RELATPROBND")
d$FAMILY_STATUS %>% table()
d$FAMILY_ID %>% table()
d$RELATPROBND %>% table()


# fill relation with familystatus where relation is missing
d$RELATPROBND[d$FAMILY_STATUS == "familyMember" & is.na(d$RELATPROBND)] <- "non-specified family"

d$RELATPROBND[is.na(d$FAMILY_STATUS) ] # relation missing where family status missing

# make logic for rows with proband
d$is.proband <- ifelse(d$FAMILY_STATUS %in% c("probandSolo", "probandWithFamily"), TRUE, FALSE)

# where proband, what is the relationshiop?
d$RELATPROBND[d$is.proband] %>% table()
# 3  Brother Daughter   Father   Mother   Sister are relationships when there should be none (assumedly) << only "3" when filtering out the NHOS pats

# make relation NA where is.proband because, e.g., brother of self?
d$RELATPROBND[d$is.proband] <- NA

# what patients are there where not proband and not a part of a proband's family
proband_familes <- d$FAMILY_ID[d$is.proband]
probandless_families <- setdiff(d$FAMILY_ID, proband_familes)

pseudo_probands <- d[d$FAMILY_ID %in% probandless_families , family_vars] %>% unique()
# ^ these relations do not have a proband to be related to

# family that shows up >1 times
frequent_probandless_family <- table(pseudo_probands$FAMILY_ID)[table(pseudo_probands$FAMILY_ID) > 1] %>% names()

# U77-MURPHY shows up 3 times, what's the deal?
d[d$FAMILY_ID%in%frequent_probandless_family , ] %>% glimpse
frequent_probandless_family_patients <- d$patID[d$FAMILY_ID%in%frequent_probandless_family ] %>% 
  unique()
# ^ three patients in this family and no proband

# which of these patients has most data and visits?
d %>% 
  filter(patID %in% frequent_probandless_family_patients) %>% 
  group_by(patID_eye) %>% 
  summarize(n_vis = n_distinct(visitID))
# ^ all have one visit

# sum of numeric variables
pat_colsums <- sapply(frequent_probandless_family_patients, function(pat){
    d %>% 
      filter(patID == pat) %>%
      select(which(sapply(d, is.numeric))) %>% 
      colSums(., na.rm = TRUE)
  
}) %>% 
  as.data.frame

# which pat has largest sum (worst disease)
best_pat <- names(pat_colsums)[colSums(pat_colsums)==max(colSums(pat_colsums))] #  025098 U77-MURPHY    familyMember  Brother   

# make pseudo_probands excluding frequentprobandlessfamily - best_pat ( is best pat or is not frequent family)
patients_to_proband <- pseudo_probands$patID[pseudo_probands$patID == best_pat | pseudo_probands$FAMILY_ID != frequent_probandless_family ]

d$is.proband[d$patID %in% patients_to_proband] <-  TRUE

# how many non-probands? any missing?
table(d$is.proband)
# FALSE  TRUE << filtering out NHOS pats
#    86  5506 
# FALSE  TRUE 
#   596 11680 

# remove non-proband patients
d <- d[d$is.proband, ] 

```

## Make VA rates
```{r}

# difference from first over difference in time, ignoring NAs in . and month
d <- d %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  mutate(across(matches("FA\\d$|FA_mean$|ACUIT|LOGMAR|SNELL"), ~{
    (dplyr::last(.[!is.na(.)])-dplyr::first(.[!is.na(.)]))/
      (dplyr::last(month[!is.na(.)])-dplyr::first(month[!is.na(.)]))
    }, 
    .names = "{.col}_rate"
    )
  )


```

## remove FA vars
```{r}

colData(sce)[ ,  grep("FA_|FA\\d", names(colData(sce))) ] <- NULL

```


## Make MT.rates
```{r}

names(d) %>% grep("MT0|parafovea", ., value = TRUE)

d <- d %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  mutate(across(matches("MT0|parafovea"), ~{
    (dplyr::last(.)-dplyr::first(.))/
      (dplyr::last(month)-dplyr::first(month))
    }, 
    .names = "{.col}_rate"
    )
  )


```


# _______________________
# what phenotypes to model?
*ensure numeric
```{r}

phenotypes <-  c(rownames(sce), grep("SNELL|ACUIT|LOGMAR|chew_grade_derived_mine_loose|RW|disease|t2d|SMKCIGSDAY|WGTKG", names(colData(sce)) , value = TRUE, ignore.case = TRUE))

# make phenotypes numeric in d
d <- d %>% 
  mutate(across(all_of(phenotypes),
                ~ as.numeric(.)))

# remove skewed vars
bp <- abs(sapply(d[, phenotypes], skew)) %>% 
  sort() %>% 
  boxplot()

cat("\n")
cat(paste(names(bp$out), collapse = ", "))
# HRDEXUDATE, RW144, RAV01, YELLOWRND, ICL03, RAV04, ICL01, RPE_LK06, ICL04

phenotypes <- phenotypes[!phenotypes %in% names(bp$out)]

```


<!-- #__  ___   ___   __ -->
<!-- # modeling vision loss -->
<!-- # *** -->

<!-- #$$ -->
<!-- # * raw values -->
<!-- ## VA_mean -->
<!-- #### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->


<!-- # #  variables (rows of value>0) < 30 where FA_mean present -->
<!-- # m <- d[!is.na(d$FA_mean),phenotypes] != 0 -->
<!-- # vars <- m %>% -->
<!-- #   colSums(., na.rm = TRUE)  -->
<!-- #  -->
<!-- # # variables with > 30 obs of value>0  -->
<!-- # deviant_vars <- vars[vars > 30] %>% names() -->
<!-- #  -->
<!-- # # how many unique months do these have where FA_mean and var available? -->
<!-- # count_unq_months <-  sapply(deviant_vars, function(x){  -->
<!-- #                     # how many unique months where both FA and var available? -->
<!-- #                     return( d $month[!is.na(d$FA_mean) & !is.na(d[,x])] %>%  -->
<!-- #                              unique() %>% -->
<!-- #                              length() -->
<!-- #                             ) -->
<!-- #                   }) -->
<!-- #  -->
<!-- # # vars with greater than one unique month observed  -->
<!-- # VA_mean_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%  -->
<!-- #   names() -->

<!-- # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0 -->

<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   # select(month, patID_eye, patID, SEXF, all_of(VA_mean_greater_than_one_month)) -->
<!--   select(month, patID_eye, patID, SEXF, all_of(phenotypes)) -->

<!-- response <- 'FA_mean' -->
<!-- # scale vars of interest -->
<!-- model_df[, phenotypes[-which(phenotypes == response)]] <- scale( -->
<!--   model_df[, phenotypes[-which(phenotypes == response)]] -->
<!--   ) -->


<!-- ``` -->


<!-- <!-- ```{r, FDR exercise} --> -->
<!-- <!-- # Rank p-values from smallest to largest. --> -->
<!-- <!-- # Compare each p-value to an increasing threshold (i/m)α, where i is the rank, m is the total number of tests, and α is the desired FDR level. --> -->
<!-- <!-- # Reject hypotheses for p-values below this threshold. --> -->
<!-- <!-- a <- 0.05 --> -->
<!-- <!-- set.seed(42) --> -->
<!-- <!-- p_null <- runif(500)             # True nulls (uniform) --> -->
<!-- <!-- p_alt <- rbeta(250, 0.5, 1)      # False nulls (skewed to 0) --> -->
<!-- <!-- p_values <- sort(c(p_null, p_alt), decreasing = F) --> -->
<!-- <!-- plot(density(p_values)) --> -->
<!-- <!-- r <- rank(p_values) --> -->
<!-- <!-- p_sim <- (r/length(r))*a --> -->
<!-- <!-- plot(density(p_sim)) --> -->

<!-- <!-- z <- data.frame(P = round(p_values,3), FDR = round(p_sim,3), Sig =  p_sim<=0.05) --> -->


<!-- <!-- ``` --> -->

<!-- #### model each variable against FA_mean, correcting for SEXF, and random effect eye and patient -->
<!-- ```{r} -->

<!-- # exclude the response variable, FA_mean and correction variables -->
<!-- input_phenotypes <- names(model_df)[-which(names(model_df) %in%  -->
<!--                                              c("FA_mean", "FA_mean_rate",  -->
<!--                                                 "month","patID_eye","patID","SEXF"))  -->
<!--                                     ] -->

<!-- # for each variable in input_phenotypes, model FA_mean  -->
<!-- formulas_char <- paste0(response," ~ ", input_phenotypes, " + SEXF + (1|patID_eye) + (1 | patID)")  -->

<!-- results_list <- mclapply(formulas_char, function(form_char){ -->

<!--   formula <- as.formula( form_char) -->

<!--   fit <- try({ -->
<!--           lmer(formula, data = model_df) -->
<!--         }) -->
<!--   if(inherits(fit, "try-error")){ -->
<!--     warning(paste0("FAILURE!!!\n",form_char,"\n")) -->
<!--   } -->

<!--   return(summary(fit)) -->

<!-- }, mc.cores = detectCores()-1) -->

<!-- # simplify names to response and predictor (replace a space followed by a + with anything after with nothing) -->
<!-- names(results_list) <- sub("\\s\\+.*", "",formulas_char) -->

<!-- # filter out models that didnt work -->
<!-- faulty_models <- sapply(results_list,function(x) all(class(x) %in% c("summaryDefault","table")) ) %>%  -->
<!--   Filter(isTRUE,.) %>%  -->
<!--   names() -->

<!-- results_list <- results_list[!names(results_list) %in% faulty_models] -->

<!-- # fill df with model values -->
<!-- results_df <- extract_results(results_list) -->


<!-- # filter control variables from results -->
<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) %>%  -->
<!--   filter(q_value>0) -->


<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Model %in% grep("MT|MT0|RW\\d+|FA\\d|SEX", results_df$Model, value = TRUE)) %>%  -->
<!--   filter(!Model %in% c("mean_parafoveal_rates")) -->

<!-- ### plot #### -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- VA_by_phenos_volcano <- ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_mean ~ SEXF + phenotype_i + (1|patID_eye) + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "VA_by_phenos_volcano.png"), VA_by_phenos_volcano, width = 20, height = 12, dpi = 200) -->


<!-- # volcano plot of vaa~phenotypes result -->
<!-- VA_by_phenos_volcano_filtered <- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_mean ~ SEXF + phenotype_i + (1|patID_eye) + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "VA_by_phenos_volcano_filtered.png"), VA_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200) -->



<!-- ``` -->

<!-- ## FA1  -->

<!-- #### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->


<!-- # #  variables (rows of value>0) < 30 where FA1 present -->
<!-- # m <- d[!is.na(d$FA1),phenotypes] != 0 -->
<!-- # vars <- m %>% -->
<!-- #   colSums(., na.rm = TRUE)  -->
<!-- #  -->
<!-- # # variables with > 30 obs of value>0  -->
<!-- # deviant_vars <- vars[vars > 30] %>% names() -->
<!-- #  -->
<!-- # # how many unique months do these have where FA1 and var available? -->
<!-- # count_unq_months <-  sapply(deviant_vars, function(x){  -->
<!-- #                     # how many unique months where both FA and var available? -->
<!-- #                     return( d $month[!is.na(d$FA1) & !is.na(d[,x])] %>%  -->
<!-- #                              unique() %>% -->
<!-- #                              length() -->
<!-- #                             ) -->
<!-- #                   }) -->
<!-- #  -->
<!-- # # vars with greater than one unique month observed  -->
<!-- # FA1_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%  -->
<!-- #   names() -->

<!-- # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0 -->

<!-- # # subset d with vars of interest  -->
<!-- # model_df <- d %>%  -->
<!-- #   select(month, patID_eye, patID, SEXF, FA1, all_of(FA1_greater_than_one_month)) -->
<!-- #  -->
<!-- # # scale vars of interest -->
<!-- # model_df[, FA1_greater_than_one_month] <- scale(model_df[, FA1_greater_than_one_month]) -->


<!-- # subset d with vars of interest  -->
<!-- response <- 'FA1' -->

<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF, all_of(c(response,phenotypes))) -->

<!-- # scale vars of interest -->
<!-- model_df[, phenotypes[-which(phenotypes == response)]] <- scale( -->
<!--   model_df[, phenotypes[-which(phenotypes == response)]]) -->

<!-- ``` -->

<!-- #### model FA1, correcting for SEXF, month, and random effect eye -->
<!-- ```{r} -->

<!-- # exclude the response variable, FA1 -->
<!-- input_phenotypes <- names(model_df)[-which(names(model_df) %in% c("FA1", "FA1_rate", "month","patID_eye","patID","SEXF" ))] -->

<!-- # for each variable in input_phenotypes, model FA1  -->
<!-- formulas_char <- paste0(response," ~ ", input_phenotypes, " + SEXF + (1|patID_eye) + (1 | patID)")  -->

<!-- # model parrallel -->
<!-- results_list <- mclapply(formulas_char, function(form_char){ -->

<!--   formula <- as.formula( form_char) -->

<!--   fit <- try({ -->
<!--           lmer(formula, data = model_df) -->
<!--         }) -->
<!--   if(inherits(fit, "try-error")){ -->
<!--     warning(paste0("FAILURE!!!\n",form_char,"\n")) -->
<!--   } -->

<!--   return(summary(fit)) -->

<!-- }, mc.cores = detectCores()-1) -->

<!-- # simplify names to response and predictor (replace a space followed by a + with anything after with nothing) -->
<!-- names(results_list) <- sub("\\s\\+.*", "",formulas_char) -->

<!-- # filter out models that didnt work -->
<!-- faulty_models <- map(results_list,function(x) all(class(x) %in% c("summaryDefault","table")) ) %>%  -->
<!--   Filter(isTRUE,.) %>%  -->
<!--   names() -->

<!-- results_list <- results_list[!names(results_list)%in%faulty_models ] -->

<!-- # fill df with model values -->
<!-- results_df <- extract_results(results_list) -->

<!-- # filter control variables from results -->
<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->

<!-- # filter out unwanted variables -->
<!-- results_df <- results_df %>%  -->
<!--   filter(!Model %in% c("FA_mean")) -->

<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Term %in% c(grep("MT|MT0|RW\\d|FA_m|FA\\d|SEX", results_df$Term, value =TRUE), "mean_parafoveal_rates" ) ) -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- FA1_by_phenos_volcano <- ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_1 ~ SEXF + phenotype_i + (1|patID_eye) + (1 | patID)") -->

<!-- ggsave(paste0( VA_modeling_path, "FA1_by_phenos_volcano.png"), FA1_by_phenos_volcano, width = 20, height = 12, dpi = 200) -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- FA1_by_phenos_volcano_filtered <- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_1 ~ SEXF + phenotype_i + (1|patID_eye) + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "FA1_by_phenos_volcano_filtered.png"), FA1_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200) -->

<!-- ``` -->


<!-- ## FA2 -->

<!-- #### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->

<!-- #  -->
<!-- # #  variables (rows of value>0) < 30 where FA2 present -->
<!-- # m <- d[!is.na(d$FA2),phenotypes] != 0 -->
<!-- # vars <- m %>% -->
<!-- #   colSums(., na.rm = TRUE)  -->
<!-- #  -->
<!-- # # variables with > 30 obs of value>0  -->
<!-- # deviant_vars <- vars[vars > 30] %>% names() -->
<!-- #  -->
<!-- # # how many unique months do these have where FA2 and var available? -->
<!-- # count_unq_months <-  sapply(deviant_vars, function(x){  -->
<!-- #                     # how many unique months where both FA and var available? -->
<!-- #                     return( d $month[!is.na(d$FA2) & !is.na(d[,x])] %>%  -->
<!-- #                              unique() %>% -->
<!-- #                              length() -->
<!-- #                             ) -->
<!-- #                   }) -->
<!-- #  -->
<!-- # # vars with greater than one unique month observed  -->
<!-- # FA2_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%  -->
<!-- #   names() -->

<!-- # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0 -->

<!-- # # subset d with vars of interest  -->
<!-- # model_df <- d %>%  -->
<!-- #   select(month, patID_eye, patID, SEXF, FA2, all_of(FA2_greater_than_one_month)) -->
<!-- #  -->
<!-- # # scale vars of interest -->
<!-- # model_df[, FA2_greater_than_one_month] <- scale(model_df[, FA2_greater_than_one_month]) -->

<!-- # subset d with vars of interest  -->
<!-- response <- 'FA2' -->

<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF, all_of(c(response,phenotypes))) -->

<!-- # scale vars of interest -->
<!-- model_df[, phenotypes[-which(phenotypes == response)]] <- scale( -->
<!--   model_df[, phenotypes[-which(phenotypes == response)]]) -->

<!-- ``` -->

<!-- #### model FA2, correcting for SEXF, month, and random effect eye -->
<!-- ```{r} -->

<!-- # exclude the response variable, FA2 -->
<!-- input_phenotypes <- names(model_df)[-which(names(model_df) %in% c("FA2", "FA2_rate", "month","patID_eye","patID","SEXF" ))] -->

<!-- # for each variable in input_phenotypes, model FA1  -->
<!-- formulas_char <- paste0(response," ~ ", input_phenotypes, " + SEXF + (1|patID_eye) + (1 | patID)")  -->

<!-- # model parrallel -->
<!-- results_list <- mclapply(formulas_char, function(form_char){ -->

<!--   formula <- as.formula( form_char) -->

<!--   fit <- try({ -->
<!--           lmer(formula, data = model_df) -->
<!--         }) -->
<!--   if(inherits(fit, "try-error")){ -->
<!--     warning(paste0("FAILURE!!!\n",form_char,"\n")) -->
<!--   } -->

<!--   return(summary(fit)) -->

<!-- }, mc.cores = detectCores()-1) -->

<!-- # simplify names to response and predictor (replace a space followed by a + with anything after with nothing) -->
<!-- names(results_list) <- sub("\\s\\+.*", "",formulas_char) -->

<!-- # filter out models that didnt work -->
<!-- faulty_models <- map(results_list,function(x) all(class(x) %in% c("summaryDefault","table")) ) %>%  -->
<!--   Filter(isTRUE,.) %>%  -->
<!--   names() -->

<!-- results_list <- results_list[!names(results_list)%in%faulty_models ] -->

<!-- # fill df with model values -->
<!-- results_df <- extract_results(results_list) -->

<!-- # filter correction terms from results -->
<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->

<!-- # filter out unwanted variables -->
<!-- results_df <- results_df %>%  -->
<!--   filter(!Model %in% c("FA_mean")) -->

<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Term %in% c(grep("MT|MT0|RW\\d|FA_m|FA\\d|SEX", results_df$Term, value =TRUE), "mean_parafoveal_rates" ) ) -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- FA2_by_phenos_volcano <- ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_2 ~ SEXF + phenotype_i + (1|patID_eye) + (1 | patID)") -->

<!-- ggsave(paste0( VA_modeling_path, "FA2_by_phenos_volcano.png"), FA2_by_phenos_volcano, width = 20, height = 12, dpi = 200) -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- FA2_by_phenos_volcano_filtered <- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_2 ~ SEXF + phenotype_i + (1|patID_eye) + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "FA2_by_phenos_volcano_filtered.png"), FA2_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200) -->

<!-- ``` -->



<!-- #$$ -->
<!-- # * Rate -->
<!-- ## Corrplot of betas for VA rate models #### -->
<!-- run all models for each factor rate  -->
<!-- ### Subset data for modeling #### -->
<!-- ```{r} -->

<!-- factor_rate_vars <- grep("FA.*_rate", names(d), value = TRUE) -->

<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF,  all_of(factor_rate_vars), all_of(phenotypes)) -->

<!-- # scale vars of interest, except responses: FA_mean, FA1:4 -->
<!-- model_df[, phenotypes[-which(phenotypes%in%factor_rate_vars)] ] <- scale(model_df[, phenotypes[-which(phenotypes%in%factor_rate_vars)]]) -->

<!-- # reduce model_df to patient's rate and first_visit phenotype value -->
<!-- ## for fa, grab first non na  -->
<!-- model_df_first <- model_df %>%  -->
<!--   arrange(patID_eye, month) %>%   -->
<!--   group_by(patID_eye) %>% -->
<!--   mutate(across( -->
<!--     !matches("FA"),  -->
<!--     ~dplyr::first(.), -->
<!--     .names  = "first_{.col}" -->
<!--     )) %>% -->
<!--   mutate(across( -->
<!--     matches("FA"),  -->
<!--     ~dplyr::first(.[!is.na(.)]), -->
<!--     .names  = "first_{.col}" -->
<!--     )) %>%  -->
<!--   ungroup() -->

<!-- # filter out non firsts and firsts that don't make sense, only unique rows -->
<!-- model_df_first_filtered_unique <- model_df_first %>%  -->
<!--   select( patID_eye, patID, SEXF, starts_with("first")) %>%    -->
<!--   select(!all_of(c("first_month", "first_patID", "first_SEXF"))) %>% -->
<!--   distinct() -->

<!-- ``` -->


<!-- ### Model #### -->
<!-- ```{r} -->

<!-- # for each rate , for each phenotype, model rate by that phenotype and add to main results_df -->
<!-- ## vars modeled -->
<!-- pred_vars <- paste("first", phenotypes, sep = "_") -->
<!-- response_vars <- paste("first", factor_rate_vars, sep = "_") -->

<!-- # loop through rates, except mean, and then model vars -->
<!-- for(response_var in response_vars){ -->


<!--   cat(paste("\n***\n", toupper(response_var), "\n***\n\n")) -->

<!--   form_strs <- paste0(response_var, " ~ ",pred_vars," + SEXF + (1|patID)" ) -->

<!--   # model -->
<!--   a_rates_summaries <- mclapply(form_strs, function (form_str){ -->
<!--       # make formula -->
<!--       a_formula <- form_str %>%  -->
<!--         as.formula() -->
<!--       # model and summarize -->
<!--       a_mod <- try({ -->
<!--         lmer(a_formula, data = model_df_first_filtered_unique) -->
<!--       }) -->
<!--       return(summary(a_mod)) -->
<!--    }, mc.cores = detectCores()-1) -->

<!--   names(a_rates_summaries) <- sub("\\s\\+.*", "", form_strs) -->

<!--   # where no error with model -->
<!--   functioning_models<-  map(a_rates_summaries, function(x) !all(class(x) %in% c("summaryDefault", "table"))) %>%  -->
<!--     unlist() -->
<!--   a_rates_summaries <- a_rates_summaries[functioning_models] -->

<!--   # make results df -->
<!--   a_results <- extract_results(a_rates_summaries) -->

<!--   # for first loop make empty results df -->
<!--   if(response_var == response_vars[[1]]) results_df <- a_results[0,] -->

<!--   results_df <- rbind(results_df, a_results) -->
<!-- } -->

<!-- ### Filter results and format data #### -->

<!-- # filter out intercepts and correcting terms -->
<!-- results_df <- results_df %>% -->
<!--   filter(!str_detect(Term, "Intercept|SEXF")) -->


<!-- # pivot data so rate vars are cols with estimates and models as rows -->
<!-- estimates_rates_d <- results_df %>%  -->
<!--   select(Response, Term, Estimate, q_value) -->

<!-- # widen data each rate col = estimates for each model -->
<!-- rate_estimates_d <- estimates_rates_d %>% -->
<!--   pivot_wider(names_from = Response, values_from = Estimate,  -->
<!--               id_cols = Term, values_fn = list(Estimate = dplyr::first)) -->

<!-- # rate_qvals_d <- estimates_rates_d %>% -->
<!-- #   pivot_wider(names_from = Response, values_from = q_value,  -->
<!-- #               id_cols = Model, values_fn = list(q_value = dplyr::last)) -->
<!-- ``` -->


<!-- ### Plot betas  -->
<!-- ```{r} -->

<!-- # plot raw values of betas -->
<!-- rate_estimates_matrix <- rate_estimates_d %>% -->
<!--   column_to_rownames("Term") %>% -->
<!--   as.matrix()  %>%  -->
<!--   t() -->

<!-- pdf(paste0(VA_modeling_path, "factor_rate_model_betas_heatmap.pdf"), width = 20, height = 10) -->
<!-- Heatmap(rate_estimates_matrix ,  -->
<!--         na_col = "black", -->
<!--         column_title = "Factor rate model betas",  -->
<!--         column_title_gp = gpar(fontsize = 20), -->
<!--         column_dend_height = unit(30, "mm"), -->
<!--         row_dend_width = unit(30, "mm"), -->
<!--         cluster_rows = hclust(dist(rate_estimates_matrix), method = "ward.D2"), -->
<!--         cluster_columns = hclust(dist(t(rate_estimates_matrix)), method = "ward.D2") -->
<!--         # cell_fun = function(j, i, x, y, width, height, fill) { -->
<!--         #               grid.text(text_annotations[i, j], x, y) -->
<!--         #       } -->
<!-- ) -->
<!-- dev.off() -->

<!-- # plot correlaation between betas -->
<!-- rate_estimates_cor <- cor(rate_estimates_d[, -1]) -->
<!-- pdf(paste0(VA_modeling_path, "factor_rate_betas_cor_heatmap.pdf"), width = 10, height = 8) -->
<!-- Heatmap(rate_estimates_cor,  -->
<!--         na_col = "black", -->
<!--         col =  plasma(5), -->
<!--         column_title = "Correlation of factor rate model betas",  -->
<!--         column_title_gp = gpar(fontsize = 20), -->
<!--         column_dend_height = unit(30, "mm"), -->
<!--         row_dend_width = unit(30, "mm"), -->
<!--         clustering_method_columns = "ward.D2" -->
<!--         ) -->
<!-- dev.off() -->


<!-- # par(mfrow = c( 2,1)) -->
<!-- # plot(type = "fan", as.phylo(hclust(dist(t(rate_estimates_matrix), ), method = "ward.D2"))) -->
<!-- # plot(type = "fan", as.phylo(hclust(dist(rate_estimates_matrix, ), method = "ward.D2"))) -->
<!-- # # plot(hclust(dist(rate_estimates_matrix), method = "ward.D2")) -->
<!-- # # plot(hclust(dist(t(rate_estimates_matrix), ), method = "ward.D2")) -->
<!-- # par(mfrow = c(1, 1)) -->




<!-- # ### plot q_values #### -->
<!-- # # plot correlations between q-vals -->
<!-- # rate_q_value_cor <- cor(rate_qvals_d[, -1]) -->
<!-- # pdf(paste0(wd_path, "results/VAA_analysis/factor_rate_qVals_cor_heatmap.pdf"), width = 10, height = 8) -->
<!-- # Heatmap(rate_q_value_cor,  -->
<!-- #         na_col = "black", -->
<!-- #         column_title = "Correlations between q-values in modeling for different factor rates",  -->
<!-- #         column_title_gp = gpar(fontsize = 20), -->
<!-- #         column_dend_height = unit(30, "mm"), -->
<!-- #         row_dend_width = unit(30, "mm") -->
<!-- #         ) -->
<!-- # dev.off() -->
<!-- #  -->
<!-- # # plot raw values of q_values -->
<!-- # pdf(paste0(wd_path, "results/VAA_analysis/factor_rate_model_qVals_heatmap.pdf"), width = 20, height = 10) -->
<!-- # Heatmap(rate_qvals_d %>% column_to_rownames("Model") %>%  mutate_all(~-log10(.)) %>% as.matrix() %>% t(), -->
<!-- #         na_col = "black", -->
<!-- #         column_title = "-log10 q-value of models for different factor rates", -->
<!-- #         column_title_gp = gpar(fontsize = 20), -->
<!-- #         column_dend_height = unit(30, "mm"), -->
<!-- #         row_dend_width = unit(30, "mm") -->
<!-- #     ) -->
<!-- # dev.off() -->


<!-- ``` -->


<!-- ## Mean Acuity rate by phenotypes -->

<!-- #### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->
<!-- # #  variables (rows of value>0) < 30 where FA_mean_rate present -->
<!-- # m <- d[!is.na(d$FA_mean_rate),phenotypes] != 0 -->
<!-- # vars <- m %>% -->
<!-- #   colSums(., na.rm = TRUE)  -->
<!-- #  -->
<!-- # # variables with > 30 obs of value>0  -->
<!-- # deviant_vars <- vars[vars > 30] %>% names() -->
<!-- #  -->
<!-- # # how many unique months do these have where FA_mean_rate and var available? -->
<!-- # count_unq_months <-  sapply(deviant_vars, function(x){  -->
<!-- #                     # how many unique months where both FA and var available? -->
<!-- #                     return( d $month[!is.na(d$FA_mean_rate) & !is.na(d[,x])] %>%  -->
<!-- #                              unique() %>% -->
<!-- #                              length() -->
<!-- #                             ) -->
<!-- #                   }) -->
<!-- # # vars with greater than one unique month observed  -->
<!-- # VA_mean_rate_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%  -->
<!-- #   names() -->
<!-- #  -->
<!-- # VA_mean_rate_greater_than_one_month <- c(VA_mean_rate_greater_than_one_month, "FA_mean_rate") -->

<!-- # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0 -->

<!-- # # subset d with vars of interest  -->
<!-- # model_df <- d %>%  -->
<!-- #   select(month, patID_eye, patID, SEXF, all_of(VA_mean_rate_greater_than_one_month)) -->
<!-- #  -->
<!-- # # scale vars of interest -->
<!-- # model_df[, VA_mean_rate_greater_than_one_month] <- scale(model_df[, VA_mean_rate_greater_than_one_month]) -->


<!-- response =  "FA_mean_rate" -->
<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF, all_of(c(response,phenotypes))) -->

<!-- # scale vars of interest, minus response: FA_mean_rate -->
<!-- model_df[, phenotypes[phenotypes!=response]] <- scale(model_df[, phenotypes[phenotypes!=response] ]) -->

<!-- # # are any first values NA? -->
<!-- # first_NA <- d %>% -->
<!-- #   filter(!is.na(eye_num)) %>%  -->
<!-- #   arrange(month) %>% -->
<!-- #   group_by(patID_eye) %>%  -->
<!-- #   select(month, patID_eye, all_of(VA_mean_greater_than_one_month)) %>% -->
<!-- #   filter(if_any(everything(), ~is.na(dplyr::first(.)) )) %>%  -->
<!-- #   ungroup() -->
<!-- #  -->
<!-- # first_NA %>% arrange(patID_eye) -->
<!-- # # ^ YES!  -->

<!-- ## reduce model_df to patient's rate and first_visit phenotype value. -->
<!-- # model_df_first <- model_df %>%  -->
<!-- #   arrange(month) %>%   -->
<!-- #   group_by(patID_eye) %>% -->
<!-- #   mutate(across( -->
<!-- #     everything(),  -->
<!-- #         ~dplyr::first(.), -->
<!-- #     .names  = "first_{.col}" -->
<!-- #     )) %>%  -->
<!-- #   ungroup() -->


<!-- # make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows -->
<!-- first_values_df <- model_df %>%  -->
<!--   select(-patID, - SEXF) %>%  -->
<!--   arrange(patID_eye, month) %>%   -->
<!--   group_by(patID_eye) %>% -->
<!--   # mutate(across( -->
<!--   #   everything(),  -->
<!--   #   ~ifelse(  -->
<!--   #       is.na(dplyr::first(.)), -->
<!--   #       dplyr::n(., n = 2),  -->
<!--   #       dplyr::first(.) -->
<!--   #   ), -->
<!--   #   .names  = "first_{.col}" -->
<!--   #   )) %>%  -->
<!--   mutate( -->
<!--     across( -->
<!--         -matches("FA"),  -->
<!--         ~dplyr::first(.[!is.na(.)]), -->
<!--         .names = "first_{.col}" -->
<!--   )) %>% -->
<!--   ungroup() %>% -->
<!--   select(-matches("month"), -matches("FA_mean")) %>%  -->
<!--   select(patID_eye,  starts_with("first_")) %>%  -->
<!--   unique() %>%  -->
<!--   arrange(patID_eye) -->

<!-- # filter out the firsts that don't make sense -->
<!-- model_df_first <- model_df %>%  -->
<!--   arrange(patID_eye) %>%  -->
<!--   select(patID_eye, patID, SEXF, FA_mean_rate) %>%  -->
<!--   na.omit() %>% -->
<!--   unique() %>%   -->
<!--   left_join(., first_values_df) %>%  -->
<!--   as.data.frame() -->

<!-- ``` -->



<!-- #### model each variable against FA_mean_rate, using zonal phenotype averages -->
<!-- ```{r} -->

<!-- # what phenos to sum -->
<!-- zonal_pheno_names <- grep("_.*0\\d$", names(model_df_first), value = TRUE) %>%  -->
<!--     gsub("0\\d",'', . ) %>%  -->
<!--     unique() -->


<!-- # for each pheno name, sum the columns in model_df_first -->
<!-- for (zonal_pheno in zonal_pheno_names){ -->
<!--   new_name <- paste0(zonal_pheno, "_sum") -->
<!--   model_df_first[[new_name]] <- model_df_first[, grep(zonal_pheno, names(model_df_first))] %>%  -->
<!--     rowSums(.) -->
<!-- } -->

<!-- # exclude the response variable, FA_mean_rate and sum variables -->
<!-- input_phenotypes <- names(model_df_first)[!names(model_df_first) %in% c("month", "patID_eye", "patID", "SEXF", "FA_mean_rate")] %>%  -->
<!--   grep("0_sum", ., value = TRUE, invert = TRUE) # ignore existent sum variables -->

<!-- # formulas -->
<!-- form_strs <- paste0(response," ~ ",input_phenotypes," + SEXF  + (1|patID)") -->

<!-- # for each variable in input_phenotypes, model FA_mean  -->
<!-- results_list <- mclapply(form_strs, function(form_str){ -->

<!--   ## linear model with random effects -->
<!--   re_formula <- as.formula(form_str) -->

<!--   fit <- try({lmer(formula = re_formula, data = model_df_first)}) -->

<!--   return(summary(fit)) -->

<!-- }, mc.cores = detectCores()-1) -->

<!-- names(results_list) <- sub("\\s\\+.*", "", form_strs) -->

<!-- # filter for models that worked -->
<!-- results_list <- results_list[unlist(lapply(results_list, function(x) !all(class(x) %in% c("summaryDefault", "table"))))] -->

<!-- ## make results df #### -->
<!-- results_df <- extract_results(results_list) -->

<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->

<!-- VA_mean_rate_results_df <- results_df -->

<!-- # filter for models that don't make sense -->
<!-- results_df_filtered <- results_df %>%  -->
<!--   # filter(!Model   %in% grep("FA|SEX", results_df$Model, value = TRUE)) %>%  -->
<!--   filter(!Term   %in% grep("avg|MT0|_.*0\\d$|RW\\d|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars -->

<!-- ### plot #### -->
<!-- VA_rate_by_phenos_volcano_summed_phenos <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_mean_rate ~ SEXF + phenotype_i + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "VA_rate_by_phenos_volcano_summed_phenos.png"), VA_rate_by_phenos_volcano_summed_phenos, width = 20, height = 12, dpi = 200) -->

<!-- VA_rate_by_phenos_volcano_summed_phenos_filtered <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_mean_rate ~ SEXF + phenotype_i + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "VA_rate_by_phenos_volcano_summed_phenos_filtered.png"), VA_rate_by_phenos_volcano_summed_phenos_filtered, width = 20, height = 12, dpi = 200) -->

<!-- VA_rate_by_phenos_volcano_summed_phenos_filtered_nominalP <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( `Pr(>|t|)` < .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_mean_rate ~ SEXF + phenotype_i + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "VA_rate_by_phenos_volcano_summed_phenos_filtered_nominalP.png"), VA_rate_by_phenos_volcano_summed_phenos_filtered_nominalP, width = 20, height = 12, dpi = 200) -->

<!-- ``` -->

<!-- <!-- ```{r} --> -->


<!-- <!-- km <- kmeans(results_df_filtered[, c("q_value", "Estimate"), ], centers =  6) --> -->
<!-- <!-- # db <- dbscan::dbscan(results_df_filtered[, c("q_value", "Estimate"), ], eps = .01,minPts = 2) --> -->

<!-- <!-- clusters <- km$cluster --> -->
<!-- <!-- # clusters <- db$cluster+1 --> -->
<!-- <!-- table(clusters) --> -->

<!-- <!-- cols <- palKelly()[seq(n_distinct(clusters))][clusters] --> -->

<!-- <!-- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`)))+ --> -->
<!-- <!--   geom_point(show.legend = TRUE, color = cols) --> -->

<!-- <!-- clusters[clusters==4] --> -->
<!-- <!-- ``` --> -->



<!-- ## FA1 Rate by phenotypes -->

<!-- #### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->

<!-- # #  variables (rows of value>0) < 30 where FA1_rate present -->
<!-- # m <- d[!is.na(d$FA1_rate),phenotypes] != 0 -->
<!-- # vars <- m %>% -->
<!-- #   colSums(., na.rm = TRUE)  -->
<!-- #  -->
<!-- # # variables with > 30 obs of value>0  -->
<!-- # deviant_vars <- vars[vars > 30] %>% names() -->
<!-- #  -->
<!-- # # how many unique months do these have where FA1_rate and var available? -->
<!-- # count_unq_months <-  sapply(deviant_vars, function(x){  -->
<!-- #                     # how many unique months where both mr and var available? -->
<!-- #                     return( d $month[!is.na(d$FA1_rate) & !is.na(d[,x])] %>%  -->
<!-- #                              unique() %>% -->
<!-- #                              length() -->
<!-- #                             ) -->
<!-- #                   }) -->
<!-- # # vars with greater than one unique month observed  -->
<!-- # FA1_rate_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%  -->
<!-- #   names() -->
<!-- #  -->
<!-- # FA1_rate_greater_than_one_month <- c(FA1_rate_greater_than_one_month, "FA1_rate") -->

<!-- # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0 -->

<!-- # # subset d with vars of interest  -->
<!-- # model_df <- d %>%  -->
<!-- #   select(month, patID_eye, patID, SEXF, all_of(FA1_rate_greater_than_one_month)) -->
<!-- #  -->
<!-- # # scale vars of interest -->
<!-- # model_df[, FA1_rate_greater_than_one_month] <- scale(model_df[, FA1_rate_greater_than_one_month]) -->

<!-- response = 'FA1_rate' -->
<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, response))) -->

<!-- # scale vars of interest, minus response: FA1_rate -->
<!-- model_df[, phenotypes[phenotypes!=response] ] <- scale(model_df[, phenotypes[phenotypes!=response] ]) -->


<!-- # # reduce model_df to patient's rate and first_visit phenotype value -->
<!-- # model_df_first <- model_df %>%  -->
<!-- #   arrange(month) %>%   -->
<!-- #   group_by(patID_eye) %>% -->
<!-- #   mutate(across( -->
<!-- #     everything(),  -->
<!-- #     ~dplyr::first(.), -->
<!-- #     .names  = "first_{.col}" -->
<!-- #     )) %>%  -->
<!-- #   ungroup() -->
<!-- #  -->
<!-- #  -->
<!-- #  -->
<!-- # model_df_first_filtered <- model_df_first %>%  -->
<!-- #   select(patID_eye, patID, SEXF, FA1_rate, starts_with("first")) %>%    -->
<!-- #   select(!matches("first.*\\d_rate")) %>%  -->
<!-- #   select(!c("first_SEXF", "first_month", "first_patID")) %>%  -->
<!-- #   unique() -->


<!-- # make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows -->
<!-- first_values_df <- model_df %>%  -->
<!--   select(-patID, - SEXF) %>%  -->
<!--   arrange(patID_eye, month) %>%   -->
<!--   group_by(patID_eye) %>% -->
<!--   mutate( -->
<!--     across( -->
<!--         everything(),  -->
<!--         ~dplyr::first(.[!is.na(.)]), -->
<!--         .names = "first_{.col}" -->
<!--   )) %>% -->
<!--   ungroup() %>% -->
<!--   select(-matches("month"), -matches("FA1_rate")) %>%  -->
<!--   select(patID_eye,  starts_with("first_")) %>%  -->
<!--   unique() %>%  -->
<!--   arrange(patID_eye) -->

<!-- # filter out the firsts that don't make sense -->
<!-- model_df_first <- model_df %>%  -->
<!--   arrange(patID_eye) %>%  -->
<!--   select(patID_eye, patID, SEXF, FA1_rate) %>%  -->
<!--   na.omit() %>% -->
<!--   unique() %>%   -->
<!--   left_join(., first_values_df) -->




<!-- ``` -->

<!-- ###### rate plots height and width -->
<!-- ```{r} -->

<!-- h_w <- c(10,15) -->

<!-- ``` -->

<!-- #### model each variable against FA1_rate, using zonal phenotype averages -->
<!-- ```{r} -->

<!-- # what phenos to sum -->
<!-- zonal_pheno_names <- grep("_.*0\\d$", names(model_df_first), value = TRUE) %>%  -->
<!--     gsub("0\\d",'', . ) %>%  -->
<!--     unique() -->


<!-- # for each pheno name, sum the columns in model_df_first -->
<!-- for (zonal_pheno in zonal_pheno_names){ -->
<!--   new_name <- paste0(zonal_pheno, "_sum") -->
<!--   model_df_first[[new_name]] <- model_df_first[, grep(zonal_pheno, names(model_df_first))] %>%  -->
<!--     rowSums(.) -->
<!-- } -->

<!-- # exclude the response variable, FA1_rate and sum variables -->
<!-- input_phenotypes <- names(model_df_first)[!names(model_df_first) %in% c("month", "patID_eye", "patID", "SEXF", "FA1_rate")] %>%  -->
<!--   grep("0_sum", ., value = TRUE, invert = TRUE) # ignore existent sum variables -->

<!-- # formulas -->
<!-- form_strs <- paste0(response," ~ ",input_phenotypes," + SEXF  + (1|patID)") -->

<!-- # for each variable in input_phenotypes, model FA_mean  -->
<!-- results_list <- mclapply(form_strs, function(form_str){ -->

<!--   ## linear model with random effects -->
<!--   re_formula <- as.formula(form_str) -->

<!--   fit <- try({lmer(formula = re_formula, data = model_df_first)}) -->

<!--   return(summary(fit)) -->

<!-- }, mc.cores = detectCores()-1) -->

<!-- names(results_list) <- sub("\\s\\+.*", "", form_strs) -->

<!-- # filter for models that worked -->
<!-- results_list <- results_list[unlist(lapply(results_list, function(x) !all(class(x) %in% c("summaryDefault", "table"))))] -->

<!-- ## make results df #### -->
<!-- results_df <- extract_results(results_list) -->

<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->

<!-- # filter for models that don't make sense -->
<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Model   %in% grep("FA|SEX", results_df$Model, value = TRUE)) %>%  -->
<!--   filter(!Model   %in% grep("avg|MT0|_.*0\\d$|SEX", results_df$Model, value = TRUE)) # exclude MT vars and zonal vars -->

<!-- FA1_rate_results_df <- results_df -->

<!-- ### volcano plot of vaa~phenotypes result #### -->
<!-- FA1_rate_by_phenos_volcano_summed_phenos <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_1_rate ~ SEXF + phenotype_i + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "FA1_rate_by_phenos_volcano_summed_phenos.png"), FA1_rate_by_phenos_volcano_summed_phenos, width = 20, height = 12, dpi = 200) -->

<!-- FA1_rate_by_phenos_volcano_summed_phenos_filtered <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_1_rate ~ SEXF + phenotype_i + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "FA1_rate_by_phenos_volcano_summed_phenos_filtered.png"), FA1_rate_by_phenos_volcano_summed_phenos_filtered, width = 20, height = 12, dpi = 200) -->

<!-- FA1_rate_by_phenos_volcano_summed_phenos_filtered_nominalP <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( `Pr(>|t|)` < .05,  Term, NA)), -->
<!--                        size = 5, -->
<!--                        # nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_1_rate ~ SEXF + phenotype_i + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "FA1_rate_by_phenos_volcano_summed_phenos_filtered_nominalP.png"), FA1_rate_by_phenos_volcano_summed_phenos_filtered_nominalP,  width = h_w[2], height = h_w[1], dpi = 200) -->

<!-- # add results to list -->
<!-- poster_results_dfs <- list() -->
<!-- FA1_rate_results_df_filtered <- results_df_filtered -->
<!-- poster_results_dfs[["FA1_rate_results_df_filtered"]] <- FA1_rate_results_df_filtered -->



<!-- ``` -->

<!-- ## FA2 Rate by phenotypes -->

<!-- #### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->

<!-- # #  variables (rows of value>0) < 30 where FA2_rate present -->
<!-- # m <- d[!is.na(d$FA2_rate),phenotypes] != 0 -->
<!-- # vars <- m %>% -->
<!-- #   colSums(., na.rm = TRUE)  -->
<!-- #  -->
<!-- # # variables with > 30 obs of value>0  -->
<!-- # deviant_vars <- vars[vars > 30] %>% names() -->
<!-- #  -->
<!-- # # how many unique months do these have where FA2_rate and var available? -->
<!-- # count_unq_months <-  sapply(deviant_vars, function(x){  -->
<!-- #                     # how many unique months where both mr and var available? -->
<!-- #                     return( d $month[!is.na(d$FA2_rate) & !is.na(d[,x])] %>%  -->
<!-- #                              unique() %>% -->
<!-- #                              length() -->
<!-- #                             ) -->
<!-- #                   }) -->
<!-- # # vars with greater than one unique month observed  -->
<!-- # FA2_rate_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%  -->
<!-- #   names() -->
<!-- #  -->
<!-- # FA2_rate_greater_than_one_month <- c(FA2_rate_greater_than_one_month, "FA2_rate") -->

<!-- # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0 -->

<!-- # # subset d with vars of interest  -->
<!-- # model_df <- d %>%  -->
<!-- #   select(month, patID_eye, patID, SEXF, all_of(FA2_rate_greater_than_one_month)) -->
<!-- #  -->
<!-- # # scale vars of interest -->
<!-- # model_df[, FA2_rate_greater_than_one_month] <- scale(model_df[, FA2_rate_greater_than_one_month]) -->

<!-- response ="FA2_rate" -->
<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, response))) -->

<!-- # scale vars of interest, minus response: FA2_rate -->
<!-- model_df[, phenotypes[phenotypes!=response] ] <- scale(model_df[, phenotypes[phenotypes!=response] ]) -->


<!-- # # reduce model_df to patient's rate and first_visit phenotype value -->
<!-- # model_df_first <- model_df %>%  -->
<!-- #   arrange(month) %>%   -->
<!-- #   group_by(patID_eye) %>% -->
<!-- #   mutate(across( -->
<!-- #     everything(),  -->
<!-- #     ~dplyr::first(.), -->
<!-- #     .names  = "first_{.col}" -->
<!-- #     )) %>%  -->
<!-- #   ungroup() -->
<!-- #  -->
<!-- #  -->
<!-- #  -->
<!-- # model_df_first_filtered <- model_df_first %>%  -->
<!-- #   select(patID_eye, patID, SEXF, FA2_rate, starts_with("first")) %>%    -->
<!-- #   select(!matches("first.*\\d_rate")) %>%  -->
<!-- #   select(!c("first_SEXF", "first_month", "first_patID")) %>%  -->
<!-- #   unique() -->


<!-- # make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows -->
<!-- first_values_df <- model_df %>%  -->
<!--   select(-patID, - SEXF) %>%  -->
<!--   arrange(patID_eye, month) %>%   -->
<!--   group_by(patID_eye) %>% -->
<!--   mutate( -->
<!--     across( -->
<!--         everything(),  -->
<!--         ~dplyr::first(.[!is.na(.)]), -->
<!--         .names = "first_{.col}" -->
<!--   )) %>% -->
<!--   ungroup() %>% -->
<!--   select(-matches("month"), -matches("FA2_rate")) %>%  -->
<!--   select(patID_eye,  starts_with("first_")) %>%  -->
<!--   unique() %>%  -->
<!--   arrange(patID_eye) -->

<!-- # filter out the firsts that don't make sense -->
<!-- model_df_first <- model_df %>%  -->
<!--   arrange(patID_eye) %>%  -->
<!--   select(patID_eye, patID, SEXF, FA2_rate) %>%  -->
<!--   na.omit() %>% -->
<!--   unique() %>%   -->
<!--   left_join(., first_values_df) -->

<!-- ``` -->


<!-- #### model each variable against FA2_rate, using zonal phenotype averages -->
<!-- ```{r} -->

<!-- # what phenos to sum -->
<!-- zonal_pheno_names <- grep("_.*0\\d$", names(model_df_first), value = TRUE) %>%  -->
<!--     gsub("0\\d",'', . ) %>%  -->
<!--     unique() -->


<!-- # for each pheno name, sum the columns in model_df_first -->
<!-- for (zonal_pheno in zonal_pheno_names){ -->
<!--   new_name <- paste0(zonal_pheno, "_sum") -->
<!--   model_df_first[[new_name]] <- model_df_first[, grep(zonal_pheno, names(model_df_first))] %>%  -->
<!--     rowSums(.) -->
<!-- } -->

<!-- # exclude the response variable, FA2_rate and sum variables -->
<!-- input_phenotypes <- names(model_df_first)[!names(model_df_first) %in% c("month", "patID_eye", "patID", "SEXF", "FA2_rate")] %>%  -->
<!--   grep("0_sum", ., value = TRUE, invert = TRUE) # ignore existent sum variables -->

<!-- # formulas -->
<!-- form_strs <- paste0(response," ~ ",input_phenotypes," + SEXF  + (1|patID)") -->

<!-- # for each variable in input_phenotypes, model FA_mean  -->
<!-- results_list <- mclapply(form_strs, function(form_str){ -->

<!--   ## linear model with random effects -->
<!--   re_formula <- as.formula(form_str) -->

<!--   fit <- try({lmer(formula = re_formula, data = model_df_first)}) -->

<!--   return(summary(fit)) -->

<!-- }, mc.cores = detectCores()-1) -->

<!-- names(results_list) <- sub("\\s\\+.*", "", form_strs) -->

<!-- # filter for models that worked -->
<!-- results_list <- results_list[unlist(lapply(results_list, function(x) !all(class(x) %in% c("summaryDefault", "table"))))] -->

<!-- ## make results df #### -->
<!-- results_df <- extract_results(results_list) -->

<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->

<!-- # filter for models that don't make sense -->
<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Model   %in% grep("FA|SEX", results_df$Model, value = TRUE)) %>%  -->
<!--   filter(!Model   %in% grep("avg|MT0|_.*0\\d$|SEX", results_df$Model, value = TRUE)) # exclude MT vars and zonal vars -->


<!-- FA2_rate_results_df <- results_df -->

<!-- ### volcano plot of vaa~phenotypes result #### -->
<!-- FA2_rate_by_phenos_volcano_summed_phenos <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_2_rate ~ SEXF + phenotype_i + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "FA2_rate_by_phenos_volcano_summed_phenos.png"), FA2_rate_by_phenos_volcano_summed_phenos, width = 20, height = 12, dpi = 200) -->

<!-- FA2_rate_by_phenos_volcano_summed_phenos_filtered <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_2_rate ~ SEXF + phenotype_i + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "FA2_rate_by_phenos_volcano_summed_phenos_filtered.png"), FA2_rate_by_phenos_volcano_summed_phenos_filtered, width = 20, height = 12, dpi = 200) -->

<!-- FA2_rate_by_phenos_volcano_summed_phenos_filtered_nominalP <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( `Pr(>|t|)` < .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        # nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("VA_factor_2_rate ~ SEXF + phenotype_i + (1 | patID)") -->

<!-- ggsave(paste0(VA_modeling_path, "FA2_rate_by_phenos_volcano_summed_phenos_filtered_nominalP.png"), FA2_rate_by_phenos_volcano_summed_phenos_filtered_nominalP, width = h_w[2], height = h_w[1], dpi = 200) -->


<!-- # add results to list -->
<!-- FA2_rate_results_df_filtered <- results_df_filtered -->
<!-- poster_results_dfs[["FA2_rate_results_df_filtered"]] <- FA2_rate_results_df_filtered -->

<!-- ``` -->


<!-- ##### corrplot of betas -->
<!-- ```{r} -->

<!-- FA_mean <- VA_mean_rate_results_df %>%  -->
<!--   select(Term, Estimate) %>% -->
<!--   filter(!grepl("SEX", Term)) %>%  -->
<!--   rename_with(~all_of(c("Term", "FA_mean_Estimate"))) -->

<!-- FA1 <- FA1_rate_results_df %>%  -->
<!--   filter(!grepl("SEX", Term)) %>%  -->
<!--   select(Term, Estimate) %>%  -->
<!--   rename_with(~all_of(c("Term", "FA1_Estimate"))) -->

<!-- estimates <- merge(FA1, FA_mean) %>%  -->
<!--   column_to_rownames("Term") %>%  -->
<!--   as.matrix() -->

<!-- Heatmap(estimates) -->
<!-- plot(estimates, type = "n");text(estimates,  -->
<!--                                  labels = rownames(estimates),  -->
<!--                                  # pos = 4,  -->
<!--                                  col = rainbow(length(estimates)) -->
<!--                                  );abline(c(0,1), lty = 3) -->


<!-- ``` -->


<!-- <!-- ##### find what's driving crystal_dep significance in the FA_mean_rate model --> -->
<!-- <!-- ```{r} --> -->


<!-- <!-- # find first value of each patient for FA rates and CRYSTALDEP --> -->
<!-- <!-- CRYSTALDEP_model_df <- d %>%  --> -->
<!-- <!--   select(month, SEXF, patID, patID_eye, grep("FA.*_rate", names(d), value = TRUE), CRYSTALDEP) %>%  --> -->
<!-- <!--   arrange(month) %>%   --> -->
<!-- <!--   group_by(patID_eye) %>% --> -->
<!-- <!--   mutate(across( --> -->
<!-- <!--     all_of(c(grep("FA.*_rate", names(d), value = TRUE), "CRYSTALDEP") ),  --> -->
<!-- <!--     ~dplyr::first(.[!is.na(.)]), --> -->
<!-- <!--     .names  = "first_{.col}" --> -->
<!-- <!--     )) %>%  --> -->
<!-- <!--   ungroup() %>%  --> -->
<!-- <!--   select(!c(grep("FA.*_rate", names(d), value = TRUE), "CRYSTALDEP"))%>%  --> -->
<!-- <!--   select(-month) %>%  --> -->
<!-- <!--   unique() --> -->

<!-- <!-- CRYSTALDEP_results_list <- list() ; for (factor_name in paste0("first_", c("FA1_rate", "FA2_rate"))){ #, "FA3_rate", "FA4_rate", "FA_mean_rate"))  ) { --> -->

<!-- <!--   re_formula <- as.formula( paste0(factor_name, " ~ SEXF  + first_CRYSTALDEP + (1|patID)") ) --> -->
<!-- <!--   re_formula %>% print() --> -->


<!-- <!--   CRYSTALDEP_results_list[[factor_name]] <- summary(lmer(formula = re_formula, data = CRYSTALDEP_model_df)) --> -->
<!-- <!--   } --> -->

<!-- <!-- CRYSTALDEP_results <- data.frame(Model = character(),  Term = character(),Estimate = numeric(), --> -->
<!-- <!--                          Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE) --> -->


<!-- <!-- for (model_name in names(CRYSTALDEP_results_list)){ --> -->

<!-- <!--   model_coef <- coef(CRYSTALDEP_results_list[[model_name]]) --> -->
<!-- <!--   a_result_df <- data.frame(Model = model_name,  --> -->
<!-- <!--                          Term = rownames(model_coef),  --> -->
<!-- <!--                          Estimate = model_coef[, "Estimate"], --> -->
<!-- <!--                          Std.Error = model_coef[, "Std. Error"], --> -->
<!-- <!--                          t.value = model_coef[, "t value"], --> -->
<!-- <!--                         `Pr(>|t|)`= model_coef[, "Pr(>|t|)"]) --> -->

<!-- <!--   CRYSTALDEP_results <- rbind(CRYSTALDEP_results, a_result_df) --> -->
<!-- <!-- } --> -->

<!-- <!-- CRYSTALDEP_results <- CRYSTALDEP_results %>%  --> -->
<!-- <!--   filter(Term == "first_CRYSTALDEP") --> -->

<!-- <!-- CRYSTALDEP_results$q_value <=- p.adjust(CRYSTALDEP_results$Pr) --> -->

<!-- <!-- ggplot(CRYSTALDEP_results, aes(Estimate, -log10(q_value), color = Model))+ --> -->
<!-- <!--   geom_point()+ --> -->
<!-- <!--   geom_label_repel(aes(label =  Model), --> -->
<!-- <!--                        size = 3, --> -->
<!-- <!--                        show.legend = FALSE)+ --> -->
<!-- <!--   theme_bw()+ --> -->
<!-- <!--   ggtitle("VA_factor_i ~ SEXF + first_CRYSTALDEP + (1|patID)") --> -->

<!-- <!-- ``` --> -->

<!-- ## regress mean rate by factors within each cluster -->
<!-- ### vision of each cluster for each trajectory -->
<!-- ```{r, eval = FALSE} -->

<!-- d <- d %>%  -->
<!--   mutate( -->
<!--     tiny_letters = rowMeans(across(matches("RW1[234]")), na.rm = TRUE), -->
<!--     small_letters = rowMeans(across(matches("RW1[01]|RW0[89]")),na.rm = TRUE), -->
<!--     medium_letters = rowMeans(across(matches("RW0[4567]")),na.rm = TRUE), -->
<!--     large_letters = rowMeans(across(matches("RW0[123]")) ,na.rm = TRUE) -->
<!--          ) -->

<!-- # For row scores --------- -->
<!-- cluster_letter_means <- d %>%  -->
<!--   select(patID_eye, month, my_cluster, matches("letters")) %>%  -->
<!--   pivot_longer(cols = matches("letters"), names_to = "variable")  -->



<!-- # Define edge list -->
<!-- # edges <- -->
<!-- metadata(sce)$trajectory_data$pca_line_data %>%  -->
<!--   select(edge, PC1:PC2) %>%  -->
<!--   ggplot(aes(PC1,PC2))+ -->
<!--     geom_point()+ -->
<!--     geom_line(aes(group = edge, colour = edge))+ -->
<!--     geom_label(aes(label = edge), size = 3)+ -->
<!--     theme_void() -->

<!-- edges <- metadata(sce)$trajectory_data$pca_line_data %>%  -->
<!--   select(edge) %>%  -->
<!--   separate(edge, c("from", "to"), sep = "--") %>%  -->
<!--   distinct() %>%  -->
<!--   as.matrix()  -->
<!-- edges <- rbind(edges, edges[,c(2,1)]) -->

<!-- # Build adjacency list -->
<!-- build_adj_list <- function(edges) { -->
<!--   adj <- split(edges[,2], edges[,1]) -->
<!--   return(adj) -->
<!-- } -->

<!-- adj_list <- build_adj_list(edges) -->

<!-- # Find all paths from root node -->
<!-- find_paths <- function(adj_list, current, visited = character(), path = character()) { -->
<!--   visited <- c(visited, current) -->
<!--   path <- c(path, current) -->

<!--   neighbors <- adj_list[[current]] -->
<!--   if (is.null(neighbors)) { -->
<!--     return(list(path)) -->
<!--   } -->

<!--   paths <- list() -->
<!--   for (neighbor in neighbors) { -->
<!--     if (!(neighbor %in% visited)) { -->
<!--       new_paths <- find_paths(adj_list, neighbor, visited, path) -->
<!--       paths <- c(paths, new_paths) -->
<!--     } -->
<!--   } -->

<!--   if (length(paths) == 0) { -->
<!--     return(list(path))  # If no children, return current path -->
<!--   } else { -->
<!--     return(paths) -->
<!--   } -->
<!-- } -->


<!-- # Run from root -->
<!-- paths <- find_paths(adj_list, 4) -->

<!-- # Print nicely -->
<!-- lapply(paths, function(p) cat(paste(p, collapse = ", "), "\n")) -->

<!-- paths <- list( -->
<!--   one_4  =  c(4, 3, 7, 5, 1, 6 ),  -->
<!--   two_4 =   c(4, 3, 7, 5, 1, 10 ),  -->
<!--   three_4 = c(4, 3, 7, 5, 1, 13 ),  -->
<!--   four_4 = c(4, 3, 7, 5, 1, 14 ),  -->
<!--   five_4 =  c(4, 3, 7, 5, 11, 2, 8 ),  -->
<!--   six_4 =  c(4, 3, 7, 5, 11, 2, 9 ),  -->
<!--   seven_4 = c(4, 3, 7, 5, 15 ) -->
<!-- ) -->


<!-- # for each path plot every each letter size -->
<!-- cluster_letters_plots <- list(); for(path in paths){ -->

<!--   plotting_d <- cluster_letter_means %>%  -->
<!--     filter(my_cluster %in% path) %>%  -->
<!--     mutate(my_cluster = factor(my_cluster, levels = path)) -->

<!--   cluster_letter_plot <- ggplot(plotting_d, aes(x = my_cluster, y = value))+ -->
<!--     # geom_violin(aes( color = my_cluster)) + -->
<!--     geom_boxplot(aes( color = my_cluster), show.legend = FALSE) + -->
<!--     geom_point(aes( color = my_cluster), alpha = .3, show.legend = FALSE)+ -->
<!--     # geom_jitter(aes( color = my_cluster), alpha = .3)+ -->
<!--     geom_smooth(method = "lm", aes(color = NULL, ,group = variable ),show.legend = FALSE)+ -->
<!--     # geom_smooth(method = "loess", aes(color = NULL, ,group = variable ),show.legend = FALSE)+ -->
<!--     facet_wrap(~variable)+ -->
<!--     ggtitle(paste0(paste(path, collapse = "-"))) -->

<!--   cluster_letters_plots[[paste(path, collapse = "-")]] <- cluster_letter_plot -->


<!-- } -->

<!-- cluster_letters_plots_grid <- patchwork::wrap_plots(cluster_letters_plots, ncol = 4) -->

<!-- ggsave(paste0(VA_modeling_path, "letter_means_by_cluster_path_grid.png"),  -->
<!--        cluster_letters_plots_grid, -->
<!--        height = 10, width = 40 -->
<!--        ) -->

<!-- # For ACUIT --------- -->
<!-- cluster_ACUIT <- d %>%  -->
<!--   select(patID_eye, month, my_cluster, ACUIT) %>%  -->
<!--   pivot_longer(cols = ACUIT, names_to = "variable")  -->



<!-- # path = paths[["four"]] -->

<!-- # for each path plot every each letter size -->
<!-- cluster_ACUIT_plots <- list(); for(path in paths){ -->

<!--   plotting_d <- cluster_ACUIT %>%  -->
<!--     filter(my_cluster %in% path) %>%  -->
<!--     mutate(my_cluster = factor(my_cluster, levels = path)) -->

<!--   cluster_ACUIT_plot <- ggplot(plotting_d, aes(x = my_cluster, y = value))+ -->
<!--     # geom_violin(aes( color = my_cluster)) + -->
<!--     geom_boxplot(aes( color = my_cluster), show.legend = FALSE) + -->
<!--     geom_point(aes( color = my_cluster), alpha = .3, show.legend = FALSE)+ -->
<!--     # geom_jitter(aes( color = my_cluster), alpha = .3)+ -->
<!--     geom_smooth(method = "lm", aes(color = NULL, ,group = variable ),show.legend = FALSE)+ -->
<!--     # geom_smooth(method = "loess", aes(color = NULL, ,group = variable ),show.legend = FALSE)+ -->
<!--     facet_wrap(~variable)+ -->
<!--     ggtitle(paste0(paste(path, collapse = "-"))) -->

<!--   cluster_ACUIT_plots[[paste(path, collapse = "-")]] <- cluster_ACUIT_plot -->


<!-- } -->

<!-- ggsave(paste0(VA_modeling_path, "ACUIT_by_cluster_path.png"),  -->
<!--        marrangeGrob(cluster_ACUIT_plots %>% sapply(., ggplotGrob), ncol =4, nrow =2 ), -->
<!--        height = 10, width = 40 -->
<!--        ) -->

<!-- ``` -->

## see how rate correlates to factors when subsetting for each trajectory a -> b, b->c, etc.

<!-- # $$ -->
<!-- # * Rate for each cluster -->

<!-- ```{r} -->




<!-- ``` -->


# ____   ___   ___   _____
# modeling vision: ACUIT

# value
## Thickness

### shape data
```{r}

response = "ACUIT"
# subset d with vars of interest
model_df <- d %>%
  select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, response)))

# scale vars of interest, minus response: FA2_rate
model_df[, phenotypes[phenotypes!=response] ] <- scale(model_df[, phenotypes[phenotypes!=response] ])


```

### model each variable against ACUIT rate, correcting for SEXF, and random effect of patient dependence
```{r}

# exclude the response variable, ACUIT_rate
input_phenotypes <- setdiff( names(model_df), c("month","patID_eye", "patID", "SEXF",response))
  
# formulas
form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID) +  (1|patID_eye)")

# for each variable in input_phenotypes, model FA_mean 
results_list <- mclapply(form_strs, function(form_str){
  ## linear model with random effects
  formula <- as.formula(form_str)

  fit <- try({lmer(formula = formula, data = model_df)})
  
  if(inherits(fit, "try-error")){
    return(NA)
  }else{
    return(summary(fit))
  }
}, mc.cores = detectCores()-1)

names(results_list) <- str_extract(form_strs, "\\S+\\s*~\\s*\\S+")

names(results_list)[is.na(results_list)]
# "OTHER_DISEASE" "SMKCIGSDAY"    "t2d"    

# filter for models that worked
results_list <- results_list[!is.na(results_list)]

## make results df ####
results_df <- extract_results(results_list)

results_df <- results_df %>% 
  filter(!grepl("(Intercept)|SEXF|month", Term))

# filter for models that don't make sense
results_df_filtered <- results_df %>% 
  filter(!Term   %in% grep("RW|SNELL|LOGMA|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars


# volcano plot of vaa~phenotypes result
ACUIT_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("ACUIT ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

ggsave(paste0(VA_modeling_path, "ACUIT_by_phenos_volcano.png"), ACUIT_by_phenos_volcano, width = 20, height = 12, dpi = 200)

# volcano plot of vaa~phenotypes result
ACUIT_by_phenos_volcano_filtered<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("ACUIT ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

ggsave(paste0(VA_modeling_path, "ACUIT_by_phenos_volcano_filtered.png"), ACUIT_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200)

ACUIT_by_phenos_volcano_filtered_nominalP<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( `Pr(>|t|)` < .05,  Term, NA)),
                       size = 4,
                       # nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("ACUIT ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

h_w <- c(10,15) 
ggsave(paste0(VA_modeling_path, "ACUIT_by_phenos_volcano_filtered_nominalP.png"), ACUIT_by_phenos_volcano_filtered_nominalP,  width = h_w[2], height = h_w[1], dpi = 200)


# # add results to list
# ACUIT_results_df_filtered <- results_df_MT_filtered
# poster_results_dfs[["ACUIT_results_df_filtered"]] <- ACUIT_results_df_filtered



```


## Rate
### Global Rate
#### subset data to scale and model
* filter out the ones with (rows>0) < 30
```{r}


response = "ACUIT_rate"
# subset d with vars of interest 
model_df <- d %>% 
  select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, response)))

# scale vars of interest, minus response: FA2_rate
model_df[, phenotypes[phenotypes!=response] ] <- scale(model_df[, phenotypes[phenotypes!=response] ])


# make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows
first_values_df <- model_df %>% 
  select(-patID, - SEXF) %>% 
  arrange(patID_eye, month) %>%  
  group_by(patID_eye) %>%
  mutate(across(everything(), 
        ~dplyr::first(.[!is.na(.)]),
        .names = "first_{.col}"
  )) %>%
  ungroup() %>%
  select(-matches("month"), -matches("ACUIT_rate")) %>% 
  select(patID_eye,  starts_with("first_")) %>% 
  unique() %>% 
  arrange(patID_eye)

# filter out the firsts that don't make sense
model_df_first <- model_df %>% 
  arrange(patID_eye) %>% 
  select(patID_eye, patID, SEXF, ACUIT_rate) %>% 
  na.omit() %>%
  unique() %>%  
  left_join(., first_values_df)



```

#### model each variable against ACUIT rate, correcting for SEXF, and random effect of patient dependence
> modeling rate by  patient's first phenotype does not have time effect, and therefore no repeated eye measures; but there are still repeated patients because each has two eyes

```{r}

# exclude the response variable, ACUIT_rate
input_phenotypes <- setdiff( names(model_df_first), c("month","patID_eye", "patID", "SEXF",response))
  
# formulas
form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID)")

# for each variable in input_phenotypes, model FA_mean 
results_list <- mclapply(form_strs, function(form_str){
  ## linear model with random effects
  formula <- as.formula(form_str)

  fit <- try({lmer(formula = formula, data = model_df_first)})
  
  if(inherits(fit, "try-error")){
    return(NA)
  }else{
    return(summary(fit))
  }
}, mc.cores = detectCores()-1)

names(results_list) <- str_extract(form_strs, "\\S+\\s*~\\s*\\S+")

names(results_list)[is.na(results_list)]
# "ACUIT_rate ~ first_OTHER_DISEASE" "ACUIT_rate ~ first_SMKCIGSDAY" "ACUIT_rate ~ first_t2d"

# filter for models that worked
results_list <- results_list[!is.na(results_list)]

## make results df ####
results_df <- extract_results(results_list)

ACUIT_global_rate_results_df <- results_df

results_df <- results_df %>% 
  filter(!grepl("(Intercept)|SEXF|month", Term))

# filter for models that don't make sense
results_df_filtered <- results_df %>% 
  filter(!Term   %in% grep("RW|SNELL|LOGMA|ACUIT|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars


# volcano plot of vaa~phenotypes result
ACUIT_rate_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("ACUIT_rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(VA_modeling_path, "ACUIT_rate_by_phenos_volcano.png"), ACUIT_rate_by_phenos_volcano, width = 20, height = 12, dpi = 200)


# volcano plot of vaa~phenotypes result
ACUIT_rate_by_phenos_volcano_filtered<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("ACUIT_rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(VA_modeling_path, "ACUIT_rate_by_phenos_volcano_filtered.png"), ACUIT_rate_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200)

ACUIT_rate_by_phenos_volcano_filtered_nominalP<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( `Pr(>|t|)` < .05,  Term, NA)),
                       size = 4,
                       # nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("ACUIT_rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(VA_modeling_path, "ACUIT_rate_by_phenos_volcano_filtered_nominalP.png"), ACUIT_rate_by_phenos_volcano_filtered_nominalP,  width = h_w[2], height = h_w[1], dpi = 200)


# # add results to list
# ACUIT_rate_results_df_filtered <- results_df_MT_filtered
# poster_results_dfs[["ACUIT_rate_results_df_filtered"]] <- ACUIT_rate_results_df_filtered



```

### Stepwise rate

```{r stepwise var rate function}
# ****** FUNCTION: calculate combination of each x with following x for each eye ******
stepwise_rate <- function(a_eye, df = NULL, var = NA, time_var = "year") {
    
  if(is.na(var)) stop("Response NA. Please provide variable for calculate steppwise rate")
  
  
  # Filter for eye of interest
  a_eye_diff_d <- df[df$patID_eye == a_eye, ]
  
  if(time_var == "year"){
    # make year var
    if(!"year" %in% names(a_eye_diff_d)){
      a_eye_diff_d <- a_eye_diff_d %>%
        mutate(year = round(month/12, 2))
    }
    
    # Create year to following year combinations
    year_combos <- data.frame(year = a_eye_diff_d$year, post_year = lead(a_eye_diff_d$year)) 
    
    # Create var to every other var combinations
    var_combos <- data.frame(var = a_eye_diff_d[[var]], post_var = lead(a_eye_diff_d[[var]])) 
      
    # Time pp combinations
    year_var_combos <- cbind(year_combos, var_combos)
    
    # make rate for each i to i+1: change in var over change in time>> postx-x/postyear-year
    year_var_combos$var_rate <- with(year_var_combos , (post_var-var)/(post_year-year))
    
    # Join back to dataset
    a_eye_diff_d <- suppressMessages(
      a_eye_diff_d %>% 
        left_join(year_var_combos, by = c("year")) 
      )
    
    # remove functional vars
    a_eye_diff_d$var <- NULL
    a_eye_diff_d$post_year <- NULL
    a_eye_diff_d$post_var <- NULL
    
    
    # rename "var" to var
    names(a_eye_diff_d) <- str_replace_all(names(a_eye_diff_d), "var", var)
  } 
    
  if(time_var == "month"){
    
    # Create month to following month combinations
    month_combos <- data.frame(month = a_eye_diff_d$month, post_month = lead(a_eye_diff_d$month)) 
    
    # Create var to every other var combinations
    var_combos <- data.frame(var = a_eye_diff_d[[var]], post_var = lead(a_eye_diff_d[[var]])) 
      
    # Time pp combinations
    month_var_combos <- cbind(month_combos, var_combos)
    
    # make rate for each i to i+1: change in var over change in time>> postx-x/postmonth-month
    month_var_combos$var_rate <- with(month_var_combos , (post_var-var)/(post_month-month))
    
    # Join back to dataset
    a_eye_diff_d <- suppressMessages(
      a_eye_diff_d %>% 
        left_join(month_var_combos, by = c("month")) 
      )
    
    # remove functional vars
    a_eye_diff_d$var <- NULL
    a_eye_diff_d$post_month <- NULL
    a_eye_diff_d$post_var <- NULL
    
    
    # rename "var" to var
    names(a_eye_diff_d) <- str_replace_all(names(a_eye_diff_d), "var", var)
  }
  
  return(a_eye_diff_d)
}
```

#### data
* filter out the ones with (rows>0) < 30
```{r}

var <- "ACUIT"
# subset d with vars of interest 
model_df <- d %>% 
  select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, var)))

# scale vars of interest, minus response: FA2_rate
model_df[, phenotypes[phenotypes!=var] ] <- scale(model_df[, phenotypes[phenotypes!=var] ])


model_df_lag <- mclapply(unique(model_df$patID_eye), function(x) stepwise_rate(x, df = model_df,  var = var, time_var = "month"), mc.cores = detectCores()-1) %>% 
  do.call(rbind,.)

response = sprintf("%s_rate", var)

```

#### model 

```{r}

# exclude the response variable, ACUIT_rate
input_phenotypes <- setdiff( names(model_df_lag), c("month","patID_eye", "patID", "SEXF",response, var, "year"))
  
# formulas
form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID) + (1|patID_eye)")
# form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID) ")

# for each variable in input_phenotypes, model FA_mean 
results_list <- mclapply(form_strs, function(form_str){
  ## linear model with random effects
  formula <- as.formula(form_str)

  fit <- try({lmer(formula = formula, data = model_df_lag)})
  
  if(inherits(fit, "try-error")){
    return(NA)
  }else{
    return(summary(fit))
  }
}, mc.cores = detectCores()-1)

names(results_list) <- str_extract(form_strs, "\\S+\\s*~\\s*\\S+")

names(results_list)[is.na(results_list)]
# "ACUIT_rate ~ first_OTHER_DISEASE" "ACUIT_rate ~ first_SMKCIGSDAY" "ACUIT_rate ~ first_t2d"

# filter for models that worked
results_list <- results_list[!is.na(results_list)]

## make results df ####
results_df <- extract_results(results_list)

ACUIT_stepwise_rate_results_df <- results_df

results_df <- results_df %>% 
  filter(!grepl("(Intercept)|SEXF|month", Term))

# filter for models that don't make sense
results_df_filtered <- results_df %>% 
  filter(!Term   %in% grep("RW|SNEL|LOGM|SEX|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars


# volcano plot of vaa~phenotypes result
ACUIT_rate_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("stepwise_ACUIT_rate ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

ggsave(paste0(VA_modeling_path, "ACUIT_stepwise_lag_rate_by_phenos_volcano.png"), ACUIT_rate_by_phenos_volcano, width = 20, height = 12, dpi = 200)

# volcano plot of vaa~phenotypes result
ACUIT_rate_by_phenos_volcano_filtered<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("stepwise_ACUIT_rate ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

ggsave(paste0(VA_modeling_path, "ACUIT_stepwise_lag_rate_by_phenos_volcano_filtered.png"), ACUIT_rate_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200)

ACUIT_rate_by_phenos_volcano_filtered_nominalP<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( `Pr(>|t|)` < .05,  Term, NA)),
                       size = 4,
                       # nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("stepwise_ACUIT_rate ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

ggsave(paste0(VA_modeling_path, "ACUIT_stepwise_lag_rate_by_phenos_volcano_filtered_nominalP.png"), ACUIT_rate_by_phenos_volcano_filtered_nominalP,  width = h_w[2], height = h_w[1], dpi = 200)



```


### compare global to stepwise
```{r}

fun <- function(x){
x %>% 
  mutate(Term = str_remove_all(Term, "first_")) %>% 
  select(Term, `Pr(>|t|)`) %>% 
  mutate(neglogP = -log10(`Pr(>|t|)`)) %>% 
  select(-`Pr(>|t|)`) %>% 
  filter(!grepl("Intercept|SEX", Term))
}

inner_join(fun(ACUIT_global_rate_results_df), fun(ACUIT_stepwise_rate_results_df), by = "Term", suffix = c("_global", "_stepwise")) %>% 
  column_to_rownames("Term") %>% 
  mutate_all(~-log10(.)) %>% 
  plot()

abline(a = 0, b = 1)

```

# ____   ___   ___   _____

# modeling thickness 

<!-- ## MT5 thickness -->
<!-- ### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->

<!-- # #  variables (rows of value>0) < 30 where MT05 present -->
<!-- # m <- d[!is.na(d$MT05),phenotypes] != 0 -->
<!-- # vars <- m %>% -->
<!-- #   colSums(., na.rm = TRUE) -->
<!-- # -->
<!-- # # variables with > 30 obs of value>0 -->
<!-- # deviant_vars <- vars[vars > 30] %>% names() -->
<!-- # -->
<!-- # # how many unique months do these have where FA_mean and var available? -->
<!-- # count_unq_months <-  sapply(deviant_vars, function(x){ -->
<!-- #                     # how many unique months where both mr and var available? -->
<!-- #                     return( d $month[!is.na(d$MT05) & !is.na(d[,x])] %>% -->
<!-- #                              unique() %>% -->
<!-- #                              length() -->
<!-- #                             ) -->
<!-- #                   }) -->
<!-- # # vars with greater than one unique month observed -->
<!-- # MT5_greater_than_one_month <- count_unq_months[count_unq_months>1] %>% -->
<!-- #   names() -->

<!-- # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0 -->

<!-- response = "MT05" -->
<!-- # subset d with vars of interest -->
<!-- model_df <- d %>% -->
<!--   select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, response))) -->

<!-- # scale vars of interest, minus response: FA2_rate -->
<!-- model_df[, phenotypes[phenotypes!=response] ] <- scale(model_df[, phenotypes[phenotypes!=response] ]) -->


<!-- ``` -->

<!-- ### model each variable against MT5 correcting for SEXF, and random effect of patient dependence -->
<!-- > modeling rate by  patient's first phenotype does not have time effect, and therefore no repeated eye measures; but there are still repeated patients because each has two eyes -->

<!-- ```{r} -->

<!-- # exclude the response variable, mt5 -->
<!-- input_phenotypes <- names(model_df)[!names(model_df) %in% c("month","patID_eye", "patID", "SEXF", "MT05")] -->

<!-- # formulas -->
<!-- form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID) + (1|patID_eye)") -->

<!-- # for each variable in input_phenotypes, model FA_mean  -->
<!-- results_list <- mclapply(form_strs, function(form_str){ -->

<!--   ## linear model with random effects -->
<!--   re_formula <- as.formula(form_str) -->

<!--   fit <- try({lmer(formula = re_formula, data = model_df)}) -->

<!--   return(summary(fit)) -->

<!-- }, mc.cores = detectCores()-1) -->

<!-- names(results_list) <- sub("\\s\\+.*", "", form_strs) -->

<!-- # filter for models that worked -->
<!-- results_list <- results_list[unlist(lapply(results_list, function(x) !all(class(x) %in% c("summaryDefault", "table"))))] -->

<!-- ## make results df #### -->
<!-- results_df <- extract_results(results_list) -->

<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->

<!-- # filter for models that don't make sense -->
<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Term   %in% grep("avg|MT0|_.*0\\d$|MICRON|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- MT5_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("MT5 ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)") -->

<!-- ggsave(paste0(MT_modeling_path, "MT5_by_phenos_volcano.png"), MT5_by_phenos_volcano, width = 20, height = 12, dpi = 200) -->

<!-- MT5_by_phenos_volcano_filtered <-ggplot(results_df_filtered , aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("MT5 ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)") -->

<!-- ggsave(paste0(MT_modeling_path, "MT5_by_phenos_volcano_filtered.png"), MT5_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200) -->

<!-- ``` -->


<!-- ## MT5 rate -->
<!-- ### subset data to scale and model -->
<!-- * filter for the vars with (rows>0) > 30 -->
<!-- ```{r} -->

<!-- # #  variables (rows of value != 0) where MT05_rate present -->
<!-- # m <- d[!is.na(d$MT05_rate),phenotypes] != 0 -->
<!-- #  -->
<!-- # # how many of each var is not 0 -->
<!-- # vars <- m %>% -->
<!-- #   colSums(., na.rm = TRUE) -->
<!-- #  -->
<!-- # # variables with > 30 obs of value>0 -->
<!-- # powered_vars <- vars[vars > 30] %>% names() -->
<!-- #  -->
<!-- # # how many unique months do these have where FA_mean and var available? -->
<!-- # count_unq_months <-  sapply(powered_vars, function(x){ -->
<!-- #                      -->
<!-- #                     # how many unique months where both response variable and var available? -->
<!-- #                     return( d$month[!is.na(d$MT05_rate) & !is.na(d[,x])] %>% -->
<!-- #                              unique() %>% -->
<!-- #                              length() -->
<!-- #                             ) -->
<!-- #                   }) -->
<!-- #  -->
<!-- # # vars with greater than one unique month observed -->
<!-- # MT5_rate_greater_than_one_month <- count_unq_months[count_unq_months>1] %>% -->
<!-- #   names() -->



<!-- # # subset d with vars of interest -->
<!-- # model_df <- d %>% -->
<!-- #   select(month, patID_eye, patID, SEXF, all_of(MT5_rate_greater_than_one_month)) -->
<!-- #  -->
<!-- # # scale vars of interest -->
<!-- # model_df[, MT5_rate_greater_than_one_month] <- scale(model_df[, MT5_rate_greater_than_one_month]) -->


<!-- # subset d with vars of interest -->

<!-- response = "MT05_rate" -->
<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, response))) -->

<!-- # scale vars of interest, minus response: FA2_rate -->
<!-- model_df[, phenotypes[phenotypes!=response] ] <- scale(model_df[, phenotypes[phenotypes!=response] ]) -->


<!-- # # reduce model_df to patient's rate and first_visit phenotype value -->
<!-- # model_df_first <- model_df %>%  -->
<!-- #   arrange(month) %>%   -->
<!-- #   group_by(patID_eye) %>% -->
<!-- #   # summarize(across( -->
<!-- #   #   .cols = !c("patID_eye", "month", "SEXF"),  -->
<!-- #   #   .fns = ~dplyr::first(.), -->
<!-- #   #   .names  = "first_{.col}") -->
<!-- #   #   ) %>% -->
<!-- #   mutate(across( -->
<!-- #     everything(),  -->
<!-- #     ~dplyr::first(.), -->
<!-- #     .names  = "first_{.col}" -->
<!-- #     ))  -->
<!-- #  -->
<!-- # # select columns for modeling -->
<!-- # model_df_first_filtered <- model_df_first %>%  -->
<!-- #   select( patID_eye, patID, SEXF, MT05_rate, starts_with("first")) %>%    -->
<!-- #   select(!matches("0[12346789]_rate")) %>% # exclude MT rates -->
<!-- #   select(!c("first_SEXF", "first_month", "first_patID")) %>%  -->
<!-- #   unique() -->

<!-- # make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows -->
<!-- first_values_df <- model_df %>%  -->
<!--   select(-patID, - SEXF) %>%  -->
<!--   arrange(patID_eye, month) %>%   -->
<!--   group_by(patID_eye) %>% -->
<!--   mutate( -->
<!--     across( -->
<!--         everything(),  -->
<!--         ~dplyr::first(.[!is.na(.)]), -->
<!--         .names = "first_{.col}" -->
<!--   )) %>% -->
<!--   ungroup() %>% -->
<!--   select(-matches("month"), -matches("MT05_rate")) %>%  -->
<!--   select(patID_eye,  starts_with("first_")) %>%  -->
<!--   unique() %>%  -->
<!--   arrange(patID_eye) -->

<!-- # filter out the firsts that don't make sense -->
<!-- model_df_first <- model_df %>%  -->
<!--   arrange(patID_eye) %>%  -->
<!--   select(patID_eye, patID, SEXF, MT05_rate) %>%  -->
<!--   na.omit() %>% -->
<!--   unique() %>%   -->
<!--   left_join(., first_values_df) -->




<!-- ``` -->

<!-- ### model each variable against MT5 correcting for SEXF, and random effect of patient dependence -->
<!-- > modeling rate by  patient's first phenotype does not have time effect, and therefore no repeated eye measures; but there are still repeated patients because each has two eyes -->

<!-- ```{r} -->


<!-- # exclude the response variable, mt5 rate -->
<!-- input_phenotypes <- names(model_df)[!names(model_df) %in% c("month","patID_eye", "patID", "SEXF", "MT05_rate")] -->

<!-- # formulas -->
<!-- form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID)") -->

<!-- # for each variable in input_phenotypes, model FA_mean  -->
<!-- results_list <- mclapply(form_strs, function(form_str){ -->

<!--   ## linear model with random effects -->
<!--   re_formula <- as.formula(form_str) -->

<!--   fit <- try({lmer(formula = re_formula, data = model_df)}) -->

<!--   return(summary(fit)) -->

<!-- }, mc.cores = detectCores()-1) -->

<!-- names(results_list) <- sub("\\s\\+.*", "", form_strs) -->

<!-- # filter for models that worked -->
<!-- results_list <- results_list[unlist(lapply(results_list, function(x) !all(class(x) %in% c("summaryDefault", "table"))))] -->

<!-- ## make results df #### -->
<!-- results_df <- extract_results(results_list) -->

<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->

<!-- # filter for models that don't make sense -->
<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Term   %in% grep("avg|MT0|_.*0\\d$|MICRON|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- MT5_rate_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("MT5_rate ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)") -->

<!-- ggsave(paste0(MT_modeling_path, "MT5_rate_by_phenos_volcano.png"), MT5_rate_by_phenos_volcano, width = 20, height = 12, dpi = 200) -->

<!-- MT5_rate_by_phenos_volcano_filtered<-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("MT5_rate ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)") -->

<!-- ggsave(paste0(MT_modeling_path, "MT5_rate_by_phenos_volcano_filtered.png"), MT5_rate_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200) -->

<!-- MT5_rate_by_phenos_volcano_filtered_nominalP <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse(`Pr(>|t|)`< .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("MT5_rate ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)") -->

<!-- ggsave(paste0(MT_modeling_path, "MT5_rate_by_phenos_volcano_filtered_nominalP.png"), MT5_rate_by_phenos_volcano_filtered_nominalP , width = 20, height = 12, dpi = 200) -->





<!-- ``` -->



# Parafovea

## Thickness

### shape data
```{r}

response = "parafovea"
# subset d with vars of interest
model_df <- d %>%
  select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, response)))

# scale vars of interest, minus response: FA2_rate
model_df[, phenotypes[phenotypes!=response] ] <- scale(model_df[, phenotypes[phenotypes!=response] ])


```

### model each variable against parafovea rate, correcting for SEXF, and random effect of patient dependence
```{r}

# exclude the response variable, parafovea_rate
input_phenotypes <- setdiff( names(model_df), c("month","patID_eye", "patID", "SEXF",response))
  
# formulas
form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID) +  (1|patID_eye)")

# for each variable in input_phenotypes, model FA_mean 
results_list <- mclapply(form_strs, function(form_str){
  ## linear model with random effects
  formula <- as.formula(form_str)

  fit <- try({lmer(formula = formula, data = model_df)})
  
  if(inherits(fit, "try-error")){
    return(NA)
  }else{
    return(summary(fit))
  }
}, mc.cores = detectCores()-1)

names(results_list) <- str_extract(form_strs, "\\S+\\s*~\\s*\\S+")

names(results_list)[is.na(results_list)]
# "OTHER_DISEASE" "SMKCIGSDAY"    "t2d"    

# filter for models that worked
results_list <- results_list[!is.na(results_list)]

## make results df ####
results_df <- extract_results(results_list)

results_df <- results_df %>% 
  filter(!grepl("(Intercept)|SEXF|month", Term))

# filter for models that don't make sense
results_df_filtered <- results_df %>% 
  filter(!Term   %in% grep("avg|MT0|_.*0\\d$|MICRON|FOVTHICK|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars


# volcano plot of vaa~phenotypes result
parafovea_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("parafovea ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

ggsave(paste0(MT_modeling_path, "parafovea_by_phenos_volcano.png"), parafovea_by_phenos_volcano, width = 20, height = 12, dpi = 200)

# volcano plot of vaa~phenotypes result
parafovea_by_phenos_volcano_MT_filtered<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("parafovea ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

ggsave(paste0(MT_modeling_path, "parafovea_by_phenos_volcano_MT_filtered.png"), parafovea_by_phenos_volcano_MT_filtered, width = 20, height = 12, dpi = 200)

parafovea_by_phenos_volcano_MT_filtered_nominalP<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( `Pr(>|t|)` < .05,  Term, NA)),
                       size = 4,
                       # nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("parafovea ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")


ggsave(paste0(MT_modeling_path, "parafovea_by_phenos_volcano_MT_filtered_nominalP.png"), parafovea_by_phenos_volcano_MT_filtered_nominalP,  width = h_w[2], height = h_w[1], dpi = 200)


# # add results to list
# parafovea_results_df_filtered <- results_df_MT_filtered
# poster_results_dfs[["parafovea_results_df_filtered"]] <- parafovea_results_df_filtered



```


## Rate
### Global rate
#### subset data to scale and model
* filter out the ones with (rows>0) < 30
```{r}


response = "parafovea_rate"
# subset d with vars of interest 
model_df <- d %>% 
  select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, response)))

# scale vars of interest, minus response: FA2_rate
model_df[, phenotypes[phenotypes!=response] ] <- scale(model_df[, phenotypes[phenotypes!=response] ])


# make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows
first_values_df <- model_df %>% 
  select(-patID, - SEXF) %>% 
  arrange(patID_eye, month) %>%  
  group_by(patID_eye) %>%
  mutate(across(everything(), 
        ~dplyr::first(.[!is.na(.)]),
        .names = "first_{.col}"
  )) %>%
  ungroup() %>%
  select(-matches("month"), -matches("parafovea_rate")) %>% 
  select(patID_eye,  starts_with("first_")) %>% 
  unique() %>% 
  arrange(patID_eye)

# filter out the firsts that don't make sense
model_df_first <- model_df %>% 
  arrange(patID_eye) %>% 
  select(patID_eye, patID, SEXF, parafovea_rate) %>% 
  na.omit() %>%
  unique() %>%  
  left_join(., first_values_df)


```

#### model each variable against parafovea rate, correcting for SEXF, and random effect of patient dependence
> modeling rate by  patient's first phenotype does not have time effect, and therefore no repeated eye measures; but there are still repeated patients because each has two eyes

```{r}

# exclude the response variable, parafovea_rate
input_phenotypes <- setdiff( names(model_df_first), c("month","patID_eye", "patID", "SEXF",response))
  
# formulas
form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID)")

# for each variable in input_phenotypes, model FA_mean 
results_list <- mclapply(form_strs, function(form_str){
  ## linear model with random effects
  formula <- as.formula(form_str)

  fit <- try({lmer(formula = formula, data = model_df_first)})
  
  if(inherits(fit, "try-error")){
    return(NA)
  }else{
    return(summary(fit))
  }
}, mc.cores = detectCores()-1)

names(results_list) <- str_extract(form_strs, "\\S+\\s*~\\s*\\S+")

names(results_list)[is.na(results_list)]
# "parafovea_rate ~ first_OTHER_DISEASE" "parafovea_rate ~ first_SMKCIGSDAY" "parafovea_rate ~ first_t2d"

# filter for models that worked
results_list <- results_list[!is.na(results_list)]

## make results df ####
results_df <- extract_results(results_list)

parafovea_global_results_df <- results_df

results_df <- results_df %>% 
  filter(!grepl("(Intercept)|SEXF|month", Term))

# filter for models that don't make sense
results_df_filtered <- results_df %>% 
  filter(!Term   %in% grep("avg|MT0|_.*0\\d$|MICRON|FOVTHICK|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars


# volcano plot of vaa~phenotypes result
parafovea_rate_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("parafovea_rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(MT_modeling_path, "parafovea_rate_by_phenos_volcano.png"), parafovea_rate_by_phenos_volcano, width = 20, height = 12, dpi = 200)

# # volcano plot of vaa~phenotypes result
# parafovea_rate_by_phenos_volcano_otherZones_filtered<- ggplot(results_df_otherZones_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+
#       geom_point(show.legend = FALSE)+
#       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
#                        size = 4,
#                        nudge_y = 0.5,
#                        show.legend = FALSE,
#                        max.overlaps = 20)+
#       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
#       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
#       theme_bw()+
#       # facet_wrap(~quintile, scales = "fixed")+
#       ggtitle("parafovea_rate ~ SEXF + phenotype_i + (1| patID)")
# 
# ggsave(paste0(MT_modeling_path, "parafovea_rate_by_phenos_volcano_otherZones_filtered.png"), parafovea_rate_by_phenos_volcano_otherZones_filtered, width = 20, height = 12, dpi = 200)

# volcano plot of vaa~phenotypes result
parafovea_rate_by_phenos_volcano_MT_filtered<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("parafovea_rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(MT_modeling_path, "parafovea_rate_by_phenos_volcano_MT_filtered.png"), parafovea_rate_by_phenos_volcano_MT_filtered, width = 20, height = 12, dpi = 200)

parafovea_rate_by_phenos_volcano_MT_filtered_nominalP<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( `Pr(>|t|)` < .05,  Term, NA)),
                       size = 4,
                       # nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("parafovea_rate ~ SEXF + phenotype_i + (1| patID)")

ggsave(paste0(MT_modeling_path, "parafovea_rate_by_phenos_volcano_MT_filtered_nominalP.png"), parafovea_rate_by_phenos_volcano_MT_filtered_nominalP,  width = h_w[2], height = h_w[1], dpi = 200)


# # add results to list
# parafovea_rate_results_df_filtered <- results_df_MT_filtered
# poster_results_dfs[["parafovea_rate_results_df_filtered"]] <- parafovea_rate_results_df_filtered



```

### step wise rate
#### subset data to scale and model
* filter out the ones with (rows>0) < 30
```{r}

var  = "parafovea"
# subset d with vars of interest 
model_df <- d %>% 
  select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, var))) %>% 
  as.data.frame()

# scale vars of interest, minus var: FA2_rate
model_df[, phenotypes[phenotypes!=var] ] <- scale(model_df[, phenotypes[phenotypes !=var] ])



model_df_lag <- mclapply(unique(model_df$patID_eye), function(x) stepwise_rate(x, df = model_df,  var = var, time_var = "month"), mc.cores = detectCores()-1) %>% 
  do.call(rbind,.)

response = sprintf("%s_rate", var)


```

#### model each variable against parafovea rate, correcting for SEXF, and random effect of patient dependence
> modeling rate by  patient's first phenotype does not have time effect, and therefore no repeated eye measures; but there are still repeated patients because each has two eyes

```{r}

# exclude the response variable, parafovea_rate
input_phenotypes <- setdiff( names(model_df_lag), c("month","patID_eye", "patID", "SEXF",response, var, "year"))
  
# formulas
form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID) + (1|patID_eye)")

# for each variable in input_phenotypes, model FA_mean 
results_list <- mclapply(form_strs, function(form_str){
  ## linear model with random effects
  formula <- as.formula(form_str)

  fit <- try({lmer(formula = formula, data = model_df_lag)})
  
  if(inherits(fit, "try-error")){
    return(NA)
  }else{
    return(summary(fit))
  }
}, mc.cores = detectCores()-1)

names(results_list) <- str_extract(form_strs, "\\S+\\s*~\\s*\\S+")

names(results_list)[is.na(results_list)]
# "parafovea_rate ~ first_OTHER_DISEASE" "parafovea_rate ~ first_SMKCIGSDAY" "parafovea_rate ~ first_t2d"

# filter for models that worked
results_list <- results_list[!is.na(results_list)]

## make results df ####
results_df <- extract_results(results_list)
parafovea_stepwise_results_df <- results_df

results_df <- results_df %>% 
  filter(!grepl("(Intercept)|SEXF|month", Term)) 

# filter for models that don't make sense
results_df_filtered <- results_df %>% 
  filter(!Term   %in% grep("avg|MT0|_.*0\\d$|MICRON|FOVTHICK|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars


# volcano plot of vaa~phenotypes result
parafovea_rate_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("stepwise_parafovea_rate ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

ggsave(paste0(MT_modeling_path, "parafovea_stepwise_lag_rate_by_phenos_volcano.png"), parafovea_rate_by_phenos_volcano, width = 20, height = 12, dpi = 200)

# volcano plot of vaa~phenotypes result
parafovea_rate_by_phenos_volcano_MT_filtered<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)),
                       size = 4,
                       nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("stepwise_parafovea_rate ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

ggsave(paste0(MT_modeling_path, "parafovea_stepwise_lag_rate_by_phenos_volcano_MT_filtered.png"), parafovea_rate_by_phenos_volcano_MT_filtered, width = 20, height = 12, dpi = 200)

parafovea_rate_by_phenos_volcano_MT_filtered_nominalP<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Term))+
      geom_point(show.legend = FALSE)+
      geom_label_repel(aes(label = ifelse( `Pr(>|t|)` < .05,  Term, NA)),
                       size = 4,
                       # nudge_y = 0.5,
                       show.legend = FALSE,
                       max.overlaps = 20)+
      geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+
      geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+
      theme_bw()+
      # facet_wrap(~quintile, scales = "fixed")+
      ggtitle("stepwise_parafovea_rate ~ SEXF + phenotype_i + (1| patID) + (1|patID_eye)")

ggsave(paste0(MT_modeling_path, "parafovea_stepwise_lag_rate_by_phenos_volcano_MT_filtered_nominalP.png"), parafovea_rate_by_phenos_volcano_MT_filtered_nominalP,  width = h_w[2], height = h_w[1], dpi = 200)


# # add results to list
# parafovea_rate_results_df_filtered <- results_df_MT_filtered
# poster_results_dfs[["parafovea_rate_results_df_filtered"]] <- parafovea_rate_results_df_filtered



```



### compare global to stepwise
```{r}

fun <- function(x){
x %>% 
  mutate(Term = str_remove_all(Term, "first_")) %>% 
  select(Term, `Pr(>|t|)`) %>% 
  mutate(neglogP = -log10(`Pr(>|t|)`)) %>% 
  select(-`Pr(>|t|)`) %>% 
  filter(!grepl("Intercept|SEX", Term))
}

inner_join(fun(parafovea_global_results_df), fun(parafovea_stepwise_results_df <- results_df), by = "Term", suffix = c("_global", "_stepwise")) %>% 
  column_to_rownames("Term") %>% 
  mutate_all(~-log10(.)) %>% 
  plot()


abline(a = 0, b = 1)


```

#___   ________   _____   ______
# model slope

<!-- ## *slope  -->
<!-- ### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->
<!-- #  -->
<!-- # #  variables (rows of value>0) < 30 where MT_slope_sans3 present -->
<!-- # m <- d[!is.na(d$MT_slope_sans3),phenotypes] != 0 -->
<!-- #  -->
<!-- # vars <- m %>% -->
<!-- #   colSums(., na.rm = TRUE)  -->
<!-- #  -->
<!-- # # variables with > 30 obs of value>0  -->
<!-- # deviant_vars <- vars[vars > 30] %>% names() -->
<!-- #  -->
<!-- # # how many unique months do these have where FA_mean and var available? -->
<!-- # count_unq_months <-  sapply(deviant_vars, function(x){  -->
<!-- #                     # how many unique months where both mr and var available? -->
<!-- #                     return( d $month[!is.na(d$MT_slope_sans3) & !is.na(d[,x])] %>%  -->
<!-- #                              unique() %>% -->
<!-- #                              length() -->
<!-- #                             ) -->
<!-- #                   }) -->
<!-- # # vars with greater than one unique month observed  -->
<!-- # slope_greater_than_one_month <- count_unq_months[count_unq_months>1] %>%  -->
<!-- #   names() -->
<!-- #  -->
<!-- # # setdiff(deviant_vars ,greater_than_one_month  ) # "BLRPATTERN"  "NOABNORMFAF" only have 1 month of observations where >30 records of rows of value > 0 -->

<!-- response = "MT_slope_sans3" -->
<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, response))) -->

<!-- # scale vars of interest, minus response: FA2_rate -->
<!-- model_df[, phenotypes[phenotypes!=response] ] <- scale(model_df[, phenotypes[phenotypes!=response] ]) -->


<!-- ``` -->

<!-- ### model each variable against slope correcting for SEXF, and random effect of patient dependence -->

<!-- ```{r} -->

<!-- # exclude the response variable, mt5 -->
<!-- input_phenotypes <- names(model_df)[!names(model_df) %in% c("month","patID_eye", "patID", "SEXF", "MT_slope_sans3")] -->

<!-- # formulas -->
<!-- form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID) + (1|patID_eye)") -->

<!-- # for each variable in input_phenotypes, model FA_mean  -->
<!-- results_list <- mclapply(form_strs, function(form_str){ -->

<!--   ## linear model with random effects -->
<!--   re_formula <- as.formula(form_str) -->

<!--   fit <- try({lmer(formula = re_formula, data = model_df)}) -->

<!--   return(summary(fit)) -->

<!-- }, mc.cores = detectCores()-1) -->

<!-- names(results_list) <- sub("\\s\\+.*", "", form_strs) -->

<!-- # filter for models that worked -->
<!-- results_list <- results_list[unlist(lapply(results_list, function(x) !all(class(x) %in% c("summaryDefault", "table"))))] -->

<!-- ## make results df #### -->
<!-- results_df <- extract_results(results_list) -->

<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->

<!-- # filter for models that don't make sense -->
<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Term   %in% grep("avg|MT0|_.*0\\d$|MICRON|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars -->


<!-- MT_slope_sans3_by_phenos_volcano_filtered <-ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("MT_slope_sans3 (-zone3) ~ SEXF + phenotype_i + (1| patID_eye) + (1|patID)") -->

<!-- ggsave(paste0(MT_modeling_path, "MT_slope_sans3_by_phenos_volcano_filtered.png"), MT_slope_sans3_by_phenos_volcano_filtered, width = 20, height = 12, dpi = 200) -->





<!-- ``` -->


<!-- ## *slope rate  -->
<!--  ### subset data to scale and model -->
<!-- * filter out the ones with (rows>0) < 30 -->
<!-- ```{r} -->


<!-- # #  variables (rows of value!=0) < 30 where MT_slope_sans3_rate present -->
<!-- # m <- d[!is.na(d$MT_slope_sans3_rate) , phenotypes] != 0 -->
<!-- # vars <- m %>% -->
<!-- #   colSums(., na.rm = TRUE) -->
<!-- #  -->
<!-- # # variables with > 30 obs of value != 0 -->
<!-- # deviant_vars <- vars[vars > 30] %>% names() -->
<!-- #  -->
<!-- # # how many unique months do these have where FA_mean and var available? -->
<!-- # count_unq_months <-  sapply(deviant_vars, function(x){ -->
<!-- #                     # how many unique months where both mr and var available? -->
<!-- #                     return( d $month[!is.na(d$MT_slope_sans3_rate) & !is.na(d[,x])] %>% -->
<!-- #                              unique() %>% -->
<!-- #                              length() -->
<!-- #                             ) -->
<!-- #                   }) -->
<!-- # # vars with greater than one unique month observed -->
<!-- # slope_rate_greater_than_one_month <- count_unq_months[count_unq_months>1] %>% -->
<!-- #   names() -->


<!-- # # filter out vars that dont change where there is an fameanrate -->
<!-- # vars_dont_change <- phenotypes[ sapply(d[  !is.na(d$MT_slope_sans3_rate), phenotypes], n_distinct) <= 1] -->
<!-- #  -->
<!-- # # how many unique months do these have where MT_slope_sans3_rate and var available?  -->
<!-- # count_unq_months <- sapply(phenotypes, function(x){ -->
<!-- #    -->
<!-- #   num_months <- d$month[!is.na(d$MT_slope_sans3_rate) & !is.na(d[[x]]) ] %>%  -->
<!-- #       unique() %>%  -->
<!-- #       length() -->
<!-- #   return(num_months) -->
<!-- # }) -->
<!-- #  -->
<!-- # vars_few_month <- count_unq_months[count_unq_months<2] %>%  -->
<!-- #   names() -->
<!-- #  -->
<!-- # vars_dont_change_and_few_month <- c(vars_dont_change, vars_few_month) %>% unique() -->
<!-- #  -->
<!-- # slope_rate_greater_than_one_month <- c("MT_slope_sans3_rate", phenotypes[!phenotypes %in% vars_dont_change_and_few_month]) -->

<!-- response = "MT_slope_sans3_rate" -->
<!-- # subset d with vars of interest  -->
<!-- model_df <- d %>%  -->
<!--   select(month, patID_eye, patID, SEXF, all_of(c(phenotypes, response))) -->

<!-- # scale vars of interest, minus response: FA2_rate -->
<!-- model_df[, phenotypes[phenotypes!=response] ] <- scale(model_df[, phenotypes[phenotypes!=response] ]) -->


<!-- # make df of first visit phenotype values;  if first visit value @ phenotype is NA, grab first non-na, else, NA; exclude duplicated rows -->
<!-- first_values_df <- model_df %>%  -->
<!--   select(-patID, - SEXF,-MT_slope_sans3_rate ) %>%  -->
<!--   arrange(patID_eye , month) %>%   -->
<!--   group_by(patID_eye) %>% -->
<!--   mutate( -->
<!--     across( -->
<!--         everything(),  -->
<!--         ~dplyr::first(.[!is.na(.)]), -->
<!--         .names = "first_{.col}" -->
<!--   )) %>% -->
<!--   ungroup() %>% -->
<!--   select(-matches("month"), -matches("slope_rate")) %>%  -->
<!--   select(patID_eye,  starts_with("first_")) %>%  -->
<!--   unique() %>%  -->
<!--   arrange(patID_eye) -->

<!-- # join the first df to the model df -->
<!-- model_df_first <- model_df %>%  -->
<!--   arrange(patID_eye) %>%  -->
<!--   select(patID_eye, patID, SEXF, MT_slope_sans3_rate) %>%  -->
<!--   na.omit() %>% -->
<!--   unique() %>%   -->
<!--   left_join(., first_values_df) -->

<!-- # # scale model_df_first -->
<!-- # model_df_first <- model_df_first %>%  -->
<!-- #   mutate(across(-c( patID_eye, patID, SEXF) , ~as.vector(scale(.))) )    -->

<!-- # # filter out vars with sd na or 0 -->
<!-- # sds <- model_df_first %>%  -->
<!-- #   select(-c(patID_eye, patID, SEXF)) %>%  -->
<!-- #   sapply(., function(x)sd(x, na.rm = TRUE) %>% round) -->
<!-- #  -->
<!-- # # model df of ids and vars with sd not na or 0 -->
<!-- # model_df_first <- model_df_first %>%  -->
<!-- #   select(patID_eye, patID, SEXF, MT_slope_sans3_rate, all_of( names(sds[sds !=0 & !is.na(sds)]) )  ) -->


<!-- ``` -->

<!-- ### model each variable against slope rate, correcting for SEXF, and random effect of patient dependence -->
<!-- ```{r} -->

<!-- # exclude the response variable, MT_slope_sans3_rate -->
<!-- input_phenotypes <- names(model_df_first)[!names(model_df_first) %in% c("month","patID_eye", "patID", "SEXF", "MT_slope_sans3_rate")] -->

<!-- # formulas -->
<!-- form_strs <- paste0(response, " ~ ",input_phenotypes," + SEXF  + (1|patID)") -->

<!-- # for each variable in input_phenotypes, model FA_mean  -->
<!-- results_list <- mclapply(form_strs, function(form_str){ -->

<!--   ## linear model with random effects -->
<!--   re_formula <- as.formula(form_str) -->

<!--   fit <- try({lmer(formula = re_formula, data = model_df_first)}) -->

<!--   return(summary(fit)) -->

<!-- }, mc.cores = detectCores()-1) -->

<!-- names(results_list) <- sub("\\s\\+.*", "", form_strs) -->

<!-- # filter for models that worked -->
<!-- results_list <- results_list[unlist(lapply(results_list, function(x) !all(class(x) %in% c("summaryDefault", "table"))))] -->

<!-- ## make results df #### -->
<!-- results_df <- extract_results(results_list) -->

<!-- results_df <- results_df %>%  -->
<!--   filter(!Term %in% c("(Intercept)", "SEXF", "month")) -->

<!-- # filter for models that don't make sense -->
<!-- results_df_filtered <- results_df %>%  -->
<!--   filter(!Term   %in% grep("avg|MT0|_.*0\\d$|MICRON|SEX", results_df$Term, value = TRUE)) # exclude MT vars and zonal vars -->




<!-- # Plotting ------- -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- MT_slope_sans3_rate_by_phenos_volcano <-ggplot(results_df, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("avg_slope rate(-3) ~ SEXF + phenotype_i + (1| patID)") -->

<!-- # ggsave(paste0(MT_modeling_path, "MT5_slope_rate_by_phenos_volcano.png"), MT5_slope_rate_by_phenos_volcano, width = 20, height = 12, dpi = 200) -->
<!-- #  -->
<!-- # # volcano plot of vaa~phenotypes result -->
<!-- # MT5_slope_rate_by_phenos_volcano_otherZones_filtered<- ggplot(results_df_otherZones_filtered, aes(x = Estimate, y = -log10(q_value), color = Model))+ -->
<!-- #       geom_point(show.legend = FALSE)+ -->
<!-- #       geom_label_repel(aes(label = ifelse( q_value <= .05,  Model, NA)), -->
<!-- #                        size = 4, -->
<!-- #                        nudge_y = 0.5, -->
<!-- #                        show.legend = FALSE, -->
<!-- #                        max.overlaps = 20)+ -->
<!-- #       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!-- #       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!-- #       theme_bw()+ -->
<!-- #       # facet_wrap(~quintile, scales = "fixed")+ -->
<!-- #       ggtitle("MT5_slope rate ~ SEXF + phenotype_i + (1| patID)") -->
<!-- #  -->
<!-- # ggsave(paste0(MT_modeling_path, "MT5_slope_rate_by_phenos_volcano_otherZones_filtered.png"), MT5_slope_rate_by_phenos_volcano_otherZones_filtered, width = 20, height = 12, dpi = 200) -->

<!-- # volcano plot of vaa~phenotypes result -->
<!-- MT_slope_sans3_rate_by_phenos_volcano_MT_filtered<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( q_value <= .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("avg_slope(-3) rate ~ SEXF + phenotype_i + (1| patID)") -->

<!-- ggsave(paste0(MT_modeling_path, "MT_slope_sans3_rate_by_phenos_volcano_MT_filtered.png"), MT_slope_sans3_rate_by_phenos_volcano_MT_filtered, width = 20, height = 12, dpi = 200) -->

<!-- MT_slope_sans3_rate_by_phenos_volcano_MT_filtered_nominalP<- ggplot(results_df_filtered, aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Term))+ -->
<!--       geom_point(show.legend = FALSE)+ -->
<!--       geom_label_repel(aes(label = ifelse( `Pr(>|t|)` < .05,  Term, NA)), -->
<!--                        size = 4, -->
<!--                        # nudge_y = 0.5, -->
<!--                        show.legend = FALSE, -->
<!--                        max.overlaps = 20)+ -->
<!--       geom_abline(slope = 0, intercept = -log10(.05), linetype = 2, color = "red")+ -->
<!--       geom_vline(xintercept = 0, color = "darkgrey", linetype = "dashed")+ -->
<!--       theme_bw()+ -->
<!--       # facet_wrap(~quintile, scales = "fixed")+ -->
<!--       ggtitle("avg_slope rate(-3) ~ SEXF + phenotype_i + (1| patID)") -->

<!-- ggsave(paste0(MT_modeling_path, "MT_slope_sans3_rate_by_phenos_volcano_MT_filtered_nominalP.png"), MT_slope_sans3_rate_by_phenos_volcano_MT_filtered_nominalP,width = h_w[2], height = h_w[1], dpi = 400) -->

<!-- #  -->
<!-- # # add results to list -->
<!-- # MT_slope_sans3_rate_results_df_filtered <- results_df_MT_filtered -->
<!-- # poster_results_dfs[["MT_slope_sans3_rate_results_df_filtered"]] <- MT_slope_sans3_rate_results_df_filtered -->



<!-- ``` -->

#________________________
# heatmap of structural  variables
```{r}

str_vars <- names(d)[sapply(d, function(x) is.numeric(x)|is.factor(x))] %>% 
  grep("eye|VAAM|month|DIAG|GRADER|msng|quint|pat|vis|SITE|SMK", .,value = TRUE, invert = TRUE)
  
setdiff(names(d), str_vars)

m <- d %>% 
  ungroup() %>% 
  select(all_of(str_vars)) %>%
  select(where(~ mean(is.na(.)) < 1)) %>% 
  # select(all_of(sub_str_vars)) %>% 
  mutate_all(~as.numeric(as.character(.))) %>% 
  as.matrix() %>% 
  scale()

cor_m <- cor(m, use = "pairwise.complete.obs")

col_dend <- as.dendrogram(hclust(dist(cor_m), 
                                 method = "ward.D2"
                                 # method = "complete"
                                 ))
row_dend <- as.dendrogram(hclust(dist(cor_m^2),
                                 method = "ward.D2"
                                 # method = "complete"
                                 ))

hm <- Heatmap(cor_m, 
              column_title = "Feature correlations",
              na_col = "grey", 
              cluster_rows = TRUE,
              cluster_columns = TRUE,
              # cluster_rows = row_dend, 
              # cluster_columns = col_dend,
              # cluster_columns = row_dend,
              clustering_method_columns = "ward.D2",
              clustering_method_rows = "ward.D2",
              row_names_gp = gpar(fontsize = 7),
              column_names_gp = gpar(fontsize = 7), 
              row_dend_width = unit(5, "cm"),
              column_dend_height = unit(5, "cm")
              )

png(paste0(wd_path, "results/modeling/str_vars_clustered_hm.png"), height = 2000, width = 2000)
draw(hm)
dev.off()

```

#____________________________
#|||||||||||||||||||||

# Save data
```{r}


saveHDF5SummarizedExperiment(sce, paste0(wd_path, "processed_data/sce003.h5"), replace = T)

notify.me("10_functionAndStructureModeling.Rmd Completed")
# D
# new_data_name <- data_name %>% 
#   gsub("6", "7", .)
# saveRDS(d, paste0(wd_path, "processed_data/", new_data_name))
# # info
# write.xlsx(info, paste0(wd_path, "processed_data/database_grading_key_05.xlsx"))
# # raw
# saveRDS(raw, paste0(wd_path , "processed_data/dfs_list_03.rds"))
```

```{r}
check_dims()
# dims : 12525 143
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ no change

### APRIL 2 2025 ###
# dims : 12276 180
# n_patIDs             3971
# n_patID_eyes         7888
# n_observations      12275
# n_patIDs_w_gr_1_vis   492


```

