---
title: "11_ClusterModeling"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---

# options
```{r}
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()
```


# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

pacman::p_load(tidyr, dplyr, reshape2, purrr, GGally, tidyverse, ggplot2, gridExtra, ComplexHeatmap, viridis, zoo, openxlsx, plotly, mgcv, zoo, tools, mice, ggrepel, networkD3, htmltools, Metrics, psych, car, stats, lme4, lmerTest, ape, SingleCellExperiment, scater, parallel, future, HDF5Array, patchwork)

rm(list = ls())
gc()

```

# functions
```{r}

source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")
how.long()

```

# Load Data
info and d from sce
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

# load experiment
sce <- loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/sce003.h5"))

d <- assays(sce)$pmm_imp_of_scaled %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  left_join( colData(sce) %>% data.frame())

```


# make new directorys for plots
```{r}
path <- paste0(wd_path, "results/modeling/")
if(!dir.exists(path)) {
  dir.create(path, recursive = TRUE)
  
}


cluster_modeling_path <- paste0(wd_path, "results/modeling/cluster_modeling/")
if(!dir.exists(cluster_modeling_path)) {
  dir.create(cluster_modeling_path, recursive = TRUE)
  
}

```

# mung data

```{r, look at dims}
check_dims <- function(){
  
  D <- as.data.frame(colData(sce)) %>% 
  left_join(
    as.data.frame(t(assays(sce)$scaled_measure_vars)) %>% 
      rownames_to_column("sample")
  ) 
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()
# dims : 12525 143
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ no change

### APRIL 2 2025 ###
# dims : 12276 180
# n_patIDs             3971
# n_patID_eyes         7888
# n_observations      12275
# n_patIDs_w_gr_1_vis   492

```


## make cluster dichotomous
```{r}


dichotomous_clusters <- data.frame(sample = d$sample, model.matrix(~0+clust, data = d %>% mutate(clust = my_cluster)))

# join dichotomous clusters to sce and d
colData(sce) <- cbind(colData(sce), dichotomous_clusters)

d <- d %>% 
  left_join(dichotomous_clusters)


```

## calculate pseudoprog rate
```{r}

# calculae rate of pseudoprog (pt) as lastpt - firstpt / lastmonth - firstmonth (where pt not na)
pseudoprog_rate_d <- d %>% 
  select(patID_eye, patID, month, pseudoprog, regr_fxd_pseudoprog) %>% 
  arrange(month) %>% 
  group_by(patID_eye) %>% 
  mutate(pseudoprog_rate =
          (dplyr::last(pseudoprog[!is.na(pseudoprog)]) - dplyr::first(pseudoprog[!is.na(pseudoprog)]))
              /
          (dplyr::last(month[!is.na(pseudoprog)]) - dplyr::first(month[!is.na(pseudoprog)]))
           ) %>% 
  mutate(pseudoprog_rate = 
           case_when(
             is.infinite(pseudoprog_rate) ~ NA, 
             .default = pseudoprog_rate
           )) %>% 
  mutate(regr_fxd_pseudoprog_rate =
          (dplyr::last(regr_fxd_pseudoprog[!is.na(regr_fxd_pseudoprog)]) - dplyr::first(regr_fxd_pseudoprog[!is.na(regr_fxd_pseudoprog)]))
              /
          (dplyr::last(month[!is.na(regr_fxd_pseudoprog)]) - dplyr::first(month[!is.na(regr_fxd_pseudoprog)]))
           ) %>% 
  mutate(regr_fxd_pseudoprog_rate = 
           case_when(
             is.infinite(regr_fxd_pseudoprog_rate) ~ NA, 
             .default = regr_fxd_pseudoprog_rate
           )) %>% 
  ungroup()

pseudoprog_rate_d %>%
  select(matches("pseudoprog")) %>%
  plot()


# join pseudoprogs to d
d <- d %>% 
  left_join(pseudoprog_rate_d)


```


#___________


# Model pseudoprog 
## model df 
select vars that change and scale them
```{r}

# vars used in modeling that arn't predictors
special_vars <- c( "patID_eye", 'eye_num',"patID", "month","year", "SEXF", "pseudoprog", "pseudoprog_rate", "regr_fxd_pseudoprog_rate", "regr_fxd_pseudoprog")

# what vars are numeric
numeric_cols <- names(d)[sapply(d, is.numeric)]

names(d)[sapply(d, is.factor)]
 
# how do vars change
sds <- d %>%
  select(all_of(numeric_cols)) %>% 
  sapply(., function(a_var) sd(a_var, na.rm = TRUE))

sd_gr_zero <- names(sds)[sds!=0]

# grater than one month of non-missing
sd_gr_zero_count_unq_months <-  sapply(sd_gr_zero, function(x){ 
                    # how many unique months where both var of interest and var available?
                    return( d $month[!is.na(d$pseudoprog) & !is.na(d[,x])] %>% 
                             unique() %>%
                             length()
                            )
                  })


good_vars <- names(sd_gr_zero_count_unq_months)[sd_gr_zero_count_unq_months>1]

good_vars <- good_vars %>% 
  grep("VAA|SITE|visit|GRADER|septile|YROF|eye|EHTN", ., value = TRUE, invert = TRUE)

# filter out vars that don't change and ones where < 1 month non-na
pp_model_df <- d %>% 
  select(all_of(special_vars), all_of( good_vars) ) %>% 
  select(-matches("cirrus|stratus|spectralis"))

# scale predictors
pp_model_df <- pp_model_df %>% 
  mutate(across(-all_of(special_vars), ~as.numeric(scale(.))))


predictor_vars <- predictors <- names(pp_model_df)[-which(names(pp_model_df) %in% special_vars)]


```

## Correlation:  model pseudoprog as function of phenotypes
### correlation modeling
```{r extract results function}

# ****** FUNCTION: make results df from summaries list ******
## take list of model summaries and fill dataframe; filter out intercept and correction variables; filter for Pr>0
extract_results <- function( summaries = list()){

  # empty df
  results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
                         Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)

  # fill df with model values
  for (model_name in names(summaries)) {

      model_coef <- coef( summaries[[model_name]] )

      a_result_df <- data.frame(Model = model_name,
                             Term = rownames(model_coef),
                             Estimate = model_coef[, "Estimate"],
                             Std.Error = model_coef[, "Std. Error"],
                             t.value = model_coef[, "t value"],
                             Pr = model_coef[, "Pr(>|t|)"])

      results_df <- rbind(results_df, a_result_df)
      }

      # calculate adjusted p value
      results_df$q_value <- p.adjust(results_df$Pr, method = "BH")

      # filter control variables from results
      results_df <- results_df %>%
      filter(!Term %in% c("(Intercept)", "SEXF", "month", "patID_eye")) %>%
      filter(!Model %in% c("patID_eye"))

      # filter pr of 0
      results_df <- results_df %>%
        filter(Pr>0)
}

```


<!-- ```{r,} -->

<!-- ## model pseudoprog raw values---- -->
<!-- # loop through all but special vars and model pseudoprog -->
<!-- response <-  "pseudoprog" -->

<!-- predictors <- names(pp_model_df)[-which(names(pp_model_df) %in% special_vars)] -->


<!-- # run models in parallel -->
<!-- pseudoprog_summaries <- mclapply(predictors, function(predictor) { -->

<!--   a_formula <- as.formula(paste0( response, "~ SEXF + ", predictor, " + (1|patID) + (1|patID_eye)"  )) -->
<!--   print(a_formula ) -->

<!--   summ <- try({summary(lmer(a_formula, data = pp_model_df))}) -->
<!--   if(inherits(summ, "try-error")){ -->
<!--     return(NA) -->
<!--   }else{ -->
<!--     return(summ) -->
<!--   } -->

<!-- },mc.cores= detectCores()-1) -->

<!-- names(pseudoprog_summaries) <- predictors -->

<!-- # filter out errors -->
<!-- names(pseudoprog_summaries)[is.na(pseudoprog_summaries)] #  "NOABNORMCF"  "NOABNORMFL"  "NOABNORMFAF" "NOABNORMOCT" -->
<!-- pseudoprog_summaries <- pseudoprog_summaries[!is.na(pseudoprog_summaries)] -->

<!-- # extract results and plot -->
<!-- pseudoprog_results = extract_results(pseudoprog_summaries) -->

<!-- pseudoprog_volcano = ggplot(pseudoprog_results, aes(Estimate, -log10(q_value), color = Model))+ -->
<!--     geom_hline(yintercept = -log10(.05), linetype = 3, color = "black")+ -->
<!--     geom_vline(xintercept = 0, linetype = 2, color = "grey")+ -->
<!--     geom_point(show.legend = FALSE)+ -->
<!--      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)), -->
<!--                          size = 4, -->
<!--                          # nudge_y = 0.5,   -->
<!--                          show.legend = FALSE,  -->
<!--                          max.overlaps = 20)+ -->
<!--       ggtitle(paste0("Model: pseudoprog ~ SEXF + _phenotype_ + (1|patID) + (1|patID_eye)"))+ -->
<!--     theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/modeling/cluster_modeling/pseudoprog_volcano.png"), pseudoprog_volcano,height = 20, width = 25) -->

<!-- ## model regressioncorrected pseudoprog raw values---- -->
<!-- # loop through all but special vars and model pseudoprog -->
<!-- response <-  "regr_fxd_pseudoprog" -->
<!-- regr_fxd_pseudoprog_summaries <- mclapply(predictors, function(predictor) { -->

<!--  a_formula <- as.formula(paste0( response, "~ SEXF + ", predictor, " + (1|patID) + (1|patID_eye)"  )) -->
<!--   print(a_formula ) -->

<!--   summ <- try({summary(lmer(a_formula, data = pp_model_df))}) -->
<!--   if(inherits(summ, "try-error")){ -->
<!--     return(NA) -->
<!--   }else{ -->
<!--     return(summ) -->
<!--   } -->

<!-- },mc.cores= detectCores()-1) -->
<!-- names(regr_fxd_pseudoprog_summaries) <- predictors -->

<!-- # filter out errors -->
<!-- names(regr_fxd_pseudoprog_summaries)[is.na(regr_fxd_pseudoprog_summaries)] #  "NOABNORMCF"  "NOABNORMFL"  "NOABNORMFAF" "NOABNORMOCT" -->
<!-- regr_fxd_pseudoprog_summaries <- regr_fxd_pseudoprog_summaries[!is.na(regr_fxd_pseudoprog_summaries)] -->


<!-- # extract results and plot -->
<!-- regr_fxd_pseudoprog_results <- extract_results(regr_fxd_pseudoprog_summaries) -->

<!-- regr_fxd_pseudoprog_volcano <- ggplot(regr_fxd_pseudoprog_results, aes(Estimate, -log10(q_value), color = Model))+ -->
<!--     geom_hline(yintercept = -log10(.05), linetype = 3, color = "black")+ -->
<!--     geom_vline(xintercept = 0, linetype = 2, color = "grey")+ -->
<!--     geom_point(show.legend = FALSE)+ -->
<!--      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)), -->
<!--                          size = 4, -->
<!--                          # nudge_y = 0.5,   -->
<!--                          show.legend = FALSE,  -->
<!--                          max.overlaps = 20)+ -->
<!--       ggtitle(paste0("Model: regr_fxd_pseudoprog ~ SEXF + _phenotype_ + (1|patID)"))+ -->
<!--     theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/modeling/cluster_modeling/regr_fxd_pseudoprog_volcano.png"), regr_fxd_pseudoprog_volcano,height = 20, width = 25) -->


<!-- ``` -->

<!-- #### compare fixed versus original pseudoprogression estimates -->
<!-- ```{r} -->

<!-- # LOOK AT RELATIONSHIP BETWEEN ESTIMATEs ---- -->
<!-- compare <- pseudoprog_results %>%  -->
<!--   filter( !grepl("SEX" , Term)) %>%  -->
<!--   mutate(response = 'pseudoprog') %>%  -->
<!--   rbind(regr_fxd_pseudoprog_results %>% filter( !grepl("SEX" , Term)) %>% mutate(response = 'regr_fxd_pseudoprog')) -->

<!-- pr <- compare %>% -->
<!--   select(Term, Estimate, response) %>% -->
<!--   pivot_wider(names_from = response, values_from = Estimate) %>%  -->
<!--   ggplot(aes(pseudoprog, regr_fxd_pseudoprog, color = Term))+ -->
<!--   geom_abline(slope = 1, lty = 3)+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   ggtitle("Estimate") -->


<!-- est <- compare %>% -->
<!--   select(Term, Pr, response) %>% -->
<!--   pivot_wider(names_from = response, values_from = Pr) %>%  -->
<!--   ggplot(aes(-log10(pseudoprog), -log10(regr_fxd_pseudoprog), color = Term))+ -->
<!--   geom_abline(slope = 1, lty = 3)+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   ggtitle("Pr") -->


<!-- pr/est -->

<!-- ``` -->



## Prognosis: model pseudoprog rate as function of phenotypes
### Global rate
#### original pseudoprogression
<!-- # ```{r} -->
<!-- # # ****** FUNCTION: calculate combination of each x with following x for each eye ****** -->
<!-- # stepwise_rate <- function(a_eye, df = NULL, var = NA) { -->
<!-- #      -->
<!-- #   if(is.na(var)) stop("Response NA. Please provide variable for calculate steppwise rate") -->
<!-- #    -->
<!-- #    -->
<!-- #   # Filter for eye of interest -->
<!-- #   a_eye_diff_d <- df[df$patID_eye == a_eye, ] -->
<!-- #    -->
<!-- #   # make year var -->
<!-- #   if(!"year" %in% names(a_eye_diff_d)){ -->
<!-- #     a_eye_diff_d <- a_eye_diff_d %>%  -->
<!-- #       mutate(year = round(month/12, 2)) -->
<!-- #   } -->
<!-- #    -->
<!-- #   # Create year to following year combinations -->
<!-- #   year_combos <- data.frame(year = a_eye_diff_d$year, post_year = lead(a_eye_diff_d$year))  -->
<!-- #    -->
<!-- #   # Create var to every other var combinations -->
<!-- #   var_combos <- data.frame(var = a_eye_diff_d[[var]], post_var = lead(a_eye_diff_d[[var]]))  -->
<!-- #      -->
<!-- #   # Time pp combinations -->
<!-- #   year_var_combos <- cbind(year_combos, var_combos) -->
<!-- #    -->
<!-- #   # make rate for each i to i+1: change in var over change in time>> postx-x/postyear-year -->
<!-- #   year_var_combos$var_rate <- with(year_var_combos , (post_var-var)/(post_year-year)) -->
<!-- #    -->
<!-- #   # Join back to dataset -->
<!-- #   a_eye_diff_d <- suppressMessages( -->
<!-- #     a_eye_diff_d %>%  -->
<!-- #       left_join(year_var_combos, by = c("year"))  -->
<!-- #     ) -->
<!-- #    -->
<!-- #   # remove functional vars -->
<!-- #   a_eye_diff_d$var <- NULL -->
<!-- #   a_eye_diff_d$post_year <- NULL -->
<!-- #   a_eye_diff_d$post_var <- NULL -->
<!-- #    -->
<!-- #    -->
<!-- #   # rename "var" to var -->
<!-- #   names(a_eye_diff_d) <- str_replace_all(names(a_eye_diff_d), "var", var) -->
<!-- #    -->
<!-- #   return(a_eye_diff_d) -->
<!-- # } -->
<!-- #  -->
<!-- # ``` -->
<!-- #  -->
<!-- #  -->

##### prognosis model df
     
     appears in 13_routeModeling this one does not use same architeture as other pr~first mods

<!-- ```{r} -->

<!-- # # make first values df ---- -->
<!-- # ## grab first value of each varaiable -->
<!-- # orig_pp_firsts <- pp_model_df %>% -->
<!-- #   filter(!is.na(pseudoprog_rate)) %>% -->
<!-- #   arrange(month) %>% -->
<!-- #   select(patID_eye,  all_of(predictor_vars) ) %>% -->
<!-- #   group_by(patID_eye) %>% -->
<!-- #   mutate(across(everything(), -->
<!-- #     # ~ dplyr::first(.[!is.na(.)]), -->
<!-- #     ~ dplyr::first(.), -->
<!-- #     .names = "first_{.col}" -->
<!-- #               ) )%>% -->
<!-- #   select(patID_eye, matches("first_")) %>% -->
<!-- #   unique() -->
<!-- #  -->
<!-- # table(orig_pp_firsts$patID_eye)[table(orig_pp_firsts$patID_eye)>1] -->
<!-- #  -->
<!-- # # join back the special vars -->
<!-- # orig_pp_model_df_first <- orig_pp_firsts %>% -->
<!-- #   left_join(pp_model_df %>% select(patID_eye,patID, SEXF, pseudoprog_rate ) %>% unique()) -->
<!-- #  -->
<!-- # table(orig_pp_model_df_first$patID_eye)[table(orig_pp_model_df_first$patID_eye)>1] -->
<!-- #  -->
<!-- # colSums(is.na(orig_pp_model_df_first)) -->

<!-- # make first values df ---- -->
<!-- ## grab first value of each varaiable -->
<!-- orig_pp_model_df_first <- data.frame(colData(sce)) %>% -->
<!--   left_join(assays(sce)$pmm_imp_of_scaled %>% t() %>% data.frame() %>% rownames_to_column("sample")) %>% -->
<!--   arrange(month) %>% -->
<!--   mutate(year = month/12) %>% -->
<!--   group_by(patID_eye) %>% -->
<!--   mutate(pp_rate = -->
<!--           (dplyr::last(pseudoprog[!is.na(pseudoprog)]) - dplyr::first(pseudoprog[!is.na(pseudoprog)])) -->
<!--               / -->
<!--           (dplyr::last(year[!is.na(pseudoprog)]) - dplyr::first(year[!is.na(pseudoprog)])) -->
<!--            ) %>% -->
<!--   # mutate(pp_rate = (dplyr::last(pseudoprog) - dplyr::first(pseudoprog))/(dplyr::last(year) - dplyr::first(year))) %>%  -->
<!--   mutate(pp_rate = ifelse(is.infinite(pp_rate), NA, pp_rate)) %>% -->
<!--   filter(!is.na(pp_rate)) %>% -->
<!--   select(patID_eye, patID,SEXF, pp_rate, any_of(predictor_vars) ) %>% -->
<!--   mutate(across(any_of(predictor_vars)), ~as.numeric(scale(.))) %>%  -->
<!--   slice(1) %>% -->
<!--   ungroup() -->



<!-- table(orig_pp_model_df_first$patID_eye)[table(orig_pp_model_df_first$patID_eye)>1] -->

<!-- colSums(is.na(orig_pp_model_df_first)) -->


<!-- ``` -->

<!-- ##### prognosis modeling -->
<!-- ```{r} -->

<!-- # model pseudoprog rate ---- -->
<!-- # loop through all but special vars and model response  -->

<!-- response <-  "pp_rate" -->
<!-- # response <-  "pseudoprog_rate" -->
<!-- pseudoprog_rate_summaries <- sapply(predictors, function(predictor){ -->
<!-- # pseudoprog_rate_summaries <- sapply(sprintf("first_%s",predictors), function(predictor){ -->
<!--     a_formula <- as.formula(paste0( response, "~ SEXF + ", predictor, " + (1|patID)"  )) -->
<!--     summ <- try({summary(lmer(a_formula, data = orig_pp_model_df_first))}) -->
<!--     if(inherits(summ, "try-error")){ -->
<!--       return(NA) -->
<!--     }else{ -->
<!--       return(summ) -->
<!--     } -->
<!-- }, simplify = FALSE) -->

<!-- names(pseudoprog_rate_summaries)[is.na(pseudoprog_rate_summaries)] #"first_NOABNORMCF"  "first_NOABNORMFL"  "first_NOABNORMFAF" "first_NOABNORMOCT" "first_chew_grade"  "first_t2d"         -->
<!-- pseudoprog_rate_summaries <- pseudoprog_rate_summaries[!is.na(pseudoprog_rate_summaries)] -->
<!-- orig_pseudoprog_rate_results <- extract_results(pseudoprog_rate_summaries)   -->


<!-- # filter out VA stuff, rates, and pt derivatives -->
<!-- orig_pseudoprog_rate_results <- orig_pseudoprog_rate_results %>%  -->
<!--   filter(!Model %in% grep("ACUIT|LOGMAR|SNELL|pseudo|severity|rate|RW|variance|FA|letters", orig_pseudoprog_rate_results$Model, value = TRUE) ) %>%  -->
<!--   mutate(Model = Model %>% gsub("avg_MT", "MT", .)) -->

<!-- orig_pseudoprog_rate_results %>%  -->
<!--   filter(Estimate<1) %>% -->
<!--   filter(!grepl("clust|chew", Term)) %>%  -->
<!-- ggplot( aes(Estimate, -log10(q_value), color = Model))+ -->
<!--     geom_hline(yintercept = -log10(.05), linetype = 3, color = "black")+ -->
<!--     geom_vline(xintercept = 0, linetype = 2, color = "grey")+ -->
<!--     geom_point(show.legend = FALSE)+ -->
<!--      geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)), -->
<!--                          size = 4, -->
<!--                          # nudge_y = 0.5,   -->
<!--                          show.legend = FALSE,  -->
<!--                          max.overlaps = 20)+ -->
<!--     ggtitle(paste0("Model: pseudoprog_rate ~ SEXF + _phenotype_ + (1|patID)"))+ -->
<!--     theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/modeling/cluster_modeling/structure_orig_pseudoprog_rate_volcano.png"), height = 15, width = 20) -->



<!-- orig_pseudoprog_rate_results %>%  -->
<!--   filter(grepl("clust", Term)) %>%  -->
<!--   ggplot(aes(Estimate, -log10(q_value), color = Model))+ -->
<!--     geom_hline(yintercept = -log10(.05), linetype = 3, color = "black")+ -->
<!--     geom_vline(xintercept = 0, linetype = 2, color = "grey")+ -->
<!--     geom_point(show.legend = FALSE)+ -->
<!--      geom_label_repel(aes( -->
<!--        # label = ifelse( q_value < .05,  Model, NA)), -->
<!--        label = Model), -->
<!--                          size = 4, -->
<!--                          # nudge_y = 0.5,   -->
<!--                          show.legend = FALSE,  -->
<!--                          max.overlaps = 20)+ -->
<!--     ggtitle(paste0("Model: pseudoprog_rate ~ SEXF + _phenotype_ + (1|patID)"))+ -->
<!--     theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/modeling/cluster_modeling/cluster_orig_pseudoprog_rate_volcano.png"), height = 10, width = 12) -->

<!-- ``` -->

<!-- #### regression fixed pseudoprogression -->
<!-- ##### prognosis model df -->
<!-- ```{r} -->

<!-- # make first values df ---- -->
<!-- ## grab first value of each varaiable  -->
<!-- regr_fxd_pseudoprog_firsts <- pp_model_df %>%  -->
<!--   filter(!is.na(regr_fxd_pseudoprog_rate)) %>%  -->
<!--   arrange(month) %>%  -->
<!--   select(patID_eye,  all_of(predictor_vars) ) %>%  -->
<!--   group_by(patID_eye) %>%  -->
<!--   mutate(across(everything(),  -->
<!--     ~ dplyr::first(.[!is.na(.)]),  -->
<!--     .names = "first_{.col}" -->
<!--               ) )%>%  -->
<!--   select(patID_eye, matches("first_")) %>%  -->
<!--   unique()  -->

<!-- table(regr_fxd_pseudoprog_firsts$patID_eye)[table(regr_fxd_pseudoprog_firsts$patID_eye)>1] -->

<!-- # join back the special vars -->
<!-- regr_fxd_pseudoprog_model_df_first <- regr_fxd_pseudoprog_firsts %>%  -->
<!--   left_join(pp_model_df %>% select(patID_eye,patID, SEXF, regr_fxd_pseudoprog_rate ) %>% unique()) -->


<!-- ``` -->

<!-- ##### prognosis modeling -->
<!-- ```{r} -->

<!-- # model pseudoprog rate ---- -->
<!-- # loop through all but special vars and model response  -->
<!-- response <-  "regr_fxd_pseudoprog_rate" -->

<!-- regr_fxd_pseudoprog_rate_summaries <- sapply(sprintf("first_%s",predictors), function(predictor){ -->
<!--     a_formula <- as.formula(paste0( response, "~ SEXF + ", predictor, " + (1|patID)"  )) -->
<!--     summ <- try({summary(lmer(a_formula, data = regr_fxd_pseudoprog_model_df_first))}) -->
<!--     if(inherits(summ, "try-error")){ -->
<!--       return(NA) -->
<!--     }else{ -->
<!--       return(summ) -->
<!--     } -->
<!-- }, simplify = FALSE) -->

<!-- names(regr_fxd_pseudoprog_rate_summaries)[is.na(regr_fxd_pseudoprog_rate_summaries)] # "NOABNORMCF"  "NOABNORMFL"  "NOABNORMFAF" "NOABNORMOCT" -->
<!-- regr_fxd_pseudoprog_rate_summaries <- regr_fxd_pseudoprog_rate_summaries[!is.na(regr_fxd_pseudoprog_rate_summaries)] -->
<!-- regr_fxd_pseudoprog_rate_results <- extract_results(regr_fxd_pseudoprog_rate_summaries)   -->


<!-- regr_fxd_pseudoprog_rate_results <- extract_results(regr_fxd_pseudoprog_rate_summaries) -->

<!-- # filter out VA stuff, rates, and pt derivatives -->
<!-- regr_fxd_pseudoprog_rate_results <- regr_fxd_pseudoprog_rate_results %>%  -->
<!--   filter(!Model %in% grep("ACUIT|LOGMAR|SNELL|pseudo|severity|rate|RW|variance|FA|letters", regr_fxd_pseudoprog_rate_results$Model, value = TRUE) ) %>%  -->
<!--   mutate(Model = Model %>% gsub("avg_MT", "MT", .)) -->

<!-- regr_fxd_pseudoprog_rate_results %>%  -->
<!--   filter(!grepl("clust", Term)) %>% -->
<!--   ggplot( aes(Estimate, -log10(q_value), color = Model))+ -->
<!--   geom_hline(yintercept = -log10(.05), linetype = 3, color = "black")+ -->
<!--   geom_vline(xintercept = 0, linetype = 2, color = "grey")+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--    geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)), -->
<!--                        size = 4, -->
<!--                        # nudge_y = 0.5,   -->
<!--                        show.legend = FALSE,  -->
<!--                        max.overlaps = 20)+ -->
<!--   ggtitle(paste0("Model: regr_fxd_pseudoprog_rate ~ SEXF + _phenotype_ + (1|patID)"))+ -->
<!--   theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/modeling/cluster_modeling/structure_regr_fxd_pseudoprog_rate_volcano.png"), height = 15, width = 20) -->

<!-- regr_fxd_pseudoprog_rate_results %>%  -->
<!--   filter(grepl("clust", Term)) %>%  -->
<!--   ggplot(aes(Estimate, -log10(q_value), color = Model))+ -->
<!--     geom_hline(yintercept = -log10(.05), linetype = 3, color = "black")+ -->
<!--     geom_vline(xintercept = 0, linetype = 2, color = "grey")+ -->
<!--     geom_point(show.legend = FALSE)+ -->
<!--      geom_label_repel(aes( -->
<!--        # label = ifelse( q_value < .05,  Model, NA)), -->
<!--        label = Model), -->
<!--                          size = 4, -->
<!--                          # nudge_y = 0.5,   -->
<!--                          show.legend = FALSE,  -->
<!--                          max.overlaps = 20)+ -->
<!--     ggtitle(paste0("Model: regr_fxd_pseudoprog_rate ~ SEXF + _phenotype_ + (1|patID)"))+ -->
<!--     theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/modeling/cluster_modeling/cluster_regr_fxd_pseudoprog_rate_volcano.png"), height = 10, width = 12) -->

<!-- ``` -->

### Stepwise rate
#### original pseudoprogression
```{r stepwise var rate function}
# ****** FUNCTION: calculate combination of each x with following x for each eye ******
stepwise_rate <- function(a_eye, df = NULL, var = NA, time_var = "year") {
    
  if(is.na(var)) stop("Response NA. Please provide variable for calculate steppwise rate")
  
  
  # Filter for eye of interest
  a_eye_diff_d <- df[df$patID_eye == a_eye, ]
  
  if(time_var == "year"){
    # make year var
    if(!"year" %in% names(a_eye_diff_d)){
      a_eye_diff_d <- a_eye_diff_d %>%
        mutate(year = round(month/12, 2))
    }
    
    # Create year to following year combinations
    year_combos <- data.frame(year = a_eye_diff_d$year, post_year = lead(a_eye_diff_d$year)) 
    
    # Create var to every other var combinations
    var_combos <- data.frame(var = a_eye_diff_d[[var]], post_var = lead(a_eye_diff_d[[var]])) 
      
    # Time pp combinations
    year_var_combos <- cbind(year_combos, var_combos)
    
    # make rate for each i to i+1: change in var over change in time>> postx-x/postyear-year
    year_var_combos$var_rate <- with(year_var_combos , (post_var-var)/(post_year-year))
    
    # Join back to dataset
    a_eye_diff_d <- suppressMessages(
      a_eye_diff_d %>% 
        left_join(year_var_combos, by = c("year")) 
      )
    
    # remove functional vars
    a_eye_diff_d$var <- NULL
    a_eye_diff_d$post_year <- NULL
    a_eye_diff_d$post_var <- NULL
    
    
    # rename "var" to var
    names(a_eye_diff_d) <- str_replace_all(names(a_eye_diff_d), "var", var)
  } 
    
  if(time_var == "month"){
    
    # Create month to following month combinations
    month_combos <- data.frame(month = a_eye_diff_d$month, post_month = lead(a_eye_diff_d$month)) 
    
    # Create var to every other var combinations
    var_combos <- data.frame(var = a_eye_diff_d[[var]], post_var = lead(a_eye_diff_d[[var]])) 
      
    # Time pp combinations
    month_var_combos <- cbind(month_combos, var_combos)
    
    # make rate for each i to i+1: change in var over change in time>> postx-x/postmonth-month
    month_var_combos$var_rate <- with(month_var_combos , (post_var-var)/(post_month-month))
    
    # Join back to dataset
    a_eye_diff_d <- suppressMessages(
      a_eye_diff_d %>% 
        left_join(month_var_combos, by = c("month")) 
      )
    
    # remove functional vars
    a_eye_diff_d$var <- NULL
    a_eye_diff_d$post_month <- NULL
    a_eye_diff_d$post_var <- NULL
    
    
    # rename "var" to var
    names(a_eye_diff_d) <- str_replace_all(names(a_eye_diff_d), "var", var)
  }
  
  return(a_eye_diff_d)
}
```

#####  model df
```{r}

var <- "pseudoprog"

# subset d with vars of interest 
orig_pp_model_df_stepwise <- pp_model_df %>% 
  select(month, patID_eye, patID, SEXF, all_of(c(predictor_vars, var)))

# scale vars of interest, minus response: FA2_rate
orig_pp_model_df_stepwise[, predictor_vars[predictor_vars!=var] ] <- scale(orig_pp_model_df_stepwise[, predictor_vars[predictor_vars!=var] ])


orig_pp_model_df_stepwise <- mclapply(unique(orig_pp_model_df_stepwise$patID_eye), function(x) stepwise_rate(x, df = orig_pp_model_df_stepwise,  var = var, time_var = "month"), mc.cores = detectCores()-1) %>% 
  do.call(rbind,.)

response = sprintf("%s_rate", var)

```

#####  modeling
```{r}

# model pseudoprog rate ----
predictor = predictors[1]
# loop through all but special vars and model response 
pseudoprog_rate_summaries <- sapply(predictors, function(predictor){
    a_formula <- as.formula(paste0( response, "~ SEXF + ", predictor, " + (1|patID)"  ))
    summ <- try({summary(lmer(a_formula, data = orig_pp_model_df_stepwise))})
    if(inherits(summ, "try-error")){
      return(NA)
    }else{
      return(summ)
    }
}, simplify = FALSE)

names(pseudoprog_rate_summaries)[is.na(pseudoprog_rate_summaries)] # "NOABNORMCF"  "NOABNORMFL"  "NOABNORMFAF" "NOABNORMOCT"
pseudoprog_rate_summaries <- pseudoprog_rate_summaries[!is.na(pseudoprog_rate_summaries)]
orig_pseudoprog_rate_results <- extract_results(pseudoprog_rate_summaries)  
  

# filter out VA stuff, rates, and pt derivatives
orig_pseudoprog_rate_results <- orig_pseudoprog_rate_results %>% 
  filter(!Model %in% grep("ACUIT|LOGMAR|SNELL|pseudo|severity|rate|RW|variance|FA|letters", orig_pseudoprog_rate_results$Model, value = TRUE) ) %>% 
  mutate(Model = Model %>% gsub("avg_MT", "MT", .))

orig_pseudoprog_rate_results %>% 
  filter(!grepl("clust", Term)) %>% 
ggplot( aes(Estimate, -log10(q_value), color = Model))+
    geom_hline(yintercept = -log10(.05), linetype = 3, color = "black")+
    geom_vline(xintercept = 0, linetype = 2, color = "grey")+
    geom_point(show.legend = FALSE)+
     geom_label_repel(aes(label = ifelse( q_value < .05,  Model, NA)),
                         size = 4,
                         # nudge_y = 0.5,  
                         show.legend = FALSE, 
                         max.overlaps = 20)+
    ggtitle(paste0("Model: stepwise_pseudoprog_rate ~ SEXF + _phenotype_ + (1|patID)"))+
    theme_bw()
ggsave(paste0(wd_path, "results/modeling/cluster_modeling/structure_orig_pseudoprog_stepwise_rate_volcano.png"), height = 15, width = 20)
  
orig_pseudoprog_rate_results %>% 
  filter(grepl("clust", Term)) %>% 
  ggplot(aes(Estimate, -log10(q_value), color = Model))+
    geom_hline(yintercept = -log10(.05), linetype = 3, color = "black")+
    geom_vline(xintercept = 0, linetype = 2, color = "grey")+
    geom_point(show.legend = FALSE)+
     geom_label_repel(aes(
       # label = ifelse( q_value < .05,  Model, NA)),
       label = Model),
                         size = 4,
                         # nudge_y = 0.5,  
                         show.legend = FALSE, 
                         max.overlaps = 20)+
    ggtitle(paste0("Model: stepwise_pseudoprog_rate ~ SEXF + _phenotype_ + (1|patID)"))+
    theme_bw()
ggsave(paste0(wd_path, "results/modeling/cluster_modeling/cluster_orig_pseudoprog_stepwise_rate_volcano.png"), height = 10, width = 12)

```

# _______________________
# correlate phenotypes to cluster 
```{r, eval = FALSE}

# what phenotypes to model?
phenotypes <- sd_gr_zero %>% 
  grep("stratus|cirrus|spectralis|clust_|month", ., value = TRUE, invert = TRUE)


cluster_d <- 
  d %>%
  select(patID, patID_eye, month, my_cluster, all_of(phenotypes)) %>%
  arrange(patID_eye,month) 

# subset for vars that have >.2 observed
## ratio of observed for each variable: sum of observed over num rows is greater than .2
cluster_d <- cluster_d[ , (colSums(!is.na(cluster_d))  / nrow(cluster_d)) > .2]

# which  vars to not scale
dont_scale_vars <- grep("patID|month|clust|SEXF", names(cluster_d), value = TRUE)

# scale all phenotypes
cluster_d <- 
  cluster_d %>% 
  ungroup() %>% 
  mutate(across(
    -all_of(dont_scale_vars),
                ~as.numeric(scale(.)) )
         )
# which vars are constant: sd == 0 or NA OR < 30 records where value > 0
var_sds <- colSds(cluster_d %>% select(where(~is.numeric(.))) %>% as.matrix(), na.rm = TRUE)
no_var_or_missing <- var_sds == 0 | is.na(var_sds)
few_records_grthn_zero <- colSums(cluster_d[, names(var_sds)] > 0, na.rm = TRUE) < 30

# variables that have no or missing variance OR have less than 30 records greater than 0
constant_vars <- c(
  names(few_records_grthn_zero)[few_records_grthn_zero],
  names(no_var_or_missing)[no_var_or_missing]
)
  
# remove vars that are constant
cluster_d <- cluster_d %>% 
  select(-all_of(constant_vars))
 
# # filter for eyes and patients with > 1 record
# next_cluster_d <- next_cluster_d %>% 
#   group_by(patID_eye) %>% 
#   filter(n()>1) %>% 
#   ungroup() %>% 
#   group_by(patID) %>% 
#   filter(n() > 1) %>% 
#   ungroup()

library(caret)
# make dummy variables for cluster presentation 
cluster_dummies <- predict(dummyVars(~my_cluster, cluster_d, sep = "_"), cluster_d) %>%
  as.data.frame() 

# add cluster dummy varaiables to cluster d
cluster_d <- cluster_d %>% 
  cbind(cluster_dummies)

# remove km_nn_clsuter
cluster_d$my_cluster <- NULL

# # function to make results df
# make.results.df <- function(mod_list){
#   
#   
#   summaries <- lapply(mod_list, summary)
#   
#   results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
#                          Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)
#     
#   # fill df with model values
#   for (model_name in names(summaries)) {
#     
#     model_coef <- coef(summaries[[model_name]])
#     
#     a_result_df <- data.frame(Model = model_name, 
#                            Term = rownames(model_coef), 
#                            Estimate = model_coef[, "Estimate"],
#                            Std.Error = model_coef[, "Std. Error"],
#                            z.value = model_coef[, "z value"],
#                            Pr = model_coef[, "Pr(>|z|)"]
#                            )
#     a_result_df$q_value <- p.adjust(a_result_df$Pr, method = "BH")
#     
#     results_df <- rbind(results_df, a_result_df)
#   }
#   
#   # filter out intercept
#   results_df <- results_df %>% 
#     filter(!Term %in% c("(Intercept)"))
#   
# }

# what vars are we predicting
responses <- grep("my_cluster", names(cluster_d), value  = TRUE)

predictors <- intersect(names(cluster_d), phenotypes)

memory_usage_summary()
rm("mod_list"); gc()

# LOOP through curr_clust and model for all predictors----
how.long(T)
curr_clust_summs <- list(); for(predictor in predictors){
  
  summ_list <- mclapply(responses, function(response){
  
    # naive formula
    log_form <-
      # paste0("next_my_cluster ~ (1|patID_eye)+ (1|month) + ", predictor) %>%
      # paste0("next_my_cluster ~  (1|month) + ", predictor) %>%
      paste0(response," ~  SEXF + ", predictor) %>%
      as.formula()
    
    # # random effects modeling for eye, patient, and month
    # me_log_form <- 
    #   # paste0(cluster_response," ~  ", predictor, " + SEXF + (1 | patID) + (1 | patID_eye) + (1 | patID_eye:month)") %>%
    #   # paste0(cluster_response," ~  ", predictor, " + SEXF + (1 | patID)  + (1 | patID_eye:month)") %>%
    #   # paste0(cluster_response," ~  ", predictor, " + SEXF + (1 | patID_eye)") %>%
    #   # paste0(cluster_response," ~  ", predictor, " + SEXF + (1 | patID)") %>%
    #   paste0(cluster_response," ~  ", predictor, " + SEXF + (1 | patID)+ (1 | patID_eye)") %>%
    #   # paste0(cluster_response," ~  ", predictor, " + SEXF + (1 | patID/patID_eye)") %>%
    #   as.formula()
    
    log_mod <- glm(log_form, data = cluster_d, family = binomial)
    
    return(summary(log_mod))
    
  }, mc.cores = detectCores()-1 )
  
  names(summ_list) <- paste(responses, predictor, sep = "~")
  
  curr_clust_summs <- append(curr_clust_summs, summ_list)
  
  cat('\n\n',response,'\n\n')
  
}
how.long()


# use new extract_results function
source('/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R')

# combine all results into one df
all_curr_clust_results <- extract_results(curr_clust_summs)

all_curr_clust_results <- all_curr_clust_results %>% 
  filter(!grepl("SEXF|Intercept", Term))

curr_clust_volcano <- all_curr_clust_results %>% 
  ggplot(aes(Estimate, -log10(q_value), color = Predictor))+
  geom_point(show.legend = FALSE) +
  facet_wrap(~Response, nrow = 3, scale = "free")+
  geom_hline(yintercept = .05, color = "grey", linetype = 2)+
  geom_vline(xintercept = 0, color = "grey", linetype = 2)+
  theme_bw()+
  ggtitle("Effect of feature on log-odds of being in cluster (current_cluster ~ SEXF + predictor)")+
  geom_label_repel(aes(label = ifelse(q_value<.05, Predictor, NA)),max.overlaps = 20, show.legend = FALSE, label.size = 1, size = 2)

ggsave(paste0(wd_path, "results/modeling/cluster_modeling/current_cluster_modeling.png"), curr_clust_volcano, height = 30, width = 60,  limitsize =FALSE)


```


#__________
# model what phenos push eye INTO each cluster
what phenos does an eye have that moves into 1-11?
```{r}

# what phenotypes to model?
phenotypes <- sd_gr_zero %>% 
  grep("month", ., value = TRUE, invert = TRUE)

# filter for eyes records where next or previous cluster not equal to current and 
# filter for where next cluster not NA
next_cluster_d <- 
  d %>%
  select(patID, patID_eye, month, SEXF,my_cluster, all_of(phenotypes)) %>%
  mutate(SEXF = SEXF*1) %>% 
  group_by(patID_eye) %>%
  arrange(patID_eye,month) %>% 
  filter(lead(my_cluster) != my_cluster | lag(my_cluster) != my_cluster) %>%
  mutate(next_my_cluster = lead(my_cluster)) %>% 
  filter(!is.na(next_my_cluster)) %>% 
  select(patID_eye, month, my_cluster, next_my_cluster, everything())

# subset for vars that have >.2 observed
## ratio of observed for each variable: sum of observed over num rows is greater than .2
((colSums(!is.na(next_cluster_d))  / nrow(next_cluster_d)) > .2 )%>%
  {!.} %>% 
  which(., useNames = T)
# GRADERID  NOABNORMCF  NOABNORMFL NOABNORMFAF NOABNORMOCT      DMYRDX    SMKAGEST   SMKAGESTP  SMKCIGSDAY       RW011 RW021       RW031       RW041       RW051       RW061      VAAMTH  chew_grade 

next_cluster_d <- next_cluster_d[ , (colSums(!is.na(next_cluster_d))  / nrow(next_cluster_d)) > .2]

# which  vars to not scale
dont_scale_vars <- grep("patID|month|clust|next_is|SEXF", names(next_cluster_d), value = TRUE)

# scale all phenotypes
next_cluster_d <- next_cluster_d %>% 
  ungroup() %>% 
  mutate(across(
    -all_of(dont_scale_vars),
                ~as.numeric(scale(.)) )
         )
# which vars are constant: sd == 0 or NA OR < 30 records where value > 0
var_sds <- colSds(next_cluster_d %>% select(where(~is.numeric(.))) %>% as.matrix(), na.rm = TRUE)
no_var_or_missing <- var_sds == 0 | is.na(var_sds)
few_records_grthn_zero <- colSums(next_cluster_d[, names(var_sds)] > 0, na.rm = TRUE) < 30

# variables that have no or missing variance OR have less than 30 records greater than 0
constant_vars <- c(
  names(few_records_grthn_zero)[few_records_grthn_zero],
  names(no_var_or_missing)[no_var_or_missing]
)
  
# remove vars that are constant
next_cluster_d <- next_cluster_d %>% 
  select(-all_of(constant_vars))
 
# # filter for eyes and patients with > 1 record
# next_cluster_d <- next_cluster_d %>% 
#   group_by(patID_eye) %>% 
#   filter(n()>1) %>% 
#   ungroup() %>% 
#   group_by(patID) %>% 
#   filter(n() > 1) %>% 
#   ungroup()

# make variable for each next cluster, e.g. next_cluster_4[1,0,1,0,1] for every next cluster
# a_next_cluster = unique(next_cluster_d$next_my_cluster)[1]
for(a_next_cluster in unique(next_cluster_d$next_my_cluster)){
  
  next_cluster_name <- paste0("next_is_", a_next_cluster)
  
  next_cluster_d[[next_cluster_name]] <- as.numeric(next_cluster_d$next_my_cluster == a_next_cluster)
  
}


# what vars are we predicting
responses <- grep("next_is", names(next_cluster_d), value  = TRUE)

predictors <- intersect(names(next_cluster_d), phenotypes)

# ****** Function to make results df ******
make.results.df <- function(mod_list){
  summaries <- lapply(mod_list, summary)
  results_df <- data.frame(Model = character(),  Term = character(),Estimate = numeric(),
                         Std.Error = numeric(),t.value = numeric(),Pr = numeric(),stringsAsFactors = FALSE)
  # fill df with model values
  for (model_name in names(summaries)) {
    model_coef <- coef(summaries[[model_name]])
    a_result_df <- data.frame(Model = model_name, 
                           Term = rownames(model_coef), 
                           Estimate = model_coef[, "Estimate"],
                           Std.Error = model_coef[, "Std. Error"],
                           z.value = model_coef[, "z value"],
                           Pr = model_coef[, "Pr(>|z|)"]
                           )
    a_result_df$q_value <- p.adjust(a_result_df$Pr, method = "BH")
    results_df <- rbind(results_df, a_result_df)
  }
  # filter out intercept
  results_df <- results_df %>% 
    filter(!Term %in% c("(Intercept)"))
  return(results_df)
}


# LOOP through next_is_xx and model for all predictors----
next_clust_results_dfs <- mclapply(responses, function(next_cluster_response){
  
  # loop through phenotypes in next_cluster_d and model next cluster
  mod_list <- list(); for(predictor in predictors){
      print(paste("modeling, ", predictor))
      # naive formula
      log_form <-
        # paste0("next_my_cluster ~ (1|patID_eye)+ (1|month) + ", predictor) %>%
        # paste0("next_my_cluster ~  (1|month) + ", predictor) %>%
        paste0(next_cluster_response," ~  SEXF + ", predictor) %>%
        as.formula()
      
      # random effects modeling for eye, patient, and month
      me_log_form <- 
        # paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID) + (1 | patID_eye) + (1 | patID_eye:month)") %>%
        # paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID)  + (1 | patID_eye:month)") %>%
        # paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID_eye)") %>%
        # paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID)") %>%
        paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID)+ (1 | patID_eye)") %>%
        # paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID/patID_eye)") %>%
        as.formula()
      
      
      # control_settings <- glmerControl(optCtrl = list(maxfun = 1e6))
      log_mod <- glm(log_form, data = next_cluster_d, family = binomial)
      # log_mod <- glmer(me_log_form, data = next_cluster_d, family = binomial)
      
      mod_list[[predictor]] <- log_mod
      
      print(paste(predictor, ":", paste(as.character(me_log_form)[c(2,1, 3)], collapse = " ") ))
      print("")
  }
  return(make.results.df(mod_list))
}, mc.cores = detectCores()-1)

names(next_clust_results_dfs) <- responses

# filter for models that worked
next_clust_results_dfs <- next_clust_results_dfs[sapply(next_clust_results_dfs,function(x)class(x) =="data.frame")]

# add which response to each df
for(df_name in names(next_clust_results_dfs)){
  next_is <-  str_split(df_name, "_", simplify = T)[3]
  next_clust_results_dfs[[df_name]]<- next_clust_results_dfs[[df_name]] %>% cbind(next_is)
}

# combine all results into one df
all_next_cluster_results <- purrr::reduce(next_clust_results_dfs, rbind)
all_next_cluster_results$q_value <- p.adjust(all_next_cluster_results$Pr)

write.csv(all_next_cluster_results, paste0(wd_path, "processed_data/next_cluster_modeling_results_d.csv"))

# all_next_cluster_results %>% 
#   select(Model, Estimate, q_value, next_is) %>% 
#   filter(!grepl("SEX", Model)) %>% 
#   # group_by(next_is) %>%
#   # slice_min(q_value, n = 50) %>%
#   # arrange(next_is, q_value) %>% 
#   filter(abs(Estimate)<=10) %>% 
#   ggplot(aes(Estimate, -log10(q_value), color = Model))+
#   geom_point(show.legend = FALSE) +
#   geom_label_repel(aes(label = ifelse(q_value<.05, Model, NA)),max.overlaps = 20, show.legend = FALSE,force = 2)+
#   facet_wrap(~factor(next_is, levels = 1:16), nrow = 2, scale = "free")+
#   geom_hline(yintercept = -log10(.05), color = "grey", linetype = 2)+
#   geom_vline(xintercept = 0, color = "grey", linetype = 2)+
#   theme_bw()+
#   ggtitle("Effect of feature on log-odds of entering cluster (next_is_n ~ SEXF + predictor)")
# 
# ggsave(paste0(wd_path, "results/modeling/cluster_modeling/next_cluster_modeling.png"), height = 10, width = 40)


all_next_cluster_results %>% 
  select(Model,Term, Estimate, Pr, next_is) %>% 
  filter(!grepl("SEX", Model)) %>% 
  filter(grepl("clust", Term)) %>% 
  ggplot(aes(Estimate, -log10(Pr), color = Term))+
    geom_point(show.legend = FALSE) +
    geom_label_repel(aes(label =  Term),max.overlaps = 20, show.legend = FALSE,force = 2)+
    facet_wrap(~factor(next_is, levels = 1:16), nrow = 2, scale = "free")+
    geom_hline(yintercept = -log10(.05), color = "grey", linetype = 2)+
    geom_vline(xintercept = 0, color = "grey", linetype = 2)+
    theme_bw()+
    ggtitle("Effect of feature on log-odds of entering cluster (next_is_n ~ SEXF + predictor)")
ggsave(paste0(wd_path, "results/modeling/cluster_modeling/next_cluster_modeling_clust_only.png"), height = 8, width = 30)


```

### does t2d predict movement INTO cluster?
```{r, eval = FALSE}

# model d with t2d
t2d_d <- next_cluster_d %>% 
  left_join(colData(sce) %>% as.data.frame() %>% select(patID,  t2d) %>% unique(), relationship = "many-to-many")

t2d_mod_list <- list(); for (next_cluster_response in responses){
  
  log_form <-
    # paste0("next_my_cluster ~ (1|patID_eye)+ (1|month) + ", predictor) %>%
    # paste0("next_my_cluster ~  (1|month) + ", predictor) %>%
    paste0(next_cluster_response," ~  SEXF + t2d") %>%
    as.formula()
  
  log_mod <- try({glm(formula = log_form, data = t2d_d, family = binomial)})
  
  if(inherits(log_mod, "try-error")){
    log_mod <- NA
  }
  
  t2d_mod_list[[next_cluster_response]] <- log_mod
      
}
  
t2d_results <- make.results.df(t2d_mod_list) %>% 
  filter(!Term %in%c("SEXF"))

t2d_results %>% 
  ggplot(aes(Estimate, -log10(q_value), color = Model))+
  geom_point()+
  geom_hline(yintercept = -log10(.05), linetype = 3, color = "grey", size = 1)+
  geom_vline(xintercept = 0, linetype = 3, color = "grey", size = 1)+
  geom_label(aes(label = ifelse(q_value<.05, Model, NA)), show.legend = FALSE)+
  ggtitle("Correlating next cluster to T2D diagnosis")

```

# model what phenos push eye OUT of each cluster
what phenos does an eye have that moves into 1-11?
```{r, eval = FALSE}

# filter for eyes records where next or previous cluster not equal to current and 
# filter for where next cluster not NA
next_cluster_d <- 
  d %>%
  select(patID, patID_eye, month, my_cluster, all_of(phenotypes)) %>%
  group_by(patID_eye) %>%
  # filter(lead(my_cluster) != my_cluster | lag(my_cluster) != my_cluster) %>% 
  arrange(patID_eye,month) %>% 
  mutate(next_my_cluster = lead(my_cluster)) %>% 
  filter(!is.na(next_my_cluster)) %>% 
  select(patID_eye, month, my_cluster, next_my_cluster, everything())

# subset for vars that have >.2 observed
## ratio of observed for each variable: sum of observed over num rows is greater than .2
next_cluster_d <- next_cluster_d[ , (colSums(!is.na(next_cluster_d))  / nrow(next_cluster_d)) > .2]

# which  vars to not scale
dont_scale_vars <- grep("patID|month|clust|next_is|SEXF", names(next_cluster_d), value = TRUE)

# scale all phenotypes
next_cluster_d <- next_cluster_d %>% 
  ungroup() %>% 
  mutate(across(
    -all_of(dont_scale_vars),
                ~as.numeric(scale(.)) )
         )
# which vars are constant: sd == 0 or NA OR < 30 records where value > 0
var_sds <- colSds(next_cluster_d %>% select(where(~is.numeric(.))) %>% as.matrix(), na.rm = TRUE)
no_var_or_missing <- var_sds == 0 | is.na(var_sds)
few_records_grthn_zero <- colSums(next_cluster_d[, names(var_sds)] > 0, na.rm = TRUE) < 30

# variables that have no or missing variance OR have less than 30 records greater than 0
constant_vars <- c(
  names(few_records_grthn_zero)[few_records_grthn_zero],
  names(no_var_or_missing)[no_var_or_missing]
)
  
# remove vars that are constant
next_cluster_d <- next_cluster_d %>% 
  select(-all_of(constant_vars))
 
# # filter for eyes and patients with > 1 record
# next_cluster_d <- next_cluster_d %>% 
#   group_by(patID_eye) %>% 
#   filter(n()>1) %>% 
#   ungroup() %>% 
#   group_by(patID) %>% 
#   filter(n() > 1) %>% 
#   ungroup()

# make variable for each next cluster, e.g. next_cluster_4[1,0,1,0,1] for every next cluster
# a_next_cluster = unique(next_cluster_d$next_my_cluster)[1]
for(a_next_cluster in unique(next_cluster_d$next_my_cluster)){
  
  next_cluster_name <- paste0("next_is_", a_next_cluster)
  
  next_cluster_d[[next_cluster_name]] <- as.numeric(next_cluster_d$next_my_cluster == a_next_cluster)
  
}

# what vars are we predicting
responses <- grep("next_is", names(next_cluster_d), value  = TRUE)

predictors <- intersect(names(next_cluster_d), phenotypes)

###
# next_cluster_response <- "next_is_1"
# predictor = "RAV04"
###

# LOOP through next_is_xx and model for all predictors----
results_dfs <- parallel::mclapply(responses, function(next_cluster_response){
  
  # loop through phenotypes in next_cluster_d and model next cluster
  mod_list <- list(); for(predictor in predictors){
      print(paste("modeling, ", predictor))
      # naive formula
      log_form <-
        # paste0("next_my_cluster ~ (1|patID_eye)+ (1|month) + ", predictor) %>%
        # paste0("next_my_cluster ~  (1|month) + ", predictor) %>%
        paste0(next_cluster_response," ~  SEXF + ", predictor) %>%
        as.formula()
      
      # random effects modeling for eye, patient, and month
      me_log_form <- 
        # paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID) + (1 | patID_eye) + (1 | patID_eye:month)") %>%
        # paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID)  + (1 | patID_eye:month)") %>%
        # paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID_eye)") %>%
        # paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID)") %>%
        paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID)+ (1 | patID_eye)") %>%
        # paste0(next_cluster_response," ~  ", predictor, " + SEXF + (1 | patID/patID_eye)") %>%
        as.formula()
      
      
      # control_settings <- glmerControl(optCtrl = list(maxfun = 1e6))
      log_mod <- glm(log_form, data = next_cluster_d, family = binomial)
      # log_mod <- glmer(me_log_form, data = next_cluster_d, family = binomial)
      # summary(log_mod)
      
      mod_list[[predictor]] <- log_mod
      
      print(paste(predictor, ":", paste(as.character(me_log_form)[c(2,1, 3)], collapse = " ") ))
      print("")
      
  }
  
  return(make.results.df(mod_list))
  
}, mc.cores = parallel::detectCores()-1)

# add names to list
names(results_dfs) <- responses

# add which response to each df
for(df_name in names(results_dfs)){
  
  next_is <-  str_split(df_name, "_", simplify = T)[3]
  results_dfs[[df_name]]<- results_dfs[[df_name]] %>% cbind(next_is)
  
  
}

# combine all results into one df
all_next_cluster_results <- purrr::reduce(results_dfs, rbind)
all_next_cluster_results$q_value <- p.adjust(all_next_cluster_results$Pr)

write.csv(all_next_cluster_results, paste0(wd_path, "processed_data/next_cluster_modeling_results_d.csv"))

all_next_cluster_results %>% 
  select(Model, Estimate, q_value, next_is) %>% 
  filter(-log10(q_value)>0) %>% 
  group_by(next_is) %>% 
  slice_min(q_value, n = 50) %>%
  arrange(next_is, q_value) %>% 
  ggplot(aes(Estimate, -log10(q_value), color = Model))+
  geom_point(show.legend = FALSE) +
  geom_label_repel(aes(label = ifelse(q_value<.05, Model, NA)),max.overlaps = 20, show.legend = FALSE)+
  facet_wrap(~factor(next_is, levels = 1:11), nrow = 2, scale = "free")+
  geom_hline(yintercept = .05, color = "grey", linetype = 2)+
  geom_vline(xintercept = 0, color = "grey", linetype = 2)+
  theme_bw()+
  ggtitle("Effect of feature on log-odds of entering cluster. next_is_n ~ SEXF + predictor")

ggsave(paste0(wd_path, "results/modeling/cluster_modeling/next_cluster_modeling.png"), height = 20, width = 40)

```


#=========================
# SAVE DATA
```{r}

# add data back to sce ----

## coldata
# colData(sce) %>% dim()
new_colData <- colData(sce) %>% 
  as.data.frame()

new_colData <- new_colData %>% 
  left_join(
    d %>% select(sample, pseudoprog_rate)
    )
# new_colData %>% dim()

colData(sce) <- colData(sce) %>% 
  cbind(new_colData %>% select(pseudoprog_rate))


# save sce ----
saveHDF5SummarizedExperiment(sce, paste0(wd_path, "processed_data/sce004.h5"), replace = T)

notify.me(paste("11_clusterModeling.Rmd complete", how.long()))
```


```{r check dims}
check_dims()
# dims : 12525 157
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ added dichotomous clusters and pseudoprog rate, and t2d column
```
