---
title: "12_Cluster_exploration"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---

Predicting routes from eye movement through clusters over real time

# options
```{r}
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()
```


# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

pacman::p_load(tidyr,dplyr,reshape2,purrr,GGally,tidyverse,ggplot2,gridExtra,ComplexHeatmap,viridis,zoo,openxlsx,plotly,mgcv,zoo,tools,mice,ggrepel,networkD3,htmltools,Metrics,psych,car,caret,stats,lme4,lmerTest,ape,SingleCellExperiment,scater,velociraptor,igraph,Rtsne,randomForest, HDF5Array, txtplot, parallel, patchwork)

rm(list = ls())
gc()


```

# functions
```{r}

source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```

# Load Data
info and d from sce
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

# load experiment
sce <- HDF5Array::loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/sce004.h5"))

```

### function to join assay and colData as df
```{r}
# ****** Function: join assay and colData as df ****
make_d <- function(assay = "pmm_imp_of_scaled"){
  suppressMessages(
    assays(sce)[[assay]] %>% 
      t() %>% 
      as.data.frame() %>% 
      rownames_to_column("sample") %>% 
      left_join(
        colData(sce) %>% 
          data.frame() 
      )
  )
}

```


# make new directorys for plots
```{r}

a_path <- paste0(wd_path, "results/clustering/12_clustering")
if(!dir.exists(a_path)) {
  dir.create(a_path, recursive = TRUE)
  
}

```

# Mung data

```{r look at dims}
check_dims <- function(){
  
  D <- as.data.frame(colData(sce)) %>% 
  left_join(
    as.data.frame(t(assays(sce)$scaled_measure_vars)) %>% 
      rownames_to_column("sample")
  ) 
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()
# dims : 12525 157
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ no change
```


## simplify clusters
```{r}

plotTSNE(sce, color_by = "my_cluster", text_by = "my_cluster")+
  geom_line(data = metadata(sce)$trajectory_data$tsne_line_data, aes(TSNE1,TSNE2, group = edge))

# make simple cluster assingment: first clusters after main division
sce$simple_cluster <- ifelse(sce$my_cluster %in% c(10,11,5,8), 111, 
                             ifelse(sce$my_cluster %in% c(13,7), 999 , sce$my_cluster)) %>% 
  factor()

table(sce$simple_cluster)

plotReducedDim(sce, "TSNE", color_by = "simple_cluster", text_by = "my_cluster")+
  geom_line(data = metadata(sce)$trajectory_data$tsne_line_data, aes(TSNE1,TSNE2, group = edge))

```

#__________________________


# quick side analyses
## pseudoprog quintile plots
```{r}

# first visit pseudoprog values
first_pseudotimes <- make_d() %>% 
  select(patID_eye, month, pseudoprog) %>% 
  arrange(month) %>% 
  group_by(patID_eye) %>% 
  filter(row_number() == 1) %>% 
  pull(pseudoprog)

# quintile first visit pseudotimes
pt_quintiles <- quantile(first_pseudotimes, seq(0,1,.2), na.rm = TRUE)

pt_quintile_d <- make_d() %>% 
  select(patID_eye, month, pseudoprog) %>% 
  group_by(patID_eye) %>% 
  filter(row_number() == 1) %>% 
  mutate(pseudotime_quintile = 
           cut(pseudoprog, breaks = pt_quintiles, include.lowest = TRUE, labels = as.character(1:5) )
         ) %>% 
  select(patID_eye, pseudotime_quintile) %>% 
  unique() 
  # mutate(across(ends_with("quintile"), ~factor(., levels = as.character(seq(5,1, -1)) )))

d <- make_d() %>% 
  left_join(pt_quintile_d)


 d %>% 
  select(patID_eye, month, pseudoprog, pseudotime_quintile) %>% 
  na.omit %>% 
  ggplot(aes(month, pseudoprog))+
  geom_smooth(method = "lm")+
  facet_wrap(~pseudotime_quintile, ncol = 5) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey")+
  ggtitle("Pseudotime over time stratified by baseline pseudotime quintiles")+
   theme_bw()

ggsave(paste0(wd_path, "results/clustering/12_clustering/pseudotime_quintile.png") , height = 4, width = 20)

 ggsave(paste0(wd_path, "results/publication_figures/pseudotime_quintile.pdf") , height = 4, width = 17)


```


## graph of eye_paths
```{r}

paths <- d %>% 
  select(patID_eye, month, pseudoprog, my_cluster) %>% 
  # arrange(pseudoprog) %>%
  arrange(month) %>%
  group_by(patID_eye) %>% 
  na.omit() %>% 
  # filter(lead(as.numeric(my_cluster)) > as.numeric(my_cluster)) %>% 
  mutate(source = my_cluster, 
         destination = lead(my_cluster)) %>% 
  ungroup() %>% 
  filter(!is.na(destination) & source != destination) 
  # filter(as.numeric(destination) > as.numeric(source))

# paths %>% 
#     arrange(patID_eye) %>% 
#     group_by(patID_eye) %>% 
#     filter(n_distinct(month)> 1) %>% 
#   na.omit()
# 
# 
# 
# for(eye in unique(paths$patID_eye)){
#   
#   
#   
# }

edges <- paths %>% 
  select(source, destination) %>% 
  na.omit() %>% 
  as.matrix()

path_graph <- graph_from_edgelist(edges)

path_graph <- set_vertex_attr(path_graph, index = V(path_graph), name = "color", value =  cluster_cols[V(path_graph)])

plot.igraph(path_graph, 
     edge.arrow.size = 1, 
     vertex.size = 10, 
     vertex.label.cex = 1,
     # vertex.color = "white",
     edge.color = "black",
     main = "Eye movement from cluster to cluster",
     col.main = "black"
     )

png(paste0(wd_path, "results/clustering/eye_movement_graph.png"), height = 2000, width = 2000)
par(bg = "black",  col.main = "white")
set.seed(12433)
plot.igraph(path_graph, 
     edge.arrow.size = 1, 
     vertex.size = 10, 
     vertex.label.cex = 5,
     layout = layout_with_fr(path_graph),
     # vertex.label.color = "blue",   
     # vertex.color = "white",
     edge.color = "white",
     main = "Eye movement from cluster to cluster",
     )
dev.off()
par(bg  = "white", col.main = "black")
  
# paths %>% 
#   filter((source ==1 & destination == 2) | (source ==2 & destination == 3)) %>% 
#   select(source, destination) %>% 
#   unique()

```

### edge counts
```{r}

char_edges <- apply(edges, 1, function(row) paste(row, collapse = "-"))

png(paste0(wd_path, "results/clustering/12_clustering/eye_movement_graph_edge_counts.png"), height = 1000, width = 1000)
barplot(
  sort(table(char_edges)),
  horiz = T,
  las = T,
  main = "Edge counts"
  )
dev.off()

```


## T2D divide clusters?
```{r, eval = FALSE}

diabetes_clusters <- cbind(cluster = sce$my_cluster,simple_cluster = sce$simple_cluster, diagnosis = sce$DM, type =  sce$DMTYPE) %>% 
  as.data.frame() %>% 
  filter(diagnosis == "Yes" & type == "Type II") %>%
  mutate()
head(diabetes_clusters)

# Prepare the data for chi-squared test
diabetes_cluster_tab <- diabetes_clusters %>%
  group_by(simple_cluster) %>%
  summarise(simple_cluster_freq = sum(cluster_count)) %>%
  ungroup()

# Create a contingency table
contingency_table <- diabetes_cluster_tab %>%
  column_to_rownames("simple_cluster") %>%
  as.matrix()

# divide cluster freqs by cluster sizes
## size of eaach cluster
cluster_sizes <- colData(sce)[ , c("my_cluster", "simple_cluster")] %>% 
  table() %>% 
  as.data.frame()
names(cluster_sizes) <- c("my_cluster",'simple_cluster', "size")

simple_cluster_sizes <- cluster_sizes %>% 
  group_by(simple_cluster) %>% 
  summarise(simple_cluster_size = sum(size)) %>%
  ungroup() %>% 
  column_to_rownames("simple_cluster") %>% 
  as.matrix()

contingency_table_corrected <- contingency_table/simple_cluster_sizes

chisq.test(contingency_table)

# proportion test; count of t2d in each cluster versus size of eacah cluster
prop.test(diabetes_cluster_tab$simple_cluster_freq, simple_cluster_sizes)




```


## cluster data for each eye
```{r}

# get tsne with info about each eye
tsne <- reducedDim(sce, "TSNE") %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  left_join(
    colData(sce)[, c("sample", "patID_eye", "patID", "month", "pseudoprog", "my_cluster", "simple_cluster")] %>% 
      as.data.frame() 
  ) %>% 
  mutate(year = round(month, 2))

```
#_________________
# Describe routes
## define routes
```{r}
## actual tree structure ----
# XXX

# simplified tree structure ----
A <- c(3,6,4,1,111)
A.termini <- c(111)
B <- c(3,6,4,9,2,999)
B.termini <- c(999)
A.unique <- setdiff(A, B)
B.unique <- setdiff(B, A)

# tree edges
edges <- matrix(as.character(c(
  3,6,
  6,4,
  4,12,
  4,9,
  9,2,
  2,999,
  4,1,
  1,111
)), ncol = 2, byrow = TRUE)

# make tree
tree <- graph_from_edgelist(edges, directed = TRUE)
plot(tree, layout = layout_as_tree(tree,root = "3"))

#  distances from the root node (node 1)
distances_from_root <- distances(tree, v = V(tree)["3"], to = V(tree), mode = "out")
```


## extract path info for each eye
```{r}
# extract each eye's cluster lineage/path ----
eye_paths <- mclapply(unique(tsne$patID_eye), function(an_eye){
  
  tsne[ tsne$patID_eye == an_eye , ] %>% 
    arrange(month) %>% 
    # pull(my_cluster)
    pull(simple_cluster) %>% 
    as.character()# simplified terminal clusters to 333 from 3 and 444 from 4
    
}, mc.cores = detectCores()-1)

# make paths same length; fill ends with NA
max.length <- max(sapply(eye_paths, length))
eye_paths_same_length <- eye_paths %>% 
  sapply(., function(a.path){
    return(c(a.path, rep(NA, max.length - length(a.path))))
  })

eye_paths_m <- eye_paths_same_length %>% 
  t() 

# add meaninful names to path m
colnames(eye_paths_m) <- paste(rep("step", max.length), 1:max.length, sep = "")
rownames(eye_paths_m) <- unique(tsne$patID_eye)

# extract features about each path----
extract.path.features <- function(a.path){
  
  a.path <- na.omit(a.path)
  char_path <- paste(a.path, collapse  = "-")
  
  # count of nodes affilieated with each path's distinct nodes
  count.in.A <- sum(a.path %in% A.unique)
  count.in.B <- sum(a.path %in% B.unique)
  
  # end node
  terminus <- a.path[length(a.path)]
  terminus.route <- ifelse(terminus %in% A.termini, "a", 
                          ifelse(terminus %in% B.termini, "b", "none"))
  
  # count termini affiliated with each path
  count.A.termini <- sum(a.path %in% A.termini)
  count.B.termini <- sum(a.path %in% B.termini)
  

  # how deep does this path go (end-ness), excluding 12
  depths <- distances_from_root [1 , match(a.path[a.path!=12], as.numeric(colnames(distances_from_root)))] %>% na.omit()
  total.depth <- sum(depths)
  
  # how long is longest run
  path.rle <- rle(as.numeric(a.path))
  longest.run <- path.rle$lengths %>% max()
  
  # which node has longest run, if two grab first
  longest.run.node <- unique((path.rle$values[path.rle$lengths == longest.run]))[1]
  
  # splitting the path in two, how many of each termini sets on left/right side
  halfway <- length(a.path)/2
  
  # left side counts of termini
  count.B.termini.left <- sum(a.path[1:halfway] %in% B.termini)
  count.A.termini.left <- sum(a.path[1:halfway] %in% A.termini)
  
  # right side termini
  count.A.termini.right <-  sum(a.path[halfway+1 : length(a.path)] %in% A.termini)
  count.B.termini.right <-  sum(a.path[halfway+1 : length(a.path)] %in% B.termini)
  
  # what's the next non-first node
  first.non.root.node = ifelse(length(a.path)>1, a.path[a.path != a.path[1]][1], NA)
  
  # path length   
  path.length = length(a.path)
  
  # dataframe to return
  features <- data.frame(
      path = char_path, 
      path.length = path.length,
      n.distinct.nodes = n_distinct(a.path),
      root.node = a.path[1],
      first.non.root.node = first.non.root.node,
      last.node = tail(a.path, 1),
      longest.run = longest.run, 
      longest.run.node = longest.run.node, 
      count.in.A= count.in.A,
      count.in.B= count.in.B,
      prop.in.A = count.in.A/ length(a.path),
      prop.in.B = count.in.B/ length(a.path),
      prop.A.to.B = count.in.A/count.in.B,
      prop.B.to.A = count.in.B/count.in.A,
      terminus = terminus, 
      terminus.route = terminus.route,
      count.A.termini = count.A.termini,
      count.B.termini = count.B.termini,
      count.B.termini.left =count.B.termini.left ,
      count.A.termini.left =count.A.termini.left ,
      count.A.termini.right=count.A.termini.right,
      count.B.termini.right=count.B.termini.right,
      total.depth = total.depth,
      norm.total.depth = total.depth / length(a.path)
      # inv.total.depth = 1/total.depth,
      # inv.norm.total.depth = 1/(total.depth / length(a.path))
    )
    
    # make step fields
    a.path_steps <- c(a.path, rep(a.path[path.length], 10-path.length))
    names(a.path_steps) <- paste0("step", 1:10)
    a.path_steps <- a.path_steps %>% 
      t() %>%
      as.data.frame()
    
    # add steps to features
    features <- features %>% cbind(a.path_steps)
  
  return(features)
}


# extract features from path matrix
eye_paths_features <- mclapply(seq(nrow(eye_paths_m)), function(i) extract.path.features(eye_paths_m[i, ]), mc.cores = detectCores()-1)

# reduce list to data.frame
eye_paths_features_d <- parr_reduce(eye_paths_features, rbind)
eye_paths_features_d$patID_eye <- rownames(eye_paths_m)
rownames(eye_paths_features_d) <- rownames(eye_paths_m)

# clean for infinities
eye_paths_features_d <-eye_paths_features_d %>% 
  mutate_all(~ifelse(is.infinite(.), NA, .)) %>% 
  group_by(path) %>% 
  mutate(count = n()) %>% 
  ungroup() %>% 
  data.frame()

# # make step columns
# path_columns <- eye_paths_features_d$path %>% 
#   str_split(., "-", simplify = T) %>% 
#   as.data.frame() %>% 
#   mutate_all(~ifelse(. == "", NA, .)) %>% 
#   mutate_all(~as.numeric(.))
#   
# 
# colnames(path_columns) <- colnames(eye_paths_m)
# 
# 
# eye_paths_features_d <- cbind(eye_paths_features_d, path_columns)

rownames(eye_paths_features_d) <- eye_paths_features_d$patID_eye

# filter for route likely eyes: 
## ones that don't have only one visit in unbifurcated clusters or outlier clusters (7); xxxand have more than one cluster affiliationxxx
select_eyes <- tsne %>% 
  select(patID_eye, my_cluster) %>% 
  group_by(patID_eye) %>% 
  filter( 
    !(n() == 1 & my_cluster %in% c(3,6,4,12)) 
    ) %>% 
  arrange(patID_eye) %>% 
  pull(patID_eye) %>% 
  unique()
  
eye_paths_features_d <- eye_paths_features_d %>% 
  filter(patID_eye %in% select_eyes)

eye_paths_features_d %>% 
  select(matches("step")) %>% 
  unlist() %>% 
  table

```

### add PC1/2 to path data
```{r}

pcs <- reducedDim(sce, "PCA") %>% 
  data.frame() %>% 
  rownames_to_column("sample") %>% 
  left_join(make_d()) %>% 
  select(matches("sample|patID_eye|month|cluster|PC1$|PC2$"))

# for each eye, average PC1,PC2,PC1*PC2
pcs <- pcs %>% 
  group_by(patID_eye) %>% 
  mutate(PC1_mean = mean(PC1)) %>% 
  mutate(PC2_mean = mean(PC2)) %>% 
  mutate(PC1x2_mean = mean(PC2*PC1)) %>% 
  ungroup()
  
# add to features d
eye_paths_features_d <- eye_paths_features_d %>% 
  left_join(pcs %>% select(patID_eye, matches("PC[12].*_mean") ) %>% distinct(), relationship = "one-to-many"  )

# try transforming interaction
eye_paths_features_d$PC1x2_mean_transformed <- eye_paths_features_d$PC1x2_mean %>% 
    {abs(.)^(1/2)*sign(.)}

pcs %>%
  ggplot(aes(PC1,PC2))+
  # scale_color_manual(values = cluster_cols)+
  geom_point(aes(color = (PC1*PC2-1) %>% {abs(.)^(1/2)*sign(.)} ), alpha = .5)+
  scale_color_viridis_c()+
  geom_line(data = metadata(sce)$trajectory_data$pca_line_data  , aes(group = edge))+
  theme_bw()+
  theme(aspect.ratio = 1)

```

#________________
# Predict route from path data
### label paths
#### sample eye paths to label
```{r}

# find perfect route followers (follows A or B; allowing runs)
# ****** function to see if path == or ~= A or B
is.good.route <- function(path.seq){
  path.seq <- unique(na.omit(path.seq))
  long_enough <- length(path.seq) >= 4
  end_2_nodes <- all(c(4, A.termini) %in% path.seq) | all(c(4, B.termini) %in% path.seq)
  perfect_path <- identical(path.seq, A) | identical(path.seq, B)
  if (long_enough & perfect_path) {
    return(TRUE)
  } else if (!long_enough & end_2_nodes) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}
# c(3,6,4,999) %>%
# c(3,6,4,111) %>%
# c(4,999) %>%
# c(4,111) %>%
c(1) %>%
  is.good.route(.)

# eyes is good rouute?
route_followers <- eye_paths_m %>% 
  apply(., 1, is.good.route)

good_route_followers <- names(route_followers)[route_followers]

# good route followers feaatures
good_route_followers_features <- eye_paths_features_d %>% 
  select(patID_eye, path) %>% 
  filter(patID_eye %in% good_route_followers) %>% 
  mutate(route = NA)
  # mutate(route = ifelse(grepl("3|333", path), "a", 
  #                       ifelse(grepl("4|444", path), "b", NA))
  #        )

# sample some eyes to label
set.seed(1937) # 1937 has very good balance of nodes
random_sample <- eye_paths_features_d %>% 
  slice_sample(n=300)

# look at proportions of nodes in sample
random_sample$path %>%  str_split(., "-") %>% unlist() %>% table

# Look at distribution of path lengths; ensure same prob for each path length?
txtplot::txtdensity(eye_paths_features_d$path.length, height = 10, '·')
#     +-·----+---------+----------+---------+------+
# 0.8 + ··                                         +
# 0.6 +  ··                                        +
#     |   ·                                        |
# 0.4 +   ·                                        +
#     |    ·                                       |
# 0.2 +    ··                                      +
#     |     ····      ························     |
#   0 +------+-········+----------+---------+-····-+
#            2         4          6         8       

```

#### label paths

```{r, eval=FALSE}

# make sample with representative number of of path lengths
#+ for each path length in eye_paths_featurs_d sample some records and store as df; rbind all 

## constant to multiply probabilities by
sub_sample_constant = 1000

## make df to add to 
sub_samples <-  eye_paths_features_d[0, ]

# ## how likely are pathlengths excluding length 1?
# probs <-  table(eye_paths_features_d$path.length[eye_paths_features_d$path.length!=1])/
#                  sum(table(eye_paths_features_d$path.length[eye_paths_features_d$path.length!=1]))
# 
# ## add length 1 probability to probs
# prob_1 <- .05
# probs <- probs-(prob_1/8)
# probs <- c(`1` = prob_1, probs)
# txtdensity(probs)

## make long paths have exponentially more representation than short paths because these are exponentially more complex than short paths, i.e. more combinations of these 
probs <- (1:n_distinct(eye_paths_features_d$path.length))^(1/1.4)
probs <- probs/sum(probs)
barplot(probs, names.arg = 1:length(probs))

n = 1
for(n in seq(max(eye_paths_features_d$path.length))){
  prob <- probs[[n]]
  sub <- eye_paths_features_d[eye_paths_features_d$path.length == n , ]
  sub_sample_size <- floor(sub_sample_constant*prob)
  if(nrow(sub) >= sub_sample_size){
    set.seed(2920)
    sub_sample <- sub[sample(nrow(sub), sub_sample_size, FALSE) , ]
  }else{ 
    sub_sample <- sub
  }
  
  sub_samples <- rbind(sub_samples, sub_sample)
}


(tab <- table(sub_samples$path.length))


# ++++++++
# manually label each path as follows (so long as the path follows this route):----
# ++++++++
#> now 111 = 333, 999 = 444
## a: 3-6-4-111; 6-4-111; 
## b: 3-6-4-999; 6-4-999; 
## c: 111
## d: 999 
## x: none of these

# df of sample to label route
labeled_paths <- data.frame(
  patID_eye = sub_samples$patID_eye,
  path = sub_samples$path,
  route = character(length = length(sub_samples$path))
  )



for(i in 1:nrow(labeled_paths)){
  print(labeled_paths[i, ])
  route = readline("enter route this path follows: ")
  labeled_paths$route[i] <- route
}


mistake_fixes <- tribble(
  ~mistake, ~fix,
  
)

mistakes <- mistake_fixes$mistake
fixes <-mistake_fixes$fix
  
cbind(labeled_paths[labeled_paths$patID_eye %in% mistakes ,], fix = fixes)

labeled_paths$route[labeled_paths$patID_eye %in% mistakes] <- fixes

labeled_paths <- labeled_paths[labeled_paths$route!='',]

# saveRDS(labeled_paths, paste0(wd_path, "processed_data/manually_labeled_paths.rds"))


```

#### mung with labeled path data 
```{r}

# manually labeled paths
labeled_sample <- readRDS(paste0(wd_path, "processed_data/manually_labeled_paths.rds"))

# make manual "NA" NA
labeled_sample <- labeled_sample %>% 
    mutate_all(~ifelse(. %in% c("", "NA"), NA, .))


# new labeled sample adding good route followers who were not already there
all_labeled_paths <- labeled_sample %>% 
  rbind(good_route_followers_features %>% filter(!patID_eye %in% labeled_sample$patID_eye))

# add labels to d----
eye_paths_features_d_labeled <- left_join(eye_paths_features_d, all_labeled_paths, by = "patID_eye") 

## coalesce path and route
all(eye_paths_features_d_labeled$path.x==eye_paths_features_d_labeled$path.y, na.rm = TRUE)
eye_paths_features_d_labeled$path <- with(eye_paths_features_d_labeled, coalesce(path.x,path.y))

# remove redundant cols
eye_paths_features_d_labeled[, c("path.y", "path.x")] <- NULL

# make route factor
eye_paths_features_d_labeled <- eye_paths_features_d_labeled %>% 
  mutate(route = factor(route))

# add numerical route score: A = 2, B = -2, C = 1, D = -1, E =  0, F = 0
eye_paths_features_d_labeled <- eye_paths_features_d_labeled %>% 
  mutate(
    route_score = 
      case_when(
        route == "a" ~ 2, 
        route == "b" ~ -2, 
        route == "c" ~ 1, 
        route == "d" ~ -1, 
        route == "x" ~ 0, 
        TRUE~ NA_real_
      )
  )


# fix where prop.A.to.B or prop.B.to.A NA and route == A or B, make it 1
eye_paths_features_d_labeled$prop.B.to.A[ is.na(eye_paths_features_d_labeled$prop.B.to.A) & eye_paths_features_d_labeled$route %in% c("b")] <- 1
eye_paths_features_d_labeled$prop.A.to.B[ is.na(eye_paths_features_d_labeled$prop.A.to.B) & eye_paths_features_d_labeled$route %in% c("a")] <- 1

# add actual paths; for each eye, grab string of unique clusters; e.g. 3-6-4-9-2-7
actual_eye_paths <- mclapply(unique(tsne$patID_eye), function(an_eye){
  an_eye_path <- tsne[ tsne$patID_eye == an_eye , ] %>% 
    arrange(month) %>% 
    pull(my_cluster) %>% 
    as.character() %>% 
    paste( collapse = "-")
  return(an_eye_path)
}) %>% 
  do.call(rbind,.) %>% 
  as.data.frame

names(actual_eye_paths) <- "actual_path"
actual_eye_paths$patID_eye <- rownames(actual_eye_paths)

eye_paths_features_d_labeled <- eye_paths_features_d_labeled %>% 
  left_join(actual_eye_paths)


# !!!!!!!!!!!!!!!!! !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! !!!!!!!!!!!!!!!!!!!!
# !!!!!!!!!!!!!!!!! convert c to a, d to b, keep x  !!!!!!!!!!!!!!!!!
# !!!!!!!!!!!!!!!!! !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! !!!!!!!!!!!!!!!!!!!!
eye_paths_features_d_labeled <- eye_paths_features_d_labeled %>%
  mutate(granular_route = route) %>%
  mutate(route = case_when(
    route == "c" ~ "a",
    route == "d" ~ "b",
    TRUE ~ route
  )) %>%
  mutate(route = factor(route, levels = c("a", "b", "x")))



```

#### PCA on labeled paths
<!-- ```{r, eval = FALSE} -->

<!-- # data to filter and feed into pca -->
<!-- into_pca <- eye_paths_features_d_labeled %>%  -->
<!--   as.data.frame() -->

<!-- rownames(into_pca) <- eye_paths_features_d_labeled$patID_eye -->

<!-- into_pca <- into_pca %>%  -->
<!--     na.omit() %>%  -->
<!--     select(where(~is.numeric(. ))) -->

<!-- # mds <- cmdscale(dist(t(into_pca)));plot(mds, type = "n"); text (mds, labels = rownames(mds)); abline(a = 0, b = 0); abline(v = 0) -->

<!-- set.seed(309) -->
<!-- pcs <- prcomp(into_pca) -->

<!-- loadings <- pcs$rotation[,1:2] %>% as.data.frame() -->
<!-- # plot(loadings, type = "n") -->
<!-- # abline(h = 0);abline(v = 0) -->
<!-- # text(x = loadings$PC1, y = jitter(loadings$PC2, 1) , label = rownames(loadings)) -->

<!-- loadings$variable <- rownames(loadings) -->

<!-- loadings <- loadings %>% -->
<!--   mutate(across(PC1:PC2, -->
<!--          ~ abs(.), -->
<!--          .names = "abs_{.col}" -->
<!--          )) -->

<!-- ggplot(loadings , aes(PC1, PC2, color = variable))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   geom_label_repel(aes(label = variable),max.overlaps = 20, show.legend = FALSE)+ -->
<!--   geom_vline(xintercept = 0, linetype= 3)+ -->
<!--   geom_hline (yintercept = 0, linetype= 3) -->

<!-- # ggbiplot::ggbiplot(pcs,choices = c(1,2), scale = TRUE, -->
<!-- #                    # ellipse = TRUE,  -->
<!-- #                    circle = TRUE, -->
<!-- #                    groups = eye_paths_features_d_labeled %>% na.omit() %>% pull(route) -->
<!-- #                    )+  -->
<!-- #   theme_bw() -->

<!-- # par(scale = 2, mar = c(4,4,4,4)) -->
<!-- # plot(pcs$rotation[, 1:2], type = "n", xlim = c(-1, 2), ylim = c(-1,2),  -->
<!-- #      xlab = paste0(round(pcs$sdev[1]^2/sum(pcs$sdev^2)*100, 3), "%"), -->
<!-- #      ylab = paste0(round(pcs$sdev[2]^2/sum(pcs$sdev^2)*100, 3), "%") -->
<!-- #      ) -->
<!-- # abline(a =0, b = 0, lty = 2); abline(v = 0, lty = 2) -->
<!-- # text(pcs$rotation[, 1:2], labels = rownames(pcs$rotation)) -->
<!-- # rot_lines <- pcs$rotation[, 1:2] -->
<!-- # # row = rot_lines[1,] -->
<!-- # apply(rot_lines, 1, function(row){ -->
<!-- #  -->
<!-- #   lines(x  = c(0, row["PC1"]), y = c(0, row['PC2']), col = "blue") -->
<!-- #    -->
<!-- # }) -->

<!-- # ggbiplot::ggbiplot(pcs,choices = c(2,3),scale = TRUE, -->
<!-- #                    groups = eye_paths_features_d_labeled %>% na.omit() %>% pull(route),  -->
<!-- #                    ellipse = TRUE, circle = TRUE)+  -->
<!-- #   theme_bw() -->


<!-- # # add first 2 pcs to eye_paths_features_d_labeled -->
<!-- # into_pca <- into_pca %>%  -->
<!-- #   cbind(pcs$x[, 1:2]) %>%  -->
<!-- #   rownames_to_column("patID_eye") -->
<!-- #  -->
<!-- #  -->
<!-- # eye_paths_features_d_labeled <- eye_paths_features_d_labeled %>%  -->
<!-- #   left_join(into_pca) -->

<!-- ``` -->


## model using RF
#### make validation and train data
    
    workflow:
      make full dataset of path features without ids
      split into 30 validation, 70 train
      with 70 train, two step data augmentation
        stratified resample each route to have N observations. this evens out route representation in training
        resample whole dataset to have 3x original size. this gives the model a better chance at learning patterns
      train rf on 70 train df
        regression on rf_route_score
        classification on rf_route
    
```{r}

# path data where route score not missing
truth <-  eye_paths_features_d_labeled %>% 
  filter(!is.na(route)) %>% 
  # remove variables that are missing a lot; e.g. NAN when no B in path, NA when only one node
  select(-c(prop.A.to.B, prop.B.to.A, first.non.root.node, count, actual_path, path), -matches("step")) %>%
  column_to_rownames("patID_eye") %>% 
  ungroup() 

# omit missing
colSums(is.na(truth))
truth <- na.omit(truth)


# split into 70/30 train/test datasets
# set.seed(129049)
set.seed(10309)
inds_validation <- sample(seq(nrow(truth)), floor(nrow(truth)*.3))
validation <- truth[inds_validation,]
train <- truth[-inds_validation,]

rbind(validation =validation$route %>% table(useNA = "ifany"),
train = train$route %>% table(useNA = "ifany"))


## resample each route >> stratified bootstrap resample
sample_size_per_route <- 100
train_resampled <- train[0, ]; 
for(a_route in unique(train$route)){
  
  sub <- train[train$route == a_route,]
  
  set.seed(120920)
  sample_inds <- sample( 1:nrow(sub), sample_size_per_route , TRUE)
  sub_sampled <- sub[sample_inds , ]
  
  train_resampled <- train_resampled %>% rbind(sub_sampled)
}

# bootstrap stepwise sampled data
size_mult <- 3
set.seed(3032)
boot_inds <- sample(1:nrow(train_resampled), size_mult* nrow(train_resampled) , replace = TRUE)
train_resampled_bootstrapped <- train_resampled[ boot_inds ,]

# plot what is going on here
ylim <- c(0, 350)
par(mfrow = c(3,1))
train$route %>% table%>% barplot(., main = "Manual route frequencies before stepwise resampling and bootstrapping", ylim = ylim)
train_resampled$route %>% table%>% barplot(., main = "Manual route frequencies after stepwise resampling", ylim = ylim)
train_resampled_bootstrapped$route %>% table %>% barplot(., main = "Manual route frequencies after stepwise resampling and bootstrapping", ylim = ylim)
par(mfrow = c(1,1))




```

#### model path features
```{r, eval = TRUE}

####################
# train and test model ----
####################
  # predict route score for continuous prediciton
  # set.seed(230)
  # route_score_rf_mod <- randomForest::randomForest(route_score ~ ., 
  #                                                data = train_resampled_bootstrapped %>% select(-route), importance = TRUE, 
  #                                                type = "regression")
  # # In randomForest.default(m, y, ...) :
  # # The response has five or fewer unique values.  Are you sure you want to do regression?
  
how.long()  
# predict route for categorical prediction
set.seed(230)
route_rf_mod <- randomForest::randomForest(route ~ ., 
                                         data = train_resampled_bootstrapped %>% select(-matches(".+route|route+.")),
                                         # data = train %>% select(-matches(".+route|route+.")),
                                         # data = select(truth, -matches(".+route|route+.")), # test on full data
                                         importance = TRUE,
                                         type = "classification")
how.long()  # 1s
route_rf_mod

# # save mods
# saveRDS(route_rf_mod,  paste0(wd_path, "processed_data/stepwise_resample_route_from_path_feature_rf_mod.rds"))
# saveRDS(route_score_rf_mod,  paste0(wd_path, "processed_data/stepwise_resample_route_score_from_path_feature_rf_mod.rds"))


```


##### Eval model
###### try other seeds to test instability
```{r, eval = FALSE}

set.seed(1234)
seeds <- runif(499, 11, 1.3e6) %>% floor()
# seeds with og seed
seeds <- c(230, seeds)

# run mod on diff seeds
other_route_rf_mods <- mclapply(seeds, function(seed){
  set.seed(seed)
  fit <- randomForest::randomForest(route ~ ., 
    # data = train_resampled_bootstrapped %>% select(-matches(".+route|route+.")),
    data = select(truth, -matches(".+route|route+.")), # test on full data
    importance = TRUE, 
    type = "classification")
  return(fit)
}, mc.set.seed = F, mc.cores = detectCores()-1)

# OOB error rates
err_rates <- sapply(other_route_rf_mods, function(mod) tail(mod$err.rate[, "OOB"], 1) * 100)
err_rates%>% txtplot::txtdensity(height = 10, width = 80)
#   +--***+-----------+------------+-----------+-----------+-----------+--------+
# 20 +    ****                                                                   +
# 15 +       ***                                                                 +
#    |         ***                                                               |
# 10 +           ***                                                       **    +
#    |             ***                                               *******     |
#  5 +               ****                                        *****           +
#    |                  ********                        **********               |
#  0 +-----+-----------+-------**************************---+-----------+--------+
#        0.56        0.58          0.6        0.62        0.64        0.66        

predictions <- sapply(other_route_rf_mods, `[[`,3) 

# how much do different models aggree?
consensus_error <- apply(predictions, 1, function(prediction){
  n_distinct(prediction)/length(prediction)
})
(consensus_error) %>% txtplot::txtdensity(height = 10, width = 80)
#      +--**---------------+---------------+---------------+----------------+----+
# 5000 +   **                                                                    +
# 4000 +    *                                                                    +
# 3000 +    **                                                                   +
#      |     *                                                                   |
# 2000 +      *                                                                  +
# 1000 +      **                                                                 +
#      |       ***                                                        ***    |
#    0 +--+------**********************************************************-+----+
#       0.002           0.0025           0.003          0.0035            0.004   


# of all samples, how many had any miss identification?
(miss_id_rate <- sum(apply(predictions, 1, function(row) n_distinct(row)!=1) )/nrow(predictions))
# ~ 2%

# how does the consensus compare to teh current main model? if two maxes, choose randomly
consensus <- apply(predictions[,-1], 1, function(prediction){
  tab <- table(prediction)
  mx <- max(tab)  
  if(sum(tab==mx)>1){
    cons <- sample(names(tab)[tab==mx],1)
  }else{
    cons <-  names(tab)[tab==mx]
  }
})

table(consensus = consensus, original  = predictions[,1]) %>% 
  {print(.);sum(diag(.))/sum(.)}
# 0.9981238


rm("other_route_rf_mods")

```

###### accuracy and rmse
```{r, eval = TRUE}

# ****** FUNCTION: calculate accuracy score of confusion matrix *****
calc.accuracy <- function(sqr_matrix){return(sum(diag(sqr_matrix)) / sum(sqr_matrix))}

# predict validation data
route_predictions <- predict(route_rf_mod, newdata = validation %>% select(-matches("^route_score$")) %>%  mutate(route = NA))
# route_score_predictions <- predict(route_score_rf_mod, newdata = validation %>% select(-route) %>%  mutate(route_score = NA))


# confusion matrix and accuracy
(route_rf_confusion_m <- table(actual = validation$route, predicted = route_predictions)) # predicted x actual
(accuracy <- calc.accuracy((route_rf_confusion_m)))
# .962>> 0.93 (diff seed on inds) >>  0.95

## accuracy without x
calc.accuracy((route_rf_confusion_m[-5,-5]))
# 0.95 >>

# (rmse <- rmse(route_score_predictions, validation$route_score))
# # ^ 0.32 >> 0.35 (diff seed on inds) >> .30
# ## without x, vis. 0
# rmse(route_score_predictions[validation$route_score!=0], validation$route_score[validation$route_score!=0])
# # 0.27>> 
# 
# png(paste0(wd_path, "results/clustering/12_clustering/rf_route_score_actualvpredicted.png"), height = 800, width = 900)
# plot(factor(validation$route_score), route_score_predictions,ylab ="predicted", xlab = "actual", main = "route score prediction from path features", ylim = c(-2,2)); 
# text(1, pos = 4, label = sprintf("RMSE = %0.2f\nAccuracy = %0.2f", rmse, accuracy))
# dev.off()

```


###### feature importance
```{r, eval = TRUE}

path_feature_importance_m <- cbind(route_rf_mod$importance)

# scale A:F decrease in accuracy across routes 
heatmap_data <- path_feature_importance_m[, which(colnames(path_feature_importance_m) %in% letters)] %>% 
  t() %>% 
  scale() %>% 
  t()

heatmap_data[is.nan(heatmap_data)] <- 0

annotation_data <- path_feature_importance_m[, which(!colnames(path_feature_importance_m) %in% colnames(heatmap_data) )] %>% 
  as.data.frame()

path_feature_importance_m[, which(!colnames(path_feature_importance_m) %in% colnames(heatmap_data) )] %>% 
  scale() %>% 
  rowSums() %>% 
  sort()


library(circlize)
num_colors <- 2
# Define custom color functions using viridis palettes
mean_decrease_accuracy_col_fun <- circlize::colorRamp2(
  c(min(annotation_data$MeanDecreaseAccuracy), max(annotation_data$MeanDecreaseAccuracy)), 
  plasma(num_colors)
)
mean_decrease_gini_col_fun <- circlize::colorRamp2(
  c(min(annotation_data$MeanDecreaseGini), max(annotation_data$MeanDecreaseGini)), 
  plasma(num_colors)
)
# inc_mse_col_fun <- circlize::colorRamp2(
#   c(min(annotation_data$`%IncMSE`), max(annotation_data$`%IncMSE`)), 
#   plasma(num_colors)
# )
# inc_node_purity_col_fun <- circlize::colorRamp2(
#   c(min(annotation_data$IncNodePurity), max(annotation_data$IncNodePurity)), 
#   plasma(num_colors)
# )

# Create annotations using custom color functions
annotation <- rowAnnotation(
  df = annotation_data,
  col = list(
    MeanDecreaseAccuracy = mean_decrease_accuracy_col_fun,
    MeanDecreaseGini = mean_decrease_gini_col_fun
    # `%IncMSE` = inc_mse_col_fun,
    # IncNodePurity = inc_node_purity_col_fun
  )
)


# Plot the heatmap with the left annotations
hm <- Heatmap(heatmap_data, 
        clustering_method_rows = "ward.D2",
        clustering_method_columns = "ward.D2",
        column_title =  "Path feature RF importance metrics", 
        name = "scaled decrease in accuracy", 
        column_names_rot = TRUE, 
        # col = redblue(100), 
        left_annotation = annotation
      )

png(paste0(wd_path, "results/clustering/12_clustering/stepwise_resampled_class_and_regress_rf_without_step_importance_hm.png"), height = 1000, width = 900)
draw(hm)
dev.off()
```

###### see how model performs on out of bag routes
```{r, eval = FALSE}

## simulate some paths: 1000 paths, of >0 <11 length, composed of 1:5, 333 and 444
set.seed(3930)
some.paths <- sapply(1:10000, function(i) sample(unique(c(A,B)), sample(1:10, 1), TRUE))
some.path.features <- mclapply(some.paths, extract.path.features, mc.cores = detectCores()-2) %>% 
  do.call(rbind,.) %>% 
  as.data.frame()

# add other features
new_features <-  grep("route",setdiff(names(train), colnames(some.path.features)), invert = TRUE, value = TRUE)
for(feature in new_features){
  set.seed(123290)
  some.path.features[[feature]] <- sample(eye_paths_features_d[[feature]], nrow(some.path.features), T)
}

some.path.features$route <- predict(route_rf_mod, some.path.features)

distrs <- sapply(as.character(unique(some.path.features$route)), function(route){
  some.path.features$path[some.path.features$route==route] %>% 
    str_split(.,"-") %>% 
    unlist %>% 
    table()
}, simplify = FALSE) %>% 
  sapply(function(tab){
    missing_routes <- setdiff(as.character(c(A,B)), names(tab)) 
    tab <- c(tab, setNames(rep(0,length(missing_routes)),missing_routes))
    return(tab)
  } )


freqs <- distrs %>% 
  as.data.frame() %>% 
  rownames_to_column("node") %>% 
  mutate(node = factor(node, levels = as.character(c(3,6,4,111,999)))) %>% 
  melt(value.name = "frequency")
  
freqs$node %>% unique()
ggplot(freqs, aes(node, frequency, color = node, fill = node))+
  facet_wrap(~variable, scales = "free")+
  geom_bar(stat = "identity")+
  geom_text(aes(label = node), color = "black")+
  ggtitle("Out of bag paths prediction node distributions")
ggsave(paste0(wd_path, "results/clustering/12_clustering/rf_route_mod_out_of_bag_prediction_node_distributions.png"), height = 10, width = 15)



```

#### model PC features
##### add pc features to train data
     for each eye, calculate avg of each PC , interaction of PC1xPC2 ^2
```{r}
# plot trees with routes ----
pc_tree <- metadata(sce)$trajectory_data$pca_line_data

PC <- reducedDim(sce, "PCA") %>% 
    as.data.frame() %>% 
    rownames_to_column("sample") %>% 
    left_join(
      colData(sce) %>% 
        data.frame() %>% 
        select(sample, patID_eye,patID, my_cluster,month ),
      .
    ) %>% 
  left_join(eye_paths_features_d_labeled %>% 
              select(patID_eye,route)) %>% 
  mutate(route = as.character(route))



pc_labs <- PC %>% 
  summarize(across(matches("PC"), median), .by = my_cluster)

pc_traj <- lapply(2:50, function(i){
  PCy <- paste0("PC", i)
  PCx <- paste0("PC", i-1)
  ggplot(PC %>% filter(!is.na(route)), aes(get(PCx), get(PCy)))+
    geom_point(alpha = .5, aes(color = route))+
    scale_color_manual(values = c("a" = "red", "b" = "blue", "x"= "grey"))+
    # geom_point(alpha = .5, aes(color = my_cluster))+
    # scale_color_manual(values = cluster_cols)+
    geom_line(data = pc_tree, aes(group = edge), linewidth = 2.0)+
    geom_label(data = pc_labs, aes(get(PCx), get(PCy),label = my_cluster), cex = 3)+
    xlab(PCx)+
    ylab(PCy)+
    theme_bw()+
    theme(aspect.ratio = 1)
})

grid <- wrap_plots(pc_traj, ncol = 10)
hw <- c(20,50)
ggsave(paste0(wd_path, "results/clustering/12_clustering/PCs_tree.pdf"),grid, width = hw[2],height = hw[1], limitsize = F)


# # find what predicts route well from pcs ----
# preds <- paste0("PC", 1:50)
# mods <- mclapply(preds, function(pc){
#   form <- "route=='b'~ (1|patID) + (1|patID_eye)" %>% 
#     sprintf("%s + %s",., pc) %>% 
#     as.formula()
#   glmer(form, data = PC %>% select(-c( sample, my_cluster)) %>% filter(route!="x"), family = binomial)
# }, mc.cores = detectCores()-1)
# 
# names(mods) <- preds
# summs <- sapply(mods, summary, simplify = FALSE)
# 
# extract_results(summs) %>% 
#   arrange(`Pr(>|z|)`) %>% 
#   filter(!grepl("Int", Term)) %>% 
#   ggplot(aes(Estimate, -log10(`Pr(>|z|)`), color = Term))+
#   geom_hline(yintercept = -log10(0.05), linetype = 3)+
#   geom_vline(xintercept = 0, linetype = 3)+
#   geom_point(show.legend = FALSE)+
#   geom_label_repel(aes(label = Term), show.legend = FALSE)+
#   theme_bw()

# MAKE PC FEATURES ----
pc_trnsfrm <- function(x){
  (x^2)*sign(x)
}

# pc features
pc_features_raw <- PC %>% 
  arrange(month) %>% 
  select(patID_eye, matches("PC")) %>%
  distinct() %>% 
  mutate(vis = row_number(), .by = patID_eye)

pc_features <- pc_features_raw%>% 
  group_by(patID_eye) %>%
  # mutate(across(PC1, ~{
  #   PC1*(max(1, 1.01*(PC1>0)))
  #   })) %>% 
  # mutate(across(matches("PC"), ~{.*vis} )) %>% 
  mutate(PC1xPC2 = PC1*PC2) %>% 
  # mutate(PC1xPC2 = log(abs(PC1xPC2))*sign(PC1xPC2)) %>%
  mutate(PC1sqrd = PC1 %>% pc_trnsfrm()) %>% 
  mutate(PC2sqrd = PC2 %>% pc_trnsfrm()) %>% 
  mutate(PC1sqrdxPC2sqrd = (PC1sqrd*PC2sqrd)) %>% 
  # mutate(PC1sqrdxPC2sqrd = log(abs(PC1sqrdxPC2sqrd))*sign(PC1sqrdxPC2sqrd)) %>%
  summarize(across(matches("^PC\\d"), mean, .names = "{.col}_mean")) %>%
  # mutate(n_vis = n()) %>% 
  ungroup() %>%
  distinct() %>% 
  left_join(PC %>% select(patID_eye, patID, route) %>% distinct(), relationship = "one-to-many")


p1 <- pc_features %>% 
  # filter(!is.na(route)) %>% 
  ggplot(aes(PC1sqrd_mean, PC2sqrd_mean))+
  geom_point(aes(color = PC1xPC2_mean))+
  # scale_color_viridis_c()+
  scale_color_gradient2()+
  theme_bw()+
  theme(aspect.ratio = 1)

p2 <- pc_features %>% 
  filter(!is.na(route)) %>%
  ggplot(aes(PC1sqrd_mean, PC2sqrd_mean))+
  geom_point(aes(color = route))+
  theme_bw()+
  theme(aspect.ratio = 1)

p3 <- pc_features %>% 
  # filter(!is.na(route)) %>% 
  ggplot(aes(PC1sqrd_mean, PC2sqrd_mean))+
  geom_point(aes(color = PC1sqrdxPC2sqrd_mean))+
  scale_color_gradient2()+
  # scale_color_viridis_c()+
  theme_bw()+
  theme(aspect.ratio = 1)

p1/p2/p3


# how do pcas look after feature engineering?
pca <- prcomp(select(pc_features,-c(patID_eye, patID, route) ))

post_pc_plots <- lapply(2:50, function(i){
  x <- paste0("PC",i-1)
  y <- paste0("PC",i)
  pca$x %>% 
    as.data.frame() %>% 
    cbind(route = pc_features$route) %>% 
    filter(!is.na(route)) %>% 
    ggplot(aes(get(x),get(y), color = route))+
    geom_point()+
    xlab(x)+
    ylab(y)+
    theme_bw()+
    theme(aspect.ratio = 1)+
    scale_color_brewer(palette = "Set1") 
})

post_pc_plots_grid <- wrap_plots(post_pc_plots, ncol = 10)
ggsave(paste0(wd_path, "results/clustering/12_clustering/pc_features_pca.png"),post_pc_plots_grid, height = 10, width = 28)

```

##### train model
```{r}

pc_vars <- grep("PC", names(pc_features), value = TRUE) %>% 
  # grep(paste(20:50, collapse = "|"), ., invert = TRUE, value = TRUE) %>%
  {}

# make pc truth, train, validate
pc_truth <- pc_features %>% 
  select(route,all_of(pc_vars)) %>%
  # select(route, PC1_mean,PC2_mean, PC1xPC2_mean, PC1sqrd_mean, PC2sqrd_mean, PC1sqrdxPC2sqrd_mean) %>%
  drop_na()  %>% 
  mutate(route = factor(route))

setdiff(names(pc_features), names(pc_truth))


# make test/train set 
# set.seed(10290)
set.seed(12039)
test_ratio <- .3

test_inds <- sample(seq(nrow(pc_truth)), floor(test_ratio*nrow(pc_truth)))
pc_test <- pc_truth[test_inds, ]
pc_train <- pc_truth[-test_inds,]

# # bootstrap training data
# set.seed(10290)
# train_inds <- sample(seq(nrow(pc_train)), 1*nrow(pc_train), T)
# pc_train <- pc_train[train_inds, ]

# RF ----
# run full model
set.seed(10390)
pc_route_rf_mod <- randomForest(route~., 
                          data = pc_truth,
                          type = "classification"
                          )


# run evaluation model
set.seed(10390)
pc_route_rf_eval_mod <- randomForest(route~., 
                          data = pc_train,
                          type = "classification"
                          )
pc_route_rf_eval_mod
#  randomForest(formula = route ~ ., data = pc_train, type = "classification") 
#                Type of random forest: classification
#                      Number of trees: 500
# No. of variables tried at each split: 7
# 
#         OOB estimate of  error rate: 24.33%
# Confusion matrix:
#    a   b   x class.error
# a 62  16  16   0.3404255
# b 10 112  17   0.1942446
# x 14  18 109   0.2269504

# LDA  ----
# run full model
set.seed(10390)
pc_route_lda_mod <- MASS::lda(route~., 
                          data = pc_truth
                          )

# class priors
priors <- table(pc_truth$route) %>% {./sum(.)}

set.seed(10390)
pc_route_lda_eval_mod <- MASS::lda(route~., 
                          data = pc_train,
                          prior = as.numeric(priors)
                          )


```


##### Eval model
###### test on validate set
```{r}

# RF 
table(predicted  = predict(pc_route_rf_eval_mod, pc_test), 
      actual =pc_test$route) %>% 
  {print(.);print(c(N = sum(.)));calc.accuracy(.)}
#         actual
# predicted  a  b  x
#         a 27  5  3
#         b  9 47  3
#         x 12  6 47
#   N 
# 159 
# [1] 0.7610063


# LDA 
pred <- predict(pc_route_lda_eval_mod, pc_test)

table(pred = pred$class, 
      actual = pc_test$route) %>% 
  {print(.);print(c(N = sum(.)));calc.accuracy(.)}
#     actual
# pred  a  b  x
#    a 34  3  1
#    b  4 52  2
#    x 10  3 50
#   N 
# 159 
# [1] 0.8553459

```

    lda is better at classifying along  linear decision boundaries. RF is good at finding non-linear patterns, but overfits in this setting
    
####### look at lda boundaries
```{r}

set.seed(10390)
pc_route_lda_eval_mod_simple <-MASS::lda(route~., 
                          data = pc_truth %>% select(route, PC1_mean, PC2_mean),
                          prior = as.numeric(priors)
                          )

# Create a prediction grid
grid <- expand.grid(
  PC1_mean = seq(min(pc_test$PC1_mean)-1, max(pc_test$PC1_mean)+1, length.out = 200),
  PC2_mean = seq(min(pc_test$PC2_mean)-1, max(pc_test$PC2_mean)+1, length.out = 200)
)

# Predict class probabilities over the grid
grid_pred <- predict(pc_route_lda_eval_mod_simple, newdata = grid)

# Combine with grid
grid$class <- grid_pred$class

# Plot
ggplot(pc_train, aes(x = PC1_mean, y = PC2_mean, color = route)) +
  geom_point(size = 2) +
  geom_contour(data = cbind(grid, z = as.numeric(grid$class)),
               aes(x = PC1_mean,y = PC2_mean, z = z), breaks = c(1.5, 2.5), color = "black") +
  theme_minimal() +
  scale_color_manual(values = c(scales::hue_pal()(3)[-3],"darkgrey"))+
  xlab("PC1 Average")+
  ylab("PC2 Average")+
  labs(color = "Trajectory")+
  theme(aspect.ratio = 1) +
  ggtitle("LDA decision boundaries")
ggsave(paste0(wd_path, "results/clustering/12_clustering/pc_lda_pc1m_pc2m_boundaries.png"), height = 5, width = 6, bg = "white")

```

   
###### error rate and importance
```{r}

cols <- c("green", "red", "blue")
pc_route_rf_eval_mod$votes[,1, drop = TRUE] %>% 
  density(.) %>% 
  plot(col = cols[1], main  = "Votes")
lapply(2:3 , function(i){
  pc_route_rf_eval_mod$votes[,i] %>% 
    density(.) %>% 
    lines(col = cols[i])
})
legend("topright", legend = colnames(pc_route_rf_eval_mod$votes), fill = cols)

pc_route_rf_eval_mod$importance %>% 
  Heatmap(col = plasma(10))

err.rate <- pc_route_rf_eval_mod$err.rate

err.rate %>% 
  as.data.frame( ) %>%
  mutate(tree = row_number()) %>% 
  pivot_longer(-tree) %>% 
  mutate(name = factor(name, levels = c("a", "b", "x", "OOB"))) %>% 
  ggplot(aes(tree,value))+
  geom_line()+
  facet_wrap(~name)+
  ggtitle("class error rates")+
  theme_bw()

err.rate %>% 
  apply(., 2, which.min) %>% 
  {print(.); mean(.)}
# OOB   a   b   x 
# 210  48 332 191 
# [1] 195.25

```

###### try other seeds to test instability
```{r, eval = FALSE}

set.seed(1234)
seeds <- runif(499, 11, 1.3e6) %>% floor()
# seeds with og seed
seeds <- c(10390, seeds)

# run mod on diff seeds
other_route_rf_mods <- mclapply(seeds, function(seed){
  set.seed(seed)
  fit <- randomForest(route~., 
                          data = pc_truth,
                          type = "classification")
  return(fit)
}, mc.set.seed = F, mc.cores = detectCores()-1)

# OOB error rates
err_rates <- sapply(other_route_rf_mods, function(mod) tail(mod$err.rate[, "OOB"], 1) * 100)
err_rates%>% txtplot::txtdensity(height = 10, width = 80)
#     +--------+----------+----------+-*****-----+----------+----------+---------+
# 0.6 +                           ******   *****                                 +
#     |                        ****            ******                            |
# 0.4 +                    *****                    ****                         +
#     |                 ****                           ****                      |
#     |             *****                                 ****                   |
# 0.2 +        ******                                        *****               +
#     |  *******                                                 ***********     |
#   0 +--------+----------+----------+-----------+----------+----------+---***---+
#            20.5        21        21.5         22        22.5        23          

predictions <- sapply(other_route_rf_mods, `[[`,3) 

# how much do different models aggree?
consensus_error <- apply(predictions, 1, function(prediction){
  n_distinct(prediction)/length(prediction)
})
(consensus_error) %>% txtplot::txtdensity(height = 10, width = 80)
# 1200 +--**---------------+---------------+---------------+----------------+----+
#      |   ***                                                                   |
# 1000 +     **                                                                  +
#  800 +      **                                                                 +
#  600 +       **                                                                +
#  400 +        **                                                               +
#  200 +         ***                    *******                                  +
#      |           ****           *******     *******                ********    |
#    0 +--+------------************--------+--------******************------+----+
#       0.002            0.003           0.004           0.005            0.006   


# of all samples, how many had any miss identification?
(miss_id_rate <- sum(apply(predictions, 1, function(row) n_distinct(row)!=1) )/nrow(predictions))
# ~ 20%

# how does the consensus compare to teh current main model? if two maxes, choose randomly
consensus <- apply(predictions[,-1], 1, function(prediction){
  tab <- table(prediction)
  mx <- max(tab)  
  if(sum(tab==mx)>1){
    cons <- sample(names(tab)[tab==mx],1)
  }else{
    cons <-  names(tab)[tab==mx]
  }
})

table(consensus = consensus, original  = predictions[,1]) %>% 
  {print(.);sum(diag(.))/sum(.)}
#         original
# consensus   a   b   x
#         a 194   1   3
#         b   1 315   1
#         x   1   2 294
# [1] 0.9889163


rm("other_route_rf_mods")

```

#__________________
# deploy models on data
```{r, eval = TRUE}

eye_paths_features_d_labeled_new <- eye_paths_features_d_labeled
eye_paths_features_d_labeled_new$rf_route <- predict(route_rf_mod, eye_paths_features_d_labeled)

# add to eye_paths_features_d_labeleled
eye_paths_features_d_labeled <- eye_paths_features_d_labeled %>%
  left_join(pc_features)

eye_paths_features_d_labeled_new$pc_rf_route <- predict(pc_route_rf_mod, eye_paths_features_d_labeled)

eye_paths_features_d_labeled_new$pc_lda_route <- suppressWarnings( predict(pc_route_lda_mod, eye_paths_features_d_labeled)$class )

table(eye_paths_features_d_labeled_new$rf_route, useNA = "always")
table(eye_paths_features_d_labeled_new$pc_rf_route, useNA = "always")
table(eye_paths_features_d_labeled_new$pc_lda_route, useNA = "always")


```


# check distribution of nodes
```{r, eval = TRUE}

which_route <- "pc_rf_route"
# which_route <- "rf_route"
rf_route_paths <- eye_paths_features_d_labeled_new %>% 
  distinct(get(which_route), patID_eye, .keep_all = TRUE) %>% 
  select(path, all_of(which_route)) %>% 
  na.omit()

graph_nodes <- c(1,9,111,999,3,6,4,12) %>% as.character()

rf_route_node_counts <- list();
for(a_route in unique(rf_route_paths[[which_route]])){
  sub <- rf_route_paths[rf_route_paths[[which_route]] == a_route, ]
  nodes <- unlist(sub$path) %>% 
    str_split(., "-") %>% 
    unlist()
  node_counts <-  as.matrix(table(nodes))
  # add nodes not present for this route
  nodes_not_here <- setdiff(graph_nodes, rownames(node_counts))
  node_counts <- rbind(node_counts, 
                       matrix( 0, ncol = 1, nrow = length(nodes_not_here), dimnames = list(nodes_not_here) )
                       )
  # node_counts <- node_counts[order(rownames(node_counts)) , ]
  rf_route_node_counts[[a_route]] <- node_counts
}

rf_route_node_count_d <-  rf_route_node_counts %>% 
  sapply(function(x){
    missing_routes <- setdiff(as.character(c(A,B)), rownames(x)) 
    y <- matrix(0, ncol = 1,nrow = length(missing_routes), dimnames = list(missing_routes,NULL))
    x <- rbind(x, y)
    x <- x[order(rownames(x)), ]
    return(x)
  } ) %>% 
  as.data.frame() 

  
rf_route_node_count_d$node <- rownames(rf_route_node_count_d)

rf_route_node_count_d %>% 
  reshape2::melt(id.vars = "node", variable.name = "route", value.name = "value") %>% 
  ggplot(aes(node, value, fill = node))+
  geom_bar(stat = "identity", show.legend = FALSE)+
  facet_wrap(~factor(route, levels = letters), scales = "free", ncol  = 2)+
  ggtitle("RF route prediction node frequencies", which_route)
ggsave(paste0(wd_path, sprintf("results/clustering/12_clustering/rf_route_prediction_node_frequencies_%s.png", which_route)), height = 10, width = 12)

```

#--
# logic based assignment
- when prop in A > prop in B -> A
- when count A termini > count B > termini -> A

## apply functions
```{r}

dens <- function(...){plot(density(...))}

# df to test logic classification
logic_test <- eye_paths_features_d_labeled_new

# STRICT FUCTION----
# make score as count of nodes unique to each route
strict.A.score <-with(logic_test, 
                count.in.A )#%>% {log(.+1)})
strict.B.score <- with(logic_test, 
                count.in.B)#%>%  {log(.+1)})
dens(strict.B.score)

# add average depth scaled by how many termini 
strict.A.score <- with(logic_test, 
                ifelse(count.A.termini>count.B.termini, strict.A.score + (norm.total.depth*count.A.termini), strict.A.score))

strict.B.score <- with(logic_test, 
                      ifelse(count.B.termini>count.A.termini, strict.B.score + (norm.total.depth*count.B.termini), strict.B.score))

dens(strict.B.score)

logic_test$strict.A.score <- strict.A.score
logic_test$strict.B.score <- strict.B.score

# make overall score
logic_test$strict.A.minus.B <- logic_test$strict.A.score -logic_test$strict.B.score

logic_test$strict.route <- with(logic_test, 
                                  ifelse(strict.A.score>strict.B.score, "a", 
                                         ifelse(strict.B.score>strict.A.score, "b", NA)))

# FLEXIBLE FUCTION----
# make score as count of nodes unique to each route plus count route specific termini 
flexible.A.score <-with(logic_test, 
                  count.in.A+ count.A.termini)
flexible.B.score <- with(logic_test, 
                  count.in.B + count.B.termini)

dens(flexible.B.score)

logic_test$flexible.A.score <- flexible.A.score
logic_test$flexible.B.score <- flexible.B.score

# make overall score
logic_test$flexible.A.minus.B <- logic_test$flexible.A.score -logic_test$flexible.B.score

logic_test$flexible.route <- with(logic_test, 
                                  ifelse(flexible.A.score>flexible.B.score, "a", 
                                         ifelse(flexible.B.score>flexible.A.score, "b", NA)))



# # TRY USING WEIGHTS FROM LOG MOD ----
# mod_d <- eye_paths_features_d_labeled_new %>% 
#   filter( route %in% c("a", "b")) %>%
#   mutate(route_a = (route =="a")*1) %>% 
#   # select(route_a,matches("length|distinct\\.nodes|longest\\.run$|count|prop|depth|PC")) 
#   select(route_a,norm.total.depth, count.in.A,count.in.B,count.A.termini,count.B.termini)
# 
# head(mod_d)
# 
# mod <- glm(route_a~ ., data = mod_d ,family = binomial)
# 
# # predict using mod; convert to probability (mod is in log odds)
# strict.A.score <- predict(mod, mod_d) %>% 
#   {1/(1+exp(.*-1))}
# 
# logic_test$strict.A.score <- strict.A.score
# logic_test$strict.B.score <- 1-strict.A.score
# 
# logic_test$strict.route <- with(logic_test, 
#                                   ifelse(strict.A.score>strict.B.score, "a", 
#                                          ifelse(strict.B.score>strict.A.score, "b", NA)))

```


## test logical assignment
```{r}

table(actual = logic_test$route,
      predicted = logic_test$strict.route)[1:2,1:2] %>%
  calc.accuracy()
table(actual =  logic_test$route,
      predicted = logic_test$flexible.route)[1:2,1:2] %>%
  calc.accuracy()
# ^ both ~0.98

```


## add logic routes to eye_paths_features_d_labeled_new 
```{r}

# add logic routes to eye features
eye_paths_features_d_labeled_new <- eye_paths_features_d_labeled_new %>% 
  left_join(logic_test) %>% 
  mutate(across(c(strict.route,flexible.route), ~factor(.)))


```
> logic good compared to rf

#--
# add routes to sce
```{r}

# make df of path and route
coldat_w_routes <- colData(sce) %>% 
  as.data.frame() %>% 
  left_join(
    eye_paths_features_d_labeled_new %>% 
      select(patID_eye, matches("route|A.minus.B|^path$|strict|flexible"),-terminus.route) %>%
      distinct(),
    relationship =  "many-to-one")

# add path and route info to sce
cols_to_add <- setdiff(names(coldat_w_routes), names(colData(sce)))
for(column in cols_to_add){
  if(!column %in% names(colData(sce))) colData(sce)[[column]] <- coldat_w_routes[[column]] else cat(column, "already in sce\n\n")
}

```

## plot Routes on reduced dims
```{r}

routes <- grep("route", names(colData(sce)), value = TRUE)

tab <-  table(as.data.frame(colData(sce)[,routes]) %>% mutate_all(as.character)) 

tab_d <- as.data.frame(tab) %>% 
  arrange(desc(Freq)) %>% 
  filter(Freq>0)

kableExtra::kable(tab_d, format = "html", table.attr = "class='table table-condensed'", caption = "<span style='color: #222; font-weight: bold;'>Route Assignment Combinations</span>",) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::row_spec(0, bold = TRUE, background = "#444", color = "white") %>%  
  kableExtra::column_spec(1:ncol(tab_d), width = "5em")



tsne_tree <- geom_line(data = metadata(sce)$trajectory_data$tsne_line_data, aes(TSNE1,TSNE2, group = edge))
umap_tree <- geom_line(data = metadata(sce)$trajectory_data$umap_line_data, aes(UMAP1,UMAP2, group = edge))
pca_tree <- geom_line(data = metadata(sce)$trajectory_data$pca_line_data, aes(PC1,PC2, group = edge))


# ****** FUNCTION to plot each route with tree
plotFunc <- function(route=character(), rdim=character(), tree = data.frame()){
  if(n_distinct(colData(sce)[[route]])>6){
    plot <- sce[, !is.na(sce[[route]])] %>% 
      plotReducedDim( dimred = rdim, color_by = route)+
      ggtitle(route)+
      tree+
      theme(aspect.ratio = 1)+
      scale_color_gradient2(mid = "grey")
  }else{
    plot <- sce[, !is.na(sce[[route]])] %>% 
      plotReducedDim( dimred = rdim, color_by = I( factor( colData(sce)[!is.na(sce[[route]]) , route] ) ))+
      ggtitle(route)+
      tree+
      theme(aspect.ratio = 1)+
      scale_color_brewer(palette = "Set1") 
  }
  return(plot)
}
# plotFunc("granular_route", "TSNE", tsne_tree)
# plotFunc(route = "route_score",rdim =  "TSNE",tree= tsne_tree)

tsnes <- lapply(routes, function(route) plotFunc(route, "TSNE", tsne_tree))

tsnes%>% 
  wrap_plots(ncol = 4)
ggsave(paste0(wd_path, "results/clustering/12_clustering/route_tsnes.png"), width = 17)

umaps <- lapply(routes, function(route) plotFunc(route, "UMAP", umap_tree))

umaps%>% 
  wrap_plots(ncol = 4)
ggsave(paste0(wd_path, "results/clustering/12_clustering/route_umaps.png"), width = 17)


pcas <- lapply(routes, function(route) plotFunc(route, "PCA", pca_tree))

pcas%>% 
  wrap_plots(ncol = 4)
ggsave(paste0(wd_path, "results/clustering/12_clustering/route_pcas.png"), width = 17)


nm = "rf_route"
mclapply(routes[grepl("route$", routes)], function(nm){
  p <- make_d() %>%
    left_join(tsne) %>%
    filter(!is.na(get(nm))) %>%
    ggplot(aes(TSNE1,TSNE2))+
    geom_point(aes(color = get(nm)))+
    scale_color_manual(values = c("a" = "red", "c" = "orange", "b" = "blue", "d" = "purple", "x"= "grey"))+
    theme_bw()+
    labs(color = nm)+
    theme(aspect.ratio = 1)+
    ggtitle(nm)+
    tsne_tree
  ggExtra::ggMarginal(p, type = "density", margins = "both", groupColour = TRUE, groupFill = TRUE)
}, mc.cores = 4) %>% 
  wrap_plots(ncol = 4)
ggsave(paste0(wd_path, "results/clustering/12_clustering/route_tsnes_w_dens.png"), width = 17)

mclapply(routes[grepl("route$", routes)], function(nm){
  p <- make_d() %>%
    left_join(reducedDim(sce, "UMAP") %>% data.frame() %>% rownames_to_column("sample")) %>% 
    filter(!is.na(get(nm))) %>%
    ggplot(aes(UMAP1,UMAP2))+
    geom_point(aes(color = get(nm)))+
    scale_color_manual(values = c("a" = "red", "c" = "orange", "b" = "blue", "d" = "purple", "x"= "grey"))+
    theme_bw()+
    labs(color = nm)+
    theme(aspect.ratio = 1)+
    ggtitle(nm)+
    umap_tree
  ggExtra::ggMarginal(p, type = "density", margins = "both", groupColour = TRUE, groupFill = TRUE)
}, mc.cores = 4) %>% 
  wrap_plots(ncol = 4)
ggsave(paste0(wd_path, "results/clustering/12_clustering/route_umaps_w_dens.png"), width = 17)



mclapply(routes[grepl("route$", routes)], function(nm){
  p <- make_d() %>%
    left_join(reducedDim(sce, "PCA") %>% data.frame() %>% rownames_to_column("sample")) %>% 
    filter(!is.na(get(nm))) %>%
    ggplot(aes(PC1,PC2))+
    geom_point(aes(color = get(nm)))+
    scale_color_manual(values = c("a" = "red", "c" = "orange", "b" = "blue", "d" = "purple", "x"= "grey"))+
    theme_bw()+
    labs(color = nm)+
    theme(aspect.ratio = 1)+
    ggtitle(nm)+
    pca_tree
  ggExtra::ggMarginal(p, type = "density", margins = "both", groupColour = TRUE, groupFill = TRUE)
}, mc.cores = 4) %>% 
  wrap_plots(ncol = 4)
ggsave(paste0(wd_path, "results/clustering/12_clustering/route_pcas_w_dens.png"), width = 17)


```


# ***look at route counts
```{r}

# plot logic score densities before and after join
par(mfrow = c(2,2))
plot(density(eye_paths_features_d_labeled_new$flexible.A.minus.B, na.rm = TRUE))
plot(density(eye_paths_features_d_labeled_new$strict.A.minus.B, na.rm = TRUE))
plot(density(sce$flexible.A.minus.B, na.rm = TRUE))
plot(density(sce$strict.A.minus.B, na.rm = TRUE))
par(mfrow = c(1,1))

# check that manual route count same 
table(eye_paths_features_d_labeled_new$flexible.route) 
table(eye_paths_features_d_labeled_new$route) 
table(sce$route)


# table of right x left eyes route
split_eye_routes <-
  eye_paths_features_d_labeled_new %>% 
  select(patID_eye, route) %>%
  mutate(
    patID = sapply(str_split(patID_eye, "_"), `[`, 1),
    eye = sapply(str_split(patID_eye, "_"), `[`, 2) ) %>% 
  mutate(eye = ifelse( eye == "1", "right", "left")) %>% 
  na.omit() %>% 
  pivot_wider(id_cols = patID, values_from = route, names_from = eye)


t <- table(split_eye_routes$right,split_eye_routes$left)
t <- t %>% 
  cbind(rowSums(t))
t <- t %>% 
  rbind(colSums(t))

# count eyes with rf_route
sce$patID_eye[!is.na(sce$rf_route)] %>% n_distinct()
eye_paths_features_d_labeled_new$patID_eye[!is.na(eye_paths_features_d_labeled_new$rf_route)] %>% n_distinct()


# table of right x left eyes rf_route
split_eye_rf_routes <-
  eye_paths_features_d_labeled_new %>% 
  select(patID_eye, rf_route) %>%
  mutate(
    patID = sapply(str_split(patID_eye, "_"), `[`, 1),
    eye = sapply(str_split(patID_eye, "_"), `[`, 2) ) %>% 
  mutate(eye = ifelse( eye == "1", "right", "left")) %>% 
  # na.omit() %>% 
  pivot_wider(id_cols = patID, values_from = rf_route, names_from = eye)

nrow(split_eye_rf_routes)
t <- table(split_eye_rf_routes$left,split_eye_rf_routes$right, useNA = "no") # useNA = "ifany")
t <- t %>% 
  cbind(rowSums(t))
t <- t %>% 
  rbind(colSums(t))
cat("RF_route counts")
t


split_eye_rf_routes %>% 
  filter(if_any(everything(), ~is.na(.)))
colData(sce)[sce$patID == "002079" , ] %>% 
  as.data.frame() %>% 
  select(rf_route, my_cluster)


```

# ____________________
# Look at vars important route classification
```{r}

response <- "route"
assay <- assays(sce)[["pmm_imp_of_scaled"]] %>% 
      t() %>% 
      as.data.frame() %>% 
      rownames_to_column("sample") %>% 
      left_join(
        colData(sce) %>% 
          data.frame() %>% 
          select(sample, all_of(response))
      ) %>% 
  column_to_rownames("sample") %>% 
  drop_na(all_of(response))

pacman::p_load(shapviz, xgboost)
params <- list(objective = "multi:softprob", num_class = n_distinct(assay[[response]]), nthread = 2)
dtrain <- xgb.DMatrix( data.matrix(select(assay, -all_of(response))) , 
                       label = as.integer(assay[[response]]) - 1)

fit <- xgb.train(params = params, data = dtrain, nrounds = 50)

# Create "mshapviz" object (logit scale)
shapvis <- shapviz(fit, X_pred = data.matrix(select(assay,  -all_of(response))), X = assay)

# # Contains "shapviz" objects for all classes
# all.equal(x[[3]], shapviz(fit, X_pred = data.matrix(select(assay, -rf_route)), X = assay, which_class =  n_distinct(assay$rf_route) ))

# Better names
names(shapvis) <- levels(assay[[response]])
shapvis
#> 'mshapviz' object representing 3 'shapviz' objects:
#>    'setosa': 150 x 4 SHAP matrix
#>    'versicolor': 150 x 4 SHAP matrix
#>    'virginica': 150 x 4 SHAP matrix

beeswarm <- sv_importance(shapvis, max_display = 20, kind = "beeswarm")
ggsave(paste0(wd_path, "results/clustering/12_clustering/rf_route_var_shaps_beeswarm.png"),beeswarm, height = 10, width = 20)

```
# ____________________


## plot path progression node x step
```{r}

# params
h = 10;w = 10

step_node <- eye_paths_features_d_labeled_new %>% 
  select(patID_eye, matches("_route|^route"), matches("step")) %>% 
  pivot_longer(matches("step"), names_to = "step", values_to = "node") %>% 
  mutate(step = gsub("step", "", step) %>% as.numeric()) %>% 
  # mutate(node = factor(node)) %>%
  mutate(across(c(step, node), ~factor(.))) %>%
  arrange(patID_eye, step) 

# plot step x node colored by patID_eye, line connecting them
step_node %>% 
  # ggplot(aes(step, node, color = patID_eye, group = patID_eye))+
  ggplot(aes(step, node,  group = patID_eye))+
  geom_hline(yintercept = factor(c(1:5, 333, 444)),  linetype = 3)+
  geom_vline(xintercept = factor(c(1:10)),  linetype = 3)+
  # geom_point(show.legend = FALSE)+
  geom_line(show.legend = FALSE, position = position_dodge(width = .3), size = .4, alpha = .4)+
  # geom_smooth(aes(group = 1), method = "loess" , show.legend = FALSE)+
  ggtitle("Eye progression from cluster to cluster\nsimplified termini")
ggsave(paste0(wd_path, "results/clustering/12_clustering/eye_progression_plot.png"), height = h, width = w)

# plot step x node colored by patID_eye, line connecting them; filter for last node not same or next node not same
step_node %>% 
  group_by(patID_eye)  %>% 
  filter(node != lead(node) | lag(node) != node) %>% 
  # ggplot(aes(step, node, color = patID_eye, group = patID_eye))+
  ggplot(aes(step, node,  group = patID_eye))+
  geom_hline(yintercept = factor(c(1:5, 333, 444)),  linetype = 3)+
  geom_vline(xintercept = factor(c(1:10)),  linetype = 3)+
  # geom_point(show.legend = FALSE)+
  geom_line(show.legend = FALSE, position = position_dodge(width = .3), size = .4, alpha = .4)+
  # geom_smooth(aes(group = 1), method = "loess" , show.legend = FALSE)+
  ggtitle("Eye progression from cluster to cluster\nsimplified termini\nrecords whose next node or previous node is not the same")
ggsave(paste0(wd_path, "results/clustering/12_clustering/filter_lag_lead_eye_progression_plot.png"), height = h, width = w)

# plot step x node colored by route, line connecting patID_eyes
step_node %>% 
  group_by(patID_eye)  %>% 
  filter(node != lead(node) | lag(node) != node) %>%
  filter(!is.na(route)) %>% 
  
  ggplot(aes(step, node, color = route, group = patID_eye))+
  geom_hline(yintercept = factor(c(1:5, 333, 444)),  linetype = 3)+
  geom_vline(xintercept = factor(c(1:10)),  linetype = 3)+
  # geom_point(show.legend = FALSE)+
  geom_line(show.legend = TRUE, position = position_dodge(width = .3), size =1 , alpha = .6)+
  scale_color_manual(values = c("a" = "red",
                                "b" = "blue",
                                "c" = "orange", 
                                "d"= "purple",
                                "E" = "green",
                                "F"= "grey")
                     )+
  ggtitle("Eye progression from cluster to cluster\nsimplified termini\nrecords whose next node or previous node is not the same")
ggsave(paste0(wd_path, "results/clustering/12_clustering/filter_lag_lead_route_eye_progression_plot.png"), height = h, width = w)

```




#_________________
# Parse paths
## plot eye paths
```{r, eval = FALSE}

# FILTER DATA ----
# exclude eyes with only one visit and it's in 1,2,3,or 5
tsne <- tsne %>% 
  left_join(colData(sce) %>% as.data.frame()) 

# # exclude eyes that only stay in one cluster
# f.tsne <- f.tsne %>% 
#   group_by(patID_eye) %>% 
#   filter( 
#     n_distinct(my_cluster) > 1
#     ) 

# f.tsne %>% 
#   filter(!is.na(rf_route)) %>% 
#   ggplot(aes(TSNE1, TSNE2, color = pseudoprog)) +
#   geom_point()+
#   scale_color_viridis_c()+
#   facet_wrap(~rf_route)
# 
# f.tsne %>% 
#   filter(!is.na(logic_route)) %>% 
#   ggplot(aes(TSNE1, TSNE2, color = pseudoprog)) +
#   geom_point()+
#   scale_color_viridis_c()+
#   facet_wrap(~logic_route)
# 
# f.tsne %>% 
#   filter(!is.na(rf_2_route)) %>% 
#   ggplot(aes(TSNE1, TSNE2, color = pseudoprog)) +
#   geom_point()+
#   scale_color_viridis_c()+
#   facet_wrap(~rf_2_route)

# filter for all routes present
f.tsne <- tsne %>% 
  filter(if_all(matches("route"), ~!is.na(.))) 

# non-singleton patients and non missing of all routes
repeat_patients <- f.tsne %>% 
  arrange(patID_eye, month) %>% 
  group_by(patID_eye) %>% 
  filter(n() > 1) %>% 
  pull(patID_eye) %>% 
  unique()

# patients with greatest pseudoprog change
patient_pt_delta <- f.tsne %>% 
  filter(patID_eye %in% repeat_patients) %>% 
  group_by(patID_eye) %>% 
  arrange(pseudoprog) %>% 
  summarise(pseudotime_delta = dplyr::last(pseudoprog) - dplyr::first(pseudoprog), .groups = "drop") %>% 
  arrange(-pseudotime_delta)

# top pt delta patients 
top_eyes <- patient_pt_delta %>% 
  slice_max(pseudotime_delta, n = 30) %>% 
  pull(patID_eye)

## which route system to use in plotting paths
which_route <- "route"
# which_route <- "rf_route"
# which_route <- "rf_2_route"

# filter tsne for sample of top eyes and where which_route not missing
sub_tsne <- f.tsne %>% 
  filter(patID_eye %in% top_eyes) %>%
  ungroup()


# # PLOT EYE PATHS----
# # edges connecting eyes
# tsne_segments <- sub_tsne %>% 
#   group_by(patID_eye) %>% 
#   arrange(month) %>%  
#   summarise(
#     TSNE1 = list(TSNE1),
#     TSNE2 = list(TSNE2),
#     .groups = 'drop'
#   )
# 
# # Flatten the list-columns to use in geom_path
# tsne_flat <- sub_tsne %>% 
#   select(sample, patID_eye,  TSNE1:TSNE2)
# 
# # colors <- viridis(n = n_distinct(sub_tsne$year))
# # names(colors) <- unique(sub_tsne$year)
# # colors_vec <- colors[match(sub_tsne$year, names(colors) )]
# 
# colors <-c( "blue", "red", "green", "purple", "pink")
# colors <- colors[1:n_distinct(sub_tsne[[which_route]])]
# names(colors) <- unique(sub_tsne[[which_route]])
# colors_vec <- colors[match(sub_tsne[[which_route]], names(colors) )]
#   
# tsne_paths <- 
#   sub_tsne %>% 
#   ggplot() +
#   geom_point(data = f.tsne , aes(TSNE1, TSNE2),alpha = .2, color = "grey", size = 2)+
#   geom_path(
#     data = tsne_flat,
#     aes(x = TSNE1, y = TSNE2, group = patID_eye, color = patID_eye), 
#     inherit.aes = FALSE, show.legend = FALSE, arrow = arrow(length = unit(0.2, "cm")), 
#     linewidth = .5, linetype = 9
#   ) +
#   scale_color_manual(values = colors_vec) +
#   geom_point(aes(x = TSNE1, y = TSNE2, color  = factor(get(which_route)), shape  = factor(get(which_route))), show.legend = TRUE, size = 3) +
#   theme_minimal()+
#   ggtitle(which_route)
# 
#   
# tsne_path_plotly <- ggplotly(tsne_paths) %>% 
#   add_annotations(
#     x = ~sub_tsne$TSNE1,
#     y = ~sub_tsne$TSNE2,
#     text = ~sub_tsne$path,
#     showarrow = TRUE,
#     arrowhead = 2,
#     ax = 20,
#     ay = -40, 
#     font = list(size = 8)
#   )
# 
# htmlwidgets::saveWidget(tsne_path_plotly,paste0(wd_path, "results/clustering/12_clustering/tsne_", which_route, "_path_plot.html"))


```

<!-- ## plot eye velocities -->
<!-- ```{r, eval  = FALSE} -->
<!-- # velo.out <- scvelo(sce, assay.spliced = "pmm_imp_of_scaled", assay.unspliced = "pmm_imp_of_scaled",assay.X="pmm_imp_of_scaled", use.dimred="PCA") -->
<!-- # velo.out -->
<!-- #  -->
<!-- # sce.sperm$pseudoprog <- velo.out$velocity_pseudotime -->
<!-- #  -->
<!-- # # Also embedding the velocity vectors, for some verisimilitude. -->
<!-- # embedded <- embedVelocity(reducedDim(sce.sperm, "TSNE"), velo.out) -->
<!-- # grid.df <- gridVectors(reducedDim(sce.sperm, "TSNE"), embedded, resolution=30) -->
<!-- #  -->
<!-- # plotTSNE(sce.sperm, colour_by="pseudoprog", point_alpha=0.3) + -->
<!-- #     geom_segment(data=grid.df,  -->
<!-- #         mapping=aes(x=start.1, y=start.2, xend=end.1, yend=end.2),  -->
<!-- #         arrow=arrow(length=unit(0.05, "inches"), type="closed")) -->

<!-- sce <- slingshot(sce, clusterLabels = "my_cluster", reducedDim = "TSNE") -->

<!-- tsne_df <- reducedDim(sce, "TSNE") %>%  -->
<!--   as.data.frame() %>%  -->
<!--   cbind(as.data.frame( colData(sce)$pseudoprog)) -->

<!-- lineages <- slingLineages(sce) -->

<!-- # max_length <- sapply(lineages, length) %>% max -->
<!-- # lineages_new <- lapply(lineages, function(a_vec){ -->
<!-- #    -->
<!-- #   if(length(a_vec) <max_length){ -->
<!-- #      -->
<!-- #     a_vec <- c(a_vec, rep(a_vec[length(a_vec)], max_length - length(a_vec))) -->
<!-- #   } -->
<!-- #    -->
<!-- #   return(a_vec) -->
<!-- # }) -->
<!-- #  -->
<!-- # lineages_d <- as.data.frame(lineages_new) -->


<!-- #  -->
<!-- # ggplot(tsne_df, aes(x = TSNE1, y = TSNE2, color = pseudoprog)) + -->
<!-- #   geom_point() + -->
<!-- #   scale_color_viridis_c() + -->
<!-- #   theme_minimal() + -->
<!-- #   labs(title = "t-SNE Plot with pseudoprog", -->
<!-- #        x = "t-SNE 1", -->
<!-- #        y = "t-SNE 2", -->
<!-- #        color = "pseudoprog") + -->
<!-- #   geom_path(data = lineages_d, aes(x = TSNE1, y = TSNE2, group = Lineage), color = "black", size = 1, alpha = 0.8) -->


<!-- sling_curves <- slingCurves(sce) -->

<!-- # Initialize an empty dataframe to store lineage points -->
<!-- lineage_df <- data.frame(TSNE1 = numeric(0), TSNE2 = numeric(0), Lineage = factor()) -->

<!-- # Loop through each lineage and extract the curve points -->
<!-- for (i in seq_along(sling_curves)) { -->
<!--   curve <- sling_curves[[i]] -->
<!--   curve_df <- data.frame(TSNE1 = curve$s[,1], TSNE2 = curve$s[,2], Lineage = as.factor(i)) -->
<!--   lineage_df <- rbind(lineage_df, curve_df) -->
<!-- } -->

<!-- ggplot(tsne_df, aes(x = TSNE1, y = TSNE2, color = pseudoprog)) + -->
<!--   geom_point() + -->
<!--   scale_color_viridis_c() + -->
<!--   theme_minimal() + -->
<!--   labs(title = "t-SNE Plot with pseudoprog and Trajectories", -->
<!--        x = "t-SNE 1", -->
<!--        y = "t-SNE 2", -->
<!--        color = "pseudoprog") + -->
<!--   geom_path(data = lineage_df, aes(x = TSNE1, y = TSNE2, group = Lineage), color = "black", size = 1, alpha = 0.8) -->


<!-- ``` -->
### plot eye 'velocities'
```{r, eval = FALSE}

df <-
  # tsne %>%
  # f.tsne %>%
  sub_tsne %>%
  arrange(patID_eye, month) %>%
  group_by(patID_eye) %>%
  mutate(
    TSNE1_next = lead(TSNE1),
    TSNE2_next = lead(TSNE2)
  ) %>%
  filter(!is.na(TSNE1_next) & !is.na(TSNE2_next)) %>%
  ungroup()

xrange <-  df$TSNE1 %>% range()
yrange <-  df$TSNE2 %>% range()
plot_velocities <- function(df, which_route = "route"){
  
  ggplot() +
  geom_point(data = tsne, aes(TSNE1, TSNE2), color = "grey", alpha = .3)+
  geom_segment(data = df, aes(x = TSNE1, y = TSNE2, xend = TSNE1_next, yend = TSNE2_next, color = pseudoprog, group = patID_eye),
               arrow = arrow(length = unit(0.2, "cm")), size = 1) +
  geom_label_repel(data = df, aes( x = TSNE1, y = TSNE2, label = ifelse(pseudoprog == min(pseudoprog), paste0(patID_eye,": \n" ,path ), NA)),size = 4 )+
  scale_color_viridis_c()+
  geom_point(data = df, aes(x = TSNE1, y = TSNE2, shape =get(which_route), group = patID_eye), size = 2) +
  theme_minimal() +
  labs(
    title = paste("t-SNE with Velocity Vectors;", which_route),
    x = "t-SNE 1",
    y = "t-SNE 2",
    color = "pseudoprog"
  )+ 
  lims(x = xrange, y = yrange)
  
}

df %>% 
  # filter(rf_route %in% c("a")) %>%
  # filter(rf_route %in% c("b")) %>%
  # filter(rf_route %in% c("c")) %>%
  # filter(rf_route %in% c("d")) %>%
  filter(rf_route %in% c("x")) %>%
  # filter(route %in% c("a", "c")) %>%
  # filter(route %in% c("b", "d")) %>%
  filter(patID_eye ==  sample(unique(patID_eye), 1)) %>% 
  plot_velocities(., which_route = "route")

```

## do patients' eyes follow the same route? 

<!-- ### count of eyes that stick to one route versus eyes that take >1 -->
<!-- ```{r} -->

<!-- library(scales) -->

<!-- count_distinct_routes <- tsne %>%  -->
<!--   filter(!is.na(rf_route)) %>%  -->
<!--   group_by(patID) %>%  -->
<!--   summarise(n_distinct_routes = n_distinct(rf_route)) -->

<!-- count_distinct_routes %>% -->
<!--   ggplot(aes(factor(n_distinct_routes))) + -->
<!--   geom_bar(stat = "count") + -->
<!--   geom_text(stat = "count", aes(label = after_stat(count)+10, y = after_stat(count)+10)) + -->
<!--   scale_y_continuous(breaks = pretty_breaks(15))+ -->
<!--   ggtitle("Count patients with 1 and 2 routes")+ -->
<!--   theme_bw() -->

<!-- distinct_routes_t <- table(count_distinct_routes$n_distinct_routes ) -->
<!-- # distinct_routes_t[2]/distinct_routes_t[1] -->
<!-- distinct_routes_t -->

<!-- ``` -->

<!-- ### paths of patients that have >1 route and 1 route  -->
<!-- ```{r} -->

<!-- par(mfrow= c(1,2),mar = c(10,2,2,2)) -->
<!-- tsne %>%  -->
<!--   filter(patID %in% count_distinct_routes$patID[count_distinct_routes$n_distinct_routes>1]) %>%  -->
<!--   select(patID, patID_eye, month, pseudoprog, rf_route_score,route_score,  my_cluster, path) %>%  -->
<!--   arrange(patID_eye, month) %>%  -->
<!--   filter(abs(rf_route_score) >1) %>%  -->
<!--   pull(path) %>%  -->
<!--   table() %>% -->
<!--   sort() %>%  -->
<!--   barplot(las= 2, main = "Eye paths of patients with > 1 route" ) -->

<!-- tsne %>%  -->
<!--   filter(patID %in% count_distinct_routes$patID[count_distinct_routes$n_distinct_routes==1]) %>%  -->
<!--   select(patID, patID_eye, month, pseudoprog, rf_route_score,route_score,  my_cluster, path) %>%  -->
<!--   arrange(patID_eye, month) %>%  -->
<!--   filter(abs(rf_route_score) >1) %>%  -->
<!--   pull(path) %>%  -->
<!--   table() %>% -->
<!--   sort() %>%  -->
<!--   barplot(las= 2, main = "Eye paths of patients with 1 route" ) -->
<!-- par(mar = c(2,2,2,2), mfrow= c(1,1)) -->

<!-- ``` -->



#=========================
# SAVE DATA
```{r, eval = TRUE}

# save sce ----
saveHDF5SummarizedExperiment(sce, paste0(wd_path, "processed_data/sce005.h5"), replace = T)
notify.me(paste("12_clusterExploration Completed"))
how.long()


```

```{r check dims}

check_dims()
# dims : 12525 176
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ added route vars: scores, routes, paths, ...

### APRIL 10 2025 ###
# dims : 12276 205
# n_patIDs             3971
# n_patID_eyes         7888
# n_observations      12275
# n_patIDs_w_gr_1_vis   492


```

