---
title: "13.2_T2D_clusterModeling"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---

# Packages
```{r}

library(tidyr)
library(reshape2)
library(purrr)
library(GGally)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(gridExtra)
library(ComplexHeatmap)
library(viridis)
library(zoo)
library(plotly)
library(mgcv)
library(tools)
library(mice)
library(ggrepel)
library(htmltools)
library(Metrics)
library(psych)
library(car)
library(stats)
library(lme4)
library(lmerTest)
library(ape)
library(SingleCellExperiment)
library(scater)
library(dplyr)
library(HDF5Array)
library(parallel)
library(emmeans)
library(gplots)
library(limma)
library(optimx)

rm(list = ls())
gc()

options(max.print = 100)

```

### functions
```{r}

source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```

# Load Data
```{r}


wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

a_path <- paste0(wd_path, "results/clustering/13_clustering/T2D_Contrasts/")
if(!dir.exists(a_path)) {
  dir.create(a_path, recursive = TRUE)
  
}
# load experiment
sce <- HDF5Array::loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/sce006.h5"))


```


#__________________________
# does t2d predict cluster transition?

## make model data
```{r}

# df of subjects transitions (N-1), clust to next clust
transitions <-
  as.data.frame(colData(sce)) %>% 
  select(subject, patID_eye, month, clust = km_nn_cluster, t2d) %>% 
  arrange(patID_eye, month) %>% 
  group_by(patID_eye) %>%
  mutate(lead_clust = lead(clust)) %>% 
  filter(!is.na(clust)) %>% 
  filter(!is.na(lead_clust)) %>% 
  mutate(
    transition = paste0("X.",clust, "to", lead_clust)
  ) %>% 
  filter(clust!=lead_clust) %>% 
  ungroup()

# factor transitions in order of clusters
transition_levels <- unique(transitions$transition[order(transitions$clust,transitions$lead_clust)] )
transitions$transition <- factor(transitions$transition, levels = transition_levels )

# summarize t2d prevalence, etc.
transitions_stats <- transitions %>% 
  group_by(transition) %>% 
  reframe(odds_t2d = sum(t2d == 1, na.rm = TRUE)/sum(t2d == 0, na.rm = TRUE), 
          ratio_t2d = sum(t2d == 1, na.rm = TRUE)/n(), 
          frequency = n()/ nrow(.)) %>% 
  # mutate(log_odds_t2d = log(odds_t2d)) %>% 
  filter(if_all(everything(), ~is.finite(.))) %>% 
  arrange(desc(odds_t2d)) 

# fits <- list(
#   `ratio_t2d ~ sqrt(odds_t2d)` = lm(ratio_t2d ~ sqrt(odds_t2d), data = transitions_stats),
#   `ratio_t2d ~ sqrt(odds_t2d+1)` = lm(ratio_t2d ~ sqrt(odds_t2d), data = transitions_stats),
#   `ratio_t2d ~ log(odds_t2d+1)` = lm(ratio_t2d ~ log(odds_t2d+1), data = transitions_stats),
#   `odds_t2d ~ ratio_t2d^3`  = lm(odds_t2d ~ I(ratio_t2d^3), data = transitions_stats)) 
# lapply(fits,summary)
# sapply(fits, AIC) %>% data.matrix()
# sapply(fits, BIC)%>% data.matrix()

transitions_stats_plot <- ggplot(transitions_stats,aes(odds_t2d, ratio_t2d))+
  geom_smooth( aes(group= 1,linetype = "y ~ log(x+1)"),method ="lm", formula = y ~ log(x+1), show.legend = TRUE)+
  scale_linetype_manual(
    name = "Fit",
    values = c("y ~ log(x+1)" = "solid")
  )+
  geom_abline(slope = 1, intercept = 0, linetype = 3)+
  geom_point(show.legend = FALSE, aes(color = transition))+
  geom_label_repel(show.legend = FALSE, aes(color = transition, label = transition))+
  theme_bw()+
  guides(color = "none")+
  geom_blank()
ggsave(paste0(wd_path, "results/clustering/13_clustering/T2D_Contrasts/transitions_stats.png"),transitions_stats_plot, height = 10, width = 12)

# make transitions dichotomous 
transitions_wide <-
  transitions %>%
  mutate(value = 1) %>% 
  pivot_wider(
    id_cols = c(patID_eye, month, t2d),  # Keep `patID_eye` as rows
    names_from = transition, # Each unique transition becomes a column
    values_from = value, 
    values_fill = 0,
  ) %>% 
  # mutate(across(-c(patID_eye, month), ~as.numeric(scale(.)) )) %>% 
  ungroup()


formulas <- paste0(
  unique(grep("to", names(transitions_wide), value = TRUE)),
  " ~ t2d"
  # ," + (1|patID_eye)"
)  %>%
  sapply(as.formula)

mods <- mclapply(formulas, function(form){
  mod <- try({
    glm(form, data = transitions_wide,family = binomial)
    # glmer(form, data = transitions_wide, family = "binomial") use diff optimization alg
  })
  
  if(inherits(mod, "try-error")) mod <- "ERROR" 
  
  return(mod)
  
}, mc.cores = detectCores()-1)

summs <- mods %>% map(summary)
summs <- summs[sapply(summs, function(x) sum(grepl("character", x)) != 2)]


results <- extract_results(summs) %>% 
  filter(!grepl("Intercept", Term))

results <- results %>% 
  filter(abs(Estimate) <5)

ggplot(results , aes(Estimate, -log10(`Pr(>|z|)`), color = Response))+
  geom_point(show.legend = FALSE)+
  geom_label_repel(aes(label = Response), show.legend = FALSE, max.overlaps = 100)+
  geom_hline(yintercept = -log10(.05), color = "grey", linetype = 2)+
  geom_vline(xintercept = 0, color = "grey", linetype = 2)+
  ggtitle("Transitions ~ T2D")+
  theme_bw()
ggsave(paste0(wd_path, "results/clustering/13_clustering/T2D_Contrasts/transitions_by_t2d.png"), height = 10, width = 12)


```

* model with Route
* if eye in cluster in route
* by cluster t2d ~ dicotomous clsuter
** then agglomorate to route 

* fit prevalence of t2d across groups (clusters)
** model matrix with no intercept, not change, or odds, prevalence
** differences in the average of teh averages 
* on LM 

### does t2d predict presence in cluster
```{r}

# dichotomize cluster
clusters <- 
  as.data.frame(colData(sce)) %>% 
  select(patID, patID_eye, month, clust = km_nn_cluster, t2d, BMI)

clusters_wide <- clusters %>% 
  mutate(value = 1) %>% 
  pivot_wider(
    id_cols = c(patID_eye, month, t2d), 
    names_from = clust,
    names_prefix = "clust_",
    values_from = value, 
    values_fill = 0
  ) 

# formulas: cluster  ~ diabetes

forms <- paste0(
  grep("clust", names(clusters_wide), value = TRUE), 
  " ~ t2d"
) %>% 
  sapply(as.formula)

mod_type <- "lm"
# mod_type <- "glm"

fits <- mclapply(forms, function(form){
  if(mod_type == "lm"){
    fit <- try({
      lm(form, data = clusters_wide)
    })
  } else if(mod_type == "glm"){
    fit <- try({
      glm(form, data = clusters_wide, family = binomial)
    })
  }
  if(inherits(fit, "try-error")) fit <- "AAHH! ERROR!"
  return(fit)
}, mc.cores =  detectCores()-1)

summs <- lapply(fits, summary)

results <- extract_results(summs) %>% 
  filter(!grepl("Intercept", Term))

names(results)[ grep("Pr\\(>", names(results)) ] <- "pval"

ggplot(results , aes(Estimate, -log10(q_value), color = Response))+
  geom_point(show.legend = FALSE)+
  geom_label_repel(aes(label = Response), show.legend = FALSE, max.overlaps = 100)+
  geom_hline(yintercept = -log10(.05), color = "grey", linetype = 2)+
  geom_vline(xintercept = 0, color = "grey", linetype = 2)+
  ggtitle("Cluster ~ T2D", mod_type)+
  theme_bw()
ggsave(paste0(wd_path, "results/clustering/13_clustering/T2D_Contrasts/cluster_by_t2d.png"), height = 10, width = 12)

```

# Contrasts of cluster~t2d

## Contrasts practice
```{r}


Data = read.table(header=TRUE, stringsAsFactors=TRUE, 
        text="Treatment Response Type 
               Merlot 5 Red
               Merlot 6 Red
               Merlot 7 Red
               Cabernet 8 Red
               Cabernet 9 Red
               Cabernet 10  Red
               Syrah  11  Red
               Syrah  12  Red
               Syrah  13  Red
               Chardonnay 1 White
               Chardonnay 2 White
               Chardonnay 3 White
               Riesling 1 White
               Riesling 2 White
               Riesling 2 White
               Gewürtztraminer  1 White
               Gewürtztraminer  2 White
               Gewürtztraminer  4 White"
        )

Data$Treatment = factor(Data$Treatment, levels = unique(Data$Treatment))
Data$Type = factor(Data$Type)

boxplot(Response ~ Treatment,data = Data,ylab="Response",xlab="Treatment")
boxplot(Response ~ Type,data = Data,ylab="Response",xlab="Type")


###  Define linear model
model = lm(Response ~ Treatment,data = Data)
# x_model = lm(Response ~ Treatment:Type,data = Data)
# summary(x_model)

library(car)

Anova(model, type="II")
summary(model)

Treatment_means <- aggregate(Response ~ Treatment, mean, data = Data)
Type_means <- aggregate(Response ~ Type, mean, data = Data)
marginal =  emmeans::emmeans(model, ~ Treatment)

levels(Data$Treatment)
# "Merlot"          "Cabernet"        "Syrah"           "Chardonnay"      "Riesling"        "Gewürtztraminer"

# IS DIFFERENCE BETWEEN REDS SIGNIFICANLTY DIFFERENT? ----
Contrasts = list(R1   = c(1, -1,  0,  0,  0,  0),
                 R2   = c(0,  1, -1,  0,  0,  0))
### The column names match the order of levels of the treatment variable
### The coefficients of each row sum to 0

(Test = emmeans::contrast(marginal, Contrasts))
emmeans::test(Test, joint=TRUE)
# df1 df2 F.ratio p.value
#   2  12  42.100  <.0001
## ***P at least one of the contrasts has a difference in response means != 0

### Note that two lines of contrasts resulted in one hypothesis test
###   using 2 degrees of freedom. This investigated the effect within
###   a group of 3 treatments.
### Results are essentially the same as those from multcomp.

# IS DIFFERENCE BETWEEN WHITES SIGNIFICANLTY DIFFERENT? ----
Contrasts = list(W1   = c(0, 0,  0,  1,  -1,  0),
                 W2   = c(0,  0, 0,  0,  1,  -1))
(Test =emmeans::contrast(marginal, Contrasts))
emmeans::test(Test, joint=TRUE)
# df1 df2 F.ratio p.value
#   2  12  72.700  <.0001


# IS THERE DIFFERENCE BETWEEN RED AND WHITE? ----
## not averaging each group average
Contrasts = list(RvW   = c(1, 1,  1,  -1,  -1,  -1))
(Test = emmeans::contrast(marginal, Contrasts))
# contrast estimate   SE df t.ratio p.value
# RvW            21 1.49 12  14.087  <.0001

emmeans::test(Test, joint=FALSE)
# contrast estimate   SE df t.ratio p.value
# RvW            21 1.49 12  14.087  <.0001
## ^ same output because only one contrast

## averaging each group average
### red mean - white mean
Contrasts = list(RvW   = c(1/3, 1/3,  1/3,  -1/3,  -1/3,  -1/3))
(Test = emmeans::contrast(marginal, Contrasts))
# contrast estimate    SE df t.ratio p.value
#  RvW             7 0.497 12  14.087  <.0001
emmeans::test(Test, joint=FALSE)
# contrast estimate    SE df t.ratio p.value
#  RvW             7 0.497 12  14.087  <.0001


```

## is mean t2d of left route significantly different from mean of right route?
### LM contrasts
```{r}
# treatment contrasts; where each estimate is difference in mean from reference (cluster 1)
form_char <- "t2d~clust"
form <- as.formula(form_char)
# avgs_lm <- lm(form, data = clusters)
avgs_lm <- glm(form, data = clusters, family = "binomial")
summary(avgs_lm)

# coefficients are difference in mean from intecept 
barplot(sort(avgs_lm$coefficients[2:10]+avgs_lm$coefficients[1]), main = "Cluster T2D prevalence (model)")
clust_means <- aggregate(t2d~clust, mean, data = clusters)
barplot( clust_means$t2d[order(clust_means$t2d)], 
         names.arg  = paste0("clust",clust_means$clust[order(clust_means$t2d)]),
         main = "Cluster T2D prevalence")
         
# ad <- data.frame(letter=c(rep("a",30),rep("b",30)),y=c(rnorm(30),rnorm(30)+0.6) )
# 
# summary(lm(d$y~d$letter))
# t.test(d$y~d$letter,var.equal	=T)
# 
# wilcox.test(d$y~d$letter)
# wilcox.test(d,paired = FALSE )
# 
# 
# Anova(d$y~d$letter)




design <- model.matrix(~0+clust, data = clusters)
# # pink is left route (3-(11,9,7)), yellow is right route (4, ...)
# contrast.matrix <- makeContrasts(pVy=((clust3+clust7+clust9+clust11)/4)-((clust4+clust6+clust8+clust10)/4),
#                                  PvY = ((clust1+clust2+clust3+clust7+clust9+clust11)/6)-((clust1+clust2+clust4+clust6+clust8+clust10)/6),
#                                  levels=design)
# contrasts = apply(contrast.matrix,2, `[`, simplify = FALSE)

levels(clusters$clust)
# [1] "1"  "2"  "3"  "4"  "5"  "6"  "7"  "8"  "9"  "10" "11"

# estimated marginal means (estimated response means for each group in given model)
marginal = emmeans(avgs_lm, ~ clust, data = clusters)

# c(0,0,0,0,0,0,0,0,0,0,0)
# contrasts; averaged response average
Contrasts <- list(
  # left and right routes unique nodes; normalized to number of nodes in each route 
  L.v.R = c(0,0,1,-1,0,-1,1,-1,1,-1,1)/4, 
  # end nodes; normalized to number of nodes in each route 
  endL.v.endR = c(0,0,0,0,0,-1,1,-1,1,-1,1)/3, 
  # average of end nodes of Left route versus cluster 3
  three.v.endL = c(0,0,1,0,0,0,-1/3,0,-1/3,0,-1/3),
  # average of end nodes of right route versus cluster 4
  four.v.endR =  c(0,0,0,1,0,-1/3,0,-1/3,0,-1/3,0),
  # 5 versus everywhere else; average else
  five.v.else = c(-1/10,-1/10,-1/10,-1/10,1,-1/10,-1/10,-1/10,-1/10,-1/10,-1/10)
)
# cbind(cluster = as.numeric(levels(clusters$clust)),weight= C$L.v.R)

# contrasts model
(Test <- contrast(marginal, Contrasts))
## contrast     estimate     SE    df t.ratio p.value
##  L.v.R          -0.167 0.0245 11843  -6.828  <.0001
##  endL.v.endR    -0.174 0.0324 11843  -5.383  <.0001
##  THREE.v.FOUR   -0.146 0.0122 11843 -11.928  <.0001
### right has significantly greater average than left in three categories

# test(contrast(marginal, C),joint = TRUE )

# # all pairwise contrasts
# pair_contrasts <- emmeans(avgs_lm, pairwise ~ clust, data = clusters, adjust=NULL)
results <- 
  as.data.frame(Test) %>% 
  # rbind(as.data.frame(pair_contrasts$contrasts)) %>%
  mutate(q_value = p.adjust(p.value, method = "BH")) %>% 
  mutate(contrast = gsub("\\.v\\.", " - ", contrast))


ggplot(results, aes(estimate, -log10(q_value)))+
  geom_vline(xintercept = 0, linetype = 2)+
  geom_hline(yintercept = -log10(.05), linetype = 2)+
  geom_point(aes(color = contrast), show.legend = FALSE)+
  geom_label_repel(aes(label = contrast,color = contrast), show.legend = FALSE)+
  ggtitle("Cluster T2D contrasts",form_char)+
  theme_bw()
ggsave(paste0(wd_path, "results/clustering/13_clustering/T2D_Contrasts/cluster_t2d_contrasts.png"), height = 10, width = 12)
```


### LMER Contrasts
```{r}

# fit t2d on clusters  ( adding eye term has no benefit)
lmer_fit <- lmer(t2d~clust + (1|patID),
                data = clusters ,
                # control = lmerControl(optimizer ="Nelder_Mead")
                # control = lmerControl(optimizer ='optimx', optCtrl=list(method='nlminb'))
                control = lmerControl(optimizer ='optimx', optCtrl=list(method='L-BFGS-B')) # smalleds re sd residual
                )
summary(lmer_fit)

# estimated marginal means (estimated response means for each group in given model)
marginal = emmeans(lmer_fit, ~ clust, data = clusters)

# contrasts model
contrast(marginal, Contrasts)

```

### GLMER contrasts
```{r, eval = FALSE}

# ****** FUNCTION to run GLM model 
glmer_model <- function(){
    glmer_fit <- glmer(
        factor(t2d)~ clust + (1|patID),
        data = clusters,
        family = binomial,
        # control = glmerControl(optimizer ="Nelder_Mead")
        # control = glmerControl(optimizer ='optimx', optCtrl=list(method='nlminb'))
        control = glmerControl(optimizer ='optimx', optCtrl=list(method='L-BFGS-B')) # smalleds re sd residual; THIS ONE WORKS!
        )
    
    # estimated marginal means (estimated response means for each group in given model)
    glmer_marginal = emmeans(glmer_fit, ~ clust, data = clusters)
    
    # contrasts model
    glmer_contrasts <- as.data.frame(contrast(glmer_marginal, Contrasts))
    
    plot <- ggplot(glmer_contrasts,aes(estimate,-log10(p.value), label = contrast, color = contrast))+
      geom_point(show.legend =FALSE)+
      geom_label_repel(show.legend =FALSE)+
      geom_vline(xintercept = 0, linetype = 2)+
      geom_hline(yintercept = -log10(.05), linetype = 2)+
      ggtitle("GLMER t2d~clust+(1|patID)")+
      theme_bw()
    ggsave(paste0(wd_path, "results/clustering/13_clustering/T2D_Contrasts/cluster_t2d_contrasts_glmer.png"), plot, height = 10, width = 12)
    
     return(list(glmer_fit = glmer_fit, glmer_marginal = glmer_marginal, glmer_contrasts = glmer_contrasts, plot =  plot))
}


jb <- mcparallel( glmer_model())

wait <- TRUE
(glmer_obj <- mccollect(jb, wait = wait)[[as.character(jb$pid)]])
if(wait) glmer_obj$plot


```

### Limma contrasts
```{r, eval = TRUE}

# all patIDs
pats <- clusters %>%
  filter(n()>1, .by = patID_eye) %>%
  pull(patID)

# eyes <- clusters %>%
#   # filter(n()>1, .by = patID_eye) %>%
#   pull(patID_eye)

# sample 300 patients, maintain their multiple observations
# set.seed(3094)
# select_eyes <- eyes[eyes %in% sample(unique(eyes), 500)]
# select_eyes <- eyes
select_pats <- pats

# response x predictor matrix
block_clusters <- clusters %>% 
  arrange(patID_eye) %>% 
  # filter(patID_eye %in% select_eyes) %>% 
  filter(patID %in% select_pats) %>% 
  select( patID, patID_eye, t2d, clust) %>% 
  drop_na() 

# design on sub clusters ~ cluster without intercept
block_design <- model.matrix(~0+clust, data = block_clusters)


# make counts matrix p X n
block_clusters_m <- t(select(block_clusters, t2d))

# check
qr(block_design)$rank
dim(block_design)
dim(block_clusters_m)

# # make duplicate correlations object ( Breaks when using singlton patients in blocks!!)
# how.long(T)
# corfit <- duplicateCorrelation(block_clusters_m,
#                              design =  block_design,
#                              block = block_clusters$patID_eye
#                              )
# how.long() # ~ 20min for all; 1m for longitudinal
# ## $consensus.correlation
# ## [1] 0.99
# ## 
# ## $cor
# ## [1] 0.99
# ## 
# ## $atanh.correlations
# ## [1] 2.646652
# # ^ same result blocking by patID and patID_eye



# Fit the linear model with the correlation structure from duplicateCorrelation
how.long(T)
limma_fit <- limma::lmFit(object = block_clusters_m,
                          design = block_design,
                          # correlation = corfit$consensus.correlation,
                          # block = block_clusters$patID_eye
                          )
how.long() # ~ 1min
# # ^ coefficients match marginal means


# Apply eBayes for statistical testing

limma_contrasts <- t(do.call(rbind, Contrasts))
cfit <- contrasts.fit(limma_fit, limma_contrasts)

limma_e_fit <- eBayes(limma_fit)
# topTable(limma_e_fit)
c_efit <- eBayes(cfit)

topTable(c_efit, 1)
limma_c_results <- sapply(1:ncol(limma_contrasts), function(i) topTable(c_efit,i)) %>% t()
rownames(limma_c_results) <- colnames(limma_contrasts)
limma_c_results <-cbind(limma_c_results, Estimate = 2^as.numeric(limma_c_results[, "logFC"]) )
# B is log  odds >> c_efit$lods

# unnest each element
limma_c_results <- apply(limma_c_results, c(1, 2), function(x) x[[1]])

limma_c_results %>% 
  as.data.frame( ) %>% 
  rownames_to_column("Contrast") %>% 
  ggplot(aes(logFC, -log10(P.Value)))+
      geom_point(aes(color = Contrast), show.legend = FALSE)+
      geom_label_repel(aes(label = Contrast, color = Contrast),show.legend =FALSE)+
      geom_vline(xintercept = 0, linetype = 2)+
      geom_hline(yintercept = -log10(.05), linetype = 2)+
      ggtitle("Limma t2d~clust with mixed effects patID block")+
      theme_bw()


```

## GLM on right eyes on last observation then on left
```{r}

# make eye and patient fields
clusters <- clusters %>%
  mutate(pat_eye= patID_eye) %>% 
  separate(pat_eye, c("patID", "eye"), "_")
  
# select left/right eyes and filter for last observation for each eye
last_obs_eyes <- lapply(c(right_eyes = 1,left_eyes = 2), function(eye_num){
  clusters %>% 
    arrange(patID_eye, month) %>% 
    filter(eye == eye_num) %>% 
    filter(row_number() == n(), .by  = patID_eye)
})

# ****** FUNCTION: model function lm on dataframe ******
model_eye_side <- function(model_dataframe, formula = character()){
  
  # treatment contrasts; where each estimate is difference in mean from reference (cluster 1)
  form <- as.formula(formula)
  # fit <- lm(form, data = clusters)
  fit <- glm(form, data = model_dataframe, family = "binomial")
  # summary(fit)
  
  # design matrix
  design <- model.matrix(~0+clust, data = model_dataframe)
  
  # estimated marginal means  
  marginal = emmeans(fit, ~ clust, data = clusters)

  # contrasts model
  (Test <- contrast(marginal, Contrasts))
    
  Test <- as.data.frame(Test)
  
  return(list(Contrasts = Test, Summary = summary(fit)))
}  

# model on each df
last_obs_eyes_contrasts <- lapply(last_obs_eyes, function(dat){
  model_eye_side(dat,  formula = "factor(t2d) ~ clust")$Contrasts
})
# lapply(last_obs_eyes, function(dat)model_eye_side(dat,  formula = "factor(t2d) ~ clust")$Summary)

# make one df
last_obs_results <- lapply(names(last_obs_eyes_contrasts), function(name){
  dat = last_obs_eyes_contrasts[[name]]
  
  dat$eye <- sub("_.*", "" ,name)
  return(dat)
}) %>% 
  do.call(rbind, .)

# plot
ggplot(last_obs_results, aes(estimate, -log10(p.value), color = contrast, label = paste(eye,contrast, sep = "::")))+
  geom_vline(xintercept = 0, linetype = 2)+
  geom_hline(yintercept = -log10(.05), linetype = 2)+
  geom_line(alpha = .5, show.legend = FALSE)+
  geom_point(show.legend = FALSE)+
  geom_label_repel(show.legend = FALSE)+
  theme_bw()+
  ggtitle("GLM of left last visits and right last visits separately", "t2d~clust")
ggsave(paste0(wd_path, "results/clustering/13_clustering/T2D_Contrasts/eye_stratified_last_visit_cluster_t2d_contrasts_glm.png"), height = 10, width = 12)


```
 
