---
title: "13_routeModeling"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---


# options
```{r}
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()

```


# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

pacman::p_load(tidyr, reshape2, purrr, GGally, tidyverse, ggplot2, gridExtra, ComplexHeatmap, viridis, zoo, plotly, mgcv, tools, mice, ggrepel, htmltools, Metrics, psych, car, stats, lme4, lmerTest, ape, SingleCellExperiment, scater, velociraptor, randomForest, skimr, dplyr, HDF5Array, pdftools, parallel, patchwork, ltjmm, rstan)


# library(conflicted)
# # set priotiry funciton for conflicts
# conflicted::conflict_prefer("select", "dplyr")
# conflicted::conflict_prefer("filter", "dplyr")
# conflicted::conflict_prefer("first", "dplyr")
# conflicted::conflict_prefer("last", "dplyr")
# conflicted::conflict_prefer("cbind", "base")

rm(list = ls())
gc()

start = Sys.time()
```

# functions
```{r}
source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")
how.long()
```


# make new directorys for plots
```{r}
wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

a_path <- paste0(wd_path, "results/clustering/13_clustering")
if(!dir.exists(a_path)) {
  dir.create(a_path, recursive = TRUE)
  
}

```

# Load Data
## info and d from sce
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}


# load experiment
sce <- HDF5Array::loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/sce005.h5"))

```


## genetics data 
#### PCs
```{r}
# PCs
pcs <- read.delim("/Users/canderson/Documents/LMRI/PRJ2024001/raw_data/Liams_genetics/20_PCs.txt", sep = "\t")
```

#### patient ID df (AID, etc.)
```{r}

pids <- read.csv("/Users/canderson/Documents/LMRI/PRJ2024001/raw_data/PatientIDs.csv") 


```


<!-- #### significant loci -->
<!-- ```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE} -->

<!-- # snps info -->
<!-- loci = read.delim("/Users/canderson/Documents/LMRI/PRJ2024001/raw_data/Liams_genetics/MacTel_GWAS2023_Report/All_MacTel_Loci.txt") -->
<!-- loci.sample.info= read.delim("/Users/canderson/Documents/LMRI/PRJ2024001/raw_data/Liams_genetics/MacTel_GWAS2023_Report/MacTel_GWAS2023_Sample_Data.txt") -->
<!-- purrr::mapped.genes <- read.delim("/Users/canderson/Documents/LMRI/PRJ2024001/raw_data/Liams_genetics/MacTel_GWAS2023_Report/purrr::mapped_Genes.txt", sep = "\t") -->


<!-- # pdf of report -->
<!-- report_pdf <- pdf_text("/Users/canderson/Documents/LMRI/PRJ2024001/raw_data/Liams_genetics/MacTel_GWAS2023_Report/MacTel_GWAS2023.pdf")[4] -->

<!-- ## extract table from page -->
<!-- ### Split the text into lines -->
<!-- lines <- str_split(report_pdf, "\n")[[1]] -->

<!-- ### remove title -->
<!-- lines <- lines[2:25] -->

<!-- ### save names -->
<!-- lines[1:3] -->
<!-- table_names <- c( -->
<!--   "SNP rsid", -->
<!--   "Chr", -->
<!--   "Bp", -->
<!--   "Effect/non-effect allele", -->
<!--   "EAF Cases", -->
<!--   "EAF Controls", -->
<!--   "OR", -->
<!--   "P value", -->
<!--   "Proximal coding genes" -->
<!-- ) %>% gsub(" ", "_", .) -->

<!-- ### shape table -->
<!-- table_lines <- lines[4:length(lines)] -->

<!-- table_vec <- table_lines %>% -->
<!--   str_trim() %>% -->
<!--   gsub( " +", "\t", .) %>% -->
<!--   str_split(., "\t") -->

<!-- # fill matrix with table items -->
<!-- for(i in 1:length(table_lines)){ -->

<!--   vec <- table_vec[[i]] -->

<!--   # if vec > 9, collapse last items to item 9, if < 9, add empty item -->
<!--   if(length(vec) < 9){ -->
<!--     vec <- c(vec, rep("", 9 - length(vec))) -->
<!--   } -->
<!--   if(length(vec) > 9){ -->
<!--     vec[9] <- paste(vec[9:length(vec)], collapse = "") -->
<!--     vec <- vec[1:9] -->
<!--   } -->

<!--   # if first iteration , make matrix as first vector with table_names -->
<!--   if(i == 1){ -->
<!--     table_m <- matrix(vec, ncol = 9, dimnames = list(NA, table_names)) -->
<!--   }else{ -->
<!--     table_m <- table_m %>% rbind(vec) -->
<!--   } -->

<!-- } -->
<!-- rownames(table_m) <- NULL -->


<!-- # table_data <- lines %>% -->
<!-- #   str_trim() %>%                 # Trim any excess whitespace -->
<!-- #   str_split_fixed(" +", n = 9) %>% # Adjust 'n' based on the number of columns in your table -->
<!-- #   as.data.frame() -->

<!-- # clean table -->
<!-- clean_table_m <- table_m -->
<!-- clean_table_m[8, ncol(clean_table_m)] <- gsub(" ", "" , paste(clean_table_m[8, ncol(clean_table_m)], clean_table_m[7, ], clean_table_m[9, ], collapse = "")) -->
<!-- clean_table_m[13, ncol(clean_table_m)] <- gsub(" ", "" , paste( paste(clean_table_m[12, ], collapse  = ""), clean_table_m[14,1 ], sep = "")) -->
<!-- clean_table_m <- clean_table_m[-c(7, 9, 12,14) , ] -->

<!-- # make data.frame -->
<!-- sig_loci_d <- clean_table_m %>% as.data.frame() -->

<!-- # remove "*" from snp id -->
<!-- sig_loci_d$SNP_rsid <- sig_loci_d$SNP_rsid %>% gsub("\\*", "", .) -->



<!-- ``` -->


#### snp info
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

# snps info
snp_info <- purrr::map(17:18,~openxlsx::read.xlsx("/Users/canderson/Documents/LMRI/PRJ2024001/raw_data/Liams_genetics/new_snp_data/Identification_of_genetic_factors_bonellli_2021.xlsx", sheet = .))

snp_info <- lapply(snp_info, function(x){
  names(x) <- janitor::make_clean_names(x[1,])
  x <- x[2:nrow(x),]
  rownames(x) <- seq(nrow(x))
  return(x)
})

names(snp_info[[2]])[names(snp_info[[2]])=='x'] <- "rsid"

purrr::map(snp_info,names)

m <- matrix(NA, ncol = length( setdiff(names(snp_info[[2]]), names(snp_info[[1]]))), nrow = nrow(snp_info[[1]]), dimnames = list(NULL, setdiff(names(snp_info[[2]]), names(snp_info[[1]]))))

snp_info[[1]] <- cbind(snp_info[[1]], 
                       m)

m <- matrix(NA, ncol = length( setdiff(names(snp_info[[1]]), names(snp_info[[2]]))), nrow = nrow(snp_info[[2]]), dimnames = list(NULL, setdiff(names(snp_info[[1]]), names(snp_info[[2]]))))

snp_info[[2]] <- cbind(snp_info[[2]], 
                       m)

snp_info_d <- do.call(rbind, snp_info)

# combine these
snp_info_d$assoc_or_prox_coding_genes <- coalesce(snp_info_d$associated_coding_genes, snp_info_d$proximal_coding_genes)

# rename snps to have meaningfull names
snp_info_d <-snp_info_d %>%
  mutate(label = janitor::make_clean_names(paste(rsid, assoc_or_prox_coding_genes, sep = "_")))

# fix characteres in p value
snp_info_d <- snp_info_d %>% 
  mutate(across(matches("p_value|eaf|odds"), ~as.numeric(gsub("\\*", "", .) )))

# add simple names
simple_names <- matrix(c(
  "rs532303_phgdh_hmgcs2",                    "phgdh_common",
  "rs146953046_phgdh_hmgcs2",                 "phgdh_rare",
  "rs2120770_ppp2r5a_tmem206",                "ppp2r5a",
  "rs1260326_mpv17_gtf3c2_eif2b4_snx17_znf513_ppm1g_nrbp1_krtcap3_ift172_gckr_znf512", "gckr",
  "rs2160387_slc1a4",                          "slc1a4",
  "rs1047891_cps1",                            "cps1",
  "rs9820465_slc4a7_eomes",                    "eomes",
  "rs17279437_slc6a20_sacm1l_limd1_lars2",     "slc6a20",
  "rs17421627_tmem161b",                       "tmem161b",
  "rs6955423_sept14_mrps17_znf713_gbas_psph_cct6a_sumf2_phkg1_chchd2_nupr1l", "psph",
  "rs2954021_trib1",                           "trib1",
  "rs677622_ttc39b",                           "ttc39b",
  "rs10995566_nrbf2_jmjd1c_reep3",             "jmjd1c",
  "rs2984814_dcdc1_dnajc24_immp1l_elp4_pax6",  "pax6",
  "rs11077850_mxra7_jmjd6_mettl23_srsf2_mfsd11", "jmjd6",
  "rs139412173_cers4_fbn3",                    "fbn3"
), ncol = 2, byrow = TRUE,
dimnames = list(NULL, c("label", "simple_label")))

setdiff_intersect(list(simple_names[,"label"],snp_info_d$label),values = T)
# rs1973480_cd109
# ^  okay

simple_names <- simple_names %>% 
  as.data.frame() %>% 
  mutate(simple_label = toupper(simple_label)) %>% 
  mutate(simple_label = 
           str_replace(simple_label, "_RARE|_COMMON", 
                       tolower(str_extract(simple_label, "_RARE|_COMMON"))
                       )
         )

snp_info_d <- snp_info_d %>% 
  left_join(simple_names)

metadata(sce)$genetics <- list(snp_info = snp_info_d)

```

### add new snp data (april 2025) to loci.sample.info
    
    new snp data from liams gwas2 
```{r}

new_snps <- read.delim("/Users/canderson/Documents/LMRI/PRJ2024001/raw_data/Liams_genetics/new_snp_data/GWAS2_SNPs.txt", sep = "\t")
new_snps <- new_snps[!duplicated(new_snps$IID),]
colSums(is.na(new_snps))
# ^ 0...

setdiff_intersect(list(dat= grep("rs",str_remove_all(names(new_snps), "_."), value = TRUE), paper = snp_info_d$rsid), value = TRUE)
# Only dat : 0 
# Only paper : 1 
# rs1973480 
# dat ∩ paper : 16 
# rs532303, rs146953046, rs2120770, rs1260326, rs2160387, rs1047891, rs9820465, rs17279437, rs17421627, rs6955423, rs2954021, rs677622, rs10995566, rs2984814, rs11077850, rs139412173 
#+
#+ ^^ rs1973480 missing from data
#+ 

# Strip suffixes from names(new_snps) to match with snp_info_d$rsid
stripped_names <- str_remove_all(names(new_snps), "_.")

# Match rsid with stripped names
match_index <- match(snp_info_d$rsid, stripped_names)

# Store the original names of new_snps that match
snp_info_d$other_name <- names(new_snps)[match_index]

# Now rename elements in new_snps that match those stored in other_name
rename_index <- match(names(new_snps), snp_info_d$other_name)

# Update names of new_snps
names(new_snps)[!is.na(rename_index)] <- snp_info_d$label[rename_index[!is.na(rename_index)]]


# new_snps %>% purrr::map(~{if(n_distinct(.)<20) unique(.)})
# 
# nrow(inner_join(new_snps, loci.sample.info))-nrow(loci.sample.info)
# # -97
# nrow(inner_join(new_snps, loci.sample.info))-nrow(new_snps)
# # -1136
# ## ^^ new_snps has overlapp with previoius data and lots more 
# 
# # join all together
# new_and_old_snps <- full_join(new_snps, loci.sample.info)
# 
# compare.columns <- function(x, y) {
#   sapply(x, function(col1) {
#     sapply(y, function(col2) {
#       length(intersect(col1, col2))})})}

# 
# comp <- compare.columns(new_and_old_snps %>% select(matches("ID")), colData(sce) %>% data.frame() )
# comp[rowSums(comp)>4, ]
# #           FID  IID  batch_id
# # patId2   2822 2822        1
# # patId4     10   10        1
# # patId2_1 2822 2822        1
# 
# new_and_old_snps %>% nrow()
# # 3733
# new_and_old_snps$FID %>% n_distinct()
# # 3636
# new_and_old_snps$IID %>% n_distinct()
# # 3626
# 
# # filter out duplicates
# new_and_old_snps <- new_and_old_snps %>% 
#   # filter(!duplicated(across(-c(FID, IID))))
#   filter(!duplicated(across(-FID)))
# 
# nrow(new_and_old_snps)
# # 3725
# 
# # some patients have more than one observation, see if they can be condensed into 1
# new_and_old_snps %>% 
#   arrange(IID) %>% 
#   filter(any(grepl(":", FID)), .by = IID) %>% 
#   select(FID,IID, matches("rs|PC")) 
# # ^ some have missing in other obs, fill those in 
# 
# # fill in missing values with where not missing for each IID
# new_and_old_snps <- new_and_old_snps %>% 
#   group_by(IID) %>% 
#   summarize_all(~dplyr::first(.[!is.na(.)]))
# 
# nrow(new_and_old_snps)
# # 3626


```


#_________________________
# Mung data
```{r look at dims}

check_dims <- function(sce_obj = NULL){
  
  if(is.null(sce_obj)) sce_obj <- sce
  
  D <- as.data.frame(colData(sce_obj)) %>% 
  left_join(
    as.data.frame(t(assays(sce_obj)$scaled_measure_vars)) %>% 
      rownames_to_column("sample")
  ) 
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()
### version 013 ###
# dims : 12525 176
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ no change

### APRIL 10 2025 ###
# dims : 12276 205
# n_patIDs             3971
# n_patID_eyes         7888
# n_observations      12275
# n_patIDs_w_gr_1_vis   492

```

## attach patIDs to colData 
```{r}

## first row is header; remove extraneous chars
pids <- pids[-1, ] %>% 
  clean.it()

compare.columns <- function(x, y) {
  sapply(x, function(col1) {
    sapply(y, function(col2) {
      length(intersect(col1, col2))})})}


# what to join on?
compare.columns(pids,colData(sce)) %>% 
  {r =rowSums(.); .[r>9, ]}
#         Patient.ID A.Number Site.ID ID.4
# patId2            0     3046       1    1
# patId3            1        1     241    1
# patId4            0        1       1   25
# patId2_1          0     3046       1    1
# patID          3867        0       0    0


# add pids to colData
colData_with_pids<- data.frame(colData(sce)) %>% 
  left_join(pids, by = join_by(patID == Patient.ID, patId2 ==A.Number), keep = T) %>% 
  mutate(samp = sample) %>% 
  column_to_rownames("samp")

colData_with_pids$sample.1 <- NULL
colData_with_pids$Patient.ID <- NULL

stopifnot(all(colnames(sce)%in% rownames(colData_with_pids))&all(colnames(sce)%in% colData_with_pids$sample))

colData(sce) <- DataFrame(colData_with_pids)

```


## remove redundant cols
```{r}

sce$sample.1 <- NULL
sce$sample.2 <- NULL

```

## impute routes from others
```{r}

colData(sce) %>% 
  data.frame() %>% 
  select(matches("route")) %>% 
  {colSums(!is.na(.))} %>% 
  sort()
# route    route_score granular_route flexible.route   strict.route    pc_rf_route   pc_lda_route       rf_route 
#  2309           2309           2309           6813           6816           8429           8429           8435 
  
# where route missing in one eye, use concensus
colData(sce) %>% 
  data.frame() %>% 
  select(patID_eye,matches("route")) %>% 
  filter(any(is.na(pc_lda_route) & !is.na(rf_route)), .by = patID_eye)  # 6 rows

where <- is.na(colData(sce)$pc_lda_route) & !is.na(colData(sce)$rf_route)
colData(sce)$pc_lda_route[where] <- colData(sce)$rf_route[where]


```


## clean genetic data and join to colData
```{r}

# par(mar = rep(0,4))
# grid.newpage()
# list(new_IID= new_snps$IID, old_IID = loci.sample.info$IID, sce_A.Number = unique(sce$A.Number)) %>% 
#   purrr::map(na.omit) %>% 
#   VennDiagram::venn.diagram(disable.logging = T, filename = NULL, offset = 1, fill = c("blue", "red", "green"), cat.col =  c(rgb(0,0,1,.8),rgb(1,0,0,.8), rgb(0,1,0,.8)), cat.cex = 1.5,scaled = TRUE) %>%
#   grid.draw()
# 
# list(new_IID= new_snps$IID, old_IID = loci.sample.info$IID, sce_A.Number = unique(sce$A.Number)) %>% 
#   purrr::map(na.omit) %>% 
#   setdiff_intersect()
# # Only new_IID: 395
# # Only old_IID: 0
# # Only sce_A.Number: 223
# # new_IID ∩ old_IID: 409
# # new_IID ∩ sce_A.Number: 650
# # old_IID ∩ sce_A.Number: 0
# # new_IID ∩ old_IID ∩ sce_A.Number: 2172

compare.columns(colData(sce), new_snps[,c('FID',  'IID')]) %>% 
  t() %>% 
  {r =rowSums(.); .[r>100, ]} 
#           FID  IID
# patId2   2822 2822
# patId2_1 2822 2822
# A.Number 2822 2822

identical(sce$patId2, sce$A.Number)
# TRUE

# join new_snps to coldata
colDat_with_genetics <- colData(sce) %>% 
  data.frame()%>% 
  left_join(new_snps, 
            by = join_by(patId2 == IID), keep = TRUE,relationship = "many-to-one") %>% 
  mutate(samp = sample) %>% 
  column_to_rownames("samp")



# unique rows
(colDat_with_genetics <- distinct(colDat_with_genetics))
nrow(colDat_with_genetics)
# 8554
# 12276

colDat_with_genetics %>% 
  select(patID, any_of(names(new_snps))) %>% 
  distinct() %>% 
  {colSums(!is.na(.))}
#        patID           FID      phenotype            sex       batch_id       platform     rs532303_A  rs146953046_G 
#         3867           2822           2822           2822           2822           2822           2822           2822 
#  rs2120770_A    rs1260326_T    rs2160387_T    rs1047891_C    rs9820465_T   rs17279437_A   rs17421627_tmem161b    rs6955423_G 
#         2822           2822           2822           2822           2822           2822           2822           2822 
#  rs2954021_A     rs677622_A   rs10995566_C    rs2984814_G   rs11077850_C  rs139412173_A            PC1            PC2 
#         2822           2822           2822           2822           2822           2822           2822           2822 
#          PC3            PC4            PC5            PC6            PC7            PC8            PC9           PC10 
#         2822           2822           2822           2822           2822           2822           2822           2822 
#         PC11           PC12           PC13           PC14           PC15           PC16           PC17           PC18 
#         2822           2822           2822           2822           2822           2822           2822           2822 
#         PC19           PC20    rs2422358_A    rs9837054_A  rs114622514_G    rs4535700_T     rs686030_C   rs10822187_A 
#         2822           2822           2172           2172           2172           2172           2172           2172 
# rs67362977_A    rs8067445_G   rs72993321_T    rs8135665_C  X6.32433276_G Normalised_PRS      Phenotype 
#         2172           2172           2172           2172           2172           2172           2172 

new_snps %>% 
  select(-FID) %>% 
  distinct() %>% 
  {colSums(!is.na(.))}

setdiff_intersect(list(coldat_w_gen = colDat_with_genetics$IID, new_snps= new_snps$IID))
# Only coldat_w_gen : 0 
# Only new_snps : 804 
# coldat_w_gen ∩ new_snps : 2822 




# ADD COLS TO SCE----

setdiff(names(colDat_with_genetics), names(colData(sce)))

colData(sce) <- DataFrame(colDat_with_genetics)


md.pattern(colDat_with_genetics %>% select(matches("X\\.|rs|PC\\d{1,2}")), rotate.names = T)
# m <- data.table::fread("rs532303_A_1_PHGDHHMGCS2 rs146953046_G_1_PHGDHHMGCS2 rs2120770_A rs1260326_T_2_GKCR rs2160387_T rs1047891_C_2_CPS1 rs9820465_T rs17279437_A_3_SLC6A20LIMD1SACM1LLARS2 rs17421627_tmem161b rs6955423_G rs2954021_A_8_TRIB1 rs677622_A rs10995566_C rs2984814_G rs11077850_C rs139412173_A  PC1  PC2  PC3  PC4  PC5  PC6  PC7  PC8  PC9 PC10 PC11 PC12 PC13 PC14 PC15 PC16 PC17 PC18 PC19 PC20 rs2422358_A_2_SLC1A4 rs9837054_A_3_SLC4A7EOMES rs17421627_tmem161b rs4535700_T_7_SEPT14MRPS17ZNF713 rs686030_C_9_TTC39B rs10822187_A_10_NRBF2JMJD1CREEP3 rs67362977_A_16_GPT2MYLK3C16orf87 rs8067445_G_17_MXRA7 rs72993321_T_19_CERS4FBN3 rs8135665_C_22_SLC16A8 Normalised_PRS
# 7098                        1                           1           1                  1           1                  1           1                                      1            1           1                   1          1            1           1            1             1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1                    1                         1                        1                                1                   1                                1                                 1                    1                         1                      1              1      0
# 1765                        1                           1           1                  1           1                  1           1                                      1            1           1                   1          1            1           1            1             1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1    1                    0                         0                        0                                0                   0                                0                                 0                    0                         0                      0              0     11
# 3413                        0                           0           0                  0           0                  0           0                                      0            0           0                   0          0            0           0            0             0    0    0    0    0    0    0    0    0    0    0    0    0    0    0    0    0    0    0    0    0                    0                         0                        0                                0                   0                                0                                 0                    0                         0                      0              0     47
#                          3413                        3413        3413               3413        3413               3413        3413                                   3413         3413        3413                3413       3413         3413        3413         3413          3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413 3413                 5178                      5178                     5178                             5178                5178                             5178                              5178                 5178                      5178                   5178           5178 179826
# ") %>% 
#   as.matrix()

```



### compare previous genetics to current version
```{r, eval = FALSE}

sum(!is.na(sce$t2d))
# 11854(v13)>>5877>>6089
table(sce$t2d)
#    0    1 
# 4007 1870
# 4039 2050 

options(max.print = 500)
data.frame(colData(sce)) %>% 
  select(matches("rs|X6|PC|NORM.*PRS")) %>% 
  skimr::skim(.) %>% 
  as.data.frame()
options(max.print = 200)



#                             skim_variable prev.version.n_missing n_missing prev.version.complete_rate complete_rate numeric.mean numeric.sd numeric.p0 numeric.p25 numeric.p50 numeric.p75 numeric.p100 numeric.hist prev.version.numeric.hist
# 1                rs532303_A_1_PHGDHHMGCS2                   4940      5180                      0.606         0.578      0.89000     0.7200      0.000      0.0000     1.00000     1.0          2.0            ▆▁▇▁▃                     ▆▁▇▁▃
# 2             rs146953046_G_1_PHGDHHMGCS2                   4940      5180                      0.606         0.578      0.12700     0.3440      0.000      0.0000     0.00000     0.0          2.0            ▇▁▁▁▁                     ▇▁▁▁▁
# 3                      rs1260326_T_2_GKCR                   4940      5180                      0.606         0.578      1.01000     0.6840      0.000      1.0000     1.00000     1.0          2.0            ▃▁▇▁▃                     ▃▁▇▁▃
# 4                    rs2422358_A_2_SLC1A4                   4940      5180                      0.606         0.578      1.29000     0.6790      0.000      1.0000     1.00000     2.0          2.0            ▂▁▇▁▇                     ▂▁▇▁▇
# 5                      rs1047891_C_2_CPS1                   4940      5180                      0.606         0.578      1.63000     0.5630      0.000      1.0000     2.00000     2.0          2.0            ▁▁▃▁▇                     ▁▁▃▁▇
# 6               rs9837054_A_3_SLC4A7EOMES                   4940      5180                      0.606         0.578      1.81000     0.4190      0.000      2.0000     2.00000     2.0          2.0            ▁▁▂▁▇                     ▁▁▂▁▇
# 7  rs17279437_A_3_SLC6A20LIMD1SACM1LLARS2                   4940      5180                      0.606         0.578      0.29400     0.4860      0.000      0.0000     0.00000     1.0          2.0            ▇▁▃▁▁                     ▇▁▃▁▁
# 8                rs17421627_tmem161b                   4940      5180                      0.606         0.578      0.27000     0.4900      0.000      0.0000     0.00000     0.0          2.0            ▇▁▂▁▁                     ▇▁▂▁▁
# 9        rs4535700_T_7_SEPT14MRPS17ZNF713                   4940      5180                      0.606         0.578      0.65300     0.6690      0.000      0.0000     1.00000     1.0          2.0            ▇▁▇▁▂                     ▇▁▇▁▂
# 10                    rs2954021_A_8_TRIB1                   4940      5180                      0.606         0.578      1.11000     0.6900      0.000      1.0000     1.00000     2.0          2.0            ▃▁▇▁▅                     ▃▁▇▁▅
# 11                    rs686030_C_9_TTC39B                   4940      5180                      0.606         0.578      0.38300     0.5570      0.000      0.0000     0.00000     1.0          2.0            ▇▁▃▁▁                     ▇▁▃▁▁
# 12       rs10822187_A_10_NRBF2JMJD1CREEP3                   4940      5180                      0.606         0.578      1.48000     0.6210      0.000      1.0000     2.00000     2.0          2.0            ▁▁▆▁▇                     ▁▁▆▁▇
# 13      rs67362977_A_16_GPT2MYLK3C16orf87                   4940      5180                      0.606         0.578      1.69000     0.5090      0.000      1.0000     2.00000     2.0          2.0            ▁▁▃▁▇                     ▁▁▃▁▇
# 14                   rs8067445_G_17_MXRA7                   4940      5180                      0.606         0.578      1.25000     0.6930      0.000      1.0000     1.00000     2.0          2.0            ▂▁▇▁▇                     ▂▁▇▁▇
# 15              rs72993321_T_19_CERS4FBN3                   4940      5180                      0.606         0.578      1.96000     0.2000      0.000      2.0000     2.00000     2.0          2.0            ▁▁▁▁▇                     ▁▁▁▁▇
# 16                 rs8135665_C_22_SLC16A8                   4940      5180                      0.606         0.578      1.70000     0.4970      0.000      1.0000     2.00000     2.0          2.0            ▁▁▃▁▇                     ▁▁▃▁▇
# 17                       X6.32433276_G_6_                   4940      5180                      0.606         0.578      1.44000     0.6320      0.000      1.0000     2.00000     2.0          2.0            ▁▁▆▁▇                     ▁▁▆▁▇
# 18                                    PC1                   3800      4140                      0.697         0.663     -0.00263     0.0153     -0.191     -0.0026     0.00243     0.00313      0.00441        ▁▁▁▁▇                     ▁▁▁▁▇
##^ with new genetics download, same snps have same number of missing, but new snps have 20% more than old

# how much genetics data?
data.frame(colData(sce))[,c('patID', 'PC1')] %>%
  drop_na() %>% 
  distinct()  %>% 
  nrow()
# 2499>>2822
data.frame(colData(sce))[,c('patID', 'PC1')] %>% 
  drop_na()  %>% 
  nrow()
# 8140>>8863
data.frame(colData(sce))[,c('patID', "rs17421627_tmem161b")] %>% 
  drop_na() %>% 
  distinct()  %>% 
  nrow()
# 2172>>2172
data.frame(colData(sce))[,c('patID', "rs17421627_tmem161b")] %>%
  drop_na()  %>% 
  nrow()
# 7098>>7098

sum(!is.na(sce$rf_route))
#  8435

loci.sample.info %>% 
  distinct() %>% 
  {!is.na(.)} %>% 
  colSums() %>% 
  as.matrix()
# FID            2585
# IID            2585
# rs532303_A     2585
# rs146953046_G  2585
# rs1260326_T    2585
# rs2422358_A    2585
# rs1047891_C    2585
# rs9837054_A    2585
# rs17279437_A   2585
# rs114622514_G  2585
# rs4535700_T    2585
# rs2954021_A    2585
# rs686030_C     2585
# rs10822187_A   2585
# rs67362977_A   2585
# rs8067445_G    2585
# rs62993321_T   2585
# rs8135665_C    2585
# X6.32433276_G  2585
# Normalised_PRS 2585
# Phenotype      2585

data.frame(colData(sce)) %>% 
  select(patID, matches(paste(names(loci.sample.info) , sep = "|"))) %>% 
  filter(row_number()==1, .by = patID) %>% 
  {!is.na(.)} %>% 
  colSums() %>% 
  as.matrix()
# patID                                  3971
# IID                                    2499
# rs532303_A_1_PHGDHHMGCS2               2172
# rs146953046_G_1_PHGDHHMGCS2            2172
# rs1260326_T_2_GKCR                     2172
# rs2422358_A_2_SLC1A4                   2172
# rs1047891_C_2_CPS1                     2172
# rs9837054_A_3_SLC4A7EOMES              2172
# rs17279437_A_3_SLC6A20LIMD1SACM1LLARS2 2172
# rs17421627_tmem161b               2172
# rs4535700_T_7_SEPT14MRPS17ZNF713       2172
# rs2954021_A_8_TRIB1                    2172
# rs686030_C_9_TTC39B                    2172
# rs10822187_A_10_NRBF2JMJD1CREEP3       2172
# rs67362977_A_16_GPT2MYLK3C16orf87      2172
# rs8067445_G_17_MXRA7                   2172
# rs72993321_T_19_CERS4FBN3              2172
# rs8135665_C_22_SLC16A8                 2172
# X6.32433276_G_6_                       2172
# Normalised_PRS                         2172
# Phenotype                              2172

(counts<- table(sce$rs17421627_tmem161b, useNA = "no"))
#    0    1    2       XX <NA> 
# 5336 1608  154       XX 5178 
# 6858 1843  162 

(ratios <- counts/sum(counts))
#      0          1          2        XX      <NA> 
#  0.4346693  0.1309873  0.0125448       XX 0.4217986 
# 0.77377863 0.20794313 0.01827824 

# previous counts
prev_counts <- c(5693, 1718,  174)

# previos ratios
prev_ratios <- prev_counts/sum(prev_counts)

counts-prev_counts
#   0    1    2 
# -357 -110  -20
#^^ -487 records with snp value

par(mar = c(5,4,4,4), mfrow = c(1,1))
plot(prev_ratios,as.numeric(ratios))
add_lines()
plot(prev_counts,as.numeric(counts))
add_lines()


saveRDS(colnames(sce),sprintf("%sprocessed_data/samples.rds", wd_path))

# read in old sce samples to compare
old_samples <- readRDS(sprintf("%sprocessed_data/samples.rds", str_replace(wd_path, "14","13")))

# function to round number at end of string
round_end <- function(x = character(), sep = "_"){
  # grab ids
  body <- str_split(x, sep) %>% 
    purrr::map(`[`,1:2)
  # grab months
  num <- str_split(x, sep) %>% 
    purrr::map(`[`,3)
  # round months
  num <- round(as.numeric(unlist(num)))
  # combine ids again
  body <- unlist(purrr::map(body, function(x) paste(x, collapse = sep)))
  # add month again
  x_new <- paste(body, num, sep = sep)
  
  return(x_new)
}

old_samples <- round_end(old_samples)
new_samples <- round_end(colnames(sce))

par(mar = rep(0,4))
plot.new()
VennDiagram::venn.diagram(list(old = old_samples ,new = new_samples),
                          filename = NULL) %>% grid.draw()
par(mar = c(5,4,4,4))

# what samples are in current, but not in old version
check_dims(sce = sce[,!new_samples%in% old_samples])
# dims : 3123 246
# n_patIDs             742
# n_patID_eyes        1469
# n_observations      3122
# n_patIDs_w_gr_1_vis  338
# ^ samples unique to version 014

```

## calculate covariates
### update SEXF
```{r}

data.frame(colData(sce)) %>% 
  select(matches("sex")) %>% 
  summarize(n = n(), .by = everything()) %>% 
  arrange(n)
#    SEXF sex    n
# 1    NA  -9   96
# 2 FALSE  NA  637
# 3  TRUE  NA  838
# 4 FALSE   1 1428
# 5    NA   1 1766
# 6    NA  NA 1938
# 7  TRUE   2 2581
# 8    NA   2 2992

colData(sce)$SEXF[sce$sex == 2] <- TRUE
colData(sce)$SEXF[sce$sex == 1] <- FALSE


```

#### and compare prev version yrofbrth to current
```{r, eval = TRUE}

colSums(!is.na(colData(sce)[, c("SEXF", "YROFBRTH")]))
# 5484 5484
# 10242  5484

new <- data.frame(colData(sce)) %>% 
  select(patID, YROFBRTH, SEXF) %>% 
  mutate(YROFBRTH = as.numeric(YROFBRTH)) %>% 
  distinct()

# # use old sce to filll dem vars ----
# old_sce <-  readRDS( "/Users/canderson/Documents/LMRI/PRJ2024001/analysis_versions/version013/processed_data/sce005.rds")
# 
# old <- data.frame(colData(old_sce)) %>% 
#   select(patID, YROFBRTH, SEX) %>% 
#   mutate(YROFBRTH = as.numeric(YROFBRTH)) %>% 
#   distinct()
# 
# # whats in both
# inn <- inner_join(old,new, by = 'patID', suffix = c(".old", ".new"))
# 
# mice::md.pattern(inn, rotate.names = T)
# 
# inn %>% 
#   drop_na() %>% 
#   filter(YROFBRTH.old!=YROFBRTH.new) %>% 
#   nrow()
# # 0. no disagreement
# 
# # add old sce yrofbiorth to new sce
# add <- inn %>%
#  filter(!is.na(YROFBRTH.old), is.na(YROFBRTH.new))
# 
# all(add$patID%in%sce$patID)
# 
# colData(sce)$YROFBRTH[match( add$patID, colData(sce)$patID) ] <-  add$YROFBRTH.old


# incorporate new download of DEM (raw_data/07/DEM.csv; 15 APRIL 2025) for YROFBRTH and SEX ----
new_dem <- read.csv("/Users/canderson/Documents/LMRI/PRJ2024001/raw_data/07/DEM.csv") %>%
  clean.it() %>%
  mutate(YROFBRTH = as.numeric(YROFBRTH))

add <- new_dem %>%
  select(patID = patientId, YROFBRTH, GENDER) %>%
  mutate(SEXF = GENDER == "Female") %>%
  select(-GENDER) %>%
  filter(patID%in%sce$patID) %>%
  distinct()



inner_join(new, add, by = 'patID', suffix = c(".old", ".new")) %>%
  md.pattern(rotate.names = T)

# dfs for what data to add; filter for where  full so not adding nas to coldata
add_SEXF <- add %>% 
  select(patID, SEXF) %>% 
  drop_na()

add_YROFBRTH <- add %>% 
  select(patID, YROFBRTH) %>% 
  drop_na()

where_add_SEXF <- match( add_SEXF$patID, colData(sce)$patID)
where_add_YROFBRTH <- match( add_YROFBRTH$patID, colData(sce)$patID)

stopifnot(!any(c(is.na(where_add_SEXF), is.na(where_add_YROFBRTH))))

colData(sce)$SEXF[where_add_SEXF ] <-  add_SEXF$SEXF
colData(sce)$YROFBRTH[where_add_YROFBRTH] <-  add_YROFBRTH$YROFBRTH

# ----
colSums(!is.na(colData(sce)[, c("SEXF", "YROFBRTH")]))
# 5484 8494 >> with old sce data
# 8617 8644 >> with new download
# 8617 8792 >> with new download and filtering add dfs for no missing
# 10978  8710 >> with new download and filtering add dfs for no missing; and using sex from new genetics download


# patID_eyes_wout_SEX <- colData(sce) %>%
#     as.data.frame() %>%
#     select( patID_eye, SEXF) %>%
#     distinct() %>%
#     filter(n()>1, .by = patID_eye) 
# 
# sce$SEXF[sce$patID_eye%in%patID_eyes_wout_SEX$patID_eye] <- rep(TRUE, 4)

# # Get non-missing SEXF values by patID
# sexf_lookup <- colData(sce) %>%
#   as.data.frame() %>%
#   select(patID, SEXF) %>%
#   filter(!is.na(SEXF)) %>%
#   distinct(patID, .keep_all = TRUE)
# 
# # Broadcast SEXF across all entries of same patID
# sexf_broadcast <- match(colData(sce)$patID, sexf_lookup$patID)
# colData(sce)$SEXF <- sexf_lookup$SEXF[sexf_broadcast]
# 


```



### callculate age vars
```{r}

sum(!is.na(sce$YROFBRTH))
# 8644
# 8710

sce$YROFBRTH <- as.Date(paste0(sce$YROFBRTH,"-01-01-"), format = "%Y-%m-%d")

(!is.na(sce$YROFBRTH)) %>% table(useNA = "ifany")
# FALSE  TRUE 
#  6792  5484
#  3632  8644 >> redownloaded dem
#  3484  8792  >> redownloaded dem and filtering 'add' dfs for no missing
 # 3566  8710  >> redownloaded dem and filtering 'add' dfs for no missing;  and using sex from new genetics download

 ## calculate baseline eye age: groupby eye; select baseline visit where month == first month; select distinct rows; remove missing; calculate difference in years
age_d <- colData(sce) %>% 
  as.data.frame() %>% 
  dplyr::select( patID_eye, visitID, YROFBRTH, DATE_coalesced) %>% 
  group_by(patID_eye) %>% 
  filter(visitID == dplyr::first(visitID)) %>% 
  unique()  %>% 
  na.omit() %>% 
  mutate(BL_eye_age = as.numeric(round(difftime( DATE_coalesced, as.Date(YROFBRTH), units = "days")/365))) %>% 
  ungroup()

## join to sce
age_to_join <- colData(sce) %>% 
  as.data.frame() %>% 
  left_join(
    age_d %>% dplyr::select( patID_eye, BL_eye_age, YROFBRTH), by = c("patID_eye", "YROFBRTH"),
    relationship = "many-to-one"
    )

## check that bl age calculated where possible
# age_to_join %>% 
#   filter(!is.na(DATE_coalesced) & !is.na(YROFBRTH) & is.na(BL_eye_age)) %>% 
#   dplyr::select( patID_eye, visitID, month, DATE_coalesced, YROFBRTH, BL_eye_age)

### add BL age to sce
sce$BL_eye_age <- age_to_join$BL_eye_age

(!is.na(sce$BL_eye_age)) %>%  table(useNA = "ifany")
# FALSE  TRUE 
#  6792  5484
#  3782  8494 
#  3850  8426 

colData(sce) %>% 
  as.data.frame() %>% 
  # filter(my_cluster =='1') %>% 
  ggplot(aes(x = BL_eye_age, color =  my_cluster, fill = my_cluster))+
  stat_density(bw = 1)+
  scale_fill_manual(values = cluster_cols)+
  scale_color_manual(values = cluster_cols)+
  # facet_wrap(~my_cluster)+
  ggtitle("Cluster Ages")

# add age + months from first visit
sce$AGE <-round(sce$BL_eye_age + sce$month/12)

plotTSNE(sce, color_by = "AGE")



plot(sce$AGE ~ sce$my_cluster, main = "Age ~ cluster")
plot(sce$AGE ~ sce$granular_route, main = "Age ~ granular_route")
plot(sce$AGE ~ sce$rf_route, main = "Age ~ rf_route")
plot(sce$AGE ~ sce$pc_lda_route, main = "Age ~ pc_lda_route")
plot(sce$BL_eye_age ~ sce$rf_route, main = "Age ~ rf_route")



```

### calculate bmi 
```{r}
# BMI= kg/m^2
# BMI= lbs * 703 / inches^2 
sce$HGTIN <- as.numeric(sce$HGTIN)
sce$WGTLBS <- as.numeric(sce$WGTLBS)
sce$BMI <- (as.numeric(sce$WGTLBS)* 703) / as.numeric(sce$HGTIN)^2


# quick model
x1<-data.frame(colData(sce)) %>% 
  mutate(rf_routeA = (rf_route == "a")*1) %>% 
  # mutate(rf_route = factor(rf_route, levels = c("a", "b", "x"))) %>% 
  mutate(rf_route = factor(rf_route, levels = c("a", "b"))) %>% 
  ungroup()


covs <- c('AGE','SEXF','BL_eye_age','BMI','HGTIN','WGTLBS')
forms <- t(combn(covs,3)) %>% 
  apply(., 1,function(x)paste(x,  collapse= "+"))

summs <- sapply(forms, function(form){
  form <- as.formula(paste0("rf_route ~ ", form))
  summary(glm(form, family = "binomial", data = x1))
}, simplify = FALSE)

summary(glm(rf_route ~ AGE+SEXF+BL_eye_age, family = "binomial", data = x1)) %>% 
  
  mock_text()
# (Intercept) -3.07886    0.22005 -13.991  < 2e-16 ***
# AGE          0.00191    0.01409   0.136 0.892121    
# SEXFTRUE     0.10654    0.06060   1.758 0.078713 .  
# BL_eye_age   0.04726    0.01430   3.305 0.000951 ***

```



#__________________________

# Side analyses ~~~~~~~~~~~~~~~~


<!-- ```{r} -->

<!-- library(igraph) -->

<!-- X <- reducedDim(sce, "PCA") -->
<!-- set.seed(100) -->
<!-- X <- X[sample(seq(nrow(X)), 5e3, F) , ] -->

<!-- distance <- as.matrix(dist(X)) -->
<!-- graph <- igraph::graph_from_adjacency_matrix(distance, weighted = T) -->
<!-- mst <- igraph::mst(graph) -->

<!-- edges <- as_edgelist(mst) -->
<!-- edges <- as.data.frame(edges) %>%  -->
<!--   dplyr::rename_with(~c("from","to")) -->

<!-- plotTSNE(sce[, colnames(sce)%in%rownames(X)], color_by = "my_cluster") -->
<!-- tsne <- as.data.frame(reducedDim(sce, "TSNE")) -->
<!-- tsne$sample <- rownames(tsne) -->
<!-- tsne$my_cluster <- sce$my_cluster -->

<!-- tsne <- tsne[tsne$sample%in%rownames(X) , ] -->

<!-- # Merge edge coordinates -->
<!-- transitions <- edges %>% -->
<!--   left_join(tsne %>% dplyr::rename(from_TSNE1 = TSNE1, from_TSNE2 = TSNE2), by = c("from" = "sample")) %>% -->
<!--   left_join(tsne %>% dplyr::rename(to_TSNE1 = TSNE1, to_TSNE2 = TSNE2), by = c("to" = "sample")) -->

<!-- # Plot t-SNE with MST edges -->
<!-- ggplot(tsne, aes(x = TSNE1, y = TSNE2, color = my_cluster)) + -->
<!--   geom_point(size = 2.0, alpha = 0.8) + -->
<!--   geom_segment(data = transitions, -->
<!--                aes(x = from_TSNE1, y = from_TSNE2, xend = to_TSNE1, yend = to_TSNE2), -->
<!--                color = "black", alpha = 0.3, linewidth = 2) + -->
<!--   theme_minimal() + -->
<!--   scale_color_manual(values = cluster_cols)+ -->
<!--   labs(color = "Cluster") -->


<!-- ``` -->


## Pseudoprog var plots

```{r, eval = FALSE}

dat <- data.frame(colData(sce)) %>% 
  left_join(data.frame(t(assays(sce)$pmm_imp_of_scaled)) %>%rownames_to_column("sample") ) 


dat$pseudoprog_quantile <- round(dat$pseudoprog,2)

dat %>% 
  filter(!is.na(pc_lda_route)) %>% 
  filter(pc_lda_route!="x") %>% 
  group_by(pseudoprog_quantile, pc_lda_route) %>% 
  summarize(across(all_of(rownames(sce)), mean)) %>% 
  ungroup() %>% 
  pivot_longer(-c(pseudoprog_quantile, pc_lda_route)) %>% 
  ggplot(aes(pseudoprog_quantile, value, color = pc_lda_route))+
  # geom_line()+
  # geom_smooth(method = "loess")+
  geom_smooth(method = "lm")+
  facet_wrap(~name, scales = "free")+
  theme_bw()
  

lm(pseudoprog~RW074*pc_lda_route ,data = dat %>% filter(pc_lda_route %in% c("a", "b")) ) %>% summary()
  


```

## chew grade quintile plots
```{r}

chew <- data.frame(colData(sce)) %>% 
  arrange(patID_eye, month) %>% 
  select(patID, patID_eye, month, AGE, chew =  chew_grade_derived_mine_loose) %>% 
  group_by(patID_eye) %>% 
  filter(n()>2) %>% 
  mutate(chew = cummax(chew)) %>%
  mutate(first_chew = dplyr::first(chew)) %>% 
  ungroup()

# chew_firsts <- chew %>% 
#   group_by(patID_eye) %>% 
#   dplyr::slice(1)
#   
# chew_quantiles <- quantile(chew_firsts$chew, seq(0,1,.3333))
# 
# chew_firsts$quantile <- cut(chew_firsts$chew, breaks = chew_quintiles, include.lowest = T, labels = F)
# 
# chew <- left_join(chew, chew_firsts %>% select(patID_eye, quantile) %>% 
#                     unique,
#                     by = "patID_eye")

ggplot(chew, aes(month, chew))+
  # geom_line( show.legend = FALSE, color = rgb(0,0,0,.4))+
  # geom_smooth(aes(group = patID_eye), show.legend = FALSE, color = rgb(0,0,0,.1),method= "lm", se = FALSE)+
  geom_smooth(method= "lm")+
  facet_wrap(~first_chew, ncol = 7)+
  ggtitle("Chew grade over time by BL Chew Grade")+
  theme_bw()
ggsave(sprintf("%sresults/publication_figures/chew_grade_bl_facet.pdf",wd_path), height = 4, width = 20)
  

```


## sigmoid time offset model
```{r, eval = FALSE}

library(brms)

offset_model_d <- data.frame(colData(sce)) %>% 
  select(patID, patID_eye, month, AGE, pseudoprog, chew = chew_grade_derived_mine_loose) %>% 
  arrange(patID_eye, month) %>% 
  group_by(patID_eye) %>% 
  filter(n()>3) %>% 
  mutate(regr_fxd_pseudoprog = cummax(pseudoprog)) %>% 
  mutate(regr_fxd_chew = cummax(chew)) %>% 
  ungroup()


set.seed(1202)
offset_model_d <- offset_model_d[offset_model_d$patID_eye%in%sample(unique(offset_model_d$patID_eye), 100, F), ]


ggplot(offset_model_d, aes(AGE, regr_fxd_chew, group = patID_eye))+
   geom_smooth(method = "gam", formula = y~s(x, k = 3), show.legend = FALSE, color = rgb(0,0,0,.2), se = FALSE)

formula <- bf(
  regr_fxd_chew ~ Amax  / (1 + ((Amax - A0) / A0) * exp(-k * (AGE+offset) )),
  offset ~ (1 | patID_eye),
  # Amax + A0 + k ~1, # intercept only fixed effecets>> only one estimate  of each for all patients
  nl = TRUE)

prior <- prior(normal(6, 1), nlpar = 'Amax') + # aruond where do you think the upper asymptote is?
  prior(normal(0, 0.1), nlpar = 'A0') + # aruond where do you think the lower asymptote is?
  prior(normal(20, 10), nlpar = 'offset') + # aruond how much do you think patients 'disease age' is shifted?
  prior(uniform(0, 10), nlpar = 'k')  # aruond how fast do you think outcome worsens?

# Fit model
fit <- brm(
  formula = formula,
  data = offset_model_d,
  prior = prior,
  chains = 10,
  cores = parallel::detectCores()-1,
  iter = 1000,
  backend = "cmdstanr",
  # algorithm ="meanfield" 
  algorithm ="sampling" ,
  control = list(adapt_delta = .8)
)

plot(fit)
summary(fit)

saveRDS(fit,sprintf("%sresults/sigmoid_offset_mod.rds", wd_path))
fit <- readRDS(sprintf("%sresults/sigmoid_offset_mod.rds", wd_path))


offset_estimate <- coef(fit)$patID_eye[,,"offset_Intercept"][, "Estimate"] %>% 
  as.data.frame()
names(offset_estimate) <- "offset"
offset_estimate$patID_eye <- rownames(offset_estimate)

plot(offset_estimate$offset)
txtdensity(offset_estimate$offset)

# add to data
offset_model_d <- left_join(offset_model_d, offset_estimate)
offset_model_d$offset_AGE <- offset_model_d$AGE+offset_model_d$offset 


offset_model_d %>% 
  pivot_longer(c(AGE,offset_AGE)) %>% 
  mutate(name = factor(name, levels = c("AGE", "offset_AGE"))) %>% 
  ggplot(aes(value, pseudoprog))+
  # geom_line(aes(group=patID_eye))+
  #geom_point(alpha = .1)+
  geom_smooth(aes(group = patID_eye),method = "gam", formula = y~s(x, k = 4), se = FALSE, color = rgb(0,0,0,.1))+
  # stat_smooth(
  #   aes(group = patID_eye),
  #   method = "nls",
  #   formula = y ~ Amax_true / (1 + ((Amax_true - A0_true) / A0_true) * exp(-k_true * x)),
  #   method.args = list(start = list(Amax_true = Amax_true, A0_true = A0_true, k_true = k_true)),
  #   # formula = y ~ 5 / (1 + ((5 - .001) / .001) * exp(-.2 * x)),
  #   show.legend = FALSE,
  #   se = FALSE,
  #   color = rgb(0, 0, 0, 0.4)
  # )+
  facet_wrap(~name, scales = "free")+
  theme_bw()

```


## latent time modeling
```{r, eval = FALSE}

dat <- data.frame(colData(sce)) %>% 
  select(patID_eye, BL_eye_age  ,month,  SEXF, pseudoprog, chew_grade_derived_mine_loose) %>% 
  mutate(SEXF = SEXF*1) %>% 
  mutate(across(c(pseudoprog, chew_grade_derived_mine_loose), ~(.-min(.))/(max(.)-min(.)) )) %>%
  melt(measure.vars = c('pseudoprog', "chew_grade_derived_mine_loose")) %>%
  ungroup() 

head(dat)

# rstan settings
options(mc.cores = parallel::detectCores())
options(auto_write = TRUE)

# Read data
# The data needs to be in long format with following columns: Patient_ID, Time.since.diag_scaled, Age_at_diagnosis_scaled, Sex, variable, value
# The column 'variable' contains the outcomes, i.e. the following values: UPDRS1_scaled, UPDRS2_scaled, UPDRS3_scaled, UPDRS4_scaled, PIGD_scaled, MCATOT_scaled, SCOPA_scaled
# All outcomes were min-max scaled regarding the theoratical minimum and maximum of the scores.
# Additionally, Age_at_diagnosis_scaled and Time.since.diag_scaled are calculated as the original times (in years) divided by 10. This speeds up the convergence of the chains.

 # Y ~ variable for observation time | fixed effects | subject id | outcome id
# fit LTJMM
fit <- ltjmm::ltjmm_stan(
  # value ~ Time.since.diag_scaled |1 + Age_at_diagnosis_scaled + Sex |Patient_ID |variable
  # formula = value ~ month |1 + BL_eye_age + SEXF |patID|variable,
  formula = value ~ month |1+SEXF+BL_eye_age| patID_eye|variable,
  # random_effects = "multivariate",
  data = (dat)
  # pars = c("beta", "delta", "alpha0", "alpha1", "gamma","sigma_delta", "sigma_y", "log_lik"),
  # open_progress = TRUE,
  # chains = 4,
  # iter = 25000,
  # thin = 5,
  # warmup = 12500,
  # cores = detectCores() - 1,
  # seed = 1,
  # refresh = 10
)
sce$unq_time <- fit$data$unq_time


data.frame(colData(sce)) %>% 
  filter(n()>1, .by = patID_eye) %>%
  filter(pc_lda_route %in% c("a", "b")) %>% 
  ggplot(aes(month, pseudoprog, color = pc_lda_route))+
  # ggplot(aes(month, chew_grade_derived_mine_loose, color = pc_lda_route))+
  geom_line(aes(group = patID_eye),show.legend = FALSE, alpha = .2)+
  # geom_smooth(formula = y~s(x, k = 4))+
  geom_smooth()+
  theme_bw()

data.frame(colData(sce)) %>% 
  filter(n()>1, .by = patID_eye) %>%
  arrange(month) %>% 
  filter(row_number()==1, .by = patID_eye) %>% 
  pull(pseudoprog) %>% 
  density() %>% 
  plot()
  

```



## Compare routes between this and previous version
```{r, eval = FALSE}

old_sce <-  loadHDF5SummarizedExperiment( "/Users/canderson/Documents/LMRI/PRJ2024001/analysis_versions/version013/processed_data/sce006.h5")

# simplify all routes
old <- data.frame(colData(old_sce)) %>% 
  select(patID_eye, matches("rf_route$"), rs17421627_tmem161b) %>% 
  mutate_all(tolower) %>% 
  mutate(across(matches("rf_route"), ~case_when(
    .=="c"~'a', 
    .=="d"~'b', 
    TRUE ~ . 
  ), .names = "{.col}_reduced")) %>% 
  select(patID_eye, rf_route, rf_route_reduced,rs17421627_tmem161b) %>% 
  distinct() 

table(old$patID_eye)[table(old$patID_eye)>1]
# 0 

new <- data.frame(colData(sce)) %>% 
  select(patID_eye, rf_route, granular_route, pc_rf_route, pc_lda_route,rs17421627_tmem161b) %>% 
  distinct()

# data.frame(colData(sce)) %>% 
#   select(route, granular_route, rf_route) %>% 
#   group_by(across(everything())) %>% 
#   summarize(n = n()) %>% 
#   arrange(n)

table(new$patID_eye)[table(new$patID_eye)>1]
# 0 
length(setdiff(old$patID_eye, new$patID_eye))
# 89
length(setdiff(new$patID_eye, old$patID_eye))
# 649
length(union(new$patID_eye, old$patID_eye))
# 7977


# obs shared by both
inn <- inner_join(new, old, by = 'patID_eye',suffix = c("_new", "_old"))


list(
  table(rf_route_new_granular=inn$granular_route, rf_route_old= inn$rf_route_old),
  table(rf_route_new=inn$rf_route_new, rf_route_old_reduced= inn$rf_route_reduced), 
  table(pc_rf_route=inn$pc_rf_route, rf_route_old_reduced= inn$rf_route_reduced),
  table(pc_lda_route=inn$pc_lda_route, rf_route_old_reduced= inn$rf_route_reduced)
  ) %>% purrr::map(~{list(x = ., accu = sum(diag(.))/sum(.))})
#                      rf_route_old
# rf_route_new_granular   a   b   c   d   x
#                     a  14   0  26   3   8
#                     b   2   2   7  26  11
#                     c   1   1  73   5   7
#                     d   2   2  12 110  19
#                     x  15   1  40  18 120
# [1] 0.607619
#             rf_route_old_reduced
# rf_route_new    a    b    x
#            a 1133  221   39
#            b  303 1285   53
#            x  111   32  234
#            rf_route_old_reduced
# [1] 0.7774846
# pc_rf_route    a    b    x
#           a 1080  109   45
#           b  307 1360   65
#           x  160   69  215
#             rf_route_old_reduced
# [1] 0.7785924
# pc_lda_route    a    b    x
#            a  931  104   32
#            b  411 1284   65
#            x  205  150  228
# [1] 0.7164223


# for where both eyes follow same route, is the route the same in both versions?
x <- inn %>% 
  mutate(patID = str_remove(patID_eye, "_\\d")) %>% 
  # filter(n_distinct(rf_route_new)==1, n_distinct(rf_route_reduced)==1, .by = patID_eye) %>% 
  group_by(patID_eye) %>% 
  filter(n_distinct(rf_route_old)==1) %>%
  # select(-patID_eye) %>% 
  distinct()

nrow(x)
# 7239

list(
  # table(rf_route_new_granular=x$granular_route, rf_route_old= x$rf_route_old),
  table(rf_route_new=x$rf_route_new, rf_route_old_reduced= x$rf_route_reduced), 
  table(pc_rf_route=x$pc_rf_route, rf_route_old_reduced= x$rf_route_reduced),
  table(pc_lda_route=x$pc_lda_route, rf_route_old_reduced= x$rf_route_reduced)
  ) %>% purrr::map(~{list(x = ., accu = sum(diag(.))/sum(.))})
#             rf_route_old_reduced
# rf_route_new    a    b    x
#            a 1133  221   39
#            b  303 1285   53
#            x  111   32  234
# 0.7774846
#            rf_route_old_reduced
# pc_rf_route    a    b    x
#           a 1080  109   45
#           b  307 1360   65
#           x  160   69  215
# 0.7785924
#             rf_route_old_reduced
# pc_lda_route    a    b    x
#            a  931  104   32
#            b  411 1284   65
#            x  205  150  228
# 0.7164223



inn[inn=="x"] <- NA
new[new=="x"] <- NA
old[old=="x"] <- NA

inn <- inn %>% 
  mutate(across(where(is.factor), as.character))

old <- old %>% 
  mutate(across(where(is.factor), as.character))

new <- new %>% 
  mutate(across(where(is.factor), as.character))


table(tmem_new=inn$rs17421627_tmem161b_new, tmem_old= inn$rs17421627_tmem161b_old)
#        tmem_old
# tmem_new    0    1    2
#        0 3248    0    0
#        1    0 1014    0
#        2    0    0   70

table(old$rs17421627_tmem161b)
#  0    1    2 
# 3263 1022   72 

table(new$rs17421627_tmem161b)
#   0    1    2 
# 3248 1014   70 

table(old$rs17421627_tmem161b, old$rf_route_reduced) %>% 
  {print(.);chisq.test(.)}
#      a   b
#   0 803 750
#   1 303 190
#   2  17  10
# X-squared = 15.192, df = 2, p-value = 0.0005023

table(new$rs17421627_tmem161b, new$rf_route) %>% 
  {print(.);chisq.test(.)}
#       a   b
#   0 724 779
#   1 271 216
#   2  12  18
# X-squared = 9.4064, df = 2, p-value = 0.009066

table(new$rs17421627_tmem161b, new$pc_rf_route) %>% 
  {print(.);chisq.test(.)}
#    a   b
#   0 607 811
#   1 233 223
#   2  11  17
# X-squared = 9.9327, df = 2, p-value = 0.006968

table(new$rs17421627_tmem161b, new$pc_lda_route) %>% 
  {print(.);chisq.test(.)}
#       a   b
#   0 521 832
#   1 212 233
#   2   9  15
# X-squared = 11.677, df = 2, p-value = 0.002914

table(inn$rs17421627_tmem161b_new>0, inn$rf_route_reduced) %>% 
  {print(.);chisq.test(.)}
#           a   b
#   FALSE 802 745
#   TRUE  317 198
# X-squared = 14.293, df = 1, p-value = 0.0001564

table(inn$rs17421627_tmem161b_new>0, inn$rf_route_new) %>% 
  {print(.);chisq.test(.)}
#           a   b
#   FALSE 724 779
#   TRUE  283 234
# 
# X-squared = 6.3788, df = 1, p-value = 0.01155

table(inn$rs17421627_tmem161b_new>0, inn$pc_rf_route) %>% 
  {print(.);chisq.test(.)}
#           a   b
#   FALSE 607 811
#   TRUE  244 240
# X-squared = 8.1395, df = 1, p-value = 0.004331

table(inn$rs17421627_tmem161b_new>0, inn$pc_lda_route) %>% 
  {print(.);chisq.test(.)}
#           a   b
#   FALSE 521 832
#   TRUE  221 248
# X-squared = 10.353, df = 1, p-value = 0.001293

str(new)

table(old$rs17421627_tmem161b) %>% 
  {./sum(.)} %>% 
  round(3)
#     0     1     2 
# 0.749 0.235 0.017 

table(new$rs17421627_tmem161b) %>% 
  {./sum(.)} %>% 
  round(3)
#     0     1     2 
# 0.750 0.234 0.016 

```


## do different routes have different pp progression patterns?
```{r}

pp_d <- colData(sce) %>% 
  data.frame() %>% 
  select(sample, patID_eye, month, pseudoprog, pc_lda_route) %>% 
  group_by(patID_eye) %>% 
  arrange(month) %>% 
  mutate(delta_bl_pp = pseudoprog - dplyr::first(pseudoprog)) %>% 
  filter(n()>1) %>% 
  ungroup()

pp_d$month_bin <- round(pp_d$month, -1)  # round month to nearest 10

medians_df <- pp_d %>%
  group_by(pc_lda_route, month_bin) %>%
  summarise(median_pp = median(delta_bl_pp), .groups = "drop")

# Step 2: plot with boxplots and median lines
ggplot(pp_d) +
  geom_boxplot(aes(x = month_bin, y = delta_bl_pp, group = interaction(month_bin, pc_lda_route), fill = pc_lda_route), color = rgb(0,0,0,.4),
               position = position_dodge(width = 5), width = 5, alpha = .3,outliers = FALSE) +
  # geom_line(data = medians_df,aes(x = month_bin, y = median_pp, group = pc_lda_route, color = pc_lda_route), linewidth = 2, alpha = .4,linetype = 2) +
  geom_smooth(aes(x = month, y = delta_bl_pp, color = pc_lda_route),
              # method = "gam", formula = y ~ s(x,k = 10, bs = "cs"), se = TRUE, cex = 2) +
              method = "lm", se = TRUE, cex = 2) +
  scale_x_continuous(breaks = sort(unique(pp_d$month_bin))) +
  ggtitle("Route Pseudoprogression Pattern") +
  xlab("Month") +
  ylab("Delta from Baseline") +
  theme_bw()
ggsave(paste0(wd_path, "results/clustering/13_clustering/rf_routes_pp_deltaBL.png"), height = 5,width = 8)

```


## does BL age determine trajectory?
```{r, eval = FALSE}

###
# what are bl ages of eyes that start in 1/2 and go to 333/444? ----
###
cluster_levels <- as.character(c(1 ,2 ,3, 4,6,9,12,111,999))

# grab eye age cluster and route info; make route C>A and D>B, filter for A/B
eyes_simplified_route_d <- 
  colData(sce) %>% 
  as.data.frame() %>% 
  dplyr::select( patID_eye, patID, month, AGE, BL_eye_age, simple_cluster, rf_route, path, SEXF)  %>% 
  mutate(origin = str_split(path, "-") %>% 
           sapply(function(x)head(x,1)) %>% 
           factor(., levels = as.character(cluster_levels))
    ) %>% 
  mutate(terminus = str_split(path, "-") %>% 
           sapply(function(x)tail(x,1)) %>% 
           factor(., levels = as.character(cluster_levels))
    ) %>% 
  mutate(rf_route = 
           case_when(
             rf_route == "c" ~ "a", 
             rf_route == "d" ~ "b", 
             TRUE ~ rf_route
           ) %>% factor()) %>% 
  filter(rf_route%in% c("a", "b")) %>% 
  mutate(count_vis = n(),.by = patID_eye)

# with ^ grab 1 record per eye; filter for first cluster in 1/2 and last in 333/444
eyes_meet_criteria_d <-
  eyes_simplified_route_d %>% 
  distinct(patID_eye, .keep_all = TRUE) %>% 
  filter(origin %in% c("3","4","6") & terminus %in% c("111", "999")) 

# look at how filtering changes age distribution
ggplot() + 
  geom_density(data = eyes_simplified_route_d%>% filter(origin %in% c("3","4","6") & terminus %in% c("111", "999")),
               aes(x = BL_eye_age, color = terminus), linetype = 3, linewidth = 1.5)+
  geom_density(data =eyes_meet_criteria_d ,
               aes(x = BL_eye_age, color = terminus))+
  ggtitle("Difference in BL age distribution when filtering for distinct eyes (solid) and when not (dotted)")

ggplot() + 
  geom_density(data =eyes_meet_criteria_d ,
               aes(x = BL_eye_age, color = terminus))+
  ggtitle("Difference in BL age distribution when filtering for distinct eyes")

# count visits for 333 at each age and for 444 at each age
eyes_simplified_route_d %>% 
  filter(origin %in% c("1","2") & terminus %in% c("333", "444")) %>% 
  ggplot(aes(BL_eye_age))+
  geom_density(aes(color = factor(count_vis)))+
  facet_wrap(~terminus, ncol = 1)+
  ggtitle("Eyes' BL age densities for different number of visits and different termini; multiple records per patient")

# # does starting age predict route?; using multiple records per eye
# glmer(rf_route ~ BL_eye_age + SEXF + (1|patID) + (1|patID_eye) ,
#       data = eyes_simplified_route_d,
#       "binomial"
#       ) %>% 
#   summary()
# 
# # does starting age predict route?; using 1 record per eye; eyes follow {1,2} >>{333,444}
# glmer(rf_route ~ BL_eye_age + SEXF + (1|patID),
#       data = eyes_meet_criteria_d,
#       "binomial"
#       ) %>% 
#   summary()

# distribution of age for eyes that start in 1/2 and end in 333/444
eyes_meet_criteria_d %>% 
  ggplot(aes(color = terminus ))+
  geom_density(aes(AGE), linetype  = 3)+
  geom_density(aes(BL_eye_age))+
  ggtitle("Eyes age at baseline (solid) and at visit (dotted) where first cluster in {1,2} and last in {333,444}")


# test if difference in age means between eyes that end in 333 versus 444 from {1,2}
t.test(eyes_meet_criteria_d$BL_eye_age[eyes_meet_criteria_d$terminus == "999"] , eyes_meet_criteria_d$BL_eye_age[eyes_meet_criteria_d$terminus == "111"])


```

## which values were imputed
```{r}

# add assay to sce
assays(sce)$was_imputed <- is.na(assays(sce)$measure_vars)

```

## how many eyes and patients
```{r}

# counts of eye demographics
counts <- as.data.frame(rbind(
count_patients = n_distinct(sce$patID),
count_eyes = n_distinct(sce$patID_eye),
count_eye_several_visits = sum(table(sce$patID_eye) > 1),
count_eye_single_visit = sum(table(sce$patID_eye)==1),
count_patients_with_route = n_distinct(sce$patID[ !is.na(sce$rf_route)]),
count_eyes_with_route = n_distinct(sce$patID_eye[ !is.na(sce$rf_route)])
))

patID_eyes <- unique(cbind(eyes = sce$patID_eye, pats = sce$patID))

count_single_eyes <- sum(table(patID_eyes[,"pats"]) ==1)
counts <- rbind(counts, count_single_eyes= count_single_eyes)

names(counts ) <- "count"
counts$stat <- rownames(counts)


par(mfrow = c(1,1), cex = 1, mar = c(4,3,3,3))
png(paste0(wd_path, "results/clustering/13_clustering/eye_patient_counts_bar.png"), 1500, 1000)
bp <- barplot(sort(counts$count) ,
        names.arg = sort(counts$stat),
        las = 2,
        main = "Count eyes and patients with and without RF route classification"
        )

text(bp, sort(counts$count) - .5*(sort(counts$count)), labels = counts$count)
dev.off()


```

## Plot Distributions of Classifications for each model
* manual
* rf on path features
* rf on phenotypic data
* logical assignment
```{r}

# eyes with route assignments and scores; unique ( still allows > 1 where one might be NA)
eye_routes <- 
  colData(sce)[, grep("patID_eye|route|score|flex|strict", names(colData(sce) ) )] %>% 
  as.data.frame() %>% 
  dplyr::select( -matches("A.score|B.score|norm|pheno")) %>% 
  unique()

eye_routes <- eye_routes[, c("patID_eye", sort(names(eye_routes)[-1]))]

eye_routes$patID <- sapply(str_split(eye_routes$patID_eye, "_"), `[`, 1)
eye_routes$eye <- sapply(str_split(eye_routes$patID_eye, "_"), `[`, 2)

eye_routes$eye <- ifelse(eye_routes$eye == "1", "right", 
                         ifelse(eye_routes$eye == "2", "left", 
                                eye_routes$eye))
str(eye_routes)

# fill missing with last observed or next non-missing observation to ensure only one record per eye
eye_routes <- eye_routes %>% 
  arrange(patID_eye) %>% 
  group_by(patID_eye) %>%
  fill(everything(), .direction = "downup") %>% 
  unique() %>% 
  ungroup(patID_eye) 
  

##---
# Distribution of eye route score assignments
##---
score_models <- names(eye_routes) %>% 
  grep("A.minus.B|score", ., value = TRUE) %>% 
  grep("\\.score|norm|pheno", ., value = TRUE, invert = TRUE)

classification_models <- names(eye_routes) %>% 
  grep("route", ., value = TRUE) %>% 
  grep("score", ., value = TRUE, invert = TRUE)

route_score_density_plot_d <- eye_routes %>% 
  dplyr::select( all_of(score_models)) %>% 
  reshape2::melt(variable.name = "model", value.name = "value") %>% 
  mutate(model = factor(model, levels = c("route_score","rf_route_score",   "flexible.A.minus.B","strict.A.minus.B") )) 

route_score_density_plot <- route_score_density_plot_d %>% 
  ggplot(aes(x = value, color = model ))+
  geom_density(show.legend = FALSE, )+
  # stat_density(show.legend = FALSE, kernel = "rectangular", alpha = 0)+
  geom_rug()+
  ggtitle("Eye route score densities")+
  facet_wrap(~model, scales = "free", ncol = 1 )+
  theme_bw()

ggsave(paste0(wd_path, "results/clustering/13_clustering/route_score_densities.png"), route_score_density_plot, width = 10, height = 10)

##---
# Distribution of Eye route classification
##---
eye_route_classifications <- eye_routes %>%
  dplyr::select( all_of(classification_models)) %>% 
  pivot_longer(cols = everything(), names_to = "model", values_to = "route") %>% 
  mutate(model = factor(model, levels = c("route","rf_route","flexible.route" ,  "strict.route") ))

route_class_bar <- eye_route_classifications %>% 
  ggplot(aes(x = route, fill = model ))+
  facet_wrap(~model, scales = "fixed", ncol = 1 )+
  geom_bar(stat = "count",show.legend = FALSE)+
  ggtitle("Eye route counts")+
  theme_bw()

ggsave(paste0(wd_path, "results/clustering/13_clustering/route_bars.png"), route_class_bar, width = 10, height = 10)


rf_route_freq_tab <- colData(sce) %>% 
  as.data.frame() %>% 
  dplyr::select( patID_eye, rf_route) %>% 
  unique() %>% 
  pull(rf_route) %>% 
  table() 
  
rf_route_bp <- barplot(rf_route_freq_tab, main = "RF route frequencies for unique eyes", ylim = c(0, max(rf_route_freq_tab)+300)) 
text(rf_route_bp, rf_route_freq_tab-(rf_route_freq_tab*.5), labels = rf_route_freq_tab )


```

## Do patients' eyes follow the same route? 
### correlation between left and right
```{r}

# ***************FUNCTIONS ************************************************************
calc.accuracy <- function(sqr_matrix){
  accuracy <-  sum(diag(sqr_matrix)) / sum(sqr_matrix)
  return(accuracy)
}

line_break <- function(...) {
  args <- list(...)
  if (length(args) > 0) {
    x <- args[[1]]
  } else {
    x <- "*   "
  }
  cat("\n")
  cat(paste(rep(x, 10), collapse = ""))
  cat("\n\n")
}

# add row, column and total sums to matrix
add_sums <- function(mat) {
  # Calculate row sums and column sums
  row_sums <- rowSums(mat)
  col_sums <- colSums(mat)
  total_sum <- sum(mat)
  
  # Create a new matrix with an extra row and column for the sums
  extended_mat <- rbind(cbind(mat, rowsums = row_sums), colsums = c( col_sums, total_sum))
  
  return(extended_mat)
}

# *************** ************************************************************

sink(file = "~/Documents/LMRI/PRJ2024001/version015/results/clustering/13_clustering/route_tables_and_correlations.txt", append = FALSE)

cat("\n")
print(Sys.time())
line_break("°°°° °°°°")

print("rows = left, cols = right")
line_break("°°°° °°°°")




#____
# route left/right correlation
#____
# make df of patID, right rf_route and left rf_route
route_eyes <- eye_routes %>%
  dplyr::select( patID, eye, route) %>% 
  unique() %>% 
  pivot_wider(id_cols = patID, names_from = eye, values_from = route)

# do right route = left routes
route_table <- table(route_eyes[, c("left", "right")])
route_table_accu <- calc.accuracy(route_table)
route_table <-route_table %>% add_sums

# make df of score for right and left
route_score_eyes <- eye_routes %>%
  dplyr::select( patID, eye, route_score) %>% 
  unique() %>%
  pivot_wider(id_cols = patID, names_from = eye, values_from = route_score)

# how to right scores correlate to left
print("My Route")
route_table
print(paste("route accuracy = ", route_table_accu))
print(paste("route_score correlation = ", cor( route_score_eyes$right, route_score_eyes$left , use = "complete.obs")))
line_break("°°°° °°°°")

#____
# rf_route left/right correlation
#____
# make df of patID, right rf_route and left rf_route
rf_route_eyes <- eye_routes %>%
  dplyr::select( patID, eye, rf_route) %>% 
  unique() %>% 
  pivot_wider(id_cols = patID, names_from = eye, values_from = rf_route)

# do right route = left routes
rf_route_table <- table(rf_route_eyes[, c("left", "right")]) 
rf_chsq <- chisq.test(rf_route_table)
rf_route_table_accu <- calc.accuracy(rf_route_table)
rf_route_table <- rf_route_table%>% add_sums

# # make df of rf_route_score for right and left
# rf_score_eyes <- eye_routes %>%
#   dplyr::select( patID, eye, rf_route_score) %>% 
#   unique() %>%
#   pivot_wider(id_cols = patID, names_from = eye, values_from = rf_route_score)

# how to right scores correlate to left
print("RF Route")
rf_route_table
rf_chsq
# print(paste("rf_route accuracy = ", rf_route_table_accu))
# print(paste("rf_route_score correlation = ", cor( rf_score_eyes$right, rf_score_eyes$left , use = "complete.obs")))
line_break("°°°° °°°°")
#____
# strict logic left/right correlation
#____

# make df of patID, right rf_route and left rf_route
strict_route_eyes <- eye_routes %>%
  dplyr::select( patID, eye, strict.route) %>% 
  unique() %>% 
  pivot_wider(id_cols = patID, names_from = eye, values_from = strict.route)

# do right route = left routes
strict_route_table <- table(strict_route_eyes[, c("left", "right")])  
strict_chsq <- chisq.test(strict_route_table)
strict_route_table_accu <- calc.accuracy(strict_route_table)
strict_route_table <-  strict_route_table  %>% add_sums

# make df of score for right and left
strict_route_score_eyes <- eye_routes %>%
  dplyr::select( patID, eye, strict.A.minus.B) %>% 
  unique() %>%
  pivot_wider(id_cols = patID, names_from = eye, values_from = strict.A.minus.B)

# how to right scores correlate to left
print("Strict Route")
strict_route_table
strict_chsq
print(paste("strict.route accuracy = ", strict_route_table_accu))
print(paste("strict.A.minus.B correlation = ", cor( strict_route_score_eyes$right, strict_route_score_eyes$left , use = "complete.obs")))
line_break("°°°° °°°°")

#____
# flexible logic left/right correlation
#____

# make df of patID, right rf_route and left rf_route
flexible_route_eyes <- eye_routes %>%
  dplyr::select( patID, eye, flexible.route) %>% 
  unique() %>% 
  pivot_wider(id_cols = patID, names_from = eye, values_from = flexible.route)

# do right route = left routes
flexible_route_table <- table(flexible_route_eyes[, c("left", "right")]) 
flex_chsq <- chisq.test(flexible_route_table)
flexible_route_table_accu <- calc.accuracy(flexible_route_table)
flexible_route_table <- flexible_route_table%>% add_sums

# make df of score for right and left
flexible_route_score_eyes <- eye_routes %>%
  dplyr::select( patID, eye, flexible.A.minus.B) %>% 
  unique() %>%
  pivot_wider(id_cols = patID, names_from = eye, values_from = flexible.A.minus.B)

# how to right scores correlate to left
print("Flexible Route")
flexible_route_table
flex_chsq
print(paste("flexible.route accuracy = ", flexible_route_table_accu))
print(paste("flexible.A.minus.B correlation = ", cor( flexible_route_score_eyes$right, flexible_route_score_eyes$left , use = "complete.obs")))
line_break("°°°° °°°°")

### MAKE NEW ROUTE classification from flexible.A.minus.B

## bin near 0 to Z and other peaks A and B respectively
flexible_score_dens <- density(eye_routes$flexible.A.minus.B, na.rm = TRUE)
# Find minima
y_values <- flexible_score_dens$y

# funtion to return inds of minima in a vector
delta <- 2
minima.ind <- function(values){ return(  which(diff(sign(diff(values))) == 2) + 1  ) }
maxima.ind <- function(values){ return(  which(diff(sign(diff(values))) == -2) + 1  ) }

# find minima of flexible score density-----

# find the minima; this is the threshold to cut for A, Z, B
minima_indices <- minima.ind(y_values)
maxima_indices <- maxima.ind(y_values)

minima_values <- flexible_score_dens$x[minima_indices]
maxima_values <- flexible_score_dens$x[maxima_indices]

# plot density with minima
plot(flexible_score_dens, main="Density Plot with Minima and Maxima")
points(minima_values, y_values[minima_indices], col="blue", pch=19)
points(maxima_values, y_values[maxima_indices], col="red", pch=19)

threshold <- round(mean(sort(abs(minima_values))[1:2]),2)
abline(v = c(-threshold, threshold), lty = 3)



# make new class X for within minima interval ~0
flexible_route_eyes_new <- eye_routes %>%
  mutate(flexible.route.new = 
           case_when(
             flexible.A.minus.B <threshold~ "b", 
             flexible.A.minus.B >threshold ~ "a", 
             flexible.A.minus.B < threshold[2] & flexible.A.minus.B > threshold[1] ~ "x", 
             
           )) %>% 
  dplyr::select( patID, eye, flexible.route.new) %>% 
  unique() %>% 
  pivot_wider(id_cols = patID, names_from = eye, values_from = flexible.route.new)

# do right route = left routes
flexible_route_eyes_new_table <- table(flexible_route_eyes_new[, c("left", "right")] ) 
flexible_route_eyes_new_chisq <- chisq.test(flexible_route_eyes_new_table)
flexible_route_eyes_new_table_accu <- calc.accuracy(flexible_route_eyes_new_table)
flexible_route_eyes_new_table <- flexible_route_eyes_new_table %>% add_sums 

# how do right scores correlate to left
print("New Flexible Route")
flexible_route_eyes_new_table
flexible_route_eyes_new_chisq
print(paste("New flexible.A.minus.B accuracy = ", flexible_route_eyes_new_table_accu))
line_break("°°°° °°°°")

sink()

```

#### plot rf_route right/left table and scores
```{r, }

#____
# pc_lda_route left/right correlation
#____
# make df of patID, right pc_lda_route and left pc_lda_route; make A and C same, ditto for B,D
pc_lda_route_eyes <- eye_routes %>%
  dplyr::select( patID, eye, pc_lda_route) %>% 
  mutate(pc_lda_route = as.character(pc_lda_route)) %>% 
  mutate(pc_lda_route = case_when(
             pc_lda_route== "c" ~"a", 
             pc_lda_route == "d" ~"b", 
             TRUE~ pc_lda_route
  )) %>% 
  unique() %>% 
  pivot_wider(id_cols = patID, names_from = eye, values_from = pc_lda_route)

# do right route = left routes
pc_lda_route_table <- table(pc_lda_route_eyes[, c("left", "right")]) 
pc_lda_chsq <- chisq.test(pc_lda_route_table)
pc_lda_route_table_accu <- calc.accuracy(pc_lda_route_table)

# # make df of pc_lda_route_score for right and left
# rf_score_eyes <- eye_routes %>%
#   dplyr::select( patID, eye, pc_lda_route_score) %>% 
#   unique() %>%
#   pivot_wider(id_cols = patID, names_from = eye, values_from = pc_lda_route_score)

# how to right scores correlate to left
print("RF Route")
pc_lda_route_table
pc_lda_chsq
print(paste("pc_lda_route accuracy = ", pc_lda_route_table_accu))
# print(paste("pc_lda_route_score correlation = ", cor( rf_score_eyes$right, rf_score_eyes$left , use = "complete.obs")))

dimnames(pc_lda_route_table) <- list(paste("Left ", c("a", "b", "x"), sep = ""), paste("Right ", c("a", "b", "x"), sep = ""))

pc_lda_route_table_tile <-pc_lda_route_table %>% 
    reshape2::melt(varnames = c("Left", "Right")) %>% 
    mutate_all(~gsub("Right|Left","",.)) %>% 
    mutate(value = as.numeric(value)) %>% 
    ggplot(aes(Right, Left, fill = value, label = value))+
    geom_tile(show.legend = FALSE)+
    geom_text()+
    theme_minimal()+
    theme(aspect.ratio = 1)+
    scale_fill_continuous(high = "red", low = "white")+
    ggtitle("Patient trajectory eye symetry",sprintf("X^2 = %.0f, p-value = %.02e", pc_lda_chsq$statistic, pc_lda_chsq$p.value) )+
    geom_blank()

ggsave(paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_assignment_contingency_table.pdf"), pc_lda_route_table_tile, height = 5, width = 5)

```

<!-- #__________________________ -->
<!-- # FA of assay  -->
<!-- ## make factors from assay -->
<!-- ```{r, eval = FALSE} -->

<!-- # grab assay of measure vars -->
<!-- assay <- assays(sce)$pmm_imp_of_scaled %>%  -->
<!--   t() %>%  -->
<!--   as.matrix() -->

<!-- # how many factors -->
<!-- # set.seed(10329) -->
<!-- # fa.parallel(assay) -->

<!-- # what variables contribute most to factors -->
<!-- msa <- psych::KMO(assay) -->
<!-- par(mar = c(8,4,2,2)) -->
<!-- msa$MSAi %>% sort() %>% barplot(las = 2, horiz = FALSE, cex.names = .5) -->
<!-- # vars <- names(msa$MSAi)[msa$MSAi>.6] -->

<!-- # Run FA wiht nfacs and specific rotation; pa to help with multicollinearity -->
<!-- nFacs <- 9 -->
<!-- method <- "mr" -->
<!-- FA <- psych::fa(assay, nfactors = nFacs, rotate = "varimax", fm =method) -->

<!-- par(mar = c(8,4,2,2)) -->
<!-- FA$communality %>% sort %>% barplot(las = 2, ylim  =c(0,1), cex.names = .5) -->

<!-- # wahich variables do factors capture -->
<!-- loads <- FA$loadings  -->
<!-- loads %>% Heatmap(.) -->
<!-- factor_names <- colnames(loads) -->

<!-- # add factors to sce -->
<!-- colData(sce) <- colData(sce) %>%  -->
<!--   cbind(FA$scores) -->

<!-- ``` -->

### plot factors on TSNE
<!-- ```{r, eval = FALSE} -->

<!-- # # factor coloring TSNE -->
<!-- # plots <- list(); for(f in factor_names ){ -->
<!-- #   # plot factor without outliers -->
<!-- #   bx <- boxplot(sce[[f]], plot = FALSE)   -->
<!-- #   plots[[f]] <- plotReducedDim(sce[,sce[[f]]>bx$stats[1] & sce[[f]]<bx$stats[5] ]  , "TSNE", color_by  = f) -->
<!-- # } -->

<!-- ########    ####    ####    ####         -->
<!-- # INTERPOLATE FACTOR VALUE ACROSS TSNE1/2---- -->
<!-- ########    ####    ####    ####         -->

<!-- # make df of factor contributing phenos -->
<!-- f_loads <- loads -->
<!-- f_loads[abs(loads)<.4] <-  0 -->
<!-- # Heatmap(f_loads) -->

<!-- # vars pasted together for each factor -->
<!-- fac_vars <- apply(f_loads, 2, function(Col){  -->
<!--   nams <- names(Col)[Col>0] -->
<!--   if(length(nams) > 5) nams <- c(nams[1:5], "\n", nams[6:length(nams)]) -->
<!--   paste0(nams, collapse = "; ")  -->

<!-- } ) -->

<!-- # cluster centroids -->
<!-- centroids <- reducedDim(sce, "PCA") %>%  -->
<!--   as.data.frame() %>%  -->
<!--   rownames_to_column("sample") %>%  -->
<!--   left_join(colData(sce) %>%  -->
<!--               as.data.frame() %>%  -->
<!--               select(sample, my_cluster)) %>%  -->
<!--   select(-sample) %>%  -->
<!--   summarize(across(PC1:PC2, ~mean(.)), .by = my_cluster) -->


<!-- factor_names <- c(factor_names, "pseudoprog") %>%  -->
<!--   unique() -->

<!-- filter_outliers = FALSE -->
<!-- var = "pseudoprog" -->
<!-- for(filter_outliers in c(TRUE, FALSE)){ -->
<!--   # loop through factors and plot interpolation across tsne -->
<!--   interp_FA_plots <- mclapply(factor_names, function(var){ -->

<!--     # select data -->
<!--     dat_df <- reducedDim(sce, "PCA") %>%  -->
<!--       as.data.frame() -->

<!--     pcs <- setdiff(grep("PC", names(dat_df),value = TRUE), c("PC1", "PC2")) -->

<!--     dat_df <- dat_df%>%  -->
<!--       select(-all_of(pcs)) %>%  -->
<!--       rownames_to_column("sample") %>%  -->
<!--       left_join(colData(sce) %>%  -->
<!--                   as.data.frame() %>%  -->
<!--                   select(sample, all_of(var))) %>%  -->
<!--       select(-sample) -->

<!--     # character string to add to image filename -->
<!--     filter_outliers_char <- if(filter_outliers) "filtered" else "unfiltered" -->

<!--     # filter out outlieres -->
<!--     if(filter_outliers) { -->
<!--       # make whiskers -->
<!--       bx <- boxplot(dat_df[,var], plot = FALSE)   -->
<!--       # filter factor -->
<!--       dat_df <-  dat_df[dat_df[,var]>bx$stats[1] & dat_df[,var]<bx$stats[5] ,] -->
<!--     } -->

<!--     # add cushion to borders of PCA by imputing values -->

<!--     # ## lm -->
<!--     # lm_form <- as.formula( -->
<!--     #   paste0(var, " ~ PC1*PC2") -->
<!--     # ) -->
<!--     # mod <- lm(form, data = dat_df) -->

<!--     ## GAM -->
<!--     gam_form <- as.formula( -->
<!--       paste0(var, " ~ s(PC1)+s(PC2)") -->
<!--     ) -->
<!--     mod <- gam(gam_form, data = dat_df) -->



<!--     # how far to simulate points to then predict wiht model -->
<!--     min_border <- min(max(abs(dat_df$PC1)), max(abs(dat_df$PC2)) ) %>% round() -->
<!--     max_border <- max(max(abs(dat_df$PC1)), max(abs(dat_df$PC2)) ) %>% round() -->

<!--     # how much past actual limit to simulate -->
<!--     add = 10 -->
<!--     # how many points -->
<!--     n = 200 -->
<!--     # simulated points -->
<!--     sample_vec1 <- sample(seq(min_border,(min_border+add),1), n, T) * sample(c(-1,1), n,T) -->
<!--     sample_vec2 <- sample(seq(min_border,(min_border+add),1), n,T) * sample(c(-1,1), n,T) -->
<!--     # predict 'var' from simulated points -->
<!--     sim_coords <- data.frame(PC1 = sample_vec1, PC2 = sample_vec2) -->
<!--     sim_coords[[var]] <- predict(mod, sim_coords) -->

<!--     # add predicted factor value for simpulated points -->
<!--     dat_df <- dat_df %>%  -->
<!--       rbind(sim_coords) -->


<!--     # Interpolate z values over a grid of x and y -->
<!--     interp_data <- akima::interp(dat_df$PC1, dat_df$PC2, dat_df[,var], duplicate = "mean") -->

<!--     # Convert the interpolated grid to a data frame -->
<!--     grid_data <- with(interp_data, data.frame( -->
<!--       x = rep(x, length(y)), -->
<!--       y = rep(y, each = length(x)), -->
<!--       z = as.vector(z) -->
<!--     )) -->

<!--     # Plot the data points and the contours for z; filter for PCA within pre specified range -->
<!--     plot <- dat_df %>% -->
<!--       # filter((PC1 > -max_border & PC2 > -max_border) & (PC1 <max_border & PC2 <max_border)) %>%  -->
<!--       ggplot( aes(PC1,PC2)) + -->
<!--         geom_point(aes(color = !!sym(var)), alpha = 0.01) + -->
<!--         scale_color_viridis_c()+ -->
<!--         geom_contour_filled(data = grid_data, -->
<!--                             # %>% -->
<!--                               # filter((x > -max_border & y > -max_border) & (x <max_border & y <max_border)) ,  -->
<!--                             aes(x,y,z=z), bins = 20, show.legend = FALSE ) + -->
<!--         # geom_point(data = dat_df %>% distinct(PC1,PC2), aes(PC1,PC2), color = "black", alpha = 1e-1, shape =1) + -->
<!--         geom_density2d(color = "black", alpha = .2, bins = 20)+ -->
<!--         # geom_point(color = "black", alpha = 1)+ -->
<!--         labs(color = var, title = paste(filter_outliers_char,'\n', var," loadings for interp. > 0.4 vars{\n",fac_vars[var],"\n}\nWith point density.") )+ -->
<!--         theme_bw()+ -->
<!--         geom_label(data = centroids, aes(PC1,PC2, label = my_cluster), fill = "black", color = "white", size = 3)+ -->
<!--         theme(aspect.ratio = 1) -->

<!--     interp_FA_plots[[var]] <- plot -->
<!--   }, mc.cores = detectCores()-1) -->

<!--   interp_FA_plots_grid <- marrangeGrob(sapply(interp_FA_plots,ggplotGrob), ncol = 5, nrow =2, top =  "") -->
<!--   ggsave(paste0(wd_path, "results/clustering/13_clustering/",filter_outliers_char,"_FA_factors_PC_interpolation_grid.png"), interp_FA_plots_grid, height = 15, width = 40) -->
<!-- } -->

<!-- ``` -->


#__________________________
# Route
## route
###~firstPhenos prognostic model
for eyes that start in early cluster
    
    - take eyes that starts in early cluster
    - with first obs, model what phenotypes correlate to route
#### make modeling data
```{r}

# x <- sapply(as.list(colData(sce)[, snps]), function(snp_vec){
#   coef <- lm(as.matrix(colData(sce)$t2d)~snp_vec + as.matrix(colData(sce)[, paste0("PC", 1:5)])) %>% 
#     summary() %>% 
#     {.$coef}  
#   rownames(coef) <- str_remove(rownames(coef), "as.matrix\\(.*\\)")
#   return(coef[2,])
# }, simplify = TRUE) %>% 
  # t()
  
# what to feed into models
predictors <-  c(rownames(assays(sce)$pmm_imp_of_scaled))
response <- c("pc_lda_route")
covariates <- c("SEXF", "(1|patID)", "(1|patID_eye)", "AGE", "BMI", "SEXF")

# make model data; scale predictors and covs in colData
pheno_route_d <-  colData(sce) %>% 
  as.data.frame() %>% 
  left_join(
    assays(sce)$pmm_imp_of_scaled %>%
      t() %>%
      as.data.frame() %>% 
      rownames_to_column("sample")
  ) %>% 
  dplyr::select( patID_eye, patID, month, my_cluster, 
                 chew_grade_derived_mine_loose,
                 all_of(c(response, predictors,covariates[covariates%in%names(colData(sce))])) )

# scale predictors and covariates
pheno_route_d_scaled <- pheno_route_d %>%
  mutate(across(all_of
                (c(
                  predictors, 
                  covariates[covariates%in%names(colData(sce))]
                  )), ~ as.numeric(scale(.))))

# seelect first observations and chew <=2 at first visit
pheno_route_d_scaled_filtered <- pheno_route_d_scaled %>% 
  arrange(patID_eye,month) %>% 
  group_by(patID_eye) %>% 
  dplyr::slice(1) %>% 
  # filter(pc_lda_route %in% c("a", "b")) %>%
  
  # # filter for starting in early cluster
  # filter(my_cluster%in%c(3,6,4)) %>% 
  
  # filte for starting with minor disease, chew < n
  filter(dplyr::first(chew_grade_derived_mine_loose)<=2) %>%
  
  # # filter for one route patients
  # ungroup() %>% group_by(patID) %>%  filter(n_distinct(pc_lda_route)==1) %>% select(-patID_eye) %>% distinct() %>% 
  
  ungroup() 

# merge routes: collapse A+C and B+D, X to NA so reference is A
pheno_route_d_scaled_filtered <-
  pheno_route_d_scaled_filtered %>%
  mutate(across(pc_lda_route, ~factor(., levels = c("a", "b", "c", "d", "x")))) %>%
  mutate(across(pc_lda_route,
                ~ case_when(
                  . %in% c("a", "c") ~ "a",
                  . %in% c("b", "d") ~ "b",
                  TRUE ~ NA_character_
                )
  )) %>%
  droplevels() %>% 
  filter(!pc_lda_route=="x")

nrow(pheno_route_d_scaled_filtered)
# 851

# check if some patients have >1 route
pheno_route_d_scaled_filtered %>% 
  filter(n_distinct(pc_lda_route) > 1, .by = patID)

pheno_route_d_scaled_filtered %>% 
  {colSums(is.na(.))} %>% table()


```

<!-- ##### plot feature distributions  -->
<!-- ```{r} -->

<!-- dat <- pheno_route_d_scaled_filtered -->

<!-- feat_route_fit <-  -->
<!--   # glm(pheno_route_d_scaled_filtered$pc_lda_route~as.matrix(pheno_route_d_scaled_filtered[, predictors]), family = "binomial") %>%  -->
<!--   lm(as.matrix(dat[, predictors]) ~ as.numeric(dat$pc_lda_route)) %>%  -->
<!--   summary() %>%  -->
<!--   coef() %>%  -->
<!--   as.list() -->

<!-- feat_route <- lapply(names(feat_route_fit), function(j){ -->
<!--   z <- as.data.frame(feat_route_fit[[j]]) -->
<!--   z$response <- str_remove(j, "Response\\s+") -->
<!--   return(z) -->
<!-- }) %>%  -->
<!--   do.call(rbind, .) -->

<!-- feat_route <- as.data.frame(feat_route) -->

<!-- feat_route$term <- janitor::make_clean_names(rownames(feat_route)) -->
<!-- rownames(feat_route) <- seq(nrow(feat_route)) -->
<!-- feat_route <- feat_route[!grepl("intercept", feat_route$term), ] -->

<!-- # feat_route$p <- feat_route$`Pr(>|z|)` -->
<!-- # feat_route$`Pr(>|z|)` <- NULL -->
<!-- feat_route$p <- feat_route$`Pr(>|t|)` -->
<!-- feat_route$`Pr(>|t|)` <- NULL -->
<!-- feat_route$q <- p.adjust(feat_route$p, "BH") -->

<!-- route_feature_long <- pheno_route_d_scaled %>% -->
<!--   select(-matches("pat|month|clust")) %>%  -->
<!--   melt(id.vars = "pc_lda_route") %>%  -->
<!--   left_join(feat_route, by = join_by(variable == response)) %>%  -->
<!--   filter(pc_lda_route%in% c("a", "b")) %>% -->
<!--   filter(p>0&!is.na(p))  -->

<!-- route_feature_boxplots <- route_feature_long %>%  -->
<!--   ggplot(aes(x = reorder(variable, -log10(q)),y = value,color = pc_lda_route))+ -->
<!--   geom_boxplot(outliers = TRUE, width = .3, size = .5)+ -->
<!--   theme_bw()+ -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1))+ -->
<!--   ggtitle("Feature distributions by route", "with q values of lm feature~route") -->

<!-- route_feature_est_bars <- route_feature_long %>%  -->
<!--   distinct(variable, Estimate,  q) %>%  -->
<!--   # ggplot(aes(x = reorder(variable, -log10(q)),y = Estimate, fill = abs(Estimate)))+ -->
<!--   ggplot(aes(x = reorder(variable, -log10(q)),y = Estimate, fill = abs(Estimate)))+ -->
<!--   geom_bar(stat = "identity", position = "dodge")+ -->
<!--   geom_text(aes(y = 0, label = sprintf("%.0e", q)), angle = 90)+ -->
<!--   theme_bw()+ -->
<!--   scale_fill_viridis_c()+ -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1))+ -->
<!--   ggtitle("Feature distributions by route", "with q values of lm feature~route") -->

<!-- # route_feature_est_bars -->

<!-- ``` -->

#### model
```{r}

resp <- factor(pheno_route_d_scaled_filtered[, response, drop = T])
resp
pred <- as.matrix(pheno_route_d_scaled_filtered[, predictors])
pred <- as.matrix(pheno_route_d_scaled_filtered[, predictors])
head(pred)


# cov <- as.matrix(pheno_route_d_scaled[, c("SEXF", "AGE", "BMI", "SEXF")])
cov <- as.matrix(pheno_route_d_scaled_filtered[, c("SEXF", "AGE")])
head(cov)

result_l <- mclapply(colnames(pred), function(nm){
  
  mod_d <- data.frame(
    response = resp, 
    predictor = pred[,nm], 
    cov, 
    patID = pheno_route_d_scaled_filtered$patID,
    chew = pheno_route_d_scaled_filtered$chew_grade_derived_mine_loose
  )
  
  # if >1 routes per patient
  mod <- try({glmer(response~predictor + SEXF+AGE + (1|patID), family = "binomial", data = mod_d)})
  # mod <- try({glmer(response~predictor*chew + SEXF+AGE + (1|patID), family = "binomial", data = mod_d)})
  
  # # if 1 route per patient, no need for ME RI
  # mod <- try({glm(response~predictor + SEXF+AGE, family = "binomial", data = mod_d)})
  # mod <- try({glm(response~predictor*chew + SEXF+AGE, family = "binomial", data = mod_d)})
  
  if(inherits(mod, "try-error")){
    return(NA)
  }else{
    coefs <- summary(mod)$coef
    rownames(coefs) <- str_replace_all(rownames(coefs), "predictor", nm)
    res <- coefs[rownames(coefs)==nm,, drop = FALSE]
    return(res)
  }
}, mc.cores = detectCores()-2 ) 

# remove failed models
names(result_l)[is.na(result_l)]

result_l <- result_l[!is.na(result_l)]

# make df
result <- do.call(rbind, result_l) %>% 
  as.data.frame
result$p <- result$`Pr(>|z|)`
result$`Pr(>|z|)` <- NULL
result$q <- p.adjust(result$p, "BH")
result$term <- rownames(result)

write_csv(result, sprintf("%sresults/publication_tables/%s",wd_path, "route_pheno.csv"))


route_pheno_volcano <-  result %>% 
  filter(p>0,p<.9) %>% 
  filter(abs(Estimate)<10) %>% 
  ggplot( aes(Estimate, -log10(q), color = term))+
  geom_point(show.legend = FALSE)+
  geom_label_repel(aes(label = ifelse(q < .05, term, NA)), 
                   show.legend = FALSE, max.overlaps = 100
                   ) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = 2) +
  theme_bw()+
  ggtitle("Route ~ Phenotype + SEXF + AGE + (1|patID)","For eyes with BL Chew <= 2" )+
  xlab("Estimate (<- = a trajectory, -> = b trajectory)")+
  ylab("-log10(q.value)")+
  theme(aspect.ratio = 1,plot.background = element_rect(fill = "transparent"))
route_pheno_volcano

ggsave(paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_pheno_volcano.pdf"),route_pheno_volcano,height = 7, width = 7)
sprintf("cp %s %s",
  paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_pheno_volcano.pdf"),
  paste0(wd_path, "results/poster_plots/pc_lda_route_pheno_volcano.pdf")
         ) %>% 
  system()


# patch <- route_feature_est_bars/route_feature_boxplots/route_pheno_volcano
# ggsave(paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_pheno_summary_grid.pdf"),patch,height = 20, width = 20)

sprintf("%sresults/publication_tables",wd_path) %>% 
  {if(!dir.exists(.)) dir.create(.)}


result %>% 
  filter(p<.05,q>0, abs(Estimate)<10) %>% 
  select(-term) %>% 
  kableExtra::kable("simple")

#                 Estimate    Std. Error     z value           p           q
# BRV02          -5.019302     2.0098506   -2.497351   0.0125125   0.0426563
# RAV02           1.991516     0.4089368    4.869985   0.0000011   0.0000042
# RPE_LK05       -7.217714     1.5416645   -4.681767   0.0000028   0.0000102

```


<!-- ### ~ covariates -->
<!-- for eyes that start in early cluster -->
<!-- #### make modeling data -->
<!-- ```{r} -->

<!-- # what to feed into models -->
<!-- response <- c("pc_lda_route") -->
<!-- covariates <- c("SEXF", "(1|patID)", "(1|patID_eye)", "AGE", "BMI", "SEXF", "t2d") -->

<!-- # make model data; scale predictors and covs in colData -->
<!-- cov_route_d <-  colData(sce) %>%  -->
<!--   as.data.frame() %>%  -->
<!--   dplyr::select( patID_eye, patID, month, my_cluster, chew_grade_derived_mine_loose, -->
<!--                  all_of(c(response, covariates[covariates%in%names(colData(sce))])) ) -->

<!-- # scale predictors and covariates -->
<!-- cov_route_d_scaled <- cov_route_d %>% -->
<!--   mutate(across(all_of(c(covariates[covariates%in%names(colData(sce))])),  -->
<!--                 ~ as.numeric(scale(.)))) -->

<!-- # seelect first observation; select eyes with first obs in early cluster or; filter for a or b route;  -->
<!-- cov_route_d_scaled_filtered <- cov_route_d_scaled %>%  -->
<!--   arrange(patID_eye,month) %>%  -->
<!--   group_by(patID_eye) %>%  -->
<!--   mutate(across(c(SEXF, t2d,AGE), ~ifelse(any(!is.na(.)), .[!is.na(.)], .))) %>% -->
<!--   dplyr::slice(1) %>%  -->
<!--   filter(pc_lda_route %in% c("a", "b")) %>%  -->

<!--   # # filter for starting in early cluster -->
<!--   # filter(my_cluster%in%c(3,6,4)) %>%  -->

<!--   # filte for starting with minor disease, chew < n -->
<!--   filter(dplyr::first(chew_grade_derived_mine_loose)<=2) %>%  -->

<!--   # # filter for one route patients -->
<!--   # ungroup() %>% group_by(patID) %>%  filter(n_distinct(pc_lda_route)==1) %>% select(-patID_eye) %>% distinct() %>%  -->

<!--   ungroup()  -->

<!-- # check if some patients have >1 route -->
<!-- cov_route_d_scaled_filtered %>%  -->
<!--   filter(n_distinct(pc_lda_route) > 1, .by = patID) -->

<!-- cov_route_d_scaled_filtered %>%  -->
<!--   {colSums(is.na(.))} -->

<!-- table(cov_route_d_scaled_filtered$patID_eye) %>% unique() -->


<!-- #### model -->

<!-- cov_route_mod_w_t2d <- glm(pc_lda_route ~ SEXF + AGE + t2d,  -->
<!--                      family = "binomial",  -->
<!--                      data = cov_route_d_scaled_filtered) # original -->

<!-- cov_route_mod <- glm(pc_lda_route ~ SEXF + AGE, -->
<!--                        family = "binomial", -->
<!--                        data = cov_route_d_scaled_filtered) # edited 20 Nov 2025 -->

<!-- # cov_route_mod <- glmer(pc_lda_route ~ scale(AGE) + SEXF + t2d  + (1|patID), family = "binomial", -->
<!-- #                        data = cov_route_d_scaled_filtered %>% -->
<!-- #                          mutate(across(c(SEXF, t2d),  ~ifelse(. == min(., na.rm = TRUE), F, ifelse(. == max(., na.rm = TRUE), T, .)) %>% as.logical()  )) -->
<!-- #                        ) -->

<!-- summary(cov_route_mod) -->
<!-- summary(cov_route_mod_w_t2d) -->

<!-- result <- data.frame(coef(summary(cov_route_mod))) -->
<!-- result$term <- rownames(result) -->
<!-- result <- janitor::clean_names(result) -->

<!-- t2d_result <- data.frame(coef(summary(cov_route_mod_w_t2d))) -->
<!-- t2d_result$term <- rownames(t2d_result) -->
<!-- t2d_result <- janitor::clean_names(t2d_result) -->

<!-- write_csv(result, sprintf("%sresults/publication_tables/%s",wd_path, "route_pheno_cov.csv")) -->
<!-- write_csv(t2d_result, sprintf("%sresults/publication_tables/%s",wd_path, "route_pheno_cov_w_t2d.csv")) -->

<!-- # route_cov_volcano <-  result %>%  -->
<!-- #   ggplot( aes(estimate, -log10(pr_z), color = term))+ -->
<!-- #   geom_point(show.legend = FALSE)+ -->
<!-- #   geom_label_repel(aes(label = ifelse(pr_z < .05, term, NA)), show.legend = FALSE) + -->
<!-- #   geom_vline(xintercept = 0, linetype = 2) + -->
<!-- #   geom_hline(yintercept = -log10(0.05), linetype = 2) + -->
<!-- #   theme_bw()+ -->
<!-- #   ggtitle("Route ~ SEXF + AGE + T2D + (1|patID)","For eyes with BL Chew <= 2" )+ -->
<!-- #   xlab("Estimate (<- = a trajectory, -> = b trajectory)")+ -->
<!-- #   ylab("-log10(q.value)")+ -->
<!-- #   theme(aspect.ratio = 1,plot.background = element_rect(fill = "transparent")) -->
<!-- #  -->
<!-- # route_cov_volcano -->
<!-- # ggsave(paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_cov_volcano.pdf"),route_pheno_volcano,height = 7, width = 7) -->
<!-- # sprintf("cp %s %s", -->
<!-- #   paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_cov_volcano.pdf"), -->
<!-- #   paste0(wd_path, "results/poster_plots/pc_lda_route_cov_volcano.pdf") -->
<!-- #          ) %>%  -->
<!-- #   system() -->


<!-- ``` -->


## route~snps
### model correcting for SEXF AGE and patID RE
```{r}

snps <- c("rs146953046_phgdh_hmgcs2","rs2120770_ppp2r5a_tmem206","rs532303_phgdh_hmgcs2","rs1047891_cps1","rs1260326_mpv17_gtf3c2_eif2b4_snx17_znf513_ppm1g_nrbp1_krtcap3_ift172_gckr_znf512","rs2160387_slc1a4","rs17279437_slc6a20_sacm1l_limd1_lars2","rs9820465_slc4a7_eomes","rs17421627_tmem161b","rs6955423_sept14_mrps17_znf713_gbas_psph_cct6a_sumf2_phkg1_chchd2_nupr1l","rs2954021_trib1","rs677622_ttc39b","rs10995566_nrbf2_jmjd1c_reep3","rs2984814_dcdc1_dnajc24_immp1l_elp4_pax6","rs11077850_mxra7_jmjd6_mettl23_srsf2_mfsd11","rs139412173_cers4_fbn3")

# make model data; 1 record per eye; fill sex and t2d, and then AGE
model_data <- colData(sce) %>% 
  as.data.frame() %>% 
  dplyr::select(pc_lda_route,month, patID_eye, patID, SEXF,AGE,t2d,
                chew_grade_derived_mine_loose,
                all_of(c(snps, paste0("PC",1:10)))) %>% 
  arrange(patID_eye,month) %>% 
  group_by(patID_eye) %>% 
  mutate(across(c(SEXF, t2d), ~ifelse(any(!is.na(.)), .[!is.na(.)], .))) %>% 
  dplyr::slice(1) %>%
  mutate(across(c(AGE), ~ifelse(any(!is.na(.)), .[!is.na(.)], .))) %>% 
  ungroup()
  
# scale snps and covs
model_data <- model_data %>% 
  mutate(across(
    all_of(c(snps, "AGE", "SEXF", "t2d")),
      ~ as.numeric(scale(.)))) 

  
# collapse A+C and B+D, X to NA so reference is A
model_data <- model_data %>% 
  mutate(across(pc_lda_route, ~factor(., levels = c("a", "b", "c", "d", "x")))) %>%
  mutate(across(pc_lda_route,
                ~ case_when(
                  . %in% c("a", "c") ~ "a",
                  . %in% c("b", "d") ~ "b",
                  TRUE ~ NA_character_
                )
  )) %>%
  filter(pc_lda_route!="x") %>%
  droplevels() 

# # filter for chew <= 2 
# model_data <- model_data %>% 
#   filter(dplyr::first(chew_grade_derived_mine_loose) <=2 )
# ^^^^ no observations

table(model_data$patID_eye) %>% table()

```



#### run model
```{r}

# GLMER PCS
res <- mclapply(snps, function(snp){
    
  mod <- glmer(
    (pc_lda_route=="b") ~ 
      model_data[,snp,drop = TRUE]  +
      SEXF + AGE+ t2d +(1|patID) + 
      PC1+PC2+PC2+PC3+PC4+PC5+PC6+PC7+PC8+PC9+PC10,
    data = model_data,
    family = "binomial")
  
  summ <- summary(mod)$coef
  summ[2,]
  }, mc.cores =(detectCores()-1)) %>% 
  do.call(rbind,.) %>% 
  as.data.frame()

rownames(res) <- snps
colnames(res) <- c( 'estimate' ,'std_error',"stat_val", "p")
res$model <- "GLMER"
res$PCs <- "YES"
res$stat <- "z"
res_all <- res[0,]
res_all <- rbind(res_all,res)
rm("res")


# # GLMER NO PCS
# res <- mclapply(seq(length(snps)), function(i){
#   mod <- glmer(model_data$pc_lda_route ~ snps_dat[,i]+ model_data$SEXF + model_data$BL_eye_age +(1|model_data$patID),family = "binomial")
#   summ <- summary(mod)$coef
#   summ[2,]
# }, mc.cores =(detectCores()-5)) %>% 
#   do.call(rbind,.) %>% 
#   as.data.frame
# rownames(res) <- snps
# colnames(res) <- c( 'estimate' ,'std_error',"stat_val", "p")
# res$model <- "GLMER"
# res$PCs <- "NO"
# res$stat <- "z"
# res_all <- rbind(res_all,res)
# rm("res")
# 
# # GLM PCS
# model_data2 <- model_data %>% 
#       filter(n_distinct(pc_lda_route)==1, .by = patID) %>%
#       select(-patID_eye) %>% 
#       distinct()
# snps_dat <- model_data2[,snps]
# pc_dat <- model_data2[,paste0("PC",1:10)]
# res <- mclapply(seq(length(snps)), function(i){
#   mod <- glm(model_data2$pc_lda_route ~ snps_dat[,i]+ model_data2$SEXF + model_data2$BL_eye_age+ as.matrix(pc_dat),family = "binomial")
#   summ <- summary(mod)$coef
#   summ[2,]
# }, mc.cores =(detectCores()-5)) %>% 
#   do.call(rbind,.) %>% 
#   as.data.frame
# rownames(res) <- snps
# colnames(res) <- c( 'estimate' ,'std_error',"stat_val", "p")
# res$model <- "GLM"
# res$PCs <- "YES"
# res$stat <- "t"
# res_all <- rbind(res_all,res)
# rm("res")
# 
# # GLM NO PCS
# res <- mclapply(seq(length(snps)), function(i){
#   mod <- glm(model_data2$pc_lda_route ~ snps_dat[,i]+ model_data2$SEXF + model_data2$BL_eye_age,family = "binomial")
#   summ <- summary(mod)$coef
#   summ[2,]
# }, mc.cores =(detectCores()-5)) %>% 
#   do.call(rbind,.) %>% 
#   as.data.frame
# rownames(res) <- snps
# colnames(res) <- c( 'estimate' ,'std_error',"stat_val", "p")
# res$model <- "GLM"
# res$PCs <- "NO"
# res$stat <- "t"
# res_all <- rbind(res_all,res)
# rm("res")

# plot data
res_all$term <- rep(snps, 1)
rownames(res_all) <- NULL
res_all <- res_all %>% 
  group_by(model, PCs) %>% 
  mutate(q_value = p.adjust(p, "BH"))

ggplot(res_all, aes(estimate, -log10(q_value), color = factor(term)))+
  geom_point()+
  geom_label_repel(aes(label = ifelse(p < .05, substr(term, 1,10), NA))) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = 2) +
  theme_bw()+
  facet_wrap(PCs~model, ncol = 2)+
  ggtitle("Route ~ snp adjusted p value")


res_all %>% 
  filter(p>0) %>% 
  filter(PCs=="YES" & model =="GLMER") %>% 
  ggplot( aes(estimate, -log10(q_value), color = factor(term)))+
  geom_point()+
  geom_label_repel(aes(label = ifelse(p < .05, substr(term, 1,10), NA))) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = 2) +
  theme_bw()+
  facet_wrap(PCs~model, ncol = 2)+
  ggtitle("Route ~ snp adjusted p value")

 ggsave(paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_snp_GLMER_PCS.pdf"),height = 10, width = 15)
 
 
 
res_all %>% 
  left_join(metadata(sce)$genetics$snp_info, by = join_by(term ==label)) %>% 
  filter(p>0) %>% 
  filter(PCs=="YES" & model =="GLMER") %>% 
  ggplot( aes(estimate, -log10(q_value), color = factor(simple_label)))+
  geom_point()+
  labs(color  = "SNP")+
  geom_label_repel(aes(label = ifelse(p < .05, substr(simple_label, 1,10), NA))) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = 2) +
  theme_bw()+
  ggtitle("Trajectory ~ SNPs")+
  xlab("Estimate (<- = a trajectory, -> = b trajectory)")+
  ylab("-log10(q.value)")+
  theme(aspect.ratio = 1,
        plot.background = element_rect(fill = "transparent")
        )
   
ggsave(paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_snp_GLMER_PCS_simple.pdf"),height = 7, width = 7)
sprintf("cp %s %s",
  paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_snp_GLMER_PCS_simple.pdf"),
  paste0(wd_path, "results/poster_plots/pc_lda_route_snp_GLMER_PCS_simple.pdf")
         ) %>% 
  system()

write_csv(res_all, sprintf("%sresults/publication_tables/%s",wd_path, "route_SNP.csv"))

res_all %>% 
  filter(p<.05) %>% 
  filter(PCs=="YES" & model =="GLMER") %>% 
  arrange(p) 



```

## route~covariates only
#### run model
```{r}

# make model data; 1 record per eye; scale predictors
model_data_full <- data.frame(colData(sce)) %>% 
  select( patID_eye, month, patID, pc_lda_route, SEXF, AGE, t2d) %>% 
  mutate(t2d = as.logical(t2d)) %>% # make t2d logical
  group_by(patID_eye) %>% 
  mutate(across(c(SEXF, t2d,AGE), ~ifelse(any(!is.na(.)), .[!is.na(.)], .))) %>% # impute constant variables (technically not correct for age, but negligible only ~ 1-2 years differences)
  arrange(patID_eye, month) %>% 
  slice_head(n = 1) %>% # take first visit observation per eye
  ungroup() %>% 
  mutate(AGEz = (AGE-mean(AGE, na.rm = T))/sd(AGE, na.rm = T)) # stdz age 

nrow(model_data_full)
# 7888
table(model_data_full$patID_eye) %>% table()
# 7888

# collapse A+C and B+D, X to NA so reference is A
model_data <- model_data_full %>%
  mutate(across(matches("route$"), 
                ~ case_when(
                  . %in% c("a", "c") ~ "a", 
                  . %in% c("b", "d") ~ "b", 
                  TRUE ~ NA_character_
                )
  )) %>% 
  mutate(across(matches("route"), ~factor(., levels = c("a", "b")))) %>% # refactor 
  drop_na()

table(model_data$patID_eye) %>% table()
# 1036 w/pcs
# 1327 w/out pcs

head(model_data)
nrow(model_data)
n_distinct(model_data$patID_eye)
n_distinct(model_data$patID)


cat("\n*** GLMER NO PCS ***\n")
summary(
  fit <- glmer(pc_lda_route ~ SEXF + AGEz +t2d+(1|patID),
                     data = model_data, 
                     family = "binomial")
        )

res <- data.frame(coef(summary(fit)))
res$term = rownames(res)
names(res) <- c("Estimate",	"Std. Error",	"z value",	"p",	"term")
write_csv(res, sprintf("%sresults/publication_tables/%s",wd_path, "route~covariates.csv"))

# OUTPUTS----
# *** GLMER NO PCS ***
# > fit <- glmer(model_data$pc_lda_route ~ model_data$SEXF + model_data$BL_eye_age +model_data$t2d+(1|model_data$patID),family = "binomial")
# > summary(fit)
# Generalized linear mixed model fit by maximum likelihood (Laplace Approximation) ['glmerMod']
#  Family: binomial  ( logit )
# Formula: model_data$pc_lda_route ~ model_data$SEXF + model_data$BL_eye_age +      model_data$t2d + (1 | model_data$patID)
# 
#      AIC      BIC   logLik deviance df.resid 
#   1317.9   1342.6   -653.9   1307.9     1031 
# 
# Scaled residuals: 
#     Min      1Q  Median      3Q     Max 
# -2.1093 -0.9233  0.5244  0.7615  2.6775 
# 
# Random effects:
#  Groups           Name        Variance Std.Dev.
#  model_data$patID (Intercept) 0.2755   0.5249  
# Number of obs: 1036, groups:  model_data$patID, 1001
# 
# Fixed effects:
#                       Estimate Std. Error z value Pr(>|z|)    
# (Intercept)            0.10629    0.12485   0.851   0.3946    
# model_data$SEXFTRUE    0.18437    0.14586   1.264   0.2062    
# model_data$BL_eye_age  0.63942    0.08202   7.795 6.42e-15 ***
# model_data$t2d         0.33113    0.14586   2.270   0.0232 *  
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Correlation of Fixed Effects:
#             (Intr) m_$SEX m_$BL_
# m_$SEXFTRUE -0.686              
# mdl_dt$BL__ -0.021  0.125       
# modl_dt$t2d -0.406 -0.036  0.009
# > cat("\n*** GLM NO PCS ***\n")
# 
# *** GLM NO PCS ***
# > summary(glm(model_data$pc_lda_route ~ model_data$SEXF + model_data$BL_eye_age + model_data$t2d,family = "binomial"))
# 
# Call:
# glm(formula = model_data$pc_lda_route ~ model_data$SEXF + model_data$BL_eye_age + 
#     model_data$t2d, family = "binomial")
# 
# Coefficients:
#                       Estimate Std. Error z value Pr(>|z|)    
# (Intercept)            0.10245    0.11749   0.872   0.3832    
# model_data$SEXFTRUE    0.16891    0.13657   1.237   0.2162    
# model_data$BL_eye_age  0.59984    0.06877   8.722   <2e-16 ***
# model_data$t2d         0.31099    0.13595   2.288   0.0222 *  
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# (Dispersion parameter for binomial family taken to be 1)
# 
#     Null deviance: 1405.5  on 1035  degrees of freedom
# Residual deviance: 1309.2  on 1032  degrees of freedom
# AIC: 1317.2
# 
# Number of Fisher Scoring iterations: 4

# PLOT COV MOD----
res <- data.frame(coef(summary(fit)) )
res <- janitor::clean_names(res)
res$term <- str_remove_all(rownames(res),"model_data\\$")

ggplot(res, aes(estimate, -log10(pr_z),  color = term))+
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = 2) +
  geom_point(show.legend = FALSE)+
  geom_label_repel(aes(label = ifelse(pr_z < .05, term, NA)),show.legend = FALSE) +
  theme_bw()+
  xlab("Estimate")+
  ylab("-log10(p.value)")+
  ggtitle("Route ~ SEXF + BL_AGE + T2D + (1|patID)")+
  theme(aspect.ratio = 1,
        plot.background = element_rect(fill = "transparent")
        )
  

   
ggsave(paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_covariates_only_GLMER_noPCS.pdf"),height = 6, width = 6)
sprintf("cp %s %s",
  paste0(wd_path, "results/clustering/13_clustering/pc_lda_route_covariates_only_GLMER_noPCS.pdf"),
  paste0(wd_path, "results/poster_plots/pc_lda_route_covariates_only_GLMER_noPCS.pdf")
         ) %>% 
  system()

write_csv(res, sprintf("%sresults/publication_tables/%s",wd_path, "route_SNP_cov.csv"))


```

#_________________
# Phenotypes 
## ACUIT_PR
###~first_phenos
#### filter data 
```{r}

#___________
# MAKE MODEL DATA
#___________


# what to feed into models
predictors <-  c(rownames(assays(sce)$pmm_imp_of_scaled))
predictors <- predictors %>% 
  grep("RW\\d",., invert = TRUE, value = TRUE)
covariates <- c("SEXF", "(1|patID)", "(1|patID_eye)", "AGE", "BMI", "SEXF")
response <- "ACUIT_rate"

# make model data; scale predictors and covs in colData
acuit_pheno_d <-  colData(sce) %>% 
  as.data.frame() %>% 
  left_join(
    assays(sce)$pmm_imp_of_scaled %>%
      t() %>%
      as.data.frame() %>% 
      rownames_to_column("sample")
  ) %>% 
  dplyr::select( patID_eye, patID, month, my_cluster, chew_grade_derived_mine_loose,ACUIT,pc_lda_route,
                 all_of(c( predictors, covariates[covariates%in%names(colData(sce))])) 
                 )

# scale predictors and covariates
acuit_pheno_d_scaled <- acuit_pheno_d %>%
  mutate(across(all_of(c(
                  predictors, 
                  covariates[covariates%in%names(colData(sce))]
                  )), ~ as.numeric(scale(.))))

# calculate AUIT rate
acuit_pheno_d_scaled <- acuit_pheno_d_scaled %>% 
  mutate(year = month/12) %>% 
  group_by(patID_eye) %>% 
  # mutate(ACUIT_rate = (dplyr::last(ACUIT)-dplyr::first(ACUIT))  /  (dplyr::last(year)-dplyr::first(year)))
  mutate(ACUIT_rate =
          (dplyr::last(ACUIT[!is.na(ACUIT)]) - dplyr::first(ACUIT[!is.na(ACUIT)]))
              /
          (dplyr::last(year[!is.na(ACUIT)]) - dplyr::first(year[!is.na(ACUIT)]))
           ) %>%
  mutate(ACUIT_rate = ifelse(is.infinite(ACUIT_rate), NA, ACUIT_rate))


# seelect first observation; select eyes with first obs in early cluster; filter for a or b route; 
acuit_pheno_d_scaled_filtered <- acuit_pheno_d_scaled %>% 
  arrange(patID_eye,month) %>% 
  group_by(patID_eye) %>% 
  filter(row_number()==1) %>% 
  # filte for starting with minor disease, chew < n
  filter(dplyr::first(chew_grade_derived_mine_loose)<=2) %>% 
  
  # # filter for one route patients
  # ungroup() %>% group_by(patID) %>% filter(n_distinct(pc_lda_route)==1) %>% select(-patID_eye) %>% distinct() %>% 
  
  ungroup() 

nrow(acuit_pheno_d_scaled_filtered)
# 5068

acuit_pheno_d_scaled_filtered$patID_eye %>% table %>% table
#    1 
# 5068

acuit_pheno_d_scaled_filtered %>% 
  {colSums(is.na(.))}

```

#### model
```{r}

resp <- acuit_pheno_d_scaled_filtered[, response, drop = T]
resp
pred <- as.matrix(acuit_pheno_d_scaled_filtered[, predictors])
head(pred)

# cov <- as.matrix(pheno_d_scaled[, c("SEXF", "AGE", "BMI", "SEXF")])
cov <- as.matrix(acuit_pheno_d_scaled_filtered[, c("SEXF", "AGE")])
head(cov)

result_l <- mclapply(colnames(pred), function(nm){
  
  mod_d <- data.frame(
    response = resp, 
    predictor = pred[,nm], 
    cov, 
    patID = acuit_pheno_d_scaled_filtered$patID, 
    chew = acuit_pheno_d_scaled_filtered$chew_grade_derived_mine_loose
  )
  
  # mod
  mod <- try({lmer(response~predictor + SEXF+AGE + (1|patID), data = mod_d)})
  
  if(inherits(mod, "try-error")){
    return(NA)
  }else{
    coefs <- summary(mod)$coef
    rownames(coefs) <- str_replace_all(rownames(coefs), "predictor", nm)
    res <- coefs[rownames(coefs)==nm,, drop = FALSE]
    return(res)
  }
}, mc.cores = detectCores()-5 ) 

# remove failed models
names(result_l)[is.na(result_l)]

result_l <- result_l[!is.na(result_l)]

# make df
result <- do.call(rbind, result_l) %>% 
  as.data.frame
result$p <- result$`Pr(>|t|)`
result$`Pr(>|t|)` <- NULL
result$q <- p.adjust(result$p, "BH")
result$term <- rownames(result)

acuit_pheno_volcano <- result %>% 
  filter(p>0,p<.9) %>%
  ggplot(aes(Estimate, -log10(q), color = term))+
  geom_point(show.legend = FALSE)+
  geom_label_repel(aes(label = ifelse(p < .05, term, NA)), show.legend = FALSE) +
  # geom_label_repel(aes(label = ifelse(q< .05, term, NA)), show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = 2) +
  scale_x_continuous(breaks = seq(-12,2,2))+
  theme_bw()+
  ggtitle("ACUIT_rate ~ phenotype + SEXF + AGE + (1|patID)", "For eyes with BL Chew <= 2")

acuit_pheno_volcano

ggsave(paste0(wd_path, "results/clustering/13_clustering/acuit_pheno_volcano.pdf"),acuit_pheno_volcano,height = 10, width = 12)


result %>% 
  filter(p>0,p<.9) %>%
  ggplot(aes(Estimate, -log10(q), color = term))+
    geom_hline(yintercept = -log10(.05), linetype = 2, color = "grey")+
    geom_vline(xintercept = 0, linetype = 2, color = "grey")+
    geom_vline(xintercept = c(-1, 1), linetype = 3, color = "grey")+
    geom_point(show.legend = FALSE)+
     geom_label_repel(aes(label = ifelse( q < .05,  term, NA)),
                         size = 4,
                         # nudge_y = 0.5,  
                         show.legend = FALSE, 
                         max.overlaps = 20)+
    ggtitle("ACUIT_rate ~ phenotype + SEXF + AGE + (1|patID)", "For eyes with BL Chew <= 2")+
    ylab("-log10(q.value)")+
    theme_bw()+
    theme(aspect.ratio = 1,
          panel.grid = element_blank(),
          panel.background = element_rect(fill = "white", color = NA),
          plot.background = element_rect(fill = "transparent", color = NA),
          legend.background = element_rect(fill = "transparent", color = NA),
          legend.box.background = element_rect(fill = "transparent", color = NA))+
    theme(aspect.ratio = 1,
          axis.title.x = element_text(),
          axis.title.y = element_text(angle = 90))
ggsave(paste0(wd_path, "results/poster_plots/acuit_pheno_volcano.pdf"), height = 5, width = 5)


```


## CHEW_PR
### ~first_phenos
#### filter data 
```{r}

#___________
# MAKE MODEL DATA
#___________



# what to feed into models
predictors <-  c(rownames(assays(sce)$pmm_imp_of_scaled))
covariates <- c("SEXF", "(1|patID)", "(1|patID_eye)", "AGE", "BMI", "SEXF")


# make model data; scale predictors and covs in colData
chew_pheno_d <-  colData(sce) %>% 
  as.data.frame() %>% 
  left_join(
    assays(sce)$pmm_imp_of_scaled %>%
      t() %>%
      as.data.frame() %>% 
      rownames_to_column("sample")
  ) %>% 
  dplyr::select( patID_eye, patID, month, my_cluster, chew_grade_derived_mine_loose,pc_lda_route,
                 all_of(c( predictors, covariates[covariates%in%names(colData(sce))])) 
                 )

# scale predictors and covariates
chew_pheno_d_scaled <- chew_pheno_d %>%
  mutate(across(all_of(c(
                  predictors, 
                  covariates[covariates%in%names(colData(sce))]
                  )), ~ as.numeric(scale(.))))

# calculate AUIT rate
chew_pheno_d_scaled <- chew_pheno_d_scaled %>% 
  mutate(year = month/12) %>% 
  group_by(patID_eye) %>% 
  # mutate(CHEW_rate = (dplyr::last(chew_grade_derived_mine_loose)-dplyr::first(chew_grade_derived_mine_loose))  /  (dplyr::last(year)-dplyr::first(year)))
  mutate(CHEW_rate =
          (dplyr::last(chew_grade_derived_mine_loose[!is.na(chew_grade_derived_mine_loose)]) - dplyr::first(chew_grade_derived_mine_loose[!is.na(chew_grade_derived_mine_loose)]))
              /
          (dplyr::last(year[!is.na(chew_grade_derived_mine_loose)]) - dplyr::first(year[!is.na(chew_grade_derived_mine_loose)]))
           ) %>%
  mutate(CHEW_rate = ifelse(is.infinite(CHEW_rate), NA, CHEW_rate))


# seelect first observation; select eyes with first obs in early cluster; filter for a or b route; 
chew_pheno_d_scaled_filtered <- chew_pheno_d_scaled %>% 
  arrange(patID_eye,month) %>% 
  group_by(patID_eye) %>% 
  filter(row_number()==1) %>% 
  # filte for starting with minor disease, chew < n
  filter(dplyr::first(chew_grade_derived_mine_loose) <= 2) %>% 
  
  # # filter for one route patients
  # ungroup() %>% group_by(patID) %>% filter(n_distinct(pc_lda_route)==1) %>% select(-patID_eye) %>% distinct() %>% 
  
  ungroup() 

nrow(chew_pheno_d_scaled_filtered)
# 4437

chew_pheno_d_scaled_filtered$patID_eye %>% table %>% table
#    1 
# 4437

chew_pheno_d_scaled_filtered %>% 
  {colSums(is.na(.))}

```

#### model
```{r}

response <- "CHEW_rate"
resp <- chew_pheno_d_scaled_filtered[, response, drop = T]
resp
pred <- as.matrix(chew_pheno_d_scaled_filtered[, predictors])
head(pred)

# cov <- as.matrix(pheno_d_scaled[, c("SEXF", "AGE", "BMI", "SEXF")])
cov <- as.matrix(chew_pheno_d_scaled_filtered[, c("SEXF", "AGE")])
head(cov)

result_l <- mclapply(colnames(pred), function(nm){
  
  mod_d <- data.frame(
    response = resp, 
    predictor = pred[,nm], 
    cov, 
    patID = chew_pheno_d_scaled_filtered$patID, 
    chew = chew_pheno_d_scaled_filtered$chew_grade_derived_mine_loose
  )
  
  # mod
  mod <- try({lmer(response~predictor + SEXF+AGE + (1|patID), data = mod_d)})
  
  if(inherits(mod, "try-error")){
    return(NA)
  }else{
    coefs <- summary(mod)$coef
    rownames(coefs) <- str_replace_all(rownames(coefs), "predictor", nm)
    res <- coefs[rownames(coefs)==nm,, drop = FALSE]
    return(res)
  }
}, mc.cores = detectCores()-5 ) 

names(result_l) <- colnames(pred)

# remove failed models
names(result_l)[is.na(result_l)]

result_l <- result_l[!is.na(result_l)]

# make df
result <- do.call(rbind, result_l) %>% 
  as.data.frame
result$p <- result$`Pr(>|t|)`
result$`Pr(>|t|)` <- NULL
result$q <- p.adjust(result$p, "BH")
result$term <- rownames(result)

chew_pheno_volcano <- result %>% 
  filter(p>0,p<.9) %>%
  ggplot(aes(Estimate, -log10(q), color = term))+
  geom_point(show.legend = FALSE)+
  geom_label_repel(aes(label = ifelse(p < .05, term, NA)), show.legend = FALSE) +
  # geom_label_repel(aes(label = ifelse(q< .05, term, NA)), show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = 2) +
  scale_x_continuous(breaks = seq(-12,2,2))+
  theme_bw()+
  ggtitle("Chew_rate ~ phenotype + SEXF + AGE + (1|patID)", "For eyes with BL Chew <= 2")

chew_pheno_volcano

ggsave(paste0(wd_path, "results/clustering/13_clustering/chew_pheno_volcano.pdf"),chew_pheno_volcano,height = 10, width = 12)


result %>% 
  filter(p>0,p<.9) %>%
  ggplot(aes(Estimate, -log10(q), color = term))+
  # ggplot(aes(Estimate, -log10(p), color = term))+
    geom_hline(yintercept = -log10(.05), linetype = 2, color = "grey")+
    geom_vline(xintercept = 0, linetype = 2, color = "grey")+
    geom_vline(xintercept = c(-.1, .1), linetype = 3, color = "grey")+
    geom_point(show.legend = FALSE)+
     geom_label_repel(aes(label = ifelse( q < .05,  term, NA)),
                         size = 4,
                         # nudge_y = 0.5,  
                         show.legend = FALSE, 
                         max.overlaps = 20)+
    ggtitle("Chew_rate ~ phenotype + SEXF + AGE + (1|patID)", "For eyes with BL Chew <= 2")+
    ylab("-log10(q.value)")+
    theme_bw()+
    theme(aspect.ratio = 1,
          panel.grid = element_blank(),
          panel.background = element_rect(fill = "white", color = NA),
          plot.background = element_rect(fill = "transparent", color = NA),
          legend.background = element_rect(fill = "transparent", color = NA),
          legend.box.background = element_rect(fill = "transparent", color = NA))+
    theme(aspect.ratio = 1,
          axis.title.x = element_text(),
          axis.title.y = element_text(angle = 90))
ggsave(paste0(wd_path, "results/poster_plots/chew_pheno_volcano.pdf"), height = 5, width = 5)


sprintf("%sresults/publication_tables/",wd_path) %>% {if(!dir.exists(.)) dir.create(.)}

write_csv(result, sprintf("%sresults/publication_tables/%s",wd_path, "chew_PR_pheno.csv"))

result %>% 
  arrange(q) %>% 
  filter(p<.05) %>% 
  select( Estimate, p,q) %>% 
  mutate(across(where(is.numeric), \(x) round(x,3))) 
#                             Estimate     p     q
# ISOS_LOC_CEN                  -0.128 0.000 0.032
# ISOS_BREAK                    -0.070 0.009 0.326
# RPE_LK05                       0.063 0.017 0.377
# OUTR_EMPTYSPC                 -0.092 0.020 0.377
# AF_INCREASED_1DD               0.053 0.031 0.469
# OUTR_EMPTYSPC_CTR_PT_INVOLV   -0.085 0.045 0.575


```

### ~covariates
#### filter data 
```{r, eval = TRUE}

#___________
# MAKE MODEL DATA
#___________

# make model data; scale predictors and covs in colData
chew_cov_d <-  colData(sce) %>% 
  as.data.frame() %>% 
  dplyr::select( patID_eye, patID, month, my_cluster, chew = chew_grade_derived_mine_loose,pc_lda_route,AGE, BL_eye_age, SEXF, t2d)

# # force no regression
# chew_cov_d <- chew_cov_d %>% 
#   group_by(patID_eye) %>% 
#   mutate(chew = cummax(chew)) %>% 
#   ungroup()
  

# calculate  rate
chew_cov_d <- chew_cov_d %>% 
  mutate(year = month/12) %>% 
  arrange(patID_eye, year) %>% 
  group_by(patID_eye) %>% 
  mutate(CHEW_rate = coef(lm(chew~year))[2]) %>% 
  # mutate(CHEW_rate = (dplyr::last(chew)-dplyr::first(chew))  /  (dplyr::last(year)-dplyr::first(year))) %>%
  # mutate(CHEW_rate =
  #         (dplyr::last(chew[!is.na(chew)]) - dplyr::first(chew[!is.na(chew)]))
  #             /
  #         (dplyr::last(year[!is.na(chew)]) - dplyr::first(year[!is.na(chew)]))
  #          ) %>%
  mutate(CHEW_rate = ifelse(is.infinite(CHEW_rate), NA, CHEW_rate)) %>%
  ungroup()


# seelect first observation; select eyes with first obs in early cluster; filter for a or b route; 
chew_cov_d_filtered <- chew_cov_d %>% 
  arrange(patID_eye,month) %>% 
  group_by(patID_eye) %>% 
  dplyr::slice(1) %>% 
  
  # filte for starting with minor disease, chew < n
  filter(dplyr::first(chew)<=2) %>% 
  
  # # filter for one route patients
  # ungroup() %>% group_by(patID) %>% filter(n_distinct(pc_lda_route)==1) %>% select(-patID_eye) %>% distinct() %>% 
  
  ungroup() 
 
nrow(chew_cov_d_filtered)
# 5068

chew_cov_d_filtered$patID_eye %>% table %>% table
#    1 
# 5068

chew_cov_d_filtered %>% 
  {colSums(is.na(.))}

chew_cov_d_filtered %>% 
  filter(if_all(c(CHEW_rate, t2d, SEXF), ~{!is.na(.)}))

```

#### model
```{r, eval = TRUE}


chew_cov_mod <- lmer(CHEW_rate ~ AGE + SEXF+ (1|patID), data = chew_cov_d_filtered)
(summ <- summary(chew_cov_mod))
effects::allEffects(chew_cov_mod) %>% plot()
result <- data.frame(coef(summ))
result <- janitor::clean_names(result)
result$term <- rownames(result)

result %>% 
  filter(pr_t>0,pr_t<.9) %>%
  ggplot(aes(estimate, -log10(pr_t), color = term))+
  # ggplot(aes(Estimate, -log10(p), color = term))+
    geom_hline(yintercept = -log10(.05), linetype = 2, color = "grey")+
    geom_vline(xintercept = 0, linetype = 2, color = "grey")+
    geom_vline(xintercept = c(-.1, .1), linetype = 3, color = "grey")+
    geom_point(show.legend = FALSE)+
     geom_label_repel(aes(label = ifelse( pr_t < .05,  term, NA)),
                         size = 4,
                         # nudge_y = 0.5,  
                         show.legend = FALSE, 
                         max.overlaps = 20)+
    ggtitle("Chew_rate ~ SEXF + AGE + (1|patID)", "For eyes with BL Chew <= 2")+
    ylab("-log10(p.value)")+
    theme_bw()+
    theme(aspect.ratio = 1,
          panel.grid = element_blank(),
          panel.background = element_rect(fill = "white", color = NA),
          plot.background = element_rect(fill = "transparent", color = NA),
          legend.background = element_rect(fill = "transparent", color = NA),
          legend.box.background = element_rect(fill = "transparent", color = NA))+
    theme(aspect.ratio = 1,
          axis.title.x = element_text(),
          axis.title.y = element_text(angle = 90))
ggsave(paste0(wd_path, "results/poster_plots/chew_cov_volcano.pdf"), height = 5, width = 5)



write_csv(result, sprintf("%sresults/publication_tables/%s",wd_path, "chew_PR_pheno_covs.csv"))

```

## PP_PR
###~first_phenos
#### filter data 
```{r}

#___________
# MAKE MODEL DATA
#___________

# what to feed into models
predictors <-  c(rownames(assays(sce)$pmm_imp_of_scaled))
covariates <- c("SEXF", "(1|patID)", "(1|patID_eye)", "AGE", "BMI", "SEXF")


# make model data; scale predictors and covs in colData
pp_pheno_d <-  colData(sce) %>% 
  as.data.frame() %>% 
  left_join(
    assays(sce)$pmm_imp_of_scaled %>%
      t() %>%
      as.data.frame() %>% 
      rownames_to_column("sample")
  ) %>% 
  dplyr::select( patID_eye, patID, month, my_cluster,pseudoprog, chew_grade_derived_mine_loose,pc_lda_route,
                 all_of(c( predictors, covariates[covariates%in%names(colData(sce))])) 
                 )

# scale predictors and covariates
pp_pheno_d_scaled <- pp_pheno_d %>%
  mutate(across(all_of( c(predictors, covariates[covariates%in%names(colData(sce))])), 
                ~ as.numeric(scale(.))
                ))

# calculate AUIT rate
pp_pheno_d_scaled <- pp_pheno_d_scaled %>% 
  mutate(year = month/12) %>% 
  group_by(patID_eye) %>% 
  # mutate(pp_rate = (dplyr::last(pseudoprog)-dplyr::first(pseudoprog))  /  (dplyr::last(year)-dplyr::first(year)))
  mutate(pp_rate =
          (dplyr::last(pseudoprog[!is.na(pseudoprog)]) - dplyr::first(pseudoprog[!is.na(pseudoprog)]))
              /
          (dplyr::last(year[!is.na(pseudoprog)]) - dplyr::first(year[!is.na(pseudoprog)]))
           ) %>%
  mutate(pp_rate = ifelse(is.infinite(pp_rate), NA, pp_rate))


# seelect first observation; select eyes with first obs in early cluster; filter for a or b route; 
pp_pheno_d_scaled_filtered <- pp_pheno_d_scaled %>% 
  arrange(patID_eye,month) %>% 
  group_by(patID_eye) %>% 
  dplyr::slice(1) %>% 
  
  # filter for starting with minor disease, pp < n
  filter(chew_grade_derived_mine_loose<=2) %>%
  
  # # filter for one route patients
  # ungroup() %>% group_by(patID) %>% filter(n_distinct(pc_lda_route)==1) %>% select(-patID_eye) %>% distinct() %>% 
  
  ungroup() 

nrow(pp_pheno_d_scaled_filtered)
# 4437

pp_pheno_d_scaled_filtered$patID_eye %>% table %>% table
#    1 
# 4437

pp_pheno_d_scaled_filtered %>% 
  {colSums(!is.na(.))}

```

#### model
```{r}

response <- "pp_rate"
resp <- pp_pheno_d_scaled_filtered[, response, drop = T]
resp
pred <- as.matrix(pp_pheno_d_scaled_filtered[, predictors])
head(pred)

# cov <- as.matrix(pheno_d_scaled[, c("SEXF", "AGE", "BMI", "SEXF")])
cov <- as.matrix(pp_pheno_d_scaled_filtered[, c("SEXF", "AGE")])
head(cov)

# nm <- colnames(pred)[1]
result_l <- mclapply(colnames(pred), function(nm){
  
  mod_d <- data.frame(
    response = resp, 
    predictor = pred[,nm], 
    cov, 
    patID = pp_pheno_d_scaled_filtered$patID
  )
  
  # mod
  mod <- try({lmer(response~predictor + SEXF+AGE + (1|patID), data = mod_d)})
  
  if(inherits(mod, "try-error")){
    return(NA)
  }else{
    coefs <- summary(mod)$coef
    rownames(coefs) <- str_replace_all(rownames(coefs), "predictor", nm)
    res <- coefs[rownames(coefs)==nm,, drop = FALSE]
    return(res)
  }
}, mc.cores = detectCores()-5 ) 

names(result_l) <- colnames(pred)

# remove failed models
names(result_l)[is.na(result_l)]

result_l <- result_l[!is.na(result_l)]

# make df
result <- do.call(rbind, result_l) %>% 
  as.data.frame
result$p <- result$`Pr(>|t|)`
result$`Pr(>|t|)` <- NULL
result$q <- p.adjust(result$p, "BH")
result$term <- rownames(result)

pp_pheno_volcano <- result %>% 
  filter(p>0,p<.9) %>%
  ggplot(aes(Estimate, -log10(q), color = term))+
  geom_point(show.legend = FALSE)+
  geom_label_repel(aes(label = ifelse(p < .05, term, NA)), show.legend = FALSE) +
  # geom_label_repel(aes(label = ifelse(q< .05, term, NA)), show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = 2) +
  scale_x_continuous(breaks = seq(-12,2,2))+
  theme_bw()+
  ggtitle("Pseudoprog Rate ~ phenotype + SEXF + AGE + (1|patID)", "For eyes with BL Chew <= 2")

pp_pheno_volcano

ggsave(paste0(wd_path, "results/clustering/13_clustering/pp_pheno_volcano.pdf"),pp_pheno_volcano,height = 10, width = 12)


result %>% 
  filter(p>0,p<.9) %>%
  ggplot(aes(Estimate, -log10(q), color = term))+
  # ggplot(aes(Estimate, -log10(p), color = term))+
    geom_hline(yintercept = -log10(.05), linetype = 2, color = "grey")+
    geom_vline(xintercept = 0, linetype = 2, color = "grey")+
    geom_vline(xintercept = c(-.1, .1), linetype = 3, color = "grey")+
    geom_point(show.legend = FALSE)+
     geom_label_repel(aes(label = ifelse( q < .05,  term, NA)),
                         size = 4,
                         # nudge_y = 0.5,  
                         show.legend = FALSE, 
                         max.overlaps = 20)+
    ggtitle("Pseudoprog Rate ~ phenotype + SEXF + AGE + (1|patID)", "For eyes with BL Chew <= 2")+
    ylab("-log10(q.value)")+
    theme_bw()+
    theme(aspect.ratio = 1,
          panel.grid = element_blank(),
          panel.background = element_rect(fill = "white", color = NA),
          plot.background = element_rect(fill = "transparent", color = NA),
          legend.background = element_rect(fill = "transparent", color = NA),
          legend.box.background = element_rect(fill = "transparent", color = NA))+
    theme(aspect.ratio = 1,
          axis.title.x = element_text(),
          axis.title.y = element_text(angle = 90))
ggsave(paste0(wd_path, "results/poster_plots/pp_pheno_volcano.pdf"), height = 5, width = 5)

write_csv(result, sprintf("%sresults/publication_tables/%s",wd_path, "PP_PR_pheno.csv"))

result %>% 
  arrange(q) %>% 
  filter(p<.05) %>% 
  select( Estimate, p,q) %>% 
  mutate(across(where(is.numeric), \(x) round(x,3))) 
#                             Estimate     p     q
# RW074                          0.367 0.000 0.006
# RW084                          0.257 0.001 0.056
# OUTR_EMPTYSPC                 -0.250 0.005 0.075
# OUTR_EMPTYSPC_CTR_PT_INVOLV   -0.264 0.005 0.075
# RW064                          0.337 0.006 0.075
# ISOS_LOC_CEN                  -0.231 0.005 0.075
# ISOS_BREAK                    -0.161 0.007 0.077
# RW094                          0.181 0.009 0.089
# ICL05                         -0.147 0.012 0.090
# RPE_LK02                       0.228 0.011 0.090
# RW054                          0.339 0.021 0.143
# RPE_LK05                       0.123 0.040 0.231
# HRDEXUDATE                    -0.409 0.038 0.231

```

 

###~covariates
#### filter data 
```{r, eval = FALSE}

#___________
# MAKE MODEL DATA
#___________

# make model data; scale predictors and covs in colData
pp_cov_d <-  colData(sce) %>% 
  as.data.frame() %>% 
  dplyr::select( patID_eye, patID, month, my_cluster, chew = chew_grade_derived_mine_loose, pseudoprog,pc_lda_route,AGE, BL_eye_age, SEXF, t2d)

# calculate  rate
pp_cov_d <- pp_cov_d %>% 
  mutate(year = month/12) %>% 
  arrange(patID_eye, year) %>% 
  group_by(patID_eye) %>% 
  # mutate(pp_rate = coef(lm(pseudoprog~year))[2]) %>% 
  # mutate(pp_rate = (dplyr::last(pp)-dplyr::first(pp))  /  (dplyr::last(year)-dplyr::first(year))) %>%
  mutate(pp_rate =
          (dplyr::last(pseudoprog[!is.na(pseudoprog)]) - dplyr::first(pseudoprog[!is.na(pseudoprog)]))
              /
          (dplyr::last(year[!is.na(pseudoprog)]) - dplyr::first(year[!is.na(pseudoprog)]))
           ) %>%
  mutate(pp_rate = ifelse(is.infinite(pp_rate), NA, pp_rate)) %>%
  ungroup()


# seelect first observation; select eyes with first obs in early cluster; filter for a or b route; 
pp_cov_d_filtered <- pp_cov_d %>% 
  arrange(patID_eye,month) %>% 
  group_by(patID_eye) %>% 
  dplyr::slice(1) %>% 
  
  # filte for starting with minor disease, pp < n
  filter(dplyr::first(chew)<=2) %>%
  
  # # filter for one route patients
  # ungroup() %>% group_by(patID) %>% filter(n_distinct(pc_lda_route)==1) %>% select(-patID_eye) %>% distinct() %>% 
  
  ungroup() 
 
nrow(pp_cov_d_filtered)
# 5068

pp_cov_d_filtered$patID_eye %>% table %>% table
#    1 
# 5068

pp_cov_d_filtered %>% 
  {colSums(is.na(.))}

pp_cov_d_filtered %>% 
  # filter(if_all(c(pp_rate, SEXF), ~{!is.na(.)}))
  filter(if_all(c(pp_rate, BL_eye_age,SEXF), ~{!is.na(.)}))

```

#### model
```{r, eval = FALSE}

pp_cov_mod <- lmer(pp_rate ~ AGE + SEXF+ (1|patID), data = pp_cov_d_filtered)
(summ <- summary(pp_cov_mod))
effects::allEffects(pp_cov_mod) %>% plot()
result <- data.frame(coef(summ))
result <- janitor::clean_names(result)
result$term <- rownames(result)

result %>% 
  filter(pr_t>0,pr_t<.9) %>%
  ggplot(aes(estimate, -log10(pr_t), color = term))+
  # ggplot(aes(Estimate, -log10(p), color = term))+
    geom_hline(yintercept = -log10(.05), linetype = 2, color = "grey")+
    geom_vline(xintercept = 0, linetype = 2, color = "grey")+
    geom_vline(xintercept = c(-.1, .1), linetype = 3, color = "grey")+
    geom_point(show.legend = FALSE)+
     geom_label_repel(aes(label = ifelse( pr_t < .05,  term, NA)),
                         size = 4,
                         # nudge_y = 0.5,  
                         show.legend = FALSE, 
                         max.overlaps = 20)+
    ggtitle(paste0("Pseudoprog Grade Rate ~ AGE + SEXF + (1|patID)"), "For eyes with BL Chew <= 2")+
    ylab("-log10(p.value)")+
    theme_bw()+
    theme(aspect.ratio = 1,
          panel.grid = element_blank(),
          panel.background = element_rect(fill = "white", color = NA),
          plot.background = element_rect(fill = "transparent", color = NA),
          legend.background = element_rect(fill = "transparent", color = NA),
          legend.box.background = element_rect(fill = "transparent", color = NA))+
    theme(aspect.ratio = 1,
          axis.title.x = element_text(),
          axis.title.y = element_text(angle = 90))
ggsave(paste0(wd_path, "results/poster_plots/pp_cov_volcano.pdf"), height = 5, width = 5)


```




 ## model phenptypes by snps
### filter data for terms that converge
```{r}

#___________
# MAKE MODEL DATA
#___________

# what to feed into models
# predictors <-  snps
predictors <-  snps

# factor_names <- grep("MR\\d", colnames(sce), value  = TRUE)
# responses <- c(rownames(assays(sce)$pmm_imp_of_scaled), factor_names)
responses <- c(rownames(assays(sce)$pmm_imp_of_scaled))
covariates <- c("SEXF", "(1|patID)", "(1|patID_eye)", "AGE", "BMI", paste0("PC",1:10), "SEXF")

# make model data; scale predictors and covs in colData
snp_pheno_data <- colData(sce) %>% 
  as.data.frame() %>% 
  left_join(
    assays(sce)$pmm_imp_of_scaled %>%
      t() %>%
      as.data.frame() %>% 
      rownames_to_column("sample")
  ) %>% 
  dplyr::select( patID_eye, patID, month, all_of(c(responses, predictors, covariates[covariates%in%names(colData(sce))])) )

snp_pheno_data_scaled <- snp_pheno_data %>%
  mutate(across(all_of(c(predictors, covariates[covariates%in%names(colData(sce))]  )),
                ~ as.numeric(scale(.))))


nrow(snp_pheno_data_scaled[!is.na(snp_pheno_data_scaled$rs1047891_cps1),])
# ] 8863 obs
n_distinct(snp_pheno_data_scaled$patID_eye[!is.na(snp_pheno_data_scaled$rs1047891_cps1)])
# ] 5611
n_distinct(snp_pheno_data_scaled$patID[!is.na(snp_pheno_data_scaled$rs1047891_cps1)])
# ] 2822

```


### Parallelized model correcting for covariates
```{r}


###    ####    ####    ####    ####
# RUN MODEL ----
###    ####    ####    ####    ####

  
# Define lists of covariates
covariates_list <- list(
  # c("SEXF", "(1|patID)" ),
  # c("SEXF", "(1|patID)" , "(1|patID_eye)")
  # c("SEXF", "(1|patID)" , "(1|patID_eye)"),
  # c("SEXF", "(1|patID)" , "(1|patID_eye)", "AGE", "BMI"),
  # c("SEXF", "(1|patID)" , "(1|patID_eye)", "AGE"),
  c("SEXF", "(1|patID)" , "(1|patID_eye)", "AGE", paste0("PC", 1:10))
)

# Create all combinations of covariate sets, phenotypes, and SNPs
formula_d <- expand.grid(
  response = responses, 
  predictor = predictors, 
  covariates = sapply(covariates_list, function(x)paste(x, collapse = " + ")), 
  stringsAsFactors = FALSE)

forms <- apply(formula_d, 1, function(x) sprintf("%s ~ %s + %s", x["response"], x["predictor"], x["covariates"])) %>% 
  as.character()

how.long(T); 
  fit_list <- mclapply(forms, function(form){
    # is this a mixed effects model
    is_me <- grepl("patID", form)
    form <- as.formula(form)
    
    # if not mixed effeects
    if(is_me) {
      # model with patID random intercept
      mod <- try({lmer(form, snp_pheno_data_scaled) })    
    }else{
      # model with fixed effects
      mod <- try({lm(form, snp_pheno_data_scaled) })
    }
    # add formula to mod
    attr(mod, "form") <- form
    if(inherits( mod, "try-error")){
      return(NA)
    }else{
      return(mod)
    }
  }, mc.cores = (detectCores() - 1))
how.long() #>> 4mins
  


####    ####    ####    ####    ####    
# ORGANIZE RESULTS LISTS ----
###    ####    ####    ####    ####
# give respective names
names(fit_list) <- forms

# which models didn't work
names(fit_list[is.na(fit_list)])

# remove models that didn't work
fit_list <- fit_list[!is.na(fit_list)]

pheno_snp_summs <- mclapply(fit_list, summary, mc.cores = detectCores()-5)

pheno_snp_results <- extract_results(pheno_snp_summs) 

pheno_snp_results <- pheno_snp_results %>% 
  filter(!grepl("Intercept|SEX|AGE|BL|BMI|PC", Term)) %>% 
  arrange(`Pr(>|t|)`) %>% 
  mutate(q_value = p.adjust(`Pr(>|t|)`, "BH"))


rm("fit_list")
rm("pheno_snp_summs")

```


#### plot Snp x phenotype filled with betas
```{r}

# results for SEXF + patID + eye + age models
# results <- snp_pheno_mods_results$`SEXF+(1|patID)+(1|patID_eye)+AGE`
# results <- snp_pheno_mods_results$`SEXF+(1|patID)+(1|patID_eye)+AGE+PC1+PC2+PC3+PC4+PC5+PC6+PC7+PC8+PC9+PC10`
results <- pheno_snp_results %>% 
  left_join(metadata(sce)$genetics$snp_info %>% select(label, simple_label), by = join_by(Term == label)) %>% 
  mutate(simple_label = toupper(simple_label)) %>% 
  mutate(simple_label = 
           str_replace(simple_label, "_RARE|_COMMON", 
                       tolower(str_extract(simple_label, "_RARE|_COMMON"))
                       )
         )

results$q_value <- p.adjust(results$`Pr(>|t|)`, method = "BH")

# add phenotypes info to results
var_names <- as.data.frame(rowData(sce) ) %>% 
  select(name, description)

results_to_save <- results %>% 
  left_join(var_names, by = join_by(Response == name)) %>% 
  mutate(description = ifelse(is.na(description), "", description))

# Save results
write.csv(results_to_save %>% select(-Predictor), paste0(wd_path, "results/clustering/13_clustering/patID_patID_eye_AGE_snp_pheno_results.csv"), row.names = FALSE)

# # Filter out factors
# results <- results %>% 
#   filter(!Response %in% factor_names)
# fill matrix with snp x pheno filed with betas
## make empty matrix
snp_pheno_beta_m <-  matrix(rep(NA, n_distinct(results$Response))) %*% rep(NA,n_distinct(results$simple_label))
dimnames(snp_pheno_beta_m) <- list(sort(unique(results$Response)), sort(unique(results$simple_label)))

# for each row column, fill cell with estimate where df var = row/col name of matrix at i,j
for(i in 1:nrow(snp_pheno_beta_m)){for(j in 1:ncol(snp_pheno_beta_m)){
  snp_pheno_beta_m[i,j] <- results$Estimate[results$Response == rownames(snp_pheno_beta_m)[i] & results$simple_label == colnames(snp_pheno_beta_m)[j]]
  }}

colLabs <- results %>% 
  select(Term, simple_label) %>% 
  mutate(new_simple_label = ifelse(simple_label=="TMEM161B", "MIR9-2", simple_label)) %>% 
  distinct() %>% 
  mutate(lab = paste(sapply(str_split(Term, "_"), `[`,1), new_simple_label, sep = "-")) %>% 
  select(-Term)

snp_pheno_beta_m_labeled <- snp_pheno_beta_m
colnames(snp_pheno_beta_m_labeled) <- colLabs$lab[match(colnames(snp_pheno_beta_m_labeled), colLabs$simple_label)]

set.seed(123919)
hm <- Heatmap(snp_pheno_beta_m_labeled, 
        clustering_method_columns = "ward.D2",
        clustering_method_rows = "ward.D2", 
        name = "Estimate",  # This is the name used by decorate_Heatmap_body
        column_title = "Phenotype ~ SNP", 
        # height = unit(50*1.5, "cm"), 
        # width = unit(20*1.5, "cm")
        )
# draw(hm,padding = unit(c(4, 2, 2, 2), "cm") )

# for each row column, fill cell with estimate where df var = row/col name of matrix at i,j
snp_pheno_q_m <- snp_pheno_beta_m*NA
for(i in 1:nrow(snp_pheno_q_m)){for(j in 1:ncol(snp_pheno_q_m)){
    snp_pheno_q_m[i,j] <- results$q_value[results$Response == rownames(snp_pheno_q_m)[i] & results$simple_label == colnames(snp_pheno_q_m)[j]]
  }}

snp_pheno_p_m <- snp_pheno_beta_m*NA
for(i in 1:nrow(snp_pheno_p_m)){for(j in 1:ncol(snp_pheno_p_m)){
    snp_pheno_p_m[i,j] <- results$`Pr(>|t|)`[results$Response == rownames(snp_pheno_p_m)[i] & results$simple_label == colnames(snp_pheno_p_m)[j]]
  }}


# # matrix of asterisks of significance
# sig_asterisks <- function(vec, none = NA_character_){
#   asterisks <- case_when(
#     vec < .001 ~ "  ***", 
#     vec < .01 ~"**",
#     vec < .05 ~ "*", 
#     vec < .1 ~ ".", 
#     TRUE ~ none)
#   return(asterisks)
# }

## convert p-values to asterisks; add rownames back
# sig_m <- apply(snp_pheno_q_m,2, sig_asterisks)
```


##### use p and q values to fill with sig asterisks
```{r}

pq_m <- snp_pheno_q_m*NA


for(i in 1:nrow(pq_m)){
  for(j in 1:ncol(pq_m)){
    q <- snp_pheno_q_m[i,j]
    p <- snp_pheno_p_m[i,j]
    pq_m[i,j] <- ifelse(
      q<.01, "  ***", 
      ifelse(
        q<.05, "**", 
        ifelse(
          p<.05, "*", 
          ""
        )      )    )
  }
}
  
rownames(pq_m) <- rownames(snp_pheno_q_m)

all(rownames(pq_m )==rownames(snp_pheno_q_m))
all(colnames(pq_m )==colnames(snp_pheno_q_m))

# reorder sig m to be that of hm
row_order_names <- rownames(pq_m)[row_order(hm)]
col_order_names <- colnames(pq_m)[column_order(hm)]
pq_m <- pq_m[row_order_names,col_order_names]

# plot
pdf(paste0(wd_path, "results/clustering/13_clustering/snp_pheno_beta_hm_pq.pdf"), height = 20, width = 12, bg = "transparent")
par(mar = c(10, 3,3,3))


draw(hm,padding = unit(c(4, 2, 2, 2), "cm"), background = "transparent" )
# Add asterisks for significance
decorate_heatmap_body("Estimate", {  # Referencing the correct Heatmap name
    # Loop over the matrix to overlay asterisks based on significance
    for(i in 1:nrow(snp_pheno_beta_m)) {
        for(j in 1:ncol(snp_pheno_beta_m)) {
            
          significance = pq_m[i,j]
          # Overlay the significance level at the center of the cell
          if (!is.na(significance)) {
            grid.text(significance,
              x = j/ncol(snp_pheno_beta_m) - .05,
              y = (nrow(snp_pheno_beta_m) - i + 1)/nrow(snp_pheno_beta_m) - .005,
              # x = j/ncol(snp_pheno_beta_m),
              # y = (nrow(snp_pheno_beta_m) - i + 1)/nrow(snp_pheno_beta_m) ,
              gp = gpar(fontsize = 14, col = "black"))
            }
          }
        }
    })

par(mar = rep(2,4))
dev.off()


```

##### use p to fill with sig asterisks
```{r}

p_m <- pq_m*NA

# *** = q < 0.05 
# ** = q < 0.1
# * = p < 0.05
# • rm

for(i in 1:nrow(p_m)){
  for(j in 1:ncol(p_m)){
    # q <- snp_pheno_q_m[i,j]
    p <- snp_pheno_p_m[i,j]
    p_m[i,j] <- ifelse(
      p<.001, "  ***", 
      ifelse(
        p<.01, "**", 
        ifelse(
          p<.05, "*", 
          ""
        )      )    )
  }
}
  
rownames(p_m) <- rownames(snp_pheno_q_m)

all(rownames(p_m )==rownames(snp_pheno_q_m))
all(colnames(p_m )==colnames(snp_pheno_q_m))

# reorder sig m to be that of hm
row_order_names <- rownames(p_m)[row_order(hm)]
col_order_names <- colnames(p_m)[column_order(hm)]
p_m <- p_m[row_order_names,col_order_names]

# plot
pdf(paste0(wd_path, "results/clustering/13_clustering/snp_pheno_beta_hm_p.pdf"), height = 20, width = 12, bg = "transparent")
par(mar = c(10, 3,3,3))


draw(hm,padding = unit(c(4, 2, 2, 2), "cm"), background = "transparent" )
# Add asterisks for significance
decorate_heatmap_body("Estimate", {  # Referencing the correct Heatmap name
    # Loop over the matrix to overlay asterisks based on significance
    for(i in 1:nrow(snp_pheno_beta_m)) {
        for(j in 1:ncol(snp_pheno_beta_m)) {
            
          significance = p_m[i,j]
          # Overlay the significance level at the center of the cell
          if (!is.na(significance)) {
            grid.text(significance,
              x = j/ncol(snp_pheno_beta_m) - .05,
              y = (nrow(snp_pheno_beta_m) - i + 1)/nrow(snp_pheno_beta_m) - .005,
              # x = j/ncol(snp_pheno_beta_m),
              # y = (nrow(snp_pheno_beta_m) - i + 1)/nrow(snp_pheno_beta_m) ,
              gp = gpar(fontsize = 14, col = "black"))
            }
          }
        }
    })

par(mar = rep(2,4))
dev.off()

```

##### use q to fill with sig asterisks
```{r}

q_m <- pq_m*NA

# *** = q < 0.05 
# ** = q < 0.1
# * = p < 0.05
# • rm

for(i in 1:nrow(q_m)){
  for(j in 1:ncol(q_m)){
    q <- snp_pheno_q_m[i,j]
    # p <- snp_pheno_p_m[i,j]
    q_m[i,j] <- ifelse(
      q<.001, "  ***", 
      ifelse(
        q<.01, "**", 
        ifelse(
          q<.05, "*", 
          ""
        )      )    )
  }
}
  
rownames(q_m) <- rownames(snp_pheno_q_m)

all(rownames(q_m )==rownames(snp_pheno_q_m))
all(colnames(q_m )==colnames(snp_pheno_q_m))

# reorder sig m to be that of hm
row_order_names <- rownames(q_m)[row_order(hm)]
col_order_names <- colnames(q_m)[column_order(hm)]
q_m <- q_m[row_order_names,col_order_names]

# plot
pdf(paste0(wd_path, "results/clustering/13_clustering/snp_pheno_beta_hm_q.pdf"), height = 20, width = 12, bg = "transparent")
par(mar = c(10, 3,3,3))


draw(hm,padding = unit(c(4, 2, 2, 2), "cm"), background = "transparent" )
# Add asterisks for significance
decorate_heatmap_body("Estimate", {  # Referencing the correct Heatmap name
    # Loop over the matrix to overlay asterisks based on significance
    for(i in 1:nrow(snp_pheno_beta_m)) {
        for(j in 1:ncol(snp_pheno_beta_m)) {
            
          significance = q_m[i,j]
          # Overlay the significance level at the center of the cell
          if (!is.na(significance)) {
            grid.text(significance,
              x = j/ncol(snp_pheno_beta_m) - .05,
              y = (nrow(snp_pheno_beta_m) - i + 1)/nrow(snp_pheno_beta_m) - .005,
              # x = j/ncol(snp_pheno_beta_m),
              # y = (nrow(snp_pheno_beta_m) - i + 1)/nrow(snp_pheno_beta_m) ,
              gp = gpar(fontsize = 14, col = "black"))
            }
          }
        }
    })

par(mar = rep(2,4))
dev.off()

```

##### move beta hms
```{r}

pths <- c(
  '~/Documents/LMRI/PRJ2024001/version015/results/clustering/13_clustering/snp_pheno_beta_hm_p.pdf',
'~/Documents/LMRI/PRJ2024001/version015/results/clustering/13_clustering/snp_pheno_beta_hm_pq.pdf', 
'~/Documents/LMRI/PRJ2024001/version015/results/clustering/13_clustering/snp_pheno_beta_hm_q.pdf')

for(p in pths){
  sprintf("cp %s %s" , 
            p, 
            sprintf("%sresults/publication_figures/%s", wd_path, basename(p))) %>% 
    system()
}

```

### modeling covariates
#### make modeling data 
```{r}
#___________
# MAKE MODEL DATA
#___________

# what to feed into models
responses <- c(rownames(assays(sce)$pmm_imp_of_scaled))
covariates <- c("SEXF",'t2d', "AGE", "BMI", paste0("PC",1:10), "(1|patID)", "(1|patID_eye)")

# make model data; scale predictors and covs in colData
cov_pheno_data <- colData(sce) %>% 
  as.data.frame() %>% 
  left_join(
    assays(sce)$pmm_imp_of_scaled %>%
      t() %>%
      as.data.frame() %>% 
      rownames_to_column("sample")
  ) %>% 
  dplyr::select( patID_eye, patID, month, all_of(c(responses, covariates[covariates%in%names(colData(sce))])) )

cov_pheno_data_scaled <- cov_pheno_data %>%
  mutate(across(matches("BMI|AGE|PC"),
                ~ as.numeric(scale(.)))
         )


```

#### run model
```{r}


###    ####    ####    ####    ####
# RUN MODEL ----
###    ####    ####    ####    ####

  
# Create all combinations of covariate sets, phenotypes, and SNPs
formula_d <- expand.grid(
  response = responses, 
  # covariates = sapply(covariates, function(x)paste(x, collapse = " + ")), 
  covariates = paste(covariates, collapse = " + "), 
  stringsAsFactors = FALSE)

forms <- apply(formula_d, 1, function(x) sprintf("%s ~  %s", x["response"], x["covariates"])) %>% 
  as.character()

forms[1]
# "IR01 ~  SEXF + (1|patID) + (1|patID_eye) + AGE + BMI + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10 + t2d"
 
how.long(T); 
  fit_list <- mclapply(forms, function(form){
    # is this a mixed effects model
    is_me <- grepl("patID", form)
    form <- as.formula(form)
    
    # if not mixed effeects
    if(is_me) {
      # model with patID random intercept
      mod <- try({lmer(form, cov_pheno_data_scaled) })    
    }else{
      # model with fixed effects
      mod <- try({lm(form, cov_pheno_data_scaled) })
    }
    # add formula to mod
    attr(mod, "form") <- form
    if(inherits( mod, "try-error")){
      return(NA)
    }else{
      return(mod)
    }
  }, mc.cores = (detectCores() - 8))
how.long() #>> 8s
  


####    ####    ####    ####    ####    
# ORGANIZE RESULTS LISTS ----
###    ####    ####    ####    ####
# give respective names
names(fit_list) <- forms

# which models didn't work
names(fit_list[is.na(fit_list)])

# remove models that didn't work
fit_list <- fit_list[!is.na(fit_list)]

pheno_cov_summs <- mclapply(fit_list, summary, mc.cores = detectCores()-5)

pheno_cov_results <- extract_results(pheno_cov_summs) 

pheno_cov_results <- pheno_cov_results %>% 
  arrange(`Pr(>|t|)`) %>% 
  mutate(q_value = p.adjust(`Pr(>|t|)`, "BH"))

rm("fit_list")
rm("pheno_cov_summs")

```


#### plot Snp x phenotype filled with betas
```{r}

pheno_cov_results <- janitor::clean_names(pheno_cov_results)
pheno_cov_results %>% 
  filter(pr_t>0,pr_t<.9) %>%
  ggplot(aes(estimate, -log10(q_value), color = term))+
  # ggplot(aes(Estimate, -log10(p), color = term))+
    geom_hline(yintercept = -log10(.05), linetype = 2, color = "grey")+
    geom_vline(xintercept = 0, linetype = 2, color = "grey")+
    geom_vline(xintercept = c(-.1, .1), linetype = 3, color = "grey")+
    geom_point(show.legend = FALSE)+
    geom_label_repel(aes(label = ifelse( q_value < .05,  term , NA)),
                         size = 3,
                         show.legend = FALSE, 
                         max.overlaps = 20)+
    ggtitle(paste0("e.g. ", forms[1]))+
    ylab("-log10(q.value)")+
    facet_wrap(~response, scales = "free_y", ncol = 15)+
    theme_bw()+
    theme(aspect.ratio = 1,
          panel.grid = element_blank(),
          panel.background = element_rect(fill = "white", color = NA),
          plot.background = element_rect(fill = "transparent", color = NA),
          legend.background = element_rect(fill = "transparent", color = NA),
          legend.box.background = element_rect(fill = "transparent", color = NA))+
    theme(aspect.ratio = 1,
          axis.title.x = element_text(),
          axis.title.y = element_text(angle = 90))
ggsave(paste0(wd_path, "results/poster_plots/pheno_cov_volcano.pdf"), height = 5, width = 5)


```

<!-- # sample difference from median across samples for each feature -->
<!-- ```{r} -->

<!-- var <- "pseudoprog" -->
<!-- plotRLE(sce[, order(colData(sce)[[var]])], "pmm_imp_of_scaled", color_by = var) -->

<!-- ``` -->


#-----#-----#-----#-----#-----
# SAVE DATA
```{r}

saveHDF5SummarizedExperiment(sce, paste0(wd_path, "processed_data/sce006.h5"), replace = TRUE)

notify.me("13_routeModeling Completed")
how.long()
```



```{r}

# untangling center point involvement vars

# raw <- metadata(sce)$raw
# 
# raw %>% 
#   lapply(function(df){ grep("CTR", names(df), value = TRUE)})
# 
# # fafgrade, octmain have CTR_PNT INVOLVED
# faf <- raw$FAFGRADE
# faf <- convert_to_numeric_if_possible(faf)
# faf_cor <- faf %>% 
#   dplyr::select( where(~is.numeric(.))) %>% 
#   as.matrix() %>% 
#   # scale() %>% 
#   cor(., use = "complete.obs")
# #^  AF_INCREASED_1DD correlates most to CTR_PT_INVOLV ==> if CTR_PT_INVOLV then AF_INCREASE at center 
# 
# 
# 
# u <- unzip("/Users/christiananderson/Downloads/archive.zip")
# h <- read.csv(u)
# h <- clean.it(h) %>% 
#   convert_to_numeric_if_possible()
# 
# h_cor <- h %>% 
#   dplyr::select( where(~is.numeric(.))) %>% 
#   as.matrix() %>% 
#   scale() %>%
#   cor(., use = "pairwise.complete.obs") 
# 
# idx <- rownames(h_cor) %>% grep("FOVTHICK|INR_EMP|OUTR_EMPT|ISOS_BRE|CTRPOINT",., value = TRUE) %>% sort
# Heatmap(h_cor[idx,idx],
#         cluster_columns= TRUE,
#         cluster_rows= TRUE,
#       col = c("white", "red")
# )



```
FAF{
• CTR_PT_INVOLV >> AF_INCREASED_1DD
OCTMAIN{
• CTRPOINT_INVOLVD >> INR_EMPTY_SPC
• CTRPOINT_INVOLV_2 >> OUTR_EMPTY_SPC

```{r look at dims}

check_dims()
# dims : 12525 200
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ 
spongebob()
```
