---
title: "Untitled"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---



# setup
```{r}
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()
options(max.print = 100)


# Packages
pacman::p_load(tidyr,reshape2,purrr,tidyverse,ggplot2,gridExtra,ComplexHeatmap,viridis,zoo,mgcv,mice,ggrepel,htmltools,Metrics,psych,car,stats,lme4,lmerTest,ape,SingleCellExperiment,scater,dplyr,HDF5Array,parallel, brms,cmdstanr,txtplot)

# sudo rm -rf /Library/Developer/CommandLineTools
# xcode-select --install
# xcode-select -p
# 
# CXX14FLAGS=-O3 -march=native -mtune=native -I/Library/Developer/CommandLineTools/SDKs/MacOSX.sdk/usr/include
# 
# CXX14 = clang++ -arch arm64 -std=c++14
# 
# check_cmdstan_toolchain()
# install_cmdstan(overwrite = TRUE)


rm(list = ls())
gc()

# functions

# script with all functions i have made
source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```



# Load Data
## sce
```{r}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

how.long()
# load experiment
sce <-  loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/sce006.h5"))
how.long()

```

# make new directorys for plots
```{r}

a_path <- paste0(wd_path, "results/clustering/14.1_progressionTimeModeling/")
if(!dir.exists(a_path)) {
  dir.create(a_path, recursive = TRUE)

}

```
 
#~~~~~~~~~~~~~~~~

# Model Time Difference by progression measure difference

> Fit model that predicts difference in time given difference in psuedotime (on longitudinal eyes)
> then run model on average cluster pseudoprog differences between each other 
  > output is predicted time difference between clusters

## chew

### model time between chew grades for each nth visit to every nth+1, nth+2, ...
```{r}

chew_time <- data.frame(colData(sce)) %>% 
  select(patID_eye, month, chew = chew_grade_derived_mine_loose) %>% 
  arrange(patID_eye, month) %>% 
  mutate(year = month /12) %>% 
  filter(n()>1, .by = patID_eye)


# MAKE STEPWISE DIFF D----
chew_eye_diff_d_raw <- structure(list(year = numeric(0), next_year = numeric(0), pp = numeric(0), 
    next_pp = numeric(0), pp_diff = numeric(0), year_diff = numeric(0), 
    pp_rate = numeric(0), patID_eye = character(0)), row.names = integer(0), class = "data.frame")

for(a_eye in unique(chew_time$patID_eye)){
  
  a_eye_diff_d <- chew_time[chew_time$patID_eye == a_eye, ]
  
  # Create year to following year combinations
  # year_combos <- data.frame(year = a_eye_diff_d$year, next_year = lead(a_eye_diff_d$year))
  year_combos <- as.data.frame(expand.grid( a_eye_diff_d$year, a_eye_diff_d$year))
  names(year_combos) <- c("year", "next_year")
  
  # Create chew to every other chew combinations
  # chew_combos <- data.frame(chew = a_eye_diff_d$chew, next_chew = lead(a_eye_diff_d$chew)) 
  chew_combos <- as.data.frame(expand.grid( a_eye_diff_d$chew, a_eye_diff_d$chew))
  names(chew_combos) <- c("chew", "next_chew")
    
  # Time chew combinations
  year_chew_combos <- cbind(year_combos, chew_combos)
  
  # filter for foward steps
  year_chew_combos <- year_chew_combos[year_chew_combos$year<year_chew_combos$next_year,]
  
  year_chew_combos$chew_diff <- with(year_chew_combos , (next_chew-chew))
  year_chew_combos$year_diff <-  with(year_chew_combos, (next_year-year))
  
  # make rate for each i to i+1: change in chew over change in time>> nextx-x/nextyear-year
  year_chew_combos$chew_over_year <- with(year_chew_combos , chew_diff/year_diff)
  year_chew_combos$year_over_chew <- with(year_chew_combos , year_diff/chew_diff)
  
  year_chew_combos$patID_eye <- a_eye
  
  year_chew_combos$first_chew <- a_eye_diff_d$chew[1]
  
  chew_eye_diff_d_raw <- rbind(chew_eye_diff_d_raw, year_chew_combos)
  
}

chew_eye_diff_d <- chew_eye_diff_d_raw


# MAKE OTHER VARS AND FILTER OUTLIERS ----
chew_eye_diff_d <- chew_eye_diff_d %>% 
  arrange(patID_eye, year, next_year) %>% 
  filter(chew<=next_chew) %>% 
  filter(chew!=6) %>% 
  ungroup()


# look at relationships
dat <- chew_eye_diff_d %>%
  # filter(dplyr::first(chew)<6, .by = patID_eye) %>%
  mutate(chew = factor(chew))

dat %>%
  ggplot( aes(year_diff, chew_diff))+
  geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .1)+
  stat_smooth(aes(color = chew, fill = chew), alpha = .5,method = "lm")+
  ggtitle("Marginal effects of chew stage on chew progression")+
  theme_bw()

# slopes <- dat %>%
#   group_by(chew) %>%
#   summarise(
#     estimate = coef(lm(chew_diff ~ year_diff))[["year_diff"]],
#     .groups = "drop"
#   ) %>%
#   mutate(years_per_chew = 1 / estimate)
# 
# slopes %>%
#   ggplot(aes(chew, years_per_chew))+
#   geom_line(aes(group = 1))+
#   geom_point(aes(group = 1))+
#   scale_y_continuous(breaks = seq(0,180, 10))+
#   theme_bw()

# # MODEL USING BAYESIAN METHODS ----
# 
# # Fit the model
# interaction_mod <- brm(
#   chew_diff~year_diff*as.character(chew),
#   data=chew_eye_diff_d ,
#   chains = 5,
#   cores = parallel::detectCores() - 1,
#   iter = 2000,
#   algorithm = "sampling",
#   backend = "cmdstanr",
#   silent = 2
# )
# 
# # prior_summary(interaction_mod)
# # plot(interaction_mod)
# summary(interaction_mod)
# # conditional_effects(interaction_mod)
# coefs <- posterior_summary(interaction_mod) %>% #, variable = c("b_Intercept" ,"b_year_diff" ,"b_chew" ,"b_year_diff:chew"  ,"sigma" ,"Intercept" )) %>%
#   as.data.frame() %>%
#   rownames_to_column("term")


# INVERT MODEL TO SOLVE FOR YEAR DIFF WITH EACH CHEW----
 
# # Fit the model
# interaction_mod <- lm(
#   chew_diff~year_diff*as.character(chew),
#   data=chew_eye_diff_d )
# 
# summary(interaction_mod)
# 
# plot(summary(interaction_mod)$coefficients[8:12,1])
# 
# 
# 
# # grab interaction terms for all levels of chew: year_diff intercept, eyar_diff:chew1, ... eyar_diff:chew5;   make named vector
# coefs <- coef(summary(interaction_mod)) %>% 
#   as.data.frame() 
# coefs$term <- rownames(coefs)
# slopes <- setNames(coefs$Estimate, coefs$term)
# 
# # effect of year change on chew change for chew == 0 plus additional effect at each level of chew
# intrxn_coefs <- slopes[grep("year_diff", names(slopes))]


# # Fit the model
# 
# dat <- chew_eye_diff_d %>%
#   mutate(chew = factor(chew)) %>%
#   ungroup()
# 
# dat <- dat[dat$patID_eye %in% sample(unique(dat$patID_eye), 100, F), ]
# 
# # Get the relevant priors for each coefficient as usual
# prior_list <- c(
#   prior(normal(0.1, 1), class = "b", coef = "year_diff"),
#   prior(normal(-0.1, 1), class = "b", coef = "year_diff:chew1"),
#   prior(normal(-0.2, 1), class = "b", coef = "year_diff:chew2"),
#   prior(normal(-0.3, 1), class = "b", coef = "year_diff:chew3"),
#   prior(normal(-0.4, 1), class = "b", coef = "year_diff:chew4"),
#   prior(normal(-0.5, 1), class = "b", coef = "year_diff:chew5")
# )
# 
# brm_interaction_mod <- brm(
#   # formula = chew_diff ~ year_diff*chew  + (year_diff|patID_eye),
#   formula = chew_diff ~ year_diff*chew  + (1|patID_eye),
#   family = gaussian(),
#   prior =prior_list,
#   data=dat,
# chains = 5,
#   cores = parallel::detectCores() - 1,
#   iter = 2000,
#   algorithm = "sampling",
#   backend = "cmdstanr",
#   silent = 2
# )
# 
# brm_interaction_mod %>%
#   # conditional_effects(effects = "year_diff:chew")
#   conditional_effects()


interaction_mod <- lm(
  chew_diff~year_diff*chew,
  data=chew_eye_diff_d )

# Get effect
eff <- effects::Effect(focal.predictors = c("year_diff", "chew"), interaction_mod)

# Extract as data frame
eff_df <- as.data.frame(eff)

# Reorder columns if needed
ggplot(eff_df, aes(x = year_diff, y = fit)) +
  geom_line(aes( group = chew,color = factor(chew)),linewidth = 1.2, alpha = .9) +
  geom_ribbon(aes(ymin = lower, ymax = upper,fill = factor(chew)), alpha = 0.3)+
  ggtitle("Marginal effect of pseudoprogression quantile on pseudoprogression-progregression")+
  theme_bw()


(summ <- summary(interaction_mod))
coefs <- summ$coefficients

# get coefs of interaction: year_diff estimate is 0:1, 1 is 1:2, ...
intrxn_coefs <- coefs[grepl("year_diff", rownames(coefs)) , "Estimate"]

# replicate chew interaction as each chew grade effect
intrxn_coefs <- c(intrxn_coefs[1],rep(intrxn_coefs[2],5))

names(intrxn_coefs) <- paste0("tochew", seq(intrxn_coefs))

# multiply coefs by index for each marginal effect
#+ first estimate is 0:1; second is 1:2, next is 2:3, which is twice the first; next 3 times teh first
intrxn_coefs <- intrxn_coefs*c(1,1:5)

# sum year diff effect with each interaction for each chew's full effect, not just marginal: slopes for each chew grade
#+ >> at each chew grade, how much does chew change with increase of one year?
full_effect <- intrxn_coefs
full_effect[-1] <- full_effect[1] + full_effect[-1]

# how much does year diff change given one unit change in chew grade?
recip_full_effect <- 1/full_effect

cumm_recip_full_effect <- cumsum(recip_full_effect)

plot(cumm_recip_full_effect, type = "n", yaxt = "n", main = "Years To Each Chew Grade", xlab = "Chew Grade", ylab = "Years")
axis(2, at = seq(-200,200,10), las = 2)
lapply(seq(1,200, 10), function(x) abline(h = x, col = rgb(0,0, 0,.2)))
lapply(seq(0,6, 1), function(x) abline(v = x, col = rgb(0,0, 0,.2)))
points(cumm_recip_full_effect, type = "b")


```

### model time between chew grades for each nth visit to every nth(chew_diff)-1

<!-- ```{r} -->

<!-- chew_time <- data.frame(colData(sce)) %>%  -->
<!--   select(patID_eye, month, chew = chew_grade_derived_mine_loose) %>%  -->
<!--   arrange(patID_eye, month) %>%  -->
<!--   mutate(year = month /12) %>%  -->
<!--   filter(n()>1, .by = patID_eye) %>%  -->
<!--   as.data.frame() -->



<!-- # FILTER FOR SALIENT TRANSITIONNS---- -->

<!-- df <- chew_time %>% -->
<!--   mutate(nvis = n(), .by = patID_eye) %>% -->
<!--   mutate(nchew = n_distinct(chew), .by = patID_eye) %>% -->
<!--   filter(nchew+nvis== max(nchew+nvis)) %>% -->
<!--   filter(patID_eye == dplyr::first(patID_eye)) -->

<!-- make_salient_differences <- function(df) { -->

<!--   df <- df %>% mutate(row = row_number()) -->
<!--   result <- list() -->

<!--   for (chew_level in unique(df$chew)) { -->
<!--     origin_row <- df %>% filter(chew == chew_level) %>% dplyr::slice(1) -->
<!--     origin_idx <- origin_row$row -->
<!--     origin_year <- origin_row$year -->

<!--     future_rows <- df %>% filter(row > origin_idx) -->

<!--     if (nrow(future_rows) > 0) { -->
<!--       out <- future_rows %>%  -->
<!--         # mutate( -->
<!--         #   chew_at_origin = chew_level, -->
<!--         #   transition = paste(origin_idx, "-", row, sep = ""), -->
<!--         #   year_diff = year - origin_year, -->
<!--         #   chew_diff = chew - chew_level -->
<!--         # ) -->
<!--         mutate( -->
<!--             chew_at_origin = chew_level, -->
<!--             transition = paste0(round(origin_year*2)/2, "-", round(year*2)/2), -->
<!--             year_diff = year - origin_year, -->
<!--             chew_diff = chew - chew_level -->
<!--           ) -->

<!--       # Longest transition with chew_diff == 0 -->
<!--       longest_zero <- out %>%  -->
<!--         filter(chew_diff == 0) %>%  -->
<!--         filter(year_diff == max(year_diff, na.rm = TRUE)) -->

<!--       # Shortest transitions for each distinct chew_diff > 0 -->
<!--       shortest_nonzero <- out %>% -->
<!--         filter(chew_diff > 0) %>% -->
<!--         group_by(chew_diff) %>% -->
<!--         dplyr::slice_min(order_by = year_diff, n = 1, with_ties = FALSE) %>% -->
<!--         ungroup() -->

<!--       filtered_out <- bind_rows(longest_zero, shortest_nonzero) %>% -->
<!--         select(patID_eye,chew_at_origin, transition, year_diff, chew_diff) -->

<!--       result[[as.character(chew_level)]] <- filtered_out -->
<!--     } -->
<!--   } -->

<!--   res <- bind_rows(result) -->
<!--   rownames(res) <- seq(nrow(res)) -->

<!--   return(res) -->
<!-- } -->



<!-- chew_eye_diff_d <- mclapply(unique(chew_time$patID_eye), function(eye){ -->
<!--   make_salient_differences(chew_time[chew_time$patID_eye==eye, ]) -->
<!-- }, mc.cores = detectCores()-1) %>%  -->
<!--   do.call(rbind,.) -->

<!-- setdiff_intersect(list(chew_eye_diff_d$patID_eye, chew_time$patID_eye)) -->
<!-- # Only Set1 : 0  -->
<!-- # Only Set2 : 0  -->
<!-- # Set1 ∩ Set2 : 975  -->


<!-- mod <- lm(chew_diff~year_diff*as.character(chew_at_origin), -->
<!--           data=chew_eye_diff_d %>%  -->
<!--             ungroup() -->
<!--             ) -->
<!-- summary(mod) -->

<!-- # grab year and interaction terms -->
<!-- coefs <- summary(mod)$coefficients -->
<!-- coefs <- coefs[grep("year|\\:",rownames(coefs)),"Estimate"] -->

<!-- # add year term to interactions; each interaction is additional slope from year_diff's slope -->
<!-- coef_sum <- c(coefs[1],coefs[2:6]+coefs[1]) -->

<!-- # years_to_chew <- cumsum(1/coef_sum[-length(coef_sum)]) -->
<!-- years_to_chew <- cumsum(1/coef_sum) -->


<!-- { -->
<!--   plot_vec <- years_to_chew  -->
<!--   # plot_vec <- years_to_chew[-length(years_to_chew)]  -->
<!--   years_to_chew_d <- data.frame(chew = seq(plot_vec), year =plot_vec) -->

<!--   par(mar = rep(5,4)) -->
<!--   plot(years_to_chew_d, ylab = "Years", xlab = "Chew Grade", main = "Time between Chew grade stages", type = "b") -->
<!--   line_m <- lm(year~poly(chew,2), years_to_chew_d) -->
<!--   line_m %>% summary() -->
<!--   line_d <- data.frame(chew = seq(0,6, .2))  -->
<!--   line_d$year <- predict(line_m, line_d) -->
<!--   curve(x^2+26, add = TRUE, col = "darkgrey", lty = 3) -->
<!--   # lines(line_d, col = "blue", lwd = 3) -->
<!--   legend("topleft",  -->
<!--          legend = c( -->
<!--            # paste0("y ~ exp(x) (adj. R^2 = ", round(summary(line_m)$adj.r.squared,2),")"), -->
<!--            paste0("cummulativeSum{1/(chew_diff ~ year_diff * chew)} "), -->
<!--            "x^2" -->
<!--            ), -->
<!--          cex = .7,  -->
<!--          fill = c( -->
<!--            # "blue", -->
<!--            "black", -->
<!--            "darkgrey" -->
<!--            ), -->
<!--          box.col = NA) -->
<!-- } -->


<!-- # lm(chew_rate~0+ transition,data = chew_eye_diff_d %>% filter(chew!=next_chew) %>% mutate(transition = paste(chew,next_chew,sep = "--"))) %>%  -->
<!-- year_diff_mod <- lm(year_diff~0+ transition, -->
<!--                     data = chew_eye_diff_d %>%  -->
<!--                       filter(chew<next_chew) %>% -->
<!--                       filter(chew_diff!=0) %>% -->
<!--                       mutate(transition = paste(chew,next_chew,sep = "--")) -->
<!--                     ) %>% -->
<!--   summary() -->
<!-- year_diff_mod -->

<!-- ``` -->



## pseudoprog
### model time diff given pp diff
```{r}

pp_time <- data.frame(colData(sce)) %>% 
  select(patID_eye, month, pp = pseudoprog) %>% 
  arrange(patID_eye, month) %>% 
  mutate(year = month /12) %>% 
  filter(n()>1, .by = patID_eye)

pp_quantiles <- quantile(pp_time$pp, seq(0,1,.1))
pp_time$pp <- cut(pp_time$pp,pp_quantiles , labels = FALSE, include.lowest = TRUE)-1


# MAKE STEPWISE DIFF D----
eye_diff_d_raw <- structure(list(year = numeric(0), next_year = numeric(0), pp = numeric(0), 
    next_pp = numeric(0), pp_diff = numeric(0), year_diff = numeric(0), 
    pp_rate = numeric(0), patID_eye = character(0)), row.names = integer(0), class = "data.frame")

for(a_eye in unique(pp_time$patID_eye)){
  
  a_eye_diff_d <- pp_time[pp_time$patID_eye == a_eye, ]
  
  # Create year to following year combinations
  # year_combos <- data.frame(year = a_eye_diff_d$year, next_year = lead(a_eye_diff_d$year))
  year_combos <- as.data.frame(expand.grid( a_eye_diff_d$year, a_eye_diff_d$year))
  names(year_combos) <- c("year", "next_year")
  
  # Create pp to every other pp combinations
  # pp_combos <- data.frame(pp = a_eye_diff_d$pp, next_pp = lead(a_eye_diff_d$pp)) 
  pp_combos <- as.data.frame(expand.grid( a_eye_diff_d$pp, a_eye_diff_d$pp))
  names(pp_combos) <- c("pp", "next_pp")
    
  # Time pp combinations
  year_pp_combos <- cbind(year_combos, pp_combos)
  
  year_pp_combos <- year_pp_combos[year_pp_combos$year<year_pp_combos$next_year,]
  
  year_pp_combos$pp_diff <- with(year_pp_combos , (next_pp-pp))
  year_pp_combos$year_diff <-  with(year_pp_combos, (next_year-year))
  
  # make rate for each i to i+1: change in pp over change in time>> nextx-x/nextyear-year
  year_pp_combos$pp_over_year <- with(year_pp_combos , pp_diff/year_diff)
  year_pp_combos$year_over_pp <- with(year_pp_combos , year_diff/pp_diff)
  
  year_pp_combos$patID_eye <- a_eye
  
  year_pp_combos$first_pp <- a_eye_diff_d$pp[1]
  
  eye_diff_d_raw <- rbind(eye_diff_d_raw, year_pp_combos)
  
}

eye_diff_d <- eye_diff_d_raw


# MAKE OTHER VARS AND FILTER OUTLIERS ----

eye_diff_d <- eye_diff_d %>% 
  arrange(patID_eye, year, next_year) %>% 
  mutate_all(~ifelse(is.infinite(.), NA, .)) %>% 
  # mutate(across(c(pp_over_year, year_over_pp), ~log(.+1), .names= "log_{.col}")) %>% 
  # mutate(across(c(pp_over_year, year_over_pp), ~sqrt(.), .names= "sqrt_{.col}")) %>% 
  filter(pp<=next_pp) %>% 
  filter(pp!=10) %>% 
  ungroup()

# eye_diff_d$pp <- factor(eye_diff_d$pp, levels = as.character(1:10))

# look at relationships
dat <- eye_diff_d %>%
  # filter(dplyr::first(chew)<6, .by = patID_eye) %>%
  mutate(pp = factor(pp))

dat %>%
  ggplot( aes(year_diff, pp_diff))+
  geom_line(aes(group = patID_eye), show.legend = FALSE, alpha = .1)+
  stat_smooth(aes(color = pp, fill = pp), alpha = .5,method = "lm")+
  ggtitle("Marginal effect of pseudoprogression quantile on pseudoprogression-progregression")+
  theme_bw()


# # MODEL ----
# interaction_mod <- lmer(pp_diff ~ year_diff * pp + (1|patID_eye), data = eye_diff_d)
interaction_mod <- lm(pp_diff ~ year_diff * pp , data = eye_diff_d )
summary(interaction_mod)

# interaction_brm <- brm(
#   pp_diff ~ year_diff * pp,
#   data = eye_diff_d,
#   chains = 5,
#   cores = parallel::detectCores() - 1,
#   prior = c(
#     prior(normal(1,1), coef = "year_diff:pp")
#   ),
#   iter = 2000,
#   algorithm = "sampling",
#   backend = "cmdstanr",
#   silent = 2
# )
# interaction_brm
# shinystan::launch_shinystan(interaction_brm)
# 
# plot(interaction_brm)
# conditional_effects(interaction_brm,effects = "year_diff:pp_quantile")
# post <- data.frame(year_diff = 0, pp_quantile = factor("2")) %>% 
#   posterior_predict(interaction_brm, .)
# plot(density(post));abline(v = mean(post));text(y = .02,x = mean(post), labels = round(mean(post), 2))
# 
# 
# as_draws_matrix(interaction_brm) %>%
#     as.data.frame() %>%
#     select(-matches("lp")) %>%
#     pivot_longer(everything(), names_to = "parameter", values_to = "estimate") %>%
#     ggplot(aes(estimate))+
#     theme_bw()+
#     geom_density(aes(fill = parameter,color = parameter), alpha = .5)+
#     facet_wrap(~parameter, scales ="free_y")

# # Get effect
# eff <- effects::Effect(focal.predictors = c("year_diff", "pp"), interaction_mod)
# 
# # Extract as data frame
# eff_df <- as.data.frame(eff)
# 
# # Reorder columns if needed
# ggplot(eff_df, aes(x = year_diff, y = fit)) +
#   geom_line(aes( color = pp),linewidth = 1.2, alpha = .9) +
#   geom_ribbon(aes(ymin = lower, ymax = upper,fill = pp), alpha = 0.3)+
#   ggtitle("Marginal effect of pseudoprogression quantile on pseudoprogression-progregression")+
#   theme_bw()

# # MODEL USING BAYESIAN METHODS ----
# # make quantiles and do same model
# # Fit the model
# interaction_mod <- brm(
#   pp_diff~year_diff*pp ,
#   # year_diff~pp_diff*pp ,
#   data=eye_diff_d,
#   prior =c(
#     # prior(normal(0.1, 0.05), coef = pp), # pp has positive effect
#     # prior(normal(0.1, 0.05), coef = pp_diff), # pp_diff has positive relationship
#     # prior(normal(-0.1, 0.05), coef = `year_diff:pp`)# pp has negative effect on the effect that longer passages of time has on the passage of pp
#   ),
#   chains = 5,
#   cores = parallel::detectCores() - 1,
#   iter = 2000,
#   # algorithm = "sampling",
#   backend = "cmdstanr",
#   silent = 2
# )
# 
# # prior_summary(interaction_mod)
# # plot(interaction_mod)
# summary(interaction_mod)
# conditional_effects(interaction_mod)
# 
# coefs <- posterior_summary(interaction_mod, variable = c("b_Intercept" ,"b_year_diff" ,"b_pp" ,"b_year_diff:pp"  ,"sigma" ,"Intercept" )) %>%
#   as.data.frame() %>%
#   rownames_to_column("term")

# # INVERT MODEL TO SOLVE FOR YEAR DIFF----
# 
# # ****** FUNCTION to predict year diff from ppdiff and pp by reversing the equation
# yeardiff_given_ppdiff_pp <- function(df){
#   
#   # pp_diff = β₀ + β₁ * year_diff + β₂ * pp + β₃ * year_diff * pp
#   # pp_diff = β₀ + β₂ * pp + year_diff * (β₁ + β₃ * pp)
#   # year_diff = (pp_diff - β₀ - β₂ * pp) / (β₁ + β₃ * pp)
#   
#   # BRM
#   B0 <- coefs$Estimate[coefs$term=="b_Intercept"]
#   B1 <- coefs$Estimate[coefs$term=="b_year_diff"]
#   B2 <- coefs$Estimate[coefs$term=="b_pp"]
#   B3 <- coefs$Estimate[coefs$term=="b_year_diff:pp"]
#   
#   # # LM
#   # B0 <- coefs$Estimate[coefs$term=="Intercept"]
#   # B1 <- coefs$Estimate[coefs$term=="year_diff"]
#   # B2 <- coefs$Estimate[coefs$term=="pp"]
#   # B3 <- coefs$Estimate[coefs$term=="year_diff:pp"]
# 
#   if( all(c("pp_diff", "pp") %in% names(df))){ 
#     year_diff <-   (df$pp_diff - B0 - (B2*df$pp)) / (B1+(B3*df$pp))
#     return(year_diff)
#   }else{
#     warning("Input data frame must contain columns 'pp_diff' and 'pp'")
#   }
# }


# # Set up range of x values
# x <- seq(-10, 10, length.out = 1000)
# 
# 
# {
#   df <- 3
#   mu = 0
#   sigma = 1
#   # Density of Student-t with df=1, mean=0, scale=1
#   y <- dt((x - mu)/sigma, df = df) / sigma  # scale the density
#   
#   # Plot
#   plot(x, y, type = "l", lwd = 2,
#        main = sprintf("Student-t(%s, %s, %s) Distribution",df,mu,sigma ),
#        xlab = "x", ylab = "Density")
#   abline(v = mu, lty = 2, col = "gray")  # vertical line at the mean
#   abline(v =0, lty = 3, col = "darkgray")  # vertical line at the mean
# }

```

#### predict time change between cluster PT averages
between pseudoprog bins, predict waht the time change is
```{r}


# FEED MODEL ACTUAL DIFFERENCES BETWEEN CLUSTER MEAN PTS ----
# make matrix of differences between each cluster
## cluster PT averages; add 0th cluster for i-1 calc later on
cluster_pp_means <- aggregate(pseudoprog~my_cluster,  data  = as.data.frame(colData(sce)), mean, simplify = TRUE)
cluster_pp_means$my_cluster <- as.numeric(as.character(cluster_pp_means$my_cluster))

cluster_pp_means$pp_quantile <- cut(cluster_pp_means$pseudoprog, pp_quantiles, labels = FALSE, include.lowest = T)-1

# difference between each cluster's means; exclude dummy cluster 0 at ind 1; times -1 to make foward positive difference
cluster_diff_m <- outer(cluster_pp_means$pseudoprog, cluster_pp_means$pseudoprog, "-") %>% abs()# * -1
dimnames(cluster_diff_m) <- list(as.character(cluster_pp_means$my_cluster), as.character(cluster_pp_means$my_cluster))

# # remove lower.triangle
# cluster_diff_m[lower.tri(cluster_diff_m)] <- NA

Heatmap(cluster_diff_m ,
        col = plasma(50),
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        row_names_side = "left",
        column_names_side = "top",
        column_title = "Difference in pseudoprog mean between clusters"
        )


# ## for each transition, model difference between current clust and next; interaction term of previous PT is taken from cluster_pp_means
#
# # d of matrix indices
# inds <- expand.grid(i = 1:nrow(cluster_diff_m), j = 1:ncol(cluster_diff_m), KEEP.OUT.ATTRS = F)
#
# # ignore diagonal
# inds <- inds[inds$i!=inds$j,]
#
# predictions_d <- lapply(seq(nrow(inds)), function(n){
#  # grab which clusters to compare between
#   current_clust = inds[n,1]
#   next_clust = inds[n,2]
#
#   # grab the difference in mean between current and next cluster
#   pp_diff <- cluster_diff_m[current_clust, next_clust]
#
#   # grab the mean of the previous cluster; if current cluster is 1 return numeric(0)
#   pp <- cluster_pp_means$pseudoprog[cluster_pp_means$my_cluster==current_clust]
#
#   # model
#   newdat <- data.frame(pp_diff=pp_diff, pp=pp, first_pp = 15)
#   newdat$i = current_clust
#   newdat$j = next_clust
#   return(newdat)
# }) %>%
#   do.call(rbind,.) %>%
#   as.data.frame()
#
# # # use brm mod to predict year diff
# # predictions_d <-cbind(predictions_d,predict(interaction_mod, predictions_d, re_formula = NA))
#
# # # use my ppdiff model and revers terms to calculate year diff
# # predictions_d$Estimate <-yeardiff_given_ppdiff_pp(predictions_d)
#
# # Create empty matrix of appropriate size
# predictions <- matrix(NA, nrow = max(predictions_d$i), ncol = max(predictions_d$j))
#
# # Fill matrix with Estimate values
# for (row in 1:nrow(predictions_d)) {
#   predictions[predictions_d$i[row], predictions_d$j[row]] <- predictions_d$Estimate[row]
# }
#
# # Optionally, assign row and column names
# rownames(predictions) <- paste0("", 1:nrow(predictions))
# colnames(predictions) <- paste0("", 1:ncol(predictions))
#
#
# Heatmap(predictions ,
#         col = plasma(100),
#         cluster_columns = FALSE,
#         cluster_rows = FALSE,
#         row_names_side = "left",
#         column_names_side = "top",
#         name = "predicted dT",
#         column_title  = "Predicted dT between clusters using difference in mean between clusters",
#         )


```

##### Calculate dT from steps between trajectory transitions using continuous pp model
<!-- ```{r} -->

<!-- # metadata(sce)$trajectory_data$pca_line_data[,"edge",drop = FALSE] %>%  -->
<!-- #   separate(edge, c("from", "to"), "--") %>%  -->
<!-- #   unique() %>%  -->
<!-- #   as.matrix() %>%  -->
<!-- #   igraph::graph_from_edgelist(directed = F) %>%  -->
<!-- #   plot() -->

<!-- cummulative_pred <- predictions*NA -->

<!-- # # start by ignoriing cluster 3 mean; it's not zero so not sure how to interpret it -->
<!-- cummulative_pred[3,6] <- predictions[3,6] -->

<!-- # # start by adding cluster 3 mean -->
<!-- # cummulative_pred[3,6] <- predictions[3,3] + predictions[3,6] -->

<!-- cummulative_pred[6,4] <- cummulative_pred[3,6] + predictions[6,4] -->

<!-- cummulative_pred[4,9] <- cummulative_pred[6,4]+predictions[4,9] -->
<!-- cummulative_pred[9,2] <- cummulative_pred[4,9]+predictions[9,2] -->
<!-- cummulative_pred[2,7] <- cummulative_pred[9,2]+predictions[2,7] -->
<!-- cummulative_pred[2,13] <- cummulative_pred[9,2]+predictions[2,13] -->

<!-- cummulative_pred[4,1] <- cummulative_pred[6,4]+predictions[4,1] -->
<!-- cummulative_pred[1,5] <- cummulative_pred[4,1]+predictions[1,5] -->
<!-- cummulative_pred[1,10] <- cummulative_pred[4,1]+predictions[1,10] -->
<!-- cummulative_pred[1,8] <- cummulative_pred[4,1]+predictions[1,8] -->
<!-- cummulative_pred[1,11] <- cummulative_pred[4,1]+predictions[1,11] -->


<!-- cummulative_pred <- cummulative_pred[,order(colSums(cummulative_pred, na.rm = TRUE))] -->
<!-- cummulative_pred <- cummulative_pred[rowSums(!is.na(cummulative_pred))!=0 , colSums(!is.na(cummulative_pred))!=0] -->


<!-- library(circlize) -->
<!-- Heatmap(cummulative_pred, -->
<!--         cluster_rows = FALSE, -->
<!--         cluster_columns = FALSE, -->
<!--         col = colorRamp2(seq(min(cummulative_pred, na.rm = TRUE), -->
<!--                              max(cummulative_pred, na.rm = TRUE), length.out = 10), -->
<!--                          plasma(10, begin = .4)), -->
<!--         heatmap_legend_param = list(title = "Time"), -->
<!--         show_heatmap_legend = FALSE, -->
<!--         cell_fun = function(j, i, x, y, width, height, fill) { -->
<!--           grid.text(sprintf("%.1f", cummulative_pred[i, j]), x, y, gp = gpar(fontsize = 12)) -->
<!--         }, -->
<!--         row_names_side = "left", # <- this is the key line -->
<!--         border = TRUE, -->
<!--         column_title = "Estimated time passage along MST from cluster 3") -->


<!-- # Heatmap(cummulative_pred ,  -->
<!-- #         col = plasma(100), -->
<!-- #         cluster_columns = FALSE,  -->
<!-- #         cluster_rows = FALSE,  -->
<!-- #         row_names_side = "left",  -->
<!-- #         column_names_side = "top",  -->
<!-- #         name = "predicted dT",  -->
<!-- #         column_title  = "Predicted cummulative time along tree",  -->
<!-- #         ) -->

<!-- cummulative_pred %>%  -->
<!--   as.data.frame() %>% -->
<!--   rownames_to_column("from") %>%  -->
<!--   pivot_longer(-from, names_to = "to", values_to =  "years") %>%  -->
<!--   mutate(transition =paste(from,to,sep= "-")) %>%  -->
<!--   drop_na() %>%  -->
<!--   ggplot(aes(reorder(transition, years), years))+ -->
<!--   geom_bar(stat = "identity", width = .05, aes(fill = transition), show.legend = FALSE)+ -->
<!--   geom_point(aes(color = transition), show.legend = FALSE, size = 6)+ -->
<!--   theme_bw() -->
<!-- ``` -->


##### Calculate dT from steps between trajectory transitions using discrete pp model
```{r}

# summ <- summary(interaction_mod)
# 
# coefs <- coef(summ) %>% 
#   as.data.frame()
# coefs$term <- rownames(coefs)
# 
# slopes <- setNames(coefs$Estimate, coefs$term)
# 
# # effect of year change on chew change for chew == 0 plus additional effect at each level of chew
# intrxn_coefs <- slopes[grep("year_diff", names(slopes))]

(summ <- summary(interaction_mod))

coefs <- summ$coefficients

# get coefs of interaction: year_diff estimate is 0:1, 1 is 1:2, ...
intrxn_coefs <- coefs[grepl("year_diff", rownames(coefs)) , "Estimate"]

# replicate chew interaction as each chew grade effect
rep_n <- 8
intrxn_coefs <- c(intrxn_coefs[1],rep(intrxn_coefs[2],rep_n))
names(intrxn_coefs) <- paste0("topp", seq(intrxn_coefs))

# multiply coefs by index for each marginal effect
#+ first estimate is 0:1; second is 1:2, next is 2:3, which is twice the first; next 3 times teh first
intrxn_coefs <- intrxn_coefs*c(1,1:rep_n)

# sum year diff effect with each interaction for each chew's full effect, not just marginal: slopes for each chew grade
#+ >> at each ppquantile, how much does ppquantile change with increase of one year?
full_effect <- intrxn_coefs
full_effect[-1] <- full_effect[1] + full_effect[-1]

# how much does year diff change given change in ppquantile?
recip_full_effect <- 1/(full_effect)

# make last 0 because so inflated?
#recip_full_effect[length(recip_full_effect)] <- 0

years_to_quantile <- cumsum(recip_full_effect)

names(years_to_quantile) <- as.character(0:(length(years_to_quantile)-1))

labs <- paste0(1:9)
ylim = c(0,round(max(years_to_quantile)+5))
xlim = c(1,length(years_to_quantile))
{
  plot(years_to_quantile,
     xlim = xlim, ylim = ylim,
     type = "n", yaxt = "n",xaxt = "n", 
     main = "Years To Each Pseudotime Quantile", xlab = "Pseudotime Quantile", ylab = "Years")
  # axis(1, at = seq(xlim[1],xlim[2],1), las = 1)
  axis(1, labels = labs,at = seq(xlim[1],xlim[2],1), las = 1)
  axis(2, at = seq(ylim[1],ylim[2],5), las = 2)
  lapply(seq(xlim[1],xlim[2],1), function(x) abline(v = x, col = rgb(0,0, 0,.2)))
  lapply(seq(ylim[1],ylim[2], 5), function(x) abline(h = x, col = rgb(0,0, 0,.2) , lty = ifelse(x%%2==0, 1,2)))
  points(years_to_quantile, type = "b")
# ^ what amount of time passes with each increase in ppquantile on top of previous time increases?
}


# how  long to each cluster?----
## ad to cluster info
years_to_quantile_x <- c("0" = 0, years_to_quantile)
names(years_to_quantile_x) <- as.character(0:(length(years_to_quantile_x)-1))

cluster_pp_means$year_to_quintile <- years_to_quantile_x[match( as.character(cluster_pp_means$pp_quantile),names(years_to_quantile_x))]
cluster_pp_means <- cluster_pp_means[order(cluster_pp_means$year_to_quintile),]
cluster_pp_means$my_cluster <- factor(cluster_pp_means$my_cluster, levels = cluster_pp_means$my_cluster)

cluster_pp_means%>% 
  # mutate(year_to_quintile = year_to_quintile-min(year_to_quintile)) %>% 
  ggplot(aes(my_cluster, year_to_quintile))+
  geom_hline(yintercept = seq(0,120, 20), linetype = 2, color = "grey")+
  geom_hline(yintercept = 0, linetype = 1, color = "darkgrey", linewidth = 1.5)+
  geom_bar(stat = 'identity', aes(fill = my_cluster), width = .1, show.legend = FALSE)+
  geom_point(aes(color = my_cluster), show.legend = FALSE, size = 8)+
  geom_smooth(data = cluster_pp_means , aes(my_cluster, year_to_quintile, group = 1), 
              # method = "gam", formula = y~ s(x,k = 4)
              method = "loess"
              )+
  ylab("Year to cluster")+
  scale_y_continuous(breaks = seq(0,120, 10))+
  ggtitle("Time to reach each cluster")+
  theme_bw()


# how  long from from clust to to clust? ----
# fill matrix with differences between these clusters
from_to <- matrix(c(3,6,6,4,4,9,9,2,2,7,2,13,4,1,1,5,1,10,1,11,1,8 ), byrow = TRUE, ncol = 2)
year_to_clust <- matrix(NA, ncol = n_distinct(sce$my_cluster), nrow = n_distinct(sce$my_cluster), 
                        dimnames = list(as.character(seq(n_distinct(sce$my_cluster))),as.character(seq(n_distinct(sce$my_cluster)))))


#+ difference in times between quantiles of fromcluster and tocluster
for(i in 1:nrow(from_to)){
  row <- from_to[i,]

  # origin cluster
  from = row[[1]]
  # destination cluster
  to = row[[2]]

  # find difference between source and dest clusters times
  from_years <- cluster_pp_means$year_to_quintile[cluster_pp_means$my_cluster==from]
  to_years <- cluster_pp_means$year_to_quintile[cluster_pp_means$my_cluster==to]
  year_diff <- to_years-from_years
  
  year_to_clust[from,to] <- year_diff
}


# accumulate time from cluster 3 to 1 and 2 (ignoring end clusters because unreliable estimates)
year_from_3 <- year_to_clust

year_from_3[6,4] <- year_from_3[3,6]+year_from_3[6,4]
  year_from_3[4,1] <- year_from_3[6,4]+year_from_3[4,1]
    year_from_3[1,5] <- year_from_3[4,1]+year_from_3[1,5]
    year_from_3[1,8] <- year_from_3[4,1]+year_from_3[1,8]
    year_from_3[1,10] <- year_from_3[4,1]+year_from_3[1,10]
    year_from_3[1,11] <- year_from_3[4,1]+year_from_3[1,11]

  year_from_3[4,9] <- year_from_3[6,4]+year_from_3[4,9]
    year_from_3[9,2] <- year_from_3[4,9] +year_from_3[9,2]
      year_from_3[2,7] <- year_from_3[9,2]+year_from_3[2,7]
      year_from_3[2,13] <- year_from_3[9,2]+year_from_3[2,13]


      
# plot
## filter out all na row/cols
year_to_clust <- year_to_clust[rowSums(!is.na(year_to_clust))!=0 , colSums(!is.na(year_to_clust))!=0]
## reorder
year_to_clust <- year_to_clust[order(rowSums(year_to_clust,na.rm = TRUE), decreasing = FALSE),order(colSums(year_to_clust,na.rm = TRUE), decreasing = FALSE)]

# stepwise
Heatmap(year_to_clust, 
        cluster_rows = FALSE, 
        cluster_columns = FALSE, 
        col = circlize::colorRamp2(seq(min(year_to_clust, na.rm = TRUE), 
                             max(year_to_clust, na.rm = TRUE), length.out = 10),
                         plasma(10, begin = .4)), 
        heatmap_legend_param = list(title = "Time"), 
        show_heatmap_legend = FALSE, 
        cell_fun = function(j, i, x, y, width, height, fill) { 
                      grid.text(sprintf("%.1f",  year_to_clust[i, j]), x, y, gp = gpar(fontsize = 12)) 
                    }, 
        row_names_side = "left",
        border = TRUE, 
        column_title = "Estimated years from cluster to cluster") 




# plot
## filter out all na row/cols
year_from_3 <- year_from_3[ as.numeric(rownames(year_to_clust)) , as.numeric(colnames(year_to_clust))]

rownames(year_from_3) <- rep("3", nrow(year_from_3))

# cummulative 
Heatmap(year_from_3, 
        cluster_rows = FALSE, 
        cluster_columns = FALSE, 
        col = circlize::colorRamp2(seq(min(year_from_3, na.rm = TRUE), 
                             max(year_from_3, na.rm = TRUE), length.out = 10),
                         plasma(10, begin = .4)), 
        heatmap_legend_param = list(title = "Time"), 
        show_heatmap_legend = FALSE, 
        cell_fun = function(j, i, x, y, width, height, fill) { 
                      grid.text(sprintf("%.1f",  year_from_3[i, j]), x, y, gp = gpar(fontsize = 12)) 
                    }, 
        row_names_side = "left",
        border = TRUE, 
        column_title = "Estimated years to each cluster from 3") 




```

# plot time from three on tsne
```{r}

from_3_1d <- year_from_3[!is.na(year_from_3)]
names(from_3_1d) <- colnames(year_from_3)

from_3_1d <- from_3_1d[-which(names(from_3_1d)%in%c("5", "8", "10", "11", "13", "7"))]
from_3_1d["3"] <- 0

tsne <- as.data.frame(reducedDim(sce, "TSNE")) %>% 
  rownames_to_column("sample") %>% 
  left_join(
    data.frame(colData(sce)) %>% select(sample, my_cluster)
  ) 

tree <- metadata(sce)$trajectory_data$tsne_line_data %>% 
  mutate(x = edge) %>% 
  separate(x, c("from", "to"), "--") 
  
labels <- tree %>% 
  distinct(across(c(TSNE1,TSNE2)), .keep_all = T) %>% 
  arrange(TSNE1,TSNE2)

labels$my_cluster = factor(c(5,10,3,11,8,1,6,2,4,9,13,7,12))

timelabels <- labels
timelabels$TSNE1 <- timelabels$TSNE1+
  c(0,0,4,0,0,-6,5,5,5,6,0,0,0)

timelabels$TSNE2 <- timelabels$TSNE2+
  c(0,0,2,0,0,-3,2,2,2,2,0,0,0)

ggplot()+
  geom_point(data = tsne, aes(TSNE1, TSNE2, color = my_cluster), alpha = .3)+
  coord_equal()+
  geom_line(data = tree, aes(TSNE1, TSNE2, group = edge), alpha = .6, linewidth = 1)+
  geom_label(data = labels, aes(TSNE1, TSNE2, label = my_cluster),  alpha = .5)+
  geom_label(data = timelabels, size = 5,
             aes(TSNE1, TSNE2, fill = my_cluster, label = from_3_1d[match(my_cluster, names(from_3_1d))] %>% round(2)))+
  scale_color_manual(values =  cluster_cols)+
  scale_fill_manual(values =  cluster_cols)+
  ggtitle("Years to each cluster")+
  labs(fill = "Cluster", color = "Cluster")+
  theme_bw()
  
ggsave(sprintf("%sresults/publication_figures/TSNE_cluster_with_time_from_3.pdf", wd_path), height = 10, width = 11)

```





