---
title: "14_GeneticProgressionModeling"
author: "Christian Anderson"
date: "`r Sys.Date()`"
output: html_document
---


# options
```{r}
# dev.off()  # Turn off the current graphics device
# options(device = "RStudioGD")
# quartz()
options(max.print = 100)
```


# Packages
```{r, message=FALSE, warning=FALSE, results='hide', echo = TRUE}

pacman::p_load(tidyr,reshape2,purrr,GGally,tidyverse,ggplot2,gridExtra,ComplexHeatmap,viridis,zoo,mgcv,tools,mice,ggrepel,htmltools,Metrics,psych,car,stats,lme4,lmerTest,ape,SingleCellExperiment,scater,velociraptor,skimr,dplyr,HDF5Array,randomForest,psych,parallel)

rm(list = ls())
gc()
```

# functions
```{r}

# script with all functions i have made
source("/Users/canderson/Documents/LMRI/PRJ2024001/version015/analysis/00_MyFunctions.R")

```



# Load Data   
## sce
```{r}

wd_path = "~/Documents/LMRI/PRJ2024001/version015/"

how.long()
# load experiment
sce <-  loadHDF5SummarizedExperiment(paste0(wd_path, "processed_data/sce006.h5"))
how.long()

```

# make new directorys for plots
```{r}

a_path <- paste0(wd_path, "results/clustering/14_geneticProgressionModeling/")
if(!dir.exists(a_path)) {
  dir.create(a_path, recursive = TRUE)
  
}

```

<!-- ## Genetics data -->
<!-- ### SNP data -->
<!-- ```{r} -->



<!-- # peak sig snp stats -->
<!-- all_mactel_loci = read.delim("/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/raw_data/Liams_genetics/MacTel_GWAS2023_Report/All_MacTel_Loci.txt") -->

<!-- # allele freqs on patients for sig snps -->
<!-- patient_allele_freqs= read.delim("/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/raw_data/Liams_genetics/MacTel_GWAS2023_Report/MacTel_GWAS2023_Sample_Data.txt") -->

<!-- # gwas tower significant snps? -->
<!-- mapped_genes <- read.delim("/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/raw_data/Liams_genetics/MacTel_GWAS2023_Report/Mapped_Genes.txt", sep = "\t")  -->

<!-- # rare-variant data -->
<!-- rare_variants <- openxlsx::read.xlsx("/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/raw_data/Liams_genetics/rare_variant_08_2023.xlsx",sep = "\t") -->

<!-- # save to sce; replace sig_loci to list -->
<!-- metadata(sce)$genetics <-list(all_mactel_loci = all_mactel_loci,  -->
<!--                               patient_allele_freqs = patient_allele_freqs, -->
<!--                               mapped_genes = mapped_genes,  -->
<!--                               rare_variants = rare_variants,  -->
<!--                               sig_loci  = metadata(sce)$sig_loci -->
<!--                               ) -->
<!-- # sigloci in metadata, added to list instead so remove -->
<!-- metadata(sce)$sig_loci <- NULL -->
<!-- names(metadata(sce)) -->

<!-- ``` -->


<!-- ### rare variant table and genetics pcs -->
<!-- ```{r} -->

<!-- # JOIN RARE VARIANT DATA ---- -->
<!-- # make dichotomous for rare variant -->
<!-- rvs <- unique(rare_variants$Gene) -->
<!-- for(rv in rvs){ -->
<!--   rv_presence <- as.numeric(rare_variants$Gene==rv) -->
<!--   # make zeros NA because we don't know for sure that they don't have this -->
<!--   rv_presence[rv_presence==0] <- NA -->
<!--   rare_variants[[rv]] <- rv_presence -->
<!-- } -->

<!-- # coldata to add to sce -->
<!-- coldat_to_add <- colData(sce) %>%  -->
<!--   as.data.frame() %>%  -->
<!--   select(patID, month, eye_num, sample) -->

<!-- # read patient ID tbale and clean for chars to add a.number to sce -->
<!-- pids <- read.csv("/Users/christiananderson/Library/CloudStorage/Egnyte-lmri/Shared/LMRI/DS_Group/PRJ2024001/raw_data/PatientIDs.csv") %>%  -->
<!--   clean.it() -->
<!-- # remove artefact -->
<!-- pids <- pids[-1,] -->

<!-- # add alternate ids to coldata -->
<!-- coldat_to_add <- coldat_to_add %>%  -->
<!--   left_join(pids, relationship = "many-to-one", by = join_by(patID ==Patient.ID)) -->

<!-- names(rare_variants)[names(rare_variants) %in% c("PHGDH", "SPTLC1", "SPTLC2")] <- paste0(c("PHGDH", "SPTLC1", "SPTLC2"), "_RV") -->

<!-- # join rare_variants by A.number (a.number = AID) -->
<!-- # coldat_to_add[coldat_to_add$A.Number %in% rare_variants$AID, ] -->
<!-- coldat_to_add <- coldat_to_add %>%  -->
<!--   left_join(., -->
<!--     rare_variants %>% select(AID, matches(paste(rvs, collapse = "|"))) %>% unique(),  -->
<!--     by  = join_by(A.Number == AID),  -->
<!--     relationship = "many-to-one" -->
<!--   ) -->

<!-- # add new cols to sce coldata -->
<!-- for(new_var in setdiff(names(coldat_to_add), names(colData(sce)) ) ){ -->
<!--   colData(sce)[[new_var]] <- coldat_to_add[[new_var]] -->
<!--   cat("\nadded",new_var,"to coldata\n") -->
<!-- } -->

<!-- ``` -->


#_____________________

# Mung data
```{r look at dims}
check_dims <- function(){
  
  D <- as.data.frame(colData(sce)) %>% 
  left_join(
    as.data.frame(t(assays(sce)$scaled_measure_vars)) %>% 
      rownames_to_column("sample")
  ) 
  
  cat(paste0("\ndims : ", paste(dim(D), collapse = " "),"\n"))
  
  n_patIDs <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    pull(patID ) %>% 
    n_distinct(na.rm = TRUE)
  
  n_patID_eyes <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    select(patID, eye_num) %>% 
    n_distinct(na.rm = TRUE)
  
  n_observations <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    n_distinct(na.rm = TRUE)
  
  n_patIDs_w_gr_1_vis <- D %>% 
    select(patID, visitID, eye_num) %>% 
    drop_na() %>% 
    filter(n()>1, .by = c(patID, eye_num)) %>% 
    pull(patID) %>% 
    n_distinct(na.rm = TRUE)
  
  print(rbind(n_patIDs,n_patID_eyes,n_observations,n_patIDs_w_gr_1_vis))
  
}

check_dims()
# dims : 12525 204
#                      [,1]
# n_patIDs             3675
# n_patID_eyes         7328
# n_observations      12525
# n_patIDs_w_gr_1_vis   516
# ^ new rare variant data

# Wed Jun 11 15:34:46 2025 
# dims : 12276 254
# n_patIDs             3971
# n_patID_eyes         7888
# n_observations      12275
# n_patIDs_w_gr_1_vis   492

```

## make dm vars numeric
```{r}


# make Diabetes mellitus 1 if yes, ...
sce$DM %>% table
sce$DM_binary <- ifelse(sce$DM == "Yes", 1, 
                        ifelse(sce$DM == "No", 0, 
                               NA))

# make type I 1 and type II 2
sce$DMTYPE %>% table
sce$DMTYPE_numeric <- ifelse(sce$DMTYPE == "Type I", 1, 
                        ifelse(sce$DMTYPE == "Type II", 2, 
                               NA))

```

# Calculate vars
## calculate progression rate variables
PR = B1 from PT ~ t (months)
FA1_PR = FA1 ~ t
MT_PR = mean(parafovea) ~ t

```{r}

# for each eye, lm of PT ~ month; extract PR as B1(2nd coefficient); 
# delta PR = change in pt over change in time
PR_vars_d <-
  colData(sce) %>% 
  as.data.frame() %>% 
  select(patID, eye_num, patID_eye, sample, pseudoprog, regr_fxd_pseudoprog, month) %>% 
  # left_join(
  #   assays(sce)$pmm_imp_of_scaled %>% 
  #     t() %>% 
  #     as.data.frame() %>% 
  #     rownames_to_column("sample") %>% 
  #     select(sample, ACUIT)
  # ) 
  {}


PR <-
  PR_vars_d %>% 
  group_by(patID_eye) %>% 
  # # VA rate; handles broken models
  # mutate(
  #   ACUIT_PR = {
  #       mod <- try(lm(ACUIT ~ month))
  #       if(!inherits(mod, "try-error")) coef(mod)[2] else  NA
  #     }) %>% 
  # # parafovea rate
  # mutate(parafovea_PR = coef(lm(parafovea ~ month))[2] ) %>%
  # PT RATE
  ## lm
  mutate(lm_PR = coef(lm(pseudoprog ~ month))[2]) %>% 
  mutate(lm_regr_fxd_PR = coef(lm(regr_fxd_pseudoprog ~ month))[2]) %>% 
  ## delta BL
  mutate(delta_regr_fxd_PR = 
           (dplyr::last(regr_fxd_pseudoprog) - dplyr::first(regr_fxd_pseudoprog)) / (dplyr::last(month) - dplyr::first(month))
         ) %>% 
  mutate(delta_PR = 
           (dplyr::last(pseudoprog) - dplyr::first(pseudoprog)) / (dplyr::last(month) - dplyr::first(month))
         ) %>% 
  ungroup()


# add to colData
if(all(colData(sce)$sample == PR$sample)) colData(sce) <- colData(sce) %>% cbind(PR[, grep("PR", names(PR), value = TRUE)]) else stop("sample doesn't match")

```


## calculate pseudoprog quintile
```{r}

# first visit pseudoprog values
first_pseudotimes <- colData(sce) %>% 
  as.data.frame() %>% 
  select(patID_eye, month, pseudoprog) %>% 
  arrange(month) %>% 
  group_by(patID_eye) %>% 
  filter(row_number() == 1) %>% 
  pull(pseudoprog)

# quintile first visit pseudotimes
pp_quintiles <- quantile(first_pseudotimes, seq(0,1,.2), na.rm = TRUE)

# quitile eyes' first visit based on first visit quintiles
pp_quintile_d <-  colData(sce) %>% 
  as.data.frame() %>% 
  select(patID_eye, month, pseudoprog) %>% 
  group_by(patID_eye) %>% 
  filter(row_number() == 1) %>% 
  mutate(BL_pseudotime_quintile = 
           cut(pseudoprog, breaks = pp_quintiles, include.lowest = TRUE, labels = as.character(1:5) )
         ) %>% 
  unique() 

# add to sce
coldata_to_add <-  colData(sce) %>% 
  as.data.frame() %>% 
  select(patID_eye, month, pseudoprog) %>% 
  left_join(
    pp_quintile_d %>% select(patID_eye, BL_pseudotime_quintile),
    relationship  = "many-to-one")

for(new_var in setdiff(names(coldata_to_add), names(colData(sce)) ) ){
  colData(sce)[[new_var]] <- coldata_to_add[[new_var]]
}


colData(sce) %>% 
  as.data.frame() %>% 
  ggplot(aes(x = month, y = pseudoprog))+
  geom_smooth(method = "lm")+
  facet_wrap(~BL_pseudotime_quintile, scales = "free_x", ncol = 5)+
  theme_bw()

# colData(sce) %>% 
#   as.data.frame() %>%
#   group_by(patID_eye) %>% 
#   arrange(patID_eye, month) %>% 
#   mutate(delta0_pp=  (dplyr::last(pseudoprog) - dplyr::first(pseudoprog)) / (dplyr::last(month) - dplyr::first(month))) %>% 
#   ggplot(aes(x = month, y = delta0_pp))+
#   geom_smooth(method = "loess")

```

## Correct pseudoprogression regression
```{r}

# Cummax_na on psueodprog
regr_fxd_pseudoprog_d <- colData(sce) %>% 
  as.data.frame() %>% 
  arrange(month) %>% 
  group_by(patID_eye) %>% 
  mutate(regr_fxd_pseudoprog = cummax_na(pseudoprog)) %>% 
  ungroup()

if(all(regr_fxd_pseudoprog_d$regr_fxd_pseudoprog == sce$regr_fxd_pseudoprog)) cat("\nall(regr_fxd_pseudoprog$regr_fxd_pseudoprog == sce$regr_fxd_pseudoprog)\n") else warning("\nPP correction in this script does not match previous PP regression correction\n")



```



# Side analyisis
## look at genetic data availability
```{r, eval=FALSE}

# counts of patients with genetics, etc.
count_patients_in_db <- n_distinct(pids$Patient.ID)
count_patients_in_assay <- n_distinct(sce$patID)
count_patients_in_assay_w_genetics <- n_distinct(sce$patID[!is.na(sce$Normalised_PRS)])
count_patients_in_patient_allele_freqs <- n_distinct(patient_allele_freqs$IID)

genetics_availability <- cbind(
    count_patients_in_db ,
    count_patients_in_assay, 
    count_patients_in_assay_w_genetics ,
    count_patients_in_patient_allele_freqs
)

# bp of genetic data availability
par(mar = c(4,15,4,4))
bp <- barplot(genetics_availability,
              las = 2, 
              horiz = TRUE,
              names.arg = c("count patients in db", "count patients in assay", "count patients in assay w/ genetics", "count patients from genetics data"), 
              col = "darkgrey", 
              main = "genetic data availability"
              )
text(genetics_availability-1000, bp, genetics_availability)


# venn diagram of data availability
## patient IDs from db, my data, and from liams genetics joined to mine
patIDs_in_db <-pids$Patient.ID%>% unique()
patIDs_in_assay <- sce$patID %>% unique()
# patIDs_in_assay_w_genetics <- sce$patID[!is.na(sce$Normalised_PRS)]%>% unique()

## patIDs from raw genetics from liam; convert to patIDs and include those that dont match db pid key
AIDs_in_patient_allele_freqs <- patient_allele_freqs$IID%>% unique()
AIDs_in_patient_allele_freqs_m <- cbind(AIDs_in_patient_allele_freqs, patID_match = pids$Patient.ID[match( AIDs_in_patient_allele_freqs, pids$A.Number)])

### make AID where pid not match (NA)
where_no_match <- is.na(AIDs_in_patient_allele_freqs_m[, "patID_match"]) 
AIDs_in_patient_allele_freqs_m[ where_no_match, "patID_match"] <- AIDs_in_patient_allele_freqs_m[ where_no_match , "AIDs_in_patient_allele_freqs"]

### patIDs in patient Allele freq data
patIDs_in_patient_allele_freqs <- AIDs_in_patient_allele_freqs_m[,"patID_match"]

venn_l <- list(patIDs_in_db=patIDs_in_db,
               patIDs_in_assay=patIDs_in_assay,
               patIDs_in_patient_allele_freqs=patIDs_in_patient_allele_freqs
               )

venn <- VennDiagram::venn.diagram( venn_l, 
                                   filename = NULL,
                                   cat.pos = c(-10, 30, 180),
                                   main = "Genetics on patients", 
                                   cat.cex = 2, 
                                   main.cex = 2, 
                                   cex= 2
                                   )

dev.off()
png(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/genetics_patients_venn.png"), height = 800, width = 800)
grid.draw(venn)
dev.off()

# are there any patients in allele_freq df that didnt join?
colData(sce) %>% 
  as.data.frame() %>% 
  filter(A.Number %in% patient_allele_freqs$IID) %>% 
  filter(Normalised_PRS %>% is.na())
# ^ no


# <- <- <- <- <- <- <- <- <- <- <- <- 
# WHO ARE THE PATIENTS WITH GENETICS IN DB BUT NOT ASSAY ~400 ----
patIDs_in_allele_freqs_and_db <- intersect(patIDs_in_patient_allele_freqs, patIDs_in_db)
patIDs_with_genetics_in_db_but_not_assay <- setdiff(patIDs_in_allele_freqs_and_db, patIDs_in_assay)

raw <- metadata(sce)$raw

# filter raw dfs for patIDs_with_genetics_in_db_but_not_assay
raw_filt <- lapply(raw, function(df){
  x <- df %>% 
    filter(patientId %in% patIDs_with_genetics_in_db_but_not_assay) 
  return(x)
})

# count or ratio observed to nrow for each df
raw_filt_colsums_obs <- lapply(raw_filt, function(df){
  x <- df %>% 
    mutate_all(~as.numeric(!is.na(.))) 
  # return(colSums(x)/ nrow(x))
  return(colSums(x))
})

length(raw_filt_colsums_obs)

# plot barplots for each df for patIDs_with_genetics_in_db_but_not_assay ratio observed
par(mfrow = c(2,6), mar = c(4,10,1,1))
for(vec in raw_filt_colsums_obs){
  
  barplot(sort(vec), las = 2, horiz = T)
  
}
par(mfrow = c(1,1), mar = rep(1,4))

# ^ CF has lots of observed

# do these patients have phenotypes?
raw_filt$CF_FFA %>% sapply(function(x){
  
  if(n_distinct(x) < 15) table(x, useNA = "always")
  
})
# ^ most are NA or 0

# are these patients controls in allele_freqs d?
AIDs_with_genetics_in_db_but_not_assay <- 
  AIDs_in_patient_allele_freqs_m[
    AIDs_in_patient_allele_freqs_m[,"patID_match"] %in% patIDs_with_genetics_in_db_but_not_assay 
    ,
    "AIDs_in_patient_allele_freqs"
    ]

# allele data on patients with genetics who are not in assay
patient_allele_freqs[patient_allele_freqs$IID %in% AIDs_with_genetics_in_db_but_not_assay ,]$Phenotype %>% table
# ^ 383 either controls or removed for no data

# <- <- <- <- <- <- <- <- <- <- <- <- 
# WHO ARE THE PATIENTS IN DB NOT IN ASSAY MINUS THE CONTROLS ABOVE? ----
patIDs_in_db_not_assay <- setdiff(setdiff(patIDs_in_db,patIDs_with_genetics_in_db_but_not_assay ), patIDs_in_assay) 

# are all these patients in raw?
patIDs_in_db_not_assay[patIDs_in_db_not_assay%in% raw$CF_FFA$patientId] %>% length()
# ^ 930 in raw of those not in assay


# look at their data
raw_filt <- lapply(raw, function(df){
  x <- df %>% 
    filter(patientId %in% patIDs_in_db_not_assay) 
  return(x)
})

# count or ratio observed to nrow for each df
raw_filt_colsums_obs <- lapply(raw_filt, function(df){
  x <- df %>% 
    mutate_all(~as.numeric(!is.na(.))) 
  # return(colSums(x)/ nrow(x))
  return(colSums(x))
})

# plot barplots for each df for patIDs_in_db_not_assay ratio observed
par(mfrow = c(2,6), mar = c(4,10,1,1))
for(vec in raw_filt_colsums_obs){
  
  barplot(sort(vec), las = 2, horiz = T)
  
}
par(mfrow = c(1,1), mar = rep(1,4))

# what values in categorical variables
raw_filt$CF_FFA %>% sapply(function(x){
  
  if(n_distinct(x) < 15) table(x, useNA = "always")
  
})
```


## Demographics of Route
```{r, eval = FALSE}

# Boxplot of each variable by route ----
df <- colData(sce) %>% 
  as.data.frame() %>% 
  select(where(~(is.numeric(.) | is.factor(.))))

png(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/boxplots_by_pc_lda_route.png"), height= 2000, width = 4000)
par(mfrow = c(9,9), mar = c(1,1,1,1))
lapply(names(df[, -which(names(df) == "pc_lda_route")]), function(var) {
  boxplot(df[[var]] ~ df$pc_lda_route, main = var, ylab = var, xlab = "pc_lda_route")
})
par(mfrow = c(1,1))
dev.off()

# TSNE of each variable -----
## for each var in df, plot red dim tsne and make grid
plots <- list(); for(var in names(df)){
  
  plot <- plotReducedDim(sce[ , !is.na(sce[[var]]) ], "TSNE", color_by = var)
  plots[[var]] <- plot
}

length(plots)

# matrix to move plot positions in grid
layout_m <- matrix(1:length(plots), ncol = 16, nrow = 5, byrow = TRUE)

tsne_grid <- marrangeGrob(
  sapply(plots, ggplotGrob), 
  ncol = 16, 
  nrow = 5, 
  layout_matrix = layout_m
)


ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/tsnes_by_vars.png"), 
       tsne_grid,
       height = 20, 
       width = 100, 
       dpi = 300, 
       limitsize = FALSE
       )



```

## Demographics of cluster
```{r, eval = TRUE}

# Boxplot of each variable by route ----
df <- colData(sce) %>% 
  as.data.frame() %>% 
  select(where(~(is.numeric(.) | is.factor(.))))
length(names(df))

png(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/boxplots_by_cluster.png"), height= 2000, width = 4000)
par(mfrow = c(9,9), mar = rep(2,4))
lapply(names(df[, -which(names(df) == "km_nn_cluster")]), function(var) {
  boxplot(df[[var]] ~ df$km_nn_cluster, main = var, ylab = var, xlab = "km_nn_cluster")
})
par(mfrow = c(1,1))
dev.off()



```

<!-- ## Demographics of rare PHGDH -->
<!-- ```{r, eval = TRUE} -->

<!-- # Boxplot of each variable by PHGDH_RV ---- -->
<!-- png(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/boxplots_by_phgdh_rare_var.png"), height= 2000, width = 4000) -->
<!-- par(mfrow = c(9,9), mar = c(1,1,1,1)) -->
<!-- for(var in names(df[, -which(names(df) == "PHGDH_RV")]) ) { -->
<!--   if( any( !is.na(df[[var]][!is.na(df$PHGDH_RV)]) )  ){ -->
<!--     boxplot(df[[var]] ~ df$PHGDH_RV, main = var, ylab = var, xlab = "PHGDH_RV") -->
<!--   } -->
<!-- } -->
<!-- par(mfrow = c(1,1), mar = rep(4,4)) -->
<!-- dev.off() -->

<!-- # Age and height distribution difference ------ -->
<!-- age_height <- sce %>%  -->
<!--   colData() %>%  -->
<!--   as.data.frame() %>%  -->
<!--   select(patID, PHGDH_RV, HGTIN, BL_eye_age) %>%  -->
<!--   distinct() -->

<!-- age_height %>%  -->
<!--   # na.omit() %>%  -->
<!--   pivot_longer(c(HGTIN, BL_eye_age)) %>%  -->
<!--   ggplot(aes(value, color = factor(PHGDH_RV)))+ -->
<!--   geom_density()+ -->
<!--   facet_wrap(~name, ncol = 1) -->


<!-- snps <- grep("^rs|^X6|_RV", names(colData(sce)), value = TRUE) -->
<!-- ``` -->

## see how patients are related based on genetics
```{r, eval = FALSE}



# each patients unique genetics and family info
gene_relat <- sce %>% 
  colData() %>% 
  as.data.frame() %>% 
  select(patID, all_of(snps), matches("FAM")) %>% 
  filter(row_number() == 1, .by = patID) %>% 
  na.omit()
rownames(gene_relat) <- gene_relat$patID


# distance on genetics(numeric) vars
gene_dist <- dist( gene_relat[, sapply(gene_relat,is.numeric) ]) 


# Heirarchical clustering ----
hc <- hclust(gene_dist,
             method = "ward.D2"
             )

library(dendextend)

# Convert to a dendrogram object
dend <- as.dendrogram(hc)

# filter for very similar leaves
## select labs where < threshold 
prune_threshold <- 1000
leaf_labs <- labels(dend)
merge_heights <- hc$height
close_sibling_leaves <- leaf_labs[which(merge_heights<prune_threshold)]
far_sibling_leaves <- leaf_labs[which(!leaf_labs %in% close_sibling_leaves )]

# prune dend
pruned_dend <- dendextend::prune(dend, leaves = far_sibling_leaves)


# # label leaves by family id
# labels(pruned_dend) <- gene_relat$FAMILY_ID[match(labels(pruned_dend), gene_relat$patID)]


# # look at name distributions
# table(labels(pruned_dend) %>% 
#         str_split(., "-") %>%
#         # unlist()) %>% 
#         sapply(., `[`, 2)
#       ) %>% 
#   sort() %>% 
#   rev() 
#   # barplot(las = 2, cex.names = .3)

# color dend
dend_colored <- color_branches(pruned_dend, k = length(labels(pruned_dend)))

# Plot the colored dendrogram
par(mar = c(8, 2,2,2))
plot(dend_colored )


# # MDS on distance ----
gene_mds <- cmdscale(gene_dist)

i = 1

(var <- names(gene_relat)[i])
cols <- setNames(viridis(n_distinct(gene_relat[[var]])), unique(gene_relat[[var]]))
colors <- grab_matches(gene_relat[[var]], cols)
plot(gene_mds, col = colors)
i = i+1



```

<!-- ## PCA on genetics -->
<!-- ```{r, eval=FALSE} -->

<!-- # all data on eyes -->
<!-- data <- assays(sce)$pmm_imp_of_scaled %>%  -->
<!--   t() %>%  -->
<!--   as.data.frame() %>%  -->
<!--   rownames_to_column("sample") %>%  -->
<!--   left_join( -->
<!--     colData(sce) %>% as.data.frame() -->
<!--   ) -->

<!-- snps <- grep("^rs|^X6", names(data), value = TRUE) -->

<!-- # grab genetics and filter for unique rows of patID and snps -->
<!-- genetics <- data %>%  -->
<!--   distinct(across(all_of(c("patID", snps))), .keep_all = T)  -->
<!-- rownames(genetics) <- genetics$patID -->

<!-- #PCA on genetics -->
<!-- gene_pca <- prcomp(genetics[, snps] %>% na.omit) -->
<!-- gene_pcs <- gene_pca$x %>%  -->
<!--   as.data.frame() -->
<!-- gene_pcs$patID <- rownames(gene_pcs) -->
<!-- rot <- gene_pca$rotation[, 1:2] -->

<!-- ## cummulative percent var explained -->
<!-- perc_var <- ((gene_pca$sdev**2 / sum(gene_pca$sdev**2)) * 100)  -->
<!-- cum_perc_var <- ((gene_pca$sdev**2 / sum(gene_pca$sdev**2)) * 100) %>% cumsum() -->
<!-- plot(cum_perc_var, main = "cumulative percent variation explained") -->


<!-- par(mar = c(2,20,2,2)) -->
<!-- rot[,1] %>% sort() %>% barplot(., las = 2, horiz = T, main = "SNP rotations on PC1") -->

<!-- # # tnse on pcs -->
<!-- # set.seed(3399) -->
<!-- # gene_tsne <- Rtsne::Rtsne(gene_pcs, pca = FALSE) -->
<!-- # gene_tsnes <- gene_tsne$Y %>%  -->
<!-- #   as.data.frame() -->
<!-- # names(gene_tsnes) <- c("TSNE1", "TSNE2") -->
<!-- # gene_tsnes$patID <- gene_pcs$patID -->

<!-- # join pcs and tsnes to genetics -->
<!-- genetics <- genetics %>%  -->
<!--   # left_join(gene_pcs %>% select(patID,PC1, PC2)) %>%  -->
<!--   left_join(gene_pcs )  -->

<!-- # plot pcs -->
<!-- plot(gene_pcs[,1:2], col = "black");abline(v= 0, lty = 3); abline(a= 0, b = 0, lty = 3) -->

<!-- # plot loadings and  points -->
<!-- plot(gene_pcs[,1:2], col = "grey");abline(v= 0, lty = 3); abline(a= 0, b = 0, lty = 3) -->
<!-- # plot(c(-1,1), c(-1,1), type = "n") -->
<!-- # points(gene_pcs[,1:2]) -->
<!-- # points(rot, pch = 2, col = "red");abline(v= 0, lty = 3); abline(a= 0, b = 0, lty = 3) -->
<!-- text(rot, labels = rownames(rot) , col = "red");abline(v= 0, lty = 3); abline(a= 0, b = 0, lty = 3) -->
<!-- apply(rot, 1, function(row){lines(c(0, row[1]), c(0, row[2]), col = "darkblue")}) -->

<!-- # correlate pcs to vars -->
<!-- cor_m <- cor(  -->
<!--   genetics %>% select(where(~is.numeric(.))) %>% select(where(~sd(., na.rm = TRUE) >0)), -->
<!--   use = "pairwise.complete.obs" -->
<!--   ) -->

<!-- # hm of corealation on first 3 pcs -->
<!-- Heatmap(cor_m[ c("PC1", "PC2", "PC3"), ],  -->
<!--         cluster_columns = FALSE,  -->
<!--         cluster_rows = FALSE) -->


<!-- # plot pcs colored by various variables -->
<!-- PC_and_var <-  gene_pca$rotation %>%  -->
<!--   apply(2, function(col){ -->
<!--     rownames(gene_pca$rotation)[col == max(col)] -->
<!--   })  -->

<!-- # # for each var that has effect on rotation and pc pair plot pcs -->
<!-- genetics_plots <- list(); for(i in 1:(length(PC_and_var)-1) ) { -->

<!--   # grab which pcs and which var to colour by -->
<!--   pcs <- names(PC_and_var)[i:(i+1)] -->
<!--   # pcs <- c("PC1", "PC2") -->
<!--   # var <- PC_and_var[[i]] -->
<!--   # var = "rs532303_A_1_PHGDHHMGCS2" -->
<!--   var = "PHGDH_RV" -->

<!--   sub_genetics <- genetics %>% -->
<!--     select(all_of(c(pcs, var))) %>%  -->
<!--     filter( !is.na(get(var))) -->

<!--   names(sub_genetics)[which(names(sub_genetics) == var)] <- "variable" -->

<!--   plot <- ggplot(sub_genetics, aes(!!sym(pcs[1]), !!sym(pcs[2])))+ -->
<!--     ggtitle(paste("PCA on SNPs, colored by", var))+ -->
<!--     xlab( paste0(pcs[1], " ", round(perc_var[i],2), "%"))+ -->
<!--     ylab( paste0(pcs[2], " ", round(perc_var[i+1],2), "%"))+ -->
<!--     geom_point(aes(color = variable) )+ -->
<!--     scale_color_viridis_c() -->

<!--   genetics_plots[[i]] <- plot -->

<!-- } -->

<!-- i = 1 -->
<!-- genetics_plots[i]; i = i+1 -->

<!-- # look at pcs for where has rare variants ---- -->

<!-- ## PCs ~ phgdh rare var -->
<!-- phgdh_rare_var_mods <- lapply(grep("PC\\d", names(genetics), value = TRUE), function(pc){ -->

<!--   formula <- paste0(pc, " ~ PHGDH_RV") %>% as.formula() -->
<!--   lm(formula, data = genetics) -->

<!-- }) %>%  -->
<!--   lapply(tidy) -->

<!-- # make df of results -->
<!-- phgdh_summary <- do.call(rbind, phgdh_rare_var_mods) %>% as.data.frame() -->

<!-- # add pc variable -->
<!-- phgdh_summary$response <- rep(grep("PC\\d", names(genetics), value = TRUE), each = 2) -->

<!-- ## PCs ~ sptlc1 rare varirant -->
<!-- sptlc1_rare_var_mods <- lapply(grep("PC\\d", names(genetics), value = TRUE), function(pc){ -->

<!--   formula <- paste0(pc, " ~ SPTLC1_RV") %>% as.formula() -->
<!--   lm(formula, data = genetics) -->

<!-- }) %>%  -->
<!--   lapply(tidy) -->

<!-- sptlc1_summary <- do.call(rbind, sptlc1_rare_var_mods) %>% as.data.frame() -->
<!-- sptlc1_summary$response <- rep(grep("PC\\d", names(genetics), value = TRUE), each = 2) -->

<!-- rare_variants_results <- rbind(phgdh_summary,sptlc1_summary) -->

<!-- rare_variants_results %>%  -->
<!--   ggplot(aes(estimate, -log10(p.value), color = response))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   facet_free(~term)+ -->
<!--   geom_hline(yintercept =-log10(.05), linetype = 3)+ geom_vline(xintercept = 0, linetype =3)+ -->
<!--   geom_label_repel(aes(label = ifelse(p.value <.05, response, response)), show.legend = FALSE) -->

<!-- # look at rare variant division in most significant pcs ---- -->
<!-- # which_pcs <- c("PC1", "PC6") -->
<!-- # which_pcs <- c("PC1", "PC6") -->
<!-- which_pcs <- c("PC6", "PC7") -->
<!-- # which_pcs <- c("PC5", "PC6") -->
<!-- genetics %>%  -->
<!--   select(matches("has"), !!sym(which_pcs[1]), !!sym(which_pcs[2])) %>%  -->
<!--   pivot_longer(matches("has"), names_to = "variant", values_to = "value" ) %>%  -->
<!--   filter(!is.na(value)) %>%  -->
<!--   ggplot(aes(!!sym(which_pcs[1]),!!sym(which_pcs[2]), color = value))+ -->
<!--   geom_point()+ -->
<!--   facet_wrap(~variant)+ -->
<!--   scale_color_viridis_c()+ -->
<!--   theme_dark() -->

<!-- ## SNPS ~ phgdh rare var -->
<!-- ### model each snp by has rare phgdh using multinomial reg -->
<!-- snps_phgdh_RV <- lapply(snps, function(snp){ -->

<!--   a.formula <- paste0(snp, " ~ PHGDH_RV") %>% as.formula() -->

<!--   lm(a.formula, data = genetics) -->

<!-- }) %>%  -->
<!--   lapply(tidy) -->

<!-- names(snps_phgdh_RV)<- grep("^rs|^X6", names(genetics), value = TRUE) -->

<!-- snps_phgdh_RV_results <- do.call(rbind,snps_phgdh_RV) -->
<!-- snps_phgdh_RV_results$response <- rep(grep("^rs|^X6", names(genetics), value = TRUE), each = 2) -->

<!-- # volcano of how phgdh correlates to each snp -->
<!-- snps_phgdh_RV_results %>%  -->
<!--   ggplot(aes(estimate, -log10(p.value), color = response))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   facet_free(~term)+ -->
<!--   geom_hline(yintercept =-log10(.05), linetype = 3)+ geom_vline(xintercept = 0, linetype =3)+ -->
<!--   geom_label_repel(aes(label = ifelse(p.value <.05, response, response)), show.legend = FALSE)+ -->
<!--   theme_bw() -->

<!-- ``` -->




#~~~~~~~~~~~~~~~
# SEE if proband relations follow same routes significantly
## see how family routes agree-> consensus
```{r, eval = FALSE}

mod_d <- colData(sce) %>% 
  as.data.frame() %>% 
  select(patID_eye, FAMILY_ID, pc_lda_route, RELATPROBND) %>% 
  filter(n()>1, .by = patID_eye) %>% # filter for longitudinal eyes (these were actually classified)
  mutate(pc_lda_route = factor(pc_lda_route), 
         FAMILY_ID = factor(gsub("-", "_",FAMILY_ID)),
         RELATPROBND = factor(RELATPROBND)
  ) %>% 
  mutate(present = 1) %>% 
  distinct() %>%
  # drop_na() %>%
  # pivot_wider(names_from = FAMILY_ID,
  #             values_from = present,
  #             values_fill = list(present = 0),
  #             names_prefix = "family_"
  #             ) %>%
  select(matches("pc_lda|family")) %>%
  # mutate(across(-pc_lda_route, ~as.numeric(scale(.)))) %>%
  ungroup() 


# each family's route distributions
family_routes <- sapply(unique(mod_d$FAMILY_ID), function(family){
  table(mod_d$pc_lda_route[mod_d$FAMILY_ID == family]) 
}) %>% 
  t()
rownames(family_routes) <- unique(mod_d$FAMILY_ID)

# remove X
family_routes<- family_routes[, -which(colnames(family_routes) == "x")]

# add A, C; B, D
simplified_family_routes <- family_routes[, c("a", "b")] 

# filtler for non-zero routes
simplified_family_routes <- simplified_family_routes[rowSums(simplified_family_routes)>0,]

# normalize by rowSums
simplified_family_routes <- sweep(simplified_family_routes, 1, rowSums(simplified_family_routes), "/")

# divide each row by the number of cells filled
penalized_consensus <- family_routes %>% 
  sweep(1,rowSums(family_routes!=0),"/") %>% 
  # round(1) %>% 
  rowSums() 
penalized_consensus[is.nan(penalized_consensus)] <- 0

# each family's total number of members
ideal_consensus <- family_routes %>% 
  rowSums()

# par(mfrow = c(2,1))
# penalized_consensus %>% hist(main = "penalized consensus", xlim = c(0, 8))
# ideal_consensus %>% hist(main = "ideal consensus", xlim = c(0, 8))
# par(mfrow = c(1,1))

# densities
penalized_consensus_dens <- penalized_consensus %>%
  density()
ideal_consensus_dens <- ideal_consensus %>% 
  density()


# par(mfrow = c(2,1))
# plot consensus densities 
x_lim <- c( min(min(ideal_consensus_dens$x), min(penalized_consensus_dens$x)),  max(max(ideal_consensus_dens$x), max(penalized_consensus_dens$x)))
y_lim <- c( min(min(ideal_consensus_dens$y), min(penalized_consensus_dens$y)),  max(max(ideal_consensus_dens$y), max(penalized_consensus_dens$y)))
par(mar = c(5,4,4,2))
plot(penalized_consensus_dens,
     col = "blue", 
     main = "Ideal consensus versus penalized consensus\n(for each family, row/#routes taken)", 
     xlim = x_lim,
     ylim = y_lim
     )
lines(ideal_consensus_dens)
legend("topright",
       fill = c("black","blue"), 
       legend = c("ideal" , "penalized")
       )

# STATISTICAL TESTS ----
# is the observed pc_lda_route agreement similar to ideal family route classification?
t.test(penalized_consensus,ideal_consensus)
# ^ significant difference between concensus means (but this assumes normality)

# ks test to see if distributions of agreements same 
ks.test(penalized_consensus ,
        ideal_consensus)
# ^ significant difference between these distributions 

# MONTE CARLO PERMUTATION TEST ----
## simulate the ideal distribution and see
# permutation test to see if difference between agreements means is significantly different from random
observed_diff <- mean(ideal_consensus) - mean(penalized_consensus)

# make vector of all agreements to sample as all known values of consensus
combined <- c(ideal_consensus, penalized_consensus)
# length to sample
n1 <- length(ideal_consensus)

# sample consesus values and find difference in means n times
perm_diffs <- replicate(1e4, {
  perm <- sample(combined)
  mean(perm[1:n1]) - mean(perm[-(1:n1)])
})

# H₀: There is no true difference between the means of ideal_consensus and penalized_consensus; the observed difference is due to random chance.
# H₁: There is a significant difference between the means of ideal_consensus and penalized_consensus.
p_value <- mean(abs(perm_diffs) >= abs(observed_diff) )
# ^ significant difference between observed and permuted differences

perm_diffs_dens <- perm_diffs %>% 
  density() 

plot(perm_diffs_dens, 
     xlim = c(min(observed_diff,min(perm_diffs)),max(observed_diff,max(perm_diffs))  ), 
     main = "permuted differences between ideal consensus and penalized consensus\nversus actual difference between ideal and penalized" )
abline(v = observed_diff,
       col = "orange", 
       lty= 3, 
       lwd= 5)

text(observed_diff - (observed_diff/8 ),
     max(perm_diffs_dens$y)/2, 
     paste("observed_diff =", round(observed_diff,1), "\np = ",p_value))
# ^^ observed difference in agreement is significantly greater than random
# THEREFORE observed concensus is does not align with ideal; eyes do not follow family members to a significant degree

par(mfrow = c(1,1))



```

<!-- ## Use XGboost to see if I can predict pc_lda_route from familyID -->
<!-- ```{r} -->

<!-- library(xgboost) -->

<!-- truth <- colData(sce)[,c("pc_lda_route", "FAMILY_ID")] %>%  -->
<!--   as.data.frame() %>%  -->
<!--   drop_na() %>%  -->
<!--   distinct() -->

<!-- inds <- sample(1:nrow(truth),.7*nrow(truth) ) -->
<!-- train <- truth[inds , ] %>% -->
<!--   mutate_all(~as.numeric(factor(.))-1) %>%  -->
<!--   as.matrix() -->
<!-- test <- truth[-inds , ] %>%  -->
<!--   mutate_all(~as.numeric(factor(.))-1) %>%  -->
<!--   as.matrix() -->


<!-- # Define parameters -->
<!-- params <- list( -->
<!--   objective = "multi:softmax",  # For classification -->
<!--   num_class = n_distinct(train[,"pc_lda_route"]), # Number of classes -->
<!--   eval_metric = "mlogloss"     # Multiclass log-loss -->
<!-- ) -->

<!-- xgb <- xgboost(as.matrix(train[, "FAMILY_ID"]), -->
<!--                label = train[,"pc_lda_route"], -->
<!--                max.depth = 5, -->
<!--                eta = 1, -->
<!--                nthread = 2, -->
<!--                nrounds = 2, -->
<!--                params = params) -->


<!-- # accuracy -->
<!-- pred <- predict(xgb, test[, "FAMILY_ID"]) -->
<!-- confusion <- table(predicted = round(pred,0), actual = test[,"pc_lda_route"]) -->
<!-- confusion -->
<!-- sum(diag(confusion))/ sum(confusion) -->

<!-- ``` -->


#~~~~~~~~~~~~~~~



#~~~~~~~~~~~~~~~
# Model Progression rate

###~snps


#### filter data 
```{r}

#___________
# MAKE MODEL DATA
#___________

# terms to model 
all_predictors <- names(colData(sce)) %>% grep("^rs|X6|chr\\d|PRS|_RV", ., value = TRUE)
all_covariates <-  c("patID", "patID_eye", "month", "BL_pseudotime_quintile", "SEXF", "AGE", 't2d', 'BMI',paste0("PC", 1:8))

# df to model with, selecting covs, responses, and predictors; filtering for each eye's first visit; and scaling predictors after
pp_pr_snp_mod_d <- colData(sce) %>% 
  as.data.frame() %>% 
  select(pseudoprog,chew  = chew_grade_derived_mine_loose, all_of(c( all_covariates, all_predictors))) %>% 
  group_by(patID_eye) %>% 
  arrange(patID_eye, month) %>% 
  ungroup() 


# calculate  rate
pp_pr_snp_mod_d <- pp_pr_snp_mod_d %>% 
  mutate(year = month/12) %>% 
  group_by(patID_eye) %>% 
  # mutate(pp_rate = (dplyr::last(pseudoprog)-dplyr::first(pseudoprog))  /  (dplyr::last(year)-dplyr::first(year)))
  mutate(pp_rate =
          (dplyr::last(pseudoprog[!is.na(pseudoprog)]) - dplyr::first(pseudoprog[!is.na(pseudoprog)]))
              /
          (dplyr::last(year[!is.na(pseudoprog)]) - dplyr::first(year[!is.na(pseudoprog)]))
           ) %>%
  mutate(pp_rate = ifelse(is.infinite(pp_rate), NA, pp_rate)) %>% 
  ungroup()


# Scale
pp_pr_snp_mod_d$SEXF <- pp_pr_snp_mod_d$SEXF*1
pp_pr_snp_mod_d_scaled <- pp_pr_snp_mod_d %>% 
  mutate(across(matches("PC|AGE|t2d|BMI|SEX"), 
                ~ as.numeric(scale(.))
                ))

# seelect first observation; select eyes with first obs in early cluster; filter for a or b route; 
pp_pr_snp_mod_d_scaled_filtered <- pp_pr_snp_mod_d_scaled %>% 
  arrange(patID_eye,month) %>% 
  group_by(patID_eye) %>% 
  dplyr::slice(1) %>% 
  
  # # filter for one route patients
  # ungroup() %>% group_by(patID) %>% filter(n_distinct(pc_lda_route)==1) %>% select(-patID_eye) %>% distinct() %>% 
  
  ungroup() 

nrow(pp_pr_snp_mod_d_scaled_filtered)
# 4437

pp_pr_snp_mod_d_scaled_filtered$patID_eye %>% table %>% table
#    1 
# 4437

pp_pr_snp_mod_d_scaled_filtered %>% 
  {colSums(!is.na(.))}

pp_pr_snp_mod_d_scaled_filtered %>% 
  filter(if_all(c(rs532303_phgdh_hmgcs2, AGE, SEXF,pp_rate, PC1), ~!is.na(.))) %>% 
  data.frame() 


```

#### model
```{r}

pc_str <- paste(paste0("PC", 1:8), collapse = " + ")
forms <- sprintf("pp_rate ~ %s + SEXF + AGE + (1|patID) + %s", all_predictors,pc_str) %>% sapply(as.formula)

mod_list <- sapply(forms, function(form){
  mod <- lmer(form, data = pp_pr_snp_mod_d_scaled_filtered)
  return(mod)
}, simplify = FALSE)
summ_list <- lapply(mod_list, summary)

result <- extract_results(summ_list) %>% janitor::clean_names()
result$q_value <- p.adjust(result$pr_t, "BH")

result %>% 
  left_join(metadata(sce)$genetics$snp_info, by = join_by(term ==label)) %>% 
  filter(pr_t>0,pr_t<.9) %>%
  filter(!grepl("Intercept|SEX|AGE|PC", term)) %>% 
  ggplot(aes(estimate, -log10(q_value), color = simple_label))+
  # ggplot(aes(Estimate, -log10(p), color = term))+
    geom_hline(yintercept = -log10(.05), linetype = 2, color = "grey")+
    geom_vline(xintercept = 0, linetype = 2, color = "grey")+
    geom_vline(xintercept = c(-.1, .1), linetype = 3, color = "grey")+
    geom_point(show.legend = FALSE)+
     geom_label_repel(aes(label = ifelse( q_value < .05,  simple_label, NA)),
                         size = 4,
                         # nudge_y = 0.5,
                          # nudge_x = -.5,
                         show.legend = FALSE, 
                         max.overlaps = 20)+
    ggtitle("Pseudoprog Rate ~ SNP + SEXF + AGE + (1|patID)")+
    ylab("-log10(q.value)")+
    theme_bw()+
    theme(aspect.ratio = 1,
          panel.grid = element_blank(),
          panel.background = element_rect(fill = "white", color = NA),
          plot.background = element_rect(fill = "transparent", color = NA),
          legend.background = element_rect(fill = "transparent", color = NA),
          legend.box.background = element_rect(fill = "transparent", color = NA))+
    theme(aspect.ratio = 1,
          axis.title.x = element_text(),
          axis.title.y = element_text(angle = 90))
ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/pp_pr_snp_volcano.pdf"), height = 5, width = 5)


sprintf("cp %s %s",
  paste0(wd_path, "results/clustering/14_geneticProgressionModeling/pp_pr_snp_volcano.pdf"),
  paste0(wd_path, "results/poster_plots/pp_pr_snp_volcano.pdf")
         ) %>% 
  system()


write_csv(result, sprintf("%sresults/publication_tables/%s",wd_path, "PP_PR_SNP.csv"))


result %>% 
  left_join(metadata(sce)$genetics$snp_info, by = join_by(term ==label)) %>% 
  filter(!grepl("Intercept|SEX|AGE|PC", term)) %>% 
  arrange(q_value) %>% 
  select(term,simple_label,estimate, pr_t, q_value) %>% 
  filter(pr_t< 0.05) 
#                                                                       term simple_label   estimate       pr_t   q_value
# 1 rs6955423_sept14_mrps17_znf713_gbas_psph_cct6a_sumf2_phkg1_chchd2_nupr1l         PSPH -0.1416031 0.04195458 0.2517275

```



###~covariates
#### model
```{r, eval = FALSE}

f <- function(x){ifelse(x == min(x, na.rm = TRUE), 0, ifelse(x == max(x, na.rm = TRUE), 1, x)) %>% factor}

pp_pr_cov_mod <- lmer(pp_rate ~ AGE + SEXF + (1|patID) ,
                      data = pp_pr_snp_mod_d_scaled_filtered %>% mutate(SEXF = SEXF %>% f())
                      )
(summ <- summary(pp_pr_cov_mod))
effects::allEffects(pp_pr_cov_mod) %>% plot()
result <- data.frame(coef(summ))
result <- janitor::clean_names(result)
result$term <- rownames(result)

result %>% 
  filter(pr_t>0,pr_t<.9) %>%
  ggplot(aes(estimate, -log10(pr_t), color = term))+
  # ggplot(aes(Estimate, -log10(p), color = term))+
    geom_hline(yintercept = -log10(.05), linetype = 2, color = "grey")+
    geom_vline(xintercept = 0, linetype = 2, color = "grey")+
    geom_vline(xintercept = c(-.1, .1), linetype = 3, color = "grey")+
    geom_point(show.legend = FALSE)+
     geom_label_repel(aes(label = ifelse( pr_t < .05,  term, NA)),
                         size = 4,
                         # nudge_y = 0.5,  
                         show.legend = FALSE, 
                         max.overlaps = 20)+
    ggtitle(paste0("Pseudoprog Grade Rate ~ AGE + SEXF + (1|patID)"), "For eyes with BL Chew <= 2")+
    ylab("-log10(p.value)")+
    theme_bw()+
    theme(aspect.ratio = 1,
          panel.grid = element_blank(),
          panel.background = element_rect(fill = "white", color = NA),
          plot.background = element_rect(fill = "transparent", color = NA),
          legend.background = element_rect(fill = "transparent", color = NA),
          legend.box.background = element_rect(fill = "transparent", color = NA))+
    theme(aspect.ratio = 1,
          axis.title.x = element_text(),
          axis.title.y = element_text(angle = 90))
ggsave(paste0(wd_path, "results/poster_plots/pp_pr_snp_cov_volcano.pdf"), height = 5, width = 5)

```


<!-- ## PR ~ Snps + BL_pp_quantile -->

<!-- ### using delta PR -->
<!-- ```{r} -->

<!-- hgt = 6; wdth = 10 -->

<!-- # ORIGINAL DELTA PR ---- -->
<!-- # terms for this model formula -->
<!-- response <-  "delta_PR" -->
<!-- predictors <- all_predictors %>% -->
<!--   grep("PRS", ., value = TRUE, invert = TRUE) -->
<!-- covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile", "BL_eye_age") -->
<!-- # covariates <- c("BL_pseudotime_quintile") -->

<!-- # for each predictors, try model response and save model and summary -->
<!-- delta_PR_snp_mod_list <- list(); delta_PR_snp_summ_list <- list(); for(a_predictor in predictors){ -->
<!--   a_form <- make.formula(response, a_predictor, covariates) -->
<!--   cat(paste("\n", a_form,"\n")) -->
<!--   mod_name <- paste0(response, " ~ ", a_predictor) -->
<!--    a_mod <- try({ -->
<!--           lmer(a_form, data = PR_mod_d) -->
<!--           # lm(a_form, data = PR_mod_d) -->
<!--         }) -->
<!--    if(inherits(a_mod, "try-error")){ -->
<!--      cat(paste0("\n\nFailure\n",response," ~ ", a_predictor, "\n")) -->
<!--      }else{ -->
<!--       cat(paste0("\n",a_form, "\n")) -->
<!--       delta_PR_snp_mod_list[[mod_name]] <- a_mod -->
<!--       delta_PR_snp_summ_list[[mod_name]] <- summary(a_mod) -->
<!--      } -->
<!-- } -->

<!-- # see if any models failed -->
<!-- all_mods <- paste("delta_PR", predictors,sep = " ~ ") -->
<!-- setdiff(all_mods, names(delta_PR_snp_mod_list)) -->
<!-- (msgs <- sapply(delta_PR_snp_summ_list, function(summ) summ$fitMsgs))[sapply(msgs, length)>0] -->
<!-- # ^ RV models rank deficient so dropped from modeling -->
<!-- # ^^ RV vars must be linearlly dependent on anaother var -->

<!-- # # loook for colinearity, near zero singular values -->
<!-- # MM <- PR_mod_d %>%  -->
<!-- #   select(delta_PR, PHGDH_RV, SEXF, BL_pseudotime_quintile, BL_eye_age) %>%  -->
<!-- #   model.matrix(~PHGDH_RV+SEXF+BL_pseudotime_quintile+BL_eye_age, data = .) -->
<!-- # MM_svd <- svd(MM) -->
<!-- #  -->
<!-- # # Identify rank-deficient singular values -->
<!-- # threshold <- 1e-8 -->
<!-- # is_deficient <- MM_svd$d < threshold -->
<!-- #  -->
<!-- # plot(MM_svd$v, type = "n") -->
<!-- # text(MM_svd$v, labels = colnames(MM)) -->
<!-- #  -->
<!-- # # Extract dependent variables from V -->
<!-- # deficient_variables <- MM_svd$v[, is_deficient, drop = FALSE] -->
<!-- #  -->
<!-- # # Map to variable names -->
<!-- # dependent_vars <- colnames(MM)[apply(deficient_variables, 1, function(x) any(abs(x) > 1e-6))] -->
<!-- # print(dependent_vars) -->
<!-- #  -->
<!-- # # MM_pc <- prcomp(MM, scale = TRUE) -->
<!-- # #  -->
<!-- # # bins <- cut(MM[, "BL_eye_age"], 10) -->
<!-- # # colors <- grab_matches(bins, setNames(viridis(10),unique(bins))) -->
<!-- # #  -->
<!-- # # plot(MM_pc$x, -->
<!-- # #      # pch = ifelse(MM[,"PHGDH_RV"]==0, 1, 6), -->
<!-- # #      pch = 16, -->
<!-- # #      cex = ifelse(MM[,"PHGDH_RV"]==0, .8, 2), -->
<!-- # #      col = colors -->
<!-- # #      ) -->

<!-- # extract and filter results -->
<!-- delta_PR_snp_results <- extract_results(delta_PR_snp_summ_list) -->
<!-- delta_PR_snp_results <- delta_PR_snp_results %>% -->
<!--   filter(!Term %in% grep("Intercept|BL_|SEXF|age", delta_PR_snp_results$Term, value = TRUE)) -->

<!-- # plot results -->
<!-- delta_PR_snp_results_plot <- delta_PR_snp_results %>% -->
<!--   ggplot(aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Term))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   geom_hline(yintercept = -log10(.05), linetype = 3)+  geom_vline(xintercept = 0, linetype = 3)+ -->
<!--   # geom_label_repel(aes(label = ifelse(`Pr(>|t|)` < .05, Term, NA)), show.legend = FALSE, max.overlaps = 100)+ -->
<!--   geom_label_repel(aes(label = Term), show.legend = FALSE, max.overlaps = 100, nudge_y=.05)+ -->
<!--   ggtitle(paste("Delta from BL progression rate ~ a_snp ", paste(covariates, collapse = " + "), sep = "+"))+ -->
<!--   theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/delta_PR_snp_results.png"),delta_PR_snp_results_plot, height = hgt, width = wdth) -->


<!-- # REGRESSION FIXED DELTA PR ---- -->
<!-- # terms for this model formula -->
<!-- response <-  "delta_regr_fxd_PR" -->
<!-- predictors <- all_predictors %>% -->
<!--   grep("PRS", ., value = TRUE, invert = TRUE) -->
<!-- covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile", "BL_eye_age") -->
<!-- # covariates <- c("BL_pseudotime_quintile") -->


<!-- # for each predictors, try model response and save model and summary -->
<!-- delta_regr_fxd_PR_snp_mod_list <- list(); delta_regr_fxd_PR_snp_summ_list <- list(); for(a_predictor in predictors){ -->
<!--   a_form <- make.formula(response, a_predictor, covariates) -->
<!--   cat(paste("\n", a_form,"\n")) -->
<!--   mod_name <- paste0(response, " ~ ", a_predictor) -->
<!--    a_mod <- try({ -->
<!--           lmer(a_form, data = PR_mod_d) -->
<!--           # lm(a_form, data = PR_mod_d) -->
<!--         }) -->
<!--    if(inherits(a_mod, "try-error")){ -->
<!--      cat(paste0("\n\nFailure\n",response," ~ ", a_predictor, "\n")) -->
<!--      }else{ -->
<!--       cat(paste0("\n",a_form, "\n")) -->
<!--       delta_regr_fxd_PR_snp_mod_list[[mod_name]] <- a_mod -->
<!--       delta_regr_fxd_PR_snp_summ_list[[mod_name]] <- summary(a_mod) -->
<!--      } -->
<!-- } -->

<!-- # see if any models failed -->
<!-- all_mods <- paste("delta_regr_fxd_PR", predictors,sep = " ~ ") -->
<!-- setdiff(all_mods, names(delta_regr_fxd_PR_snp_mod_list)) -->

<!-- # extract and filter results -->
<!-- delta_regr_fxd_PR_snp_results <- extract_results(delta_regr_fxd_PR_snp_summ_list) -->
<!-- delta_regr_fxd_PR_snp_results <- delta_regr_fxd_PR_snp_results %>% -->
<!--   filter(!Term %in% grep("Intercept|BL_|SEXF|age", delta_regr_fxd_PR_snp_results$Term, value = TRUE)) -->

<!-- # plot results -->
<!-- delta_regr_fxd_PR_snp_results_plot <- delta_regr_fxd_PR_snp_results %>% -->
<!--   ggplot(aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Term))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   geom_hline(yintercept = -log10(.05), linetype = 3)+  geom_vline(xintercept = 0, linetype = 3)+ -->
<!--   # geom_label_repel(aes(label = ifelse(`Pr(>|t|)` < .05, Term, NA)), show.legend = FALSE, max.overlaps = 100)+ -->
<!--   geom_label_repel(aes(label = Term), show.legend = FALSE, max.overlaps = 100, nudge_y=.05)+ -->
<!--   ggtitle(paste("Delta from BL progression rate (regression corrected) ~ a_snp ", paste(covariates, collapse = " + "), sep = "+"))+ -->
<!--   theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/delta_regr_fxd_PR_snp_results.png"),delta_regr_fxd_PR_snp_results_plot, height = hgt, width = wdth) -->

<!-- ``` -->

<!-- ### using lm PR -->
<!-- ```{r} -->

<!-- # ORIGINAL LM PR ---- -->
<!-- # terms for this model formula -->
<!-- response <-  "lm_PR" -->
<!-- predictors <- all_predictors %>% -->
<!--   grep("PRS", ., value = TRUE, invert = TRUE) -->
<!-- # covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile") -->
<!-- covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile", "BL_eye_age") -->
<!-- # covariates <- c("BL_pseudotime_quintile") -->


<!-- # for each predictors, try model response and save model and summary -->
<!-- lm_PR_snp_mod_list <- list(); lm_PR_snp_summ_list <- list(); for(a_predictor in predictors){ -->

<!--   a_form <- make.formula(response, a_predictor, covariates) -->
<!--   cat(paste("\n", a_form,"\n")) -->

<!--   mod_name <- paste0(response, " ~ ", a_predictor) -->

<!--    a_mod <- try({ -->
<!--           lmer(a_form, data = PR_mod_d) -->
<!--           # lm(a_form, data = PR_mod_d) -->
<!--         }) -->
<!--    if(inherits(a_mod, "try-error")){ -->
<!--      cat(paste0("\n\nFailure\n",response," ~ ", a_predictor, "\n")) -->
<!--      }else{ -->
<!--       cat(paste0("\n",a_form, "\n")) -->
<!--       lm_PR_snp_mod_list[[mod_name]] <- a_mod -->
<!--       lm_PR_snp_summ_list[[mod_name]] <- summary(a_mod) -->
<!--      } -->

<!-- } -->

<!-- # see if any models failed -->
<!-- all_mods <- paste("lm_PR", predictors,sep = " ~ ") -->
<!-- setdiff(all_mods, names(lm_PR_snp_mod_list)) -->
<!-- (sapply(lm_PR_snp_summ_list, function(summ) summ$fitMsgs))[sapply(msgs, length)>0] -->


<!-- # extract and filter results -->
<!-- lm_PR_snp_results <- extract_results(lm_PR_snp_summ_list) -->
<!-- lm_PR_snp_results <- lm_PR_snp_results %>% -->
<!--   filter(!Term %in% grep("Intercept|BL_|SEXF|age", lm_PR_snp_results$Term, value = TRUE)) -->

<!-- # plot results -->
<!-- lm_PR_snp_results_plot <- lm_PR_snp_results %>% -->
<!--   ggplot(aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   geom_hline(yintercept = -log10(.05), linetype = 3)+  geom_vline(xintercept = 0, linetype = 3)+ -->
<!--   # geom_label_repel(aes(label = ifelse(q_value < .05, Term, NA)), show.legend = FALSE, max.overlaps = 100)+ -->
<!--   geom_label_repel(aes(label = Term), show.legend = FALSE, max.overlaps = 100, nudge_y=0, )+ -->
<!--   ggtitle(paste("LM progression rate (pt~month) ~ a_snp ", paste(covariates, collapse = " + "), sep = "+"))+  -->
<!--   theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/LM_PR_snp_results.png"),lm_PR_snp_results_plot, height = hgt, width = wdth) -->


<!-- # REGRESSION FIXED LM PR ---- -->
<!-- # terms for this model formula -->
<!-- response <-  "lm_regr_fxd_PR" -->
<!-- predictors <- all_predictors %>% -->
<!--   grep("PRS", ., value = TRUE, invert = TRUE) -->
<!-- # covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile") -->
<!-- covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile", "BL_eye_age") -->
<!-- # covariates <- c("BL_pseudotime_quintile") -->


<!-- # for each predictors, try model response and save model and summary -->
<!-- lm_regr_fxd_PR_snp_mod_list <- list(); lm_regr_fxd_PR_snp_summ_list <- list(); for(a_predictor in predictors){ -->

<!--   a_form <- make.formula(response, a_predictor, covariates) -->
<!--   cat(paste("\n", a_form,"\n")) -->

<!--   mod_name <- paste0(response, " ~ ", a_predictor) -->

<!--    a_mod <- try({ -->
<!--           lmer(a_form, data = PR_mod_d) -->
<!--           # lm(a_form, data = PR_mod_d) -->
<!--         }) -->
<!--    if(inherits(a_mod, "try-error")){ -->
<!--      cat(paste0("\n\nFailure\n",response," ~ ", a_predictor, "\n")) -->
<!--      }else{ -->
<!--       cat(paste0("\n",a_form, "\n")) -->
<!--       lm_regr_fxd_PR_snp_mod_list[[mod_name]] <- a_mod -->
<!--       lm_regr_fxd_PR_snp_summ_list[[mod_name]] <- summary(a_mod) -->
<!--      } -->

<!-- } -->

<!-- # see if any models failed -->
<!-- all_mods <- paste("lm_regr_fxd_PR", predictors,sep = " ~ ") -->
<!-- setdiff(all_mods, names(lm_regr_fxd_PR_snp_mod_list)) -->

<!-- # extract and filter results -->
<!-- lm_regr_fxd_PR_snp_results <- extract_results(lm_regr_fxd_PR_snp_summ_list) -->
<!-- lm_regr_fxd_PR_snp_results <- lm_regr_fxd_PR_snp_results %>% -->
<!--   filter(!Term %in% grep("Intercept|BL_|SEXF|age", lm_regr_fxd_PR_snp_results$Term, value = TRUE)) -->

<!-- # plot results -->
<!-- lm_regr_fxd_PR_snp_results_plot <- lm_regr_fxd_PR_snp_results %>% -->
<!--   ggplot(aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   geom_hline(yintercept = -log10(.05), linetype = 3)+  geom_vline(xintercept = 0, linetype = 3)+ -->
<!--   # geom_label_repel(aes(label = ifelse(q_value < .05, Term, NA)), show.legend = FALSE, max.overlaps = 100)+ -->
<!--   geom_label_repel(aes(label = Term), show.legend = FALSE, max.overlaps = 100, nudge_y=0, )+ -->
<!--   ggtitle(paste("LM progression rate (pt~month) ~ a_snp ", paste(covariates, collapse = " + "), sep = "+"))+  -->
<!--   theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/lm_regr_fxd_PR_snp_results.png"),lm_regr_fxd_PR_snp_results_plot, height = hgt, width = wdth) -->

<!-- ``` -->

<!-- ### using ACUIT PR -->
<!-- ```{r} -->

<!-- # terms for this model formula -->
<!-- response <-  "ACUIT_PR" -->
<!-- predictors <- all_predictors %>% -->
<!--   grep("PRS", ., value = TRUE, invert = TRUE) -->
<!-- covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile", "BL_eye_age") -->
<!-- # covariates <- c("BL_pseudotime_quintile") -->


<!-- # for each predictors, try model response and save model and summary -->
<!-- ACUIT_PR_snp_mod_list <- list(); ACUIT_PR_snp_summ_list <- list(); for(a_predictor in predictors){ -->

<!--   a_form <- make.formula(response, a_predictor, covariates) -->
<!--   cat(paste("\n", a_form,"\n")) -->

<!--   mod_name <- paste0(response, " ~ ", a_predictor) -->

<!--    a_mod <- try({ -->
<!--           lmer(a_form, data = PR_mod_d) -->
<!--           # VA(a_form, data = PR_mod_d) -->
<!--         }) -->
<!--    if(inherits(a_mod, "try-error")){ -->
<!--      cat(paste0("\n\nFailure\n",response," ~ ", a_predictor, "\n")) -->
<!--      }else{ -->
<!--       cat(paste0("\n",a_form, "\n")) -->
<!--       ACUIT_PR_snp_mod_list[[mod_name]] <- a_mod -->
<!--       ACUIT_PR_snp_summ_list[[mod_name]] <- summary(a_mod) -->
<!--      } -->

<!-- } -->

<!-- # see if any models failed -->
<!-- all_mods <- paste("ACUIT_PR", predictors,sep = " ~ ") -->
<!-- setdiff(all_mods, names(ACUIT_PR_snp_mod_list)) -->

<!-- # extract and filter results -->
<!-- ACUIT_PR_snp_results <- extract_results(ACUIT_PR_snp_summ_list) -->
<!-- ACUIT_PR_snp_results <- ACUIT_PR_snp_results %>% -->
<!--   filter(!Term %in% grep("Intercept|BL_|SEXF", ACUIT_PR_snp_results$Term, value = TRUE)) -->

<!-- # plot results -->
<!-- ACUIT_PR_snp_results_plot <- ACUIT_PR_snp_results %>% -->
<!--   ggplot(aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   geom_hline(yintercept = -log10(.05), linetype = 3)+  geom_vline(xintercept = 0, linetype = 3)+ -->
<!--   # geom_label_repel(aes(label = ifelse(q_value < .05, Term, NA)), show.legend = FALSE, max.overlaps = 100, nudge_y=.05)+ -->
<!--   geom_label_repel(aes(label = Term), show.legend = FALSE, max.overlaps = 100, nudge_y=.05)+ -->
<!--   ggtitle(paste("ACUIT progression rate (ACUIT~month) ~ a_snp ", paste(covariates, collapse = " + "), sep = "+"))+  -->
<!--   theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/ACUIT_PR_snp_results.png"), ACUIT_PR_snp_results_plot, height = hgt, width = wdth) -->

<!-- ``` -->


<!-- ### using pafafovea PR -->
<!-- ```{r} -->

<!-- # terms for this model formula -->
<!-- response <-  "parafovea_PR" -->
<!-- predictors <- all_predictors %>% -->
<!--   grep("PRS", ., value = TRUE, invert = TRUE) -->
<!-- covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile", "BL_eye_age") -->
<!-- # covariates <- c("BL_pseudotime_quintile") -->


<!-- # for each predictors, try model response and save model and summary -->
<!-- parafovea_PR_snp_mod_list <- list(); parafovea_PR_snp_summ_list <- list(); for(a_predictor in predictors){ -->

<!--   a_form <- make.formula(response, a_predictor, covariates) -->
<!--   cat(paste("\n", a_form,"\n")) -->

<!--   mod_name <- paste0(response, " ~ ", a_predictor) -->

<!--    a_mod <- try({ -->
<!--           lmer(a_form, data = PR_mod_d) -->
<!--           # parafovea(a_form, data = PR_mod_d) -->
<!--         }) -->
<!--    if(inherits(a_mod, "try-error")){ -->
<!--      cat(paste0("\n\nFailure\n",response," ~ ", a_predictor, "\n")) -->
<!--      }else{ -->
<!--       cat(paste0("\n",a_form, "\n")) -->
<!--       parafovea_PR_snp_mod_list[[mod_name]] <- a_mod -->
<!--       parafovea_PR_snp_summ_list[[mod_name]] <- summary(a_mod) -->
<!--      } -->

<!-- } -->

<!-- # see if any models failed -->
<!-- all_mods <- paste("parafovea_PR", predictors,sep = " ~ ") -->
<!-- setdiff(all_mods, names(parafovea_PR_snp_mod_list)) -->

<!-- # extract and filter results -->
<!-- parafovea_PR_snp_results <- extract_results(parafovea_PR_snp_summ_list) -->
<!-- parafovea_PR_snp_results <- parafovea_PR_snp_results %>% -->
<!--   filter(!Term %in% grep("Intercept|BL_|SEXF", parafovea_PR_snp_results$Term, value = TRUE)) -->

<!-- # plot results -->
<!-- parafovea_PR_snp_results_plot <- parafovea_PR_snp_results %>% -->
<!--   # ggplot(aes(x = Estimate, y = -log10(`Pr(>|t|)`), color = Term))+ -->
<!--   ggplot(aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   geom_hline(yintercept = -log10(.05), linetype = 3)+  geom_vline(xintercept = 0, linetype = 3)+ -->
<!--   # geom_label_repel(aes(label = ifelse(`Pr(>|t|)` < .05, Term, NA)), show.legend = FALSE, max.overlaps = 100)+ -->
<!--   geom_label_repel(aes(label = ifelse(q_value < .05, Term, NA)), show.legend = FALSE, max.overlaps = 100)+ -->
<!--   # geom_label_repel(aes(label = Term), show.legend = FALSE, max.overlaps = 100)+ -->
<!--   ggtitle(paste("parafovea progression rate (parafovea~month) ~ a_snp ", paste(covariates, collapse = " + "), sep = "+"))+  -->
<!--   theme_bw() -->
<!-- ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/parafovea_PR_snp_results.png"),parafovea_PR_snp_results_plot, height = hgt, width = wdth) -->

<!-- ``` -->

#### look at difference between delta and lm models; which one to use for PR_PRS calculation?
```{r, eval = FALSE}
par(mfrow = c(1,2),mar = rep(5,4))


# look at  betas and p value relationships ----
# look at  betas relationship
bs <- cbind(
  lm_b = lm_PR_snp_results %>% arrange(Predictor) %>% pull(Estimate),
  delta_b = delta_PR_snp_results %>% arrange(Predictor) %>% pull(Estimate)
  )
plot(bs)
abline(c(0,1),lty = 2)
abline(lm(bs[,1]~ bs[,2]), col = "red")


# p value relationship
ps <- cbind(
  lm_q=
    -log10(lm_PR_snp_results$q_value[order(lm_PR_snp_results$Predictor)]),
  delta_q =
    -log10(delta_PR_snp_results$q_value[order(delta_PR_snp_results$Predictor)])
  )
plot(ps)
abline(c(0,1), lty = 2)
abline(lm(ps[,1] ~ ps[,2] ), col = "red")

# # q value relationship
# qs <- cbind(
#   lm_q=
#     -log10(lm_PR_snp_results$q_value[order(lm_PR_snp_results$Predictor)]),
#   delta_q =
#     -log10(delta_PR_snp_results$q_value[order(delta_PR_snp_results$Predictor)])
#   )
# plot(qs)
# abline(c(0,1), lty = 2)
# abline(lm(qs[,1] ~ qs[,2] ), col = "red")
par(mfrow = c(1,1))

#  look at stat correlataions ----
plot(lm_PR_snp_results$`Std. Error` , delta_PR_snp_results$`Std. Error`  ); abline(c(0,1), lty = 2)
plot(delta_PR_snp_results$`t value` , lm_PR_snp_results$`t value`  ); abline(c(0,1), lty = 2)

delta_PR_snp_results$`Std. Error` %>% mean
lm_PR_snp_results$`Std. Error` %>% mean
# ^ delta_PR mean std error less so using delta models

```


## PR ~ PRS + BL_pp_quantile

### using delta PR
```{r}

# ORIGINAL ----
# terms for this model formula
response <-  "delta_PR"
predictors <- "Normalised_PRS"
covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile")
# covariates <- c("BL_pseudotime_quintile")


# for each predictors, try model response and save model and summary
delta_PR_PRS_mod <- lmer(make.formula(response, predictors, covariates), data = PR_mod_d)

delta_PR_PRS_mod %>% summary()
  
# REG FIXED ----
# terms for this model formula
response <-  "delta_regr_fxd_PR"
predictors <- "Normalised_PRS"
covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile")
# covariates <- c("BL_pseudotime_quintile")


# for each predictors, try model response and save model and summary
delta_regr_fxd_PR_PRS_mod <- lmer(make.formula(response, predictors, covariates), data = PR_mod_d)

delta_regr_fxd_PR_PRS_mod %>% summary()
  
```

### using lm PR
```{r}

# ORIGINAL ----
# terms for this model formula
response <-  "lm_PR"
predictors <- "Normalised_PRS"
covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile")
# covariates <- c("BL_pseudotime_quintile")


# for each predictors, try model response and save model and summary
lm_PR_PRS_mod <- lmer(make.formula(response, predictors, covariates), data = PR_mod_d)

lm_PR_PRS_mod %>% summary()

# REG FIXED ----
# terms for this model formula
response <-  "lm_regr_fxd_PR"
predictors <- "Normalised_PRS"
covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile")
# covariates <- c("BL_pseudotime_quintile")


# for each predictors, try model response and save model and summary
lm_regr_fxd_PR_PRS_mod <- lmer(make.formula(response, predictors, covariates), data = PR_mod_d)

lm_regr_fxd_PR_PRS_mod %>% summary()

```

## PR ~ PR_PRS + BL_pp_quantile
```{r}


# ORIGINAL  ----
delta_PR_est <- delta_PR_snp_results %>% 
  select( Estimate) %>% 
  as.matrix()

## calculate PR_PRS
### make matrix of snp presence
PR_mod_d_at_snps <- PR_mod_d %>% 
  select(all_of(rownames(delta_PR_est))) %>% 
  as.matrix()
  
### check that names match
all(colnames(PR_mod_d_at_snps) == rownames(delta_PR_est))

### calculate dot product of snps x snp betas
PR_mod_d$PR_PRS <- PR_mod_d_at_snps %*% delta_PR_est

# model with PR_PRS
# terms for this model formula
response <-  "delta_PR"
predictors <- "PR_PRS"
covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile")

PR_PRS_mod <- lmer(make.formula(response, predictors, covariates), data = PR_mod_d)

PR_PRS_mod %>% summary()

# REGRESSION FIXED  ----
delta_regr_fxd_PR_est <- delta_regr_fxd_PR_snp_results %>% 
  select( Estimate) %>% 
  as.matrix()

## calculate PR_PRS
### make matrix of snp presence
PR_mod_d_at_snps <- PR_mod_d %>% 
  select(all_of(rownames(delta_regr_fxd_PR_est))) %>% 
  as.matrix()
  
### check that names match
all(colnames(PR_mod_d_at_snps) == rownames(delta_regr_fxd_PR_est))

### calculate dot product of snps x snp betas
PR_mod_d$regr_fxd_PR_PRS <- PR_mod_d_at_snps %*% delta_regr_fxd_PR_est

# model with PR_PRS
# terms for this model formula
response <-  "delta_regr_fxd_PR"
predictors <- "regr_fxd_PR_PRS"
covariates <- c("(1|patID)", "SEXF", "BL_pseudotime_quintile")

delta_regr_fxd_PR_PRS_mod <- lmer(make.formula(response, predictors, covariates), data = PR_mod_d)

delta_regr_fxd_PR_PRS_mod %>% summary()



```

## PRS MODELS plot

```{r}

delta_PR_PRS_mod %>% summary()
delta_regr_fxd_PR_PRS_mod %>% summary()

p.values <- c(PRS =coef(summary(delta_PR_PRS_mod))["Normalised_PRS", "Pr(>|t|)"], 
           PR_PRS =  coef(summary(PR_PRS_mod))["PR_PRS", "Pr(>|t|)"],
           regr_fxd_PR_PRS =  coef(summary(delta_regr_fxd_PR_PRS_mod))["regr_fxd_PR_PRS", "Pr(>|t|)"]
           )
betas <- c(PRS =coef(summary(delta_PR_PRS_mod))["Normalised_PRS", "Estimate"], 
           PR_PRS =  coef(summary(PR_PRS_mod))["PR_PRS", "Estimate"],
           regr_fxd_PR_PRS =  coef(summary(delta_regr_fxd_PR_PRS_mod))["regr_fxd_PR_PRS", "Estimate"]
           )

PRS_mods <- cbind(p.values, betas) %>% 
  as.data.frame()

PRS_mods$p.values <- -log10(PRS_mods$p.values)

par(mar = c(4,4,4,2))
plot(PRS_mods$p.values~PRS_mods$betas, 
     type = "n",
     xlim = c(-.5,1),
     ylim = c(0, 3),
     xlab = "average effect on progression rate",
     ylab = "-log10(p.value)", 
     main = "PRS versus effect corrected PR_PRS, and regression and effect corrected PR_PRS "
     ); add_lines("y")
abline(h = -log10(.05), lty = 3, col = "darkgrey")
abline(v = 0, lty = 3, col = "darkgrey")
text(PRS_mods$betas,PRS_mods$p.values, rownames(PRS_mods))
# arrows(x0 = PRS_mods$betas[1],
#        x1 = PRS_mods$betas[2],
#        y0 =PRS_mods$p.values[1],
#        y1 =PRS_mods$p.values[2],
#        lty = 3, 
#        # col = "#282A35"
#        col = "#5E18A3",
#        lwd= 3
# )
# abline(lm(p.values~betas,x ))
```


<!-- ## PT ~ month + SNP_i + SNP_i:month + BL_PR_Q + BL_PR_Q:month + (1|patID) + (1|patID_eye) + SEXF + AGE -->
<!-- ### make model data -->
<!-- ```{r} -->

<!-- # ORIGINAL PSEUDOPROG ---- -->
<!-- # terms to model -->
<!-- responses <- c("pseudoprog", "regr_fxd_pseudoprog") -->
<!-- predictors <-snps %>% grep("RV", ., invert = TRUE, value = TRUE) -->
<!-- covariates <-  c("patID", "patID_eye", "month", "BL_pseudotime_quintile", "SEXF", "AGE") -->

<!-- # df to model with, selecting covs, responses, and predictors; scaling select covariates?; setting quintile to numeric not factor -->
<!-- pp_interaction_mod_d <- colData(sce) %>%  -->
<!--   as.data.frame() %>%  -->
<!--   select(all_of(c( covariates, responses, predictors)))  %>%  -->
<!--   arrange(patID_eye, pseudoprog)  %>%  -->
<!--   drop_na() %>%  -->
<!--   mutate(BL_pseudotime_quintile = as.numeric(as.character(BL_pseudotime_quintile))) -->
<!--   # mutate(across(where(~is.numeric(.)), ~as.numeric(scale(.)))) -->

<!-- pp_interaction_mod_d$patID_eye %>% n_distinct() -->
<!-- # ^ 4185 -->
<!-- # ^ 4301 older -->


<!-- ``` -->

<!-- ###  look at BL_pp_Q interaction visually -->
<!-- ```{r, eval = TRUE} -->

<!-- # detach("package:igraph", unload = TRUE)  -->
<!-- library(future) -->
<!-- plan(multisession) -->

<!-- # offload this plot to it's own fork -->
<!-- BL_pp_Q_year_plot_future %<-% { -->

<!--   # make test df for interaection term -->
<!--   pp_interaction_mod_d_test <- pp_interaction_mod_d %>%  -->
<!--     mutate(year = round(month/12)) -->

<!--   # Fit the linear model including the interaction term -->
<!--   form <- paste("pseudoprog ~ factor(BL_pseudotime_quintile) * year", sep = " ") %>%  -->
<!--     as.formula() -->

<!--   simple_pp_interaction_mod <- lm( -->
<!--     form, -->
<!--     data = pp_interaction_mod_d_test -->
<!--   ) -->

<!--   # Create new data for predictions -->
<!--   newdata <- expand.grid( -->
<!--     BL_pseudotime_quintile = unique(pp_interaction_mod_d_test$BL_pseudotime_quintile), -->
<!--     year = unique(round(pp_interaction_mod_d_test$year)) -->
<!--   )   -->

<!--   # Generate predictions -->
<!--   newdata$pseudotime_pred <- predict(simple_pp_interaction_mod, newdata ) -->

<!--   # PLot -->
<!--   month_BL_pp_Q_interaction_plot <- -->
<!--     ggplot(pp_interaction_mod_d_test, aes(x = factor(BL_pseudotime_quintile),y = pseudoprog,color = factor(year) )) + -->
<!--     geom_boxplot(aes(group = interaction(BL_pseudotime_quintile, year)), alpha = 0.5, show.legend = FALSE) + -->
<!--     geom_smooth(data = newdata, -->
<!--       aes( -->
<!--         x = factor(BL_pseudotime_quintile), -->
<!--         y = round(pseudotime_pred), -->
<!--         color = factor(round(year)),   -->
<!--         fill = factor(round(year)),   -->
<!--         group = factor(round(year))    -->
<!--       ), -->
<!--       size = 1.5, -->
<!--       alpha = .3, -->
<!--       method = "lm", -->
<!--       se = TRUE, -->
<!--       show.legend = TRUE -->
<!--     ) + -->
<!--     labs( -->
<!--     title= paste0("Interaction between year and BL_pp_Q"), -->
<!--     x = "Baseline pseudoprog Quintile", -->
<!--     y = "pseudoprog", -->
<!--     color = "year" -->
<!--     ) + -->
<!--     theme_bw() + -->
<!--     scale_color_viridis_d()+  -->
<!--     scale_fill_viridis_d()  -->

<!--   ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/month_BL_pp_Q_interaction_plot.png"), -->
<!--          month_BL_pp_Q_interaction_plot, height = 10, width = 10 ) -->
<!-- } -->

<!-- ``` -->

<!-- ### plot pt ~ time with line for each level of each snp -->
<!-- ```{r} -->

<!-- pp_interaction_mod_d %>%  -->
<!--   select(regr_fxd_pseudoprog, month, all_of(predictors)) %>%  -->
<!--   pivot_longer(-c(regr_fxd_pseudoprog, month), names_to = "SNP", values_to = "level") %>%  -->
<!--   ggplot(aes(month, regr_fxd_pseudoprog, color = factor(level)))+ -->
<!--   geom_smooth(method = "lm")+ -->
<!--   facet_wrap(~SNP, ncol = 6)+ -->
<!--   theme_bw()+ -->
<!--   ggtitle("Rate of progression at different levels of SNP") -->

<!-- ``` -->

<!-- ### Model each PT by each snp with interactions, and covs -->
<!-- ```{r} -->

<!-- # ORIGINAL ---- -->
<!-- # for each predictors, try model response and save model and summary -->
<!-- response <- "pseudoprog" -->
<!-- pp_interaction_snp_mod_list <- list(); pp_interaction_snp_summ_list <- list(); for(a_predictor in predictors){ -->

<!--   a_form <- make.formula(response, a_predictor, c("BL_pseudotime_quintile*month", paste(a_predictor, "month", sep = "*"), "(1|patID)","(1|patID_eye)", " SEXF", "AGE") ) -->

<!--   cat(paste("\n******\ntrying:")) -->
<!--   cat(paste("", a_form,"")) -->

<!--   mod_name <- paste0(response, " ~ ", a_predictor) -->

<!--    a_mod <- try({ -->
<!--           lmer(a_form, data = pp_interaction_mod_d) -->
<!--           # lmer(a_form, data = pp_interaction_mod_d) %>% summary() -->
<!--         }) -->
<!--    if(inherits(a_mod, "try-error")){ -->
<!--      cat(paste0("\n\nFailure\n",response," ~ ", a_predictor, "\n")) -->
<!--      }else{ -->
<!--       cat(paste0("\n",a_form, "\n")) -->
<!--       pp_interaction_snp_mod_list[[mod_name]] <- a_mod -->
<!--       pp_interaction_snp_summ_list[[mod_name]] <- summary(a_mod) -->
<!--      } -->
<!-- } -->

<!-- # extract and filter results -->
<!-- pp_interaction_snp_results <- extract_results(pp_interaction_snp_summ_list)  -->
<!-- pp_interaction_snp_results$seq <-rep(1:8,  length(predictors))  -->
<!-- pp_interaction_snp_results <- pp_interaction_snp_results %>%  -->
<!--   filter(seq == 8) -->


<!--   # filter(!Term %in% grep("Intercept|BL_pseudo|SEXF|month", pp_interaction_snp_results$Term, value = TRUE)) -->
<!--   # filter(!Term %in% grep("Intercept|BL_pseudo|SEXF|month", pp_interaction_snp_results$Term, value = TRUE)) -->
<!-- pp_interaction_snp_results %>%  -->
<!--   arrange(q_value) %>%  -->
<!--   select(Estimate, `Pr(>|t|)`, q_value) -->

<!-- # plot results -->
<!-- pp_interaction_snp_results %>%  -->
<!--   tidyr::separate_wider_delim(Term,":", names = c("Term", "interaction")) %>%  # get rid of :month on snp names -->
<!--   mutate(Term =  -->
<!--            case_when(Term == "rs1260326_T_2_GKCR" ~  "rs1260326_T_2_GCKR", # fix typo -->
<!--                      TRUE ~ Term )) %>%  -->
<!--   ggplot(aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   geom_hline(yintercept = -log10(.05), linetype = 3)+  geom_vline(xintercept = 0, linetype = 3)+ -->
<!--   geom_label_repel(aes(label = ifelse(q_value < .05, Term, NA)), show.legend = FALSE, max.overlaps = 100)+ -->
<!--   ggtitle(paste("E.G.:",as.character(a_form), collapse = " "))+  -->
<!--   theme_bw() -->

<!-- ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/pp_interaction_snp_results.png"), height = hgt, width = wdth) -->


<!-- # REGRESSION FIXED PP ---- -->
<!-- # for each predictors, try model response and save model and summary -->
<!-- response <- "regr_fxd_pseudoprog" -->
<!-- regr_fxd_interaction_snp_mod_list <- list(); regr_fxd_interaction_snp_summ_list <- list(); for(a_predictor in predictors){ -->

<!--   a_form <- make.formula(response, a_predictor, c("BL_pseudotime_quintile*month", paste(a_predictor, "month", sep = "*"), "(1|patID)","(1|patID_eye)", " SEXF", "AGE") ) -->

<!--   cat(paste("\n******\ntrying:")) -->
<!--   cat(paste("", a_form,"")) -->

<!--   mod_name <- paste0(response, " ~ ", a_predictor) -->

<!--    a_mod <- try({ -->
<!--           lmer(a_form, data = pp_interaction_mod_d) -->
<!--           # lmer(a_form, data = pp_interaction_mod_d) %>% summary() -->
<!--         }) -->
<!--    if(inherits(a_mod, "try-error")){ -->
<!--      cat(paste0("\n\nFailure\n",response," ~ ", a_predictor, "\n")) -->
<!--      }else{ -->
<!--       cat(paste0("\n",a_form, "\n")) -->
<!--       regr_fxd_interaction_snp_mod_list[[mod_name]] <- a_mod -->
<!--       regr_fxd_interaction_snp_summ_list[[mod_name]] <- summary(a_mod) -->
<!--      } -->
<!-- } -->

<!-- # extract and filter results -->
<!-- regr_fxd_interaction_snp_results <- extract_results(regr_fxd_interaction_snp_summ_list)  -->
<!-- regr_fxd_interaction_snp_results$seq <-rep(1:8,  length(predictors))  -->
<!-- regr_fxd_interaction_snp_results <- regr_fxd_interaction_snp_results %>%  -->
<!--   filter(seq == 8) -->


<!--   # filter(!Term %in% grep("Intercept|BL_pseudo|SEXF|month", regr_fxd_interaction_snp_results$Term, value = TRUE)) -->
<!--   # filter(!Term %in% grep("Intercept|BL_pseudo|SEXF|month", regr_fxd_interaction_snp_results$Term, value = TRUE)) -->
<!-- regr_fxd_interaction_snp_results %>%  -->
<!--   arrange(q_value) %>%  -->
<!--   select(Estimate, `Pr(>|t|)`, q_value) -->

<!-- # plot results -->
<!-- regr_fxd_interaction_snp_results %>%  -->
<!--   tidyr::separate_wider_delim(Term,":", names = c("Term", "interaction")) %>%  # get rid of :month on snp names -->
<!--   mutate(Term =  -->
<!--            case_when(Term == "rs1260326_T_2_GKCR" ~  "rs1260326_T_2_GCKR", # fix typo -->
<!--                      TRUE ~ Term )) %>%  -->
<!--   ggplot(aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   geom_hline(yintercept = -log10(.05), linetype = 3)+  geom_vline(xintercept = 0, linetype = 3)+ -->
<!--   geom_label_repel(aes(label = ifelse(q_value < .05, Term, NA)), show.legend = FALSE, max.overlaps = 100)+ -->
<!--   ggtitle(paste("E.G.:",as.character(a_form), collapse = " "))+  -->
<!--   theme_bw() -->

<!-- ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/regr_fxd_interaction_snp_results.png"), height = hgt, width = wdth) -->


<!-- ``` -->

<!-- ### Model each PT by each snp with interactions, and covs, stratifying by quitile -->
<!-- #### MODEL---- -->
<!-- ```{r} -->

<!-- # do for each quintile, plot each snp effect at each quintile (hm snp x quintile) -->
<!-- at_quintile_pp_interaction_snp_summ_list_list <- list();  -->
<!-- at_quintile_pp_interaction_snp_mod_list_list <- list(); -->
<!-- for(a_quintile in sort(unique(pp_interaction_mod_d$BL_pseudotime_quintile) ) ) { -->

<!--   # for each predictors, try model response and save model and summary -->
<!--   at_quintile_pp_interaction_snp_mod_list <- list(); -->
<!--   at_quintile_pp_interaction_snp_summ_list <- list();  -->
<!--   for(a_predictor in predictors){ -->

<!--     a_form <- make.formula("pseudoprog", a_predictor,  -->
<!--                            c("month", paste(a_predictor, "month", sep = "*"), "(1|patID)","(1|patID_eye)", " SEXF", "AGE") ) -->
<!--     cat(paste0("\n*****\ntrying:\n")) -->
<!--     cat(paste(a_form)) -->

<!--     mod_name <- paste0(response, " ~ ", a_predictor) -->

<!--      a_mod <- try({ -->
<!--             lmer(a_form, data = pp_interaction_mod_d %>% filter(BL_pseudotime_quintile == a_quintile), REML = FALSE) -->
<!--           }) -->
<!--      if(inherits(a_mod, "try-error")){ -->
<!--        cat(paste0("\n\nFailure\n",response," ~ ", a_predictor, "\n")) -->
<!--        }else{ -->
<!--         cat(paste0("\n",a_form, "\n")) -->
<!--         at_quintile_pp_interaction_snp_mod_list[[mod_name]] <- a_mod -->
<!--         at_quintile_pp_interaction_snp_summ_list[[mod_name]] <- summary(a_mod) -->
<!--        } -->
<!--   } -->

<!--   at_quintile_pp_interaction_snp_summ_list_list[[a_quintile]] <- at_quintile_pp_interaction_snp_summ_list -->
<!--   at_quintile_pp_interaction_snp_mod_list_list[[a_quintile]] <- at_quintile_pp_interaction_snp_mod_list -->

<!-- } -->



<!-- ``` -->


<!-- #### EVALUATE RESULTS ---- -->
<!-- ```{r, eval = FALSE} -->

<!-- ## WHY DO SOME HAVE IS.SINGULAR WARNING  -->
<!-- ## LOOK AT HOW SKEW, KURTOSIS, AND SD INFLUENCE MODEL ---- -->
<!-- singular_mods <- sapply(at_quintile_pp_interaction_snp_mod_list_list, function(a.list){ -->
<!--        sapply(a.list, isSingular) -->
<!-- }) -->
<!-- colnames(singular_mods) <- paste0("Q",as.character(1:5)) -->

<!-- # hm of singular models  -->
<!-- singular_mods_hm <- Heatmap((singular_mods * 1) , -->
<!--         cluster_rows = FALSE,  -->
<!--         cluster_columns = FALSE,  -->
<!--         col= c("white", "red"), -->
<!--         column_title = "Models that have singular fit" -->
<!--         ) -->
<!-- ## look at skew of each snp and pseudoprog at each quintile -->
<!-- pp_snp_skews <- sapply(sort(unique(pp_interaction_mod_d$BL_pseudotime_quintile)), function(Q){ -->

<!--   pp_skew = skew(pp_interaction_mod_d$pseudoprog[pp_interaction_mod_d$BL_pseudotime_quintile == Q]) -->

<!--   snp_skews = sapply(pp_interaction_mod_d[pp_interaction_mod_d$BL_pseudotime_quintile == Q , predictors], skew) -->

<!--   # return vector of skews -->
<!--   # skews <- c(snp_skews, pp_skew = pp_skew) -->
<!--   skews <- snp_skews -->

<!--   # # return vector of snp_skews scaled by |PT|_skew -->
<!--   # skews <-snp_skews * abs(pp_skew) -->

<!--   return(skews) -->
<!-- }) -->

<!-- colnames(pp_snp_skews) <- paste0("Q", as.character(1:5)) -->

<!-- ### look at pattern of skews -->
<!-- skew_hm <- Heatmap(pp_snp_skews,  -->
<!--         clustering_method_rows = "ward.D2", -->
<!--         clustering_method_columns = "ward.D2", -->
<!--         cluster_rows = FALSE, -->
<!--         cluster_columns = FALSE,  -->
<!--         column_title = "skews at each quintile", -->
<!--         # col = c("white", "red") -->
<!-- ); draw(skew_hm) -->

<!-- ## look at kurtosi of each snp and pseudoprog at each quintile -->
<!-- pp_snp_kurtosis <- sapply(sort(unique(pp_interaction_mod_d$BL_pseudotime_quintile)), function(Q){ -->

<!--   pp_kurtosi = kurtosi(pp_interaction_mod_d$pseudoprog[pp_interaction_mod_d$BL_pseudotime_quintile == Q]) -->

<!--   snp_kurtosis = sapply(pp_interaction_mod_d[pp_interaction_mod_d$BL_pseudotime_quintile == Q , predictors], kurtosi) -->

<!--   # return vector of kurtosis -->
<!--   # kurtosis <- c(snp_kurtosis, pp_kurtosi = pp_kurtosi) -->
<!--   kurtosis <- snp_kurtosis -->

<!--   # # return vector of snp_kurtosis scaled by |PT|_kurtosi -->
<!--   # kurtosis <-snp_kurtosis * abs(pp_kurtosi) -->

<!--   return(kurtosis) -->
<!-- }) -->

<!-- colnames(pp_snp_kurtosis) <- paste0("Q", as.character(1:5)) -->

<!-- ### look at pattern of kurtosis -->
<!-- kurtosi_hm <- Heatmap(pp_snp_kurtosis,  -->
<!--         clustering_method_rows = "ward.D2", -->
<!--         clustering_method_columns = "ward.D2", -->
<!--         cluster_rows = FALSE, -->
<!--         cluster_columns = FALSE,  -->
<!--         column_title = "kurtosis at each quintile", -->
<!--         # col = c("white", "red") -->
<!-- );draw(kurtosi_hm) -->


<!-- ## look at sd of each  -->
<!-- pp_snp_sds <- sapply(sort(unique(pp_interaction_mod_d$BL_pseudotime_quintile)), function(Q){ -->

<!--   pp_sd = sd(pp_interaction_mod_d$pseudoprog[pp_interaction_mod_d$BL_pseudotime_quintile == Q]) -->

<!--   snp_sds = sapply(pp_interaction_mod_d[pp_interaction_mod_d$BL_pseudotime_quintile == Q , predictors], sd) -->

<!--   # return vector of sds -->
<!--   # sds <- c(snp_sds, pp_sd = pp_sd) -->
<!--   sds <- snp_sds -->

<!--   # # return vector of snp_sds scaled by |PT|_sd -->
<!--   # sds <-snp_sds * abs(pp_sd) -->

<!--   return(sds) -->
<!-- }) -->

<!-- colnames(pp_snp_sds) <- paste0("Q", as.character(1:5)) -->

<!-- ### look at pattern of sds -->
<!-- sd_hm <- Heatmap(pp_snp_sds ,  -->
<!--         clustering_method_rows = "ward.D2", -->
<!--         clustering_method_columns = "ward.D2", -->
<!--         cluster_rows = FALSE, -->
<!--         cluster_columns = FALSE,  -->
<!--         column_title = "sds at each quintile",  -->
<!--         col = c("white", "red") -->
<!-- ) -->

<!-- ## for snps and pt, at each quintile, count records of eyes with >1 visit where var !is.na -->
<!-- pp_snp_Ns <- sapply(sort(unique(pp_interaction_mod_d$BL_pseudotime_quintile)), function(Q){ -->

<!--   # pseudoprog N when Q = Q -->
<!--   pp_N = pp_interaction_mod_d %>%  -->
<!--     filter(BL_pseudotime_quintile == Q) %>%  -->
<!--     filter(n()>1, .by = patID_eye) %>%  -->
<!--     filter(!is.na(pseudoprog)) %>%  -->
<!--     nrow() -->

<!--   snp_Ns = sapply(predictors, function(a_snp){ -->

<!--     pp_interaction_mod_d %>%  -->
<!--       filter(BL_pseudotime_quintile== Q) %>%  -->
<!--       filter(n()>1, .by = patID_eye) %>%  -->
<!--       filter(!is.na(all_of(a_snp))) %>%  -->
<!--       nrow() -->

<!--   }) -->

<!--   # return vector of Ns -->
<!--   # Ns <- c(snp_Ns, pp_N = pp_N) -->
<!--   Ns <- snp_Ns -->

<!--   # # return vector of snp_Ns scaled by |PT|_N -->
<!--   # Ns <-snp_Ns * abs(pp_N) -->

<!--   return(Ns) -->
<!-- }) -->
<!-- colnames(pp_snp_Ns) <- paste0("Q", as.character(1:5)) -->

<!-- N_hm <- Heatmap(pp_snp_Ns %>%log10(),  -->
<!--         clustering_method_rows = "ward.D2", -->
<!--         clustering_method_columns = "ward.D2", -->
<!--         cluster_rows = FALSE, -->
<!--         cluster_columns = FALSE,  -->
<!--         column_title = "count !is.na of var for eyes with >1 visit at each quintile",  -->
<!--         col = c("white", "red") -->
<!-- ) -->
<!-- # draw(N_hm) -->

<!-- # view them together -->
<!-- singular_mods_grob <- grid.grabExpr(draw(singular_mods_hm, newpage = FALSE)) -->
<!-- skew_hm_grob <- grid.grabExpr(draw(skew_hm, newpage = FALSE)) -->
<!-- sd_hm_grob <- grid.grabExpr(draw(sd_hm, newpage = FALSE)) -->
<!-- kurtosi_hm_grob <- grid.grabExpr(draw(kurtosi_hm, newpage = FALSE)) -->
<!-- N_hm_grob <- grid.grabExpr(draw(N_hm, newpage = FALSE)) -->
<!-- grid.arrange(singular_mods_grob, skew_hm_grob, kurtosi_hm_grob, sd_hm_grob, N_hm_grob, ncol = 3) -->
<!-- ``` -->


<!-- #### EXTRACT RESULTS AND PLOT ---- -->
<!-- ```{r} -->

<!-- ## extract and add quintile variable and then rbind all together -->
<!-- at_quintile_pp_interaction_snp_results <- lapply(1:length(at_quintile_pp_interaction_snp_summ_list_list), function(i){ # for each summary list -->
<!--   # extract results as data.frame -->
<!--   raw_summ_list <- at_quintile_pp_interaction_snp_summ_list_list[[i]] -->
<!--   x <- extract_results(raw_summ_list) -->

<!--   # add var for what quintile stratified for  -->
<!--   x$quintile_filter <- i -->

<!--   return(x) -->

<!-- }) %>%  -->
<!--   Reduce(rbind,.) -->


<!-- # filter results; every 5th row, num_snps * ith quintiles times -->
<!-- ith = 6 -->
<!-- at_quintile_pp_interaction_snp_results$seq <- rep(1:ith,  length(predictors) ) -->
<!-- at_quintile_pp_interaction_snp_results <- at_quintile_pp_interaction_snp_results %>%  -->
<!--   filter(seq == ith) -->

<!-- # recalculate Q value to weight based on all results together -->
<!-- at_quintile_pp_interaction_snp_results <- at_quintile_pp_interaction_snp_results %>% -->
<!--   mutate(q_value = p.adjust(`Pr(>|t|)`)) -->

<!--   # filter(!Term %in% grep("Intercept|BL_pseudo|SEXF|month", pp_interaction_snp_results$Term, value = TRUE)) -->
<!--   # filter(!Term %in% grep("Intercept|BL_pseudo|SEXF|month", pp_interaction_snp_results$Term, value = TRUE)) -->

<!-- # Volcano ---- -->
<!-- at_quintile_pp_interaction_snp_results %>%  -->
<!--   ggplot(aes(x = Estimate, y = -log10(q_value), color = Term))+ -->
<!--   geom_hline(yintercept = -log10(.05), linetype = 3)+  geom_vline(xintercept = 0, linetype = 3)+ -->
<!--   geom_point(show.legend = FALSE)+ -->
<!--   geom_label_repel(aes(label = ifelse(q_value < .05, Term, NA)), nudge_y = .5, show.legend = FALSE, max.overlaps = 100)+ -->
<!--   ggtitle(paste("Modeling pseudoprog at BL_pseudotime_quintile\nE.G.:",paste(as.character(a_form)[c(2,1,3)], collapse = " ")))+  -->
<!--   facet_wrap(~quintile_filter)+ -->
<!--   theme_bw() -->

<!-- ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/at_quintile_pp_interaction_snp_results.png"), -->
<!--        height = 10, width = 15) -->


<!-- my_transform <- function(vec){ -->

<!--   # # sqrt tranformation -->
<!--   # vec = sign(vec) * sqrt(abs(vec)) -->

<!--   # # 3rd root tranformation -->
<!--   # vec= sign(vec) * abs(vec)^(1/3) -->

<!--     # log tranformation -->
<!--   vec = sign(vec) * log10(abs(vec)) -->
<!-- } -->

<!-- # look at how estimate changes for each snp at different quintiles -->
<!-- estimate_at_different_quintiles_plot <- at_quintile_pp_interaction_snp_results %>%  -->
<!--   ggplot(aes(quintile_filter, my_transform(Estimate),color = Term))+ # transforming estimate for easier viewing -->
<!--   # ggplot(aes(quintile_filter, Estimate,color = Term))+ -->
<!--   geom_line(show.legend = FALSE)+ -->
<!--   geom_label_repel(aes(label = Term, size = -log10(q_value)), show.legend = TRUE)+ -->
<!--   geom_point(show.legend = FALSE, )+ -->
<!--   geom_hline(yintercept = 0, linetype = 3)+ -->
<!--   ggtitle("SNP estimates at different quintiles")  -->

<!-- ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/at_quintile_pp_interaction_snp_estimates_at_different_quintiles.png"), -->
<!--        estimate_at_different_quintiles_plot, -->
<!--        height = 10, width = 20) -->


<!-- # Heatmap of results---- -->
<!-- # Heatmap of Snp x quintile filled wiht estimates -->
<!-- estimate_at_different_quintiles_m <- at_quintile_pp_interaction_snp_results %>%  -->
<!--   select(Predictor, Estimate, quintile_filter) %>%  -->
<!--   pivot_wider(id_cols = Predictor, names_from = quintile_filter, values_from = Estimate) %>%  -->
<!--   column_to_rownames("Predictor") %>%  -->
<!--   as.matrix() -->

<!-- set.seed(303908) -->
<!-- hm <- Heatmap(estimate_at_different_quintiles_m,  -->
<!--               cluster_columns = FALSE,  -->
<!--               cluster_rows = TRUE,  -->
<!--               clustering_method_rows = "ward.D2",  -->
<!--               name = "Estimate",  -->
<!--               column_title =  -->
<!--                 paste( -->
<!--                       "PT ~ SNP + interactions progression model estimates stratifying by quintile\nE.g.: ", -->
<!--                       paste(as.character(a_form)[c(2,3)], collapse =" ~ "), -->
<!--                       "\n(., *, **, and ***, = qval < .1, .05, .01, and .001, respectively)\n" -->
<!--                       ) -->
<!--               ) -->
<!-- draw(hm) -->

<!-- # for each row column, fill cell with estimate where df var = row/col name of matrix at i,j -->
<!-- Q_at_different_quintiles_m <- estimate_at_different_quintiles_m -->
<!-- for(i in 1:nrow(Q_at_different_quintiles_m)){ -->
<!--   for(j in 1:ncol(Q_at_different_quintiles_m)){ -->
<!--     # Q value where colnames matches results quintile filter and where rownames matches results predictor -->
<!--     Q_at_different_quintiles_m[i,j] <-  -->
<!--       at_quintile_pp_interaction_snp_results$q_value[ -->
<!--           at_quintile_pp_interaction_snp_results$Predictor == rownames(Q_at_different_quintiles_m)[i] &  -->
<!--           at_quintile_pp_interaction_snp_results$quintile_filter == colnames(Q_at_different_quintiles_m)[j]  -->
<!--         ] -->
<!--   } -->
<!-- } -->

<!-- # matrix of asterisks of significance -->
<!-- sig_m <- Q_at_different_quintiles_m %>% -->
<!--   apply(., 2, as.character) -->
<!-- rownames(sig_m) <- rownames(Q_at_different_quintiles_m) -->

<!-- # fill matrix with asterisks -->
<!-- for(i in 1:nrow(sig_m)){ -->
<!--   for(j in 1:ncol(sig_m)){ -->
<!--     p <- Q_at_different_quintiles_m[i,j] -->
<!--     sig_m[i,j] <- ifelse(p<.001, "***",  -->
<!--                          ifelse(p<.01, "**",  -->
<!--                                 ifelse(p<.05, "*",  -->
<!--                                        ifelse(p<.1, ".", NA_character_)))) -->
<!--   } -->
<!-- } -->

<!-- # reorder sig m to be that of hm -->
<!-- row_order_names <- rownames(sig_m)[row_order(hm)] -->
<!-- col_order_names <- colnames(sig_m)[column_order(hm)] -->
<!-- sig_m <- sig_m[row_order_names,col_order_names] -->

<!-- # plot -->
<!-- png(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/at_quintile_pp_interaction_snp_estimates_and_at_different_quintiles_hm.png"), height = 1000, width = 870) -->
<!-- par(mar = c(10, 3,3,3)) -->

<!-- # Add asterisks for significance -->
<!-- draw(hm) -->

<!-- decorate_heatmap_body("Estimate", {  # Referencing the correct heatmap name -->
<!--     # Loop over the matrix to overlay asterisks based on significance -->
<!--     for(i in 1:nrow(estimate_at_different_quintiles_m)) { -->
<!--         for(j in 1:ncol(estimate_at_different_quintiles_m)) { -->

<!--           significance = sig_m[i,j] -->
<!--           # Overlay the significance level at the center of the cell -->
<!--           if (!is.na(significance)) { -->
<!--             grid.text(significance, -->
<!--               x = j/ncol(estimate_at_different_quintiles_m) - .1, -->
<!--               y = (nrow(estimate_at_different_quintiles_m) - i + 1)/nrow(estimate_at_different_quintiles_m) - .025, -->
<!--               # x = j/ncol(estimate_at_different_quintiles_m), -->
<!--               # y = (nrow(estimate_at_different_quintiles_m) - i + 1)/nrow(estimate_at_different_quintiles_m) , -->
<!--               gp = gpar(fontsize = 14, col = "black")) -->
<!--             } -->
<!--           } -->
<!--         } -->
<!--     }) -->

<!-- par(mar = rep(2,4)) -->
<!-- dev.off() -->


<!-- ``` -->

## DELTA_PP/DELATA_T ~ lag_PP + SNP
```{r}

as.data.frame(colData(sce)) %>% 
  mutate(across(matches("_RV"), ~ ifelse(!is.na(Normalised_PRS) & is.na(.), 0, .))) %>% 
  filter(n()>1, .by = patID_eye) %>% 
  select(matches("_RV")) %>% 
  distinct()
# ^^ patients with rarevariant data are not longitudinal

# MAKE MODELING DATAFRAME----
# df of each eye's cluster and pseudoprog and real time; filter for longitudinal eyes; make year; fix regressing pseudoprogression
regr_fxd_pp_delta_t_snp_mod_d <- colData(sce) %>%
  as.data.frame() %>% 
  select(sample, patID, patID_eye,month, regr_fxd_pseudoprog, Normalised_PRS, SEXF, AGE, all_of(snps)) %>% 
  filter(!is.na(Normalised_PRS)) %>% 
# impute Rare variants to 0 where genetics taken >>(hasty assumption)
  mutate(across(matches("_RV"), ~ ifelse(!is.na(Normalised_PRS) & is.na(.), 0, .))) %>% 
  arrange(sample) %>% 
  # filter for >1 visit
  filter(n()>1, .by = patID_eye) %>% 
  mutate(year = month / 12) %>% 
  mutate(regr_fxd_pseudoprog = as.numeric(scale(regr_fxd_pseudoprog))) %>% 
  ungroup()

# ****** FUNCTION: reorder cols by moving select ones to end ******
put_cols_at_end <-function(df, cols_to_move = character()){
  reorder_vec <- names(df)[!names(df)%in%cols_to_move]
  df <- df[,c(reorder_vec,cols_to_move)]
  return(df)
} 

regr_fxd_pp_delta_t_snp_mod_d <- put_cols_at_end(regr_fxd_pp_delta_t_snp_mod_d,snps)

# MAKE LAGS

# ****** FUNCTION: calculate combination of each x with every next x for each eye ******
# a_eye= regr_fxd_pp_delta_t_snp_mod_d$patID_eye[1]
process_eye <- function(a_eye) {
    # Filter for eye of interest
    a_eye_diff_d <- regr_fxd_pp_delta_t_snp_mod_d[regr_fxd_pp_delta_t_snp_mod_d$patID_eye == a_eye, ]
    
    # Create year to every other year combinations
    year_combos <- combn(a_eye_diff_d$year, 2) %>% t() %>% as.data.frame()
    colnames(year_combos) <- c("year", "post_year")
    
    # Create pp to every other pp combinations
    pseudoprog_combos <- combn(a_eye_diff_d$regr_fxd_pseudoprog, 2) %>% t() %>% as.data.frame()
    colnames(pseudoprog_combos) <- c("regr_fxd_pseudoprog", "post_pseudoprog")
    
    # Time pp combinations
    t_pp_combos <- cbind(year_combos, pseudoprog_combos)
    
    # Join back to dataset
    a_eye_diff_d <- suppressMessages(
      a_eye_diff_d %>% 
        left_join(t_pp_combos, relationship = "one-to-many")
    )
    
    # Calculate differences
    a_eye_diff_d$lag_diff_year <- a_eye_diff_d$post_year - a_eye_diff_d$year
    a_eye_diff_d$lag_diff_pp <- a_eye_diff_d$post_pseudoprog - a_eye_diff_d$regr_fxd_pseudoprog
    
    # calculate rate
    a_eye_diff_d$regr_fxd_pseudoprog_rate <- a_eye_diff_d$lag_diff_pp / a_eye_diff_d$lag_diff_year 
    
    # reorder cols
    a_eye_diff_d <- put_cols_at_end(a_eye_diff_d,snps)
    
    return(a_eye_diff_d)
  }
  
# process each eye in parallel
unique_eyes <- unique(regr_fxd_pp_delta_t_snp_mod_d$patID_eye)

eye_diff_list <- mclapply(unique_eyes, process_eye, mc.cores = detectCores() - 1)

# Combine all results into a single data frame
regr_fxd_pp_delta_t_snp_mod_d <- do.call(rbind, eye_diff_list)

# # filter for only consecutive chanages 0>1, 1>2, not 0>2...
# regr_fxd_pp_delta_t_snp_mod_d <- regr_fxd_pp_delta_t_snp_mod_d %>% 
#   group_by(patID_eye, year) %>% 
#   filter((post_year - year) == min((post_year - year))) %>% # for each year, which record has the minimimu difference between current and post year
#   as.data.frame()

####  ####   ####   ####   
# model ----
####   ####   ####   ####   

response = "regr_fxd_pseudoprog_rate"
predictors <- grep("^rs\\d|^X\\d|Normalised_PRS|_RV", names(colData(sce)), value =TRUE)
# predictors = c(predictors, "Normalised_PRS")

# for each predictors, try model response and save model and summary
lag_pp_rate_snp_mod_list <- list(); lag_pp_rate_snp_summ_list <- list(); for(a_predictor in predictors){
  
  # a_predictor <- paste0(a_predictor, "*regr_fxd_pseudoprog")
  
  # a_form <- make.formula(response, a_predictor, c("regr_fxd_pseudoprog","(1|patID)","(1|patID_eye)", "SEXF","AGE"))
  a_form <- make.formula(response, a_predictor, c("regr_fxd_pseudoprog","(1|patID_eye)", "SEXF","AGE"))
  # a_form <- make.formula(response, a_predictor)
  
  cat(paste("\n******\ntrying:"))
  cat(paste("", a_form,""))
  
  # name of model to name in list
  mod_name <- paste0(response, " ~ ", a_predictor)
  
  # try modeling
  a_mod <- try({
          lmer(a_form, data = regr_fxd_pp_delta_t_snp_mod_d)
          # lm(a_form, data = regr_fxd_pp_delta_t_snp_mod_d)
        })
   if(inherits(a_mod, "try-error")){
     cat(paste0("\n\nFailure\n",response," ~ ", a_predictor, "\n"))
     }else{
      cat(paste0("\n",a_form, "\n"))
      lag_pp_rate_snp_mod_list[[mod_name]] <- a_mod
      lag_pp_rate_snp_summ_list[[mod_name]] <- summary(a_mod)
     }
}
lag_pp_rate_snp_summ_list[21]

msg <- sapply(lag_pp_rate_snp_summ_list, function(summ) summ$fitMsgs)
msgs[sapply(msgs, length)>0]

####  ####   ####   ####   
# extract and filter results ----
####   ####   ####   ####   
lag_pp_rate_snp_results <- extract_results(lag_pp_rate_snp_summ_list) %>% 
  filter(!grepl("Int|AGE|SEXF|regr",Term))
# lag_pp_rate_snp_results$seq <- 1:6
# lag_pp_rate_snp_results <- lag_pp_rate_snp_results %>% 
#   filter(seq ==6)


# plot
eg_form <- paste(a_form[c(2,1,3)], collapse = "")
lag_pp_rate_snp_results %>% 
  ggplot(aes(Estimate,-log10(q_value), color = Term))+
  geom_hline(yintercept = -log10(.05), linetype = 3)+
  geom_vline(xintercept = 0, linetype = 3)+
  geom_point(show.legend = FALSE)+
  geom_label_repel(aes(label= Term, color = Term), show.legend = FALSE)+
  ggtitle(eg_form)+
  theme_bw()+
  geom_blank()
ggsave(paste0(wd_path, "results/clustering/14_geneticProgressionModeling/lag_pp_rate_snp_results.pdf"), height = 10, width = 12)

```


####
# Save data
```{r}

# save hd5 array
saveHDF5SummarizedExperiment(sce,paste0(wd_path, "processed_data/sce007.h5") )

notify.me("14_genetics modeling done")

```

